---
title: "Bivariate Description: One Categorical (Nominal or Ordinal) Variable and One Interval Variable"
tutorial:
  id: "08-Bivariate-Description-Categorical-Interval-Data"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to describe the relationship between a categorical and an interval level variable.
---

## Overview {#main-content}

[**Bivariate description**]{.important-text} is the practice of examining the relationship between two variables at a time --- asking whether the distribution of one variable differs systematically across categories or values of another. Tutorial 7 introduced the tools for describing the relationship between two categorical variables: cross-tabulations and grouped bar plots. This tutorial extends that logic to a second combination: one categorical variable and one interval variable. As in Tutorial 7, each section starts with a research question and a hypothesis, then selects the appropriate tool to evaluate whether the data are consistent with it. The specific question here is how an interval outcome varies across the categories of a categorical explanatory variable. The main tools are `group_by()` and `summarise()` from [dplyr]{.package-name} to compare group means, and histograms and box plots broken down by group from [ggplot2]{.package-name} to compare the full distribution across groups.

In this tutorial, you will learn how to:

* Describe the relationship between a categorical and an interval variable by comparing means across groups using `group_by()` and `summarise()`
* Use histograms broken down by group with `facet_wrap()` from [ggplot2]{.package-name} to compare the full distribution of an interval variable across categories
* Use box plots with a categorical x-axis using `geom_boxplot()` from [ggplot2]{.package-name} to compare groups efficiently


```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "QPATutorialsCourse")
if (max(counties$mortality_risk_25_45_change, na.rm = TRUE) < 2)
  counties$mortality_risk_25_45_change <- counties$mortality_risk_25_45_change * 100
if (max(counties$employ_pop_ratio_25_64, na.rm = TRUE) < 10)
  counties$employ_pop_ratio_25_64 <- counties$employ_pop_ratio_25_64 * 100
if (max(counties$wage_growth, na.rm = TRUE) < 2)
  counties$wage_growth <- counties$wage_growth * 100
counties$dem2p_percent <- counties$dem2p_vote_share * 100
counties$rural <- 1
counties$rural[counties$rural_urban <= 3] <- 0
counties$rural[counties$rural_urban >= 8] <- 2
data("world", package = "poliscidata")
```


## How Does an Interval Outcome Vary Across Groups?

Many questions in political science ask whether some measurable quantity differs across political or social categories. Do counties with a Black racial majority cast a higher Democratic vote percentage than White-majority counties? Are full democracies wealthier on average than autocracies? Do militarized interstate disputes last longer when one side has nuclear weapons? In each case, the question has the same structure: take an interval variable --- Democratic vote percentage, GDP per capita, conflict duration --- and ask whether its distribution differs across categories of a second variable --- racial majority, regime type, nuclear status. This is the second type of [**bivariate description**]{.important-text} you will learn, and it applies across every subfield of political science.

There are two ways to examine this kind of relationship. The first is to compare the **mean** of the interval variable across the categories of the explanatory variable. If democracies are wealthier than autocracies, the mean GDP per capita should be higher in the democratic category. The mean gives a concise, single-number summary of the average value in each group. The second is to compare the full **distribution** of the interval variable across groups using histograms or box plots. A mean can hide important features --- two groups can have similar means but very different spreads, or one group's mean can be pulled by a small number of extreme outliers. Looking at the full distribution shows what the mean alone cannot.

Both approaches require setting up the relationship in the right direction: the interval variable is the outcome and the categorical variable defines the groups. The key is identifying which variable is the outcome. For example, asking whether mean wage growth differs between Trump-majority and Clinton-majority counties fits this tutorial because Trump-majority status is the categorical explanatory variable defining the groups and wage growth is the interval outcome being compared across them. A different question --- does wage growth predict the *probability* that a county supported Trump? --- has an interval explanatory variable and a binary categorical outcome, which requires a different tool. The distinction is not about which variables are involved but about which one is the outcome: when the outcome is interval and the explanatory variable is categorical, compare group means. When the outcome is categorical, a different approach is needed.

## Is College Attainment Lower in Rural Counties?

Consider the following hypothesis: 

::: {.hypothesis}
In a comparison of [counties], those that are [more rural] will have a [lower share of college graduates] than will those that are [more urban].
:::

The theoretical explanation is that rural areas tend to have fewer local jobs requiring a college degree and less access to higher education, so college graduates may be more likely to live in or move to urban areas.

The data frame **counties** contains the ordinal variable **rural**, which takes the value 0 for urban counties, 1 for counties that are neither distinctly urban nor rural, and 2 for rural counties. It also contains the interval variable **prop_college_grad**, the proportion of a county's population that has completed a college degree. Our hypothesis leads us to expect that the mean proportion of college graduates should be lowest in rural counties and highest in urban counties.

### Calculate the Means

You have used `group_by()` and `summarise()` from [dplyr]{.package-name} since Tutorial 5, where you used them to count the number of cases in each category. The same two-step structure works for any summary statistic --- here, in addition to `n()` to count, you also use `mean()` to calculate the group mean within each group. The logic is identical: `group_by()` defines the groups and `summarise()` computes statistics within each one. You will use this same structure throughout this tutorial.

The run-and-observe below calculates the mean proportion of college graduates separately for urban, neither-urban-nor-rural, and rural counties. `freq = n()` counts the number of counties in each group, and `mean = mean(prop_college_grad, na.rm = TRUE)` calculates the mean within each group. Filtering both variables before grouping ensures `freq = n()` accurately reflects the number of cases contributing to the mean --- if either variable has missing values, those cases are removed before `n()` counts the groups, so `freq` and the mean are computed on exactly the same set of observations. Load [dplyr]{.package-name} at the top with `library(dplyr)` --- once loaded, it is available for all subsequent exercises in this tutorial.

**Run and observe: Calculate mean proportion of college graduates by rural**

```{r frequency-collegegrad-by-rural, exercise=TRUE, exercise.cap="Run and observe: Mean college graduation rate by rurality"}
library(dplyr)
counties %>%
  filter(!is.na(rural), !is.na(prop_college_grad)) %>%
  group_by(rural) %>%
  summarise(freq = n(), mean = mean(prop_college_grad, na.rm = TRUE))
```

This output looks similar to the frequency tables from Tutorial 5 --- the rows name the categories of **rural**, the **freq** column counts the number of counties in each group just as it did before, and the **mean** column gives the mean proportion of college graduates within each group. The only new piece is that final column.

### Is the Pattern Consistent with the Hypothesis?

Compare the three group means to see whether the pattern matches what the hypothesis predicts.

**Practice: Evaluate the rural/college-graduate hypothesis**

Compare the three group means, then decide whether the pattern matches the hypothesis and whether it is a smooth decline or something more nuanced.

```{r quiz-collegegrad-rural-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells us about the hypothesis?",
  correct = "Right. The pattern is consistent with the hypothesis, but with nuance.",
  answer("The pattern is consistent with the hypothesis, but mainly because urban counties stand out --- the neither-urban-nor-rural and rural categories are close to each other",
         correct = TRUE,
         message = "Urban counties have a mean proportion of college graduates of 0.255 --- well above neither-urban-nor-rural (0.181) and rural (0.177) counties, which are nearly identical to each other. The hypothesis predicts a smooth decline across all three categories, but the data show a large urban/non-urban gap with very little difference between the two non-urban categories."),
  answer("The pattern is consistent with the hypothesis, with the mean proportion of college graduates dropping by roughly the same amount at each step from urban to neither-urban-nor-rural to rural",
         message = "The order is correct --- urban is highest, rural is lowest --- but the size of the steps is not equal. The drop from urban (0.255) to neither-urban-nor-rural (0.181) is large, while the drop from neither-urban-nor-rural (0.181) to rural (0.177) is small. Most of the difference comes from urban counties standing apart, not from an even decline."),
  answer("The hypothesis is confirmed --- the difference in means proves that rurality causes lower educational attainment",
         message = "Descriptive comparison of means cannot confirm a hypothesis. Finding a pattern in the predicted direction tells us the data are consistent with the hypothesis, but a formal hypothesis test is needed to assess statistical significance."),
  answer("The pattern is not consistent with the hypothesis --- neither-urban-nor-rural and rural counties have nearly identical means, which contradicts the predicted decline across all three categories",
         message = "A hypothesis about direction does not require a perfectly smooth decline across every category. This hypothesis predicts that more rural counties will have lower college graduation rates than more urban counties --- and they do. Urban counties have a mean of 0.255, while both non-urban categories are well below that at around 0.18. The fact that neither-urban-nor-rural and rural counties are close to each other is a nuance worth noting, but it does not make the overall pattern inconsistent with the hypothesis."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

## Is the Democratic Vote Percentage Related to a County's Racial Majority?

Racial composition has been one of the most consistent factors associated with vote choice in recent American elections. In particular, Black Americans have historically voted for Democratic candidates over Republican candidates by wide margins. Was this true in the 2016 presidential election? To answer this question, we examine the following hypothesis:

::: {.hypothesis}
In a comparison of [counties], those with a [Black racial majority] will have a [higher mean Democratic vote percentage] than will those with a [White racial majority].
:::

The theoretical explanation is that the Democratic Party's policy positions on civil rights and voting rights have made it the party more closely associated with addressing the political concerns of Black voters, while the Republican Party has drawn disproportionate support from White voters over the same period.

The data frame **counties** contains the nominal variable **racial_majority**, which takes the values "White Majority," "Black Majority," "Hispanic Majority," and "Other/No Majority," depending on which racial group makes up a majority of the county's population. It is stored as a character variable. It also contains the interval variable **dem2p_percent**, the Democratic candidate's share of the two-party vote in the county (as a percent).

### Calculate the Means

To evaluate the hypothesis, we need the mean Democratic vote percentage for each racial majority category. Even though the hypothesis only compares Black and White Majority counties, we calculate the mean for all four categories at once --- that is how `group_by()` always works. Filter out missing values in both variables before grouping --- this ensures `freq = n()` and `mean()` operate on exactly the same set of observations.

**Practice: Compare mean Democratic vote percentage by racial majority**

Complete the code below using the same `group_by()`/`summarise()` structure as the previous example, adapted for these variables. Filter missing values in both variables before grouping. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice: Mean Democratic vote percentage by racial majority"}
counties %>%
  filter(!is.na(___), !is.na(dem2p_percent)) %>%
  ___(___) %>%
  ___(freq = ___(), mean = mean(___, na.rm = TRUE))
```

```{r exercise-dem2p-by-racialmajority-hint-1}
# Hint 1 of 2
# Use the same group_by() and summarise() structure as
# the college graduation example. Filter missing values
# in both variables before grouping. Which variable
# defines the groups? Which is the outcome?
```

```{r exercise-dem2p-by-racialmajority-hint-2}
# Hint 2 of 2
# group_by() takes the explanatory variable.
# Inside summarise(), use n() to count cases and name it freq,
# and mean() with na.rm = TRUE to compute the group mean
# and name it mean.
```

```{r exercise-dem2p-by-racialmajority-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing %>% between steps. Inside summarise(), you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the =.")
  }
  if (!grepl("\\bcounties\\b", code_no_comments)) {
    fail("Use the counties data frame.")
  }
  if (!grepl("filter\\s*\\(\\s*!\\s*is\\.na\\s*\\(\\s*racial_majority\\s*\\)", code_no_comments)) {
    fail("Filter out missing racial majority data before grouping: filter(!is.na(racial_majority), !is.na(dem2p_percent)).")
  }
  if (!grepl("is\\.na\\s*\\(\\s*dem2p_percent\\s*\\)", code_no_comments)) {
    fail("Also filter out missing dem2p_percent values: filter(!is.na(racial_majority), !is.na(dem2p_percent)). This ensures freq = n() and mean() operate on the same set of observations.")
  }
  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use group_by() to group counties by racial_majority.")
  }
  if (!grepl("group_by\\s*\\(\\s*racial_majority\\s*\\)", code_no_comments)) {
    fail("Group by racial_majority --- that is the variable that defines each county's racial majority category.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use summarise() to calculate the mean within each group.")
  }
  if (!grepl("freq\\s*=\\s*n\\s*\\(\\s*\\)", code_no_comments)) {
    fail("Name the count variable freq using freq = n() inside summarise().")
  }
  if (!grepl("mean\\s*=\\s*mean\\s*\\(\\s*dem2p_percent\\s*,\\s*na\\.rm\\s*=\\s*TRUE\\s*\\)", code_no_comments)) {
    fail("Name the mean variable mean and compute it from dem2p_percent: mean = mean(dem2p_percent, na.rm = TRUE).")
  }

  expected <- counties %>%
    filter(!is.na(racial_majority), !is.na(dem2p_percent)) %>%
    group_by(racial_majority) %>%
    summarise(freq = n(), mean = mean(dem2p_percent, na.rm = TRUE), .groups = "drop")

  result <- as.data.frame(.result)
  expected <- as.data.frame(expected)

  if (!all(c("racial_majority", "freq", "mean") %in% names(result))) {
    fail("Your table should include racial_majority, freq, and mean.")
  }

  result <- result[, c("racial_majority", "freq", "mean")]
  result <- result[order(result$racial_majority), ]
  expected <- expected[order(expected$racial_majority), ]

  if (!isTRUE(all.equal(result, expected, tolerance = 1e-6, check.attributes = FALSE))) {
    fail("Check that you filtered missing values, grouped by racial_majority, and calculated freq = n() and mean = mean(dem2p_percent, na.rm = TRUE).")
  }

  pass("Your table is correct. Examine it carefully before answering the next question.")
})
```

### Is the Pattern Consistent with the Hypothesis?

The hypothesis predicts that Black Majority counties will have a higher mean Democratic vote percentage than White Majority counties. Your table lets you check that directly.

**Practice: Evaluate the racial-majority hypothesis**

Compare the mean for Black Majority counties to the mean for White Majority counties in the table you just produced.

```{r quiz-racialmajority-dem2p-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells us about the hypothesis?",
  correct = "Right. Black Majority counties have a substantially higher mean Democratic vote percentage than White Majority counties.",
  answer("The pattern is consistent with the hypothesis --- Black Majority counties have a higher mean Democratic vote percentage than White Majority counties",
         correct = TRUE,
         message = "Black Majority counties have a mean Democratic vote percentage of 65.21% compared to 31.34% for White Majority counties --- a difference of more than thirty percentage points, consistent with the hypothesis. The comparison of means alone cannot tell us whether the difference is larger than chance or what caused it."),
  answer("The pattern is not consistent with the hypothesis --- White Majority counties have a higher mean Democratic vote percentage than Black Majority counties",
         message = "Look at the table again. Black Majority counties (65.21%) have a higher mean Democratic vote percentage than White Majority counties (31.34%), not the reverse."),
  answer("The hypothesis is confirmed --- the difference in means proves racial composition determines vote choice",
         message = "Descriptive comparison of means cannot confirm a hypothesis. It can only show whether the pattern is consistent with it."),
  answer("The hypothesis cannot be evaluated because racial_majority has more than two categories",
         message = "Comparing means of an interval variable across more than two categories of a nominal variable is exactly the right descriptive method here --- the hypothesis simply focuses on two of those categories."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```


## Is Democracy Associated with Economic Development?

Democratic institutions are often credited with producing better economic outcomes, through stronger property rights, more accountable governance, and policies that respond to broad public demand rather than to a narrow elite. Do more democratic countries actually have higher GDP per capita? To answer this question, we examine the following hypothesis:

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have [higher levels of economic development] than will those that are [less democratic].
:::

The theoretical explanation is that democratic accountability gives leaders an incentive to pursue broadly beneficial economic policy, since they can be removed from office if the economy performs poorly, while leaders in non-democratic systems face fewer such constraints.

The data frame **world** contains the interval variable **gdppcap08**, GDP per capita in 2008 measured in US dollars. We use GDP per capita rather than total GDP because it accounts for country size --- China and the United States have large total GDPs partly because of their large populations, not necessarily because their citizens are more prosperous. Dividing by population gives a better sense of average living standards. The ordinal variable **dem_level4** records regime type on a four-category scale: Full Democracy, Partial Democracy, Hybrid, and Authoritarian. It is stored as a factor.

### Calculate the Means

To evaluate the hypothesis we need the mean GDP per capita for each regime type. As in the previous examples, filter out missing values in both the grouping variable and the outcome variable before grouping --- this ensures `freq = n()` and `mean()` operate on exactly the same set of observations.

**Apply: Calculate mean GDP by democracy level**

Write the full pipeline from scratch using **gdppcap08** and **dem_level4** in the **world** data frame. Filter missing values in both variables, name the counts `freq` and the means `mean`. Try not to look at the hints. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-gdp-by-democracy-level, exercise=TRUE, exercise.cap="Apply: Mean GDP per capita by democracy level"}

```

```{r exercise-gdp-by-democracy-level-hint-1}
# Hint 1 of 2
# Start by removing cases with missing values in both
# variables you are working with --- the grouping variable
# and the outcome variable.
# Then use the same group_by() and summarise() structure
# as the previous exercises.
```








```{r exercise-gdp-by-democracy-level-hint-2}
# Hint 2 of 2
# Use the world data frame.
# filter(!is.na(dem_level4), !is.na(gdppcap08)) removes missing values in both variables.
# group_by(dem_level4) defines the groups.
# Inside summarise(), use n() to count and name it freq,
# and mean(gdppcap08, na.rm = TRUE) to get the mean
# and name it mean.
```

```{r exercise-gdp-by-democracy-level-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing %>% between steps. Inside summarise(), you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the =.")
  }
  if (!grepl("\\bworld\\b", code_no_comments)) {
    fail("Use the world data frame.")
  }
  if (!grepl("filter\\s*\\(\\s*!\\s*is\\.na\\s*\\(\\s*dem_level4\\s*\\)", code_no_comments)) {
    fail("Filter out missing regime type data before grouping: filter(!is.na(dem_level4), !is.na(gdppcap08)).")
  }
  if (!grepl("is\\.na\\s*\\(\\s*gdppcap08\\s*\\)", code_no_comments)) {
    fail("Also filter out missing gdppcap08 values: filter(!is.na(dem_level4), !is.na(gdppcap08)). This ensures freq = n() and mean() operate on the same set of observations.")
  }
  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use group_by() to group countries by dem_level4.")
  }
  if (!grepl("group_by\\s*\\(\\s*dem_level4\\s*\\)", code_no_comments)) {
    fail("Group by dem_level4 --- that is the variable that defines regime type.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use summarise() to calculate the mean within each group.")
  }
  if (!grepl("freq\\s*=\\s*n\\s*\\(\\s*\\)", code_no_comments)) {
    fail("Name the count variable freq using freq = n() inside summarise().")
  }
  if (!grepl("mean\\s*=\\s*mean\\s*\\(\\s*gdppcap08\\s*,\\s*na\\.rm\\s*=\\s*TRUE\\s*\\)", code_no_comments)) {
    fail("Name the mean variable mean and compute it from gdppcap08: mean = mean(gdppcap08, na.rm = TRUE).")
  }

  expected <- world %>%
    filter(!is.na(dem_level4), !is.na(gdppcap08)) %>%
    group_by(dem_level4) %>%
    summarise(freq = n(), mean = mean(gdppcap08, na.rm = TRUE), .groups = "drop")

  result <- as.data.frame(.result)
  expected <- as.data.frame(expected)

  if (!all(c("dem_level4", "freq", "mean") %in% names(result))) {
    fail("Your table should include dem_level4, freq, and mean.")
  }

  result <- result[, c("dem_level4", "freq", "mean")]
  result <- result[order(result$dem_level4), ]
  expected <- expected[order(expected$dem_level4), ]

  if (!isTRUE(all.equal(result, expected, tolerance = 1e-6, check.attributes = FALSE))) {
    fail("Check that you filtered missing values, grouped by dem_level4, and calculated freq = n() and mean = mean(gdppcap08, na.rm = TRUE).")
  }

  pass("Your table is correct. Examine the means carefully before answering the next question.")
})
```

### Is the Pattern Consistent with the Hypothesis?

**Apply: Evaluate the democracy/GDP hypothesis**

You have now seen two examples of evaluating a hypothesis using a table of means. Apply the same logic here --- look at the four group means you just produced and decide whether the pattern matches what the hypothesis predicts.

```{r quiz-gdp-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells us about the hypothesis?",
  correct = "Right. The pattern is partially consistent with the hypothesis.",
  answer("The pattern is fully consistent with the hypothesis --- mean GDP per capita increases consistently from authoritarian to full democracy",
         message = "Look at the table carefully. Full democracies do have the highest mean GDP per capita, but the ordering across the other three regime types does not follow the predicted pattern."),
  answer("The pattern is partially consistent with the hypothesis --- full and partial democracies have higher mean GDP than authoritarian regimes, but the ordering across the other regime types does not follow a clean democratic hierarchy",
         correct = TRUE,
         message = "Full democracies have the highest mean GDP per capita and hybrid regimes have the lowest, which is not the predicted order. Some authoritarian states --- particularly those with natural resource wealth --- push the authoritarian mean above hybrid democracies, complicating the picture. The pattern is partially consistent with the hypothesis but not cleanly."),
  answer("The pattern is not consistent with the hypothesis --- authoritarian regimes outperform at least one democratic regime type",
         message = "Partially correct, but the full picture is more nuanced. Full and partial democracies do have higher mean GDP per capita than authoritarian regimes on average. The complication is only with hybrid regimes, which have the lowest mean GDP per capita."),
  answer("GDP per capita cannot be compared across regime types because countries differ in size",
         message = "GDP per capita is specifically designed to account for country size by dividing total GDP by population. Comparing it across groups is entirely appropriate."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```


## Why Aren't Means Enough?

A table of means tells us whether groups differ on average, but it can hide a great deal. Two groups can have similar means even though one group is much more spread out than the other. A group's mean can also be pulled away from its typical value by a small number of extreme outliers. Two groups can even have the same mean while having very different shapes --- for example, one distribution might be symmetric while another is skewed. None of this is visible in a table of means alone.

To see these features, we use the same two tools introduced in Tutorial 6 for visualizing a single interval variable --- histograms and box plots --- but now broken down by categories of the explanatory variable, so that we can compare the full distribution across groups rather than just the average.

## What Does the Full Distribution Look Like Across Groups?

A table of means tells us whether groups differ on average, but it cannot show us whether that difference reflects a shift in the whole distribution or just a few unusual counties. A faceted histogram lets us see the full picture for each group.

The only new step beyond a single-variable histogram is showing a separate panel for each category of the explanatory variable. We do this in two ways: mapping the categorical variable to `fill`, so each group's bars are colored differently, and adding `facet_wrap()`, which places each group's histogram in its own panel. This technique is called [**faceting**]{.important-text} --- each panel is called a [**facet**]{.important-text}, and `facet_wrap()` is the function that creates them.

You have used the `fill` aesthetic before in two ways: as a fixed color set outside `aes()` (like `fill = "firebrick"` in Tutorial 5), and as a variable mapping inside `aes()` to color bars by group (like `fill = racial_majority` in Tutorial 7). Here it serves the same purpose as in Tutorial 7 --- coloring each panel's bars by group so the panels are visually distinct at a glance.

When `fill` maps a variable to color, ggplot automatically adds a legend identifying which color corresponds to which category. In a faceted histogram, that legend is redundant --- the panel titles already identify which group each histogram shows. Adding `guides(fill = "none")` removes the legend and keeps the plot clean. The general rule: whenever the fill variable is already identified elsewhere in the plot --- whether by panel titles, x-axis labels, or a `scale_fill_discrete()` call --- use `guides(fill = "none")` to remove the legend.

Use **dem2p_percent** and **racial_majority** from the **counties** data frame. Because **racial_majority** is stored as a character variable, ggplot treats it as discrete automatically and orders its categories alphabetically --- no `factor()` wrapper is needed. The tilde (`~`) before the variable name in `facet_wrap()` tells R which variable defines the panels --- the same syntax you will use in every `facet_wrap()` call. You do not need `scale_x_discrete()` here because the panel titles come directly from the values of **racial_majority**, which are already readable text. You will need `scale_x_discrete()` when the stored values are not self-explanatory --- for example, numeric codes like 0 and 1.

Load [ggplot2]{.package-name} with `library(ggplot2)` before the first plot. Like [dplyr]{.package-name}, it only needs to be loaded once --- place both calls at the top of your script or Quarto document. Note that the `%>%` pipe comes from [dplyr]{.package-name} --- always load both packages when piping into `ggplot()`.

**Practice: Compare the distribution of Democratic vote percentage across racial majority categories**

The starter code shows the full structure of a faceted histogram --- the filter, the geom, the labels, and the guides are all provided. Fill in the two blanks: one controls which variable colors the bars, and the other is the function that creates one panel per category. Two hints are available. Click "Submit Answer" to check your work.

```{r histogram-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice: Faceted histogram of Democratic vote percentage by racial majority", fig.alt="Faceted histograms showing the distribution of Democratic vote percentage on the x-axis and frequency on the y-axis, with separate panels for Black Majority, Hispanic Majority, Other/No Majority, and White Majority counties. The White Majority panel has far more counties and is centered around 25 percent, while the other three panels have far fewer counties and are centered at higher vote percentages, consistent with the difference in means."}
library(ggplot2)
counties %>%
  filter(!is.na(racial_majority), !is.na(dem2p_percent)) %>%
  ggplot(mapping = aes(x = dem2p_percent, ___ = racial_majority)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Distribution of Democratic Vote Percentage by Racial Majority",
       x = "Percent of Two-Party Vote",
  guides(fill = "none")
```

```{r histogram-dem2p-by-racialmajority-hint-1}
# Hint 1 of 2
# One blank goes inside aes() --- it controls which
# variable colors the bars.
# The other blank is a function name that creates
# one panel per category of a variable.
# The filter() step is already provided --- keep it.
```

```{r histogram-dem2p-by-racialmajority-hint-2}
# Hint 2 of 2
# fill = racial_majority
# facet_wrap(...)
```

```{r histogram-dem2p-by-racialmajority-check}
grade_this({
  code_lines <- gsub("#.*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing + at the end of a line, unmatched parentheses, a label string missing its quotation marks, or a missing %>% in the pipeline.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*racial_majority\\s*\\)", code_no_comments)) {
    fail("Keep !is.na(racial_majority) inside filter() so missing values are removed before plotting.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*dem2p_percent\\s*\\)", code_no_comments)) {
    fail("Also keep !is.na(dem2p_percent) inside filter() so missing values in both variables are removed before plotting.")
  }
  if (!grepl("counties", code_no_comments)) {
    fail("Use the counties data frame.")
  }
  if (!grepl("x\\s*=\\s*dem2p_percent", code_no_comments)) {
    fail("Map dem2p_percent to the x-axis inside aes().")
  }
  if (!grepl("geom_histogram\\s*\\(", code_no_comments)) {
    fail("Use geom_histogram() to create the histogram.")
  }
  if (!grepl("fill\\s*=\\s*racial_majority", code_no_comments)) {
    fail("Map racial_majority to fill inside aes().")
  }
  if (!grepl("facet_wrap\\s*\\(", code_no_comments)) {
    fail("Add a facet_wrap() layer to give each category its own panel.")
  }
  if (!grepl("facet_wrap\\s*\\(\\s*~\\s*racial_majority", code_no_comments)) {
    fail("Inside facet_wrap(), precede racial_majority with a tilde (~).")
  }
  if (!grepl("guides\\s*\\(\\s*fill\\s*=\\s*['\"]none['\"]", code_no_comments)) {
    fail('Add guides(fill = "none") to remove the redundant legend --- the panel titles already identify each group.')
  }

  pass("Your plot is correct.")
})
```

The shared y-axis makes it easy to compare the number of counties across groups, but the Black Majority, Hispanic Majority, and Other/No Majority panels are so small relative to White Majority that their distributions are almost invisible. To see the shape of each group's distribution clearly, use `scales = "free_y"` inside `facet_wrap()`. This gives each panel its own y-axis scale so the bars fill the panel regardless of how many counties are in that group. The trade-off is that you can no longer compare bar heights across panels --- but you can now see the center and shape of each distribution.

**Run and observe: Use scales = "free\_y" to compare distributions**

```{r histogram-dem2p-racialmajority-freey, exercise=TRUE, exercise.cap="Run and observe: Free y-axis scale for better shape comparison"}
counties %>%
  filter(!is.na(racial_majority), !is.na(dem2p_percent)) %>%
  ggplot(mapping = aes(x = dem2p_percent, fill = racial_majority)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Distribution of Democratic Vote Percentage by Racial Majority",
       x = "Percent of Two-Party Vote",
       y = "Frequency",
       caption = "Source: Linn, Nagler, Zilinsky") +
  facet_wrap(~racial_majority, scales = "free_y") +
  guides(fill = "none")
```

Now each panel shows the shape of its own distribution clearly. When group sizes are roughly similar, the shared y-axis is usually the better choice --- it lets you compare both shape and group size in a single plot. Use `scales = "free_y"` when groups differ so dramatically in size that the smaller panels become unreadable at the shared scale, as happened here. In that situation, report group sizes separately --- the table of means already includes the count in each group, so nothing is lost.

### What Does the Histogram Add?

The free-axis histogram confirms that the difference in means is not driven by a few unusual counties --- the entire distribution is shifted across groups. White Majority counties form a concentrated distribution around 25-30%, while Black Majority counties are centered near 60-70%. The means are reliable summaries here. The Hispanic Majority and Other/No Majority panels are harder to read because those groups contain so few counties (96 and 48 respectively) that their distributions are lumpy and difficult to characterize --- this is a limitation of small group sizes, not a feature of the underlying distribution.

## Does the Whole Distribution of College Graduation Rates Differ by Rurality?

Earlier we found that urban counties have notably higher mean college graduation rates than neither-urban-nor-rural or rural counties. A faceted histogram lets us check whether that difference reflects a shift in the whole distribution or something more subtle --- a difference in spread, shape, or just a few outliers pulling the mean.

Because **rural** is stored as numbers (0, 1, 2), the panel titles in `facet_wrap()` would show those codes rather than meaningful labels. The `as_labeller()` function solves this: it takes a named vector that maps each stored value to a readable label --- `c("stored_value" = "readable label", ...)` --- and you pass it to the `labeller` argument inside `facet_wrap()`: `facet_wrap(~rural, labeller = as_labeller(c(...)))`.

**Practice: Compare the distribution of college graduation rates by rural**

Use **prop_college_grad** and **rural** from the **counties** data frame. Filter out missing values before plotting. Use `as_labeller()` inside `facet_wrap()` to replace the numeric codes with readable panel titles. Four hints are available. Click "Submit Answer" to check your work.

Your plot should have:

- Title: "Distribution of College Graduation Rates by Rurality"
- No subtitle --- omit it from `labs()` entirely
- x-axis label: "Proportion of College Graduates"
- y-axis label: "Frequency"
- Caption: "Source: Linn, Nagler, Zilinsky"

```{r histogram-collegegrad-by-rural, exercise=TRUE, exercise.cap="Practice: Faceted histogram of college graduation rates by rurality", fig.alt="Faceted histograms showing the distribution of the proportion of college graduates on the x-axis and frequency on the y-axis, with separate panels labeled Urban, Neither Urban nor Rural, and Rural. The urban panel is widest with the longest right tail, the neither-urban-nor-rural panel is the tallest and narrowest, and the rural panel falls in between."}

```

```{r histogram-collegegrad-by-rural-hint-1}
# Hint 1 of 4
# This is the same structure as the racial_majority histogram:
# filter both variables, pipe into ggplot(), map the interval
# variable to x and the grouping variable to fill, add
# geom_histogram() and facet_wrap(). Because rural is stored
# as a number, one extra step is needed when mapping to fill.
```

```{r histogram-collegegrad-by-rural-hint-2}
# Hint 2 of 4
# Numeric variables need factor() when mapped to fill so
# ggplot treats them as discrete categories rather than
# a continuous scale. facet_wrap() does not need factor() ---
# use labeller = as_labeller() there instead to give panels
# readable titles.
```

```{r histogram-collegegrad-by-rural-hint-3}
# Hint 3 of 4
# as_labeller() takes a named vector mapping stored values
# to readable labels. The three values of rural and their
# labels are given in the prose above. Use guides(fill = "none")
# to remove the redundant legend.
```

```{r histogram-collegegrad-by-rural-hint-4}
# Hint 4 of 4
# Check that your filter removes missing values in both
# variables, your aes() has x = prop_college_grad and
# fill = factor(rural), and your facet_wrap() includes
# the labeller argument with all three labels.
```

```{r histogram-collegegrad-by-rural-check}
grade_this({
  code_lines <- gsub("#.*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing + at the end of a line, unmatched parentheses, or a label string missing its quotation marks.")
  }

  # Check data frame
  if (!any(grepl("\\bcounties\\b", .user_code))) {
    fail("Use the counties data frame.")
  }

  # Check filter for both variables — require !is.na() not is.na()
  code_no_comments_lines <- paste(gsub("#.*", "", .user_code), collapse = "\n")
  has_filter_grad <- grepl(
    "!\\s*is\\.na\\s*\\(\\s*prop_college_grad\\s*\\)",
    code_no_comments_lines
  )
  has_filter_rural <- grepl(
    "!\\s*is\\.na\\s*\\(\\s*rural\\s*\\)",
    code_no_comments_lines
  )
  if (!has_filter_grad || !has_filter_rural) {
    fail("Filter out missing values in both variables before plotting: filter(!is.na(prop_college_grad), !is.na(rural)).")
  }

  # Check aesthetics from plot object
  get_aes_name <- function(plot, aes_name) {
    out <- character(0)
    global_mapping <- plot$mapping[[aes_name]]
    if (!is.null(global_mapping)) {
      out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(global_mapping)), error = function(e) NA_character_))
    }
    for (layer in plot$layers) {
      layer_mapping <- layer$mapping[[aes_name]]
      if (!is.null(layer_mapping)) {
        out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(layer_mapping)), error = function(e) NA_character_))
      }
    }
    out[!is.na(out)]
  }

  x_vars <- get_aes_name(.result, "x")
  if (!"prop_college_grad" %in% x_vars) {
    fail("Map prop_college_grad to the x-axis inside aes().")
  }

  # Check fill requires factor(rural) — not bare rural
  fill_code <- any(grepl("fill\\s*=\\s*factor\\s*\\(\\s*rural\\s*\\)", code_no_comments))
  if (!fill_code) {
    fail("Because rural is stored as a number, wrap it in factor() when mapping it to fill: fill = factor(rural). Without factor(), ggplot treats rural as continuous and applies a color gradient rather than discrete group colors.")
  }

  # Check layer types
  layer_geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  if (!"GeomBar" %in% layer_geoms) {
    fail("Use geom_histogram() to create the histograms.")
  }

  # Check faceting uses rural
  if (!any(grepl("facet_wrap", .user_code))) {
    fail("Add facet_wrap() to give each rurality category its own panel.")
  }
  if (!any(grepl("facet_wrap\\s*\\(\\s*~\\s*(factor\\s*\\(\\s*)?rural", code_no_comments))) {
    fail("Inside facet_wrap(), use ~rural (or ~factor(rural)) so each panel corresponds to a rurality category.")
  }
  if (!any(grepl("as_labeller", .user_code))) {
    fail("Use as_labeller() inside facet_wrap() to give panels readable titles.")
  }

  # Check labels (accept single or double quotes)
  code_sq <- gsub('"', "'", paste(.user_code, collapse = "\n"))
  code_all <- paste(.user_code, collapse = "\n")
  if (!grepl('"0"\\s*=\\s*"Urban"|\'0\'\\s*=\\s*\'Urban\'', code_all)) {
    fail('Map "0" to "Urban" inside as_labeller().')
  }
  if (!grepl('"1"\\s*=\\s*"Neither Urban nor Rural"|\'1\'\\s*=\\s*\'Neither Urban nor Rural\'', code_all)) {
    fail('Map "1" to "Neither Urban nor Rural" inside as_labeller().')
  }
  if (!grepl('"2"\\s*=\\s*"Rural"|\'2\'\\s*=\\s*\'Rural\'', code_all)) {
    fail('Map "2" to "Rural" inside as_labeller().')
  }
  if (!grepl('guides\\s*\\(\\s*fill\\s*=\\s*["\']none["\']', code_all)) {
    fail('Add guides(fill = "none") to remove the redundant legend.')
  }

  # Check plot labels from object
  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Distribution of College Graduation Rates by Rurality")) {
    fail('Set title = "Distribution of College Graduation Rates by Rurality" inside labs().')
  }
  if (!identical(plot_labels$x, "Proportion of College Graduates")) {
    fail('Set x = "Proportion of College Graduates" inside labs().')
  }
  if (!identical(plot_labels$y, "Frequency")) {
    fail('Set y = "Frequency" inside labs().')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set caption = "Source: Linn, Nagler, Zilinsky" inside labs().')
  }

  pass("Your plot is correct.")
})
```

### What Does the Histogram Add?

**Practice: Interpret the rural histogram**

The table of means already showed that urban counties have higher mean college graduation rates. Look at what the histogram adds --- focus on shape and spread, not just center.

```{r quiz-histogram-collegegrad-rural, echo=FALSE}
question("The table of means told us urban counties have higher average college graduation rates. What does the histogram reveal that the means could not?",
  correct = "Right. The histogram shows the urban advantage is about spread as well as center.",
  answer("Urban counties not only have a higher average graduation rate but a much wider spread --- including a long right tail of counties with very high graduation rates --- while neither-urban-nor-rural counties have the most tightly concentrated distribution",
         correct = TRUE,
         message = "The means told you urban counties are higher on average. The histogram adds that urban counties are also far more variable --- the long right tail means some urban counties have extremely high graduation rates that pull the mean up, while the neither-urban-nor-rural panel is remarkably narrow, suggesting those counties are very consistent with each other. The rural panel falls in between on spread."),
  answer("The histogram confirms the means exactly --- urban is highest, rural is lowest, and neither-urban-nor-rural falls in between with no additional information",
         message = "The histogram does confirm the ordering from the means, but it also shows something the means could not: how spread out each distribution is. The shape and width of each panel reveal whether the difference is about a uniformly higher center or about a wider range of outcomes."),
  answer("Neither-urban-nor-rural and rural have nearly identical shapes, suggesting the means overstated the difference between them",
         message = "Look closely at the width of each panel. Neither-urban-nor-rural is noticeably taller and narrower than rural --- the two distributions are not nearly identical in shape even though their means are close."),
  answer("The histogram is less informative than the means table because it does not show exact values",
         message = "The histogram shows something the means table cannot: the shape and spread of each distribution. Knowing that urban counties have a long right tail, for example, tells you something about which counties are driving the urban mean upward that no table of means can reveal."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

## How Do Box Plots Compare Groups Efficiently?

A box plot summarizes a distribution using the median, the 25th and 75th percentiles, and whiskers that extend to the most extreme values that are not flagged as outliers. The box itself spans the interquartile range (IQR) --- the middle 50% of the data --- with a line marking the median. Unusually extreme values are shown separately as outlier points. You built single box plots in Tutorial 6. Box plots are especially useful for comparing groups, because several boxes placed side by side let you compare medians, spreads, and outliers across categories at a glance, in a way that is harder to do with overlapping histograms.

Consider a new research question to motivate the use of box plots. Democratic governance is often linked not just to economic outcomes but to public health and well-being. Healthier populations may reflect stronger public institutions --- better sanitation, more reliable healthcare access, and governments more accountable for the basic welfare of their citizens. Does life expectancy actually differ by level of democracy? To answer this question, we examine the following hypothesis:

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have a [higher life expectancy] than will those that are [less democratic].
:::

The theoretical explanation is that democratic accountability creates pressure on governments to invest in public goods like healthcare and sanitation, since leaders who fail to deliver these benefits can be removed from office, while leaders in less democratic systems face less pressure to do so.

The data frame **world** contains the interval variable **lifeex_total**, life expectancy at birth, and the ordinal variable **dem_level4**, the same regime-type variable we used to compare GDP per capita. Because **dem_level4** is already stored as a factor, you do not need to wrap it in `factor()` when mapping it inside `aes()`.

In a grouped box plot, `x` takes the categorical variable that defines the groups --- one box per category --- and `y` takes the interval variable whose distribution you are examining. Map the same categorical variable to `fill` to color each box differently. Since the x-axis already labels each box, `guides(fill = "none")` removes the redundant legend.

The stored values of **dem_level4** are abbreviated: `"Full Democ"`, `"Part Democ"`, `"Hybrid"`, `"Authoritarian"`. Use `scale_x_discrete()` after `labs()` to replace them with readable labels. When you supply labels without names, ggplot assigns them positionally --- first label to first factor level, second to second. Check the level order with `levels(world$dem_level4)` before writing the labels.

**Practice: Compare the distribution of life expectancy by democracy level**

Filter missing values in both variables before plotting, fill in the three blanks in `aes()`, and add the two layers after `labs()`. Three hints are available. Click "Submit Answer" to check your work.


```{r boxplot-lifeex-by-democracylevel, exercise=TRUE, exercise.cap="Practice: Box plots of life expectancy by democracy level", fig.alt="Box plots showing life expectancy at birth on the y-axis and democracy level on the x-axis, with separate boxes for Full Democracy, Partial Democracy, Hybrid, and Authoritarian regimes."}
world %>%
  filter(!is.na(___), !is.na(___)) %>%
  ggplot(mapping = aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot() +
  labs(title = "Distribution of Life Expectancy by Level of Democracy",
       x = "Level of Democracy",
       y = "Life Expectancy at Birth",
       caption = "Source: World Bank") +
```







```{r boxplot-lifeex-by-democracylevel-hint-1}
# Hint 1 of 3
# Fill in the two filter() blanks with the two variables
# you are working with --- both need to be non-missing.
# Inside aes(), which variable defines the categories?
# That goes on x and fill. Which variable's distribution
# are you examining? That goes on y.
```


```{r boxplot-lifeex-by-democracylevel-hint-2}
# Hint 2 of 3
# x = dem_level4, y = lifeex_total, fill = dem_level4.
# After labs(), add scale_x_discrete() with the four
# category labels in the order shown by levels(world$dem_level4):
# "Full Democracy", "Partial Democracy", "Hybrid", "Authoritarian".
```

```{r boxplot-lifeex-by-democracylevel-hint-3}
# Hint 3 of 3
# scale_x_discrete(labels = c("Full Democracy", "Partial Democracy",
#                              "Hybrid", "Authoritarian"))
# Then add guides(fill = "none") to remove the redundant legend.
```

```{r boxplot-lifeex-by-democracylevel-check}
grade_this({
  code_lines <- gsub("#.*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing + at the end of a line, unmatched parentheses, or a label string missing its quotation marks.")
  }

  # Check filter for both variables — require !is.na() not is.na()
  code_no_comments_lines <- paste(gsub("#.*", "", .user_code), collapse = "\n")
  has_filter_dem <- grepl(
    "!\\s*is\\.na\\s*\\(\\s*dem_level4\\s*\\)",
    code_no_comments_lines
  )
  has_filter_life <- grepl(
    "!\\s*is\\.na\\s*\\(\\s*lifeex_total\\s*\\)",
    code_no_comments_lines
  )
  if (!has_filter_dem || !has_filter_life) {
    fail("Filter out missing values in both variables: filter(!is.na(dem_level4), !is.na(lifeex_total)).")
  }

  if (!any(grepl("\\bworld\\b", .user_code))) {
    fail("Use the world data frame.")
  }

  # Check aesthetics from plot object
  get_aes_name <- function(plot, aes_name) {
    out <- character(0)
    global_mapping <- plot$mapping[[aes_name]]
    if (!is.null(global_mapping)) {
      out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(global_mapping)), error = function(e) NA_character_))
    }
    for (layer in plot$layers) {
      layer_mapping <- layer$mapping[[aes_name]]
      if (!is.null(layer_mapping)) {
        out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(layer_mapping)), error = function(e) NA_character_))
      }
    }
    out[!is.na(out)]
  }

  x_vars <- get_aes_name(.result, "x")
  y_vars <- get_aes_name(.result, "y")
  fill_vars <- get_aes_name(.result, "fill")

  if (!"dem_level4" %in% x_vars) {
    fail("Map dem_level4 to x inside aes() to place regime types on the x-axis.")
  }
  if (!"lifeex_total" %in% y_vars) {
    fail("Map lifeex_total to y inside aes() to show life expectancy on the y-axis.")
  }
  if (!"dem_level4" %in% fill_vars) {
    fail("Map dem_level4 to fill inside aes() to color each box differently.")
  }

  # Check layer types
  layer_geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  if (!"GeomBoxplot" %in% layer_geoms) {
    fail("Use geom_boxplot() to create the box plots.")
  }

  # Check scale_x_discrete with labels in correct order
  code_all <- paste(.user_code, collapse = "\n")
  if (!any(grepl("scale_x_discrete", .user_code))) {
    fail("Add scale_x_discrete() to replace the abbreviated stored values with readable labels.")
  }
  label_order_ok <- grepl(
    "Full Democracy.*Partial Democracy.*Hybrid.*Authoritarian",
    gsub("\\s+", " ", code_all)
  )
  if (!label_order_ok) {
    fail('The four labels must appear in this order: "Full Democracy", "Partial Democracy", "Hybrid", "Authoritarian" --- matching the stored factor level order.')
  }
  if (!grepl('guides\\s*\\(\\s*fill\\s*=\\s*["\']none["\']', code_all)) {
    fail('Add guides(fill = "none") to remove the redundant fill legend.')
  }

  pass("Your plot is correct.")
})
```


### What Does the Box Plot Add?

The hypothesis predicts a clean ordering from full democracy to authoritarian rule. The box plot gives you more information than the means alone --- look at the full picture before deciding how well the pattern matches.

**Practice: Interpret the life expectancy box plot**

The table of means already showed that full democracies have higher mean life expectancy. Remember that the line inside each box marks the **median**, not the mean, so the box plot does not display the group means from the earlier table --- but for roughly symmetric distributions the two will be close. Look at what the box plot adds beyond the means --- focus on spread, overlap, and outliers across the four groups.

```{r quiz-boxplot-lifeex-democracylevel, echo=FALSE}
question("The table of means told us full democracies have higher average life expectancy. What does the box plot reveal that the means could not?",
  correct = "Right. The box plot reveals that the pattern is not a clean step-by-step decline.",
  answer("Hybrid and authoritarian regimes have very similar medians and substantially overlapping boxes, while partial democracies have several low outliers --- the pattern is not a clean step-by-step decline from full democracy to authoritarian rule",
         correct = TRUE,
         message = "The means suggested a monotonic decline, but the box plot shows the bottom of the ordering is much messier than the top. Hybrid and authoritarian regimes overlap so substantially that the difference between them is weak. Partial democracies have several unusually low outliers that pull their picture down. Full democracy stands clearly apart from the rest, but the other three categories do not form a clean hierarchy."),
  answer("The box plot confirms the means exactly --- life expectancy declines monotonically from full democracy to authoritarian rule with clear separation between all four groups",
         message = "Look carefully at the hybrid and authoritarian boxes. Their medians are close and their boxes overlap substantially --- there is not clear separation between all four groups the way the means might suggest."),
  answer("Full democracies have the widest spread of all four groups, suggesting their high mean is driven by a few outliers rather than a uniformly high life expectancy",
         message = "Look at the width of the full democracy box compared to the others. Full democracy actually has the narrowest box --- its high life expectancy is fairly consistent across countries, not driven by a few outliers."),
  answer("Authoritarian regimes have the lowest median and the most outliers, confirming they are the most variable group",
         message = "Look at the authoritarian box plot carefully. While authoritarian regimes do have the lowest median, their whiskers extend to cover most of their range without flagging outliers --- it is partial democracy, not authoritarian rule, that shows the most outlier points."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

**Practice: Compare the distribution of Democratic vote percentage by racial majority**

You have already computed the group means and visualized the distributions with a faceted histogram for these variables. Now build a box plot to compare them. Use **dem2p_percent** and **racial_majority** from the **counties** data frame. Filter missing values in both variables before plotting. Because **racial_majority** is stored as a character variable, no `factor()` wrapper is needed and the x-axis labels come directly from the stored values. Three hints are available. Click "Submit Answer" to check your work.

Your plot should have:

- Title: "Democratic Vote Percentage by Racial Majority"
- No subtitle
- x-axis label: "Racial Majority"
- y-axis label: "Percent of Two-Party Vote"
- Caption: "Source: Linn, Nagler, Zilinsky"

```{r boxplot-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice: Box plots of Democratic vote percentage by racial majority"}

```

```{r boxplot-dem2p-by-racialmajority-hint-1}
# Hint 1 of 3
# Use the same structure as the life expectancy box plot.
# Filter missing values in both variables before plotting.
# Think about which variable defines the groups and which
# holds the values you are comparing across groups.
```

```{r boxplot-dem2p-by-racialmajority-hint-2}
# Hint 2 of 3
# The categorical variable goes on x and fill.
# The interval variable goes on y.
# Because racial_majority is stored as a character variable,
# no factor() wrapper is needed and no scale_x_discrete()
# is needed --- the stored values are already readable.
```

```{r boxplot-dem2p-by-racialmajority-hint-3}
# Hint 3 of 3
# The categorical variable that defines the groups goes on
# x and fill. The interval variable whose distribution you
# are examining goes on y. Add guides(fill = "none") to
# remove the redundant legend.
```

```{r boxplot-dem2p-by-racialmajority-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing + at the end of a line or unmatched parentheses.")
  }

  plot_mapping <- .result$mapping
  x_mapping   <- rlang::as_label(plot_mapping$x)
  y_mapping   <- rlang::as_label(plot_mapping$y)
  fill_mapping <- rlang::as_label(plot_mapping$fill)

  if (!identical(x_mapping, "racial_majority")) {
    fail("Map racial_majority to x inside aes().")
  }
  if (!identical(y_mapping, "dem2p_percent")) {
    fail("Map dem2p_percent to y inside aes().")
  }
  if (!identical(fill_mapping, "racial_majority")) {
    fail("Map racial_majority to fill inside aes().")
  }

  has_boxplot <- any(vapply(.result$layers, function(l) inherits(l$geom, "GeomBoxplot"), logical(1)))
  if (!has_boxplot) {
    fail("Use geom_boxplot() to create the box plots.")
  }

  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Democratic Vote Percentage by Racial Majority")) {
    fail('Set title = "Democratic Vote Percentage by Racial Majority" inside labs().')
  }
  if (!identical(plot_labels$y, "Percent of Two-Party Vote")) {
    fail('Set y = "Percent of Two-Party Vote" inside labs().')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set caption = "Source: Linn, Nagler, Zilinsky" inside labs().')
  }

  pass("Your plot is correct.")
})
```

### What Does the Box Plot Add?

```{r quiz-boxplot-dem2p-racialmajority, echo=FALSE}
question("The table of means suggested that both Black Majority and Hispanic Majority counties vote heavily Democratic. What does the box plot reveal about how reliably that characterization holds within each group?",
  correct = "Right. The two groups look similar in the means but differ dramatically in spread.",
  answer("Black Majority counties have a much smaller interquartile range than Hispanic Majority counties --- the narrow box shows the middle 50% of Black Majority counties cluster tightly around a high Democratic vote percentage, while the wide Hispanic Majority box reveals far greater spread across the middle half of that group",
         correct = TRUE,
         message = "The means told you both groups lean heavily Democratic. The box plot adds that this characterization is much more reliable for Black Majority counties, whose IQR spans roughly 55-70%. Hispanic Majority counties have a far wider IQR --- roughly 25-65% --- meaning the middle half of those counties spans nearly 40 percentage points. A mean of around 50% masks that enormous spread within the group."),
  answer("Both groups have similar interquartile ranges --- their means are close and their boxes are roughly the same width",
         message = "Look carefully at the widths of the two boxes. The Black Majority box is narrow while the Hispanic Majority box is much wider --- the IQRs are very different even though the means are similar."),
  answer("The many high outliers in White Majority counties suggest those counties are a genuine swing group that could vote either way",
         message = "The outliers in White Majority counties are a small number of heavily urban counties, not evidence that White Majority counties as a whole are competitive. The box itself is narrow and concentrated at low Democratic vote percentages, which better characterizes the spread of the group."),
  answer("Hispanic Majority counties have the lowest median, which explains why their interquartile range is wider than Black Majority counties",
         message = "Spread and center are separate features of a distribution. A group can have a high or low median and still have a wide or narrow IQR --- they are independent. The Hispanic Majority median is around 45%, well above White Majority, but the IQR is wide regardless of where the median falls."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

## When Should You Use a Histogram vs. a Box Plot?

Histograms and box plots both show the distribution of an interval variable across categories, but they reveal different things and involve different trade-offs. A histogram shows the full shape of each group's distribution --- whether it is symmetric or skewed, whether it has one peak or several, and roughly how the values are spread across the range. That detail comes at a cost: histograms become increasingly cumbersome as the number of groups grows, since many faceted panels are difficult to compare at a glance.

A box plot sacrifices that fine-grained shape for efficiency. It compactly summarizes each group using the median, the middle 50% (IQR), whiskers extending to the most extreme non-outlier values, and any outlier points --- making it easy to compare many groups side by side and to quickly spot differences in central tendency, spread, and outliers. What a box plot cannot show you is whether a distribution is bimodal, or exactly how skewed it is; two very differently shaped distributions can produce nearly identical-looking boxes. The rule of thumb: use a histogram when you want to see the full shape of each group's distribution, especially when you have only a few groups. Use a box plot when you want to compare many groups efficiently, or when you care most about the median, spread, and outliers rather than fine-grained shape.

**Practice: Choose the right visualization**

Apply the rule of thumb to the scenario below before selecting an answer.

```{r quiz-histogram-vs-boxplot, echo=FALSE}
question("A researcher wants to compare the distribution of household income across the 50 U.S. states. Which tool would be the better choice, and why?",
  correct = "Right. A box plot scales to 50 groups far better than a faceted histogram.",
  answer("A box plot, because comparing 50 groups requires an efficient summary, and a faceted histogram with 50 panels would be hard to read",
         correct = TRUE,
         message = "With this many groups, the fine-grained shape a histogram shows becomes hard to use --- 50 panels are difficult to scan and compare. A box plot's compact summary scales much better to many groups while still showing medians, spreads, and outliers."),
  answer("A histogram, because it shows more detail about each state's distribution than a box plot would",
         message = "A histogram does show more detail, but that level of detail becomes a liability with 50 groups --- 50 histogram panels are too many to compare effectively. A box plot trades some detail for the ability to compare many groups at once."),
  answer("It does not matter, since histograms and box plots always show the same information",
         message = "Histograms and box plots reveal different features of a distribution. A histogram shows the full shape, including skew and modality; a box plot summarizes that shape more compactly and loses some detail in exchange for being easier to compare across many groups."),
  answer("A box plot, but only because income distributions are always skewed --- if income were symmetric, a histogram would be the better choice",
         message = "The number of groups, not the shape of the distribution, is the main reason to prefer a box plot here. With 50 states, 50 histogram panels are too many to compare effectively regardless of whether the distributions are skewed or symmetric."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

## Check Your Understanding

Now let’s step back and practice the full workflow with new examples. You do not need to write any code. Focus on the logic: decide whether the research question fits this tutorial’s setup, compare means across categories, and use plots to see what the means alone might hide.

**Apply: Classify the research question**

For each question, identify which variable is the outcome and which defines the groups, then decide whether that setup matches what this tutorial covers.

```{r question-forward-reverse, echo=FALSE}
question("Which of the following research questions fits the pattern covered in this tutorial --- a categorical explanatory variable and an interval outcome --- rather than the reverse case?",
  correct = "Right. Region is categorical and commute time is interval --- exactly the setup this tutorial covers.",
  answer("Do commute times differ by region (Northeast, South, Midwest, West)?",
         correct = TRUE,
         message = "Region is categorical and commute time is interval, with commute time as the outcome --- exactly the pattern this tutorial covers. The other two questions flip the direction: an interval explanatory variable and a categorical outcome, which requires a different approach."),
  answer("Is a student's SAT score associated with whether they are admitted to a selective college?",
         message = "Here SAT score (interval) is the explanatory variable and admission (categorical) is the outcome --- not the setup this tutorial focuses on. You could compare mean SAT scores between admitted and non-admitted students --- that describes an association --- but it does not directly address how the probability of admission changes as SAT scores rise, which requires a different tool."),
  answer("Is a county's median household income associated with whether it voted for the incumbent party?",
         message = "Here income (interval) is the explanatory variable and vote choice (categorical) is the outcome --- not the setup this tutorial focuses on. You could compare mean income across voting groups --- that describes an association --- but it does not directly address how the probability of voting for the incumbent changes with income, which requires a different approach."),
  answer("Does a country's regime type predict its probability of experiencing civil war?",
         message = "Here regime type (categorical) is the explanatory variable and civil war occurrence (categorical --- yes or no) is the outcome --- not the setup this tutorial focuses on. When the outcome is categorical, comparing group means is not the right approach."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

Consider a researcher studying whether commute times differ by region. The table below shows the mean daily commute time, in minutes, for a sample of cities in four regions.

| Region    | Mean Commute Time (minutes) | N  |
|-----------|------------------------------|----|
| Northeast | 34.2                          | 48 |
| Midwest   | 24.6                          | 52 |
| South     | 26.1                          | 61 |
| West      | 28.9                          | 45 |

**Apply: Evaluate the commute-time hypothesis**

Read the table above and compare the Northeast's mean to the other three regions before answering.

```{r question-commute-table, echo=FALSE}
question("A researcher hypothesizes that cities in the Northeast have longer commute times than cities in the other three regions, due to denser populations and older transit infrastructure. Based on the table, which statement best describes the evidence for this hypothesis?",
  correct = "Right. The Northeast stands clearly apart while the other three regions are close to each other.",
  answer("The pattern is consistent with the hypothesis --- the Northeast's mean commute time is several minutes higher than any other region, while the other three regions are fairly close to each other",
         correct = TRUE,
         message = "The Northeast's mean (34.2) is at least 5 minutes higher than any other region, while the Midwest, South, and West are all within about 4 minutes of each other. This is consistent with the hypothesis, but this is a descriptive comparison, not a formal test."),
  answer("The hypothesis is confirmed, since the Northeast clearly has the highest mean commute time",
         message = "A descriptive comparison of means cannot confirm a hypothesis, no matter how large the difference looks. Confirming it would require a formal hypothesis test."),
  answer("The pattern is not consistent with the hypothesis, since commute times in all four regions are fairly similar",
         message = "Look at the actual numbers again. The Northeast's mean is several minutes higher than the other three regions, which are much closer to each other than any of them are to the Northeast."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```

The same researcher now produces box plots of commute time for each region, shown below.

```{r cyu-boxplot-commute, echo=FALSE, fig.alt="Box plots showing daily commute time in minutes on the y-axis and region on the x-axis. The Northeast box is tall, ranging from about 30 to 40 minutes, with three high outliers between 60 and 70 minutes. The Midwest, South, and West boxes are all short and similarly sized, roughly 23 to 30 minutes, with no outliers."}
set.seed(7)
northeast <- c(rnorm(40, mean = 33, sd = 8), 62, 65, 68)
midwest <- rnorm(45, mean = 24.5, sd = 2.5)
south <- rnorm(50, mean = 26, sd = 2.5)
west <- rnorm(40, mean = 29, sd = 2.7)

commute_df <- data.frame(
  region = factor(c(rep("Northeast", length(northeast)),
                     rep("Midwest", length(midwest)),
                     rep("South", length(south)),
                     rep("West", length(west))),
                   levels = c("Northeast", "Midwest", "South", "West")),
  commute = c(northeast, midwest, south, west)
)

ggplot(commute_df, aes(x = region, y = commute, fill = region)) +
  geom_boxplot() +
  labs(title = "Distribution of Commute Time by Region",
       x = "Region",
       y = "Daily Commute Time (minutes)",
       caption = "Hypothetical data") +
  guides(fill = "none")
```

**Apply: Interpret the commute-time box plots**

Think about what the table already told you, then identify what the box plot adds that the table could not.

```{r question-commute-boxplot, echo=FALSE}
question("What does this plot tell us that the table of means alone could not?",
  correct = "Right. The box plot reveals spread and outliers that the means alone could not show.",
  answer("The Northeast has a much wider spread of commute times and several unusually high outliers, while the other three regions are both shorter and far more consistent with each other",
         correct = TRUE,
         message = "The Northeast's box spans roughly 30 to 40 minutes, more than three times the width of the other regions' boxes, and it has three outlier cities above 60 minutes --- features completely invisible in a table that only reports means. The means told you the Northeast is higher on average; the box plot reveals it is also far more variable."),
  answer("This information is redundant with the table of means and tells us nothing new",
         message = "The table of means only tells us the Northeast's typical commute time is higher. The box plot reveals that the Northeast is also far more variable, with outlier cities --- information the means alone do not provide."),
  answer("Because the Northeast has outliers, its mean commute time is unreliable and should be ignored",
         message = "The presence of outliers does not make the mean meaningless; it helps explain why the mean is elevated. The right response is to consider the whole distribution, not to disregard the mean."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Incorrect. Try again."
)
```


## The Takeaways

This tutorial covered [**bivariate description**]{.important-text} of a categorical and an interval variable --- asking whether the distribution of an interval outcome differs across the categories of a categorical explanatory variable. Tutorial 7 covered two categorical variables; Tutorial 9 will cover two interval variables. The core question is the same throughout --- does the pattern in the data point in the direction the hypothesis predicts? --- but the tools change depending on the variable types.

- [**Comparing group means**]{.important-text} is the first move when one variable is categorical and one is interval. Use `group_by()` followed by `summarise()` to compute the count and mean within each group.
  - Always filter missing values in **both** the grouping variable and the outcome variable before `group_by()` --- this ensures `freq = n()` and `mean()` operate on exactly the same set of observations, so the count accurately reflects the N contributing to each mean.
  - Always set the analysis up in the right direction: the interval variable is the outcome, the categorical variable defines the groups. Reversing this answers a different question.
  - A comparison of means is not a hypothesis test --- finding that group means differ in the predicted direction tells you the data are consistent with the hypothesis, not that the difference is larger than chance. That requires a formal test, covered in Tutorial 13.

- [**Visualizing distributions across groups**]{.important-text} shows what the means cannot. A table of means hides differences in spread, shape, and outliers. Always filter missing values in the variables used in the plot before passing data to `ggplot()`. This keeps missing categories out of the display and avoids ggplot silently removing observations or producing warnings.
  - For **faceted histograms**, map the categorical variable to `fill` inside `aes()` and add `facet_wrap(~variable)`. By default all panels share the same y-axis, making group size easy to compare; adding `scales = "free_y"` gives each panel its own scale so the shape of every distribution is visible, at the cost of no longer being able to compare bar heights across panels.
  - Whether you need `factor()` depends on how the categorical variable is stored. If it is stored as a **number**, wrap it in `factor()` when mapping to `fill` --- without it, ggplot applies a continuous color gradient rather than discrete colors, and categories are ordered numerically (0, 1, 2, ...). If it is stored as a **character**, `factor()` is not needed --- ggplot already treats it as discrete and orders categories alphabetically. If it is already stored as a **factor**, `factor()` is not needed --- the stored factor levels define the order. Inside `facet_wrap()`, `factor()` is never needed; use `labeller = as_labeller(c(...))` to give panels readable titles when the stored values are not self-explanatory.
  - When the panel titles or x-axis already identify the groups, the fill legend is redundant --- add `guides(fill = "none")` to remove it. This applies to both faceted histograms and box plots.
  - For **box plots**, map the categorical variable to both `x` and `fill`, and the interval variable to `y`. The same `factor()` rules apply: numeric variables need wrapping, character and factor variables do not. Box plots sacrifice distributional detail for efficiency: a compact summary that scales to many groups and makes medians, spreads, and outliers easy to compare at a glance.

The choice between a histogram and a box plot comes down to what you want to show. Histograms reveal the full shape of each group's distribution --- skew, modality, spread --- but become increasingly cumbersome as the number of groups grows. Box plots are more efficient but show less detail. In practice, you will often want both.

In R, load each package once at the top of your script or Quarto document with `library(dplyr)` and `library(ggplot2)` before any code that uses functions from those packages. You do not need to repeat the `library()` call every time you use a function from it.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. Remember, the code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-08, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-08, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-08, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Great job! You have completed this tutorial and can now compare interval variables across groups using means, histograms, and box plots.">
  <span aria-hidden="true">🎉</span>
  Great job! You can now compare interval variables across groups using means, histograms, and box plots.
</p>
')
```
