---
title: "Bivariate Description: Categorical and Interval Variables"
tutorial:
  id: "08-Bivariate-Description-Categorical-Interval-Data"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Group means, faceted histograms, and grouped box plots.
---

## Overview {#main-content}

News reports and political research frequently compare averages across groups: whether turnout is higher in one kind of community, whether citizens in different parties express different levels of trust, or whether economic outcomes differ across political systems. An average provides a useful starting point, but it cannot show whether the difference characterizes most cases, reflects substantial overlap between the groups, or is driven by a few unusual observations.

An average can be checked against the distribution it summarizes. You will compute the mean of an interval variable within each group, then look at the whole distribution group by group --- with faceted histograms and with grouped box plots --- to see what the means leave out. The statistics come from Tutorial 6, which introduces the mean, the standard deviation, the histogram and the box plot for a single interval variable.

In this tutorial you will learn:

* How to compare the mean of an interval variable across the categories of a categorical variable using `group_by()` and `summarise()` from the [dplyr]{.package-name} package
* How to recognize what a table of means conceals about spread, shape, and outliers
* How to use `ggplot()` from the [ggplot2]{.package-name} package with `geom_histogram()` and `facet_wrap()` to compare the full distribution of an interval variable across categories
* How to use `ggplot()` from the [ggplot2]{.package-name} package with `geom_boxplot()` and a categorical x-axis to compare groups efficiently
* How to choose between a histogram and a box plot based on how many groups you are comparing and how much distributional detail you need


```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)
data("counties", package = "QPATutorialsCourse")
data("world", package = "QPATutorialsCourse")
```


## How Does an Interval Outcome Vary Across Groups?

Many questions in political science ask whether some measurable quantity differs across political or social categories. Do counties with a Black racial majority cast a higher Democratic vote percentage than White-majority counties? Are full democracies wealthier on average than autocracies? Do militarized interstate disputes last longer when one side has nuclear weapons? In each case, the question has the same structure: take an interval variable --- Democratic vote percentage, GDP per capita, conflict duration --- and ask whether its distribution differs across categories of a second variable --- racial majority, regime type, nuclear status. That is [bivariate description]{.important-text}: examining two variables at a time to see whether the values of one differ systematically across the values of the other. This is the second of three cases, sorted by the levels of measurement involved: Tutorial 7 took two categorical variables, and Tutorial 9 takes two interval ones. All three apply across every subfield of political science.

There are two ways to examine this kind of relationship. The first is to compare [group means]{.important-text} --- the mean of the interval variable computed separately within each category of the explanatory variable. If democracies are wealthier than autocracies, the mean GDP per capita should be higher in the democratic category. A group mean gives a concise, single-number summary of the average value in each group. The second is to compare the full **distribution** of the interval variable across groups using histograms or box plots. A mean can hide important features of a distribution, which is why this tutorial turns to the full distribution later.

Both approaches require setting up the relationship in the right direction: the interval variable is the outcome and the categorical variable defines the groups. The key is identifying which variable is the outcome. For example, asking whether mean wage growth differs between Trump-majority and Clinton-majority counties fits this tutorial because Trump-majority status is the categorical explanatory variable defining the groups and wage growth is the interval outcome being compared across them. A different question --- does wage growth predict the *probability* that a county supported Trump? --- has an interval explanatory variable and a binary categorical outcome, which requires a different tool. The distinction is not about which variables are involved but about which one is the outcome: when the outcome is interval and the explanatory variable is categorical, compare group means. When the outcome is categorical, a different approach is needed. Each example that follows opens with a research question and a hypothesis, and asks whether the pattern matches what the hypothesis predicts.

## Is College Attainment Lower in Rural Counties?

A bachelor's degree is one of the strongest predictors of a person's income, occupation and political attitudes in the United States, which makes the question of where graduates live more than a demographic curiosity. The share of a county's adults holding a degree shapes which employers will consider locating there, which public services it can afford, and whether its young people expect to build a life nearby or somewhere else. Over the past four decades that share has grown steadily more uneven across American communities rather than converging.

The theoretical argument for why the unevenness should track the rural-urban divide turns on migration rather than schooling. Work that requires a degree --- in professional services, technology, research, health administration, finance --- clusters where firms, clients and universities sit close together, which in practice means metropolitan areas. A student who leaves a rural county to attend college therefore finds the jobs her degree qualifies her for somewhere other than home, and a graduate choosing where to settle finds little reason to move to a county that offers few of them. Repeated across cohorts, that sorting concentrates graduates in and around cities.

It is worth being clear about what this argument does not say. It makes no claim about who is capable of earning a degree, and it does not predict that rural high schools send fewer students to college. It predicts only where graduates end up living --- which is exactly what a county-level measure of educational attainment captures, and all it captures.

That argument leads to a testable hypothesis:

::: {.hypothesis}
In a comparison of [counties], those that are [more rural] will have a [lower share of college graduates] than will those that are [more urban].
:::

The data frame **counties** contains the ordinal variable **rural**, which takes the value 0 for urban counties, 1 for counties that are neither distinctly urban nor rural, and 2 for rural counties. It also contains the interval variable **prop_college_grad**, the proportion of a county's residents age 25 or older who have completed a bachelor's degree or higher. Every county has a recorded value of **rural**, while **prop_college_grad** is missing for 1 county, so any summary you compute from it describes the 3,111 counties with a recorded value rather than all 3,112 in the data. The hypothesis leads you to expect that the mean proportion of college graduates should be lowest in rural counties and highest in urban counties.

### Calculate the Means

Tutorial 5 uses `group_by()` and `summarise()` from [dplyr]{.package-name} to count the number of cases in each category. The same two-step structure works for any summary statistic --- here, in addition to `n()` to count, you also use `mean()` to calculate the group mean within each group. The logic is identical: `group_by()` defines the groups and `summarise()` computes statistics within each one. You will use this same structure throughout this tutorial.

The run-and-observe below calculates the mean proportion of college graduates separately for urban, neither-urban-nor-rural, and rural counties. `freq = n()` counts the number of counties in each group, and `mean = mean(prop_college_grad)` calculates the mean within each group. Filtering **prop_college_grad** before grouping ensures `freq = n()` reflects the number of counties contributing to each mean --- without it, `n()` would count that missing county while `mean()` returned `NA` for its group. Filtering first is why `mean()` needs no `na.rm = TRUE` here: by the time it runs, there is nothing left to remove. **rural** needs no condition of its own. Filter the variables that have gaps, and only those.

**Run and observe 1: Calculate mean proportion of college graduates by rural**

```{r frequency-collegegrad-by-rural, exercise=TRUE, exercise.cap="Run and observe 1"}
library(dplyr)  # first use in this tutorial; it stays loaded from here on
counties %>%
  filter(!is.na(prop_college_grad)) %>%
  group_by(rural) %>%
  summarise(freq = n(), mean = mean(prop_college_grad))
```

This output looks similar to the frequency tables from Tutorial 5 --- the rows name the categories of **rural**, the **freq** column counts the number of counties in each group just as it did before, and the **mean** column gives the mean proportion of college graduates within each group. The only new piece is that final column.

### Is the Pattern Consistent with the Hypothesis?

**Apply 2: Evaluate the rural/college-graduate hypothesis**


```{r quiz-collegegrad-rural-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells you about the hypothesis?",
  correct = "Right. The pattern is consistent with the hypothesis, but with nuance.",
  answer("The pattern is consistent with the hypothesis, but mainly because urban counties stand out --- the neither-urban-nor-rural and rural categories are close to each other",
         correct = TRUE,
         message = "Urban counties have a mean proportion of college graduates of 0.255 --- well above neither-urban-nor-rural (0.181) and rural (0.177) counties, which are nearly identical to each other. The hypothesis predicts a smooth decline across all three categories, but the data show a large urban/non-urban gap with very little difference between the two non-urban categories."),
  answer("The pattern is consistent with the hypothesis, with the mean proportion of college graduates dropping by roughly the same amount at each step from urban to neither-urban-nor-rural to rural",
         message = "The order is correct --- urban is highest, rural is lowest --- but the size of the steps is not equal. The drop from urban (0.255) to neither-urban-nor-rural (0.181) is large, while the drop from neither-urban-nor-rural (0.181) to rural (0.177) is small. Most of the difference comes from urban counties standing apart, not from an even decline."),
  answer("The hypothesis is confirmed --- the difference in means proves that rurality causes lower educational attainment",
         message = "Descriptive comparison of means cannot confirm a hypothesis. Finding a pattern in the predicted direction tells you the data are consistent with the hypothesis, but a formal hypothesis test is needed to assess statistical significance."),
  answer("The pattern is not consistent with the hypothesis --- neither-urban-nor-rural and rural counties have nearly identical means, which contradicts the predicted decline across all three categories",
         message = "A hypothesis about direction does not require a perfectly smooth decline across every category. This hypothesis predicts that more rural counties will have lower college graduation rates than more urban counties --- and they do. Urban counties have a mean of 0.255, while both non-urban categories are well below that at around 0.18. The fact that neither-urban-nor-rural and rural counties are close to each other is a nuance worth noting, but it does not make the overall pattern inconsistent with the hypothesis."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Look at all three group means, not just the highest and the lowest. Consider the direction of the pattern and whether the gap between each adjacent pair of categories is the same size. Consider too what a comparison of means can and cannot establish."
)
```

## Is the Democratic Vote Percentage Related to a County's Racial Majority?

In 2016 Hillary Clinton won 51.1% of the national two-party vote and lost the election --- an outcome that turned on where her votes were rather than how many of them there were. County results are where that geography becomes visible, and the racial composition of a county is one of the sharpest lines along which it runs.

The theoretical argument points to the parties' positions rather than to the voters. Since the civil rights realignment of the 1960s the Democratic Party has been the one more closely identified with civil rights and voting rights enforcement, and Black Americans have supported its candidates by margins rarely seen for any group in any democracy --- routinely above eighty percent, and often above ninety. The 2016 campaign gave voters little reason to depart from that pattern: immigration, policing and racial grievance were unusually prominent, and both candidates made explicit appeals to racially distinct constituencies.

Notice what changes from the last tutorial. Tutorial 7 asked whether a county gave Trump a majority --- a yes-or-no outcome, which is why a cross-tab could answer it. Here the outcome is the Democratic share of the two-party vote, measured as a percentage, so the question is not only which way a county leaned but by how much. Comparing group means is what lets you ask that.

Stated as a hypothesis:

::: {.hypothesis}
In a comparison of [counties], those with a [Black racial majority] will have a [higher mean Democratic vote percentage] than will those with a [White racial majority].
:::

The data frame **counties** contains the nominal variable **racial_majority**, which takes the values "White Majority," "Black Majority," "Hispanic Majority," and "Other/No Majority," depending on which racial group makes up a majority of the county's population. It is stored as a character variable. It also contains the interval variable **dem2p_percent**, the Democratic candidate's share of the two-party presidential vote in the county in 2016, as a percentage. Neither variable has any missing values, so summaries computed from them describe all 3,112 counties.

### Calculate the Means

To evaluate the hypothesis, you need the mean Democratic vote percentage for each racial majority category. Even though the hypothesis only compares Black and White Majority counties, you calculate the mean for all four categories at once --- that is how `group_by()` always works.

**Practice 3: Compare mean Democratic vote percentage by racial majority**

Complete the code below using the same `group_by()`/`summarise()` structure as the previous example, adapted for these variables. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice 3"}
counties %>%
  ___(___) %>%
  ___(freq = ___(), mean = mean(___))
```

```{r exercise-dem2p-by-racialmajority-hint-1}
# Hint 1 of 2
# You want one row for each racial majority category,
# showing how many counties fall in it and their average
# Democratic vote percentage --- the same shape as the
# college graduation table. Which variable defines the
# groups, and which holds the values you are averaging?
```

```{r exercise-dem2p-by-racialmajority-hint-2}
# Hint 2 of 2
# group_by() takes the explanatory variable.
# Inside summarise(), use n() to count cases and name it freq,
# and mean() to compute the group mean and name it mean.
```

```{r exercise-dem2p-by-racialmajority-check}
grade_this({
  code_no_comments <- paste(gsub("#[^\r\n]*", "", .user_code), collapse = "\n")

  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing %>% between steps. Inside `summarise()`, you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the `=`.")
  }
  if (!grepl("\\bcounties\\b", code_no_comments)) {
    fail("Use the **counties** data frame.")
  }
  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use `group_by()` to group counties by **racial_majority**.")
  }
  if (!grepl("group_by\\s*\\(\\s*racial_majority\\s*\\)", code_no_comments)) {
    fail("Group by **racial_majority** --- that is the variable that defines each county's racial majority category.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use `summarise()` to calculate the mean within each group.")
  }
  if (!grepl("freq\\s*=\\s*n\\s*\\(\\s*\\)", code_no_comments)) {
    fail("Name the count variable **freq** using `freq = n()` inside `summarise()`.")
  }
  if (!grepl("mean\\s*=\\s*mean\\s*\\(\\s*dem2p_percent", code_no_comments)) {
    fail("Name the mean variable **mean** and compute it from **dem2p_percent**: `mean = mean(dem2p_percent)`.")
  }

  expected <- counties %>%
    group_by(racial_majority) %>%
    summarise(freq = n(), mean = mean(dem2p_percent), .groups = "drop")

  result <- as.data.frame(.result)
  expected <- as.data.frame(expected)

  if (!all(c("racial_majority", "freq", "mean") %in% names(result))) {
    fail("Your table should include **racial_majority**, freq, and mean.")
  }

  result <- result[, c("racial_majority", "freq", "mean")]
  result <- result[order(result$racial_majority), ]
  expected <- expected[order(expected$racial_majority), ]

  if (!isTRUE(all.equal(result, expected, tolerance = 1e-6, check.attributes = FALSE))) {
    fail("Check that you grouped by **racial_majority** and calculated `freq = n()` and `mean = mean(dem2p_percent)`.")
  }

  extra <- character(0)
  if (grepl("filter\\s*\\(", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Correct. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but neither of these variables has a missing value, so there was nothing to remove. Check the counts before reaching for either."))
  }
  pass("Your table is correct --- one row per racial majority category, with the count and the mean of **dem2p_percent** within each.")
})
```

### Is the Pattern Consistent with the Hypothesis?


**Apply 4: Evaluate the racial-majority hypothesis**


```{r quiz-racialmajority-dem2p-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells you about the hypothesis?",
  correct = "Right. Black Majority counties have a substantially higher mean Democratic vote percentage than White Majority counties.",
  answer("The pattern is consistent with the hypothesis --- Black Majority counties have the higher mean",
         correct = TRUE,
         message = "Black Majority counties have a mean Democratic vote percentage of 65.21% compared to 31.34% for White Majority counties --- a difference of more than thirty percentage points, consistent with the hypothesis. The comparison of means alone cannot tell you whether the difference is larger than chance or what caused it."),
  answer("The pattern is not consistent with the hypothesis --- White Majority counties have a higher mean Democratic vote percentage than Black Majority counties",
         message = "Look at the table again. Black Majority counties (65.21%) have a higher mean Democratic vote percentage than White Majority counties (31.34%), not the reverse."),
  answer("The hypothesis is confirmed --- the difference in means proves racial composition determines vote choice",
         message = "Descriptive comparison of means cannot confirm a hypothesis. It can only show whether the pattern is consistent with it."),
  answer("The hypothesis cannot be evaluated because **racial_majority** has more than two categories",
         message = "Comparing means of an interval variable across more than two categories of a nominal variable is exactly the right descriptive method here --- the hypothesis simply focuses on two of those categories."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Compare the mean for each category in your table against what the hypothesis predicts. Two things go wrong here most often: reading the comparison in the wrong direction, and treating a difference in means as more than evidence consistent with the hypothesis."
)
```


## Is Democracy Associated with Economic Development?

Whether democracy is good for growth is among the most argued-over questions in comparative politics, and it has been answered both ways by serious people. The optimistic case credits democratic institutions with secure property rights, accountable administration, and policy aimed at broad publics rather than narrow elites. The pessimistic case notes that several of the fastest-growing economies of the last half-century did their growing under authoritarian rule.

The theoretical argument on the optimistic side runs through accountability. A leader who can be removed for poor economic performance has reason to protect contracts, invest in infrastructure and let markets function; a leader who cannot faces no such discipline and can extract from the economy rather than build it.

One caution before you look. Wealthy societies are also more likely to become democracies and to remain so, which means a difference in average GDP between democracies and non-democracies is consistent with the causal arrow running either way, or with both. A comparison of group means cannot settle that. It can tell you whether the two travel together.

If the accountability argument is right:

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have [higher levels of economic development] than will those that are [less democratic].
:::

The data frame **world** contains the interval variable **gdppcap08**, GDP per capita in 2008 measured in US dollars. This example uses GDP per capita rather than total GDP because it accounts for country size --- China and the United States have large total GDPs partly because of their large populations, not necessarily because their citizens are more prosperous. Dividing by population gives a better sense of average living standards. The ordinal variable **dem_level4** records regime type on a four-category scale running from full democracy to authoritarian rule. It is stored as a factor whose values are abbreviated: `"Full Democ"`, `"Part Democ"`, `"Hybrid"`, `"Authoritarian"`. Every country has a recorded regime type, while **gdppcap08** is missing for 15 countries, so any summary you compute from it describes the 152 countries with a recorded value rather than all 167 in the data.

### Calculate the Means

To evaluate the hypothesis you need the mean GDP per capita for each regime type.

**Apply 5: Calculate mean GDP by democracy level**

Write the full pipeline from scratch using **gdppcap08** and **dem_level4** in the **world** data frame, naming the counts **freq** and the means **mean**. Try not to look at the hints. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-gdp-by-democracy-level, exercise=TRUE, exercise.cap="Apply 5"}

```

```{r exercise-gdp-by-democracy-level-hint-1}
# Hint 1 of 2
# You want one row per regime type, showing how many
# countries fall in it and their average GDP per capita.
# That is the same two-step structure as the previous
# exercises. One of these two variables has gaps --- check
# which, and deal with it before the counting starts.
```








```{r exercise-gdp-by-democracy-level-hint-2}
# Hint 2 of 2
# Use the world data frame. gdppcap08 is the one with
# missing values, so filter on it before group_by(); once
# those rows are gone, mean() needs no extra argument.
# dem_level4 defines the groups. Inside summarise(), n()
# counts and mean() averages --- name them freq and mean.
```

```{r exercise-gdp-by-democracy-level-check}
grade_this({
  code_no_comments <- paste(gsub("#[^\r\n]*", "", .user_code), collapse = "\n")

  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing %>% between steps. Inside `summarise()`, you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the `=`.")
  }
  if (!grepl("\\bworld\\b", code_no_comments)) {
    fail("Use the **world** data frame.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*gdppcap08\\s*\\)", code_no_comments)) {
    fail("Filter out the missing **gdppcap08** values before grouping: `filter(!is.na(gdppcap08))`. This keeps `freq = n()` and `mean()` describing the same set of countries.")
  }
  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use `group_by()` to group countries by **dem_level4**.")
  }
  if (!grepl("group_by\\s*\\(\\s*dem_level4\\s*\\)", code_no_comments)) {
    fail("Group by **dem_level4** --- that is the variable that defines regime type.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use `summarise()` to calculate the mean within each group.")
  }
  if (!grepl("freq\\s*=\\s*n\\s*\\(\\s*\\)", code_no_comments)) {
    fail("Name the count variable **freq** using `freq = n()` inside `summarise()`.")
  }
  if (!grepl("mean\\s*=\\s*mean\\s*\\(\\s*gdppcap08", code_no_comments)) {
    fail("Name the mean variable **mean** and compute it from **gdppcap08**.")
  }
  if (any(is.na(.result$mean))) {
    fail("Your mean column contains NA. **gdppcap08** has missing values, so they have to be dealt with --- filter them out before grouping, which also keeps `freq` honest.")
  }

  expected <- world %>%
    filter(!is.na(gdppcap08)) %>%
    group_by(dem_level4) %>%
    summarise(freq = n(), mean = mean(gdppcap08), .groups = "drop")

  result <- as.data.frame(.result)
  expected <- as.data.frame(expected)

  if (!all(c("dem_level4", "freq", "mean") %in% names(result))) {
    fail("Your table should include **dem_level4**, freq, and mean.")
  }

  result <- result[, c("dem_level4", "freq", "mean")]
  result <- result[order(result$dem_level4), ]
  expected <- expected[order(expected$dem_level4), ]

  if (!isTRUE(all.equal(result, expected, tolerance = 1e-6, check.attributes = FALSE))) {
    fail("Check that you filtered the missing **gdppcap08** values, grouped by **dem_level4**, and calculated `freq = n()` and `mean = mean(gdppcap08)`.")
  }

  extra <- character(0)
  if (grepl("!\\s*is\\.na\\s*\\(\\s*dem_level4\\s*\\)", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Correct. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but every country has a recorded **dem_level4**, and filtering **gdppcap08** already removed the missing values before `mean()` ran."))
  }
  pass("Exactly. You grouped by regime type and computed a count and a mean within each group, filtering **gdppcap08** first so the two columns describe the same countries.")
})
```

### Is the Pattern Consistent with the Hypothesis?

**Apply 6: Evaluate the democracy/GDP hypothesis**


```{r quiz-gdp-hypothesis, echo=FALSE}
question("Based on the table you produced, which statement best describes what the pattern tells you about the hypothesis?",
  correct = "Right. The overall direction holds; one category is out of order.",
  answer("The pattern is fully consistent with the hypothesis --- mean GDP per capita increases consistently from authoritarian to full democracy",
         message = "Look at the table carefully. Full democracies do have the highest mean GDP per capita, but the ordering across the other three regime types does not follow the predicted pattern."),
  answer("The pattern is partially consistent with the hypothesis --- the ordering across regime types does not follow a clean democratic hierarchy",
         correct = TRUE,
         message = "Full democracies have the highest mean GDP per capita and hybrid regimes have the lowest, which is not the predicted order. Some authoritarian states --- particularly those with natural resource wealth --- push the authoritarian mean above hybrid democracies, complicating the picture. The pattern is partially consistent with the hypothesis but not cleanly."),
  answer("The pattern is not consistent with the hypothesis --- authoritarian regimes outperform at least one democratic regime type",
         message = "The premise is right: authoritarian regimes do outperform hybrid ones. But that makes the pattern partially consistent, not inconsistent. Full and partial democracies both average higher than authoritarian regimes, which is what the hypothesis predicts; the ordering only breaks down at hybrid regimes. One misplaced category weakens a pattern rather than reversing it."),
  answer("GDP per capita cannot be compared across regime types because countries differ in size",
         message = "GDP per capita is specifically designed to account for country size by dividing total GDP by population. Comparing it across groups is entirely appropriate."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Work through each regime type in turn rather than only the two extremes. A pattern can run in the predicted direction overall and still break down between two adjacent categories."
)
```


## What Does a Table of Means Hide?

A table of means tells you whether groups differ on average, but it can hide a great deal. Two groups can have similar means even though one group is much more spread out than the other. A group's mean can also be pulled away from its typical value by a small number of extreme outliers. Two groups can even have the same mean while having very different shapes --- for example, one distribution might be symmetric while another is skewed. None of this is visible in a table of means alone.

To see these features, you use the same two tools introduced in Tutorial 6 for visualizing a single interval variable --- histograms and box plots --- but now broken down by categories of the explanatory variable, so that you can compare the full distribution across groups rather than just the average.

## Visualizing Distributions Across Categories: Faceted Histograms

A [faceted histogram]{.important-text} draws one histogram per category, side by side, so you can see the shape of each group's distribution rather than only its average.

Only one thing is new here, and it is the panels. Everything else you have done before. Two additions produce them: mapping the categorical variable to `fill`, which colors each group's bars differently, and adding `facet_wrap()`, which gives each group a panel of its own. Splitting a plot up this way is called [faceting]{.important-text}, and each panel is a [facet]{.important-text}.

`facet_wrap()` does not take a bare variable name. It takes a **formula** --- a tilde followed by the variable that defines the panels, as in `facet_wrap(~dem_level4)`. A formula in R states a relationship between variables rather than a value to compute, and R functions that need to know which variable plays which role ask for one. Faceting needs only the grouping variable, so nothing appears to the left of the tilde. Later tutorials use two-sided formulas, where the variable on the left is the outcome and the one on the right explains it.

When `fill` maps a variable to color, ggplot automatically adds a legend identifying which color corresponds to which category. In a faceted histogram, that legend is redundant --- the panel titles already identify which group each histogram shows. Adding `guides(fill = "none")` removes the legend and keeps the plot clean.

Filter missing values when leaving them in would change the answer, and not otherwise. Before a plot like this one it would not: **gdppcap08** is missing for 15 countries, and a missing number has no place on a continuous axis, so ggplot drops those rows itself and prints a message saying how many. Read that count and report it. Two cases are different. A missing *category* gets a panel of its own, sitting there in plain sight --- which is why Tutorials 4 and 5 filter before plotting proportions. And a table of means still needs its filter, so that `n()` and `mean()` describe the same countries.

**Run and observe 7: The distribution of GDP per capita within each regime type**

```{r histogram-gdp-by-demlevel, exercise=TRUE, exercise.cap="Run and observe 7", fig.alt="Four histogram panels of GDP per capita, one per regime type. Full Democracy has a small clump at the low end, a gap, then most of its countries between about thirty-three and forty-seven thousand, with two further out at roughly fifty-seven and seventy-eight thousand. Part Democracy spreads across the lower range with a thin tail. Hybrid and Authoritarian both pile up against the left edge and thin out to the right, and one authoritarian country sits past eighty-five thousand with nothing near it."}
library(ggplot2)  # first use in this tutorial; it stays loaded from here on
ggplot(data = world, mapping = aes(x = gdppcap08, fill = dem_level4)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of GDP per Capita by Level of Democracy",
       x = "GDP per Capita, 2008 (US$)",
       y = "Frequency",
       caption = "Source: World Bank; Economist Intelligence Unit") +
  facet_wrap(~dem_level4) +
  guides(fill = "none")
```

R prints the message it promised: 15 rows removed, the 15 countries with no recorded GDP per capita. The panels below describe the rest. Look at the four shapes. Not one of them is what a single number per group would have led you to expect. Hybrid and Authoritarian countries pile up against the left edge and thin out to the right --- most poor, a few much less so. Full Democracies look different again: a small clump at the low end, a gap, then the bulk of the group between roughly \$33,000 and \$47,000, with two countries further out still. And in the Authoritarian panel, one country sits past \$85,000 with nothing near it. A mean is one point along these axes. None of these four shapes is well described by a single point.

## Does the Whole Distribution of Democratic Vote Percentage Differ by Racial Majority?

Earlier you found that Black Majority counties have a much higher mean Democratic vote percentage than White Majority counties. A faceted histogram lets you see whether that gap reflects a shift in the whole distribution or something narrower.

The variables are **dem2p_percent** and **racial_majority**, from the **counties** data frame, and neither has a missing value. Because **racial_majority** is stored as a character variable, ggplot treats it as discrete automatically and orders its categories alphabetically --- no `factor()` wrapper is needed. The panel titles here come straight from the values of **racial_majority**, which are already readable text.


**Practice 8: Compare the distribution of Democratic vote percentage across racial majority categories**

This is the plot you just saw, on different variables --- so the labels and the geom are already filled in. What is left is the three decisions: which variable varies along the x-axis, which one colors and separates the groups, and what goes inside `facet_wrap()` to make the panels happen. Three hints are available. Click "Submit Answer" to check your work.

```{r histogram-dem2p-by-racialmajority, exercise=TRUE, message=FALSE, exercise.cap="Practice 8", fig.alt="Faceted histograms showing the distribution of Democratic vote percentage on the x-axis and frequency on the y-axis, with separate panels for Black Majority, Hispanic Majority, Other/No Majority, and White Majority counties."}
ggplot(data = counties, mapping = aes(x = ___, fill = ___)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Democratic Vote Percentage by Racial Majority",
       x = "Percent of Two-Party Vote",
       y = "Frequency",
       caption = "Source: Linn, Nagler, Zilinsky") +
  facet_wrap(___) +
  guides(fill = "none")
```

```{r histogram-dem2p-by-racialmajority-hint-1}
# Hint 1 of 3
# You want one panel per racial majority category, each
# showing how Democratic vote percentage is distributed
# within that group --- the same thing the GDP plot showed
# for regime types. Which variable varies along the x-axis,
# and which one separates the panels?
```

```{r histogram-dem2p-by-racialmajority-hint-2}
# Hint 2 of 3
# The interval variable is the one whose distribution you
# want to see, so it goes to x. The categorical one goes to
# fill, which colors each group and gives the panels their
# titles. The third blank belongs to facet_wrap(), and it
# takes the same categorical variable.

```

```{r histogram-dem2p-by-racialmajority-hint-3}
# Hint 3 of 3
# facet_wrap() wants a formula, not a bare name: write a
# tilde and then the variable, as in ~variable. Democratic
# vote percentage is dem2p_percent and the grouping variable
# is racial_majority.
```

```{r histogram-dem2p-by-racialmajority-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Your code did not produce a plot. Make sure the pipeline ends with a `ggplot()` call and that nothing is assigned to an object instead of being displayed.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing `+` at the end of a line, unmatched parentheses, or a label string missing its quotation marks.")
  }
  if (!grepl("counties", code_no_comments)) {
    fail("Use the **counties** data frame.")
  }
  if (!grepl("x\\s*=\\s*dem2p_percent", code_no_comments)) {
    fail("Map **dem2p_percent** to the x-axis inside `aes()`.")
  }
  if (!grepl("geom_histogram\\s*\\(", code_no_comments)) {
    fail("Use `geom_histogram()` to create the histogram.")
  }
  if (!grepl("fill\\s*=\\s*racial_majority", code_no_comments)) {
    fail("Map **racial_majority** to `fill` inside `aes()`.")
  }
  if (!grepl("facet_wrap\\s*\\(", code_no_comments)) {
    fail("Add a `facet_wrap()` layer to give each category its own panel.")
  }
  if (!grepl("facet_wrap\\s*\\(\\s*~\\s*racial_majority", code_no_comments)) {
    fail("Inside `facet_wrap()`, precede **racial_majority** with a tilde (~).")
  }
  if (!grepl("guides\\s*\\(\\s*fill\\s*=\\s*['\"]none['\"]", code_no_comments)) {
    fail('Add `guides(fill = "none")` to remove the redundant legend --- the panel titles already identify each group.')
  }

  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Distribution of Democratic Vote Percentage by Racial Majority")) {
    fail('Set `title = "Distribution of Democratic Vote Percentage by Racial Majority"` inside `labs()`.')
  }
  if (!identical(plot_labels$x, "Percent of Two-Party Vote")) {
    fail('Set `x = "Percent of Two-Party Vote"` inside `labs()`.')
  }
  if (!identical(plot_labels$y, "Frequency")) {
    fail('Set `y = "Frequency"` inside `labs()`.')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set `caption = "Source: Linn, Nagler, Zilinsky"` inside `labs()`.')
  }

  extra <- character(0)
  if (grepl("filter\\s*\\(", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Correct. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but neither of these variables has a missing value, so there was nothing to remove."))
  }
  pass("Well done. Mapping **racial_majority** to `fill` colors the bars by group, `facet_wrap()` gives each group its own panel, and `guides(fill = \"none\")` drops the legend the panel titles made redundant.")
})
```

The shared y-axis makes it easy to compare the number of counties across groups, but the Black Majority, Hispanic Majority, and Other/No Majority panels are so small relative to White Majority that their distributions are almost invisible. To see the shape of each group's distribution clearly, add `scales = "free_y"` inside `facet_wrap()`, as a second argument after the variable name:

```r
facet_wrap(~racial_majority, scales = "free_y")
```

This gives each panel its own y-axis scale so the bars fill the panel regardless of how many counties are in that group. The trade-off is that you can no longer compare bar heights across panels --- but you can now see the center and shape of each distribution. The plot below is the one you just built, with that single argument added.

```{r histogram-dem2p-racialmajority-freey, echo=FALSE, fig.alt="The same faceted histograms of Democratic vote percentage by racial majority, now with each panel using its own y-axis scale. The three smaller panels are stretched to the full height of their panels, so the shape of each distribution is visible, but bar heights can no longer be compared across panels."}
ggplot(data = counties, mapping = aes(x = dem2p_percent, fill = racial_majority)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Democratic Vote Percentage by Racial Majority",
       x = "Percent of Two-Party Vote",
       y = "Frequency",
       caption = "Source: Linn, Nagler, Zilinsky") +
  facet_wrap(~racial_majority, scales = "free_y") +
  guides(fill = "none")
```

Now each panel shows the shape of its own distribution clearly. When group sizes are roughly similar, the shared y-axis is usually the better choice --- it lets you compare both shape and group size in a single plot. Use `scales = "free_y"` when groups differ so dramatically in size that the smaller panels become unreadable at the shared scale, as happened here. In that situation, report group sizes separately --- the table of means already includes the count in each group, so nothing is lost.

### What Does the Histogram Add?

The histogram shows what a pair of group means cannot: the shape of each distribution. White Majority counties peak near 25 percent and trail off to the right, a long thin tail reaching counties that gave Clinton a large majority --- so the mean for that group sits below where the tail runs, pulled up a little by those cases rather than describing them. Black Majority counties peak near 65 percent and are far more symmetric. The two groups do overlap at the edges, but the bulk of each sits where the other has almost nothing, which is why the gap between their means is a fair summary of the difference. Read the Hispanic Majority and Other/No Majority panels more cautiously --- with 96 and 48 counties, their shapes are lumpy enough that what looks like structure may be small-sample noise.

## Does the Whole Distribution of College Graduation Rates Differ by Rurality?

Earlier you found that urban counties have notably higher mean college graduation rates than neither-urban-nor-rural or rural counties. A faceted histogram lets you check whether that difference reflects a shift in the whole distribution or something more subtle --- a difference in spread, shape, or just a few outliers pulling the mean.

Because **rural** is stored as numbers (0, 1, 2), the panels will come out titled 0, 1, and 2, which tells a reader nothing. Build it that way first and see for yourself --- the fix comes right after.

Recall as well that **prop_college_grad** is missing for one county while every county has a recorded **rural**.

**Practice 9: Compare the distribution of college graduation rates by rural**

Use **prop_college_grad** and **rural** from the **counties** data frame. Title the plot \"Distribution of College Graduation Rates by Rurality\", label the axes \"Proportion of College Graduates\" and \"Frequency\", caption it \"Source: Linn, Nagler, Zilinsky\", and omit the subtitle from `labs()` entirely. Three hints are available. Click "Submit Answer" to check your work.

```{r histogram-collegegrad-by-rural, exercise=TRUE, message=FALSE, exercise.cap="Practice 9", fig.alt="Faceted histograms showing the distribution of the proportion of college graduates on the x-axis and frequency on the y-axis, with three panels titled 0, 1, and 2 --- the stored values of the rurality variable rather than readable names."}

```

```{r histogram-collegegrad-by-rural-hint-1}
# Hint 1 of 3
# You want a separate view of the college graduation
# distribution for each rurality category --- the same shape
# as the racial majority histogram. One variable supplies the
# values along the horizontal axis; the other divides counties
# into panels. Two things differ from that example: one of
# these variables has a missing value, and rural is stored as
# a number rather than as text.
```

```{r histogram-collegegrad-by-rural-hint-2}
# Hint 2 of 3
# Numeric variables need factor() when mapped to fill so
# ggplot treats them as discrete categories rather than
# a continuous scale. facet_wrap() does not need it.
```

```{r histogram-collegegrad-by-rural-hint-3}
# Hint 3 of 3
# The legend repeats what the panel titles already say, so
# add guides(fill = "none") to drop it. Then check the four
# labs() strings against the ones named in the prompt ---
# they are compared exactly, punctuation included.
```

```{r histogram-collegegrad-by-rural-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Your code did not produce a plot. Make sure the pipeline ends with a `ggplot()` call and that nothing is assigned to an object instead of being displayed.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string missing its quotation marks.")
  }

  # Check data frame
  if (!any(grepl("\\bcounties\\b", .user_code))) {
    fail("Use the **counties** data frame.")
  }

  # Check aesthetics from plot object
  get_aes_name <- function(plot, aes_name) {
    out <- character(0)
    global_mapping <- plot$mapping[[aes_name]]
    if (!is.null(global_mapping)) {
      out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(global_mapping)), error = function(e) NA_character_))
    }
    for (layer in plot$layers) {
      layer_mapping <- layer$mapping[[aes_name]]
      if (!is.null(layer_mapping)) {
        out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(layer_mapping)), error = function(e) NA_character_))
      }
    }
    out[!is.na(out)]
  }

  x_vars <- get_aes_name(.result, "x")
  if (!"prop_college_grad" %in% x_vars) {
    fail("Map **prop_college_grad** to the x-axis inside `aes()`.")
  }

  # Check fill requires factor(rural) --- not bare rural
  fill_code <- any(grepl("fill\\s*=\\s*factor\\s*\\(\\s*rural\\s*\\)", code_no_comments))
  if (!fill_code) {
    fail("Because **rural** is stored as a number, wrap it in `factor()` when mapping it to `fill`: `fill = factor(rural)`. Without `factor()`, ggplot treats **rural** as continuous and applies a color gradient rather than discrete group colors.")
  }

  # Check layer types
  layer_geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  layer_stats <- sapply(.result$layers, function(l) class(l$stat)[1])
  if (!"GeomBar" %in% layer_geoms || !"StatBin" %in% layer_stats) {
    fail("Use `geom_histogram()` to create the histograms. A bar layer built another way will not bin the interval variable.")
  }

  # Check faceting uses rural
  if (!any(grepl("facet_wrap", .user_code))) {
    fail("Add `facet_wrap()` to give each rurality category its own panel.")
  }
  if (!any(grepl("facet_wrap\\s*\\(\\s*~\\s*(factor\\s*\\(\\s*)?rural", code_no_comments))) {
    fail("Inside `facet_wrap()`, use `~rural` (or `~factor(rural)`) so each panel corresponds to a rurality category.")
  }
  if (!grepl('guides\\s*\\(\\s*fill\\s*=\\s*["\']none["\']', code_no_comments)) {
    fail('Add `guides(fill = "none")` to remove the redundant legend.')
  }

  # Check plot labels from object
  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Distribution of College Graduation Rates by Rurality")) {
    fail('Set `title = "Distribution of College Graduation Rates by Rurality"` inside `labs()`.')
  }
  if (!identical(plot_labels$x, "Proportion of College Graduates")) {
    fail('Set `x = "Proportion of College Graduates"` inside `labs()`.')
  }
  if (!identical(plot_labels$y, "Frequency")) {
    fail('Set `y = "Frequency"` inside `labs()`.')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set `caption = "Source: Linn, Nagler, Zilinsky"` inside `labs()`.')
  }
  if (!is.null(plot_labels$subtitle)) {
    fail("Omit the subtitle from `labs()` --- the prompt asks for the title, axis labels, and caption only.")
  }

  extra <- character(0)
  if (grepl("!\\s*is\\.na\\s*\\(", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Correct. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but a histogram has no bar to give a missing value, so ggplot drops those rows itself and tells you how many."))
  }
  pass("Nicely done. Because **rural** is stored as numbers you wrapped it in `factor()` for the fill, and `facet_wrap()` gave each rurality category a panel of its own.")
})
```

Those panel titles are the stored values, and they are no use to a reader. The `labeller` argument inside `facet_wrap()` replaces them. It takes a function rather than plain text, and `as_labeller()` builds one from a named vector pairing each stored value with the label you want:

```r
facet_wrap(~rural, labeller = as_labeller(c("0" = "Urban",
                                            "1" = "Neither Urban nor Rural",
                                            "2" = "Rural")))
```

The names on the left must match the stored values; the text on the right supplies the panel titles. Because the matching is done by name, the entries do not have to appear in any particular order --- but each stored value has to be paired with the label that belongs to it. Here is the plot you just built with that one argument added.

```{r histogram-collegegrad-by-rural-labelled, echo=FALSE, message=FALSE, fig.alt="The same three faceted histograms of college graduation rates, now titled Urban, Neither Urban nor Rural, and Rural instead of 0, 1, and 2. All three are right-skewed and peak near a proportion of 0.2. The Neither Urban nor Rural panel is much the tallest, reaching about 270 counties; Urban peaks near 130 and has the longest tail to the right; Rural peaks near 110 with the shortest tail."}
ggplot(data = counties, mapping = aes(x = prop_college_grad, fill = factor(rural))) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of College Graduation Rates by Rurality",
       x = "Proportion of College Graduates",
       y = "Frequency",
       caption = "Source: Linn, Nagler, Zilinsky") +
  facet_wrap(~rural, labeller = as_labeller(c("0" = "Urban",
                                              "1" = "Neither Urban nor Rural",
                                              "2" = "Rural"))) +
  guides(fill = "none")
```

### What Does the Histogram Add?

**Apply 10: Interpret the rural histogram**


```{r quiz-histogram-collegegrad-rural, echo=FALSE}
question("The table of means told you urban counties have higher average college graduation rates. What does the histogram reveal that the means could not?",
  correct = "Right. The histogram shows the urban advantage is about spread as well as center.",
  answer("Urban counties have a much wider spread as well as a higher average, while neither-urban-nor-rural counties have the most tightly concentrated distribution",
         correct = TRUE,
         message = "The means told you urban counties are higher on average. The histogram adds that urban counties are also far more variable --- the long right tail means some urban counties have extremely high graduation rates that pull the mean up, while the neither-urban-nor-rural panel is remarkably narrow, suggesting that their college-graduation rates are relatively similar. The rural panel sits between them, with the shortest tail to the right of the three."),
  answer("The histogram confirms the means exactly --- urban is highest, rural is lowest, and neither-urban-nor-rural falls in between with no additional information",
         message = "The histogram does confirm the ordering from the means, but it also shows something the means could not: how spread out each distribution is. The shape and width of each panel reveal whether the difference is about a uniformly higher center or about a wider range of outcomes."),
  answer("Neither-urban-nor-rural and rural have nearly identical shapes, suggesting the means overstated the difference between them",
         message = "Look closely at the width of each panel. Neither-urban-nor-rural is noticeably taller and narrower than rural --- the two distributions are not nearly identical in shape even though their means are close."),
  answer("The histogram is less informative than the means table because it does not show exact values",
         message = "The histogram shows something the means table cannot: the shape and spread of each distribution. Knowing that urban counties have a long right tail, for example, tells you something about which counties are driving the urban mean upward that no table of means can reveal."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The means already told you which group is highest on average. Ask what the panels show that a single number per group cannot --- look at the width of each distribution and the length of its tails."
)
```

## Visualizing Distributions Across Categories: Grouped Box Plots

A [grouped box plot]{.important-text} says much the same thing in far less space. The box covers the middle 50% of cases, with a line inside it marking the median. The whiskers reach out to the most extreme values that are not flagged as outliers, and anything past them is drawn as its own point. Tutorial 6 introduces box plots for a single interval variable. Set several of them side by side and you can compare medians, spreads, and outliers across groups in one glance.

The layer that draws the boxes is `geom_boxplot()`, in place of the `geom_histogram()` you used above. The grouping variable goes on `x`, giving one box per category, and the interval variable goes on `y`. Mapping the grouping variable to `fill` as well colors each box; since the x-axis already labels them, `guides(fill = "none")` drops the legend for the same reason as before.

That is the second time a legend has been redundant, for two different reasons: in the histogram the panel titles already named the groups, and here the x-axis does. The general rule covers both --- whenever the fill variable is identified somewhere else in the plot, `guides(fill = "none")` takes the legend away.

The `fill` aesthetic works in two ways. Set outside `aes()`, as `fill = "firebrick"`, it gives every element the same fixed color. Mapped inside `aes()` to a variable, as `fill = dem_level4`, it gives each group its own color. Here it is mapped to the grouping variable, so each category's box is visually distinct.

**Run and observe 11: The same four groups as box plots**

```{r boxplot-gdp-by-demlevel, exercise=TRUE, exercise.cap="Run and observe 11", fig.alt="Box plots of GDP per capita by regime type. Full Democracy has a high narrow box with two low outliers and two high ones. Part Democracy sits lower with a wider box. Hybrid has a very low narrow box. Authoritarian has a low box that contains Hybrid's entirely, with four separate points above it, the highest past eighty-five thousand. Part Democracy's box is the widest of the four."}
ggplot(data = world, mapping = aes(x = dem_level4, y = gdppcap08, fill = dem_level4)) +
  geom_boxplot() +
  labs(title = "Distribution of GDP per Capita by Level of Democracy",
       x = "Level of Democracy",
       y = "GDP per Capita, 2008 (US$)",
       caption = "Source: World Bank; Economist Intelligence Unit") +
  guides(fill = "none")
```

Now put each box next to the mean you calculated for that group. For Full Democracies the two agree: a mean of \$35,891 against a median of \$35,677, so the line in the box sits essentially where the mean does and the mean is a fair summary of a typical country. Authoritarian regimes are the opposite. Their mean is \$9,412 and their median is \$4,388 --- less than half of it. The four countries drawn as separate points at the top of that column have pulled the mean well clear of the median, up into the top half of the box. Read the box instead and the picture holds steady: half of these countries fall between \$1,455 and \$11,315, and the typical one is among the poorest here.

That has consequences for the hypothesis. On the means, Authoritarian regimes looked better off than Hybrid ones --- \$9,412 against \$3,985. On the medians the gap is far smaller, \$4,388 against \$2,644, and Hybrid's box sits entirely inside the lower part of Authoritarian's. What looked like a difference between two regime types is largely a handful of resource-rich states inside one of them.

## Do More Democratic Countries Have Higher Life Expectancy?

Life expectancy is built out of unglamorous things: clean water, childhood vaccination, roads that do not kill people, clinics that are open and staffed. Because these accumulate over decades and reach nearly everyone, average life expectancy is one of the better available measures of whether a state actually delivers to its population rather than to a favored part of it.

The theoretical argument is the same accountability logic as in the previous example, applied to a different outcome --- and it fits this one more tightly. Public health goods benefit many people a little rather than a few people a great deal, which makes them easy for an unaccountable leader to skip and politically expensive for an elected one to neglect. GDP can be raised by enriching a narrow sector; life expectancy cannot.

If that logic holds:

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have a [higher life expectancy] than will those that are [less democratic].
:::

The data frame **world** contains the interval variable **lifeex_total**, life expectancy at birth, and the ordinal variable **dem_level4**, the same regime-type variable you used to compare GDP per capita. **lifeex_total** is missing for 2 countries, so any summary you compute from it describes the 165 countries with a recorded value rather than all 167 in the data. Because **dem_level4** is already stored as a factor, you do not need to wrap it in `factor()` when mapping it inside `aes()`.

The stored values of **dem_level4** are abbreviated: `"Full Democ"`, `"Part Democ"`, `"Hybrid"`, `"Authoritarian"`. Use `scale_x_discrete()` after `labs()` to pair each stored value with a readable label: \"Full Democ\" with \"Full Democracy\", \"Part Democ\" with \"Partial Democracy\", and \"Hybrid\" and \"Authoritarian\" with themselves. This is the same named-vector idiom `as_labeller()` used for the facet titles earlier, and writing the pairs by name means the order you list them in does not matter.

**Practice 12: Compare the distribution of life expectancy by democracy level**

Fill in the three blanks in `aes()`, then add the two layers after `labs()`. Three hints are available. Click "Submit Answer" to check your work.


```{r boxplot-lifeex-by-democracylevel, exercise=TRUE, exercise.cap="Practice 12", fig.alt="Box plots showing life expectancy at birth on the y-axis and democracy level on the x-axis, with separate boxes for Full Democracy, Partial Democracy, Hybrid, and Authoritarian regimes."}
ggplot(data = world, mapping = aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot() +
  labs(title = "Distribution of Life Expectancy by Level of Democracy",
       x = "Level of Democracy",
       y = "Life Expectancy at Birth",
       caption = "Source: CIA; Economist Intelligence Unit") +
```







```{r boxplot-lifeex-by-democracylevel-hint-1}
# Hint 1 of 3
# You want one box per regime type, each showing how life
# expectancy is distributed within that group, so that the
# medians and the spreads can be compared side by side.
# Which of these two variables defines the groups, and
# which holds the values being compared across them?
```


```{r boxplot-lifeex-by-democracylevel-hint-2}
# Hint 2 of 3
# Only one of the two variables has missing values --- the
# paragraph above says which --- and that is the one the
# filter condition names. Inside aes(), the variable that
# defines the groups goes to both x and fill, and the one
# whose distribution you are comparing goes to y.
```

```{r boxplot-lifeex-by-democracylevel-hint-3}
# Hint 3 of 3
# Two layers still come after labs(). scale_x_discrete()
# takes a labels argument pairing each stored value with
# its readable name, as in "Part Democ" = "Partial
# Democracy". Then drop the redundant legend with guides().
```

```{r boxplot-lifeex-by-democracylevel-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Your code did not produce a plot. Make sure the pipeline ends with a `ggplot()` call and that nothing is assigned to an object instead of being displayed.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string missing its quotation marks.")
  }

  if (!any(grepl("\\bworld\\b", .user_code))) {
    fail("Use the **world** data frame.")
  }

  # Check aesthetics from plot object
  get_aes_name <- function(plot, aes_name) {
    out <- character(0)
    global_mapping <- plot$mapping[[aes_name]]
    if (!is.null(global_mapping)) {
      out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(global_mapping)), error = function(e) NA_character_))
    }
    for (layer in plot$layers) {
      layer_mapping <- layer$mapping[[aes_name]]
      if (!is.null(layer_mapping)) {
        out <- c(out, tryCatch(rlang::as_name(rlang::get_expr(layer_mapping)), error = function(e) NA_character_))
      }
    }
    out[!is.na(out)]
  }

  x_vars <- get_aes_name(.result, "x")
  y_vars <- get_aes_name(.result, "y")
  fill_vars <- get_aes_name(.result, "fill")

  if (!"dem_level4" %in% x_vars) {
    fail("Map **dem_level4** to `x` inside `aes()` to place regime types on the x-axis.")
  }
  if (!"lifeex_total" %in% y_vars) {
    fail("Map **lifeex_total** to `y` inside `aes()` to show life expectancy on the y-axis.")
  }
  if (!"dem_level4" %in% fill_vars) {
    fail("Map **dem_level4** to `fill` inside `aes()` so each regime category defines a separate box.")
  }

  # Check layer types
  layer_geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  if (!"GeomBoxplot" %in% layer_geoms) {
    fail("Use `geom_boxplot()` to create the box plots.")
  }

  # Check scale_x_discrete pairs every stored value with its readable label
  x_scales <- Filter(function(sc) "x" %in% sc$aesthetics, .result$scales$scales)
  if (length(x_scales) == 0) {
    fail("Add `scale_x_discrete()` to replace the abbreviated stored values with readable labels.")
  }
  expected_labels <- c("Full Democ" = "Full Democracy",
                       "Part Democ" = "Partial Democracy",
                       "Hybrid" = "Hybrid",
                       "Authoritarian" = "Authoritarian")
  if (!identical(x_scales[[1]]$labels[names(expected_labels)], expected_labels)) {
    fail('Pair each stored value with its readable label, as in "Part Democ" = "Partial Democracy". All four categories need a label.')
  }
  if (!grepl('guides\\s*\\(\\s*fill\\s*=\\s*["\']none["\']', code_no_comments)) {
    fail('Add `guides(fill = "none")` to remove the redundant fill legend.')
  }

  extra <- character(0)
  if (grepl("!\\s*is\\.na\\s*\\(", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Correct. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but a box plot has no place on its axis for a missing value, so ggplot drops those rows itself and tells you how many."))
  }
  pass("Correct. The categorical variable goes to both `x` and `fill`, the interval variable to `y`, and `scale_x_discrete()` replaces the abbreviated stored values with readable regime labels while preserving their order.")
})
```


### What Does the Box Plot Add?

The hypothesis predicts a clean ordering from full democracy to authoritarian rule. The box plot gives you more information than the means alone --- look at the full picture before deciding how well the pattern matches.

**Apply 13: Interpret the life expectancy box plot**

The line inside each box marks the **median**, not the mean, so the box plot does not display the group means from the earlier table --- though for roughly symmetric distributions the two will be close.

```{r quiz-boxplot-lifeex-democracylevel, echo=FALSE}
question("The table of means told you full democracies have higher average life expectancy. What does the box plot reveal that the means could not?",
  correct = "Right. The medians decline in order, but the groups are not cleanly separated.",
  answer("Hybrid and authoritarian regimes overlap substantially, while partial democracies have several low outliers",
         correct = TRUE,
         message = "The means suggested a monotonic decline, but the box plot shows the bottom of the ordering is much messier than the top. Hybrid and authoritarian regimes overlap so substantially that the difference between them is weak. Partial democracies have several unusually low outliers that pull their picture down. Full democracy stands clearly apart from the rest, but the other three categories do not form a clean hierarchy."),
  answer("The box plot confirms the means exactly --- life expectancy declines monotonically from full democracy to authoritarian rule with clear separation between all four groups",
         message = "Look carefully at the hybrid and authoritarian boxes. Their medians are close and their boxes overlap substantially --- there is not clear separation between all four groups the way the means might suggest."),
  answer("Full democracies have the widest spread of all four groups, suggesting their high mean is driven by a few outliers rather than a uniformly high life expectancy",
         message = "Look at the width of the full democracy box compared to the others. Full democracy actually has the narrowest box --- its high life expectancy is fairly consistent across countries, not driven by a few outliers."),
  answer("Authoritarian regimes have the lowest median and the most outliers, confirming they are the most variable group",
         message = "Look at the authoritarian box plot carefully. While authoritarian regimes do have the lowest median, their whiskers extend to cover most of their range without flagging outliers --- it is partial democracy, not authoritarian rule, that shows the most outlier points."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The means already told you which group is highest on average. Look at where the boxes and whiskers overlap, and at which groups become hard to tell apart once you can see their full spread."
)
```

## Does a Box Plot Sharpen the Racial Majority Comparison?

Earlier you compared these same counties two ways: a table of means, and then a faceted histogram showing the whole distribution within each racial majority category. A box plot gives a third view, and a more compact one. Neither **dem2p_percent** nor **racial_majority** has a missing value.

**Practice 14: Compare the distribution of Democratic vote percentage by racial majority**

Build a box plot comparing the two. Title it \"Democratic Vote Percentage by Racial Majority\", label the y-axis \"Percent of Two-Party Vote\", caption it \"Source: Linn, Nagler, Zilinsky\", and omit the subtitle from `labs()` entirely. Set the x-axis label to `NULL` --- the stored values of **racial_majority** are already readable, so the tick labels already identify all four groups, so an axis title saying "Racial Majority" only repeats them. Three hints are available. Click "Submit Answer" to check your work.

```{r boxplot-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice 14", fig.alt="Box plots showing Democratic vote percentage on the y-axis and racial majority category on the x-axis, with separate boxes for Black Majority, Hispanic Majority, Other/No Majority, and White Majority counties."}

```

```{r boxplot-dem2p-by-racialmajority-hint-1}
# Hint 1 of 3
# Use the same structure as the life expectancy box plot,
# minus the step that dealt with its missing values ---
# neither variable here has any. Think about which variable
# defines the groups and which holds the values you are
# comparing across groups.
```

```{r boxplot-dem2p-by-racialmajority-hint-2}
# Hint 2 of 3
# The categorical variable goes on x and fill.
# The interval variable goes on y.
# Because racial_majority is stored as a character variable,
# no factor() wrapper is needed and no scale_x_discrete()
# is needed --- the stored values are already readable.
```

```{r boxplot-dem2p-by-racialmajority-hint-3}
# Hint 3 of 3
# The x-axis already names each group, so the fill legend repeats
# what the plot has already shown --- guides(fill = "none") removes it.
# In labs(), set x to NULL and use the exact title, y-axis label, and
# caption given in the prompt, with no subtitle.
```

```{r boxplot-dem2p-by-racialmajority-check}
grade_this({
  if (is.null(.result)) {
    fail("Your code did not produce a plot. Make sure the pipeline ends with a `ggplot()` call and that nothing is assigned to an object instead of being displayed.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing `+` at the end of a line or unmatched parentheses.")
  }

  code_no_comments <- paste(gsub("#[^\r\n]*", "", .user_code), collapse = "\n")
  if (!grepl("\\bcounties\\b", code_no_comments)) {
    fail("Use the **counties** data frame.")
  }

  plot_mapping <- .result$mapping
  x_mapping   <- rlang::as_label(plot_mapping$x)
  y_mapping   <- rlang::as_label(plot_mapping$y)
  fill_mapping <- rlang::as_label(plot_mapping$fill)

  if (!identical(x_mapping, "racial_majority")) {
    fail("Map **racial_majority** to `x` inside `aes()`.")
  }
  if (!identical(y_mapping, "dem2p_percent")) {
    fail("Map **dem2p_percent** to `y` inside `aes()`.")
  }
  if (!identical(fill_mapping, "racial_majority")) {
    fail("Map **racial_majority** to `fill` inside `aes()`.")
  }

  has_boxplot <- any(vapply(.result$layers, function(l) inherits(l$geom, "GeomBoxplot"), logical(1)))
  if (!has_boxplot) {
    fail("Use `geom_boxplot()` to create the box plots.")
  }

  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Democratic Vote Percentage by Racial Majority")) {
    fail('Set `title = "Democratic Vote Percentage by Racial Majority"` inside `labs()`.')
  }
  if (!is.null(plot_labels$x)) {
    fail('Set `x = NULL` inside `labs()` --- the tick labels already name each group, so an axis title repeats them.')
  }
  if (!identical(plot_labels$y, "Percent of Two-Party Vote")) {
    fail('Set `y = "Percent of Two-Party Vote"` inside `labs()`.')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set `caption = "Source: Linn, Nagler, Zilinsky"` inside `labs()`.')
  }
  if (!is.null(plot_labels$subtitle)) {
    fail("Omit the subtitle from `labs()` --- the prompt asks for the title, axis labels, and caption only.")
  }

  if (!grepl('guides\\s*\\(\\s*fill\\s*=\\s*["\']none["\']', code_no_comments)) {
    fail('Add `guides(fill = "none")` to remove the redundant legend --- the x-axis already names each group.')
  }
  extra <- character(0)
  if (grepl("filter\\s*\\(", code_no_comments)) extra <- c(extra, "a filter")
  if (grepl("na\\.rm", code_no_comments)) extra <- c(extra, "`na.rm = TRUE`")
  if (length(extra) > 0) {
    pass(paste0("Good work --- the plot is right. You included ", paste(extra, collapse = " and "),
                ", which does no harm --- but neither of these variables has a missing value, so there was nothing to remove."))
  }
  pass("Good work. You built the same box plot structure on a character variable this time, so no `factor()` was needed and the x-axis labels came straight from the stored values.")
})
```

### What Does the Box Plot Add?

**Apply 15: Interpret the racial majority box plot**

```{r quiz-boxplot-dem2p-racialmajority, echo=FALSE}
question("The table of means shows Black Majority counties at 65.2% and Hispanic Majority counties at 46.3%. What does the box plot add that those two numbers do not?",
  correct = "Right. One mean summarizes its group well; the other barely summarizes anything.",
  answer("Black Majority counties cluster tightly around their mean, while Hispanic Majority counties span nearly the full range",
         correct = TRUE,
         message = "The Black Majority box runs from about 57 to 70, so the middle half of those counties sits inside roughly 13 points and 65.2 describes them well. The Hispanic Majority box runs from about 28 to 65 --- an interquartile range near 37 points. Their mean of 46.3 falls in the middle of that range without describing many counties in it: the middle half alone runs from clearly Republican-leaning to clearly Democratic-leaning. A mean summarizes a group well when the group is concentrated and poorly when it is not, and the box plot is what tells you which case you are looking at."),
  answer("Black Majority and Hispanic Majority counties have similar interquartile ranges --- their boxes are roughly the same width",
         message = "Look carefully at the widths of the two boxes. The Black Majority box spans roughly 13 percentage points while the Hispanic Majority box spans roughly 37 --- the interquartile ranges are very different."),
  answer("The many high outliers in White Majority counties suggest those counties are a genuine swing group that could vote either way",
         message = "The outliers in White Majority counties are a small number of heavily urban counties, not evidence that White Majority counties as a whole are competitive. The box itself is narrow and concentrated at low Democratic vote percentages, which better characterizes the spread of the group."),
  answer("Hispanic Majority counties have the lowest median, which explains why their interquartile range is wider than Black Majority counties",
         message = "Spread and center are separate features of a distribution. A group can have a high or low median and still have a wide or narrow IQR --- they are independent. The Hispanic Majority median is around 45%, well above White Majority, but the IQR is wide regardless of where the median falls."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The two means differ by nearly 19 points, so this is not about which group is more Democratic. Compare the widths of those two boxes --- each spans the middle 50% of its group."
)
```

## When Should You Use a Histogram vs. a Box Plot?

Histograms and box plots both show the distribution of an interval variable across categories, but they reveal different things and involve different trade-offs. A histogram shows the full shape of each group's distribution --- whether it is symmetric or skewed, whether it has one peak or several, and roughly how the values are spread across the range. That detail comes at a cost: histograms become increasingly cumbersome as the number of groups grows, since many faceted panels are difficult to compare at a glance.

A box plot sacrifices that fine-grained shape for efficiency. It compactly summarizes each group using the median, the middle 50% (IQR), whiskers extending to the most extreme non-outlier values, and any outlier points --- making it easy to compare many groups side by side and to quickly spot differences in central tendency, spread, and outliers. What a box plot cannot show you is whether a distribution is bimodal, or exactly how skewed it is; two very differently shaped distributions can produce nearly identical-looking boxes. The rule of thumb: use a histogram when you want to see the full shape of each group's distribution, especially when you have only a few groups. Use a box plot when you want to compare many groups efficiently, or when you care most about the median, spread, and outliers rather than fine-grained shape.

**Apply 16: Choose the right visualization**


```{r quiz-histogram-vs-boxplot, echo=FALSE}
question("A researcher wants to compare the distribution of household income across the 50 U.S. states. Which tool would be the better choice, and why?",
  correct = "Right. A box plot scales to 50 groups far better than a faceted histogram.",
  answer("A box plot, because 50 panels of a faceted histogram would be unreadable",
         correct = TRUE,
         message = "With this many groups, the fine-grained shape a histogram shows becomes hard to use --- 50 panels are difficult to scan and compare. A box plot's compact summary scales much better to many groups while still showing medians, spreads, and outliers."),
  answer("A histogram, because it shows more detail about each state's distribution than a box plot would",
         message = "A histogram does show more detail, but that level of detail becomes a liability with 50 groups --- 50 histogram panels are too many to compare effectively. A box plot trades some detail for the ability to compare many groups at once."),
  answer("It does not matter, since histograms and box plots always show the same information",
         message = "Histograms and box plots reveal different features of a distribution. A histogram shows the full shape, including skew and modality; a box plot summarizes that shape more compactly and loses some detail in exchange for being easier to compare across many groups."),
  answer("A box plot, but only because income distributions are always skewed --- if income were symmetric, a histogram would be the better choice",
         message = "The number of groups, not the shape of the distribution, is the main reason to prefer a box plot here. With 50 states, 50 histogram panels are too many to compare effectively regardless of whether the distributions are skewed or symmetric."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Apply the rule of thumb from the section above. Think about how many groups are being compared and how readable each display stays as that number grows."
)
```

## Check Your Understanding

Now let's step back and practice the full workflow with new examples. You do not need to write any code. Focus on the logic: decide whether the research question fits this tutorial's setup, compare means across categories, and use plots to see what the means alone might hide.

**Apply 17: Classify the research question**


```{r question-forward-reverse, echo=FALSE}
question("Which of the following research questions fits the pattern covered in this tutorial --- a categorical explanatory variable and an interval outcome --- rather than the reverse case?",
  correct = "Right. Region is categorical and commute time is interval --- exactly the setup this tutorial covers.",
  answer("Do commute times differ by region (Northeast, South, Midwest, West)?",
         correct = TRUE,
         message = "Region is categorical and commute time is interval, with commute time as the outcome --- exactly the pattern this tutorial covers. Two of the alternatives flip the direction, pairing an interval explanatory variable with a categorical outcome, and the civil-war question has two categorical variables --- each needs a different approach."),
  answer("Is a student's SAT score associated with whether they are admitted to a selective college?",
         message = "Here SAT score (interval) is the explanatory variable and admission (categorical) is the outcome --- not the setup this tutorial focuses on. You could compare mean SAT scores between admitted and non-admitted students --- that describes an association --- but it does not directly address how the probability of admission changes as SAT scores rise, which requires a different tool."),
  answer("Is a county's median household income associated with whether it voted for the incumbent party?",
         message = "Here income (interval) is the explanatory variable and vote choice (categorical) is the outcome --- not the setup this tutorial focuses on. You could compare mean income across voting groups --- that describes an association --- but it does not directly address how the probability of voting for the incumbent changes with income, which requires a different approach."),
  answer("Does a country's regime type predict its probability of experiencing civil war?",
         message = "Here regime type (categorical) is the explanatory variable and civil war occurrence (categorical --- yes or no) is the outcome --- not the setup this tutorial focuses on. When the outcome is categorical, comparing group means is not the right approach."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "For each option, work out which variable is the outcome and what its level of measurement is. Only one option has an interval outcome compared across categorical groups."
)
```

Commuting is the part of American infrastructure policy most people experience daily, and how long it takes depends less on distance than on the shape of the metropolitan area a trip happens in.

The theoretical argument is about density and vintage together. The Northeast urbanized before the automobile, so its metropolitan areas run on street grids and rail corridors laid out for a different century, and they concentrate more people into that fixed capacity than newer regions do into theirs. Old infrastructure is also harder to widen than new infrastructure is to build.

Which gives something to check:

::: {.hypothesis}
In a comparison of [cities], those located in the [Northeast] will have [longer daily commute times] than will those located in [other regions of the country].
:::

The table below shows the mean daily commute time, in minutes, for a sample of cities in four regions.

| Region    | Mean Commute Time (minutes) | N  |
|-----------|------------------------------|----|
| Northeast | 34.2                          | 48 |
| Midwest   | 24.6                          | 52 |
| South     | 26.1                          | 61 |
| West      | 28.9                          | 45 |

**Apply 18: Evaluate the commute-time hypothesis**


```{r question-commute-table, echo=FALSE}
question("A researcher hypothesizes that cities in the Northeast have longer commute times than cities in the other three regions, due to denser populations and older transit infrastructure. Based on the table, which statement best describes the evidence for this hypothesis?",
  correct = "Right. The Northeast stands clearly apart while the other three regions are close to each other.",
  answer("The pattern is consistent with the hypothesis --- the Northeast's mean is several minutes higher than any other region",
         correct = TRUE,
         message = "The Northeast's mean (34.2) is at least 5 minutes higher than any other region, while the Midwest, South, and West are all within about 4 minutes of each other. This is consistent with the hypothesis, but this is a descriptive comparison, not a formal test."),
  answer("The pattern is consistent with the hypothesis --- commute times step down by about the same amount from each region to the next",
         message = "The direction is right, but the steps are very uneven. From the Northeast (34.2) to the West (28.9) is 5.3 minutes; from the West to the South (26.1) is 2.8; from the South to the Midwest (24.6) is 1.5. Almost all of the difference comes from the Northeast standing apart, not from an even decline --- the same shape you saw in the rural college-graduation example."),
  answer("The hypothesis is confirmed, since the Northeast clearly has the highest mean commute time",
         message = "A descriptive comparison of means cannot confirm a hypothesis, no matter how large the difference looks. Confirming it would require a formal hypothesis test."),
  answer("The pattern is not consistent with the hypothesis, since commute times in all four regions are fairly similar",
         message = "Look at the actual numbers again. The Northeast's mean is several minutes higher than the other three regions, which are much closer to each other than any of them are to the Northeast."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Read all four regional means, not only the Northeast. Then ask what a comparison of means on its own can and cannot establish."
)
```

The same researcher now produces box plots of commute time for each region, shown below.

```{r cyu-boxplot-commute, echo=FALSE, fig.alt="Box plots showing daily commute time in minutes on the y-axis and region on the x-axis. The Northeast box is tall, ranging from about 30 to 40 minutes, with three high outliers between 60 and 70 minutes. The Midwest, South, and West boxes are all short and similarly sized, roughly 23 to 30 minutes, with no outliers."}
set.seed(7)
northeast <- c(rnorm(40, mean = 33, sd = 8), 62, 65, 68)
midwest <- rnorm(45, mean = 24.5, sd = 2.5)
south <- rnorm(50, mean = 26, sd = 2.5)
west <- rnorm(40, mean = 29, sd = 2.7)

commute_df <- data.frame(
  region = factor(c(rep("Northeast", length(northeast)),
                     rep("Midwest", length(midwest)),
                     rep("South", length(south)),
                     rep("West", length(west))),
                   levels = c("Northeast", "Midwest", "South", "West")),
  commute = c(northeast, midwest, south, west)
)

ggplot(commute_df, aes(x = region, y = commute, fill = region)) +
  geom_boxplot() +
  labs(title = "Distribution of Commute Time by Region",
       x = "Region",
       y = "Daily Commute Time (minutes)",
       caption = "Hypothetical data") +
  guides(fill = "none")
```

**Apply 19: Interpret the commute-time box plots**


```{r question-commute-boxplot, echo=FALSE}
question("What does this plot tell you that the table of means alone could not?",
  correct = "Right. The box plot reveals spread and outliers that the means alone could not show.",
  answer("The Northeast has a much wider spread of commute times and several unusually high outliers",
         correct = TRUE,
         message = "The Northeast's box spans roughly 30 to 40 minutes, more than three times the width of the other regions' boxes, and it has three outlier cities above 60 minutes --- features completely invisible in a table that only reports means. The means told you the Northeast is higher on average; the box plot reveals it is also far more variable."),
  answer("This information is redundant with the table of means and tells you nothing new",
         message = "The table of means only tells you the Northeast's typical commute time is higher. The box plot reveals that the Northeast is also far more variable, with outlier cities --- information the means alone do not provide."),
  answer("All four regions have similar spreads --- the Northeast's box only looks taller because its median is higher",
         message = "Spread and center are separate features. The Northeast's box runs from about 30 to 41 minutes, while the Midwest, South, and West boxes each span only about three minutes. The Northeast is genuinely more variable, not merely shifted upward."),
  answer("The Northeast's higher mean is caused entirely by its three outlier cities",
         message = "The outliers do pull the mean up, but they are not the whole story. The Northeast's box runs from about 30 to 41 minutes and its median sits near 35, both well above the other regions --- the difference would survive removing those three cities."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The table already gave you the average for each region. Ask what the boxes, whiskers, and individual points add about spread and unusual cases that a single average per region hides."
)
```


## The Takeaways

Comparing [group means]{.important-text} is where [bivariate description]{.important-text} of a categorical and an interval variable begins --- asking whether the distribution of an interval outcome differs across the categories of a categorical explanatory variable. The table below adds that combination to the one from Tutorial 7. Read across the columns rather than down: the question is the same in both --- does the pattern point in the direction the hypothesis predicts? --- and what changes is the tool that answers it.

::: {.qpa-infographic}
<table class="qpa-table">
  <caption>Describing Categorical and Interval Variables</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Aspect of the relationship</span></th>
      <th scope="col">Two Categorical Variables</th>
      <th scope="col">One Categorical, One Interval Variable</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Numerical Summary</th>
      <td>Cross-tabulation with Column Proportions</td>
      <td>Group Means and Group Sizes</td>
    </tr>
    <tr>
      <th scope="row">Visualization</th>
      <td>Grouped Bar Plot of Proportions</td>
      <td>Faceted Histogram, Box Plot</td>
    </tr>
    <tr>
      <th scope="row">What to Look For</th>
      <td>Differences in Column Proportions Across Categories</td>
      <td>Differences in Center, Spread, and Shape Across Categories</td>
    </tr>
    <tr>
      <th scope="row">R Functions</th>
      <td><code>CrossTable()</code>, <code>geom_bar()</code></td>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>geom_histogram()</code>, <code>facet_wrap()</code>, <code>geom_boxplot()</code></td>
    </tr>
  </tbody>
</table>
:::

The direction of the setup matters: the interval variable is the outcome, and the categorical variable defines the groups. Reversing the two answers a different question. Filter missing values from the interval outcome before building the table, so that `freq = n()` and `mean()` describe exactly the same cases.

**What a table of means hides** is the reason to go further. Two groups can have similar means and very different spreads, and a group's mean can sit well away from where most of its cases actually are --- comparing the mean against the median is the quickest check. A table reports one number per group; the distributions behind those numbers can differ in ways no single number records.

**Visualizing distributions across categories** shows what the table leaves out, in one of two forms. A [faceted histogram]{.important-text} gives each category its own panel and shows the full shape of every group's distribution --- skew, modality, spread --- at the cost of becoming cumbersome as the number of groups grows. A [grouped box plot]{.important-text} says much the same thing in far less space, sacrificing distributional detail for a compact summary that scales to many groups and makes medians, spreads, and outliers easy to compare at a glance. Both are built with [ggplot2]{.package-name}, and neither needs its missing values filtered out first: a missing number has no place on a continuous axis, so ggplot drops those rows itself and reports how many. Read that count and report it. A missing category would be different --- it gets a panel of its own.

- **Faceted histograms:** [Faceting]{.important-text} splits the plot by category. Map the categorical variable to `fill` inside `aes()` and add `facet_wrap(~variable)`, which gives each category its own [facet]{.important-text}, or panel. By default all facets share a y-axis, making group size easy to compare; `scales = "free_y"` gives each facet its own scale so every shape is visible, at the cost of comparing bar heights across facets. Panel titles come from `labeller = as_labeller(c(...))` inside `facet_wrap()`.
- **Grouped box plots:** Map the categorical variable to both `x` and `fill`, and the interval variable to `y`. The tick labels along the x-axis come from `scale_x_discrete()`.
- **Both plots:** Wrap the categorical variable in `factor()` when it is stored as a **number**, and leave it alone when it is stored as a **character** or a **factor** --- inside `facet_wrap()` it is never needed. Each label argument takes a named vector pairing every stored value with its replacement, so the order you list them in cannot matter. When the panel titles or the x-axis already identify the groups, `guides(fill = "none")` removes the redundant legend.

**Choosing between a histogram and a box plot** comes down to how many groups you have and how much of each distribution you need to show. With a handful of groups and real interest in their shapes, the histogram earns its space; with many groups, or when the comparison is about center and spread, the box plot is easier to read. In practice you will often want both.

One important limitation: none of this is a hypothesis test. Finding that group means differ in the predicted direction tells you the data are consistent with the hypothesis, not that the difference is larger than chance. That requires a formal test, which Tutorial 13 takes up for a categorical and an interval variable.

Tutorial 9 completes the sequence with two interval variables. With no categories to group by, the comparison of group means gives way to a scatter plot and a correlation coefficient, and the question shifts from whether an amount differs across groups to whether two quantities move together.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. Remember, the code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-08, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-08, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-08, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Great job! You can now compare interval variables across groups using means, histograms, and box plots.">
  <span aria-hidden="true">🎉</span>
  Great job! You can now compare interval variables across groups using means, histograms, and box plots.
</p>
')
```
