---
title: "Interactions in Multiple Regression"
tutorial:
  id: "17-Interactions"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to estimate and interpret conditional relationships in multiple regression.
---

## Overview {#main-content}

Every regression model in Tutorials 15 and 16 made a quiet assumption: the effect of each independent variable on the outcome is the same regardless of the values taken by any other variable. That is often a reasonable starting point, but it is frequently wrong. In many political and social relationships, the effect of one variable depends on the level of another --- and a model that ignores this will misrepresent what is actually going on.

When the effect of one variable depends on the value of another, we say the relationship is [**conditional**]{.important-text} or [**interactive**]{.important-text}. Estimating a conditional relationship in R requires adding one new ingredient to the model: a term multiplying the two variables together, created with the `*` operator. Everything else --- `lm()`, `summary()`, the interpretation logic --- works exactly as before. The challenge is interpreting the output, which is what this tutorial focuses on.

::: {.tip}
**Plots are coming.** Conditional relationships are most vividly communicated with a prediction plot showing how one variable's effect changes across values of the other. We will build those in Tutorial 18. In this tutorial we focus on interpreting the coefficients numerically first, which is the foundation you need to read those plots meaningfully.
:::

These tutorials work with three datasets --- **counties**, **qog**, and **world** --- that are close to complete populations for their unit of analysis. None is a random sample in the survey sense. The justification for hypothesis testing is that we are asking not about these specific counties or countries, but about the underlying processes that generate political outcomes. That logic holds across all three datasets and does not need restating in each example.

In this tutorial, you will learn to estimate and interpret conditional relationships in multiple regression. Specifically, we will cover:

* How to recognize when a relationship is likely to be conditional and what to ask before estimating an interaction model
* How to include an interaction term in a regression formula using the `*` operator in `lm()`
* How to interpret the coefficients on the interacting variables and the interaction term itself
* How to compute and interpret the conditional effect of one variable at meaningful values of another
* How to recognize when a conditional effect changes direction across the observed range of the data, and how to find the crossover point

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "QPATutorialsCourse")
counties$dem2p_percent <- counties$dem2p_vote_share * 100
counties$wage_growth <- counties$wage_growth * 100
counties$Black_percent <- counties$prop_black * 100
counties$white_percent <- counties$prop_white * 100
counties$hispanic_percent <- counties$prop_hispanic * 100
counties$college_grad_percent <- counties$prop_college_grad * 100
data("qog", package = "QPATutorialsCourse")
qog$rights[qog$fh_pr <= 2] <- 0
qog$rights[qog$fh_pr > 2] <- 1
qog$democ[qog$fh_ipolity2 < 4] <- 0
qog$democ[qog$fh_ipolity2 > 6] <- 2
qog$democ[qog$fh_ipolity2 >= 4 & qog$fh_ipolity2 <= 6] <- 1
qog$democ_f <- factor(qog$democ, labels = c("Autocracy", "Partial Democracy", "Democracy"))
```

## What Is a Conditional Relationship?

If the effect of $X_1$ on $Y$ differs based on the value of $X_2$, the relationship is conditional. $X_2$ moderates or conditions the effect of $X_1$ --- and because the relationship is symmetric, $X_1$ conditions the effect of $X_2$ at the same time.

```{r interaction-diagram, echo=FALSE, results='asis'}
cat('
<div style="text-align:center; margin: 1em 0;">
<svg role="img"
     aria-label="Diagram illustrating a conditional relationship. X1 has a horizontal arrow pointing right to Y, showing X1 predicts Y. X2 has a vertical arrow pointing downward onto the middle of the X1-to-Y arrow, showing that X2 conditions or moderates the effect of X1 on Y."
     viewBox="0 0 360 200"
     xmlns="http://www.w3.org/2000/svg"
     style="max-width:320px; width:100%;">
  <!-- X1 label -->
  <text x="18" y="133" font-family="Arial, sans-serif" font-size="46" font-weight="bold" fill="#1d4e89">X</text>
  <text x="52" y="145" font-family="Arial, sans-serif" font-size="20" fill="#1d4e89">1</text>
  <!-- Horizontal arrow X1 to Y -->
  <line x1="78" y1="115" x2="262" y2="115" stroke="#1d4e89" stroke-width="5"/>
  <polygon points="262,107 280,115 262,123" fill="#1d4e89"/>
  <!-- Y label -->
  <text x="288" y="133" font-family="Arial, sans-serif" font-size="46" font-weight="bold" fill="#1d4e89">Y</text>
  <!-- X2 label -->
  <text x="148" y="28" font-family="Arial, sans-serif" font-size="46" font-weight="bold" fill="#1d4e89">X</text>
  <text x="182" y="40" font-family="Arial, sans-serif" font-size="20" fill="#1d4e89">2</text>
  <!-- Vertical arrow X2 down to X1-Y arrow -->
  <line x1="170" y1="48" x2="170" y2="99" stroke="#1d4e89" stroke-width="5"/>
  <polygon points="162,99 170,115 178,99" fill="#1d4e89"/>
</svg>
</div>
')
```

If you estimate a model that includes $X_1$ and $X_2$ as separate predictors but omits the interaction term $X_1 \times X_2$, the model will misrepresent the effect of both variables. You cannot identify a conditional relationship by looking at the main effects alone.

### Two questions to ask before specifying an interaction

Before estimating a conditional model, think through what you expect:

1. [**Does the effect of each variable always run in the same direction?**]{.important-text} Is the effect of $X_1$ always positive (or always negative) for all values of $X_2$? Or could it flip sign? Asking the same question in reverse: is the effect of $X_2$ always the same direction for all values of $X_1$?

2. [**When is the effect expected to be strongest and weakest?**]{.important-text} For what values of $X_2$ do you expect the effect of $X_1$ to be largest? Smallest? These expectations give you something concrete to check against the actual results.

You may not have clear expectations for both questions. Your expectations may turn out to be wrong. When they are, report what you find honestly and think carefully about alternative explanations --- an unexpected sign or a surprisingly large coefficient may signal a real conditional relationship, but it may also reflect an omitted variable, a measurement issue, or a sample that is not representative of the population you care about.

## Interpreting Interaction Coefficients

For a model with two interval predictors and an interaction term:

$$Y = \hat{\alpha} + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \hat{\beta}_3 X_1 X_2 + \ldots + \varepsilon$$

the coefficients are interpreted as follows:

1. $\hat{\beta}_1$ is the effect of $X_1$ on $Y$ [**when $X_2 = 0$**]{.important-text}.
2. $\hat{\beta}_2$ is the effect of $X_2$ on $Y$ [**when $X_1 = 0$**]{.important-text}.
3. $\hat{\beta}_3$ is the additional change in the effect of $X_1$ for each one-unit increase in $X_2$ --- and equivalently, the additional change in the effect of $X_2$ for each one-unit increase in $X_1$.
4. The [**total conditional effect of $X_1$**]{.important-text} at any given value of $X_2$ is: $\hat{\beta}_1 + \hat{\beta}_3 X_2$.
5. The [**total conditional effect of $X_2$**]{.important-text} at any given value of $X_1$ is: $\hat{\beta}_2 + \hat{\beta}_3 X_1$.

Rules 4 and 5 are the ones you will use most. They let you answer questions like: "What is the effect of mortality risk in a county where wage growth is at the mean?" --- by substituting the relevant value into the formula.

Two important things to notice about rules 1 and 2: the main-effect coefficients $\hat{\beta}_1$ and $\hat{\beta}_2$ are no longer "the effect of $X_1$" in the ordinary sense. They are the effect specifically when the other variable equals zero. If zero is far outside the observed range of the data, those coefficients may not be substantively meaningful on their own --- the conditional effect formula in rules 4 and 5 is what you actually want to report.

## Do Mortality Risk and Wage Growth Conditionally Predict Clinton's Vote?

In Tutorial 16, you saw that both **mortality_risk_25_45** and **wage_growth** significantly predict **dem2p_percent** when treated as separate predictors. But is their joint effect simply additive? Or does the effect of one depend on the level of the other?

There is a plausible reason to expect a conditional relationship. In counties where wages are growing, residents may be more economically resilient, and conditions that would otherwise drive votes away from the incumbent --- like elevated mortality risk --- may matter more because those conditions stand out against an otherwise positive backdrop. Or the relationship may run differently. Before looking at the data, let's think through the two questions:

1. **Direction:** Do we expect mortality risk to have a negative effect on Clinton's vote for all values of wage growth? Most likely yes --- high mortality is never a political positive for an incumbent party. Similarly, do we expect wage growth to be positive for all values of mortality risk? Probably, but less certainly --- in counties with very high mortality risk, even economic gains may not be enough to overcome a broader sense of distress.

2. **Strength:** When is mortality risk's effect likely to be strongest? One possibility is that it is amplified in counties where wages are also growing, because the contrast between economic progress and poor health outcomes is especially stark. Another is that it is muted when wages are growing because economic gains provide a counterweight. We have competing intuitions, and we should let the data decide.

::: {.hypothesis}
In a comparison of counties, the effect of mortality risk on Hillary Clinton's percent of the two-party vote depends on the level of wage growth, and the effect of wage growth depends on the level of mortality risk.
:::

### Estimate the Regression

In R, the `*` operator between two variables in a formula tells `lm()` to include both main effects and the interaction term automatically. Writing `X1 * X2` is exactly equivalent to writing `X1 + X2 + X1:X2`. The output will include a row labeled `X1:X2` for the interaction coefficient.

::: {.tip}
**You must always include the constituent terms --- $X_1$ and $X_2$ separately --- alongside the interaction term $X_1 \times X_2$.** The interaction term measures the *additional* effect above and beyond what each variable contributes on its own. If you drop $X_1$ or $X_2$ from the model while keeping their product, the product term will absorb those main effects into itself, and the coefficients will no longer mean what the interpretation rules above describe. The `*` operator handles this automatically in R, but if you ever write an interaction formula by hand using `X1:X2`, you must add `+ X1 + X2` yourself.
:::

**Run and observe: Estimate the interaction model**

```{r exercise-interaction-clintonvote, exercise=TRUE, exercise.cap="Run and observe: Estimate the interaction model for Clinton vote and county characteristics"}
model_int <- lm(dem2p_percent ~ mortality_risk_25_45 * wage_growth +
                  college_grad_percent + hispanic_percent +
                  white_percent + Black_percent, data = counties)
summary(model_int)
```

Find the row labeled `mortality_risk_25_45:wage_growth` in the Coefficients table --- that is the interaction term. Look at its sign and significance before moving on.

### Interpret the Interaction Term

`lm()` and `summary()` are part of base R --- no `library()` call is needed for any exercise in this tutorial.

The interaction term is the new element. Its row in the Coefficients table — labeled `mortality_risk_25_45:wage_growth` — tells you whether and how the two variables condition each other's effect. Look at its sign and think about what direction you expected before answering.

**Practice: Interpret the interaction coefficient**

Look at the row labeled `mortality_risk_25_45:wage_growth` in your output before answering.

```{r quiz-interaction-sign, echo=FALSE}
question("Look at the row labeled mortality_risk_25_45:wage_growth in your output. Which statement correctly identifies what the interaction coefficient means for the relationship between mortality risk and Clinton's percent of the two-party vote?",
  correct = "Right. The interaction is negative and significant --- economic gains amplify rather than cushion the political effect of health distress.",
  answer("In counties where wages are growing more strongly, higher mortality risk is associated with an even larger drop in Clinton's vote --- economic gains do not cushion the political effect of health distress, they appear to amplify it",
         correct = TRUE,
         message = "The interaction coefficient is -0.139. The simple additive model in Tutorial 16 missed this: there, mortality risk and wage growth each had a fixed effect regardless of the other's value. Here, we see that the two variables are not independent --- the political meaning of health distress changes depending on the economic context in which it occurs."),
  answer("This coefficient is the effect of wage growth on Clinton's percent of the two-party vote, controlling for mortality risk and the other variables",
         message = "The row labeled mortality_risk_25_45:wage_growth is the interaction term, not the slope for wage_growth. The slope for wage_growth appears on its own separate row. The interaction term tells us how that slope changes as mortality risk changes, not what the slope is."),
  answer("This coefficient represents the combined effect of both mortality risk and wage growth on Clinton's vote when both variables are non-zero",
         message = "The interaction term is not the combined effect of both variables. It is the additional adjustment to one variable's slope for each unit change in the other. The individual slopes already capture each variable's main effect; the interaction term tells us how those slopes change depending on each other's value."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the interaction term tells you how the effect of one variable changes as the other changes --- not what either variable's effect is on its own."
)
```

### Interpret the Coefficients

Now apply the interpretation rules from above. Remember that $\hat{\beta}_1$ --- the coefficient on **mortality_risk_25_45** --- is the effect of mortality risk specifically when **wage_growth** equals zero. Zero is not outside the observed range (**wage_growth** ranges from about −69 to +63 in this data), but it is not representative of typical counties. Keep that in mind when reading the next question.

**Practice: Interpret the main-effect coefficients**

Look at the **mortality_risk_25_45** and **wage_growth** rows in your Coefficients table before answering.

```{r quiz-interpret-maineffects, echo=FALSE}
question("Which statements correctly describe what the coefficients on mortality_risk_25_45 and wage_growth tell us in this model? Select all that apply.",
  correct = "Right. Both main-effect coefficients describe effects at zero, which is either outside or at the edge of the observed range.",
  answer("The coefficient on mortality_risk_25_45 is the effect of mortality risk on Clinton's percent of the two-party vote when wage growth is zero --- not its effect in a typical county",
         correct = TRUE,
         message = "The coefficient is -2.494. In a county with flat wages, each additional percentage point of mortality risk is associated with about 2.5 fewer percentage points of Clinton's percent of the two-party vote. This is the starting point for understanding the conditional effect, not the full story."),
  answer("The coefficient on wage_growth is the effect of wage growth on Clinton's percent of the two-party vote when mortality risk is zero --- which never actually occurs",
         correct = TRUE,
         message = "Mortality risk ranges from about 1.1 to 8.4 in this dataset --- it is never zero. The coefficient on wage_growth (0.753) is therefore a mathematical anchor point, not a substantively meaningful prediction for any real county. The conditional effect formula in the next question is more useful."),
  answer("The coefficient on mortality_risk_25_45 is the average effect of mortality risk across all counties in the sample",
         message = "In an interaction model, the main-effect coefficient is not an average across all conditions. It is the effect specifically when the other interacting variable equals zero. The average conditional effect depends on the distribution of wage_growth across counties."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

### Compute Conditional Effects

This is where rules 4 and 5 come in. To find the effect of **mortality_risk_25_45** in a county with low **wage_growth** --- say, 2 percentage points --- substitute into the formula:

$$\text{Effect of mortality\_risk} = \hat{\beta}_{\text{mortality}} + \hat{\beta}_{\text{interaction}} \times \text{wage\_growth} = -2.494 + (-0.139)(2) = -2.772$$

So in a county with modest wage growth, each additional percentage point of mortality risk is associated with about a 2.8 percentage point drop in Clinton's vote. Now apply the same formula to a different value.

**Practice: Compute the conditional effect of mortality risk at high wage growth**

Find the **mortality_risk_25_45** and interaction coefficients in your output and substitute into the conditional effect formula before answering.

```{r quiz-conditional-mortality, echo=FALSE}
question("What is the conditional effect of a one-percentage-point increase in mortality risk in a county where wage growth is 8 percentage points --- a strong-growth county?",
  correct = "Right: -2.494 + (-0.139)(8) = -3.606, larger in magnitude than at low growth.",
  answer("About -3.61 percentage points",
         correct = TRUE,
         message = "In a county where wages are growing strongly, the mortality risk effect is actually larger in magnitude (-3.61) than in a low-growth county (-2.77). This tells us that mortality risk is a more powerful predictor of Clinton's vote in economically prosperous counties --- the contrast between rising wages and poor health outcomes may make voters more, not less, likely to punish the incumbent."),
  answer("About -2.494 percentage points",
         message = "That is the coefficient on mortality_risk_25_45 itself, which is its effect only when wage_growth equals zero. Substitute 8 into the conditional effect formula: -2.494 + (-0.139)(8)."),
  answer("About -1.382 percentage points",
         message = "Check the sign on the interaction term. The coefficient is -0.139, not +0.139. Substituting: -2.494 + (-0.139)(8) = -2.494 - 1.112 = -3.606, not -2.494 + 1.112."),
  answer("About -3.09 percentage points",
         message = "That would be the conditional effect at the mean wage growth (4.3), not at 8. Substitute 8: -2.494 + (-0.139)(8) = -3.606."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: substitute 8 into the conditional effect formula using the coefficients from your output: $\\hat{\\beta}_{\\text{mortality}} + \\hat{\\beta}_{\\text{interaction}} \\times 8$."
)
```

**Practice: Compute the conditional effect of wage growth at high mortality risk**

Find the **wage_growth** and interaction coefficients in your output and substitute into the conditional effect formula before answering.

```{r quiz-conditional-wagegrowth, echo=FALSE}
question("What is the conditional effect of a one-percentage-point increase in wage growth in a county where mortality risk is 5 (higher than average, well above the mean of 3.1)?",
  correct = "Right: 0.753 + (-0.139)(5) = 0.058, nearly zero in highly distressed counties.",
  answer("About 0.06 percentage points",
         correct = TRUE,
         message = "In a county with elevated mortality risk (5%), wage growth barely moves Clinton's percent of the two-party vote --- a 0.06 point gain per percentage point of wage growth. Compare this to the main-effect coefficient of 0.753, which applied when mortality risk was zero. The distress of high mortality nearly cancels out the benefit of economic growth."),
  answer("About 0.753 percentage points",
         message = "That is the main-effect coefficient on wage_growth, which applies only when mortality risk equals zero. Use the conditional effect formula: 0.753 + (-0.139)(5) = 0.753 - 0.695 = 0.058."),
  answer("About -0.219 percentage points",
         message = "That would be the conditional effect at mortality_risk = 7, not 5. Substitute 5: 0.753 + (-0.139)(5) = 0.753 - 0.695 = 0.058."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: substitute 5 into the conditional effect formula using the coefficients from your output: $\\hat{\\beta}_{\\text{wage}} + \\hat{\\beta}_{\\text{interaction}} \\times 5$."
)
```

The interaction coefficient is negative, and the main effect of **wage_growth** is positive. You can find the point where the conditional effect of **wage_growth** crosses zero by setting the conditional effect formula equal to zero and solving for **mortality_risk**:

$$0.753 + (-0.139) \times \text{mortality\_risk} = 0$$

**Practice: Find and interpret the crossover point**

Set the conditional effect of **wage_growth** equal to zero and solve for **mortality_risk_25_45** using the coefficients from your output before answering.

```{r quiz-crossover, echo=FALSE}
question("Solve the equation above for mortality_risk. At what mortality risk level does wage growth's effect cross zero, and what does it mean?",
  correct = "Right. The crossover occurs at about 5.4% --- in the most distressed counties, wage growth works against rather than for the incumbent.",
  answer("About 5.4% --- in counties where mortality risk exceeds 5.4%, wage growth is predicted to be associated with lower Clinton two-party vote percent, not higher",
         correct = TRUE,
         message = "Solving: 0.753 / 0.139 ≈ 5.4. This is where the conditional effect flips from positive to negative. In the most health-distressed counties --- where more than 1 in 18 adults aged 25-45 is at risk of dying --- even rising wages are associated with a shift away from the incumbent party. Economic gains no longer translate into incumbent support when distress reaches this level. Notice that about 5% of counties in the data have mortality risk above 5.4."),
  answer("About 4.3% --- the crossover happens at the mean of mortality risk",
         message = "The mean of mortality_risk_25_45 is about 3.1, not 4.3. Set the conditional effect equal to zero and solve: mortality_risk = 0.753 / 0.139 ≈ 5.4."),
  answer("About 0.753% --- the crossover equals the main effect of wage_growth",
         message = "The crossover is not the main effect itself. Set 0.753 + (-0.139)(mortality_risk) = 0 and solve for mortality_risk: mortality_risk = 0.753 / 0.139 ≈ 5.4."),
  answer("There is no crossover --- the interaction term changes the size of the effect but not its direction",
         message = "With a negative interaction term and a positive main effect, the conditional effect of wage_growth does eventually cross zero as mortality_risk increases. Set 0.753 + (-0.139)(mortality_risk) = 0 and solve."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: set the conditional effect formula equal to zero and solve for mortality_risk: $0.753 + (-0.139) \\times \\text{mortality\\_risk} = 0$."
)
```

### Assess Model Fit

Adding an interaction term raises a natural question: does it meaningfully improve how well the model predicts the outcome, or does the gain in understanding come without much gain in fit?

**Practice: Assess the fit of the interaction model**

Find the Adjusted R-squared for this model and compare it to the additive model's value before answering.

```{r quiz-modelfit-interaction, echo=FALSE}
question("Look at the Adjusted R-squared and the residual standard error in your output. Which statements are true? Select all that apply.",
  correct = "Right. The interaction adds a significant conditional relationship without dramatically changing overall fit.",
  answer("The Adjusted R-squared tells us the model explains about 60.7% of the variation in Clinton's vote across counties",
         correct = TRUE,
         message = "The same model without the interaction term has an Adjusted R-squared of 0.606 --- adding the interaction barely changes the overall fit. That is not surprising: one additional term among seven predictors rarely moves the fit dramatically. The p-value on the interaction term (0.004) is the primary evidence that the conditional relationship is real; the small R-squared change is simply consistent with what we expect when a significant but narrow effect is added to a model that already explains most of the variance."),
  answer("Because the Adjusted R-squared is about 0.607, the interaction term should be dropped --- it is not adding much",
         message = "The size of the Adjusted R-squared is not a reason to drop the interaction term. If theory and evidence suggest a conditional relationship exists --- and here both the p-value (0.004) and the crossover at mortality_risk ≈ 5.4 suggest it does --- the interaction belongs in the model regardless of how much variance it adds."),
  answer("The residual standard error tells us that under roughly normal errors, about two-thirds of counties have a predicted percent of the two-party vote within about 10 percentage points of their actual one",
         correct = TRUE,
         message = "Adding the interaction term does not meaningfully change the prediction error relative to the additive model --- the improvement is in understanding the conditional nature of the relationship, not in reducing prediction error overall."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```


### What We Learned About Mortality Risk, Wage Growth, and Clinton's Vote

The interaction is statistically significant and substantively important. Mortality risk does more damage to Clinton's vote in economically growing counties, not less. In places where wages are rising, the contrast between economic progress and poor health outcomes appears to amplify voter dissatisfaction with the incumbent party — economic gains do not buffer the political consequences of health distress, they make those consequences more visible. This finding was invisible in Tutorial 16's additive model, which assumed the effect of each variable was fixed regardless of the other. The crossover at about 5.4% mortality risk is also worth noting: in the most distressed counties, even wage growth works against the incumbent. The additive model would have reported wage growth as uniformly positive for Clinton — the interaction model reveals that is only true in counties that have not crossed a threshold of health distress.

## Interpreting Categorical × Interval Interactions

When one of the interacting variables is categorical and the other is interval, the interpretation rules change in an important way. Consider a model with a binary categorical variable $X_1$ (coded 0 for the reference category, 1 for the other) and an interval variable $X_2$:

$$Y = \hat{\alpha} + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \hat{\beta}_3 X_1 X_2 + \ldots + \varepsilon$$

1. $\hat{\alpha}$ is the expected value of $Y$ for the [**reference category**]{.important-text} when $X_2 = 0$.
2. $\hat{\beta}_1$ is the difference in the expected value of $Y$ between the two categories [**when $X_2 = 0$**]{.important-text}. Like the main-effect coefficients in interval×interval models, this is the gap at one specific (possibly unobserved) value of $X_2$, not the average gap.
3. $\hat{\beta}_2$ is the slope of $X_2$ [**for the reference category**]{.important-text} ($X_1 = 0$).
4. $\hat{\beta}_2 + \hat{\beta}_3$ is the slope of $X_2$ [**for the non-reference category**]{.important-text} ($X_1 = 1$). This is the step students most often miss: to get the non-reference category's slope, you must add the interaction term to the main effect. The main effect alone gives you the reference category's slope, but not the non-reference category's slope.
5. $\hat{\beta}_3$ is the difference in slopes between the two categories --- how much steeper or shallower the $X_2$ relationship is for the non-reference category compared to the reference.

::: {.tip}
**Think of it this way.** In a model without an interaction, the two categories share the same slope on $X_2$; only the intercept differs. An interaction term lets the slopes differ too. $\hat{\beta}_2$ is the reference category's slope; $\hat{\beta}_2 + \hat{\beta}_3$ is the non-reference category's slope; and $\hat{\beta}_3$ measures how different those two slopes are.
:::

## Do Political Rights Condition the Effect of Trade Openness on Environmental Performance?

Let's consider a second example. Countries that are open to international trade may face both economic pressures and global norms that encourage environmental regulation. But does trade openness matter equally in all political contexts? In countries where citizens have political rights to vote, organize, and hold governments accountable, environmental standards may be maintained regardless of trade. In countries where political rights are limited, trade openness may be one of the few forces capable of pushing governments toward better environmental performance.

Before looking at the data, let's apply the two questions.

1. **Direction:** Trade openness is probably positive for both groups --- more economic integration correlates with access to international environmental standards and the resources to meet them. Political rights are also probably positive --- accountable governments face electoral pressure to protect public goods like clean air and water. The signs are likely the same direction for both groups, but the magnitudes may differ substantially.

2. **Strength:** This is where the conditional logic gets interesting. In countries where citizens already have broad political rights, governments face ongoing accountability for environmental outcomes regardless of trade. Trade openness may add relatively little to environmental performance in that context. In countries where political rights are limited, trade openness may be one of the few external channels through which international environmental norms reach domestic policy --- making its effect potentially much stronger there. We expect the trade openness slope to be larger for Few Rights countries than for Many Rights countries.

::: {.hypothesis}
In a comparison of countries, the effect of trade openness on environmental performance depends on the extent of citizens' political rights, and the effect of political rights depends on trade openness.
:::

The data frame **qog** contains **epi_epi**, an overall environmental performance index ranging from about 27 to 87, where higher values indicate better performance. The independent variables are:

- **rights_f**: a two-category factor --- **Few Rights** (reference) for countries where citizens have limited political rights, and **Many Rights** for countries where citizens have broad political rights
- **ipi_tradeopen**: trade openness, scored 1 (least open) to 10 (most open)
- **hf_efiscore**: overall economic freedom, scored 0 to 100 --- included as a control to reduce omitted variable bias; we will not interpret its coefficient in detail

Note that 78 of the 194 countries in **qog** are missing at least one of these variables and will be dropped automatically when `lm()` is called.

### Create the Factor Variable

The variable **rights** is coded 0 (few political rights) and 1 (many political rights). We could wrap it in `factor()` directly inside the `lm()` formula, as you did in Tutorial 15 --- but that produces output rows labeled `factor(rights)1` and `factor(rights)1:ipi_tradeopen`. Those are hard to track when you need to combine two rows to find the non-reference slope. Creating a named factor first means the output will show `rights_fMany Rights` and `rights_fMany Rights:ipi_tradeopen`, making the connection between them much clearer.

**Run and observe: Create rights_f and check its levels**

```{r exercise-factor-rights, exercise=TRUE, exercise.cap="Run and observe: Create rights_f and confirm the reference category"}
qog$rights_f <- factor(qog$rights, labels = c("Few Rights", "Many Rights"))
levels(qog$rights_f)
```

**Few Rights** is listed first, so it is the reference category --- the one whose slope on **ipi_tradeopen** will appear directly as the main-effect coefficient $\hat{\beta}_2$ in the output.

### Estimate the Regression

**Practice: Estimate the interaction model**

Fill in the blanks to estimate the model, assigning it to **model_env**, then display the results. Two hints are available. Click Submit Answer to check your work.

```{r exercise-regression-env-interaction, exercise=TRUE, exercise.cap="Practice: Estimate the environmental performance interaction model"}
model_env <- lm(___ ~ rights_f * ___ + ___, data = qog)
summary(model_env)
```

```{r exercise-regression-env-interaction-hint-1}
# Hint 1 of 2
# Look back at the hypothesis. Which variable is the outcome?
# Which two variables are interacting, and which is the control?
# The outcome goes on the left of ~; predictors go on the right.
```

```{r exercise-regression-env-interaction-hint-2}
# Hint 2 of 2
# Use the * operator between rights_f and the trade openness
# variable --- the same * syntax as in Example 1. Add the
# economic freedom control as a separate + term after the
# interaction. The outcome goes on the left of ~.
```

```{r exercise-regression-env-interaction-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use lm() to estimate the regression.")
  }
  if (!grepl("model_env\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model_env.")
  }
  if (!grepl("epi_epi\\s*~", code_no_comments)) {
    fail("The dependent variable should be epi_epi --- environmental performance is what we are trying to explain.")
  }
  if (!grepl("rights_f\\s*\\*\\s*ipi_tradeopen|ipi_tradeopen\\s*\\*\\s*rights_f", code_no_comments)) {
    fail("Use * between rights_f and ipi_tradeopen to include both the main effects and the interaction term.")
  }
  if (!grepl("hf_efiscore", code_no_comments)) {
    fail("Include hf_efiscore as a control variable.")
  }
  if (!grepl("data\\s*=\\s*qog", code_no_comments)) {
    fail("Set data = qog so R knows where to find the variables.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call summary() on your saved model object to display the results.")
  }
  expected <- lm(epi_epi ~ rights_f * ipi_tradeopen + hf_efiscore, data = qog)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling summary() on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check your formula. The model should regress epi_epi on rights_f * ipi_tradeopen + hf_efiscore using the qog data.")
  }
  pass("Correct! Look at the row labeled rights_fMany Rights:ipi_tradeopen --- that is the interaction term. Find its sign and compare the two rows for ipi_tradeopen and rights_fMany Rights:ipi_tradeopen carefully before moving on.")
})
```

### Interpret the Interaction Term

The interaction term `rights_fMany Rights:ipi_tradeopen` tells us how different the trade openness slope is for Many Rights countries compared to Few Rights countries. Look at its sign and magnitude relative to the main effect of `ipi_tradeopen` before answering.

**Practice: Interpret the interaction coefficient**

Look at the row labeled `rights_fMany Rights:ipi_tradeopen` in your output before answering.

```{r quiz-env-interaction-sign, echo=FALSE}
question("Look at the row labeled rights_fMany Rights:ipi_tradeopen in your output. Which statement correctly identifies what the interaction coefficient means for the relationship between trade openness and environmental performance?",
  correct = "Right. The interaction is large, negative, and nearly cancels the trade openness slope for Many Rights countries.",
  answer("This coefficient tells us the slope of trade openness is substantially smaller for Many Rights countries than for Few Rights countries --- controlling for economic freedom, trade openness matters much less where political accountability is already strong",
         correct = TRUE,
         message = "The interaction coefficient is -4.629. The near-cancellation is the key story here: the main effect of trade openness (4.750) and the interaction (-4.629) are almost equal in magnitude, which means the slope for Many Rights countries is nearly zero. That is what the next section will ask you to confirm by computing both slopes."),
  answer("This coefficient is the slope of trade openness for Many Rights countries, controlling for economic freedom",
         message = "The interaction term tells you how different the two slopes are from each other --- not what the Many Rights slope actually is. To find the Many Rights slope, you need to add this term to the main-effect coefficient on ipi_tradeopen, which is a different row in the output."),
  answer("This coefficient means trade openness adds EPI points per unit for all countries in the sample",
         message = "Look at the correct row in the output. The ipi_tradeopen row and the interaction term row are different. The interaction term is rights_fMany Rights:ipi_tradeopen --- find that specific row and read its Estimate."),
  answer("This coefficient is the total combined effect of having both many political rights and high trade openness, compared to having neither",
         message = "The interaction term is not a combined effect. The main effects of rights_f and ipi_tradeopen already capture each variable's individual contribution; the interaction term tells us how those contributions change depending on each other's value."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the interaction term measures how different the two groups' slopes are from each other --- not what either slope actually is."
)
```

### Compute the Conditional Slopes

The interaction coefficient tells you how different the two slopes are from each other --- but not what either slope actually is. For that, apply rules 3 and 4. Rule 3: the **ipi_tradeopen** coefficient in your output is the slope for the reference category, **Few Rights**, directly. Rule 4: the slope for **Many Rights** is that value plus the interaction coefficient. Find both numbers from your output and apply the formula.

**Practice: Compute the slope of trade openness for Many Rights countries**

Find the **ipi_tradeopen** and interaction coefficients in your output and apply the conditional slope formula before answering.

```{r quiz-env-slopes, echo=FALSE}
question("Using the coefficients in your output, what is the slope of ipi_tradeopen for countries with Many Rights?",
  correct = "Right. Trade openness barely matters where political accountability is already strong --- the Many Rights slope is nearly zero.",
  answer("About 0.12 --- trade openness barely affects environmental performance in Many Rights countries",
         correct = TRUE,
         message = "For Few Rights countries the slope is 4.750: each additional point of trade openness predicts about 4.75 more points on the EPI. For Many Rights countries the slope is 4.750 + (-4.629) = 0.121 --- nearly zero. Trade openness does not meaningfully predict environmental performance once countries already have broad political rights. The two slopes could hardly be more different."),
  answer("About 4.750 --- the ipi_tradeopen coefficient is the slope for all countries regardless of rights status",
         message = "In an interaction model, the ipi_tradeopen coefficient is the slope only for the reference category (Few Rights). To find the Many Rights slope, add the interaction term to that coefficient."),
  answer("About -4.629 --- the interaction coefficient is the slope for Many Rights countries",
         message = "The interaction coefficient is the difference between the two slopes, not the Many Rights slope itself. Add the interaction coefficient to the Few Rights slope to find the Many Rights slope."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the Many Rights slope = the ipi_tradeopen main-effect coefficient + the interaction coefficient."
)
```

### Interpret the Category Gap

The **rights_fMany Rights** coefficient (29.226) is the gap between Many Rights and Few Rights countries when **ipi_tradeopen** equals zero. As with Example 1's main-effect coefficients, zero is outside the observed range of the data (trade openness runs from 1 to 10 in this sample), so this coefficient is a mathematical anchor rather than a substantively meaningful prediction. The pattern at observed values is more informative.

**Practice: Interpret the conditional relationship at observed trade levels**

Look at the `rights_fMany Rights` coefficient and the interaction row in your output before answering.

```{r quiz-env-interpretation, echo=FALSE}
question("At low trade openness (ipi_tradeopen = 2), Many Rights countries are predicted to score about 24 points higher than Few Rights countries on the EPI. At high trade openness (ipi_tradeopen = 8), Few Rights countries are predicted to slightly outperform Many Rights countries. Which statement best explains this pattern?",
  correct = "Right: trade openness substitutes for political accountability where rights are absent.",
  answer("In Few Rights countries, trade openness is a strong driver of environmental improvement --- possibly because exposure to international markets and standards substitutes for the political accountability that is absent. In Many Rights countries, environmental performance is sustained regardless of trade openness, so the slope is nearly flat.",
         correct = TRUE,
         message = "The finding illustrates a classic conditional relationship: a variable matters a lot in one context and almost not at all in another. Trade openness appears to be a substitute for political accountability, not a complement to it --- it helps where governance is weakest and adds little where it is already strong."),
  answer("Many Rights countries always outperform Few Rights countries on environmental protection, regardless of trade openness",
         message = "Look again at the crossover: at high trade openness, Few Rights countries slightly outperform Many Rights countries. The gap reverses across the observed range."),
  answer("The interaction term is too small to be meaningful even though it is statistically significant",
         message = "The interaction coefficient (-4.629) is actually quite large relative to the main effect of ipi_tradeopen (4.750) --- it nearly cancels the slope entirely for Many Rights countries. A small p-value and a large practical effect go together here."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what the near-zero Many Rights slope and the large Few Rights slope mean for which countries benefit most from trade openness."
)
```

The interaction term is the coefficient that matters most here. Find the row labeled `rights_fMany Rights:ipi_tradeopen` in your output and use its t-statistic and p-value to draw a conclusion.

**Practice: Test the interaction term**

Find the t-statistic and p-value for the `rights_fMany Rights:ipi_tradeopen` row in your output before answering.

```{r quiz-env-hypothesistest, echo=FALSE}
question("Based on the t-statistic and p-value for the interaction term rights_fMany Rights:ipi_tradeopen in your output, what conclusion do we draw?",
  correct = "Right. The evidence strongly supports rejecting the null hypothesis --- the effect of trade openness on environmental performance differs significantly between Few Rights and Many Rights countries.",
  answer("We reject the null hypothesis --- the effect of trade openness on environmental performance differs significantly between Few Rights and Many Rights countries",
         correct = TRUE,
         message = "The t-statistic is -4.533 and the p-value is 1.48e-05 --- strong evidence against a zero difference in slopes. We can be confident the slope for Many Rights countries is substantially lower than for Few Rights countries."),
  answer("We fail to reject the null hypothesis --- the interaction is not statistically significant",
         message = "Look at the t-statistic and p-value in the rights_fMany Rights:ipi_tradeopen row. A p-value well below 0.05 means we reject the null hypothesis."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: apply the decision rule --- if the p-value for the interaction term is below 0.05, we reject the null hypothesis that the two slopes are equal."
)
```

### Assess Model Fit

With the interaction term significant and the conditional slopes strikingly different, how well does the model capture overall variation in environmental performance?

**Practice: Assess the fit of the environmental performance model**

Find the Adjusted R-squared and Residual standard error at the bottom of your output before answering.

```{r quiz-env-modelfit, echo=FALSE}
question("Look at the Adjusted R-squared and the residual standard error in your output. Which statements are true? Select all that apply.",
  correct = "Right. The model explains a solid share of the variation in environmental performance, with a moderate typical prediction error.",
  answer("The model explains about 59.9% of the variation in environmental performance across the 116 countries with complete data",
         correct = TRUE,
         message = "That is solid explanatory power for a three-predictor model, though 78 countries were excluded due to missing data --- a limitation worth keeping in mind."),
  answer("The residual standard error of about 8.6 tells us that under roughly normal errors, about two-thirds of countries fall within about 8.6 points of their predicted EPI score",
         correct = TRUE,
         message = "Given that the EPI ranges from about 27 to 87 (a 60-point range), an 8.6-point typical error is moderate --- the model captures broad patterns but leaves substantial country-specific variation unexplained."),
  answer("The Adjusted R-squared of 0.599 tells us the interaction term explains 59.9% of the variation in environmental performance",
         message = "The Adjusted R-squared reflects the combined explanatory power of all four predictors --- rights_f, ipi_tradeopen, their interaction, and hf_efiscore --- not the interaction term alone."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

### What We Learned About Trade Openness, Political Rights, and Environmental Performance

The interaction is striking. Trade openness is a strong predictor of environmental performance in countries with few political rights — each additional point of trade openness predicts about 4.75 more points on the EPI. In countries with many political rights, trade openness barely matters — the slope collapses to nearly zero. The pattern suggests that trade openness functions as a substitute for political accountability, not a complement to it. Where citizens already have the rights to organize, vote, and hold governments accountable for environmental outcomes, trade adds little. Where those rights are absent, exposure to international markets and norms may be one of the few external forces capable of pushing governments toward better environmental performance. An additive model with both variables as separate predictors would have reported average slopes that masked this completely.

## Does the Effect of Fertility on Rule of Law Depend on Regime Type?

Let's now put everything together. You have estimated interaction models from scratch, computed conditional slopes for two-category factors, and interpreted both significant and near-zero interactions. This example asks you to do all of that again, with a three-category factor that produces two interaction rows instead of one.

From Tutorial 16 you already know that **wdi_fertility** and **democ_f** both predict **fh_rol** as separate predictors. The question here is whether their effects are simply additive, or whether the relationship between fertility and rule of law depends on regime type.

Let's apply the two questions:

1. **Direction:** Higher fertility is probably associated with weaker rule of law for all regime types --- it is a broad proxy for development conditions that constrain legal institutions regardless of political context. The sign is likely consistently negative. Do we expect regime type to moderate this? Possibly: democracies, with stronger accountability mechanisms, may be better at sustaining rule of law even under demographic pressure, while autocracies may show less of a relationship because rule of law is weak throughout.

2. **Strength:** If the interaction exists, we might expect the negative fertility effect to be weakest in democracies (where institutions are more resilient) and strongest in autocracies or partial democracies (where institutions are more fragile). But competing arguments are possible, and we may find the interaction is not clearly supported.

::: {.tip}
**An honest caveat before you run the model.** Not every theoretically motivated interaction is supported by the data. We have a plausible story for why regime type might condition the fertility effect, but the data may not bear it out clearly. A null or ambiguous finding is an informative finding --- it tells us the interaction is not strong enough to detect reliably with this data.
:::

::: {.hypothesis}
In a comparison of countries, the effect of fertility rate on rule of law depends on regime type.
:::

The data frame **qog** contains **fh_rol** (Freedom House Rule of Law, 0--16) and **democ_f** (the three-category factor you used in Tutorial 16, with **Autocracy** as the reference). **wdi_fertility** is the total fertility rate. Note that **democ_f** has three categories, which means `democ_f * wdi_fertility` will produce two interaction rows in the output --- one for Partial Democracy vs. Autocracy and one for Democracy vs. Autocracy.

### Estimate the Regression

**Practice: Estimate the interaction model**

Write the code from scratch to estimate the model, assigning it to **model_fhrol**, then display the results. Three hints are available. Click Submit Answer to check your work.

```{r exercise-regression-fhrol-interaction, exercise=TRUE, exercise.cap="Practice: Estimate the rule of law interaction model"}

```

```{r exercise-regression-fhrol-interaction-hint-1}
# Hint 1 of 3
# Look back at the hypothesis. Which variable is the outcome?
# Which two variables are interacting? Think about what you
# did in Examples 1 and 2.
```

```{r exercise-regression-fhrol-interaction-hint-2}
# Hint 2 of 3
# The * operator and the data frame are the same as in
# Example 2. democ_f is already in the qog environment.
# Assign your model to model_fhrol and pass it to summary().
```

```{r exercise-regression-fhrol-interaction-hint-3}
# Hint 3 of 3: fill in the blanks
model_fhrol <- lm(___ ~ ___ * ___, data = qog)
summary(model_fhrol)
```

```{r exercise-regression-fhrol-interaction-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use lm() to estimate the regression.")
  }
  if (!grepl("model_fhrol\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model_fhrol.")
  }
  if (!grepl("fh_rol\\s*~", code_no_comments)) {
    fail("The dependent variable should be fh_rol.")
  }
  if (!grepl("wdi_fertility\\s*\\*\\s*democ_f|democ_f\\s*\\*\\s*wdi_fertility", code_no_comments)) {
    fail("Use * between wdi_fertility and democ_f to include both main effects and the interaction terms.")
  }
  if (!grepl("data\\s*=\\s*qog", code_no_comments)) {
    fail("Set data = qog.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call summary() on your saved model object to display the results.")
  }
  expected <- lm(fh_rol ~ wdi_fertility * democ_f, data = qog)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling summary() on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check that your model includes fh_rol as the outcome and wdi_fertility * democ_f as the interaction, using the qog data.")
  }
  pass("Correct! Your output now has seven rows in the Coefficients table. Find the two interaction rows --- wdi_fertility:democ_fPartial Democracy and wdi_fertility:democ_fDemocracy --- and look at their signs, magnitudes, and p-values before moving on.")
})
```

### Interpret the Interaction Terms

With a three-category factor, the output has two interaction rows — one for each non-reference category compared to Autocracy. The question is whether either slope differs meaningfully from the autocracy baseline, and if so, in which direction. The note on the marginal significance of the main effect of **wdi_fertility** is worth keeping in mind as you interpret these.

**Practice: Interpret the two interaction coefficients**

Look at both interaction rows in your output before answering.

```{r quiz-fhrol-interactions, echo=FALSE}
question("Look at the two interaction rows in your output. Which statements are true? Select all that apply.",
  correct = "Right: each interaction term tells us whether and how the fertility slope differs from the autocracy baseline.",
  answer("The wdi_fertility:democ_fPartial Democracy coefficient tells us whether the slope of fertility for partial democracies differs from the slope for autocracies",
         correct = TRUE,
         message = "This interaction coefficient is 0.261 (p = .649) --- not statistically significant by any conventional standard. There is no reliable evidence that the fertility slope differs between partial democracies and autocracies."),
  answer("The wdi_fertility:democ_fDemocracy coefficient tells us whether the slope of fertility for full democracies differs from the slope for autocracies",
         correct = TRUE,
         message = "This interaction coefficient is -0.706 (p = .055) --- just above the conventional 0.05 threshold. The evidence for a different slope in full democracies is weak, though the negative sign suggests a steeper (more negative) fertility slope in democracies than in autocracies."),
  answer("The two interaction coefficients together tell us the average effect of fertility on rule of law across all countries",
         message = "The average effect across all countries is not what the interaction terms measure. The main effect of wdi_fertility gives the slope for the reference category (autocracies); the interaction terms measure how that slope differs for the other two categories."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

### Compute the Three Conditional Slopes

Apply rules 3 and 4 from the categorical × interval section. The **wdi_fertility** main effect gives the slope for **Autocracies** directly. For **Partial Democracy**, add the Partial Democracy interaction coefficient. For **Democracy**, add the Democracy interaction coefficient.

**Practice: Compute and compare the three conditional slopes**

Apply the conditional slope rules to each democracy category using the coefficients from your output before answering.

```{r quiz-fhrol-slopes, echo=FALSE}
question("Using the coefficients in your output, which set of conditional slopes is correct?",
  correct = "Right. All three slopes are negative, with the democracy slope notably steeper --- though interpret that with caution given the marginal significance.",
  answer("Autocracy: about -0.54; Partial Democracy: about -0.28; Democracy: about -1.25",
         correct = TRUE,
         message = "All three slopes are negative, which matches the expectation that higher fertility is associated with weaker rule of law regardless of regime type. The democracy slope (-0.544 + (-0.706) = -1.250) is notably steeper --- but as you saw above, the interaction term that produces it is only marginally significant. Interpret this pattern with appropriate caution."),
  answer("Autocracy: about -0.54; Partial Democracy: about -0.54; Democracy: about -0.54",
         message = "These would be the three slopes if there were no interaction --- if all three regime types shared the same fertility slope. In an interaction model, each category has its own slope: add the relevant interaction coefficient to the main effect to find it."),
  answer("Autocracy: about 0.26; Partial Democracy: about -0.71; Democracy: about -0.54",
         message = "These look like the interaction coefficients themselves, not the conditional slopes. The slopes are the main effect of wdi_fertility plus each interaction coefficient: add 0.261 to the main effect for Partial Democracy, and add -0.706 to the main effect for Democracy."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the autocracy slope is the wdi_fertility main effect directly; for the other categories, add the relevant interaction coefficient to that main effect."
)
```

### Assess the Evidence for the Interaction

**Practice: Evaluate whether the interaction is supported**

Find the p-values for both interaction terms and the Adjusted R-squared in your output before answering.

```{r quiz-fhrol-evidence, echo=FALSE}
question("Based on the p-values for both interaction terms and the change in Adjusted R-squared compared to the additive model from Tutorial 16 (0.640 vs. about 0.647 here), which statement best describes what the data tell us?",
  correct = "Right. Neither interaction term is significant and the R-squared barely moves --- both signals point the same direction.",
  answer("The data provide weak evidence for a conditional relationship: neither interaction term is significant at the conventional 0.05 level, and the small change in Adjusted R-squared corroborates that picture --- the theoretical expectation is not clearly supported",
         correct = TRUE,
         message = "This is a common and instructive outcome. We had a plausible theoretical reason to expect the interaction, and the three slopes do show a pattern --- but the pattern is not strong enough to distinguish reliably from chance variation with this sample. Note the distinction from Example 1: there, a significant interaction also produced a small R-squared change, which was simply expected given one new term among many. Here, the R-squared barely moves because the interaction terms themselves are not significant --- both signals point the same direction. A null or ambiguous interaction finding tells us something real: regime type does not appear to dramatically reshape the demographic constraints on rule of law."),
  answer("The interaction is confirmed because all three conditional slopes are negative and in the expected direction",
         message = "Slopes going in the expected direction is a necessary but not sufficient condition for confirming an interaction. The question is whether the slopes differ significantly across categories --- and neither interaction term reaches the conventional significance threshold."),
  answer("The interaction should be dropped entirely because the p-values are above 0.05",
         message = "A p-value just above 0.05 does not automatically mean a term should be dropped. The decision depends on theory, the strength of the evidence, and whether the costs of omitting the term outweigh the costs of including it. Here, the weak evidence simply means we should report the finding cautiously rather than treating the interaction as established."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at both the p-values for the interaction terms and the change in Adjusted R-squared relative to Tutorial 16's additive model --- do they point in the same direction?"
)
```

### What We Learned About Fertility, Regime Type, and Rule of Law

The theoretically motivated interaction is not clearly supported by the data — and that absence is itself informative. We had a plausible reason to expect that the demographic constraints on rule of law would look different across regime types: that democracies, with stronger accountability mechanisms, might sustain rule of law better under fertility pressure than autocracies. The three conditional slopes do show a gradient — the democracy slope is steepest at -1.25 — but neither interaction term reaches conventional significance, and the R-squared barely moves relative to the additive model from Tutorial 16. The honest interpretation is that regime type does not appear to dramatically reshape the relationship between fertility and rule of law with this data. That is a real finding: demographic constraints on legal institutions may operate similarly regardless of political context, at least at the level of variation captured here. A null or weak interaction is not a failure of the analysis — it is an answer to the research question.


## Check Your Understanding

These exercises ask you to apply the interpretation skills from this tutorial to new scenarios, without running any new code.

**Apply: Interpret a categorical × interval interaction**

A researcher models support for immigration policy reform ($Y$) as a function of contact with immigrants ($X_1$, a binary factor: **No Contact** as reference, **Contact**) and economic anxiety ($X_2$, a continuous scale from 0 to 10). The fitted equation is:

$$\hat{Y} = 4.2 + 2.8 \, X_1 + 0.6 \, X_2 - 0.9 \, X_1 X_2$$

```{r quiz-cyu-catint, echo=FALSE}
question("Which statements correctly interpret this equation? Select all that apply.",
  correct = "Right: contact raises support when anxiety is low but the advantage shrinks as anxiety rises.",
  answer("Among people with no contact, each one-unit increase in economic anxiety is associated with 0.6 more points of support for reform",
         correct = TRUE,
         message = "The X2 coefficient (0.6) is the slope for the reference category, No Contact --- exactly what rule 3 says."),
  answer("Among people with contact, each one-unit increase in economic anxiety is associated with -0.3 more points of support for reform",
         correct = TRUE,
         message = "The Contact slope is 0.6 + (-0.9) = -0.3. People who have contact with immigrants actually become less supportive of reform as their economic anxiety rises --- a striking reversal of the pattern for those without contact."),
  answer("The interaction coefficient (-0.9) is the slope of economic anxiety for people with contact",
         message = "-0.9 is not the Contact slope. It is the difference between the two slopes: the Contact slope is 0.6 + (-0.9) = -0.3, not -0.9 alone."),
  answer("At economic anxiety = 0, people with contact are expected to score 2.8 points higher on support for reform than people with no contact",
         correct = TRUE,
         message = "The X1 coefficient (2.8) is the gap between categories when X2 = 0 --- rule 2 directly."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

**Apply: Find the crossover point in an interval × interval interaction**

A researcher models voter turnout ($Y$) as a function of unemployment rate ($X_1$) and mean years of education in the district ($X_2$). The fitted equation is:

$$\hat{Y} = 55.0 - 1.8 \, X_1 + 2.1 \, X_2 + 0.4 \, X_1 X_2$$

Education ranges from 8 to 16 years in the sample.

```{r quiz-cyu-intint, echo=FALSE}
question("At what value of education would the conditional effect of unemployment cross zero, and does this occur within the observed range?",
  correct = "Right: 1.8 / 0.4 = 4.5 years, well below the observed minimum of 8 --- unemployment has a positive effect on turnout throughout the data.",
  answer("The crossover is at education = 4.5 years, which is outside the observed range --- unemployment has a positive effect on turnout throughout the data",
         correct = TRUE,
         message = "Setting -1.8 + 0.4(X2) = 0 and solving gives X2 = 1.8 / 0.4 = 4.5. Since education ranges from 8 to 16 in the data, this crossover falls below the minimum observed value. At education = 8, the conditional effect of unemployment is -1.8 + 0.4(8) = 1.4 --- already positive. Throughout the observed range, unemployment is associated with higher turnout, and higher education amplifies that effect further. When interpreting a crossover, always check whether it falls inside the data."),
  answer("The crossover is at education = 4.5 years, which is inside the observed range --- unemployment becomes positive in highly educated districts",
         message = "The crossover is at 4.5, but the observed range starts at 8 --- so the crossover is well below the minimum. Within the observed data, unemployment already has a positive effect on turnout at all education levels."),
  answer("The crossover is at education = 0.4 years",
         message = "The interaction coefficient alone is not the crossover point. Set the conditional effect equal to zero: -1.8 + 0.4(education) = 0, and solve for education: education = 1.8 / 0.4 = 4.5."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: set the conditional effect equal to zero and solve for education: $-1.8 + 0.4(\\text{education}) = 0$. Then check whether that value falls inside the observed range of 8 to 16."
)
```

**Apply: Decide when an interaction is warranted**

The two questions above gave you specific scenarios to work through. This one asks you to step back and think about the general principle.

```{r quiz-cyu-judgment, echo=FALSE}
question("A researcher estimates a model with two predictors and finds both are statistically significant. A colleague suggests adding an interaction term. Which of the following is the strongest reason to do so?",
  correct = "Right: theory is the primary justification for specifying an interaction.",
  answer("The researcher has a theoretical reason to expect that the effect of one predictor depends on the level of the other",
         correct = TRUE,
         message = "Theory is the primary justification for specifying an interaction. Adding an interaction term because it might improve fit, without a substantive reason to expect a conditional relationship, is a form of specification fishing --- it capitalizes on chance variation in the sample."),
  answer("The adjusted R-squared of the additive model is low, suggesting the model is missing something",
         message = "A low R-squared might mean any number of things --- an omitted variable, a nonlinear relationship, measurement error --- most of which an interaction term would not fix. Low R-squared alone is not a reason to add an interaction."),
  answer("Both predictors have large, positive coefficients in the additive model",
         message = "The sign or size of the main-effect coefficients tells us nothing about whether their effects are conditional on each other. Large coefficients are consistent with either additive or interactive relationships."),
  answer("The sample is large enough that the interaction term will likely be significant",
         message = "Statistical power is a practical consideration, but it is not a justification for adding an interaction. A large sample makes it easier to detect a real interaction but also easier to detect a spurious one. The justification must be theoretical, not statistical."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what distinguishes a principled reason to add an interaction term from a purely statistical one."
)
```


## The Takeaways

Conditional relationships add one new element to multiple regression: a term multiplying two predictors together, created with the `*` operator in R. That operator also includes the constituent main effects automatically --- you must never include the interaction term without them, because the interaction measures an *additional* effect above what the main effects already capture.

The interpretation rules differ depending on the level of measurement:

For [**interval × interval**]{.important-text} interactions, each main-effect coefficient is the slope of that variable *when the other equals zero*. Use the conditional effect formula --- $\hat{\beta}_1 + \hat{\beta}_3 X_2$ --- to find the slope at any substantive value of $X_2$.

For [**categorical × interval**]{.important-text} interactions, the interval main-effect coefficient is the slope for the *reference category*. The non-reference slope is that coefficient plus the interaction term. The categorical main-effect coefficient is the gap between categories *when the interval variable equals zero* --- often not a substantively meaningful value.

In both cases, $\hat{\beta}_3$ measures the difference in slopes between conditions, and you should always ask two questions before specifying an interaction: does the effect always run in the same direction across conditions? When should it be strongest and weakest?

[**A significant interaction does not always produce a large improvement in fit**]{.important-text}, as Example 1 showed. And [**an interaction that theory predicts may not be statistically supported**]{.important-text}, as Example 3 showed. Both are informative findings. The R² change from adding an interaction is expected to be small when a significant but narrow effect joins a model that already explains most of the variance; a genuinely absent interaction will show both non-significant terms and a small R² change together.

For [**categorical × categorical**]{.important-text} interactions, the same `*` syntax applies. The output will have $(k_1 - 1) \times (k_2 - 1)$ interaction rows, where $k_1$ and $k_2$ are the number of categories in each factor. Interpretation follows the same logic as categorical × interval, but with category differences on both sides rather than slopes. These models can become complex quickly, and prediction plots --- introduced in Tutorial 18 --- are especially useful for communicating their results.

Tutorial 18 introduces `plot_model()` from the **sjPlot** package, which visualizes all types of conditional effects as prediction plots --- a much more intuitive way to communicate what the slopes and crossover points mean to an audience.


## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial.

```{r encode-logic-17, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-17, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-17, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You have completed this tutorial and can now estimate and interpret interaction models with interval and categorical predictors.">
  <span aria-hidden="true">🔀</span>
  Excellent work! You can now estimate and interpret conditional relationships in multiple regression.
</p>
')
```
