---
title: "Univariate Description: Ordinal Variables"
tutorial:
  id: "05-Ordinal-Description"
output:
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Adds the median, and putting ordered categories in the right order.
---

## Overview {#main-content}

A great deal of political measurement records order without recording distance. A survey asks whether respondents strongly agree, agree, neither agree nor disagree, disagree, or strongly disagree; a questionnaire records income in brackets rather than in dollars. Each tells you that one case stands higher than another on some scale while leaving open how much separates them. The sequence is real information --- more than a set of unordered labels would give you --- but nothing guarantees that the step from one category to the next is the same size everywhere on the scale. Knowing the ranking without knowing the spacing is the position an analyst is in more often than not, and it changes what can honestly be said about a distribution.

That order is still worth something. Where unordered categories let you report only which was most common, ranked categories let you also ask where the middle of the distribution falls, and answer it with the **median**. Frequency tables, proportion tables, and bar plots all carry over, but they now have a direction to read across. The habit this tutorial builds is treating an ordered scale as ordered rather than as a set of labels that happen to be numbered: a result of 6 or 1 says nothing until you translate it back into the category it stands for.

This tutorial is the middle step of three on **univariate description**, and it covers **ordinal variables**, whose categories have a meaningful order but no guaranteed spacing. Tutorial 4 covers the nominal case, where categories are different but not ranked, and is where the frequency table, the proportion table and the bar plot are introduced --- the tools you keep using here, alongside the new one, `median()`.

In this tutorial you will learn:

* How to describe central tendency for ordinal variables using both the mode and the median
* How to explain why the median becomes meaningful when categories have a rank order
* How to use `median()`, and how to add `na.rm = TRUE` when a variable has missing values
* How to use frequency and proportion tables to describe dispersion and shape
* How to use `ggplot()` from the [ggplot2]{.package-name} package with `geom_bar()` to create bar plots for ordinal variables while preserving the substantive order of the categories
* How to check what order R will use for a variable's categories, and how to set the order yourself with `factor(levels = ...)`

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)
data("counties", package = "QPATutorialsCourse")
data("fHouse", package = "QPATutorialsCourse")
```

## Univariate Description

Tutorial 4 uses frequency tables, proportion tables, and bar plots to describe nominal variables to summarize central tendency, dispersion, and the shape of the distribution. The same tools apply to ordinal variables, but the rank order of the categories gives you one additional tool.

Before working through the examples, take a moment to identify which variables have ordered categories.

**Practice 1: Identify ordinal level variables**

```{r concepts-quiz-ordinal, echo=FALSE}
question("Which of the following variables are ordinal level measures? Select all that apply.",
  correct = "Right. Both are ordinal variables.",
  answer("The GDP of nations in the European Union, measured in dollars",
         message = "GDP is an interval (and ratio) measure. It is measured in dollars, so the values are ordered and the differences between them are equal at all points on the scale."),
  answer("Freedom House measure of whether a country is free, partly free, or not free", correct = TRUE,
         message = "The Freedom House measure is ordinal. Its response options are rank-ordered by how free a country is, but the differences between categories are not necessarily equal --- there is no guarantee that the political distance from Not Free to Partly Free equals the distance from Partly Free to Free."),
  answer("A respondent's education level, from less than high school to beyond college", correct = TRUE,
         message = "Education level is ordinal. Each successive value indicates more education, but the differences between categories are not equal --- the gap between high school and college is not the same as the gap between college and beyond college. This is the defining feature of ordinal measurement: a meaningful order without equal spacing."),
  answer("A variable measuring whether a country supported or opposed a particular UN resolution",
         message = "This is a nominal variable. Supported or did not support are unordered categories with no meaningful rank."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Two of these are ordinal. For each, ask whether the categories can be ranked, and whether the gaps between them are equal."
)
```

## Central Tendency

You can describe the [**central tendency**]{.important-text} of an ordinal variable using both the [**mode**]{.important-text} and the **median**. The mode tells you the most commonly occurring category --- the same tool used for nominal variables. The median is new here --- it was not available for nominal variables. You will work through the mode first, then the median as the new tool that ordinal measurement makes possible.

### The Mode

To illustrate both statistics, you will work with the variable **rural_urban** in the **counties** data frame. This variable is stored as a numeric variable, classifying US counties using a 9-point scale developed by the US Department of Agriculture, ranging from the most urban counties at one end to the most rural at the other. Knowing how counties are distributed across this scale tells you something important about the context in which the 2016 presidential election played out --- and about whose communities and concerns are most represented in county-level data.

The scale is coded as follows:

- 1: Counties in metro areas of 1 million population or more
- 2: Counties in metro areas of 250,000 to 1 million population
- 3: Counties in metro areas of fewer than 250,000 population
- 4: Urban population of 20,000 or more, adjacent to a metro area
- 5: Urban population of 20,000 or more, not adjacent to a metro area
- 6: Urban population of 2,500 to 19,999, adjacent to a metro area
- 7: Urban population of 2,500 to 19,999, not adjacent to a metro area
- 8: Completely rural or less than 2,500 urban population, adjacent to a metro area
- 9: Completely rural or less than 2,500 urban population, not adjacent to a metro area

Each successive category is less urban than the one before it, but the differences between categories are not equal --- making this an ordinal variable. That order is what lets you talk about a middle county, even though the numbered categories should not be treated as equally spaced units.

Before describing the "typical" county, you need to see how counties are distributed across the full urban-rural scale. A frequency and proportion table lets you see both the count in each category and the share of counties it represents --- and gives you the evidence you need to identify the mode.

**rural_urban** has a recorded value for every county, so the frequency and proportion table describes all 3,112 counties.

**Practice 2: Create a frequency and proportion table for rural_urban**

You built frequency and proportion tables in Tutorial 4 using `group_by()`, `summarise()`, and `mutate()` --- the same pipeline applies here. Name your count column **freq** and your proportion column **prop**. Load [dplyr]{.package-name} with `library(dplyr)` at the top. Three hints are available. Click "Submit Answer" to check your work.

```{r analysis-practice-rural-table, exercise=TRUE, exercise.lines=6, exercise.cap="Practice 2"}
library(dplyr)
counties %>%
```

```{r analysis-practice-rural-table-hint-1}
# Hint 1 of 3
# A frequency and proportion table answers two questions
# at once: how many counties fall in each category, and
# what share of all counties that count represents.
# The share matters because it tells you how dominant
# the most common category really is.
```

```{r analysis-practice-rural-table-hint-2}
# Hint 2 of 3
# Continue the pipeline with three steps: group the
# counties by the ordinal variable, count the cases in
# each group, and then express each count as a share of
# the total. The three functions are group_by(),
# summarise(), and mutate().
```

```{r analysis-practice-rural-table-hint-3}
# Hint 3 of 3
# Inside summarise(), n() counts the rows in each group;
# name that column freq. Inside mutate(), each proportion
# is that group's count divided by the sum of all the
# counts; name that column prop.
```

```{r analysis-practice-rural-table-check}
grade_this({
  expected <- counties %>%
    group_by(rural_urban) %>%
    summarise(freq = n(), .groups = "drop") %>%
    mutate(prop = freq / sum(freq))

  if (is.null(.result)) {
    fail("Make sure your code produces a table the grader can check.")
  }
  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing `%>%` between steps. Inside `summarise()` and `mutate()`, you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the `=`.")
  }

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use `group_by()` to group the data by **rural_urban** before summarising.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use `summarise()` with `n()` to create the **freq** column.")
  }
  if (!grepl("mutate\\s*\\(", code_no_comments)) {
    fail("Use `mutate()` to add the **prop** column.")
  }

  needed <- c("rural_urban", "freq", "prop")
  if (!all(needed %in% names(.result))) {
    fail("Make sure your table includes **rural_urban**, **freq**, and **prop**.")
  }

  result <- as.data.frame(.result)[needed]
  result <- result[order(result$rural_urban), ]
  expected <- as.data.frame(expected)[needed]
  expected <- expected[order(expected$rural_urban), ]
  rownames(result) <- NULL
  rownames(expected) <- NULL

  if (!identical(result$rural_urban, expected$rural_urban)) {
    fail("Your table should have one row for each value of **rural_urban**.")
  }
  if (!identical(as.numeric(result$freq), as.numeric(expected$freq))) {
    fail("Check your counts. Use `n()` inside `summarise()` to count counties in each **rural_urban** category.")
  }
  if (!isTRUE(all.equal(as.numeric(result$prop), as.numeric(expected$prop), tolerance = 1e-8))) {
    fail("Check your proportions. Each proportion should be **freq** divided by the total number of counties.")
  }

  pass("Correct! Your frequency and proportion table is complete. The **freq** column gives the count in each of the nine categories and **prop** gives each category's share of all counties.")
})
```

The highest count identifies the mode, but the analytic work is not finished until that numeric category is translated into words a reader can understand.

**Practice 3: Interpret the modal category**

Find the mode of **rural_urban** from the table you just created and translate it into a substantive description.

```{r concepts-quiz-mode-interpretation, echo=FALSE}
question("Which of the following statements *best* describes the mode of **rural_urban**?",
  correct = "Right. The substantive label is always more informative than the numeric code.",
  answer("There is no meaningful modal category, because the counties are spread fairly evenly across the nine categories",
         message = "One category has the highest count, even though the counts are fairly similar across categories. Compare the frequencies to identify it. Look at your frequency table and compare the counts."),
  answer("The modal category is 6, and reporting that numeric category is a sufficient description",
         message = "Category 6 is indeed the most frequent, but reporting a numeric code without explanation is not informative. What does category 6 actually represent?"),
  answer("The modal county is completely rural, with no urban population, and is not adjacent to a metro area",
         message = "That describes category 9, which is not the most common. Go back to the frequency table and find the category with the highest count before translating it into words."),
  answer("The modal county has an urban population of 2,500 to 19,999 and is adjacent to a metro area", correct = TRUE,
         message = "The most frequently occurring category is 6, but it is more informative to give the substantive meaning. Reporting 'the mode is 6' conveys nothing to a reader who does not have the codebook in front of them. Reporting that the most common county has an urban population of 2,500 to 19,999 and is adjacent to a metro area tells readers something real about the geographic character of US counties."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The mode is the category with the highest count. Once you have it, describe what that category is rather than reporting its number."
)
```

### The Median

Does the middle of the distribution fall in the same category as the mode? The [**median**]{.important-text} is the middle value when all observations are listed in rank order --- the value at the 50th percentile, where at least half the cases fall at or below it and at least half fall at or above it. Because ordinal categories have a meaningful rank order, you can ask which category sits at the midpoint of the distribution. That is what makes the median available for ordinal variables but not for nominal ones: without a rank order, there is no meaningful middle value.

One practical note: if the two middle observations fall in different categories, R's `median()` function returns the arithmetic mean of their numeric codes --- for example, 1.5 if the two middle values are coded 1 and 2. That number does not correspond to a real category. In that case, report the two bracketing categories rather than treating the numeric result as a meaningful value.

R's `median()` function computes the median directly. Pass it the variable using the `$` operator. Every county has a recorded value of **rural_urban**, so the median describes all 3,112 counties. The numeric codes must correctly represent the rank order of the categories --- always translate the numeric result back into the substantive category label when you report it.

**Run and observe 4: Calculate the median of rural_urban**

```{r analysis-calculate-median, exercise=TRUE, exercise.cap="Run and observe 4"}
median(counties$rural_urban)
```

The median is 6, which is not a quantity but a category: counties with an urban population of 2,500 to 19,999, adjacent to a metro area. Because the median depends on how the categories are ordered and coded, state how the scale runs whenever you report it. Here category 1 is the most urban and category 9 the most rural, so a median of 6 places the midpoint of the county distribution toward the more rural end of the scale.

**Practice 5: Interpret the mode and median of rural_urban**

```{r question-median-rural, echo=FALSE}
question("You just computed the mode and median of **rural_urban**. What does it mean when the mode and median fall in the same category?",
  correct = "Right. The mode and median both point to the same category.",
  answer("The distribution is perfectly symmetric",
         message = "Having the same mode and median does not guarantee symmetry. It means the most common value and the middle value coincide, but the distribution could still be skewed in either direction."),
  answer("The most common and middle county types are the same", correct = TRUE,
         message = "The mode identifies category 6 as the most common category, while the median identifies it as the category holding the midpoint of the ordered distribution. The two point to the same part of the scale, even though counties remain spread across all nine categories."),
  answer("All counties have an urban population of 2,500 to 19,999",
         message = "The mode and median describe the center of the distribution, not all values. Many counties fall in other categories."),
  answer("The median cannot be computed for an ordinal variable",
         message = "The median is appropriate for ordinal variables because their values have a meaningful rank order."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Both statistics are pointing at the same category. Ask what each one measures --- most common, and middle --- rather than what they imply about the whole distribution."
)
```

## Dispersion and the Shape of the Distribution

[**Dispersion**]{.important-text} refers to how spread out the values of a variable are across its categories. A variable with low dispersion has most cases concentrated in one or a few categories; a variable with high dispersion has cases spread more evenly across many categories. The more dispersed a variable, the less informative the mode is as a summary --- when cases are spread evenly, knowing the most common category tells you relatively little about the typical case. That is why ordinal description works best when the mode, median, and full distribution are read together.

The frequency table you built shows that **rural_urban** is fairly dispersed: the counts range across all nine categories, and even the largest holds only about 19% of counties. This means the mode is a less informative summary than it would be if one category dominated --- knowing the most common county type gives only a partial picture of how varied US counties actually are.

[**Shape**]{.important-text} is the other half of what a table can tell you: the overall pattern the categories form. The same four labels from Tutorial 4 apply here --- a distribution is [**unimodal**]{.important-text} when one category clearly leads, [**bimodal**]{.important-text} when two are similarly prominent, [**uniform**]{.important-text} when cases are spread approximately evenly, and [**multimodal**]{.important-text} when several stand out. For an ordinal variable, though, the pattern also has a [**direction across the scale**]{.important-text}, because the categories are ranked: you can ask not just whether one category dominates but whether cases pile up at one end of the scale, thin out toward the other, or cluster in the middle. A frequency table contains all of that information, but it is much easier to see in a bar plot, where the relative size of the categories is visible at once.

## Visualizing an Ordinal Variable

A bar plot shows the same distribution as the frequency table, but it makes the pattern across categories visible at once rather than requiring you to compare numbers. A nominal bar chart could not show direction at all, because its categories have no order to run along --- here the order carries meaning, so the picture does too.

This is why bar plots for ordinal variables follow the same layered structure as for nominal variables, but the categories must appear in their rank order on the x-axis --- not alphabetically or by frequency --- so that the shape of the distribution is visually meaningful.

### County Urban-Rural Classification

A bar plot lets you compare all nine categories at once and ask whether counties cluster at one end of the urban-rural scale, spread fairly evenly across it, or concentrate around a particular category.

**Practice 6: Create a frequency bar plot for rural_urban**

You built frequency bar charts in Tutorial 4 using `ggplot()`, `geom_bar()`, `labs()`, and `scale_x_discrete()` --- the same structure applies here. The one addition is `coord_flip()` as a final layer, which flips the bars horizontal and makes the nine category labels readable. Remember that it is stored as a numeric variable, even though its nine values are ordered categories rather than quantities. Complete the code below: map the variable to the x-axis, add the bar layer with its fill color, and add the final layer that flips the bars. The plot labels and the nine shortened category labels are already supplied. Set the bar fill to \"firebrick\". Three hints are available. Click "Submit Answer" to check your work.

```{r visual-practice-bar-rural, exercise=TRUE, exercise.cap="Practice 6", fig.alt="Frequency bar chart of the nine-point urban-rural classification of US counties, drawn horizontally, with the most urban category at the bottom and the most rural at the top. All nine ordered categories carry substantial bars and none dominates the rest: the longest is 2,500-19,999 near metro, in the middle of the scale, and the shortest is 20k+ not near metro."}
library(ggplot2)
ggplot(data = counties, mapping = aes(x = ___)) +
  ___ +
  labs(title = "USDA Urban/Rural Classification of US Counties",
       subtitle = "2016",
       x = NULL,
       y = "Frequency",
       caption = "Source: Linn, Nagler, Zilinsky") +
  scale_x_discrete(labels = c("1 million+", "250,000 to 1 million", "< 250,000",
                              "20k+ near metro", "20k+ not near metro",
                              "2,500-19,999 near metro", "2,500-19,999 not near metro",
                              "< 2,500 near metro", "< 2,500 not near metro")) +
  ___
```

```{r visual-practice-bar-rural-hint-1}
# Hint 1 of 3
# ggplot decides how to draw an axis from the class of
# whatever you map to it. Map something numeric and you
# get a continuous axis --- a number line. Map something
# categorical and you get one bar per category.
# You have nine ordered categories stored as numbers.
```

```{r visual-practice-bar-rural-hint-2}
# Hint 2 of 3
# First blank: wrap rural_urban in factor() inside aes().
# Second blank: geom_bar().
# Third blank: coord_flip().
```

```{r visual-practice-bar-rural-hint-3}
# Hint 3 of 3
# The fill color is a fixed setting applied to every bar,
# not a mapping from a variable, so it goes inside
# geom_bar() but outside any aes() call: fill = "firebrick".
# coord_flip() takes no arguments --- empty parentheses
# are all it needs.
```

```{r visual-practice-bar-rural-check}
grade_this({
  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }

  bar_layers <- Filter(function(layer) inherits(layer$geom, "GeomBar"), .result$layers)
  if (length(bar_layers) == 0) {
    fail("Add `geom_bar()` to create a frequency bar plot.")
  }

  if (!is.data.frame(.result$data) || !"rural_urban" %in% names(.result$data)) {
    fail("Use **rural_urban** from the **counties** data frame for this plot.")
  }

  expected_values <- sort(as.character(counties$rural_urban[!is.na(counties$rural_urban)]))
  actual_values   <- sort(as.character(.result$data$rural_urban[!is.na(.result$data$rural_urban)]))
  if (!identical(actual_values, expected_values)) {
    fail("Use all nonmissing observations of **rural_urban** from the **counties** data frame. Do not remove any valid observations.")
  }

  x_expr <- paste(deparse(rlang::get_expr(.result$mapping$x)), collapse = "")
  if (!grepl("factor\\s*\\(\\s*rural_urban\\s*\\)", x_expr) &&
      !grepl("as.factor\\s*\\(\\s*rural_urban\\s*\\)", x_expr)) {
    fail("Map `factor(rural_urban)` to the x-axis so the ordinal categories appear as categories.")
  }

  bar_layer <- bar_layers[[1]]
  y_map <- tryCatch(rlang::get_expr(bar_layer$mapping$y), error = function(e) NULL)
  y_expr <- if (!is.null(y_map)) paste(deparse(y_map), collapse = "") else ""
  if (grepl("after_stat\\s*\\(\\s*prop\\s*\\)", y_expr) || grepl("\\.\\.prop\\.\\.", y_expr)) {
    fail("This exercise asks for a frequency bar chart, not a proportion chart. Remove the `y = after_stat(prop)` and `group = 1` arguments from `geom_bar()`.")
  }

  fill_value <- bar_layer$aes_params$fill
  if (is.null(fill_value) || !identical(tolower(fill_value), "firebrick")) {
    fail('Set `fill = "firebrick"` inside `geom_bar()` to color the bars.')
  }

  has_flip <- isTRUE(tryCatch({
    coord_class <- class(.result$coordinates)[1]
    grepl("flip|Flip", coord_class, ignore.case = TRUE) ||
      isTRUE(.result$coordinates$is_flipped())
  }, error = function(e) {
    any(grepl("flip|Flip", class(.result$coordinates), ignore.case = TRUE))
  }))
  if (!has_flip) {
    fail("Add `coord_flip()` as the final layer to flip the bars horizontal.")
  }

  pass("Well done! Your chart has the right variable, the bar layer, the fill color, and the horizontal orientation that makes the nine category labels readable.")
})
```

The bars are fairly similar in length across all nine categories, with no single category clearly standing out. This tells you three things:

- **Central tendency:** The modal category is "Urban population of 2,500 to 19,999, adjacent to a metro area" (category 6) --- the same as the median. Category 6 is both the most common category and the midpoint of the distribution, though counties remain spread across all nine categories. Note that this describes the distribution of *counties*, not of *people* --- because each county counts once here, the finding tells you about units of local government, not about where most Americans live.
- **Dispersion:** The counts are spread across all nine categories without any single category clearly dominating. The data are fairly dispersed, which means the mode is a less informative summary than it would be if one category dominated.
- **Shape:** The distribution is spread fairly evenly across categories with no strong peak and no obvious concentration at either the urban or rural end. If half the counties were very urban and half very rural the distribution would be bimodal, but that is not what the data show.

Note that the median cannot be read directly from a bar plot --- that requires the computation you performed earlier.

One thing worth noticing: the bars appear in the correct rank order because the numeric codes for **rural_urban** run from 1 (most urban) to 9 (most rural) in rank order, and `factor()` preserves that numeric ordering. After `coord_flip()`, the most urban category appears at the bottom and the most rural at the top. If the variable were stored as character labels rather than ordered numbers, R would display the categories alphabetically instead. Always check that the axis reflects the substantive rank order of the categories before interpreting the shape of the distribution.

### Freedom House Political Freedom Ratings

Now let's apply the same tools to a second ordinal variable. The variable **FHStatus_2022** in the **fHouse** data frame records Freedom House's 2022 classification of countries as Not Free, Partly Free, or Free. Freedom House is a non-governmental organization that rates political freedom and civil liberties in countries worldwide; their assessments are widely used in political science research as measures of democracy and human rights. The variable is stored as a numeric variable, taking three values: "Not Free" (coded 0), "Partly Free" (coded 1), and "Free" (coded 2). As you go from "Not Free" to "Free" a country becomes progressively more politically free, making this an ordinal variable.

Unlike **rural_urban**, **FHStatus_2022** is missing for one of the 194 countries. One is enough for the `filter()` line in the starter code to matter. Without it that country would form a group of its own and land in the denominator, making the substantive proportions sum to less than 1. With it, the proportions describe the countries that have a rating rather than all countries in the data. The frequency and proportion table will tell you which category is most common and how much of the distribution falls below or above it.

**Practice 7: Create a frequency and proportion table for FHStatus_2022**

Write a pipeline that produces a frequency and proportion table for **FHStatus_2022**. The first two lines are provided --- complete the pipeline by adding `group_by()`, `summarise()`, and `mutate()`. Name your count column **freq** and your proportion column **prop**. Three hints are available. Click "Submit Answer" to check your work.

```{r analysis-practice-fhstatus-freq, exercise=TRUE, exercise.lines=8, exercise.cap="Practice 7"}
fHouse %>%
  filter(!is.na(FHStatus_2022)) %>%

```

```{r analysis-practice-fhstatus-freq-hint-1}
# Hint 1 of 3
# You are building the same kind of table as before, for
# a different variable: the count of countries in each
# freedom category, and each count as a share of all
# countries with a recorded rating.
```

```{r analysis-practice-fhstatus-freq-hint-2}
# Hint 2 of 3
# The filter() step is already done for you. Continue the
# pipeline with group_by(), summarise(), and mutate().
# Use n() inside summarise() to count the countries in
# each freedom category.
```

```{r analysis-practice-fhstatus-freq-hint-3}
# Hint 3 of 3
# Group by FHStatus_2022, name the count column freq, and name
# the proportion column prop. Each proportion is freq
# divided by the sum of freq.
```

```{r analysis-practice-fhstatus-freq-check}
grade_this({
  expected <- fHouse %>%
    filter(!is.na(FHStatus_2022)) %>%
    group_by(FHStatus_2022) %>%
    summarise(freq = n(), .groups = "drop") %>%
    mutate(prop = freq / sum(freq))

  if (is.null(.result)) {
    fail("Make sure your code produces a table the grader can check.")
  }
  if (!is.data.frame(.result)) {
    fail("Your result should be a data frame. If you're seeing an error, check for unmatched parentheses or a missing `%>%` between steps. Inside `summarise()` and `mutate()`, you are creating new columns with new names --- do not use existing variable names from the data frame on the left side of the `=`.")
  }

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("group_by\\s*\\(", code_no_comments)) {
    fail("Use `group_by()` to group the data by **FHStatus_2022** before summarising.")
  }
  if (!grepl("summari[sz]e\\s*\\(", code_no_comments)) {
    fail("Use `summarise()` with `n()` to create the **freq** column.")
  }
  if (!grepl("mutate\\s*\\(", code_no_comments)) {
    fail("Use `mutate()` to add the **prop** column.")
  }

  needed <- c("FHStatus_2022", "freq", "prop")
  if (!all(needed %in% names(.result))) {
    fail("Make sure your table includes **FHStatus_2022**, **freq**, and **prop**.")
  }

  result <- as.data.frame(.result)[needed]
  result <- result[order(result$FHStatus_2022), ]
  expected <- as.data.frame(expected)[needed]
  expected <- expected[order(expected$FHStatus_2022), ]
  rownames(result) <- NULL
  rownames(expected) <- NULL

  if (!identical(result$FHStatus_2022, expected$FHStatus_2022)) {
    fail("Your table should have one row for each non-missing value of **FHStatus_2022**. Make sure missing **FHStatus_2022** values are excluded.")
  }
  if (!identical(as.numeric(result$freq), as.numeric(expected$freq))) {
    fail("Check your counts. Use `n()` inside `summarise()` after grouping by **FHStatus_2022**.")
  }
  if (!isTRUE(all.equal(as.numeric(result$prop), as.numeric(expected$prop), tolerance = 1e-8))) {
    fail("Check your proportions. Each proportion should be **freq** divided by the total number of non-missing **FHStatus_2022** cases.")
  }

  pass("Good work! Your **FHStatus_2022** table is complete, and the one country with no recorded status was excluded before the proportions were computed --- so each proportion is a share of the countries that have a rating.")
})
```

**Practice 8: Interpret the mode of FHStatus_2022**

Find the mode of **FHStatus_2022** from the table you just created and translate it into a substantive description.

```{r concepts-quiz-fhstatus-mode, echo=FALSE}
question("Which of the following statements *best* describes the mode of **FHStatus_2022**?",
  correct = "Right. The modal category is Free.",
  answer("The mode is 2",
         message = "Rather than report a numeric code, describe the modal category substantively. What does a value of 2 represent?"),
  answer("The modal category is Partly Free because it lies in the middle of the ordered scale",
         message = "The middle of the ordered scale is where the median falls, not the mode. The mode is simply the most frequently occurring category --- it has nothing to do with position on the scale. Check your frequency table for the category with the highest count."),
  answer("The modal category is Free", correct = TRUE,
         message = "The most frequently occurring category is coded 2, which represents countries assessed as Free. Reporting this substantively is always more informative than citing the numeric code --- 'the modal category is Free' tells readers something meaningful, while 'the mode is 2' does not."),
  answer("The distribution is multimodal",
         message = "Look at your frequency table. One category clearly has more countries than the others."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Read the count column, then say the category in words."
)
```

Because **FHStatus_2022** is coded in rank order --- 0 for Not Free through 2 for Free --- you can calculate its median directly.

**Practice 9: Calculate the median for FHStatus_2022**

Now ask whether the midpoint of the ordered distribution falls in the same category as the mode. Calculate the median of **FHStatus_2022** in the **fHouse** data frame. One country has no recorded status, and `median()` returns `NA` if any value is missing --- your call needs to account for that. Three hints are available. Click "Submit Answer" to check your work.

```{r analysis-practice-fhstatus-median, exercise=TRUE, exercise.cap="Practice 9"}

```

```{r analysis-practice-fhstatus-median-hint-1}
# Hint 1 of 3
# The median is the value at the midpoint of the ordered
# distribution --- rank every country from least to most
# free, and the median is the category the middle country
# falls in. You want a single number back.
```

```{r analysis-practice-fhstatus-median-hint-2}
# Hint 2 of 3
# The function is median(). Pass it the variable through
# the $ operator, and include na.rm = TRUE --- FHStatus_2022
# has missing values, so without it the result is NA.
```

```{r analysis-practice-fhstatus-median-hint-3}
# Hint 3 of 3
# The data frame is fHouse and the variable is FHStatus_2022.
# Reach the variable the same way you reached rural_urban
# in the county example, and remember to translate the
# number you get back into its category label.
```

```{r analysis-practice-fhstatus-median-check}
grade_this({
  expected <- median(fHouse$FHStatus_2022, na.rm = TRUE)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("median\\s*\\(", code_no_comments)) {
    fail("Use the `median()` function to compute the median.")
  }
  if (!grepl("fHouse\\s*\\$\\s*FHStatus_2022", code_no_comments)) {
    fail("Access the variable with the dollar-sign operator: `fHouse$FHStatus_2022`.")
  }
  if (!grepl("na\\.rm\\s*=\\s*TRUE", code_no_comments)) {
    fail("Include `na.rm = TRUE` so `median()` excludes the missing **FHStatus_2022** values.")
  }

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number returned by `median()`.")
  }
  if (is.na(.result)) {
    fail("Your result is `NA`. Include `na.rm = TRUE` so R ignores missing values.")
  }
  if (!isTRUE(all.equal(as.numeric(.result), as.numeric(expected)))) {
    fail("Check that you are computing the median of **FHStatus_2022** in the **fHouse** data frame.")
  }

  pass(paste0("Correct! The median is ", expected, ", which corresponds to the category Partly Free."))
})
```

If the mode and median point to different categories, the full distribution helps explain why. A proportion bar chart will show how the countries are spread across the three categories, making the relationship between the mode and median visible. Proportions are more useful than counts here because they tell you each category's share of all countries --- making the distribution easier to describe and compare.

**Practice 10: Create a proportion bar chart for FHStatus_2022**

Build a proportion bar chart for **FHStatus_2022**. Filter out the missing values before piping the data into `ggplot()`, and use the proportion mapping inside `geom_bar()` that you saw earlier in this tutorial. Remember that **FHStatus_2022** is stored as a numeric variable even though its three values are ordered categories --- that has a consequence for how you map it to the x-axis. As in Tutorial 4, the data arrive via a pipe, so only `mapping =` is needed, not `data =`. In `scale_x_discrete()`, label the categories in order from code 0 to 2: \"Not Free\", \"Partly Free\", \"Free\".

Title: \"Political Freedom Around the World\", x-axis label: `NULL`, y-axis: \"Proportion\", caption: \"Source: Freedom House\", fill: \"steelblue\". Three hints are available. Click "Submit Answer" to check your work.

```{r visual-practice-fhstatus-prop-bar, exercise=TRUE, exercise.cap="Practice 10", fig.alt="Proportion bar chart of countries by Freedom House status, with three bars labeled Not Free, Partly Free, and Free, in that order along the horizontal axis. Free is the tallest at about 0.44. Not Free and Partly Free are close to each other and much shorter, at about 0.29 and 0.27. The vertical axis is labeled Proportion."}

```

```{r visual-practice-fhstatus-prop-bar-hint-1}
# Hint 1 of 3
# A proportion bar chart shows each category's share of
# all cases rather than its raw count, which is what lets
# you see how the countries are distributed across the
# three freedom categories at a glance.
```

```{r visual-practice-fhstatus-prop-bar-hint-2}
# Hint 2 of 3
# Pipe the data through filter() before ggplot(), map
# factor(FHStatus_2022) to the x-axis, and add geom_bar()
# with the proportion mapping inside it, then labs()
# for the title, axis labels, and caption.
```

```{r visual-practice-fhstatus-prop-bar-hint-3}
# Hint 3 of 3
# Inside geom_bar(), map y to after_stat(prop) and set
# group = 1 in the same aes() call. Put fill = "steelblue"
# outside aes(). Add labs() and scale_x_discrete() with the
# three category labels in order: Not Free, Partly Free, Free.
```

```{r visual-practice-fhstatus-prop-bar-check}
grade_this({
  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }

  if (!is.data.frame(.result$data) || !"FHStatus_2022" %in% names(.result$data)) {
    fail("Use **FHStatus_2022** from the **fHouse** data frame for this plot.")
  }
  if (any(is.na(.result$data$FHStatus_2022))) {
    fail("Filter out missing **FHStatus_2022** values before plotting.")
  }

  expected_values <- sort(as.character(fHouse$FHStatus_2022[!is.na(fHouse$FHStatus_2022)]))
  actual_values   <- sort(as.character(.result$data$FHStatus_2022))
  if (!identical(actual_values, expected_values)) {
    fail("Use all nonmissing observations of **FHStatus_2022** from the **fHouse** data frame. Do not remove any valid observations.")
  }

  bar_layers <- Filter(function(layer) inherits(layer$geom, "GeomBar"), .result$layers)
  if (length(bar_layers) == 0) {
    fail("Add `geom_bar()` to create the bar plot.")
  }

  x_expr <- paste(deparse(rlang::get_expr(.result$mapping$x)), collapse = "")
  if (!grepl("factor\\s*\\(\\s*FHStatus_2022\\s*\\)", x_expr) &&
      !grepl("as.factor\\s*\\(\\s*FHStatus_2022\\s*\\)", x_expr)) {
    fail("Map `factor(FHStatus_2022)` to the x-axis so the categories are treated as discrete rather than numeric.")
  }

  layer_map <- bar_layers[[1]]$mapping
  y_expr <- paste(deparse(rlang::get_expr(layer_map$y)), collapse = "")

  if (!grepl("after_stat\\s*\\(\\s*prop\\s*\\)", y_expr) && !grepl("\\.\\.prop\\.\\.", y_expr)) {
    fail("Inside `geom_bar()`, map y to `after_stat(prop)` so the bars show proportions.")
  }

  group_quo <- bar_layers[[1]]$mapping$group
  group_ok <- tryCatch({
    !is.null(group_quo) && identical(rlang::eval_tidy(group_quo), 1)
  }, error = function(e) FALSE)
  if (!group_ok) {
    fail("Inside `geom_bar()`, include `group = 1` so proportions are computed across all **FHStatus_2022** categories.")
  }

  fill_value <- bar_layers[[1]]$aes_params$fill
  if (is.null(fill_value) || !identical(tolower(fill_value), "steelblue")) {
    fail('Set `fill = "steelblue"` inside `geom_bar()` to color the bars.')
  }

  if (!identical(.result$labels$title, "Political Freedom Around the World")) fail("Check the plot title.")
  if (!is.null(.result$labels$x)) fail("Set the x-axis label to `NULL`.")
  if (!identical(.result$labels$y, "Proportion")) {
    fail("Set the y-axis label to \"Proportion\".")
  }
  if (!identical(.result$labels$caption, "Source: Freedom House")) {
    fail("Check the caption. Use \"Source: Freedom House\".")
  }

  x_scales <- Filter(function(s) "x" %in% s$aesthetics, .result$scales$scales)
  labels <- if (length(x_scales) > 0) x_scales[[1]]$labels else NULL
  if (!identical(unname(labels), c("Not Free", "Partly Free", "Free"))) {
    fail("Use `scale_x_discrete()` to label the categories Not Free, Partly Free, and Free in that order.")
  }

  pass("Exactly! Your proportion bar chart shows the three **FHStatus_2022** categories in their substantive order, from Not Free through Free.")
})
```

**Practice 11: Describe the distribution of FHStatus_2022**

```{r question-shape-fhstatus, echo=FALSE}
question("Based on the proportion bar plot, how would you best describe the shape of the distribution of **FHStatus_2022**?",
  correct = "Right. The distribution is unimodal.",
  answer("Unimodal --- Free is the clear modal category", correct = TRUE,
         message = "Free is the modal category, so the distribution has one clear peak. A distribution is unimodal when one category leads --- the modal category does not need to hold a majority of the cases."),
  answer("Uniform --- the three categories are approximately equally common",
         message = "The bars are not equal in height. Free is noticeably the most common category, so the distribution is not uniform."),
  answer("Bimodal --- two categories are similarly prevalent and more common than the third",
         message = "Bimodal requires two categories that are similarly prominent and both standing above the rest. Not Free and Partly Free are close to each other, but both sit well below Free --- similar heights near the bottom do not make two peaks."),
  answer("Multimodal --- each category represents a separate peak",
         message = "Multimodal means several categories are notably more common than the rest. Having observations in three categories does not make a distribution multimodal."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Compare the three bar heights. Is one clearly tallest, are two similar and both above the third, or are all three close?"
)
```

## What to Do When Categories Are Not in Substantive Order

Both variables in this tutorial cooperated. **rural_urban** runs from 1 to 9 and **FHStatus_2022** from 0 to 2, and in both cases the numeric order is the substantive order --- so `factor()` preserved it and the bars in each plot came out in rank order. That will not always be true, and when it is not, a plot is wrong in a way that is easy to miss.

R never guesses what your categories mean. It orders them by how they are stored:

| Stored as | Order R uses | How to check |
|:---|:---|:---|
| character | alphabetical | `sort(unique(x))` |
| factor | its stored levels | `levels(x)` |
| numeric | ascending | read the codebook |

The check is the same question in three forms: what order will R use, and is that the order I mean?

```{r ordinal-survey-setup, include=FALSE}
survey <- data.frame(
  education = c(rep("Less than high school", 12), rep("High school", 31),
                rep("College degree", 24), rep("Beyond college", 8))
)
```

**Run and observe 12: See what alphabetical order does to an ordinal variable**

The code below builds a small survey data frame from scratch, with education recorded as text rather than as codes. Run it and look at the order of the bars.

```{r ordinal-alphabetical-problem, exercise=TRUE, exercise.cap="Run and observe 12", fig.alt="Bar chart of educational attainment with four bars in alphabetical order: Beyond college, College degree, High school, Less than high school. The categories do not run from least to most education, so the shape of the distribution across the ordered scale is impossible to read."}
survey <- data.frame(
  education = c(rep("Less than high school", 12), rep("High school", 31),
                rep("College degree", 24), rep("Beyond college", 8))
)

ggplot(data = survey, mapping = aes(x = education)) +
  geom_bar() +
  labs(title = "Educational Attainment",
       x = NULL,
       y = "Frequency")
```

The categories have an obvious substantive order, and the plot ignores it. Stored as text, they come out alphabetically --- Beyond college, College degree, High school, Less than high school --- which puts the most educated group first and the least educated group last. The shape of the distribution is unreadable, because the order of the bars means nothing.

The fix is to tell R what the order is. Alongside the variable, `factor()` takes a `levels` argument that fixes the sequence:

**Example code:**

```{r factor-levels-template, exercise=FALSE, eval=FALSE}
factor(variable, levels = c("first category", "second category", "third category"))
```

Every category must appear in `levels`, spelled exactly as it is stored --- anything left out becomes `NA` and disappears from the plot.

**Practice 13: Put the categories in substantive order**

Rebuild the same bar plot from the **survey** data frame, but wrap **education** in `factor()` inside `aes()` and give it a `levels` argument listing the four categories from least to most education. Two hints are available. Click "Submit Answer" to check your work.

```{r ordinal-alphabetical-fix, exercise=TRUE, exercise.cap="Practice 13", exercise.setup="ordinal-survey-setup", fig.alt="The same bar chart with the categories in substantive order, from Less than high school through High school and College degree to Beyond college. The bars now rise to a peak at High school and fall away on either side."}
ggplot(data = survey, mapping = aes(x = ___)) +
  geom_bar() +
  labs(title = "Educational Attainment",
       x = NULL,
       y = "Frequency")
```

```{r ordinal-alphabetical-fix-hint-1}
# Hint 1 of 2
# The order you want is the substantive one: from the least education to the
# most. Write the four category labels in that order, exactly as they appear
# in the data.
```

```{r ordinal-alphabetical-fix-hint-2}
# Hint 2 of 2
# Wrap education in factor() and give factor() a second argument, levels,
# set to a character vector of the four labels in the order you want them.
```

```{r ordinal-alphabetical-fix-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!grepl("factor\\s*\\(", code_no_comments)) {
    fail("Wrap **education** in `factor()` inside `aes()`.")
  }
  if (!grepl("levels\\s*=", code_no_comments)) {
    fail("Add the `levels` argument inside `factor()` to set the order of the categories.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing `+` at the end of a line or an unmatched parenthesis.")
  }

  shown <- as.character(
    ggplot2::ggplot_build(.result)$layout$panel_params[[1]]$x$get_labels()
  )
  expected <- c("Less than high school", "High school",
                "College degree", "Beyond college")

  if (!identical(shown, expected)) {
    fail("Check the order inside `levels`. The bars should run from least **education** to most: Less than high school, High school, College degree, Beyond college.")
  }

  pass("Well done! Setting `levels` puts the bars in substantive order, so the shape of the distribution is readable.")
})
```

::: {.caution}
**Numeric codes can be out of order too.** Nothing guarantees that a variable's numeric codes run in substantive order. A survey might code "no opinion" as 5 on a 1-to-4 agreement scale, or assign 1 to high school and 2 to less than high school. R will sort them 1, 2, 3, 4, 5 regardless, and the plot will look orderly while being wrong. This is the case that needs a deliberate check, because nothing on screen looks broken.
:::

## Check Your Understanding

The following questions focus on the main new ideas in this tutorial: what the median adds to your description of an ordered variable, and what it means when the mode and median agree or diverge.

**Apply 14: Choose the right tool**

```{r question-tool-choice-ordinal, echo=FALSE}
question("A researcher wants to describe the central tendency of a 5-point satisfaction scale (1 = very dissatisfied, 5 = very satisfied). Which statistics are appropriate?",
  correct = "Right. Both the mode and the median are appropriate for ordinal variables.",
  answer("The mean only, because the variable is coded numerically",
         message = "A satisfaction scale is ordinal --- the differences between points are not necessarily equal. The mean assumes equal intervals and is not technically appropriate, though it is sometimes used in practice."),
  answer("The mode only, because the variable is categorical",
         message = "The mode is appropriate, but it is not the only option. Because the categories have a meaningful rank order, the median is also available."),
  answer("Both the mode and the median, because the categories have a meaningful rank order", correct = TRUE,
         message = "Both statistics are appropriate for ordinal variables. The mode identifies the most common response and the median locates the midpoint of the distribution. A 5-point satisfaction scale has a meaningful rank order, which makes the median available. Together they give a fuller picture than either alone."),
  answer("Neither the mode nor the median, because the variable has too few categories",
         message = "The number of categories does not determine which statistics are appropriate. Level of measurement does. Both mode and median are appropriate for ordinal variables regardless of how many categories they have."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Ask which statistics the level of measurement makes available, not which one is best. More than one may be appropriate."
)
```

**Apply 15: Interpret a gap between mode and median**

```{r question-synthesis-ordinal, echo=FALSE}
question("A researcher studying job satisfaction reports that for a 5-point scale (1 = very dissatisfied, 5 = very satisfied), the mode is 5 and the median is 3. Which of the following best interprets this result?",
  correct = "Right. The mode and median together reveal something neither statistic shows alone.",
  answer("The typical worker is very satisfied, because 5 is the most common response and the mode identifies the typical case.",
         message = "The mode identifies the most common response, but it does not tell you where the middle of the distribution falls. Here the median is 3, meaning at least half of all workers reported satisfaction of 3 or lower. A mode of 5 with a median of 3 tells a more complicated story than 'the typical worker is very satisfied.'"),
  answer("The most common response is high satisfaction, but at least half of all workers reported 3 or lower.", correct = TRUE,
         message = "When the mode is above the median, the most common response sits higher on the scale than the midpoint of the distribution. Here, 5 is the most common response, but at least half of all workers reported 3 or lower. Reporting only the mode would therefore give an incomplete and potentially misleading picture of worker satisfaction. To say more about the overall pattern --- how concentrated the responses are, or whether they trail off gradually toward the low end --- you would need to see the frequencies across all five categories, not just the mode and median."),
  answer("The distribution is symmetric, because the mode and median are both in the middle of the scale.",
         message = "The median of 3 is in the middle of the scale, but the mode of 5 is at the top. The reported statistics do not establish symmetry. The median sits at the middle of the scale while the modal response sits at the top, so the two summaries are describing different features of the distribution."),
  answer("The median is a better summary than the mode here, because the mode is always less informative for ordinal variables.",
         message = "The mode is not always less informative --- it identifies the most common response, which is meaningful. The point is that neither statistic alone tells the full story. Together they reveal something the mode alone conceals: that the most common response is at the high end while the midpoint of the distribution is considerably lower."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The mode is 5 and the median is 3. Ask what each tells you about where cases sit, and what it means that they disagree."
)
```

**Apply 16: Interpret an unchanged median**

```{r question-interpret-ordinal, echo=FALSE}
question("A 5-category ordinal variable has a median of 3. After several observations move from category 1 to category 2, the median remains 3. What can you conclude?",
  correct = "Right. The median stayed the same, but the distribution changed.",
  answer("The distribution changed even though the median did not --- observations shifted between two categories below the midpoint",
         correct = TRUE,
         message = "The median identifies the midpoint of the ordered distribution, but it is insensitive to changes that happen entirely on one side of that midpoint. Here, observations moved from category 1 to category 2 --- both below the median of 3 --- so the midpoint did not shift. The distribution is less concentrated at the lowest end than it was before, which matters for describing shape and dispersion, but the median alone cannot tell you that. This is exactly why reporting the full distribution alongside the median is more informative than reporting the median alone."),
  answer("The distribution did not change, because the median stayed the same",
         message = "The median staying the same does not mean the distribution stayed the same. Observations moved from category 1 to category 2 --- a real change in the distribution --- but because both categories lie below the median of 3, the midpoint did not shift. The distribution is now less concentrated at the lowest end, even though the median is unchanged."),
  answer("The mode must also be 3, since the median is 3 and both statistics describe the center of the distribution",
         message = "The median and mode are separate measures of central tendency. The median being 3 tells you where the midpoint of the distribution falls; it says nothing about which category is most common. The mode could be any category."),
  answer("The variable should be treated as interval rather than ordinal, since its median is stable",
         message = "The stability of the median has nothing to do with level of measurement. Ordinal variables can have stable medians across different distributions. What makes a variable ordinal is that its categories have a meaningful rank order without necessarily equal spacing between them."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The observations moved from category 1 to category 2 --- both below the median. Ask whether that could move the midpoint."
)
```

**Apply 17: Diagnose a misleading bar plot**

```{r question-alphabetical-order, echo=FALSE}
question("A student builds a bar plot of an education variable with five ordered categories, from less than high school through graduate degree. The code runs without error, and the height of every bar is correct. But the categories appear along the axis in alphabetical order. What is wrong with the graph?",
  correct = "Right. The bars are right; the order they sit in is not.",
  answer("The alphabetical order breaks the ranked scale, so the shape of the distribution cannot be read",
         correct = TRUE,
         message = "Every count is accurate, which is what makes this easy to miss. What alphabetical order destroys is the one thing ordinal measurement adds: with the categories out of rank order, you cannot see whether cases pile up at the low end, thin out toward the high end, or cluster in the middle. A reader looking for the direction across the scale sees a pattern that is an artifact of the alphabet."),
  answer("The bar heights must be wrong, because a correctly built plot orders its categories automatically",
         message = "R orders categories by how they are stored, not by what they mean --- alphabetically for a character variable --- and it does that whether or not the heights are right. The counts here are correct; only their arrangement is not."),
  answer("Nothing is wrong, since the categories are all present and every count is accurate",
         message = "All five categories are there and every count is right. But a bar plot of an ordinal variable is read left to right, and that reading is only meaningful when the categories run in their substantive order."),
  answer("The plot needs a title and axis labels before the distribution can be interpreted",
         message = "Labels matter for communication, but they would not fix this plot. Even fully labeled, the categories would still sit in an order that has nothing to do with education."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The counts are all correct. Ask what a reader loses when ordered categories are arranged by something other than their order."
)
```

## The Takeaways

The table below collects the tools for describing nominal and ordinal variables side by side. The nominal column is carried over from Tutorial 4; central tendency gains the median, and the R functions gain `median()` and `factor()`. Everything that follows explains why, and how to produce these in R. Tutorial 6 adds a third column.

::: {.qpa-infographic}
<table class="qpa-table">
  <caption>Describing Nominal and Ordinal Variables</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Aspect of the distribution</span></th>
      <th scope="col">Nominal Level Variable</th>
      <th scope="col">Ordinal Level Variable</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Central Tendency</th>
      <td>Mode</td>
      <td>Mode, Median</td>
    </tr>
    <tr>
      <th scope="row">Dispersion</th>
      <td>Frequency or Proportion Distribution</td>
      <td>Frequency or Proportion Distribution</td>
    </tr>
    <tr>
      <th scope="row">Distribution Shape</th>
      <td>Unimodal, Bimodal, Multimodal, Uniform</td>
      <td>Unimodal, Bimodal, Multimodal, Uniform; Direction Across the Scale</td>
    </tr>
    <tr>
      <th scope="row">Visualization</th>
      <td>Bar Plot of Frequencies or Proportions</td>
      <td>Bar Plot of Frequencies or Proportions</td>
    </tr>
    <tr>
      <th scope="row">R Functions</th>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>mutate()</code>, <code>filter()</code>, <code>geom_bar()</code></td>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>mutate()</code>, <code>filter()</code>, <code>median()</code>, <code>factor()</code>, <code>geom_bar()</code></td>
    </tr>
  </tbody>
</table>
:::

This tutorial is the middle step in the three-tutorial sequence on univariate description. The same three-part framework applies --- **central tendency**, **dispersion**, and **shape** --- but ordinal measurement adds one new tool: the **median**. Because ordinal categories can be ranked, you can now ask not just which category is most common but where the middle of the distribution falls.

For ordinal variables, the three dimensions are described as follows:

- [**Central tendency**]{.important-text} is described with both the [**mode**]{.important-text} and the [**median**]{.important-text}. The mode identifies the most common category; the median identifies the midpoint of the ordered distribution. In R, `median(data$variable)` gives you the median directly, with `na.rm = TRUE` added when the variable has missing values --- without it, a single `NA` makes the whole result `NA`. When the mode and median agree, the most common category and the midpoint of the distribution coincide. When they diverge, as they did for **FHStatus_2022**, that difference is part of the story and should be explained.

- [**Dispersion**]{.important-text} is described with frequency and proportion tables. As established in Tutorial 4, filter out missing values before computing proportions whenever a variable has any --- an NA group would otherwise be included in the denominator, making the substantive proportions sum to less than 1. After filtering, use `group_by()`, `summarise()`, and `mutate()` to count cases and calculate proportions. The more spread out the cases are across categories, the less you should rely on the mode or median alone.

- [**Shape**]{.important-text} is easiest to see in a bar plot. For ordinal variables, the bars must appear in the substantive rank order of the categories, because ordered categories add a [**direction across the scale**]{.important-text} to the modal pattern: cases can pile up at one end, thin out toward the other, or cluster in the middle. A distribution is [**unimodal**]{.important-text} when one category clearly leads, [**bimodal**]{.important-text} when two categories are similarly prominent, [**uniform**]{.important-text} when cases are spread approximately evenly, and [**multimodal**]{.important-text} when several categories stand out. If the numeric codes already run in the correct direction, `factor()` preserves that ordering. When they do not --- or when the categories are stored as text, which R sorts alphabetically --- give `factor()` a `levels` argument listing every category in substantive order, spelled exactly as it is stored. Check first what order R will use: `levels()` for a factor and `sort(unique())` for a character variable --- and for numeric codes, nothing guarantees they run in the order you expect. Use `scale_x_discrete()` to add readable labels while maintaining the rank order. Use `coord_flip()` when category labels are long.

In R, load [dplyr]{.package-name} and [ggplot2]{.package-name} once at the top of your script or Quarto document with `library(dplyr)` and `library(ggplot2)` before any code that uses functions from those packages. When building a proportion bar chart, filter out missing values before passing data to `ggplot()` if the variable has any --- for the same reason as proportion tables, an NA bar would otherwise be included in the proportion computation, making the substantive proportions sum to less than 1. Pipe the filtered data frame directly into `ggplot()` --- when data arrive via a pipe, the `data =` argument is supplied automatically and only `mapping =` is needed. As in Tutorial 4, `geom_bar()` draws the bars, and `aes(y = after_stat(prop), group = 1)` inside it turns counts into proportions.

In Tutorial 6, you will move to interval variables, where the distances between values are equal and meaningful. That additional property makes the **mean** available as a measure of central tendency --- the payoff for all the "not appropriate for ordinal" caveats in this tutorial. The median remains useful for interval variables too, but the mean becomes the primary tool.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-05, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-05, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-05, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Great work! You can now describe and visualize ordinal variables using their substantive rank order.">
  <span aria-hidden="true">🎉</span>
  Great work! You can now describe and visualize ordinal variables using their substantive rank order.
</p>
')
```
