---
title: "Simple Regression"
tutorial:
  id: "15-Simple-Regression"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Estimating a line, reading its two coefficients, and using a factor as a predictor.
---

## Overview {#main-content}

News reports, campaign strategy and political research are full of questions about how much: how much more legislation gets passed in Congresses where the president's party holds more Senate seats, how many percentage points of turnout separate states that register voters on election day from states that do not, how much of the difference in party support between older and younger voters any single explanation can account for. Each asks for a number rather than a yes or no, and each is a comparison across cases --- places, states or years that differ, with outcomes that differ alongside them.

This tutorial introduces **simple linear regression**, a model with one interval-level outcome variable and one variable used to explain it. Where Tutorials 12 through 14 asked whether a relationship holds up against chance, regression estimates a line through the data and reports two numbers that say how large the relationship is, what the model predicts for any case you name, and how much of the variation it leaves unexplained.

In this tutorial you will learn:

* How to use the `lm()` function to estimate a regression model and the `summary()` function to display the results
* How to interpret the intercept and slope when the predictor is measured at the interval level
* How to interpret the slope when the predictor is binary, and how that differs from interpreting a slope on an interval predictor
* How to calculate the predicted value of the outcome for a given value of the predictor
* How to use the regression output to test hypotheses about the intercept and slope, interpreting the t-statistic and p-value for each coefficient
* How to calculate and interpret a residual, and use the residual standard error to judge how large the residuals are
* How to assess model fit using $R^2$ and the residual standard error
* How to create a labeled factor variable with the `factor()` function and use it as a binary predictor in a regression

::: {.important-note}
**Before you begin.** This tutorial follows the reading **Simple Regression**, which introduces the ideas it puts to work. [Open the reading](qpa-readings/SimpleRegression.html){target="_blank"} in a new tab, or run `open_reading("simple-regression")` in the R console.
:::

::: {.tip}
**A note on language throughout the regression tutorials.** In regression, explanatory variables are commonly called [predictors]{.important-text} because the model uses them to predict values of the outcome, and a slope is commonly called the [effect]{.important-text} of its predictor. These tutorials use *predictor*, *predicts* and *effect* in this statistical sense; none of them by itself implies that one variable causes the other. A predictor with only two categories also goes by several names --- *dichotomous*, *dummy*, and [binary]{.important-text} --- all three of which appear in published work. These tutorials use *binary*.
:::

```{r setup, include=FALSE}
# Make the packaged readings reachable from inside the tutorial
local({
  readings <- system.file("readings", package = "QPATutorialsCourse")
  if (nzchar(readings)) shiny::addResourcePath("qpa-readings", readings)
})
library(learnr)
library(dplyr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
options(lifecycle_verbosity = "quiet")
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "QPATutorialsCourse")
data("fHouse", package = "QPATutorialsCourse")
counties$majority_white_f <- factor(counties$majority_white, labels = c("Not Majority White", "Majority White"))
```

## Estimating and Interpreting a Regression Model

[Simple linear regression]{.important-text} estimates a line through the data. When the predictor is interval, the slope gives the expected change in the outcome for a one-unit increase in the predictor --- information that correlation does not provide. Examples 1 and 2 illustrate this case. When the predictor is a binary categorical variable, the slope equals the difference between the two group means, while the intercept equals the mean of the reference group --- the same comparison made by a difference-of-means test, though estimated through a regression model. Example 3 illustrates that connection.

A note on the data. The data frames **counties** and **fHouse** include nearly all US counties and most of the world's countries for the relevant year rather than random samples like those used in public opinion surveys. The hypothesis tests in this tutorial still ask whether the observed pattern is consistent with a more general relationship that could recur across comparable places or periods. That interpretation treats the observed cases as one realization of a broader political or social process.

### The Regression Equation

In the population, the relationship between a predictor, $X$, and an outcome, $Y$, is written as

$$Y = \alpha + \beta X + \epsilon,$$

where $\alpha$ is the intercept, $\beta$ is the slope, and $\epsilon$ is an error term capturing everything besides $X$ that affects $Y$. No relationship in real data is ever perfect --- counties with the same wage growth cast different vote shares, countries with the same GDP have different levels of corruption --- and $\epsilon$ is what accounts for that. You never observe $\alpha$, $\beta$, or $\epsilon$ directly; you estimate them from the data.

The fitted line for the sample is written as

$$\hat{Y} = \hat{\alpha} + \hat{\beta} X,$$

where the hats mark estimated rather than true values. The [intercept]{.important-text} ($\hat{\alpha}$) is the predicted value of $Y$ when $X = 0$. The [slope]{.important-text} ($\hat{\beta}$) is the predicted change in $Y$ for a one-unit increase in $X$. The gap between an observed value of $Y$ and the value the line predicts for that case is called a **residual**.

### Estimating a Model with `lm()`

The `lm()` function in base R estimates a linear regression model. It takes two main arguments. The first is a formula: the outcome variable, followed by a tilde (`~`), read as "is modeled as a function of," followed by the predictor. So `y ~ x` means "$y$ is modeled as a function of $x$." The second is `data`, which tells R which data frame to search for the variables. Because you supply `data`, you write bare variable names in the formula --- not `dataframe$variable`.

`lm()` returns a model object containing the estimated coefficients and everything else R computed. Assign it to a named object with `<-` so you can refer to it again, then pass that object to `summary()` to display the formatted results.

### Reading the Output

`summary()` produces several pieces of output. Near the top, `Call` repeats the model formula, so you can confirm it matches what you intended. The `Coefficients` table is the part you will use most. It has one row for the `(Intercept)` and one row for each predictor, with four columns:

- `Estimate`: the estimated intercept or slope
- `Std. Error`: the uncertainty in that estimate
- `t value`: the test statistic for testing whether that coefficient is zero in the population
- `Pr(>|t|)`: the p-value for that test

R also prints significance stars next to each p-value as a visual shorthand: `***` for p < .001, `**` for p < .01, `*` for p < .05, and `.` for p < .1. A key at the bottom of the output defines them. The stars are a quick visual cue, but always rely on the actual p-value rather than the stars alone --- the stars tell you which threshold a p-value clears, not its exact value.

Below the coefficients table, the [residual standard error]{.important-text} gives the estimated standard deviation of the residuals --- the approximate typical distance between observed outcomes and the fitted regression line. [R-squared]{.important-text}, printed as `Multiple R-squared`, gives the proportion of the variance in $Y$ the model explains; a value of 0.20 means the model accounts for 20% of the variation in the outcome. Finally, there is an `F-statistic` at the very bottom; with only one predictor it tests the same thing as the slope's t-statistic, so it adds nothing here. It becomes useful once a model has several predictors.

The three examples that follow put all of this to work. Examples 1 and 2 use an interval predictor --- one from American electoral politics, one from a cross-national survey --- and Example 3 introduces a binary predictor, connecting regression back to the difference-of-means framework of Tutorial 13.

## Is Wage Growth Associated with Clinton's Vote Share?

Tutorial 14 finds a statistically significant positive correlation between wage growth and Clinton's two-party vote share --- counties where wages grew faster gave her more of the vote, consistent with retrospective voting theory. That test provided statistically significant evidence of a positive relationship. Regression now lets you ask a more precise question: how much does Clinton's vote share change for each percentage-point increase in wage growth? The hypothesis captures that directional prediction:

::: {.hypothesis}
In a comparison of counties, those with higher wage growth will cast a higher percentage of the two-party vote for Hillary Clinton than those counties with lower wage growth.
:::

The data frame **counties** contains **wage_growth**, the percent by which wages grew or fell in a county between the third quarter of 2015 and the third quarter of 2016, and **dem2p_percent**, the percent of the two-party presidential vote cast for Hillary Clinton in the November 2016 election. Both are measured in percentage points, so a one-unit increase in either is a one-percentage-point increase. You will estimate a regression with **dem2p_percent** as the outcome variable and **wage_growth** as the predictor.

**dem2p_percent** has no missing values. **wage_growth** is missing for 2 of the 3,112 counties, so any summary you compute from it describes the 3,110 counties with a recorded value rather than all of them.

### Estimate the Regression

Let's put `lm()` to work on this hypothesis.

**Run and observe 1: Estimate a regression of Clinton's two-party vote percent on wage growth**

```{r exercise-regression-voteclinton-wagegrowth, exercise=TRUE, exercise.cap="Run and observe 1"}
model1 <- lm(dem2p_percent ~ wage_growth, data = counties)
summary(model1)
```

The model object here is called `model1`. In these tutorials the name is always given to you, because the exercises check for it; in your own work you can call a model whatever you like, as long as you use the same name every time you refer to it.

### Interpret the Estimated Coefficients

The Coefficients table translates the regression line into substantive terms --- two numbers that together describe the relationship between wage growth and Clinton's vote. The questions below ask what each one means for counties in this data.

**Apply 2: Interpret the intercept**

```{r quiz-interpret-intercept-voteclinton-wagegrowth, echo=FALSE}
question("Look at the Coefficients table in your output. Which statement correctly describes what the intercept tells us about this model?",
  correct = "Right. The intercept is the expected percent of the two-party vote for Clinton when wage growth is zero.",
  answer("The intercept gives the expected percent of the two-party vote for Clinton in a county where wage growth is zero",
         correct = TRUE,
         message = "The estimate is 31.3521 --- in counties where wage growth was flat, we expect about 31.4% of the two-party vote to go to Hillary Clinton, the Democratic candidate. Because zero is a value **wage_growth** actually takes in the data, this is a substantively meaningful prediction, not just a mathematical formality."),
  answer("The intercept predicts how much Clinton's percent of the two-party vote changes for a one-percentage-point increase in wage growth",
         message = "That is the slope's job, not the intercept's. The intercept is the predicted value of the outcome when the predictor equals zero --- it is the starting point of the regression line. Look at the **wage_growth** row for the rate of change."),
  answer("The intercept gives the expected percent of the two-party vote for Clinton in a county with one percentage point of wage growth",
         message = "The intercept is the predicted value when the predictor is zero, not one. A county with one percentage point of wage growth would have a predicted percent of the two-party vote for Hillary Clinton higher than the intercept by the amount of the slope --- that is the fitted line evaluated at X = 1, not the intercept."),
  answer("The intercept gives the expected level of wage growth when Clinton's percent of the two-party vote is zero",
         message = "This reverses the outcome and predictor variables. We modeled Clinton's percent of the two-party vote as a function of wage growth --- so the intercept describes the outcome (vote percent) at a given value of the predictor (wage growth = 0), not the predictor at a given value of the outcome."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the intercept is in the row labeled (Intercept). What does it mean when the predictor equals zero?"
)
```

**Apply 3: Interpret the slope**

```{r quiz-interpret-slope-voteclinton-wagegrowth, echo=FALSE}
question("Look at the Coefficients table in your output. Which statement correctly describes what the slope tells us?",
  correct = "Right. The slope predicts how much Clinton's percent of the two-party vote changes for a one-percentage-point increase in wage growth.",
  answer("The slope predicts how much Clinton's percent of the two-party vote changes for a one-percentage-point increase in wage growth",
         correct = TRUE,
         message = "The estimate is 0.45582 --- in counties where wages grew one percentage point faster, Hillary Clinton received about 0.46 percentage points more of the two-party vote, on average. The positive sign matches the hypothesis, though the effect is modest."),
  answer("The slope gives the expected change in Clinton's percent of the two-party vote for each one-percent increase in wage growth",
         message = "**wage_growth** is already measured in percentage points. A one-unit increase is a one-percentage-point increase --- moving from, say, 4% to 5% --- not a 1% relative increase on top of whatever it already was. Percentage points and percents are different units."),
  answer("The slope gives the expected percent of the two-party vote for Clinton when wage growth is zero",
         message = "That description belongs to the intercept, not the slope. The slope tells us how Clinton's vote percent changes as wage growth changes --- it is the rate of change. The starting value when wage growth is zero is the intercept."),
  answer("The slope gives the expected change in wage growth for each one-percentage-point increase in Clinton's percent of the two-party vote",
         message = "This reverses the outcome and predictor variables. We modeled **dem2p_percent** as a function of **wage_growth**, not the reverse. The reverse regression would produce a different slope entirely."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the slope is in the row labeled **wage_growth**. What does it mean for a one-unit change in wage growth?"
)
```

::: {.tip}
**Percentage points and percents are not the same thing.** A variable moving from 20 to 21 has risen by one percentage point; a 1% relative increase would take it only to 20.2. A slope always describes the first. Watch for this in any variable whose name ends in *percent* or *rate*.
:::

### Calculate Predicted Values

The regression line isn't just a summary of the data --- it's a prediction machine. Substituting any value of **wage_growth** into the fitted line gives the expected percent of the two-party vote for Clinton in a county with that level of wage growth. At the average value of **wage_growth**, 4.3:

$$\widehat{\text{dem2p_percent}} = \hat{\alpha} + \hat{\beta}\,\text{wage_growth} = 31.3521 + 0.45582\times(4.3) = 33.31213.$$

Notice what this arithmetic reveals about the slope. If wage growth were 5.3 instead --- one percentage point higher --- the prediction would be $31.3521 + 0.45582 \times 5.3 = 33.768$. The difference between the two predictions is exactly 0.456 --- the slope. That is not a coincidence: the slope *is* the expected change in $\hat{Y}$ for a one-unit increase in X, by definition. For any other change of size $c$, the expected change in $\hat{Y}$ is $\hat{\beta} \times c$ --- the same multiplication, scaled to the size of the change you care about.

**Practice 4: Calculate a predicted Clinton vote share**


```{r quiz-prediction-voteclinton-wagegrowth, echo=FALSE}
question("Using the intercept and slope from your output, what is the predicted percent of the two-party vote for Hillary Clinton in a county with wage growth at the 75th percentile, 6.5?",
  correct = "Right. Substituting into the fitted line: 31.3521 + 0.45582 * 6.5 = 34.315.",
  answer("About 34.3 percentage points", correct = TRUE,
         message = "$\\hat{Y} = 31.3521 + 0.45582\\times 6.5 = 34.31493$. This is only about one percentage point above the prediction at the mean."),
  answer("About 29.6 percentage points",
         message = "Check the calculation: substitute 6.5 for **wage_growth** in the fitted equation. Since 6.5 is above the mean (4.3) and the slope is positive, the prediction should be above the prediction at the mean (33.31), not below it."),
  answer("About 42.5 percentage points",
         message = "Check the order of operations: multiply the slope by 6.5 first, then add the intercept. The slope is small (about 0.456), so a wage growth value of 6.5 does not push the prediction very far above the intercept."),
  answer("About 31.4 percentage points",
         message = "That is the intercept --- the predicted value when wage growth is zero. Substituting 6.5 into the fitted line requires multiplying the slope by 6.5 and adding the intercept, not reading the intercept alone."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: substitute 6.5 into the fitted equation $\\hat{Y} = \\hat{\\alpha} + \\hat{\\beta} \\times 6.5$ using the estimates from your output."
)
```

### Calculate a Residual

A predicted value tells you what the line expects. A [residual]{.important-text} tells you how far off it was for a county that actually exists --- the difference between the observed value and the value the line predicted.

Take Centre County, Pennsylvania. Its wage growth was 4.672224 and Hillary Clinton received 51.25% of its two-party vote. The wage growth figure is carried at full precision rather than rounded, so that the residual you calculate matches the one the model computes. The fitted line predicts:

$$\hat{Y} = 31.3521 + 0.45582 \times 4.672224 = 33.48.$$

The residual is what is left over:

$$\hat{\epsilon} = Y - \hat{Y} = 51.25 - 33.48 = 17.77.$$

Clinton did about 17.8 percentage points better in Centre County than a model built on wage growth alone expected. A positive residual means the model [underpredicted]{.important-text}; a negative one means it [overpredicted]{.important-text}.

**Practice 5: Calculate a residual**

Now do the same for a different county. Martin County, Kentucky sits in the eastern coalfield and voted Democratic for generations; in 2016 it gave Clinton 9.39% of the two-party vote. There is no function for this yet, so use R as a calculator: read the intercept and slope off your `summary(model1)` output, type them into an expression that computes Martin County's predicted value, and subtract that from the observed value. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-residual-centre, exercise=TRUE, exercise.cap="Practice 5"}
# Martin County, Kentucky: wage_growth = 6.4206898, dem2p_percent = 9.386503
# Use the intercept and slope from your regression output.


```

```{r exercise-residual-centre-hint-1}
# Hint 1 of 3
# A residual is the error in a prediction for a single case:
# how far the observed value sits from the value the line
# predicts. You need two numbers --- what the model expected
# for Martin County, and what actually happened there.
```

```{r exercise-residual-centre-hint-2}
# Hint 2 of 3
# The predicted value comes from the fitted line: the intercept
# plus the slope times Martin County's wage growth. Both
# estimates are in the Coefficients table from your regression.
```

```{r exercise-residual-centre-hint-3}
# Hint 3 of 3
# The order of the subtraction matters. Observed minus predicted
# --- so a negative answer means the model predicted too high.
```

```{r exercise-residual-centre-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("\\*", code_no_comments)) {
    fail("Build the predicted value from the fitted line first: the slope multiplied by Martin County's wage growth, added to the intercept. Then subtract that prediction from the observed value.")
  }
  if (!grepl("31\\.3", code_no_comments) || !grepl("0\\.45", code_no_comments)) {
    fail("Use the intercept and the slope from the output --- 31.35 and 0.456 --- to build the predicted value, rather than entering a number you worked out elsewhere.")
  }
  if (!grepl("-", code_no_comments)) {
    fail("Compute the residual in the code box rather than entering a number: subtract the predicted value from the observed value.")
  }
  if (is.null(.result)) {
    fail("Your code did not produce a result. Make sure the last line prints the residual.")
  }
  value <- suppressWarnings(as.numeric(.result))
  if (length(value) != 1 || is.na(value)) {
    fail("Your code should produce a single number --- the residual for Martin County.")
  }
  if (isTRUE(all.equal(value, 24.89, tolerance = 0.05))) {
    fail("Check the order of the subtraction. A residual is the observed value minus the predicted value, not the other way around.")
  }
  if (isTRUE(all.equal(value, 34.28, tolerance = 0.05))) {
    fail("That is the predicted value. The residual is what is left after subtracting the predicted value from the observed value.")
  }
  if (!isTRUE(all.equal(value, -24.89, tolerance = 0.05))) {
    fail("Not quite. Compute the predicted value from the fitted line first, then subtract it from the observed value.")
  }
  pass("Right. The model overpredicts Clinton's vote share in Martin County by about 24.9 percentage points --- the opposite direction from Centre County, and a larger miss.")
})
```

Every county has a residual, and R will compute all of them at once. `fitted()` returns the predicted value for each case; `resid()` returns the residual. Both come from the model object, not from the data frame --- and that distinction matters here. Two counties are missing **wage_growth**, so `lm()` fitted the model on 3,110 counties while **counties** has 3,112 rows. The residuals will not line up with the full data frame until you drop those two rows yourself.

One detail about how place names are recorded. **county_name** holds the county's name without the word "County" --- "Centre", not "Centre County" --- and **state_abb** holds the two-letter abbreviation, "PA". Both are recorded for every county. The filter below uses both: county names repeat across states, and this data has three counties named Cook and two named Harris.

The code below uses `filter()` and `mutate()` from [dplyr]{.package-name}.

**Run and observe 6: Compare residuals across counties**

```{r exercise-residuals-compare, exercise=TRUE, exercise.cap="Run and observe 6"}
library(dplyr)  # first use in this tutorial; it stays loaded from here on

model1 <- lm(dem2p_percent ~ wage_growth, data = counties)

counties_used <- filter(counties, !is.na(wage_growth))
counties_used <- mutate(counties_used,
                        predicted = fitted(model1),
                        residual  = resid(model1))

filter(counties_used,
       (county_name == "Centre" & state_abb == "PA") |
       (county_name == "Bronx"  & state_abb == "NY"))
```

**Apply 7: Interpret the two residuals**

```{r quiz-residual-interpretation, echo=FALSE}
question("Centre County's residual is about +17.8 and the Bronx's is about +56.4, while the residual standard error for this model is 15.99. Which statements are correct? Select all that apply.",
  correct = "Right. Centre County is a typical miss for this model; the Bronx is an extreme one.",
  answer("The model underpredicts Clinton's vote share in both counties",
         correct = TRUE,
         message = "Both residuals are positive, so Clinton did better in both counties than wage growth alone predicts."),
  answer("Centre County's residual is close to a typical miss for this model",
         correct = TRUE,
         message = "The residual standard error, 15.99, is roughly the typical size of a residual. Centre County's 17.8 is right about there --- and being wrong by 18 percentage points is what typical looks like for this model."),
  answer("The Bronx is a county the model fits poorly even by this model's standards",
         correct = TRUE,
         message = "A residual of 56.4 is more than three times the typical miss. Wage growth tells us almost nothing about how the Bronx voted."),
  answer("A positive residual means the model overpredicted Clinton's vote share",
         message = "It means the opposite. The residual is the observed value minus the predicted value, so a positive value means the observed vote share was higher than the model expected --- the model underpredicted."),
  answer("Because both residuals are positive, the model is biased upward",
         message = "Two counties cannot tell us that. Residuals from a fitted line always sum to zero, so for every positive residual there are negative ones elsewhere --- Carlisle County, Kentucky, has one of about -42."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there are three correct statements. Compare each residual to the residual standard error, and check what the sign means."
)
```

::: {.tip}
**A note on assumptions.** Linear regression rests on a set of assumptions --- about the form of the relationship, the behavior of the errors, and the absence of certain kinds of measurement problems. The reading lists them, and residuals are the main tool for checking them: patterns in a plot of residuals are what reveal an assumption that has failed. This course does not go further than that. You will not test the assumptions or diagnose violations, and you should not assume they hold for the models in these tutorials --- for several of them they almost certainly do not. What matters here is knowing that the assumptions exist, that they can fail, and that residuals are where you would look.
:::

### Conduct Hypothesis Tests

The columns in the `summary()` output under `Coefficients` give everything a hypothesis test needs: the estimate, its standard error, the t-statistic, and the p-value. The alternative in these tutorials is always two-tailed --- the coefficient is not equal to zero --- even when your theoretical hypothesis is directional, because that is standard practice for regression coefficients. This is more conservative than a one-tailed test --- but if you reject the two-tailed null and the sign of the coefficient matches your theoretical prediction, you have implicitly rejected the one-tailed null as well, and you can still make a directional claim in your substantive interpretation.

As in Tutorials 12, 13, and 14, hypotheses are stated in terms of population parameters, not sample statistics. In regression, the population parameters are $\alpha$ (the true intercept) and $\beta$ (the true slope) --- the quantities the hypotheses are about. The estimated values $\hat{\alpha}$ and $\hat{\beta}$ are sample statistics that estimate them. So the null and alternative hypotheses are stated in terms of $\alpha$ and $\beta$, not $\hat{\alpha}$ and $\hat{\beta}$.

The intercept test is largely a formality here --- what really matters is whether the slope is different from zero in the population, which is what the test on **wage_growth** addresses.

From here on the five-step procedure is worked through in prose rather than enumerated one step at a time.

**The intercept**

The null hypothesis is $H_0: \alpha = 0$ and the alternative is $H_A: \alpha \ne 0$. Find the t-statistic and p-value for the intercept in your output.

**Apply 8: Test the intercept**


```{r quiz-hypothesistesting-voteclinton-wagegrowth-intercept-5, echo=FALSE}
question("Based on the t-statistic and p-value for the intercept in your output, what conclusion do we draw?",
  correct = "Right. With a p-value far below 0.05 we reject the null hypothesis that the intercept is zero.",
  answer("We reject the null hypothesis that the intercept is zero",
         correct = TRUE,
         message = "The t-statistic is 80.249 and the p-value is less than 2e-16 --- overwhelming evidence against the null. Substantively, this just confirms that counties don't have a baseline percent of the two-party vote for Clinton of literally zero when wage growth is flat --- not a surprising finding, and not something the slope's interpretation depends on."),
  answer("We fail to reject the null hypothesis that the intercept is zero",
         message = "Look at the p-value for the intercept in your output. If it is below 0.05, we reject the null. Apply the decision rule: p < 0.05 → reject."),
  answer("We reject the null hypothesis, which proves that the intercept is not zero in the population",
         message = "Rejecting the null hypothesis provides statistically significant evidence that the intercept is not zero --- but it does not prove it. A p-value tells us how surprising the result would be if the null were true; a very small p-value is strong evidence against the null, not certainty that the alternative is correct."),
  answer("We cannot draw a conclusion about the intercept because the intercept test is not substantively meaningful",
         message = "We can always test any coefficient statistically, including the intercept --- the output provides a t-statistic and p-value for it just as it does for the slope. Whether the intercept test is substantively interesting is a separate question from whether it can be tested."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: apply the decision rule from Tutorial 11 --- if the p-value is below 0.05, we reject the null hypothesis."
)
```

**The slope**

The null hypothesis is $H_0: \beta_{WG} = 0$ and the alternative is $H_A: \beta_{WG} \ne 0$, where $\beta_{WG}$ is the population slope on **wage_growth**. Find the t-statistic and p-value for the slope in your output.

**Apply 9: Test the slope**

```{r quiz-hypothesistesting-voteclinton-wagegrowth-slope-34, echo=FALSE}
question("Which conclusions follow from the hypothesis test on the slope? Select all that apply.",
  correct = "Right. Both the p-value and the t-statistic lead us to reject the null hypothesis that the slope is zero.",
  answer("The p-value is less than 0.05, so we reject the null hypothesis --- there is a statistically significant relationship between wage growth and Clinton's percent of the two-party vote",
         correct = TRUE,
         message = "The p-value is 1.39e-13, far below 0.05. The data provide evidence against a zero relationship in the population."),
  answer("The absolute value of the t-statistic exceeds 1.96, which leads to the same conclusion as the p-value",
         correct = TRUE,
         message = "The t-statistic is 7.431, well above 1.96. The t-statistic and p-value always agree for the same two-tailed test, since both are calculated from the same estimate and standard error."),
  answer("We fail to reject the null hypothesis because the slope is small",
         message = "The size of the coefficient alone does not determine statistical significance. A small coefficient with a small standard error can be highly significant. What matters is the t-statistic and the p-value, not the magnitude of the estimate itself."),
  answer("The p-value gives the probability that the null hypothesis is true, so a p-value below 0.05 means there is less than a 5% chance the slope is zero in the population",
         message = "This is a common misreading of the p-value. The p-value gives the probability of observing a t-statistic this large *if the null hypothesis were true* --- it does not give the probability that the null hypothesis is true. A small p-value means the data would be surprising under the null, not that the null is probably false."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there is more than one correct answer. Check both the p-value and the t-statistic in the **wage_growth** row."
)
```

### Assess Model Fit

A significant slope tells you the relationship is statistically distinguishable from zero. But how well does this model actually predict which counties gave Clinton more of their vote? The fit statistics at the bottom of the output address that question.

**Apply 10: Assess the fit of the wage growth model**

Find `Multiple R-squared` and `Residual standard error` at the bottom of your output before answering.

```{r quiz-modelfit-voteclinton-wagegrowth, echo=FALSE}
question("Which of the following statements about model fit are correct? Select all that apply.",
  correct = "Right. An R-squared below 0.02 and a residual standard error of about 16 percentage points both indicate very poor fit.",
  answer("The Multiple R-squared shows that wage growth explains less than 2% of the variation in Clinton's percent of the two-party vote across counties",
         correct = TRUE,
         message = "The Multiple R-squared is 0.01746 --- wage growth accounts for almost none of the county-to-county variation in Clinton's vote percent. A single economic variable was never going to tell the whole story of how 3,100 counties voted."),
  answer("The Multiple R-squared shows that wage growth explains most of the variation in Clinton's percent of the two-party vote, leaving only about 2% unexplained",
         message = "This reads R-squared backwards. An R-squared of 0.01746 means wage growth explains about 1.75% of the variation, not that 1.75% is left unexplained. About 98% of the variation in Clinton's vote percent is not accounted for by wage growth alone."),
  answer("The residual standard error shows that in a typical county, the model's prediction of Clinton's percent of the two-party vote is off by about 16 percentage points",
         correct = TRUE,
         message = "The residual standard error is 15.99 percentage points --- roughly the size of a substantial swing in a county's vote percent. That is a large typical error, consistent with the near-zero R-squared. The model detects a statistically distinguishable relationship but predicts individual counties poorly."),
  answer("The residual standard error shows that the model explains about 16% of the variation in Clinton's percent of the two-party vote",
         message = "The residual standard error is not a proportion --- it is measured in the same units as the outcome, here percentage points. It gives the approximate typical distance between observed outcomes and the fitted regression line, not the share of variation explained. That is what R-squared does."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there are two correct statements --- one about R-squared and one about the residual standard error. Both point to very poor model fit."
)
```

### Reporting the Results

So is wage growth associated with Clinton's vote share? The short answer is yes, but the longer answer is more interesting.

The shaded box below is a write-up: the way this result would be reported in a paper or a research memo, rather than the way it is worked out in a tutorial. Each example in this tutorial ends with one. Read them for the shape of the argument as much as for the finding --- what gets stated, in what order, and how much is claimed.

::: {.research-example}
Wage growth had a positive relationship with the percent of the two party vote Clinton received in 2016 --- counties where wages grew faster gave her more of the vote, consistent with retrospective voting theory. The slope is highly significant: with over 3,100 counties, we are very confident the positive relationship is statistically distinguishable from zero. But the model fit is very poor. The R-squared of 0.017 means wage growth accounts for less than 2% of the county-to-county variation in Clinton's vote, and the residual standard error of about 16 percentage points --- the approximate typical distance between observed vote shares and the fitted regression line --- is large. A significant slope and poor fit are not in tension --- they are answering different questions. The slope tells us the relationship is statistically distinguishable from zero; R-squared tells us how much of the variation it explains. In political science, where outcomes have many causes, a single predictor almost never explains most of the variation, and that is not a sign that something went wrong.
:::


## Is National Freedom Associated with Approval of a Strong Leader?

The wage growth example showed regression at work on a question about US counties. The same logic applies to any research question with an interval outcome variable --- different data, different variables, different part of the world, same tool. Here the question is whether a country's level of political freedom is associated with its citizens' appetite for concentrated executive power.

Is a country's level of freedom associated with whether its citizens approve of an unconstrained strong leader --- one who doesn't have to bother with elections or a legislature? The theoretical argument is that living under genuinely free institutions builds support for the constraints those institutions impose. Citizens who experience functioning elections, civil liberties, and checks on executive power develop both an appreciation for what those constraints protect and a wariness of leaders who would dismantle them. Countries with lower levels of freedom, conversely, may have populations more accustomed to --- or more open to --- concentrated executive power.

::: {.hypothesis}
In a comparison of countries, those with higher levels of political freedom will have lower public approval of an unconstrained strong leader than those with lower levels of political freedom.
:::

The data frame **fHouse** contains **FHTotalAggr_2016**, a Freedom House score combining political rights and civil liberties (roughly 0, least free, to 100, most free), and **Pro.Strong.Leader**, the percent of a country's survey respondents who approve of a strong leader who doesn't have to bother with elections or a legislature. You will model **Pro.Strong.Leader** as a function of **FHTotalAggr_2016**.

The two come from different years, and deliberately so. The Freedom House scores assess 2016; the survey responses come from World Values Survey Wave 7, fielded between 2017 and 2023. So freedom is measured *before* the attitudes it is being asked to predict, which is the order a claim like this one needs. A predictor measured after its outcome cannot explain it, whatever the regression output says.

::: {.caution}
**Heads up on missing data.** **FHTotalAggr_2016** is missing for 1 of the 194 countries in **fHouse**. **Pro.Strong.Leader** is missing for 132 of them --- Wave 7 was fielded in only 62 countries, so two thirds of the file has no survey data at all. R drops any country missing either variable when estimating the model, which leaves 62. You'll see that noted at the foot of the output.
:::

### Estimate the Regression

The hypothesis makes a clear prediction --- freer countries should show less appetite for unconstrained leadership. Whether that holds across countries is an empirical question. Let's see what a simple regression model shows.

**Practice 11: Estimate a regression of strong leader approval on national freedom**

Fill in the blanks to estimate the model, assigning it to **model2**, then display the results. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-regression-strongleader-freedom, exercise=TRUE, exercise.cap="Practice 11"}
model2 <- ___

summary(model2)
```

```{r exercise-regression-strongleader-freedom-hint-1}
# Hint 1 of 2
# Look back at the hypothesis. Which variable is the outcome
# we are trying to explain, and which variable do we think
# drives it? In a regression formula, the outcome goes on
# the left of the ~ and the predictor goes on
# the right.
```

```{r exercise-regression-strongleader-freedom-hint-2}
# Hint 2 of 2
# lm() estimates the model. Its first argument is the formula --- the
# outcome, a tilde, then the predictor --- and data names the frame
# holding them, which here is fHouse. Save the model to a named object,
# then call summary() on it so the summary is what prints.
```

```{r exercise-regression-strongleader-freedom-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (is.null(.result)) {
    fail("Your code did not produce a result. Check for a missing tilde (~) in the formula, a misspelled variable name, or unmatched parentheses.")
  }
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use `lm()` to estimate the regression.")
  }
  if (!grepl("model2\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model2.")
  }
  if (!grepl("Pro\\.Strong\\.Leader\\s*~\\s*FHTotalAggr_2016", code_no_comments)) {
    fail("The formula should be `Pro.Strong.Leader ~ FHTotalAggr_2016` --- approval of a strong leader as a function of national freedom. Check that the outcome variable is on the left of the ~ and the predictor is on the right.")
  }
  if (!grepl("data\\s*=\\s*fHouse", code_no_comments)) {
    fail("Set `data = fHouse` so R knows where to find the variables.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call `summary()` on your saved model object to display the results.")
  }

  expected <- lm(Pro.Strong.Leader ~ FHTotalAggr_2016, data = fHouse)
  if (!inherits(.result, "summary.lm")) {
    fail("The last line should display the summary. Save the model with `lm()`, then call `summary()` on it so the summary is what prints.")
  }
  result_coef <- setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling `summary()` on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check that your model regresses Pro.Strong.Leader on **FHTotalAggr_2016** using the **fHouse** data. The outcome variable goes on the left of the ~.")
  }

  pass("Correct! The model regresses approval of a strong leader on a country's overall Freedom House score.")
})
```

### Interpret the Coefficients

The Coefficients table now describes the relationship between political freedom and appetite for unconstrained leadership. Keep in mind what the Freedom House scale represents --- 0 is the least free end, 100 the most --- as you interpret what each estimate tells you.

**Apply 12: Interpret the national freedom coefficients**

Look at your Coefficients table before answering.

```{r quiz-regression-strongleader-freedom, echo=FALSE}
question("Which of the following statements correctly interpret the intercept and slope? Select all that apply.",
  correct = "Right. The intercept gives predicted approval when freedom is zero; the slope gives the change in approval for each additional point of freedom.",
  answer("The intercept gives the predicted approval of a strong leader in a country with a Freedom House score of zero",
         correct = TRUE,
         message = "The intercept estimate is 70.107 --- in a hypothetically unfree country at **FHTotalAggr_2016** = 0, about 70.1% of respondents are predicted to approve of a strong leader. No country in the data is exactly at zero, so the intercept is a model-based benchmark rather than a prediction for an observed case."),
  answer("The slope gives the expected change in approval of a strong leader for a one-point increase in a country's Freedom House score",
         correct = TRUE,
         message = "The slope estimate is -0.283 --- countries with a one-point higher freedom score are predicted to have about 0.283 percentage points lower approval of a strong leader, on average. The negative sign matches the hypothesis."),
  answer("The slope gives the expected change in a country's Freedom House score for a one-percentage-point increase in approval of a strong leader",
         message = "This reverses the outcome and predictor variables. We modeled **Pro.Strong.Leader** as a function of **FHTotalAggr_2016**. The slope tells us the expected change in approval for a one-point increase in the freedom score, not the reverse."),
  answer("The intercept tells us the average Freedom House score among countries that approve of a strong leader",
         message = "This reverses the outcome and predictor variables. The intercept is a predicted value of the outcome (approval of a strong leader) when the predictor (Freedom House score) equals zero --- it says nothing about the average value of the predictor."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there is more than one correct answer. Apply the definitions of intercept and slope to this specific model --- the intercept is the predicted outcome when X = 0, the slope is the change per one-unit increase in X."
)
```

### Calculate Predicted Values

The same logic applies here. Substituting any value of **FHTotalAggr_2016** into the fitted line gives the predicted approval level for a country with that freedom score. At the mean freedom score among countries with survey data, 59:

$$\widehat{\text{Pro.Strong.Leader}} = 70.107 + (-0.283)\times 59 = 70.107 - 16.697 = 53.410.$$

So the model predicts about 53.4% approval of a strong leader in a country near the middle of the freedom distribution.

**Practice 13: Calculate a predicted strong-leader approval**

```{r quiz-prediction-strongleader-freedom, echo=FALSE}
question("Using the intercept and slope from your output, what is the predicted approval of a strong leader in a highly free country with **FHTotalAggr_2016** = 90?",
  correct = "Right. Substituting into the fitted line: 70.107 + (-0.283)(90) = 44.62.",
  answer("About 44.6 percentage points", correct = TRUE,
         message = "$\\widehat{\\text{Pro.Strong.Leader}} = 70.107 + (-0.283)\\times 90 = 44.62$. Compare this to the predicted 63.03% at **FHTotalAggr_2016** = 25 --- that is a difference of about 18 percentage points across the range from not-free to very free."),
  answer("About 63.0 percentage points",
         message = "That would be the predicted approval at **FHTotalAggr_2016** = 25, a not-free country --- not at 90. Substitute 90 into the equation: $70.107 + (-0.283) \\times 90$."),
  answer("About 53.4 percentage points",
         message = "That is the predicted approval at **FHTotalAggr_2016** = 59, the mean among the countries with survey data. The question asks for the prediction at 90, which is above the mean. Because the slope is negative, a higher freedom score produces a lower predicted approval."),
  answer("About 70.1 percentage points",
         message = "That is the intercept --- the predicted approval when **FHTotalAggr_2016** is zero. A country with a score of 90 is near the free end of the scale, and the negative slope means its predicted approval is well below the intercept."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: substitute **FHTotalAggr_2016** = 90 into the fitted line using the intercept and slope from your output."
)
```

### Conduct Hypothesis Tests

That slope is negative, matching the hypothesis --- but with only 62 countries having data on both variables, does it provide sufficient evidence that the population slope differs from zero? Those 62 are not a random sample of the world's countries: Wave 7 of the World Values Survey was fielded where it could be, which skews toward states willing to permit a national survey. Any conclusion applies to countries like these, not to every country. The null hypothesis is $H_0: \beta_{F} = 0$ and the alternative is $H_A: \beta_{F} \ne 0$, where $\beta_F$ is the population slope on **FHTotalAggr_2016**. Find the t-statistic and p-value for the slope in your output.

**Apply 14: Conduct the hypothesis test**

Find the t-statistic and p-value for the slope in your output before answering.

```{r quiz-hypothesistesting-strongleader-freedom-2, echo=FALSE}
question("Which conclusions follow from the hypothesis test on the slope? Select all that apply.",
  correct = "Right. Both the p-value and the t-statistic lead to the same conclusion: reject the null hypothesis.",
  answer("The p-value is less than 0.05, so we reject the null hypothesis",
         correct = TRUE,
         message = "The p-value is 0.00116 --- a relationship this size would be very unlikely if the population slope were zero. Well below the 0.05 threshold, and on only 62 countries."),
  answer("The absolute value of the t-statistic exceeds 1.96, which leads to the same conclusion as the p-value",
         correct = TRUE,
         message = "The t-statistic is -3.411, with absolute value well above 1.96. The t-statistic and p-value always agree for the same two-tailed test, since both are built from the same estimate and standard error."),
  answer("The significant result proves that living in a free country causes citizens to oppose strong leaders",
         message = "A significant coefficient tells us the relationship is statistically distinguishable from zero. It does not establish causation --- freer countries may differ from less free ones in many other ways that also shape political attitudes."),
  answer("A p-value of 0.00116 means there is a 0.1% probability that the null hypothesis is true",
         message = "This reverses what the p-value measures. The p-value is the probability of observing a result this extreme if the null were true --- not the probability that the null itself is true. A p-value of 0.00116 means results at least this extreme would occur about 0.1% of the time if the population slope were zero --- not that there is a 0.1% chance the null is correct."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there is more than one correct answer. Check the p-value and the t-statistic in the **FHTotalAggr_2016** row."
)
```

### Assess Model Fit

Significance and fit address different questions. The slope test told you the relationship is statistically distinguishable from zero --- the fit statistics now tell you how much of the country-to-country variation in strong leader approval that relationship accounts for.

**Apply 15: Assess the fit of the national freedom model**

Find `Multiple R-squared` and `Residual standard error` at the foot of your output. This model runs on 62 countries rather than 3,110 counties, which is worth keeping in mind as you read them.

```{r quiz-modelfit-strongleader-freedom, echo=FALSE}
question("Which of the following statements about model fit are correct? Select all that apply.",
  correct = "Right. An R-squared of about 0.16 and a residual standard error of about 19 percentage points indicate weak fit --- the slope is statistically distinguishable from zero, but freedom alone explains little of the variation in strong leader approval.",
  answer("The Multiple R-squared shows that national freedom explains about 16% of the variation in approval of a strong leader across countries",
         correct = TRUE,
         message = "The Multiple R-squared is 0.1624. That is a small proportion --- about 84% of the country-to-country variation in approval remains unexplained by freedom alone. The relationship is statistically distinguishable from zero but freedom is far from the whole story."),
  answer("The Multiple R-squared shows that the model explains about 84% of the variation in approval of a strong leader, leaving only 16% unexplained",
         message = "This reads R-squared backwards. An R-squared of 0.1624 means the model explains about 16% of the variation, not 84%. About 84% is left unexplained."),
  answer("The residual standard error shows that in a typical country, the model's prediction of strong leader approval is off by about 19 percentage points",
         correct = TRUE,
         message = "The residual standard error is 18.60 percentage points. Given that approval rates range from roughly 9% to 93% across countries, a typical error of 19 points is large. Weak fit and a significant slope coexist here because significance answers a different question than fit."),
  answer("The residual standard error of about 19 means the model explains about 19% of the variation in approval",
         message = "The residual standard error is not a proportion --- it is measured in the same units as the outcome, here percentage points. It gives the approximate typical distance between observed outcomes and the fitted regression line, not the share of variation explained. R-squared does that, and here it is about 0.16."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there are two correct statements --- one about R-squared and one about the residual standard error."
)
```

### Reporting the Results

So is political freedom associated with citizens' appetite for unconstrained leadership? The data say yes --- but as in Example 1, the full picture is more nuanced than a significant slope alone suggests.

::: {.research-example}
Countries with higher levels of political freedom show meaningfully lower public support for an unconstrained strong leader --- a negative relationship that is statistically significant even with only 62 countries in the sample. But the fit is weak: R-squared is 0.16, meaning freedom accounts for only about 16% of the cross-national variation in approval, and the residual standard error of about 19 percentage points means the estimated standard deviation of the residuals is large --- country-level fitted values are often far from the observed values. That combination --- a significant slope alongside weak fit --- is not a contradiction. The hypothesis test asks whether the relationship between freedom and approval is reliably non-zero in the population, and the evidence says yes. R-squared asks how much of the variation in approval freedom alone can account for, and the honest answer is: not much. That is not surprising once you consider everything else associated with whether someone wants an unconstrained leader --- economic security, generational experience, media environment, recent political history --- nearly all of which a single national freedom score cannot capture. What the model tells us is that political freedom is associated with lower appetite for unconstrained leadership; it does not tell us it is all that is relevant.
:::

## Do Majority-White Counties Differ in Their Support for Hillary Clinton?

Tutorial 13 tests whether Democratic vote share differs by racial-majority category using a two-sample t-test, comparing Black Majority and White Majority counties specifically. That test provided statistically significant evidence of a difference between those two groups. Here the question is a related but different one: is having a white majority at all --- collapsing the four racial majority categories into a simple binary --- associated with Clinton's vote? Regression answers that question and gives the exact magnitude of the gap between the two groups. Both examples so far have used an interval-level predictor; what's new here is how to create a properly labeled categorical variable in R and how to read the slope that results. The mechanics of `lm()` are identical to what you've already seen.

::: {.hypothesis}
In a comparison of counties, those with a majority-white population will cast a lower percentage of the two-party vote for Hillary Clinton than those counties without a majority-white population.
:::

The variable **majority_white** is coded 1 for the counties **racial_majority** calls White Majority and 0 for the other three categories. Both variables are recorded for every one of the 3,112 counties, and the split is lopsided: 2,866 counties are majority white and 246 are not. With only two values, running `lm()` on the raw numeric variable produces identical results to running it on a factor --- the intercept is still the mean of the 0 group and the slope is still the difference between the two group means. The only difference is cosmetic: the output labels the slope `majority_white`, the name of the numeric variable, rather than the more readable `majority_white_fMajority White` that the factor version produces.

This equivalence breaks down the moment a categorical variable has more than two values. If **majority_white** took values 0, 1, and 2 for three racial composition categories, passing the raw numeric variable to `lm()` would treat it as a continuous predictor with a single slope --- implying that the difference between categories 0 and 1 equals the difference between categories 1 and 2, which is almost never a reasonable assumption for a nominal variable. A factor tells R to estimate a separate coefficient for each non-reference category instead.

There are two reasons to use `factor()` here: readable output labels, and to establish the practice before you encounter variables with three or more categories in Tutorial 16, where it becomes essential.

### Create a Labeled Factor

The `factor()` function takes the name of the variable to recode, plus a `labels` argument: a vector of names given in order from the smallest to the largest original value. The `levels()` function then lets you check your work --- it returns a factor's category names, in the exact order R will use them, with the first one treated as the [reference category]{.important-text}: the category R omits from the regression output, against which the other category is compared.

**Example code:**

```{r factor-template, exercise=FALSE, eval=FALSE}
yourdata$new_factor <- factor(yourdata$numeric_variable,
                              labels = c("Label for the lowest code",
                                         "Label for the next code"))
levels(yourdata$new_factor)
```

**Apply 16: Create a labeled factor variable and check its levels**

In the **counties** data frame, create **majority_white_f** from **majority_white**, labeling the two categories "Not Majority White" and "Majority White", then display its levels to confirm which one R will treat as the reference. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-factor-white, exercise=TRUE, exercise.cap="Apply 16"}

```

```{r exercise-factor-white-hint-1}
# Hint 1 of 3
# A factor stores a categorical variable by its category names
# rather than by the numbers standing in for them, and the order
# of those names decides which category R treats as the baseline.
# You need to attach names to the two values of majority_white,
# confirm which name comes first.
```

```{r exercise-factor-white-hint-2}
# Hint 2 of 3
# factor() does the recoding and takes a labels argument.
# levels() reports the result, and takes only the factor itself.
```

```{r exercise-factor-white-hint-3}
# Hint 3 of 3
# The labels go in order from the smallest original value to the
# largest --- so the name for 0 comes first, then the name for 1.
# Assign the result back into the data frame, so the regression
# below can use it.
```

```{r exercise-factor-white-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("factor\\s*\\(", code_no_comments)) {
    fail("Use `factor()` to convert **majority_white** into a labeled factor.")
  }
  if (!grepl("majority_white_f\\s*(<-|=)", code_no_comments)) {
    fail("Assign the new factor to **majority_white_f**, so the regression below can use it.")
  }
  if (!grepl("labels\\s*=", code_no_comments)) {
    fail("Pass a labels argument to `factor()` so the categories print with names rather than 0 and 1.")
  }
  if (!grepl("Not Majority White", code_no_comments) ||
      !grepl("Majority White", code_no_comments)) {
    fail("Use the two labels exactly as written: \"Not Majority White\" and \"Majority White\".")
  }
  if (!grepl("levels\\s*\\(", code_no_comments)) {
    fail("Call `levels()` on your new factor to display the category names in the order R will use them.")
  }

  if (!grepl("counties\\$majority_white_f\\s*(<-|=)\\s*factor\\s*\\(\\s*(x\\s*=\\s*)?counties\\$majority_white\\b",
             code_no_comments)) {
    fail("Build the factor and assign it in one step: **counties$majority_white_f** on the left, `factor()` applied to **counties$majority_white** on the right.")
  }

  if (is.null(.result)) {
    fail("Your code did not produce a result. Make sure the last line calls `levels()` on your factor.")
  }
  if (!identical(as.character(.result),
                 c("Not Majority White", "Majority White"))) {
    fail("The levels are not in the expected order. The labels are assigned from the smallest original value upward, so the label for 0 must come first.")
  }
  built <- factor(counties$majority_white,
                  labels = c("Not Majority White", "Majority White"))
  if (!identical(as.character(counties$majority_white_f), as.character(built))) {
    fail("The factor does not match **majority_white** case for case. Build it from that variable rather than constructing the categories another way.")
  }

  pass("Correct --- **majority_white_f** is built from **majority_white**, with both categories named.")
})
```

Running `levels(counties$majority_white_f)` confirms "Not Majority White" is listed first, so it is the reference category --- exactly what the label order was chosen to produce. Because **majority_white_f** is already a factor, you can use it directly in a regression formula --- no need to wrap it in `factor()` the way you would with the raw, unlabeled variable.

### Estimate the Regression

With **majority_white_f** created and its levels confirmed, estimate the regression.

**Practice 17: Estimate a regression of Clinton's two-party vote percent on majority-white status**

Write the code to estimate the model, assigning it to **model3**, then display the results. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-regression-voteclinton-majoritywhite, exercise=TRUE, exercise.cap="Practice 17"}

```

```{r exercise-regression-voteclinton-majoritywhite-hint-1}
# Hint 1 of 2
# Look back at the hypothesis. Which variable is the outcome
# we are trying to explain, and which is the predictor?
# In a regression formula, the outcome goes on the left of the ~
# and the predictor goes on the right.
```

```{r exercise-regression-voteclinton-majoritywhite-hint-2}
# Hint 2 of 2
# lm() estimates the model and takes the formula first, then data
# to name the frame holding the variables. Save the model to an object,
# then call summary() on it so the summary is what prints. Use the
# labeled factor rather than the raw numeric variable.
```

```{r exercise-regression-voteclinton-majoritywhite-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (is.null(.result)) {
    fail("Your code did not produce a result. Check for a missing tilde (~) in the formula, a misspelled variable name, or unmatched parentheses.")
  }
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use `lm()` to estimate the regression.")
  }
  if (!grepl("model3\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model3.")
  }
  if (!grepl("dem2p_percent\\s*~\\s*majority_white_f", code_no_comments)) {
    fail("The formula should be `dem2p_percent ~ majority_white_f`. The outcome variable (**dem2p_percent**) goes on the left of the tilde and the predictor (**majority_white_f**) goes on the right.")
  }
  if (!grepl("data\\s*=\\s*counties", code_no_comments)) {
    fail("Set data = counties so R knows where to find the variables.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call `summary()` on your saved model object to display the results.")
  }

  expected <- lm(dem2p_percent ~ majority_white_f, data = counties)
  if (!inherits(.result, "summary.lm")) {
    fail("The last line should display the summary. Save the model with `lm()`, then call `summary()` on it so the summary is what prints.")
  }
  result_coef <- setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling `summary()` on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check that your model regresses **dem2p_percent** on **majority_white_f** using the counties data. Make sure **majority_white_f** is on the right-hand side of the ~.")
  }

  pass("Correct! Because **majority_white_f** is already a factor, R automatically created an indicator for the Majority White category and used Not Majority White as the reference.")
})
```

### Interpret the Coefficients

Notice the slope in the output is now labeled something like `majority_white_fMajority White` --- R names factor coefficients after the variable and the non-reference category, rather than printing a bare 0 or 1. The intercept and slope still have precise meanings, but those meanings are specific to the factor structure --- what the reference category is and how the two groups compare.

**Apply 18: Interpret the majority-white coefficients**

Look at your Coefficients table before answering.

```{r quiz-regression-voteclinton-majoritywhite, echo=FALSE}
question("Which of the following statements correctly interpret the regression output? Select all that apply.",
  correct = "Right. The intercept gives the non-majority-white mean; the slope gives the gap between the two groups; adding them gives the majority-white mean.",
  answer("The slope gives the difference in the average percent of the two-party vote for Clinton between majority-white and non-majority-white counties",
         correct = TRUE,
         message = "The slope estimate is -25.357 --- majority-white counties cast about 25 percentage points fewer votes for Hillary Clinton, the Democratic candidate, than non-majority-white counties on average. The negative sign means majority-white counties gave her less of the vote."),
  answer("The intercept gives the average percent of the two-party vote for Clinton in non-majority-white counties",
         correct = TRUE,
         message = "The intercept estimate is 56.694 --- non-majority-white counties gave Hillary Clinton about 56.7% of the two-party vote on average. For majority-white counties, add the slope to the intercept to get their predicted mean."),
  answer("The slope gives the expected decrease in Clinton's percent of the two-party vote for each one-percentage-point increase in the proportion of white residents",
         message = "**majority_white_f** is a binary variable, not a continuous percentage. Its slope gives the difference between the two groups --- majority white versus not majority white --- not the effect of a one-unit change in a continuous variable."),
  answer("The slope tells us that majority-white counties are about 25 percentage points more likely to be majority white than non-majority-white counties",
         message = "This confuses the outcome and predictor variables. The slope describes the difference in the outcome --- Clinton's percent of the two-party vote --- between the two groups defined by the predictor (majority-white status), not the reverse."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there is more than one correct answer. The intercept is the reference category mean; the slope is the difference between groups."
)
```


### Conduct Hypothesis Tests

The slope for **majority_white_fMajority White** is the coefficient of interest here --- does that gap hold up to formal scrutiny? The null hypothesis is $H_0: \beta_{W} = 0$ and the alternative is $H_A: \beta_{W} \ne 0$, where $\beta_{W}$ is the population slope on **majority_white_f**. Find the t-statistic and p-value for the slope in your output.

**Apply 19: Test the slope on majority_white_f**

```{r quiz-hypothesistesting-voteclinton-majoritywhite, echo=FALSE}
question("Which conclusions follow from the hypothesis test? Select all that apply.",
  correct = "Right. Both the p-value and the t-statistic lead to the same conclusion: reject the null hypothesis.",
  answer("The p-value is less than 0.05, so we reject the null of no relationship",
         correct = TRUE,
         message = "The p-value is less than 2e-16, far below 0.05. A difference this large between majority-white and non-majority-white counties would be extremely unlikely if there were truly no relationship in the population."),
  answer("The absolute value of the t-statistic exceeds 1.96, so we reject the null hypothesis",
         correct = TRUE,
         message = "The t-statistic is -26.07, with absolute value far above 1.96. For this same two-tailed test, the t-statistic and p-value always agree."),
  answer("There is less than a five percent probability the null hypothesis is true",
         message = "This is a common misreading of the p-value. The p-value gives the probability of observing a coefficient this far from zero *if the null hypothesis were true* --- it does not give the probability that the null hypothesis itself is true."),
  answer("The significant result proves that racial composition causes counties to vote differently",
         message = "A significant coefficient tells us the relationship is statistically distinguishable from zero. It does not establish causation --- majority-white counties differ from non-majority-white counties in many other ways that could also shape the vote."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there is more than one correct answer. Check the p-value and the t-statistic in the majority_white_fMajority White row."
)
```

### Assess Model Fit

The slope told you majority-white counties voted differently for Clinton --- but how much of the county-to-county variation in her vote does that single distinction capture? The fit statistics address that question.

**Apply 20: Assess the fit of the majority-white model**

Find `Multiple R-squared` and `Residual standard error` at the foot of your output, then compare what you find with the two earlier examples.

```{r quiz-diagnostics-voteclinton-majoritywhite, echo=FALSE}
question("Which of the following statements about model fit are correct? Select all that apply.",
  correct = "Right. An R-squared of about 0.18 and a residual standard error of about 15 percentage points indicate weak fit --- majority-white status captures some of the variation in Clinton's percent of the two-party vote but leaves most unexplained.",
  answer("The Multiple R-squared shows that majority-white status explains about 18% of the variation in Clinton's percent of the two-party vote across counties",
         correct = TRUE,
         message = "The Multiple R-squared is 0.18 --- a modest proportion. Knowing whether a county is majority white moves us in the right direction, but 82% of the variation in Clinton's percent of the two-party vote is driven by other factors."),
  answer("The Multiple R-squared shows that the model explains about 82% of the variation in Clinton's percent of the two-party vote, leaving only 18% unexplained",
         message = "This reads R-squared backwards. An R-squared of 0.18 means the model explains 18% of the variation, not 82%. About 82% remains unexplained."),
  answer("The residual standard error shows that in a typical county, the model's prediction of Clinton's percent of the two-party vote is off by roughly 14 to 15 percentage points",
         correct = TRUE,
         message = "The residual standard error is about 14.64 percentage points. That is a large typical error, consistent with an R-squared of only 0.18. The binary majority-white split captures a large, statistically distinguishable difference between groups but still leaves most county-to-county variation unaccounted for."),
  answer("The residual standard error of about 15 means the model explains about 15% of the variation in Clinton's percent of the two-party vote",
         message = "The residual standard error is not a proportion --- it is in the same units as the outcome, percentage points. It gives the approximate typical distance between observed outcomes and the fitted regression line. R-squared gives the proportion of variance explained, and here that is 0.18."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: there are two correct statements --- one about R-squared and one about the residual standard error."
)
```

### Reporting the Results

So is majority-white status associated with Clinton's vote? Yes, and by a wide margin --- but even a 25-percentage-point gap between groups leaves most of the county-to-county variation unexplained.

::: {.research-example}
Majority-white counties cast about 25 percentage points fewer votes for Clinton than non-majority-white counties --- a large, highly significant difference. But the fit is weak: the R-squared of 0.18 means majority-white status accounts for only about 18% of the county-to-county variation in Clinton's vote, and the residual standard error of about 15 percentage points means individual county predictions are often substantially off. A statistically significant relationship can coexist with poor to weak fit, because most political outcomes have many causes and a single predictor rarely accounts for much of the variation. Here one predictor explains about 18% of the variation in the outcome --- enough to provide strong evidence that the population relationship is not zero, but far from the whole story.
:::

## Check Your Understanding

The three questions below ask you to apply what you have learned to new scenarios. Work through them without looking back at the examples.

**Apply 21: Predict a value from a hypothetical regression**

Substitute X = 5 into the fitted line $\hat{Y} = \hat{\alpha} + \hat{\beta} \times X$ before answering.

```{r quiz-cyu-predicted-value, echo=FALSE}
question("A regression of Y on X produces an intercept of 10 and a slope of -2. What is the predicted value of Y when X = 5?",
  correct = "Right. Substituting into the fitted line: 10 + (-2)(5) = 0.",
  answer("0", correct = TRUE,
         message = "Substituting into the fitted line: $\\hat{Y} = 10 + (-2)(5) = 10 - 10 = 0$."),
  answer("8",
         message = "Check the substitution: the slope is -2 and X = 5, so the calculation is $10 + (-2)(5)$, not $10 + (-2)(1)$. The slope gets multiplied by the value of X."),
  answer("-10",
         message = "Check the order of operations: multiply first, then add. 10 + (-2)(5) = 10 + (-10) = 0, not 10 - 2 - 5."),
  answer("20",
         message = "Check the sign of the slope: it is -2, not +2. $10 + (-2)(5) = 0$, not $10 + 2(5) = 20$."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: substitute X = 5 into $\\hat{Y} = 10 + (-2) \\times X$."
)
```

**Apply 22: Significance and importance are different questions**

Think about what the p-value and $R^2$ are each measuring before answering.

```{r quiz-cyu-significance-importance, echo=FALSE}
question("Across the examples above, the slopes were statistically significant, yet $R^2$ never rose above about 0.18. What does this combination tell us?",
  correct = "Right. Significance and explanatory power answer different questions.",
  answer("A relationship can be statistically significant while still leaving most of the variation in the outcome unexplained", correct = TRUE,
         message = "Significance tells us whether the relationship is statistically distinguishable from zero. $R^2$ tells us how much of the outcome's variation the model accounts for. A relationship can be precisely estimated while still accounting for only a small share of the variation in the outcome."),
  answer("A statistically significant slope means the variable explains most of the variation in the outcome",
         message = "Significance and the amount of variation explained are different questions. A coefficient can be estimated very precisely --- and therefore be highly significant --- while the variable it belongs to still leaves most of the outcome's variation unexplained."),
  answer("The results must be wrong, since a real relationship should explain most of the variation in the outcome",
         message = "Relationships in political science often account for only a modest share of the variation in an outcome. A small $R^2$ alongside a statistically significant slope is a common pattern, not a sign that the results are wrong."),
  answer("A low $R^2$ means the slopes were not actually statistically significant --- the significance stars must be misleading",
         message = "Statistical significance and $R^2$ are measuring different things. A slope can be highly significant --- meaning we are confident the relationship is not zero --- while $R^2$ remains low, meaning the variable explains little of the total variation. These are not contradictory."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: significance asks whether the relationship is statistically distinguishable from zero; R-squared asks how much of the variation in the outcome the model explains. These are different questions."
)
```

**Apply 23: Interpret a slope correctly**

```{r quiz-cyu-pct-vs-pctpoint, echo=FALSE}
question("A regression of county Democratic vote share on **wage_growth** produces a slope of 0.456. A student reports: 'For each 1% increase in wage growth, Clinton's vote share increases by 0.456 percentage points.' What is wrong with this interpretation?",
  correct = "Right. **wage_growth** is measured in percentage points, so a one-unit change is a one-percentage-point change, not a one-percent relative change.",
  answer("The slope describes a one-percentage-point increase in wage growth, not a one-percent relative increase",
         correct = TRUE,
         message = "Because **wage_growth** is measured in percentage points, a one-unit change is a one-percentage-point change. A 1% relative increase would be a much smaller change --- from 4% to 4.04%, for example. The student's phrasing implies the latter, which misrepresents what the slope actually measures."),
  answer("The slope should be negative, since higher wage growth in Republican-leaning counties would reduce Clinton's vote share",
         message = "The sign of the slope is an empirical finding, not something we can determine from the units alone. The positive slope means that, on average across all counties, higher wage growth is associated with a higher Clinton vote share."),
  answer("The interpretation is correct --- a 1% increase and a one-percentage-point increase mean the same thing when the variable is measured as a percent",
         message = "They are not the same. A one-percentage-point increase is an additive change --- from 4% to 5%. A 1% relative increase is a multiplicative change --- from 4% to 4.04%. The slope describes the former, not the latter."),
  answer("The direction of the relationship is stated correctly but the magnitude of 0.456 is too small to be meaningful",
         message = "The size of the slope does not determine whether the interpretation is correct. The error is in the units: describing a one-percentage-point change as a '1% increase' conflates two different quantities."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: how is **wage_growth** measured? Is a one-unit change the same as a 1% relative increase?"
)
```

## The Takeaways

[Simple linear regression]{.important-text} unifies two frameworks you already know. When the predictor is interval, regression extends what correlation gave you --- a description of direction and strength --- by adding the quantity correlation cannot: how much the outcome changes for a one-unit increase in the predictor. When the predictor is categorical, regression produces the same result as a difference-of-means test, but inside a framework that generalizes to any level of measurement and any number of predictors. The three examples in this tutorial covered an interval predictor with a positive slope, an interval predictor with a negative slope, and a [binary]{.important-text} predictor --- three different substantive questions, one framework.

**Estimating a model.** Use `lm()` to estimate a regression model. The first argument is a formula: the outcome variable, a tilde (`~`), and the explanatory variable --- called a [predictor]{.important-text} throughout the regression tutorials. The second argument is `data`, which names the data frame containing the variables --- this means you write bare variable names in the formula, not `dataframe$variable`. Assign the result to a named object with `<-`, then pass that object to `summary()` to display the formatted results. When the predictor is categorical, use `factor()` to create a properly labeled factor variable before estimating the model. The `factor()` function takes the variable name and a `labels` argument giving category names in order from smallest to largest original value. Use `levels()` afterward to confirm which category R will treat as the reference --- the one it omits from the output, against which all others are compared.

**Interpreting the coefficients.** The intercept and slope have precise meanings that depend on what kind of predictor you have:

- *Interval predictor:* The [intercept]{.important-text} ($\hat{\alpha}$) is the predicted value of $Y$ when $X = 0$. The [slope]{.important-text} ($\hat{\beta}$) is the estimated [effect]{.important-text} of $X$ on $Y$: the expected change in $Y$ for a one-unit increase in $X$.
- *Binary predictor:* The intercept is the mean of the [reference category]{.important-text}. The slope is the difference between the two group means --- the mean of the non-reference category minus the mean of the reference category.

In either case, a predicted value of $Y$ for any specific value of $X$ is obtained by substituting that value into the fitted line: $\hat{Y} = \hat{\alpha} + \hat{\beta} X$.

**Testing the coefficients.** The `summary()` output provides a t-statistic and p-value for each coefficient. These two pieces of evidence always agree for the same two-tailed test: if the p-value is below 0.05, reject the null hypothesis that the coefficient is zero in the population; if it is above 0.05, fail to reject. The absolute value of the t-statistic provides the same information --- for the large samples in this tutorial, a t-statistic above approximately 1.96 corresponds to a p-value below 0.05 --- but in practice, use the p-value R reports directly.

The alternative in these tutorials is always two-tailed ($H_A: \beta \ne 0$), which is standard practice for regression coefficients. If you reject the two-tailed null and the sign of the coefficient matches your theoretical prediction, you have implicitly rejected the one-tailed null as well and can make a directional claim in your substantive interpretation.

**Residuals.** The [residual]{.important-text} for a case is the difference between what was observed and what the line predicted --- observed minus predicted. A positive residual means the model [underpredicted]{.important-text} that case; a negative one means it [overpredicted]{.important-text}. Residuals are worth looking at case by case, not just in aggregate: a model can fit acceptably overall and still miss a particular county or country by a wide margin.

**Assessing model fit.** Two statistics in the `summary()` output describe how well the model fits the data. [R-squared]{.important-text} ($R^2$) gives the proportion of the variation in $Y$ explained by the model --- bigger is better, and a value of 0.20 means the model accounts for 20% of the outcome's variation. The [residual standard error]{.important-text} gives the standard deviation of the residuals, which means that when the residuals are approximately normally distributed, roughly two-thirds of the observed values fall within that distance of the regression line --- smaller is better, as it means the data cluster more tightly around the regression line. All three examples in this tutorial produced significant slopes alongside small $R^2$ values, which is not a contradiction. Significance tells you the relationship is statistically distinguishable from zero; $R^2$ tells you how much of the variation it accounts for. In political science, a single predictor rarely tells the whole story.

**Percentage points and percents.** They sound like the same thing, and they are not. When a predictor is measured in percentage points --- as **wage_growth** is here --- a one-unit change is a one-percentage-point change, full stop. Wage growth moving from 4% to 5% is a one-percentage-point increase, but it is a 25% relative increase, which is a very different claim about the world. The slope tells you how the outcome changes for the former, not the latter. This sounds pedantic until you make the mistake once.

Tutorial 16 takes the next step. Every model here left most of the variation in the outcome unexplained, and other variables contribute to it. When those variables are also related to the predictor you included, the simple-regression slope may partly reflect those relationships rather than the predictor alone. Multiple regression allows you to examine whether the original relationship remains after accounting for measured competing explanations.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial.

```{r encode-logic-15, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-15, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-15, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now estimate and interpret simple regression models, including models with a factor variable as the predictor.">
  <span aria-hidden="true">📉</span>
  Excellent work! You can now estimate and interpret simple regression models, including models with a factor variable as the predictor.
</p>
')
```
