---
title: "R Basics: Part 1"
tutorial:
  id: "01-R-Part1"
output:
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn the basics of how R works.
---

## Overview {#main-content}

This is the first in a series of tutorials you will work through over the course of the semester. The tutorials are designed to teach you the ins and outs of data analysis in R. This tutorial introduces some basic principles of using R and provides background knowledge that will underlie the code you write throughout the semester.

Each tutorial typically begins with examples you will run to see how code works, followed by exercises in which you complete or modify code, and finally exercises in which you produce code from scratch. A **Hint** button is available on graded exercises if you get stuck.

::: {.tip}
**Run Code vs. Submit Answer:** These two buttons do very different things. **Run Code** simply executes whatever is in the box and shows you the result --- you can use it as many times as you like to experiment without any consequences. Only **Submit Answer** checks your work. Get in the habit of using **Run Code** freely to try things out before you submit. If something does not work the way you expected, that is useful information --- try to understand why before moving on.
:::

You should work through the tutorials from beginning to end, completing each exercise. If you run out of time, or just need a break, you can exit and return later. Your progress will be saved when you return.

You will make mistakes as you work through these tutorials. R, like any coding language, requires careful attention to matching parentheses, including commas in the right places, and spelling functions, arguments, and object names correctly. When you encounter errors --- and you will do so often --- look for these types of mistakes first. If this feels picky at first, that is because R is picky; you are learning the rules of the language, and that takes practice.

::: {.tip}
**Take notes as you work.** A good strategy is to keep a running list of the functions introduced in each tutorial, the name of the tutorial that introduced them, and a short description of what they do. This will give you a reference sheet to consult when you need to write code from scratch in class, or for problem sets.
:::

In this tutorial you will learn:

* How to use R as a calculator
* How to assign a value or values to an object with `<-`
* How to use functions
* How to determine the class of an object with the `class()` function
* How to create and use vectors using the `c()` function
* How to select and replace elements of a vector using indexing
* How to install packages with `install.packages()` and load them with `library()`

::: {.important-note}
**Function restriction:** For the graded exercises in this tutorial, stay within the tools introduced here. Use only the functions, arguments, and approaches covered in this tutorial, and do not use functions from packages that are not loaded here. This helps ensure the grader is checking the same skills the tutorial is teaching. If you are unsure whether a function or approach is permitted, check whether it appears in this tutorial before using it.
:::

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(gradethis)
library(learnrhash)
options(lifecycle_verbosity = "quiet")   # suppress tidyselect deprecation warnings from gradethis
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(error = TRUE)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513, 0.493)
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry","Obama","Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
num.list<-c(1,2,3,4,5)
my.first.object <- 5 + 3
another.object <- "100"
```

## R as a calculator

It is easy to use R like a calculator. Here's an example of adding 2 numbers.
You can add/subtract, multiply/divide, or raise a number to a power using `+`,
`-`, `*`, `/`, `^`.

**Explore: Basic arithmetic in R**

Try running the code below, then change the numbers and operators to practice
different calculations. Use the "Run Code" button to execute your code. Note
that R ignores spaces, so 14 + 2000 will produce the same output as 14+2000.

```{r calculator-practice, exercise=TRUE, exercise.cap="Explore: Basic arithmetic in R"}
14 + 2000
```

The `[1]` you see with the line of output is just R telling you that you have just
one number produced by the code. Later we will see cases where a command we run
produces an output that contains multiple numbers (or words). Each line of
output will begin with some number in square brackets [ ], with the number
denoting the index value of the output that starts that line.

## Objects

**R can store information as an [object]{.important-text} with a name of your
choice. Once you have created an R object, you can refer to it by name.**

Picking a name for your object:

* Choose an informative name
* Object names cannot start with a number
* Object names cannot contain spaces
* Object names cannot include special characters that have specific meanings in
R. This includes $ and # and %.
* Periods and underscores are permitted and commonly used to separate words ---
for example, `my.first.object` or `my_first_object` are both valid names.

In a regular RStudio session, the Global Environment pane will show a list of
all the objects you have created. This pane is not visible inside these
tutorials, but you will use it constantly when working on your own.

**We use the [assignment operator]{.important-text} `<-` to assign some value to
an object.**

**Run and observe: Create your first object**

Let's store the sum of 5+3 by assigning it an object called **my.first.object**. Use the
"Run Code" button to execute the following code and see how assignment works.

```{r objects-create-first, exercise=TRUE, exercise.cap="Run and observe: Creating and assigning an object"}
my.first.object <- 5 + 3
```

Notice that the result is not displayed in this case. We can see the value of an
object only by asking R to show it to us. Simply enter the object name.


**Run and observe: Display an object's value**

Use the "Run Code" button to see how to display what's stored in an object:

```{r objects-print-value, exercise=TRUE, exercise.cap="Run and observe: Displaying an object's value"}
my.first.object
```

If we assign a different value to the same object name, the value of the object
will change.

**Run and observe: Reassign an object**

Use the "Run Code" button to see what happens when you assign a new value to an
existing object:

```{r objects-reassign, exercise=TRUE, exercise.cap="Run and observe: Reassigning an object to a new value"}
my.first.object <- 50 + 30
my.first.object
```

The second line displays `80` because the new assignment replaces the old value
stored in `my.first.object`. R does not keep both values under the same object
name; the most recent assignment is the one that matters.

Object names are case sensitive: **my.first.object** is different from
**MY.FIRST.OBJECT**. Try asking R to tell you the value of **MY.FIRST.OBJECT**.
What happens?

**Run and observe: Test case sensitivity**

Use the "Run Code" button to see what error occurs with incorrect
capitalization:

```{r objects-case-error, exercise=TRUE, exercise.cap="Run and observe: R's case sensitivity produces an error"}
MY.FIRST.OBJECT
```

The error says that the object **MY.FIRST.OBJECT** was not found. R is telling
you exactly what went wrong: it cannot find an object with that exact spelling
and capitalization. In a regular RStudio session you could confirm this by
looking at the Global Environment pane, where you would see that
**MY.FIRST.OBJECT** does not exist.

### Storing Character Strings as Objects

To assign a string of characters to an object, we use quotation marks. Spaces are allowed inside the quotation marks.

**Run and observe: Create a character object**

Use the **Run Code** button to see how to store text as an object using quotation marks:

```{r objects-create-string, exercise=TRUE, exercise.cap="Run and observe: Storing text as a character object"}
this.class <- "PLSC 309"
this.class
```

R treats numbers like characters if we tell it to do so by enclosing them in
quotes:

**Run and observe: Numbers as characters**

Observe what happens when you put numbers in quotation marks. Use the "Run Code"
button to execute the code:

```{r objects-number-as-string, exercise=TRUE, exercise.cap="Run and observe: Numbers stored as character strings"}
another.object <- "100"
another.object
```

If you refer to an object that is a character string in an arithmetic operation,
you will see an error message. See what happens if you try to divide "100" by
10:

**Run and observe: Arithmetic error with character strings**

Use the "Run Code" button to see the error when trying arithmetic with character
strings:

```{r objects-string-arithmetic-error, exercise=TRUE, exercise.cap="Run and observe: Arithmetic error with a character object"}
another.object / 10
```

This error message is not particularly intuitive. It means R cannot apply
arithmetic to a character object. Even though `"100"` looks like a number to us,
the quotation marks tell R to treat it as text.

---

## Functions

Functions are very important in R. They can be used to perform a variety of
tasks. **A [function]{.important-text} takes one or more inputs and returns some
output.** For example, the `sqrt()` function takes as its input some number or
numeric object for which we want to know the square root. The output is the
square root of that number or object.

An **[argument]{.important-text}** is the information you put inside the
parentheses of a function so R knows what to do. In R, the syntax is generally
`function_name(input)`, where `function_name` is the name of the function and
`input` is the argument you are giving the function. Some functions take no
arguments, some take one argument, and others take several arguments. When there
is more than one argument, they are separated by a comma. Omitting a required
comma or having unbalanced parentheses are common sources of error.

The function `sqrt()` takes one argument, named `x`, which is the number or
object for which we want the square root.

**Practice: Calculate a square root**

Enter code that calculates the square root of 4 using the `sqrt()` function.
Use **Run Code** to test your code before clicking **Submit Answer** to check
your work. Two hints are available if you get stuck.

```{r functions-square-root, exercise=TRUE, exercise.cap="Practice: Calculate the square root of 4 using sqrt()"}

```

```{r functions-square-root-hint-1}
# Hint 1 of 2
# Functions follow the pattern: function_name(input)
# The function you need has a name that tells you exactly what it does.
# What function was introduced in the text just above this exercise?
```

```{r functions-square-root-hint-2}
# Hint 2 of 2
# Fill in the blank:
___(4)
```

```{r functions-square-root-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("sqrt\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the sqrt() function.")
  }
  if (!grepl("sqrt\\s*\\(\\s*4\\s*\\)", code_no_comments)) {
    fail("Use sqrt() to calculate the square root of 4.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be one number. Check that you have only one number inside the parentheses.")
  }
  if (!isTRUE(all.equal(.result, 2))) {
    fail("That is not the square root of 4. Check the number inside the parentheses.")
  }
  pass("Correct! The square root of 4 is 2.")
})
```


To see what arguments a function accepts, use the `args()` function. Type the name of the function you want to learn about inside its parentheses, without adding parentheses after that name. Here we use `args()` to examine the `round()` function, which rounds a number to a
specified number of decimal places.

**Run and observe: Explore function arguments**

Use the **Run Code** button to see what arguments the `round()` function accepts.

```{r functions-explore-args, exercise=TRUE, exercise.cap="Run and observe: Examining function arguments with args()"}
args(round)
```

The output shows four things worth knowing:

- `x` is the required argument --- the number or object you want to round.
- `digits = 0` is an optional argument with a *default value* of 0, meaning
that if you do not specify it, R will round to zero decimal places. You only
need to include it if you want a different number of decimal places.
- `...` means the function can accept additional arguments that get passed along
to other functions internally. As a beginner you will see this often and can
safely ignore it.
- The `NULL` at the bottom can also be ignored --- it is just an artifact of how
`args()` displays its output and does not mean anything is wrong.

We will return to default values and how to use them when we need them later
in this tutorial.

---

Anytime you want to learn more about a function --- what it does, what its
arguments mean, and what it returns --- type `?` followed immediately by the
function name with no space and no parentheses. Get in the habit of checking the
help page for new functions you encounter.

**Run and observe: Access function help**

Run the code below to pull up help for the `round()` function.

```{r functions-get-help, exercise=TRUE, exercise.cap="Run and observe: Accessing help documentation with ?"}
?round
```

::: {.tip}
**Help looks different here than in RStudio.** Inside this tutorial `?` produces
plain text output that can be hard to read. When you use `?` on your own machine
in RStudio, it opens a cleanly formatted Help pane. The most useful part is the
**Arguments** section, which describes what each argument does. Start with
**Usage**, **Arguments**, and **Examples**; you can skim the rest for now.
:::

If you do not know the name of a function, search for it online --- for example,
"R function to calculate standard deviation" or "R round number function."

If you receive an error such as "No documentation for 'funcname'," the function
may belong to a package that is not loaded. In that case you can try `??funcname`
to search more broadly, or install and load the relevant package first.


## Classes of objects

Every object in R has a *[class]{.important-text}* --- a label that tells R what
kind of value it contains and what operations it can perform on it. Numbers and
text, for example, are treated very differently.

You can check the class of any object using the `class()` function: type the object's name inside the parentheses. In R, giving a function its input this way is called *passing* an argument to the function --- you will see that term throughout this course. For `my.first.object` and `another.object`, this is straightforward. For `sqrt`, there is one trap to watch out for: because you are used to calling `sqrt` by writing `sqrt()`, it is tempting to write `class(sqrt())` --- but that tells R to run `sqrt` first and then check the class of its output, which is not what you want. Write `class(sqrt)` instead, with no parentheses after `sqrt`, to ask about the function itself.

**Practice: Determine object classes**

Use the `class()` function to find the class of each of the following. Write one line of code for each --- you can enter them in any order.

* `my.first.object`
* `another.object`
* `sqrt`

Click "Submit Answer" to check your work. Two hints are available.

```{r prepare-classes, echo=FALSE}
my.first.object <- 50 + 30
another.object <- "100"
```

```{r classes-determine-types, exercise=TRUE, exercise.setup="prepare-classes", exercise.cap="Practice: Determine the class of three R objects"}

```

```{r classes-determine-types-hint-1}
# Hint 1 of 2
# The class() function works the same way for all three.
# For sqrt, check the function itself --- no parentheses after sqrt.
```

```{r classes-determine-types-hint-2}
# Hint 2 of 2: fill in the blanks
class(___)
class(___)
class(___)
```

```{r classes-determine-types-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  has_first  <- grepl("class\\s*\\(\\s*my\\.first\\.object\\s*\\)", code_no_comments)
  has_second <- grepl("class\\s*\\(\\s*another\\.object\\s*\\)", code_no_comments)
  has_third  <- grepl("class\\s*\\(\\s*sqrt\\s*\\)", code_no_comments)

  if (!has_first) {
    fail("Make sure you are checking the class of my.first.object using class().")
  }
  if (!has_second) {
    fail("Make sure you are checking the class of another.object using class().")
  }
  if (!has_third) {
    fail("Make sure you are checking the class of sqrt --- no parentheses after sqrt.")
  }
  pass("Exactly right! my.first.object is numeric, another.object is character, and sqrt is a function. These three classes --- numeric, character, and function --- are the ones you will encounter most often in R.")
})
```

There are many other classes of objects in R. We will encounter more in future lessons.

## Vectors

Next, we will enter some data into R as a *numeric vector object*. A
**[vector]{.important-text}** is a way to store multiple values together in a
single object, in a specific order. A variable in a data set, for example, is a vector --  a list of values the variable takes for every unit in the data set. In this section, we will work with numeric
vectors, which contain numbers. 

To create a vector in R, we use the `c()` function (short for "concatenate") and place multiple values inside the parentheses, separated by commas. For example, `c(1, 2, 3)` creates a numeric vector with three values. Simply using parentheses alone (like `(1, 2, 3)`) will **not** create a vector --- `c()` is required.

We then assign the result to an object using `<-` so we can work with it later.

**Any time you want to work with more than one value at once, you will use this approach to combine them into a single object.** This is one of the small building blocks that will show up again and again, so it is worth getting comfortable with it now.

**Practice: Create your first vector**

The code below is partially written for you. Replace the `___` inside `c()` with 5 numbers of your choice, separated by commas. (Remember: Including the object name on the second line displays the results on the screen.) Then click "Submit Answer" to check your work. Two hints are available.

```{r vectors-create-first, exercise=TRUE, exercise.cap="Practice: Create a numeric vector with five elements"}
my.first.vector <- c(___)
my.first.vector
```

```{r vectors-create-first-hint-1}
# Hint 1 of 2
# Replace the ___ with numbers separated by commas.
# For example, c(10, 20, 30, 40, 50) would work,
# but use any 5 numbers you like.
```

```{r vectors-create-first-hint-2}
# Hint 2 of 2: fill in the blank
my.first.vector <- c(___, ___, ___, ___, ___)
my.first.vector
```

```{r vectors-create-first-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("my\\.first\\.vector\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign your vector to an object called my.first.vector.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (is.null(.result)) {
    fail("Make sure you display my.first.vector by typing its name on the last line so the grader can check it.")
  }
  if (!is.numeric(.result)) {
    fail("Your vector should contain numbers, not character strings.")
  }
  if (length(.result) != 5) {
    fail("Your vector should contain exactly 5 numbers.")
  }
  pass("Good start! Your vector has five elements, each accessible by index position.")
})
```

Now let’s try something a bit more interesting: using a vector to store real political data.

Here are the two-party vote shares won by the Democratic presidential candidate in the last 12 presidential elections:

<div class="table-container">
  <table role="table" aria-label="Presidential Election Results from 1980 to 2024 showing Democratic vote shares and candidates">
    <caption>
      <strong>Presidential Election Results 1980-2024</strong><br>
      This table displays the two-party vote share won by Democratic presidential candidates, along with the names of both Democratic and Republican candidates for each election year.
    </caption>
    <thead>
      <tr>
        <th scope="col" id="year">Election Year</th>
        <th scope="col" id="dem-share">Democratic 2-Party Vote Share</th>
        <th scope="col" id="dem-candidate">Democratic Candidate</th>
        <th scope="col" id="rep-candidate">Republican Candidate</th>
      </tr>
    </thead>
    <tbody>
      <tr><th scope="row" headers="year">1980</th><td headers="dem-share">0.448</td><td headers="dem-candidate">Jimmy Carter</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1984</th><td headers="dem-share">0.408</td><td headers="dem-candidate">Walter Mondale</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1988</th><td headers="dem-share">0.461</td><td headers="dem-candidate">Michael Dukakis</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1992</th><td headers="dem-share">0.535</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1996</th><td headers="dem-share">0.547</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">Bob Dole</td></tr>
      <tr><th scope="row" headers="year">2000</th><td headers="dem-share">0.503</td><td headers="dem-candidate">Al Gore</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2004</th><td headers="dem-share">0.488</td><td headers="dem-candidate">John Kerry</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2008</th><td headers="dem-share">0.537</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">John McCain</td></tr>
      <tr><th scope="row" headers="year">2012</th><td headers="dem-share">0.520</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">Mitt Romney</td></tr>
      <tr><th scope="row" headers="year">2016</th><td headers="dem-share">0.511</td><td headers="dem-candidate">Hillary Clinton</td><td headers="rep-candidate">Donald Trump</td></tr>
      <tr><th scope="row" headers="year">2020</th><td headers="dem-share">0.513</td><td headers="dem-candidate">Joe Biden</td><td headers="rep-candidate">Donald Trump</td></tr>
      <tr><th scope="row" headers="year">2024</th><td headers="dem-share">0.493</td><td headers="dem-candidate">Kamala Harris</td><td headers="rep-candidate">Donald Trump</td></tr>
    </tbody>
  </table>
</div>

We can enter the share of the two-party vote won by the Democratic presidential candidates in the last 12 elections as elements of a single vector we will name **dem.pres.vote.share**.

**Practice: Create a data vector**

The code below gives you the structure. Replace `___(___)` with the correct function call containing all 12 vote share values from the table above, in chronological order. Listing the object name in the second line displays the vector on the screen so you can check your work. Click "Submit Answer" when ready. Three hints are available.

```{r vectors-election-data, exercise=TRUE, exercise.cap="Practice: Create the Democratic vote share vector"}
dem.pres.vote.share <- ___(___)
dem.pres.vote.share
```

```{r vectors-election-data-hint-1}
# Hint 1 of 3
# You need to store a collection of 12 numbers together
# in a single object. What have you learned about how R
# stores multiple values in one place?
```

```{r vectors-election-data-hint-2}
# Hint 2 of 3
# In the scaffold, the outer ___ is the function name
# and the inner ___ is where your values go.
# What function combines multiple values into a vector?
# How do you separate values inside that function?
```

```{r vectors-election-data-hint-3}
# Hint 3 of 3: fill in the missing values
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535,
                         0.547, ___, ___, ___,
                         ___, ___, ___, ___)
dem.pres.vote.share
```


```{r vectors-election-data-check}
grade_this({
  correct_vals <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                    0.537, 0.520, 0.511, 0.513, 0.493)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("dem\\.pres\\.vote\\.share\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.pres.vote.share.")
  }

  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }

  if (is.null(.result)) {
    fail("Make sure you display dem.pres.vote.share by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should contain 12 values, one per election from 1980 to 2024.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check that each vote share is typed as a number.")
  }

  wrong_pos <- which(abs(.result - correct_vals) > 0.001)
  if (length(wrong_pos) > 0) {
    fail(paste0("Check your value at position ", wrong_pos[1],
                " --- that corresponds to the ", 1979 + wrong_pos[1] * 4,
                " election."))
  }

  pass("Exactly right! Your vector captures the Democratic two-party vote share across all 12 elections, from Carter's 44.8% in 1980 to Harris's 49.3% in 2024. You will use this vector in the exercises that follow.")
})
```

The `c()` function can also be used to combine multiple vectors into a single larger one.

**Practice: Create the first group of candidate names**

Using the Democratic candidate column in the table above, create a character vector called **first.6.candidates** containing the last names of the Democratic candidates for the first 6 elections (1980--2000). Display it by typing its name to check your work. Click "Submit Answer" when ready. Three hints are available.

```{r vectors-first-candidates, exercise=TRUE, exercise.cap="Practice: Create a character vector of the first six Democratic candidates"}

```

```{r vectors-first-candidates-hint-1}
# Hint 1 of 3
# You need to store 6 text values in a single object.
# Think about what you have learned about how R stores
# text, and how you combine multiple values into one object.
```

```{r vectors-first-candidates-hint-2}
# Hint 2 of 3
# Character strings need quotation marks around each name.
# Use c() to combine them, separated by commas.
# Assign the result to first.6.candidates and display it by typing its name.
```

```{r vectors-first-candidates-hint-3}
# Hint 3 of 3: fill in the blanks
first.6.candidates <- c("Carter", "___", "___",
                        "___", "___", "___")
first.6.candidates
```

```{r vectors-first-candidates-check}
grade_this({
  correct_vals <- c("Carter", "Mondale", "Dukakis",
                    "Clinton", "Clinton", "Gore")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("first\\.6\\.candidates\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called first.6.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (is.null(.result)) {
    fail("Make sure you display first.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your vector should contain character strings. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 6) {
    fail("Your vector should contain exactly 6 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The names do not match. Check spelling and order against the table.")
  }
  pass("Correct! Notice that Clinton appears twice --- Bill Clinton ran in both 1992 and 1996. You will combine this vector with the second group in the next exercise.")
})
```

**Practice: Create the second group of candidate names**

Now create a character vector called **last.6.candidates** containing the last names of the Democratic candidates for the final 6 elections (2004--2024). Display it by typing its name to check your work. Click "Submit Answer" when ready. Two hints are available.

```{r vectors-last-candidates, exercise=TRUE, exercise.cap="Practice: Create a character vector of the last six Democratic candidates"}

```

```{r vectors-last-candidates-hint-1}
# Hint 1 of 2
# Use the same approach as the previous exercise.
# Look at the Democratic candidate column in the table
# for elections from 2004 to 2024.
```

```{r vectors-last-candidates-hint-2}
# Hint 2 of 2: fill in the blanks
last.6.candidates <- c("Kerry", "___", "___",
                       "___", "___", "___")
last.6.candidates
```

```{r vectors-last-candidates-check}
grade_this({
  correct_vals <- c("Kerry", "Obama", "Obama",
                    "Clinton", "Biden", "Harris")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("last\\.6\\.candidates\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called last.6.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (is.null(.result)) {
    fail("Make sure you display last.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your vector should contain character strings. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 6) {
    fail("Your vector should contain exactly 6 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The names do not match. Check spelling and order against the table.")
  }
  pass("Right! Notice that Obama appears twice and Clinton appears once --- you will see repeated names when combining these vectors.")
})
```

**Practice: Combine the two vectors**

Now we can put the two pieces together. Combine **first.6.candidates** and **last.6.candidates** into a single vector called **dem.candidates** and display it by typing its name. Click "Submit Answer" when ready. Three hints are available.


```{r prepare-vectors-first-last, echo=FALSE}
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
```

```{r vectors-combine-names, exercise=TRUE, exercise.setup="prepare-vectors-first-last", exercise.cap="Practice: Combine two vectors into dem.candidates"}

```

```{r vectors-combine-names-hint-1}
# Hint 1 of 3
# You have two separate vectors and need to join them
# into one longer vector. Think about what function you
# have used to combine multiple values into a single
# object --- does the same approach work when the inputs
# are whole vectors rather than individual values?
```

```{r vectors-combine-names-hint-2}
# Hint 2 of 3
# The c() function works on whole vectors the same way
# it works on individual values. Pass both vector names
# as arguments, separated by a comma. Assign the result
# to dem.candidates and display it by typing its name.
```

```{r vectors-combine-names-hint-3}
# Hint 3 of 3: fill in the blanks
dem.candidates <- c(___, ___)
dem.candidates
```

```{r vectors-combine-names-check}
grade_this({
  correct_vals <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore",
                    "Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("dem\\.candidates\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to combine the two vectors.")
  }
  if (is.null(.result)) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your result should be a character vector.")
  }
  if (length(.result) != 12) {
    fail("Your combined vector should have 12 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The combined vector does not look right. Make sure you are combining first.6.candidates and last.6.candidates in that order.")
  }
  pass("Exactly! The combined vector has all 12 candidates in chronological order, ready to use alongside the vote share data.")
})
```

```{r prepare-vectors-election-data, echo=FALSE}
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
```

```{r prepare-vectors-combine-names, echo=FALSE}
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
```


## Indexing Vectors

Once you have values stored in a vector, the next useful skill is learning how to pull out the pieces you need. **To select a specific element in a vector, you use square brackets `[ ]`.** The
number inside the brackets is called the *[index]{.important-text}* --- it tells
R which position in the vector you want. Position 1 is the first element,
position 2 is the second, and so on. Here is a quick reference showing the index
position of each value in **dem.pres.vote.share**:

```{r vectors-position-table, echo=FALSE, results='asis'}
vals <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503,
          0.488, 0.537, 0.520, 0.511, 0.513, 0.493)

make_pos_table <- function(positions, values, label) {
  cat('<div class="table-container">')
  cat(sprintf('<table role="table" aria-label="%s">', label))
  cat('<thead><tr>')
  for (p in positions) cat(sprintf('<th scope="col">[%d]</th>', p))
  cat('</tr></thead><tbody><tr>')
  for (v in values) cat(sprintf('<td>%s</td>', v))
  cat('</tr></tbody></table></div>')
}

make_pos_table(1:6, vals[1:6],
  "Index positions 1 through 6 of dem.pres.vote.share")
make_pos_table(7:12, vals[7:12],
  "Index positions 7 through 12 of dem.pres.vote.share")
```

To extract an element, place its index number inside square brackets after the vector name: `dem.pres.vote.share[3]` returns the third value, 0.461.

**Practice: Access a single vector element**

The code below extracts the first element. Modify it to extract Hillary Clinton's vote share instead. Click "Submit Answer" to check your work. Two hints are available.

```{r vectors-access-single, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice: Extract Hillary Clinton's vote share by index position"}
dem.pres.vote.share[1]
```

```{r vectors-access-single-hint-1}
# Hint 1 of 2
# Use the position table above to find which index
# position corresponds to the 2016 election.
# Then replace the 1 in the brackets with that number.
```

```{r vectors-access-single-hint-2}
# Hint 2 of 2: fill in the blank
dem.pres.vote.share[___]
```

```{r vectors-access-single-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("\\[\\s*10\\s*\\]", code_no_comments)) {
    fail("Make sure you are using index position 10 inside the square brackets.")
  }

  if (is.null(.result)) {
    fail("Make sure you display the value you extracted so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("You should be extracting a numeric value.")
  }

  if (length(.result) != 1) {
    fail("You should be extracting a single value.")
  }

  if (is.na(.result)) {
    fail("Your result is NA. Check the index position inside the square brackets.")
  }

  if (!isTRUE(all.equal(.result, 0.511, tolerance = 0.001))) {
    fail("That is not quite right. Use the position table to find which index corresponds to 2016.")
  }

  pass("Correct! Hillary Clinton received 51.1% of the two-party vote in 2016.")
})
```

We can extract more than one element by placing multiple index positions inside the square brackets. Because R expects a single object inside the brackets, we combine those positions using the `c()` function: `vectorname[c(2, 5, 7)]` extracts the elements at positions 2, 5, and 7 in that order.

**Practice: Access multiple vector elements**

What was the vote share won by the Democrat in the last 5 elections? Use the position table above to find the correct index positions. Click "Submit Answer" to check your work. Three hints are available.

```{r vectors-access-multiple, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice: Extract the last five Democratic vote shares"}

```

```{r vectors-access-multiple-hint-1}
# Hint 1 of 3
# You need to extract 5 values at once.
# How do you tell R to look at more than one
# position at the same time?
```

```{r vectors-access-multiple-hint-2}
# Hint 2 of 3
# Use c() inside square brackets to specify
# multiple positions. Use the position table
# above to find which index numbers you need.
```

```{r vectors-access-multiple-hint-3}
# Hint 3 of 3: fill in the blanks
dem.pres.vote.share[c(___, ___, ___, ___, ___)]
```

```{r vectors-access-multiple-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  expected <- c(0.537, 0.520, 0.511, 0.513, 0.493)
  if (!grepl("\\[", code_no_comments) || !grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using square brackets with c() inside to select multiple elements.")
  }
  if (is.null(.result)) {
    fail("Make sure you display the extracted values so the grader can check them.")
  }
  if (!is.numeric(.result) || length(.result) != 5) {
    fail("Your result should be a numeric vector with exactly 5 values.")
  }
  if (!identical(.result, expected)) {
    fail("The values are not quite right or are out of order. Make sure you are listing the last 5 elections in chronological order.")
  }
  pass("Nicely done! The last five elections returned 0.537, 0.520, 0.511, 0.513, and 0.493 --- a narrow range showing how competitive recent presidential elections have been.")
})
```

Note that the order in which you list the index numbers determines the order of the output --- `dem.pres.vote.share[c(9, 5)]` and `dem.pres.vote.share[c(5, 9)]` return the same two values but in opposite order.

**Run and observe: Order of index numbers**

Use the **Run Code** button to see this in action:

```{r vectors-order-matters, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Run and observe: Index order determines output order"}
dem.pres.vote.share[c(9, 5)]
dem.pres.vote.share[c(5, 9)]
```

We can also use a minus sign in the index to remove an element from a vector. For example, to return all values except the last election (2024, position 12):

**Run and observe: Remove vector elements**

Use the **Run Code** button to see how negative indexing works:

```{r vectors-remove-elements, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Run and observe: Negative indexing removes an element"}
dem.pres.vote.share[-12]
```

## Vector Arithmetic

We can apply arithmetic operations on a numeric vector. *The operation will be repeated for each element of the vector.* We will often want to do this as part of creating new variables that are a function of variables in a data set.

**Practice: Convert vote shares to percentages**

Create a new vector called **dem.pres.vote.percent** that converts **dem.pres.vote.share** to the percent of the 2-party vote won by the Democratic candidate. Display it by typing its name to check your work. Click "Submit Answer" when ready. Three hints are available.

```{r vectors-convert-percent, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice: Convert vote shares to percentages"}

```

```{r vectors-convert-percent-hint-1}
# Hint 1 of 3
# A proportion and a percentage express the same value
# in different units. What single arithmetic operation
# converts a proportion to a percentage?
```

```{r vectors-convert-percent-hint-2}
# Hint 2 of 3
# Apply that operation to the entire dem.pres.vote.share
# vector. R will apply it to every element automatically.
# Assign the result to dem.pres.vote.percent and display it by typing its name.
```

```{r vectors-convert-percent-hint-3}
# Hint 3 of 3: fill in the blanks
dem.pres.vote.percent <- dem.pres.vote.share * ___
dem.pres.vote.percent
```

```{r vectors-convert-percent-check}
grade_this({
  expected <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                0.537, 0.520, 0.511, 0.513, 0.493) * 100
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("dem\\.pres\\.vote\\.percent\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.pres.vote.percent.")
  }
  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are applying the operation to dem.pres.vote.share.")
  }
  if (!grepl("\\*\\s*100|100\\s*\\*", code_no_comments)) {
    fail("Think about what arithmetic operation converts a proportion to a percentage.")
  }
  if (is.null(.result)) {
    fail("Make sure you display dem.pres.vote.percent by typing its name on the last line so the grader can check it.")
  }
  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }
  if (length(.result) != 12) {
    fail("Your result should have 12 values.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.01))) {
    fail("Your values do not look right. Check that you are multiplying by 100.")
  }
  pass("Exactly right! The values now range from about 40.8 to 54.7 --- the same information expressed as percentages rather than proportions.")
})
```

Vector arithmetic can also express each value relative to a reference point — subtracting a single baseline value from every element. This is useful any time you want to know not just what a value is, but how much it differs from a particular starting point, such as the first year in a series, a control group mean, or a historical benchmark.

**Practice: Calculate differences from baseline**

Create a vector showing how each election's Democratic vote share compares to 1980. Click "Submit Answer" when ready. Three hints are available.

```{r vectors-compare-baseline, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice: Calculate each election's vote share relative to 1980"}

```

```{r vectors-compare-baseline-hint-1}
# Hint 1 of 3
# "Compared to 1980" means each value minus the 1980
# value. What do you know about how R applies arithmetic
# operations across an entire vector?
```

```{r vectors-compare-baseline-hint-2}
# Hint 2 of 3
# You need to subtract a single value --- the 1980
# vote share --- from every element. You know how to
# extract a single element from a vector by its index.
# What index position holds the 1980 value?
```

```{r vectors-compare-baseline-hint-3}
# Hint 3 of 3: fill in the blanks
compared.to.1980 <- dem.pres.vote.share - dem.pres.vote.share[___]
compared.to.1980
```

```{r vectors-compare-baseline-check}
grade_this({
  baseline <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- baseline - baseline[1]

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("compared\\.to\\.1980\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called compared.to.1980.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using dem.pres.vote.share in your calculation.")
  }

  if (grepl("\\[\\s*1980\\s*\\]", code_no_comments)) {
    fail("Use the position of the 1980 value, not the year itself. The 1980 election is the first value in dem.pres.vote.share, so its index position is 1.")
  }

  if (!grepl("\\[\\s*1\\s*\\]", code_no_comments)) {
    fail("Use indexing to extract the 1980 value. It is at position 1 in the vector.")
  }

  if (is.null(.result)) {
    fail("Make sure you display compared.to.1980 by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your result should have 12 values, one for each election.")
  }

  if (!isTRUE(all.equal(.result[1], 0, tolerance = 0.001))) {
    fail("The first value should be 0 since you are comparing 1980 to itself.")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Make sure you are subtracting the 1980 value from every element.")
  }

  pass("Correct! Positive values mean the candidate did better than Carter in 1980; negative values mean worse.")
})
```


## Modifying Vectors

We can also use indexing to change selected values in a vector. Instead of using square brackets only to look at certain positions, we can use them together with the assignment operator to replace the values stored in those positions. The general pattern is `vector_name[position] <- new_value`. When we want to replace more than one value at once, we can put several positions inside `c()` and provide the same number of replacement values.

Below is a position reference table for **dem.candidates**.

```{r vectors-candidate-position-table, echo=FALSE, results='asis'}
vals <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore",
          "Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")

make_pos_table <- function(positions, values, label) {
  cat('<div class="table-container">')
  cat(sprintf('<table role="table" aria-label="%s">', label))
  cat('<thead><tr>')
  for (p in positions) cat(sprintf('<th scope="col">[%d]</th>', p))
  cat('</tr></thead><tbody><tr>')
  for (v in values) cat(sprintf('<td>%s</td>', v))
  cat('</tr></tbody></table></div>')
}

make_pos_table(1:6, vals[1:6],
  "Index positions 1 through 6 of dem.candidates")
make_pos_table(7:12, vals[7:12],
  "Index positions 7 through 12 of dem.candidates")
```

**Practice: Replace vector elements**

Replace the three `"Clinton"` entries in **dem.candidates** with their full names: `"Bill Clinton"` for the two earlier Clinton entries and `"Hillary Clinton"` for the later Clinton entry. Use the position reference table above to find the correct index positions. Display the result by typing its name to check your work. Click "Submit Answer" when ready. Three hints are available.

```{r vectors-replace-elements, exercise=TRUE, exercise.setup="prepare-vectors-combine-names", exercise.cap="Practice: Replace the Clinton entries with full names"}

```

```{r vectors-replace-elements-hint-1}
# Hint 1 of 3
# You need to find which positions contain "Clinton"
# and then overwrite those positions with new values.
# Print dem.candidates first to see its contents,
# then use the position table to find the right indices.
```

```{r vectors-replace-elements-hint-2}
# Hint 2 of 3
# You can replace multiple positions at once using
# the same c() indexing you used to select elements.
# The replacement values must be in the same order
# as the positions you are replacing.
```

```{r vectors-replace-elements-hint-3}
# Hint 3 of 3: fill in the blanks
dem.candidates[c(___, ___, ___)] <- c("Bill Clinton", "___", "Hillary Clinton")
dem.candidates
```


```{r vectors-replace-elements-check}
grade_this({
  expected <- c("Carter", "Mondale", "Dukakis", "Bill Clinton",
                "Bill Clinton", "Gore", "Kerry", "Obama", "Obama",
                "Hillary Clinton", "Biden", "Harris")

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("dem\\.candidates\\s*\\[", code_no_comments)) {
    fail("Use square bracket indexing on dem.candidates to replace specific positions, like dem.candidates[...] <- ...")
  }

  if (!grepl("<-", code_no_comments)) {
    fail("Use the assignment operator <- to replace values at the indexed positions.")
  }

  if (is.null(.result)) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }

  if (!is.character(.result)) {
    fail("Your result should be a character vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should still have 12 elements. You are replacing values, not adding or removing them.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check the replacement values you typed.")
  }

  if (any(.result == "Clinton", na.rm = TRUE)) {
    fail("You still have entries that say only 'Clinton'. Make sure you replaced all three with full names.")
  }

  if (!identical(.result, expected)) {
    fail("Your vector does not match the expected result. Check the full names and their index positions.")
  }

  pass("Nice work! Using full names avoids ambiguity --- your vector now distinguishes Bill Clinton (1992, 1996) from Hillary Clinton (2016).")
})
```


## Summary Functions

Now that you have worked with vectors, you are ready for a few functions that summarize them. These functions are simple to use, and they will become part of your regular R toolkit.

* `length()`: returns the number of elements in the vector.
* `min()`: returns the minimum value.
* `max()`: returns the maximum value.
* `range()`: returns a two-element vector containing the minimum and maximum values --- not a single "range" number, but both endpoints together.
* `mean()`: returns the arithmetic mean.
* `sum()`: returns the sum of all values.
* `abs()`: returns the absolute value of a number or each element of a numeric vector, removing any negative sign.

Many summary functions include an argument called `na.rm`, which stands for "remove missing values" and has a default value of `FALSE`. When `na.rm = FALSE`, missing values are left in place, and functions such as `mean()` will usually return `NA` if any value is missing. The `length()` function is different: it simply counts elements, including any `NA` values.

**Run and observe: Apply summary functions**

Use the **Run Code** button to see each function applied to our vote data. This
exercise produces several lines of output, so read them one at a time:

```{r functions-summary-stats, exercise=TRUE, exercise.setup="prepare-vectors-combine-names", exercise.cap="Run and observe: Summary functions applied to vote data"}
length(dem.candidates)
min(dem.pres.vote.share)
max(dem.pres.vote.share)
range(dem.pres.vote.share)
mean(dem.pres.vote.share)
sum(dem.pres.vote.share) / length(dem.pres.vote.share)
abs(-0.055)
```

The last two expressions are worth noting. The second-to-last shows that the mean can also be calculated by hand as the sum of all values divided by the number of values --- it should match the `mean()` output just above it. The final expression shows that `abs()` removes the negative sign: `-0.055` and `0.055` both return `0.055`. This is useful any time you need a difference but do not care about direction, such as a margin of victory. Notice also that `range()` returns two numbers --- the minimum and maximum --- not a single value.

### Default values and named arguments

Every argument either has a **[default value]{.important-text}** or is
**required**. When an argument has a default, you only need to specify it if you
want something different from the default behavior. When an argument has no
default, you must supply a value or R will return an error.

You saw this earlier when you ran `args(round)`. The output showed two arguments: `x`, which has no default and is required, and `digits`, which has a default of `0`. This means `round(3.14159)` works fine --- R rounds to zero decimal places by default. But `round(digits = 2)` would fail because `x` is required and you have not supplied it.

You can supply arguments in two ways. The first is **positional** --- R matches your inputs to arguments in the order they are listed in `args()`. The second is **by name** --- you explicitly write `argument = value`, which allows you to supply arguments in any order. For all arguments after the first, it is good practice to use names, both to avoid mistakes and to make your code easier to read.

**Run and observe: Default values and named arguments**

Use the **Run Code** button to see these ideas in action:

```{r functions-named-args-demo, exercise=TRUE, exercise.cap="Run and observe: Default values and named arguments with round()"}
round(3.14159)
round(3.14159, digits = 2)
round(digits = 2, x = 3.14159)
```

All three calls produce sensible results. The first uses the default for `digits`. The second overrides it. The third supplies arguments by name in reverse order --- R still knows what to do because the argument names make the intent unambiguous.

## Missing Values

Real data often contains **[missing values]{.important-text}**, which R
represents as `NA`. When a vector contains `NA`, most summary functions return
`NA` by default rather than attempting to compute a result from incomplete data.

**Run and observe: Missing values**

Use the **Run Code** button to see what happens when a vector contains a missing value, and how `na.rm = TRUE` fixes it:

```{r functions-missing-values, exercise=TRUE, exercise.cap="Run and observe: How NA affects summary functions and how na.rm fixes it"}
vote.with.missing <- c(0.448, 0.408, NA, 0.535, 0.547)
mean(vote.with.missing)
mean(vote.with.missing, na.rm = TRUE)
```

The first call returns `NA` because R cannot compute a mean when some values are unknown. The second call returns the mean of the four non-missing values by telling R to remove the `NA` before computing. The argument `na.rm = TRUE` overrides the default of `FALSE`.

None of the vectors we created have missing values, so the default behavior is fine for the exercises in this tutorial. But in real data you will encounter `NA` regularly, and `na.rm = TRUE` is how you handle it.

## Functions with Multiple Inputs

As we just saw, functions can take more than one argument. The `seq()` function is a good example --- it generates a sequence of numbers using three arguments: `from` (where to start), `to` (where to end), and `by` (the step size).

**Run and observe: Create a number sequence**

Use the **Run Code** button to see `seq()` in action:

```{r functions-create-sequence, exercise=TRUE, exercise.cap="Run and observe: Creating a sequence with seq()"}
my.sequence <- seq(from = 1, to = 10, by = 2)
my.sequence
```

Notice that all three arguments are named explicitly. This is good practice with multi-argument functions --- it makes the code easy to read and protects against mistakes from argument order.

**Practice: Create election year sequence**

Using `seq()`, create a vector called **elect.year** containing every presidential election year from 1980 to 2024. Display it by typing its name to check your work. Click "Submit Answer" when ready. Three hints are available.

```{r functions-election-years, exercise=TRUE, exercise.cap="Practice: Create a sequence of presidential election years"}

```

```{r functions-election-years-hint-1}
# Hint 1 of 3
# Think about the three things seq() needs to know:
# where to start, where to end, and how much to
# increase by each step. How often are elections held?
```

```{r functions-election-years-hint-2}
# Hint 2 of 3
# Presidential elections occur every 4 years.
# The sequence starts at 1980 and ends at 2024.
# Use named arguments for all three inputs.
```

```{r functions-election-years-hint-3}
# Hint 3 of 3: fill in the blanks
elect.year <- seq(from = ___, to = ___, by = ___)
elect.year
```

```{r functions-election-years-check}
grade_this({
  expected <- seq(from = 1980, to = 2024, by = 4)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("elect\\.year\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called elect.year.")
  }

  if (!grepl("seq\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the seq() function.")
  }

  if (grepl("by\\s*=\\s*2(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("seq\\s*\\([^)]*,\\s*[^)]*,\\s*2\\s*\\)", code_no_comments)) {
    fail("Check your step size --- presidential elections are 4 years apart, so the by argument should be 4.")
  }

  if (grepl("from\\s*=\\s*1(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("to\\s*=\\s*12(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("seq\\s*\\(\\s*1\\s*,\\s*12", code_no_comments)) {
    fail("Use the election years themselves, not the position numbers. The sequence should start at 1980 and end at 2024.")
  }

  if (!grepl("1980", code_no_comments)) {
    fail("Check your starting value --- what year did this series of elections begin?")
  }

  if (!grepl("2024", code_no_comments)) {
    fail("Check your ending value --- what is the most recent election year in the table?")
  }

  if (is.null(.result)) {
    fail("Make sure you display elect.year by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != length(expected)) {
    fail("Check your step size --- presidential elections are 4 years apart.")
  }

  if (!isTRUE(all.equal(.result, expected))) {
    fail("Your sequence does not match. Check your starting value, ending value, and step size.")
  }

  pass("Good work! Your sequence covers all presidential election years from 1980 to 2024.")
})
```

Note that decreasing sequences work the same way with a negative `by` value.
This is useful when the order itself matters, such as listing years from most
recent to oldest.

**Run and observe: Create a decreasing sequence**

Use the **Run Code** button to see a sequence that counts backwards:

```{r functions-decreasing-sequence, exercise=TRUE, exercise.cap="Run and observe: A sequence counting backwards"}
seq(from = 2024, to = 1980, by = -4)
```

The `names()` function lets you assign labels to the elements of a vector. Names are not part of the data itself but are a useful attribute that makes output easier to read.

```{r functions-prepare-names, echo=FALSE}
elect.year <- seq(from = 1980, to = 2024, by = 4)
```

**Run and observe: Assign names to a vector**

Use the **Run Code** button to see how names are assigned and displayed:

```{r functions-assign-names, exercise=TRUE, exercise.setup="functions-prepare-names", exercise.cap="Run and observe: Assigning names to vector elements"}
names(dem.pres.vote.share)
names(dem.pres.vote.share) <- elect.year
names(dem.pres.vote.share)
dem.pres.vote.share
```

The first line confirms there are no names yet. The second assigns the election years as names. The third and fourth lines show the result --- the vector now displays with year labels above each value.

## Packages

Base R contains many functions, but it does not contain everything we might want to use. Much of R's power comes from **[packages]{.important-text}**, which are add-ons that bundle together code, data, and help files. You can think of a package like an app on your phone: installing it puts it on your device, but opening it is what makes it available to use. In R, `install.packages()` installs a package, and `library()` loads it for the current R session.

There are tens of thousands of packages written for R, many of them contributed to the Comprehensive R Archive Network (CRAN). R does not automatically install or load all of them, because that would take unnecessary space and memory. Instead, you install the packages you need and load them when you want to use them.

For example, we will use the `ggplot()` function from the [ggplot2]{.package-name} package. You only need to install [ggplot2]{.package-name} once on your computer, but you need to load it with `library(ggplot2)` each time you start a new R session and want to use its functions.

If you try to use a function from a package that is not loaded, R will return an error.

**Run and observe: See a package error**

Use the **Run Code** button to see what happens when you call a function from an unloaded package:

```{r packages-error-demo, exercise=TRUE, exercise.cap="Run and observe: Error from calling a function in an unloaded package"}
ggplot()
```

Any time you see an error that begins "could not find function," the most likely cause is that you have not loaded the package that contains it.

**Run and observe: Load a package and use a function**

Use the **Run Code** button to see what happens when you load the package first:

```{r packages-load-demo, exercise=TRUE, exercise.cap="Run and observe: Loading a package with library() and calling a function"}
library(ggplot2)
ggplot()
```

You will see a blank gray plotting canvas --- that is normal. The `ggplot()` function creates an empty canvas waiting for data and plot specifications. The important thing is that it runs without error, confirming the package loaded successfully.

To get help with an entire package, use the `help()` function with the `package`
argument set to the package name in quotes. The output looks cleaner in RStudio
than it does here. Do not try to read every line now; just notice that package
help gives you a list of available functions, documentation, and examples.

**Run and observe: Get package help**

Use the **Run Code** button to access help for the [ggplot2]{.package-name} package:

Note: The output lists the help resources available for [ggplot2]{.package-name}. You do not need to read all of it now; just notice that package-level help is a way to see what documentation and functions come with a package. In RStudio, this will appear in the Help pane and be easier to browse.

```{r packages-get-help, exercise=TRUE, exercise.cap="Run and observe: Package-level help with help(package = ...)"}
help(package = "ggplot2")
```

## Comments

You can add comments to your code using the `#` symbol. Anything on a line after `#` is ignored by R and will not be executed. Comments are useful for noting why you wrote a particular piece of code, what a function does, or what arguments it takes.

**Run and observe: Use code comments**

Use the **Run Code** button to see how comments work alongside executable code:

```{r comments-demo, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Run and observe: Comments alongside executable code"}
# Calculate the mean Democratic vote share across all elections
mean(dem.pres.vote.share, na.rm = TRUE)
```

Notice that the comment line is ignored entirely and only the `mean()` call produces output. Since our vector has no missing values, the result is the same whether or not we include `na.rm = TRUE` --- but it is good practice to be explicit.

## Check Your Understanding

The following exercises combine concepts you have learned throughout this tutorial. Take your time and think through each problem step by step. You have seen all of the pieces already; the work now is deciding how to put them together.

**Apply: Calculate Republican vote shares**

Create a vector called **rep.pres.vote.share** containing the share of the two-party vote received by the Republican presidential candidate. Display it by typing its name. You do not need to type in the vote shares --- in a two-party race you can calculate them directly from the Democratic vote shares. Click "Submit Answer" when ready. Three hints are available.

```{r practice-republican-shares, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Apply: Calculate Republican two-party vote shares"}

```

```{r practice-republican-shares-hint-1}
# Hint 1 of 3
# In a two-party race, every vote goes to one party
# or the other. What must be true about how the two
# vote shares relate to each other?
```

```{r practice-republican-shares-hint-2}
# Hint 2 of 3
# The two shares sum to 1. You already have the
# Democratic vote share vector. Express the Republican
# share as an arithmetic operation on that vector ---
# R will apply it to every element automatically.
```

```{r practice-republican-shares-hint-3}
# Hint 3 of 3: fill in the blanks
rep.pres.vote.share <- ___ - dem.pres.vote.share
rep.pres.vote.share
```


```{r practice-republican-shares-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- 1 - dem

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("rep\\.pres\\.vote\\.share\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called rep.pres.vote.share.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using dem.pres.vote.share in your calculation rather than typing the values manually.")
  }

  if (!grepl("1\\s*-\\s*dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Use the form 1 - dem.pres.vote.share to calculate the Republican vote share.")
  }

  if (is.null(.result)) {
    fail("Make sure you display rep.pres.vote.share by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should have 12 values, one for each election.")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Remember that Republican share equals 1 minus the Democratic share.")
  }

  pass("Well done! Republican candidates averaged about 50.3% of the two-party vote across these 12 elections.")
})
```

**Apply: Calculate victory margins**

Create a vector called **victory.margin** that contains the margin of victory for each election --- that is, the absolute difference between the two parties' vote shares. You will need the `abs()` function, which returns the absolute value of a number. Display it by typing its name. Click "Submit Answer" when ready. Three hints are available.

```{r practice-setup-for-margins, echo=FALSE}
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
rep.pres.vote.share <- 1 - dem.pres.vote.share
```

```{r practice-victory-margins, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply: Calculate the margin of victory in each election"}

```

```{r practice-victory-margins-hint-1}
# Hint 1 of 3
# A margin of victory is how much the winner beat
# the loser by. You have both vote share vectors.
# Why might the sign of the difference be a problem,
# given that you do not know which party won each year?
```

```{r practice-victory-margins-hint-2}
# Hint 2 of 3
# The difference will be positive in some years and
# negative in others depending on which party won.
# A margin is always positive, so you need a function
# that removes the sign. The prompt names it for you.
```

```{r practice-victory-margins-hint-3}
# Hint 3 of 3: fill in the blanks
victory.margin <- abs(___ - ___)
victory.margin
```

```{r practice-victory-margins-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  rep_share <- 1 - dem
  expected <- abs(rep_share - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("victory\\.margin\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the result to an object called victory.margin.")
  }

  if (!grepl("abs\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the abs() function to ensure all margins are positive.")
  }

  if (!grepl("rep\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using rep.pres.vote.share in your calculation.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using dem.pres.vote.share in your calculation.")
  }

  if (is.null(.result)) {
    fail("Make sure you display victory.margin by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should have 12 values.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check that your calculation uses the two vote-share vectors correctly.")
  }

  if (any(.result < 0, na.rm = TRUE)) {
    fail("Victory margins should always be positive. Make sure you are using abs().")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Make sure you are taking the absolute value of the difference between the two vote share vectors.")
  }

  pass("That is correct! The closest election in this set was 2000, when Gore and Bush were separated by less than half a percentage point of the two-party vote.")
})
```

**Apply: Create Republican candidate names**

Create a character vector called **rep.names** containing the last names of the Republican candidates in order from 1980 to 2024: Reagan, Reagan, Bush, Bush, Dole, Bush, Bush, McCain, Romney, Trump, Trump, Trump. Display it by typing its name. Click "Submit Answer" when ready. Three hints are available.

```{r practice-republican-names, exercise=TRUE, exercise.cap="Apply: Create a vector of Republican candidate names"}

```

```{r practice-republican-names-hint-1}
# Hint 1 of 3
# You need to store 12 names as a single vector.
# What have you learned about how to store text in R,
# and how to combine multiple values into one object?
```

```{r practice-republican-names-hint-2}
# Hint 2 of 3
# Each name is a character string and needs quotation
# marks. Use c() to combine all 12 into a single vector.
# Assign the result to rep.names and display it by typing its name.
```

```{r practice-republican-names-hint-3}
# Hint 3 of 3: fill in the blanks
rep.names <- c("Reagan", "Reagan", "___", "___",
               "___", "___", "___", "___",
               "___", "___", "___", "___")
rep.names
```

```{r practice-republican-names-check}
grade_this({
  expected <- c("Reagan", "Reagan", "Bush", "Bush", "Dole", "Bush",
                "Bush", "McCain", "Romney", "Trump", "Trump", "Trump")

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("rep\\.names\\s*(<-|=)", code_no_comments)) {
    fail("Make sure you assign the names to an object called rep.names.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to combine the names into a vector.")
  }
  if (is.null(.result)) {
    fail("Make sure you display rep.names by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your result should be a character vector. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 12) {
    fail("Your vector should contain 12 names, one for each election from 1980 to 2024.")
  }
  if (!identical(.result, expected)) {
    fail("Some names do not match. Check spelling and order against the election table.")
  }
  pass("Right! Notice that Trump appears three times --- the only candidate in this set to run in three consecutive elections.")
})
```

**Apply: Calculate mean Republican vote share**

Using the object **rep.pres.vote.share** you just created, find the mean share of the two-party vote received by Republican candidates. There are no missing values, so you do not need to specify `na.rm`. Click "Submit Answer" when ready. Two hints are available.

```{r practice-mean-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply: Find the mean Republican two-party vote share"}

```

```{r practice-mean-republican-hint-1}
# Hint 1 of 2
# You need a single number summarizing the typical
# value across all elements of a vector. Which function
# from the "Summary Functions" section does this?
```

```{r practice-mean-republican-hint-2}
# Hint 2 of 2: fill in the blank
___(rep.pres.vote.share)
```

```{r practice-mean-republican-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- mean(1 - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("mean\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the mean() function.")
  }
  if (!grepl("mean\\s*\\(\\s*rep\\.pres\\.vote\\.share\\s*(,|\\))", code_no_comments)) {
    fail("Make sure you are applying mean() to rep.pres.vote.share.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("That is not quite right. Make sure you are applying mean() to the Republican vote share vector.")
  }
  pass(paste0("Exactly! The mean is a useful single-number summary --- Republican candidates averaged ",
              round(expected * 100, 1),
              "% of the two-party vote. Compare this to the Democratic mean to see how evenly matched the two parties have been."))
})
```

**Apply: Find Republican vote share range**

Using the object **rep.pres.vote.share** you just created, find the range of the two-party vote share received by Republican candidates. Click "Submit Answer" when ready. Two hints are available.

```{r practice-range-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply: Find the range of Republican two-party vote shares"}

```

```{r practice-range-republican-hint-1}
# Hint 1 of 2
# You need both the smallest and largest values in
# the vector at once. Which function from the "Summary
# Functions" section returns both endpoints together?
```

```{r practice-range-republican-hint-2}
# Hint 2 of 2: fill in the blank
___(rep.pres.vote.share)
```

```{r practice-range-republican-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- range(1 - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("range\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the range() function.")
  }
  if (!grepl("range\\s*\\(\\s*rep\\.pres\\.vote\\.share\\s*(,|\\))", code_no_comments)) {
    fail("Make sure you are applying range() to rep.pres.vote.share.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }
  if (length(.result) != 2) {
    fail("Remember that range() returns two values: the minimum and the maximum.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("That is not quite right. Make sure you are applying range() to the Republican vote share vector.")
  }
  pass(paste0("Correct! Republican candidates' two-party vote share ranged from ",
              round(expected[1] * 100, 1),
              "% to ",
              round(expected[2] * 100, 1),
              "% in these elections."))
})
```

## The Takeaways

This tutorial introduced you to the fundamental building blocks of working with data in R. These concepts will underlie everything you do in this course.

R serves two roles simultaneously: it is a calculator that performs mathematical operations, and a programming environment that stores, manipulates, and analyzes data. The bridge between these two roles is the **object** --- a named container that holds any kind of value. You assign values to objects with `<-`, and R keeps track of each object's **class** so it knows what operations are valid. The three classes you encountered here are numeric, character, and function.

Most real data analysis involves working with many values at once, which is why **vectors** are so central to R. You create them with `c()`, access individual elements with square bracket indexing [], and apply arithmetic operations that automatically repeat across every element. These simple building blocks --- creating, indexing, and transforming vectors --- will appear in nearly every piece of code you write this semester.

**Functions** are what make R powerful. Every function takes one or more arguments and returns an output. You learned to use `args()` to see what arguments a function accepts, `?` to read what a function does, and the distinction between required arguments and optional ones with default values. When a function lives in a package rather than base R, you install it once with `install.packages()` and load it each session with `library()`.

One theme runs through all of this: R requires precision. Mismatched parentheses, a missing comma, a misspelled object name --- any of these will produce an error. This is not a sign that you are bad at R. It is just how programming works, and you will get faster at spotting these issues with practice. The fact that you have worked through this tutorial means you already have the foundation you need to keep building.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic, context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-01, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Welcome to R, you are off to a great start!">
  <span aria-hidden="true">🎉</span>
  Welcome to R, you are off to a great start!
</p>
')
```
