---
title: "R Basics: Part 1"
tutorial:
  id: "01-R-Part1"
output:
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn the basics of how R works.
---

## Overview {#main-content}

This is the first in a series of tutorials on analyzing political data. The series begins with the foundations of working in R and moves through describing variables, examining relationships, statistical inference, regression, and prediction. Throughout, the goal is not simply to produce code or statistical output, but to develop the statistical reasoning needed to choose an appropriate approach, understand what the analysis is doing, interpret the results accurately, and recognize what the evidence does --- and does not --- support.

No previous experience with R is assumed. Each tutorial builds on concepts and skills introduced earlier in the series, so you should work through them in order the first time. You can also return to a completed tutorial later to review a statistical idea or remind yourself how to use a particular function.

In this tutorial you will learn:

* How to use R as a calculator
* How to assign a value or values to an object with `<-`
* How to use functions
* How to determine the class of an object with the `class()` function
* How to create and use vectors using the `c()` function
* How to select and replace elements of a vector using indexing
* How to apply an arithmetic operation to every element of a vector at once
* How to summarize a vector with `length()`, `min()`, `max()`, `range()`, `mean()`, and `sum()`
* How to recognize missing values and handle them with the `na.rm` argument
* How to supply arguments by name and rely on default values, as with `seq()`
* How to install packages with `install.packages()` and load them with `library()`
* How to annotate your code with comments

## Before You Begin

The tutorials are built around learning by doing. Every activity is labeled and numbered, and the label tells you what is being asked of you. **Run and observe** gives you code to execute and examine. **Explore** does the same, then asks you to change something and run it again. **Practice** asks you to complete or write code yourself, or to answer a question about what you have just seen. **Apply** asks you to bring several skills together --- writing code that draws on more than one idea, judging whether a pattern supports a hypothesis, or explaining what a result means.

::: {.tip}
**Run Code vs. Submit Answer:** These two buttons do very different things. "Run Code" simply executes whatever is in the box and shows you the result --- you can use it as many times as you like to experiment without any consequences. Only "Submit Answer" checks your work. Get in the habit of using "Run Code" freely to try things out before you submit. If something does not work the way you expected, that is useful information --- try to understand why before moving on.
:::

Work through each tutorial from beginning to end. If you run out of time, or simply need a break, you can exit and return later; your progress is saved. A "Hint" button is available on graded exercises if you get stuck. Each tutorial closes with a "Submit Your Work" section that generates a code recording what you completed --- if you are not using these tutorials for a graded course, you can ignore it.

You will make mistakes as you work through these tutorials. R, like any coding language, requires careful attention to matching parentheses, including commas in the right places, and spelling functions, arguments, and object names correctly. When you encounter errors --- and you will do so often --- look for these types of mistakes first. If this feels picky at first, that is because R is picky; you are learning the rules of the language, and that takes practice.

::: {.tip}
**Take notes as you work.** A good strategy is to keep a running list of the functions introduced in each tutorial, the name of the tutorial that introduced them, and a short description of what they do. This will give you a reference sheet to consult whenever you need to write code on your own, or when you return to something introduced earlier in the series.
:::

That running list will also help you with one rule that applies throughout the series.

::: {.important-note}
**Function restriction:** In the graded exercises, stay within the tools the tutorials have introduced up to that point. Use only the functions, arguments, and approaches you have been taught, and do not use functions from packages the tutorial does not load. The point of an exercise is to practice the method just introduced rather than to reach the answer by another route --- and because the grader checks for that method, code that produces the requested result using different tools may be marked wrong. If you are unsure whether something is permitted, check whether it has appeared in this or an earlier tutorial before using it.
:::

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(gradethis)
library(learnrhash)
options(lifecycle_verbosity = "quiet")   # suppress tidyselect deprecation warnings from gradethis
options(help_type = "text")
tutorial_options(exercise.checker = gradethis::grade_learnr, exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513, 0.493)
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry","Obama","Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
my.first.object <- 5 + 3
another.object <- "100"
make_pos_table <- function(positions, values, label) {
  cat('<div class="table-container">')
  cat(sprintf('<table role="table" aria-label="%s">', label))
  cat('<thead><tr>')
  for (p in positions) cat(sprintf('<th scope="col">[%d]</th>', p))
  cat('</tr></thead><tbody><tr>')
  for (v in values) cat(sprintf('<td>%s</td>', v))
  cat('</tr></tbody></table></div>')
}
```

## Roadmap

Almost everything you will do in R comes down to two moves: storing values in named objects, and applying functions to those objects. This tutorial builds both from the ground up.

We start with R as a calculator, because it asks nothing new of you --- type an expression, R evaluates it, the answer appears. From there we introduce objects, so that a result can be kept and reused rather than recomputed every time you need it. Once values have names, we introduce functions, which are how you do something to them. Vectors come next, and they are the reason the rest of the series works the way it does: political data arrive as many values of the same kind, and R is built to operate on all of them at once rather than one at a time. The remaining sections fill in what you need to work with vectors comfortably --- pulling out particular elements, transforming them, summarizing them, and handling values that are missing.

By the end you will have written code that stores real election data, modifies it, and summarizes it. Almost everything in the tutorials that follow is a variation on those same moves.

## R as a Calculator

It is easy to use R like a calculator. Here's an example of adding 2 numbers.
You can add/subtract, multiply/divide, or raise a number to a power using `+`,
`-`, `*`, `/`, `^`.

**Explore 1: Basic arithmetic in R**

Try running the code below, then change the numbers and operators to practice
different calculations. Use the "Run Code" button to execute your code. Note
that spaces around the numbers and operators do not affect the calculation, so
14 + 2000 produces the same result as 14+2000.

```{r calculator-practice, exercise=TRUE, exercise.cap="Explore 1"}
14 + 2000
```

The `[1]` you see at the start of the output line is just R telling you that the first item on that line is item number 1. In this example there is only one number. Later, when output contains many numbers or words, R may start later lines with a different number to show which item begins that line.

## Objects

**R can store information as an [**object**]{.important-text} with a name of your
choice. Once you have created an R object, you can refer to it by name.**

Picking a name for your object:

* Choose an informative name
* Object names cannot start with a number
* Object names cannot contain spaces
* Object names cannot include special characters that have specific meanings in
R. This includes $ and # and %.
* Periods and underscores are permitted and commonly used to separate words ---
for example, `my.first.object` or `my_first_object` are both valid names.

In a regular RStudio session, the Global Environment pane will show a list of
all the objects you have created. This pane is not visible inside these
tutorials, but you will use it constantly when working on your own.

**We use the [**assignment operator**]{.important-text} `<-` to assign some value to
an object.**

**Run and observe 2: Create your first object**

Let's store the sum of 5+3 by assigning it an object called **my.first.object**. Use the
"Run Code" button to execute the following code and see how assignment works.

```{r objects-create-first, exercise=TRUE, exercise.cap="Run and observe 2"}
my.first.object <- 5 + 3
```

Notice that the result is not displayed in this case. We can see the value of an
object only by asking R to show it to us. Simply enter the object name.

**Run and observe 3: Display an object's value**

```{r objects-print-value, exercise=TRUE, exercise.cap="Run and observe 3"}
my.first.object
```

If we assign a different value to the same object name, the value of the object
will change.

**Run and observe 4: Reassign an object**

```{r objects-reassign, exercise=TRUE, exercise.cap="Run and observe 4"}
my.first.object <- 50 + 30
my.first.object
```

The second line displays `80` because the new assignment replaces the old value
stored in **my.first.object**. R does not keep both values under the same object
name; the most recent assignment is the one that matters.

Object names are case sensitive: **my.first.object** is different from
**MY.FIRST.OBJECT**. Try asking R to tell you the value of **MY.FIRST.OBJECT**.
What happens?

**Run and observe 5: Test case sensitivity**

```{r objects-case-error, exercise=TRUE, exercise.cap="Run and observe 5"}
MY.FIRST.OBJECT
```

The error says that the object **MY.FIRST.OBJECT** was not found. R is telling
you exactly what went wrong: it cannot find an object with that exact spelling
and capitalization. In a regular RStudio session you could confirm this by
looking at the Global Environment pane, where you would see that
**MY.FIRST.OBJECT** does not exist.

### Storing Character Strings as Objects

To assign a string of characters to an object, we use quotation marks. Spaces are allowed inside the quotation marks. R will treat numbers as characters too if we enclose them in quotes --- watch what the second pair of lines below produces.

**Run and observe 6: Create character objects**

```{r objects-create-string, exercise=TRUE, exercise.cap="Run and observe 6"}
this.country <- "Costa Rica"
this.country
another.object <- "100"
another.object
```

If you refer to an object that is a character string in an arithmetic operation,
you will see an error message. See what happens if you try to divide "100" by
10:

**Run and observe 7: Arithmetic error with character strings**

```{r objects-string-arithmetic-error, exercise=TRUE, exercise.cap="Run and observe 7"}
another.object / 10
```

This error message is not particularly intuitive. It means R cannot apply
arithmetic to a character object. Even though `"100"` looks like a number to us,
the quotation marks tell R to treat it as text.

---

## Functions

Functions are very important in R. They can be used to perform a variety of
tasks. **A [**function**]{.important-text} carries out a task and returns a result;
most functions use one or more inputs.** For example, the `sqrt()` function takes as its input some number or
numeric object for which we want to know the square root. The output is the
square root of that number or object.

An [**argument**]{.important-text} is the information you put inside the
parentheses of a function so R knows what to do. In R, the syntax is generally
`function_name(input)`, where `function_name` is the name of the function and
`input` is the argument you are giving the function. Some functions take no
arguments, some take one argument, and others take several arguments. When there
is more than one argument, they are separated by a comma. Omitting a required
comma or having unbalanced parentheses are common sources of error.

The function `sqrt()` takes one argument, named `x`, which is the number or
object for which we want the square root.

**Practice 8: Calculate a square root**

Enter code that calculates the square root of 4 using the `sqrt()` function.
Two hints are available. Click "Submit Answer" to check your work.

```{r functions-square-root, exercise=TRUE, exercise.cap="Practice 8"}

```

```{r functions-square-root-hint-1}
# Hint 1 of 2
# Functions follow the pattern: function_name(input)
# The function you need has a name that tells you exactly what it does.
# What function was introduced in the text just above this exercise?
```

```{r functions-square-root-hint-2}
# Hint 2 of 2
# Fill in the blank:
___(4)
```

```{r functions-square-root-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("sqrt\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the sqrt() function.")
  }
  if (!grepl("sqrt\\s*\\(\\s*4\\s*\\)", code_no_comments)) {
    fail("Use sqrt() to calculate the square root of 4.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be one number. Check that you have only one number inside the parentheses.")
  }
  if (!isTRUE(all.equal(.result, 2))) {
    fail("That is not the square root of 4. Check the number inside the parentheses.")
  }
  pass("Correct! The square root of 4 is 2.")
})
```

To see what arguments a function accepts, use the `args()` function. Type the name of the function you want to learn about inside its parentheses, without adding parentheses after that name. Here we use `args()` to examine the `round()` function, which rounds a number to a
specified number of decimal places.

**Run and observe 9: Inspect function arguments**

```{r functions-explore-args, exercise=TRUE, exercise.cap="Run and observe 9"}
args(round)
```

The output shows four things worth knowing:

- `x` is the required argument --- the number or object you want to round.
- `digits = 0` is an optional argument with a *default value* of 0, meaning
that if you do not specify it, R will round to zero decimal places. You only
need to include it if you want a different number of decimal places.
- `...` means the function can accept additional arguments that get passed along
to other functions internally. As a beginner you will see this often and can
safely ignore it.
- The `NULL` at the bottom can also be ignored --- it is just an artifact of how
`args()` displays its output and does not mean anything is wrong.

We will return to default values and how to use them when we need them later
in this tutorial.

---

Anytime you want to learn more about a function --- what it does, what its
arguments mean, and what it returns --- type `?` followed immediately by the
function name with no space and no parentheses. Get in the habit of checking the
help page for new functions you encounter.

**Run and observe 10: Access function help**

Run the code below to pull up help for the `round()` function.

```{r functions-get-help, exercise=TRUE, exercise.cap="Run and observe 10"}
?round
```

::: {.tip}
**Help looks different here than in RStudio.** Inside this tutorial `?` produces
plain text output that can be hard to read. When you use `?` on your own machine
in RStudio, it opens a cleanly formatted Help pane. The most useful part is the
**Arguments** section, which describes what each argument does. Start with
**Usage**, **Arguments**, and **Examples**; you can skim the rest for now.
:::

If you do not know the name of a function, search for it online --- for example,
"R function to calculate standard deviation" or "R round number function."

If you receive an error such as "No documentation for 'funcname'," the function
may belong to a package that is not loaded. In that case you can try `??funcname`
to search more broadly, or install and load the relevant package first.

## Classes of Objects

Every object in R has a [**class**]{.important-text} --- a label that tells R what
kind of value it contains and what operations it can perform on it. Numbers and
text, for example, are treated very differently.

You can check the class of any object using the `class()` function: type the object's name inside the parentheses. In R, giving a function its input this way is called *passing* an argument to the function --- you will see that term throughout these tutorials. For **my.first.object** and **another.object**, this is straightforward. For `sqrt`, there is one trap to watch out for: because you are used to calling `sqrt` by writing `sqrt()`, it is tempting to write `class(sqrt())` --- but that tells R to run `sqrt` first and then check the class of its output, which is not what you want. Write `class(sqrt)` instead, with no parentheses after `sqrt`, to ask about the function itself.

**Practice 11: Determine object classes**

Use the `class()` function to find the class of each of the following. Write one line of code for each --- you can enter them in any order.

* **my.first.object**
* **another.object**
* `sqrt`

Two hints are available. Click "Submit Answer" to check your work.

```{r prepare-classes, echo=FALSE}
my.first.object <- 50 + 30
another.object <- "100"
```

```{r classes-determine-types, exercise=TRUE, exercise.setup="prepare-classes", exercise.cap="Practice 11"}

```

```{r classes-determine-types-hint-1}
# Hint 1 of 2
# The class() function works the same way for all three.
# For sqrt, check the function itself --- no parentheses after sqrt.
```

```{r classes-determine-types-hint-2}
# Hint 2 of 2: fill in the blanks
class(___)
class(___)
class(___)
```

```{r classes-determine-types-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  has_first  <- grepl("class\\s*\\(\\s*my\\.first\\.object\\s*\\)", code_no_comments)
  has_second <- grepl("class\\s*\\(\\s*another\\.object\\s*\\)", code_no_comments)
  has_third  <- grepl("class\\s*\\(\\s*sqrt\\s*\\)", code_no_comments)

  if (!has_first) {
    fail("Make sure you are checking the class of my.first.object using class().")
  }
  if (!has_second) {
    fail("Make sure you are checking the class of another.object using class().")
  }
  if (!has_third) {
    fail("Make sure you are checking the class of sqrt --- no parentheses after sqrt.")
  }
  pass("Exactly right! my.first.object is numeric, another.object is character, and sqrt is a function. These are the three classes introduced here --- you will meet several others as the series continues, including logical values and data frames.")
})
```

There are many other classes of objects in R. We will encounter more in future lessons.

## Vectors

Next, we will enter some data into R as a *numeric vector object*. A [**vector**]{.important-text} stores multiple values together in a single object, in a specific order. For example, a variable in a data set is usually stored as a vector: one value for each case in the data set. In this section, we will work with numeric vectors, which contain numbers.

To create a vector in R, we use the `c()` function (short for "concatenate") and place multiple values inside the parentheses, separated by commas. For example, `c(1, 2, 3)` creates a numeric vector with three values. Simply using parentheses alone (like `(1, 2, 3)`) will **not** create a vector --- `c()` is required.

We then assign the result to an object using `<-` so we can work with it later.

**When you want to store several values of the same kind together, a vector is usually the right starting point.** This is one of the small building blocks that will show up again and again, so it is worth getting comfortable with it now.

**Practice 12: Create your first vector**

The code below is partially written for you. Replace the `___` inside `c()` with 5 numbers of your choice, separated by commas. (Remember: Including the object name on the second line displays the results on the screen.) Two hints are available. Click "Submit Answer" to check your work.

```{r vectors-create-first, exercise=TRUE, exercise.cap="Practice 12"}
my.first.vector <- c(___)
my.first.vector
```

```{r vectors-create-first-hint-1}
# Hint 1 of 2
# Replace the ___ with five numbers, separating
# adjacent values with commas. Which five numbers
# you choose is entirely up to you.
```

```{r vectors-create-first-hint-2}
# Hint 2 of 2: fill in the blank
my.first.vector <- c(___, ___, ___, ___, ___)
my.first.vector
```

```{r vectors-create-first-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("my\\.first\\.vector\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign your vector to an object called my.first.vector.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (!grepl("my\\.first\\.vector\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display my.first.vector by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display my.first.vector by typing its name on the last line so the grader can check it.")
  }
  if (!is.numeric(.result)) {
    fail("Your vector should contain numbers, not character strings.")
  }
  if (length(.result) != 5) {
    fail("Your vector should contain exactly 5 numbers.")
  }
  pass("Good start! Your vector has five elements, each accessible by index position.")
})
```

Now let's try something a bit more interesting: using a vector to store real political data.

Here are the two-party vote shares won by the Democratic presidential candidate in the last 12 presidential elections:

<div class="table-container">
  <table role="table" aria-label="Presidential Election Results from 1980 to 2024 showing Democratic vote shares and candidates">
    <caption>
      <strong>Presidential Election Results 1980-2024</strong><br>
      This table displays the two-party vote share won by Democratic presidential candidates, along with the names of both Democratic and Republican candidates for each election year.
    </caption>
    <thead>
      <tr>
        <th scope="col" id="year">Election Year</th>
        <th scope="col" id="dem-share">Democratic 2-Party Vote Share</th>
        <th scope="col" id="dem-candidate">Democratic Candidate</th>
        <th scope="col" id="rep-candidate">Republican Candidate</th>
      </tr>
    </thead>
    <tbody>
      <tr><th scope="row" headers="year">1980</th><td headers="dem-share">0.448</td><td headers="dem-candidate">Jimmy Carter</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1984</th><td headers="dem-share">0.408</td><td headers="dem-candidate">Walter Mondale</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1988</th><td headers="dem-share">0.461</td><td headers="dem-candidate">Michael Dukakis</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1992</th><td headers="dem-share">0.535</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1996</th><td headers="dem-share">0.547</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">Bob Dole</td></tr>
      <tr><th scope="row" headers="year">2000</th><td headers="dem-share">0.503</td><td headers="dem-candidate">Al Gore</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2004</th><td headers="dem-share">0.488</td><td headers="dem-candidate">John Kerry</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2008</th><td headers="dem-share">0.537</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">John McCain</td></tr>
      <tr><th scope="row" headers="year">2012</th><td headers="dem-share">0.520</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">Mitt Romney</td></tr>
      <tr><th scope="row" headers="year">2016</th><td headers="dem-share">0.511</td><td headers="dem-candidate">Hillary Clinton</td><td headers="rep-candidate">Donald Trump</td></tr>
      <tr><th scope="row" headers="year">2020</th><td headers="dem-share">0.513</td><td headers="dem-candidate">Joe Biden</td><td headers="rep-candidate">Donald Trump</td></tr>
      <tr><th scope="row" headers="year">2024</th><td headers="dem-share">0.493</td><td headers="dem-candidate">Kamala Harris</td><td headers="rep-candidate">Donald Trump</td></tr>
    </tbody>
  </table>
</div>

We can enter the share of the two-party vote won by the Democratic presidential candidates in the last 12 elections as elements of a single vector we will name **dem.pres.vote.share**.

**Practice 13: Create a data vector**

The code below gives you the structure. Replace `___(___)` with the correct function call containing all 12 vote share values from the table above, in chronological order. Listing the object name in the second line displays the vector on the screen so you can check your work. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-election-data, exercise=TRUE, exercise.cap="Practice 13"}
dem.pres.vote.share <- ___(___)
dem.pres.vote.share
```

```{r vectors-election-data-hint-1}
# Hint 1 of 3
# You need to store a collection of 12 numbers together
# in a single object. What have you learned about how R
# stores multiple values in one place?
```

```{r vectors-election-data-hint-2}
# Hint 2 of 3
# In the scaffold, the outer ___ is the function name
# and the inner ___ is where your values go.
# What function combines multiple values into a vector?
# How do you separate values inside that function?
```

```{r vectors-election-data-hint-3}
# Hint 3 of 3: fill in the missing values
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535,
                         0.547, ___, ___, ___,
                         ___, ___, ___, ___)
dem.pres.vote.share
```

```{r vectors-election-data-check}
grade_this({
  correct_vals <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                    0.537, 0.520, 0.511, 0.513, 0.493)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("dem\\.pres\\.vote\\.share\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.pres.vote.share.")
  }

  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display dem.pres.vote.share by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display dem.pres.vote.share by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should contain 12 values, one per election from 1980 to 2024.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check that each vote share is typed as a number.")
  }

  wrong_pos <- which(abs(.result - correct_vals) > 0.001)
  if (length(wrong_pos) > 0) {
    fail(paste0("Check your value at position ", wrong_pos[1],
                " --- that corresponds to the ", 1979 + wrong_pos[1] * 4,
                " election."))
  }

  pass("Exactly right! Your vector captures the Democratic two-party vote share across all 12 elections, from Carter's 44.8% in 1980 to Harris's 49.3% in 2024. You will use this vector in the exercises that follow.")
})
```

The `c()` function can also be used to combine multiple vectors into a single larger one.

**Practice 14: Create the first group of candidate names**

Using the Democratic candidate column in the table above, create a character vector called **first.6.candidates** containing the last names of the Democratic candidates for the first 6 elections (1980--2000). Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-first-candidates, exercise=TRUE, exercise.cap="Practice 14"}

```

```{r vectors-first-candidates-hint-1}
# Hint 1 of 3
# You need to store 6 text values in a single object.
# Think about what you have learned about how R stores
# text, and how you combine multiple values into one object.
```

```{r vectors-first-candidates-hint-2}
# Hint 2 of 3
# Character strings need quotation marks around each name.
# Use c() to combine them, separated by commas.
# Assign the result to first.6.candidates and display it by typing its name.
```

```{r vectors-first-candidates-hint-3}
# Hint 3 of 3: fill in the blanks
first.6.candidates <- c("Carter", "___", "___",
                        "___", "___", "___")
first.6.candidates
```

```{r vectors-first-candidates-check}
grade_this({
  correct_vals <- c("Carter", "Mondale", "Dukakis",
                    "Clinton", "Clinton", "Gore")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (!grepl("first\\.6\\.candidates\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called first.6.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (!grepl("first\\.6\\.candidates\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display first.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display first.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your vector should contain character strings. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 6) {
    fail("Your vector should contain exactly 6 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The names do not match. Check spelling and order against the table.")
  }
  pass("Correct! Notice that Clinton appears twice --- Bill Clinton ran in both 1992 and 1996. You will combine this vector with the second group in the next exercise.")
})
```

**Practice 15: Create the second group of candidate names**

Now create a character vector called **last.6.candidates** containing the last names of the Democratic candidates for the final 6 elections (2004--2024). Display it by typing its name. Two hints are available. Click "Submit Answer" to check your work.

```{r vectors-last-candidates, exercise=TRUE, exercise.cap="Practice 15"}

```

```{r vectors-last-candidates-hint-1}
# Hint 1 of 2
# Use the same approach as the previous exercise.
# Look at the Democratic candidate column in the table
# for elections from 2004 to 2024.
```

```{r vectors-last-candidates-hint-2}
# Hint 2 of 2: fill in the blanks
last.6.candidates <- c("Kerry", "___", "___",
                       "___", "___", "___")
last.6.candidates
```

```{r vectors-last-candidates-check}
grade_this({
  correct_vals <- c("Kerry", "Obama", "Obama",
                    "Clinton", "Biden", "Harris")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (!grepl("last\\.6\\.candidates\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called last.6.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to create your vector.")
  }
  if (!grepl("last\\.6\\.candidates\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display last.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display last.6.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your vector should contain character strings. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 6) {
    fail("Your vector should contain exactly 6 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The names do not match. Check spelling and order against the table.")
  }
  pass("Right! Notice that Obama appears twice and Clinton appears once --- you will see repeated names when combining these vectors.")
})
```

**Practice 16: Combine the two vectors**

Now we can put the two pieces together. Combine **first.6.candidates** and **last.6.candidates** into a single vector called **dem.candidates** and display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r prepare-vectors-first-last, echo=FALSE}
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
```

```{r vectors-combine-names, exercise=TRUE, exercise.setup="prepare-vectors-first-last", exercise.cap="Practice 16"}

```

```{r vectors-combine-names-hint-1}
# Hint 1 of 3
# You have two separate vectors and need to join them
# into one longer vector. Think about what function you
# have used to combine multiple values into a single
# object --- does the same approach work when the inputs
# are whole vectors rather than individual values?
```

```{r vectors-combine-names-hint-2}
# Hint 2 of 3
# The c() function works on whole vectors the same way
# it works on individual values. Pass both vector names
# as arguments, separated by a comma. Assign the result
# to dem.candidates and display it by typing its name.
```

```{r vectors-combine-names-hint-3}
# Hint 3 of 3: fill in the blanks
dem.candidates <- c(___, ___)
dem.candidates
```

```{r vectors-combine-names-check}
grade_this({
  correct_vals <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore",
                    "Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (!grepl("dem\\.candidates\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.candidates.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to combine the two vectors.")
  }
  if (!grepl("dem\\.candidates\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your result should be a character vector.")
  }
  if (length(.result) != 12) {
    fail("Your combined vector should have 12 names.")
  }
  if (!identical(.result, correct_vals)) {
    fail("The combined vector does not look right. Make sure you are combining first.6.candidates and last.6.candidates in that order.")
  }
  pass("Exactly! The combined vector has all 12 candidates in chronological order, ready to use alongside the vote share data.")
})
```

```{r prepare-vectors-election-data, echo=FALSE}
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
```

```{r prepare-vectors-combine-names, echo=FALSE}
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
```

## Indexing Vectors

Once you have values stored in a vector, the next useful skill is learning how to pull out the pieces you need. **To select a specific element in a vector, you use square brackets `[ ]`.** The
number inside the brackets is called the [**index**]{.important-text} --- it tells
R which position in the vector you want. Position 1 is the first element,
position 2 is the second, and so on. Here is a quick reference showing the index
position of each value in **dem.pres.vote.share**:

```{r vectors-position-table, echo=FALSE, results='asis'}
vals <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503,
          0.488, 0.537, 0.520, 0.511, 0.513, 0.493)

make_pos_table(1:6, vals[1:6],
  "Index positions 1 through 6 of dem.pres.vote.share")
make_pos_table(7:12, vals[7:12],
  "Index positions 7 through 12 of dem.pres.vote.share")
```

To extract an element, place its index number inside square brackets after the vector name: `dem.pres.vote.share[3]` returns the third value, 0.461.

**Practice 17: Access a single vector element**

The code below extracts the first element. Modify it to extract Hillary Clinton's vote share instead. Two hints are available. Click "Submit Answer" to check your work.

```{r vectors-access-single, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice 17"}
dem.pres.vote.share[1]
```

```{r vectors-access-single-hint-1}
# Hint 1 of 2
# Use the position table above to find which index
# position corresponds to the 2016 election.
# Then replace the 1 in the brackets with that number.
```

```{r vectors-access-single-hint-2}
# Hint 2 of 2: fill in the blank
dem.pres.vote.share[___]
```

```{r vectors-access-single-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("\\[\\s*10\\s*\\]", code_no_comments)) {
    fail("Not quite. Use the position table above to find which index position corresponds to the 2016 election.")
  }

  if (is.null(.result)) {
    fail("Make sure you display the value you extracted so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("You should be extracting a numeric value.")
  }

  if (length(.result) != 1) {
    fail("You should be extracting a single value.")
  }

  if (is.na(.result)) {
    fail("Your result is NA. Check the index position inside the square brackets.")
  }

  if (!isTRUE(all.equal(.result, 0.511, tolerance = 0.001))) {
    fail("That is not quite right. Use the position table to find which index corresponds to 2016.")
  }

  pass("Correct! Hillary Clinton received 51.1% of the two-party vote in 2016.")
})
```

We can extract more than one element by placing multiple index positions inside the square brackets. Because R expects a single object inside the brackets, we combine those positions using the `c()` function: `vectorname[c(2, 5, 7)]` extracts the elements at positions 2, 5, and 7 in that order.

**Practice 18: Access multiple vector elements**

What was the vote share won by the Democrat in the last 5 elections? Use the position table above to find the correct index positions. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-access-multiple, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice 18"}

```

```{r vectors-access-multiple-hint-1}
# Hint 1 of 3
# You need to extract 5 values at once.
# How do you tell R to look at more than one
# position at the same time?
```

```{r vectors-access-multiple-hint-2}
# Hint 2 of 3
# Use c() inside square brackets to specify
# multiple positions. Use the position table
# above to find which index numbers you need.
```

```{r vectors-access-multiple-hint-3}
# Hint 3 of 3: fill in the blanks
dem.pres.vote.share[c(___, ___, ___, ___, ___)]
```

```{r vectors-access-multiple-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  expected <- c(0.537, 0.520, 0.511, 0.513, 0.493)
  if (!grepl("\\[", code_no_comments) || !grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using square brackets with c() inside to select multiple elements.")
  }
  if (is.null(.result)) {
    fail("Make sure you display the extracted values so the grader can check them.")
  }
  if (!is.numeric(.result) || length(.result) != 5) {
    fail("Your result should be a numeric vector with exactly 5 values.")
  }
  if (!identical(.result, expected)) {
    fail("The values are not quite right or are out of order. Make sure you are listing the last 5 elections in chronological order.")
  }
  pass("Nicely done! The last five elections returned 0.537, 0.520, 0.511, 0.513, and 0.493 --- a narrow range showing how competitive recent presidential elections have been.")
})
```

The brackets can do two more things. First, the order in which you list index numbers determines the order of the output, so `dem.pres.vote.share[c(9, 5)]` and `dem.pres.vote.share[c(5, 9)]` return the same two values in opposite order. Second, a minus sign removes an element rather than selecting it --- `dem.pres.vote.share[-12]` returns every value except the last election.

**Run and observe 19: Index order and negative indexing**

```{r vectors-order-matters, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Run and observe 19"}
dem.pres.vote.share[c(9, 5)]
dem.pres.vote.share[c(5, 9)]
dem.pres.vote.share[-12]
```

## Vector Arithmetic

We can apply arithmetic operations on a numeric vector. *The operation will be repeated for each element of the vector.* We will often want to do this as part of creating new variables that are a function of variables in a data set.

**Practice 20: Convert vote shares to percentages**

Create a new vector called **dem.pres.vote.percent** that converts **dem.pres.vote.share** to the percent of the 2-party vote won by the Democratic candidate. Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-convert-percent, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice 20"}

```

```{r vectors-convert-percent-hint-1}
# Hint 1 of 3
# A proportion and a percentage express the same value
# in different units. What single arithmetic operation
# converts a proportion to a percentage?
```

```{r vectors-convert-percent-hint-2}
# Hint 2 of 3
# Apply that operation to the entire dem.pres.vote.share
# vector. R will apply it to every element automatically.
# Assign the result to dem.pres.vote.percent and display it by typing its name.
```

```{r vectors-convert-percent-hint-3}
# Hint 3 of 3: fill in the blanks
dem.pres.vote.percent <- dem.pres.vote.share * ___
dem.pres.vote.percent
```

```{r vectors-convert-percent-check}
grade_this({
  expected <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                0.537, 0.520, 0.511, 0.513, 0.493) * 100
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (!grepl("dem\\.pres\\.vote\\.percent\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called dem.pres.vote.percent.")
  }
  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are applying the operation to dem.pres.vote.share.")
  }
  if (!grepl("\\*\\s*100|100\\s*\\*", code_no_comments)) {
    fail("Think about what arithmetic operation converts a proportion to a percentage.")
  }
  if (!grepl("dem\\.pres\\.vote\\.percent\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display dem.pres.vote.percent by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result) || !is.numeric(.result)) {
    fail("Make sure you display dem.pres.vote.percent by typing its name on the last line so the grader can check it.")
  }
  if (length(.result) != 12) {
    fail("Your result should have 12 values.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.01))) {
    fail("Your values do not look right. Check that you are multiplying by 100.")
  }
  pass("Exactly right! The values now range from about 40.8 to 54.7 --- the same information expressed as percentages rather than proportions.")
})
```

Vector arithmetic can also express each value relative to a reference point --- subtracting a single baseline value from every element. This is useful any time you want to know not just what a value is, but how much it differs from a particular starting point, such as the first year in a series, a control group mean, or a historical benchmark.

**Practice 21: Calculate differences from baseline**

Create a vector called **compared.to.1980** showing how each election's Democratic vote share compares to 1980. Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-compare-baseline, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Practice 21"}

```

```{r vectors-compare-baseline-hint-1}
# Hint 1 of 3
# "Compared to 1980" means each value minus the 1980
# value. What do you know about how R applies arithmetic
# operations across an entire vector?
```

```{r vectors-compare-baseline-hint-2}
# Hint 2 of 3
# You need to subtract a single value --- the 1980
# vote share --- from every element. You know how to
# extract a single element from a vector by its index.
# What index position holds the 1980 value?
```

```{r vectors-compare-baseline-hint-3}
# Hint 3 of 3: fill in the blanks
compared.to.1980 <- dem.pres.vote.share - dem.pres.vote.share[___]
compared.to.1980
```

```{r vectors-compare-baseline-check}
grade_this({
  baseline <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- baseline - baseline[1]

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("compared\\.to\\.1980\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called compared.to.1980.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using dem.pres.vote.share in your calculation.")
  }

  if (grepl("\\[\\s*1980\\s*\\]", code_no_comments)) {
    fail("Use the position of the 1980 value, not the year itself. The 1980 election is the first value in dem.pres.vote.share, so its index position is 1.")
  }

  if (!grepl("\\[\\s*1\\s*\\]", code_no_comments)) {
    fail("Use indexing to extract the 1980 value. It is at position 1 in the vector.")
  }

  if (!grepl("compared\\.to\\.1980\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display compared.to.1980 by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display compared.to.1980 by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your result should have 12 values, one for each election.")
  }

  if (!isTRUE(all.equal(.result[1], 0, tolerance = 0.001))) {
    fail("The first value should be 0 since you are comparing 1980 to itself.")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Make sure you are subtracting the 1980 value from every element.")
  }

  pass("Correct! Positive values mean the candidate did better than Carter in 1980; negative values mean worse.")
})
```

## Modifying Vectors

We can also use indexing to change selected values in a vector. Instead of using square brackets only to look at certain positions, we can use them together with the assignment operator to replace the values stored in those positions. The general pattern is `vector_name[position] <- new_value`. When we want to replace more than one value at once, we can put several positions inside `c()` and provide the same number of replacement values.

Below is a position reference table for **dem.candidates**.

```{r vectors-candidate-position-table, echo=FALSE, results='asis'}
vals <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore",
          "Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")

make_pos_table(1:6, vals[1:6],
  "Index positions 1 through 6 of dem.candidates")
make_pos_table(7:12, vals[7:12],
  "Index positions 7 through 12 of dem.candidates")
```

**Practice 22: Replace vector elements**

Replace the three `"Clinton"` entries in **dem.candidates** with their full names: `"Bill Clinton"` for the two earlier Clinton entries and `"Hillary Clinton"` for the later Clinton entry. Use the position reference table above to find the correct index positions. Display the result by typing its name to check your work. Three hints are available. Click "Submit Answer" to check your work.

```{r vectors-replace-elements, exercise=TRUE, exercise.setup="prepare-vectors-combine-names", exercise.cap="Practice 22"}

```

```{r vectors-replace-elements-hint-1}
# Hint 1 of 3
# You need to find which positions contain "Clinton"
# and then overwrite those positions with new values.
# Print dem.candidates first to see its contents,
# then use the position table to find the right indices.
```

```{r vectors-replace-elements-hint-2}
# Hint 2 of 3
# You can replace multiple positions at once using
# the same c() indexing you used to select elements.
# The replacement values must be in the same order
# as the positions you are replacing.
```

```{r vectors-replace-elements-hint-3}
# Hint 3 of 3: fill in the blanks
dem.candidates[c(___, ___, ___)] <- c("Bill Clinton", "___", "Hillary Clinton")
dem.candidates
```

```{r vectors-replace-elements-check}
grade_this({
  expected <- c("Carter", "Mondale", "Dukakis", "Bill Clinton",
                "Bill Clinton", "Gore", "Kerry", "Obama", "Obama",
                "Hillary Clinton", "Biden", "Harris")

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("dem\\.candidates\\s*\\[", code_no_comments)) {
    fail("Use square bracket indexing on dem.candidates to replace specific positions, like dem.candidates[...] <- ...")
  }

  if (!grepl("<-", code_no_comments)) {
    fail("Use the assignment operator <- to replace values at the indexed positions.")
  }

  if (grepl("<-\\s*dem\\.candidates\\s*\\[", code_no_comments)) {
    fail("Look at the right side of your assignment. It should be the replacement values themselves, not another expression that indexes dem.candidates.")
  }

  if (!grepl("dem\\.candidates\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display dem.candidates by typing its name on the last line so the grader can check it.")
  }

  if (!is.character(.result)) {
    fail("Your result should be a character vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should still have 12 elements. You are replacing values, not adding or removing them.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check the replacement values you typed.")
  }

  if (any(.result == "Clinton", na.rm = TRUE)) {
    fail("You still have entries that say only 'Clinton'. Make sure you replaced all three with full names.")
  }

  if (!identical(.result, expected)) {
    fail("Your vector does not match the expected result. Check the full names and their index positions.")
  }

  pass("Nice work! Using full names avoids ambiguity --- your vector now distinguishes Bill Clinton (1992, 1996) from Hillary Clinton (2016).")
})
```

Replacing values at particular positions is one of the most common things you will do with real data. A country's name is spelled two different ways across years, a survey response was coded wrong, a state abbreviation is missing for a single case --- each of these comes down to locating the position and assigning a new value there. In Tutorial 2 you will do the same thing inside a data frame, where the positions correspond to particular observations.

## Summary Functions

Now that you have worked with vectors, you are ready for a few functions that summarize them. These functions are simple to use, and they will become part of your regular R toolkit.

* `length()`: returns the number of elements in the vector.
* `min()`: returns the minimum value.
* `max()`: returns the maximum value.
* `range()`: returns a two-element vector containing the minimum and maximum values --- not a single "range" number, but both endpoints together.
* `mean()`: returns the arithmetic mean.
* `sum()`: returns the sum of all values.
* `abs()`: returns the absolute value of a number or each element of a numeric vector, removing any negative sign.

Many summary functions include an argument called `na.rm`, which stands for "remove missing values" and has a default value of `FALSE`. When `na.rm = FALSE`, missing values are left in place, and functions such as `mean()` will usually return `NA` if any value is missing. The `length()` function is different: it simply counts elements, including any `NA` values.

**Run and observe 23: Use summary functions**

This exercise produces several lines of output, so read them one at a time:

```{r functions-summary-stats, exercise=TRUE, exercise.setup="prepare-vectors-combine-names", exercise.cap="Run and observe 23"}
length(dem.candidates)
min(dem.pres.vote.share)
max(dem.pres.vote.share)
range(dem.pres.vote.share)
mean(dem.pres.vote.share)
sum(dem.pres.vote.share) / length(dem.pres.vote.share)
abs(-0.055)
```

The last two expressions are worth noting. The second-to-last shows that the mean can also be calculated by hand as the sum of all values divided by the number of values --- it should match the `mean()` output just above it. The final expression shows that `abs()` removes the negative sign: `-0.055` and `0.055` both return `0.055`. This is useful any time you need a difference but do not care about direction, such as a margin of victory. Notice also that `range()` returns two numbers --- the minimum and maximum --- not a single value.

### Default Values and Named Arguments

Every argument either has a [**default value**]{.important-text} or is
**required**. When an argument has a default, you only need to specify it if you
want something different from the default behavior. When an argument has no
default, you must supply a value or R will return an error.

You saw this earlier when you ran `args(round)`. The output showed two arguments: `x`, which has no default and is required, and `digits`, which has a default of `0`. This means `round(3.14159)` works fine --- R rounds to zero decimal places by default. But `round(digits = 2)` would fail because `x` is required and you have not supplied it.

You can supply arguments in two ways. The first is **positional** --- R matches your inputs to arguments in the order they are listed in `args()`. The second is **by name** --- you explicitly write `argument = value`, which allows you to supply arguments in any order. For all arguments after the first, it is good practice to use names, both to avoid mistakes and to make your code easier to read.

**Run and observe 24: Default values and named arguments**

```{r functions-named-args-demo, exercise=TRUE, exercise.cap="Run and observe 24"}
round(3.14159)
round(3.14159, digits = 2)
round(digits = 2, x = 3.14159)
```

All three calls produce sensible results. The first uses the default for `digits`. The second overrides it. The third supplies arguments by name in reverse order --- R still knows what to do because the argument names make the intent unambiguous.

## Missing Values

Real data often contains [**missing values**]{.important-text}, which R
represents as `NA`. Values go missing for all sorts of reasons: a survey respondent declines to report their income, a country does not submit a statistic to the agency that compiles it, a measure was not collected in the earlier years of a series, or a record was simply lost along the way. When a vector contains `NA`, most summary functions return
`NA` by default rather than attempting to compute a result from incomplete data.

**Run and observe 25: Missing values**

Watch what happens when a vector contains a missing value, and how `na.rm = TRUE` changes the result:

```{r functions-missing-values, exercise=TRUE, exercise.cap="Run and observe 25"}
vote.with.missing <- c(0.448, 0.408, NA, 0.535, 0.547)
mean(vote.with.missing)
mean(vote.with.missing, na.rm = TRUE)
```

The first call returns `NA` because R cannot compute a mean when some values are unknown. The second call returns the mean of the four non-missing values by telling R to remove the `NA` before computing. The argument `na.rm = TRUE` overrides the default of `FALSE`.

None of the vectors created here have missing values, so the default behavior is fine for the exercises in this tutorial. But in real data you will encounter `NA` regularly, and `na.rm = TRUE` is how you handle it.

## Functions with Multiple Inputs

As we just saw, functions can take more than one argument. The `seq()` function is a good example --- it generates a sequence of numbers using three arguments: `from` (where to start), `to` (where to end), and `by` (the step size).

**Run and observe 26: Create a number sequence**

```{r functions-create-sequence, exercise=TRUE, exercise.cap="Run and observe 26"}
my.sequence <- seq(from = 1, to = 10, by = 2)
my.sequence
```

Notice that all three arguments are named explicitly. This is good practice with multi-argument functions --- it makes the code easy to read and protects against mistakes from argument order.

**Practice 27: Create election year sequence**

Using `seq()`, create a vector called **elect.year** containing every presidential election year from 1980 to 2024. Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r functions-election-years, exercise=TRUE, exercise.cap="Practice 27"}

```

```{r functions-election-years-hint-1}
# Hint 1 of 3
# Think about the three things seq() needs to know:
# where to start, where to end, and how much to
# increase by each step. How often are elections held?
```

```{r functions-election-years-hint-2}
# Hint 2 of 3
# Presidential elections occur every 4 years.
# The sequence starts at 1980 and ends at 2024.
# Use named arguments for all three inputs.
```

```{r functions-election-years-hint-3}
# Hint 3 of 3: fill in the blanks
elect.year <- seq(from = ___, to = ___, by = ___)
elect.year
```

```{r functions-election-years-check}
grade_this({
  expected <- seq(from = 1980, to = 2024, by = 4)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("elect\\.year\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called elect.year.")
  }

  if (!grepl("seq\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the seq() function.")
  }

  if (grepl("by\\s*=\\s*2(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("seq\\s*\\([^)]*,\\s*[^)]*,\\s*2\\s*\\)", code_no_comments)) {
    fail("Check your step size --- how many years pass between one presidential election and the next?")
  }

  if (grepl("from\\s*=\\s*1(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("to\\s*=\\s*12(\\s*,|\\s*\\)|$)", code_no_comments) ||
      grepl("seq\\s*\\(\\s*1\\s*,\\s*12", code_no_comments)) {
    fail("Use the election years themselves, not the position numbers. The sequence should start at 1980 and end at 2024.")
  }

  if (!grepl("1980", code_no_comments)) {
    fail("Check your starting value --- what year did this series of elections begin?")
  }

  if (!grepl("2024", code_no_comments)) {
    fail("Check your ending value --- what is the most recent election year in the table?")
  }

  if (!grepl("elect\\.year\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display elect.year by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display elect.year by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != length(expected)) {
    fail("Your sequence has the wrong number of values. Check the interval you used between elections.")
  }

  if (!isTRUE(all.equal(.result, expected))) {
    fail("Your sequence does not match. Check your starting value, ending value, and step size.")
  }

  pass("Good work! Your sequence covers all presidential election years from 1980 to 2024.")
})
```

Note that decreasing sequences work the same way with a negative `by` value.
This is useful when the order itself matters, such as listing years from most
recent to oldest.

**Run and observe 28: Create a decreasing sequence**

```{r functions-decreasing-sequence, exercise=TRUE, exercise.cap="Run and observe 28"}
seq(from = 2024, to = 1980, by = -4)
```

## Packages

Base R contains many functions, but it does not contain everything we might want to use. Much of R's power comes from [**packages**]{.important-text}, which are add-ons that bundle together code, data, and help files. You can think of a package like an app on your phone: installing it puts it on your device, but opening it is what makes it available to use. In R, `install.packages()` installs a package, and `library()` loads it for the current R session.

There are tens of thousands of packages written for R, many of them contributed to the Comprehensive R Archive Network (CRAN). R does not automatically install or load all of them, because that would take unnecessary space and memory. Instead, you install the packages you need and load them when you want to use them.

For example, we will use the `ggplot()` function from the [ggplot2]{.package-name} package. You only need to install [ggplot2]{.package-name} once on your computer, but you need to load it with `library(ggplot2)` each time you start a new R session and want to use its functions.

If you try to use a function from a package that is not loaded, R will return an error.

**Run and observe 29: See a package error**

```{r prepare-package-error, echo=FALSE}
if ("package:ggplot2" %in% search()) detach("package:ggplot2")
```

```{r packages-error-demo, exercise=TRUE, exercise.setup="prepare-package-error", exercise.cap="Run and observe 29"}
ggplot()
```

Any time you see an error that begins "could not find function," there are two likely causes: you have misspelled the function name, or you have not loaded the package that contains it.

**Run and observe 30: Load a package and use a function**

```{r packages-load-demo, exercise=TRUE, exercise.cap="Run and observe 30"}
library(ggplot2)
ggplot()
```

You will see a blank gray plotting canvas --- that is normal. The `ggplot()` function creates an empty canvas waiting for data and plot specifications. The important thing is that it runs without error, confirming the package loaded successfully.

To get help with an entire package, use the `help()` function with the `package` argument set to the package name in quotes. The output looks cleaner in RStudio than it does here. Do not try to read every line now; just notice that package help gives you a list of available functions, documentation, and examples.

**Run and observe 31: Get package help**

```{r packages-get-help, exercise=TRUE, exercise.cap="Run and observe 31"}
help(package = "ggplot2")
```

## Comments

You can add comments to your code using the `#` symbol. Anything on a line after `#` is ignored by R and will not be executed. Comments are useful for noting why you wrote a particular piece of code, what a function does, or what arguments it takes.

**Run and observe 32: Use code comments**

```{r comments-demo, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Run and observe 32"}
# Calculate the mean Democratic vote share across all elections
mean(dem.pres.vote.share, na.rm = TRUE)
```

Notice that the comment line is ignored entirely and only the `mean()` call produces output. Since our vector has no missing values, the result is the same whether or not we include `na.rm = TRUE`. When a vector does have missing values you will need this argument --- there is no way to compute a mean otherwise. Just be clear about what the result then describes: the mean among the values that are present, not among all of them.

## Check Your Understanding

The following exercises combine concepts you have learned throughout this tutorial. Take your time and think through each problem step by step. You have seen all of the pieces already; the work now is deciding how to put them together.

**Apply 33: Calculate Republican vote shares**

Create a vector called **rep.pres.vote.share** containing the share of the two-party vote received by the Republican presidential candidate. Display it by typing its name. You do not need to type in the vote shares --- in a two-party race you can calculate them directly from the Democratic vote shares. Three hints are available. Click "Submit Answer" to check your work.

```{r practice-republican-shares, exercise=TRUE, exercise.setup="prepare-vectors-election-data", exercise.cap="Apply 33"}

```

```{r practice-republican-shares-hint-1}
# Hint 1 of 3
# In a two-party race, every vote goes to one party
# or the other. What must be true about how the two
# vote shares relate to each other?
```

```{r practice-republican-shares-hint-2}
# Hint 2 of 3
# The two shares sum to 1. You already have the
# Democratic vote share vector. Express the Republican
# share as an arithmetic operation on that vector ---
# R will apply it to every element automatically.
```

```{r practice-republican-shares-hint-3}
# Hint 3 of 3: fill in the blanks
rep.pres.vote.share <- ___ - dem.pres.vote.share
rep.pres.vote.share
```

```{r practice-republican-shares-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- 1 - dem

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("rep\\.pres\\.vote\\.share\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called rep.pres.vote.share.")
  }

  if (!grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are using dem.pres.vote.share in your calculation rather than typing the values manually.")
  }

  if (!grepl("1\\s*-\\s*dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Your calculation does not have the right form. In a two-party race, what must the two shares add up to?")
  }

  if (!grepl("rep\\.pres\\.vote\\.share\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display rep.pres.vote.share by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display rep.pres.vote.share by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should have 12 values, one for each election.")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Check what the two vote shares must sum to in a two-party race.")
  }

  pass("Well done! Republican candidates averaged about 50.3% of the two-party vote across these 12 elections.")
})
```

**Apply 34: Calculate victory margins**

Create a vector called **victory.margin** that contains the margin of victory for each election --- that is, the absolute difference between the two parties' vote shares. You will need the `abs()` function, which returns the absolute value of a number. Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r practice-setup-for-margins, echo=FALSE}
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
                         0.537, 0.520, 0.511, 0.513, 0.493)
rep.pres.vote.share <- 1 - dem.pres.vote.share
```

```{r practice-victory-margins, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply 34"}

```

```{r practice-victory-margins-hint-1}
# Hint 1 of 3
# A margin of victory is how much the winner beat
# the loser by. You have both vote share vectors.
# Why might the sign of the difference be a problem,
# given that you do not know which party won each year?
```

```{r practice-victory-margins-hint-2}
# Hint 2 of 3
# The difference will be positive in some years and
# negative in others depending on which party won.
# A margin is always positive, so you need a function
# that removes the sign. The prompt names it for you.
```

```{r practice-victory-margins-hint-3}
# Hint 3 of 3: fill in the blanks
victory.margin <- abs(___ - ___)
victory.margin
```

```{r practice-victory-margins-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  rep_share <- 1 - dem
  expected <- abs(rep_share - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("victory\\.margin\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the result to an object called victory.margin.")
  }

  if (!grepl("abs\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the abs() function to ensure all margins are positive.")
  }

  if (!grepl("rep\\.pres\\.vote\\.share", code_no_comments) &&
      !grepl("dem\\.pres\\.vote\\.share", code_no_comments)) {
    fail("Make sure you are calculating the margin from the vote share vectors rather than typing the values manually.")
  }

  if (!grepl("victory\\.margin\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display victory.margin by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display victory.margin by typing its name on the last line so the grader can check it.")
  }

  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }

  if (length(.result) != 12) {
    fail("Your vector should have 12 values.")
  }

  if (any(is.na(.result))) {
    fail("Your vector contains NA. Check that your calculation uses the two vote-share vectors correctly.")
  }

  if (any(.result < 0, na.rm = TRUE)) {
    fail("Victory margins should always be positive. Make sure you are using abs().")
  }

  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("Your values do not look right. Make sure you are taking the absolute value of the difference between the two vote share vectors.")
  }

  pass("That is correct! The closest election in this set was 2000, when Gore and Bush were separated by less than half a percentage point of the two-party vote.")
})
```

**Apply 35: Create Republican candidate names**

Create a character vector called **rep.names** containing the last names of the Republican candidates in order from 1980 to 2024: Reagan, Reagan, Bush, Bush, Dole, Bush, Bush, McCain, Romney, Trump, Trump, Trump. Display it by typing its name. Three hints are available. Click "Submit Answer" to check your work.

```{r practice-republican-names, exercise=TRUE, exercise.cap="Apply 35"}

```

```{r practice-republican-names-hint-1}
# Hint 1 of 3
# You need to store 12 names as a single vector.
# What have you learned about how to store text in R,
# and how to combine multiple values into one object?
```

```{r practice-republican-names-hint-2}
# Hint 2 of 3
# Each name is a character string and needs quotation
# marks. Use c() to combine all 12 into a single vector.
# Assign the result to rep.names and display it by typing its name.
```

```{r practice-republican-names-hint-3}
# Hint 3 of 3: fill in the blanks
rep.names <- c("Reagan", "Reagan", "___", "___",
               "___", "___", "___", "___",
               "___", "___", "___", "___")
rep.names
```

```{r practice-republican-names-check}
grade_this({
  expected <- c("Reagan", "Reagan", "Bush", "Bush", "Dole", "Bush",
                "Bush", "McCain", "Romney", "Trump", "Trump", "Trump")

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("rep\\.names\\s*(<-|=[^=])", code_no_comments)) {
    fail("Make sure you assign the names to an object called rep.names.")
  }
  if (!grepl("c\\s*\\(", code_no_comments)) {
    fail("Make sure you are using c() to combine the names into a vector.")
  }
  if (!grepl("rep\\.names\\s*$", trimws(code_no_comments))) {
    fail("Make sure you display rep.names by typing its name on the last line so the grader can check it.")
  }
  if (is.null(.result)) {
    fail("Make sure you display rep.names by typing its name on the last line so the grader can check it.")
  }
  if (!is.character(.result)) {
    fail("Your result should be a character vector. Make sure each name is in quotation marks.")
  }
  if (length(.result) != 12) {
    fail("Your vector should contain 12 names, one for each election from 1980 to 2024.")
  }
  if (!identical(.result, expected)) {
    fail("Some names do not match. Check spelling and order against the election table.")
  }
  pass("Right! Notice that Trump appears three times --- the only candidate in this set to run in three consecutive elections.")
})
```

**Apply 36: Calculate mean Republican vote share**

Using the object **rep.pres.vote.share** you just created, find the mean share of the two-party vote received by Republican candidates. There are no missing values, so you do not need to specify `na.rm`. Two hints are available. Click "Submit Answer" to check your work.

```{r practice-mean-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply 36"}

```

```{r practice-mean-republican-hint-1}
# Hint 1 of 2
# You need a single number summarizing the typical
# value across all elements of a vector. Which function
# from the "Summary Functions" section does this?
```

```{r practice-mean-republican-hint-2}
# Hint 2 of 2: fill in the blank
___(rep.pres.vote.share)
```

```{r practice-mean-republican-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- mean(1 - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("mean\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the mean() function.")
  }
  if (!grepl("mean\\s*\\(\\s*rep\\.pres\\.vote\\.share\\s*(,|\\))", code_no_comments)) {
    fail("Make sure you are applying mean() to rep.pres.vote.share.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("That is not quite right. Make sure you are applying mean() to the Republican vote share vector.")
  }
  pass(paste0("Exactly! The mean is a useful single-number summary --- Republican candidates averaged ",
              round(expected * 100, 1),
              "% of the two-party vote. Compare this to the Democratic mean to see how evenly matched the two parties have been."))
})
```

**Apply 37: Find Republican vote share range**

Using the object **rep.pres.vote.share** you just created, find the range of the two-party vote share received by Republican candidates. Two hints are available. Click "Submit Answer" to check your work.

```{r practice-range-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Apply 37"}

```

```{r practice-range-republican-hint-1}
# Hint 1 of 2
# You need both the smallest and largest values in
# the vector at once. Which function from the "Summary
# Functions" section returns both endpoints together?
```

```{r practice-range-republican-hint-2}
# Hint 2 of 2: fill in the blank
___(rep.pres.vote.share)
```

```{r practice-range-republican-check}
grade_this({
  dem <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488,
           0.537, 0.520, 0.511, 0.513, 0.493)
  expected <- range(1 - dem)

  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("range\\s*\\(", code_no_comments)) {
    fail("Make sure you are using the range() function.")
  }
  if (!grepl("range\\s*\\(\\s*rep\\.pres\\.vote\\.share\\s*(,|\\))", code_no_comments)) {
    fail("Make sure you are applying range() to rep.pres.vote.share.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result)) {
    fail("Your result should be a numeric vector.")
  }
  if (length(.result) != 2) {
    fail("Remember that range() returns two values: the minimum and the maximum.")
  }
  if (!isTRUE(all.equal(.result, expected, tolerance = 0.001))) {
    fail("That is not quite right. Make sure you are applying range() to the Republican vote share vector.")
  }
  pass(paste0("Correct! Republican candidates' two-party vote share ranged from ",
              round(expected[1] * 100, 1),
              "% to ",
              round(expected[2] * 100, 1),
              "% in these elections."))
})
```

The remaining questions ask you to step back from writing code and think about
what R is actually doing. In each one the code runs without an error --- the
problem is that the result is not what the person thinks it is.

Start with what happens when a variable is stored in a form that does not match
what you want to do with it.

**Apply 38: Diagnose a failed calculation**

```{r question-class-mismatch, echo=FALSE}
question("A student enters responses to a 5-point survey question by typing them with quotation marks --- \"1\", \"2\", \"3\", \"4\", \"5\" --- and stores them in a vector called **responses**. When they run `mean(responses)`, R returns `NA` along with a warning that the argument is not numeric. What went wrong?",
  correct = "Right. The quotation marks made these character strings.",
  answer("The quotation marks stored the values as character strings, and `mean()` requires numeric values.", correct = TRUE,
         message = "Anything typed inside quotation marks is stored as text, even when it looks like a number. R has no way to know the student meant these as quantities --- `class(responses)` would return `character`, and arithmetic on character values is not defined. The values would need to be entered without quotation marks to be treated as numbers."),
  answer("R recognizes that a 5-point survey scale is a set of categories and refuses to average categories.",
         message = "R does not know anything about survey scales or measurement levels --- it only knows the class of the object. The outcome here comes from the values being stored as text, not from any judgment R made about what the variable means. If the same responses were entered without quotation marks, `mean()` would run without complaint, whether or not averaging them is a good idea substantively."),
  answer("The vector contains a missing value, so `na.rm = TRUE` is needed.",
         message = "A missing value would also make `mean()` return `NA`, but without the warning that the argument is not numeric. That warning is what tells you the problem is the class of the values rather than a gap in them."),
  answer("`mean()` cannot be applied to a vector this short.",
         message = "The length of a vector does not affect whether `mean()` works. `mean()` will happily compute the mean of a single numeric value. The problem is the class of the values, not how many there are."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The stem tells you exactly what the warning says. What would make R consider these values non-numeric?"
)
```

Next, think about what actually happens at the moment you assign one object
using another.

**Apply 39: Diagnose a stale result**

```{r question-assignment-snapshot, echo=FALSE}
question("A student creates **dem.share** with the 12 Democratic vote shares, then creates **rep.share** by running `rep.share <- 1 - dem.share`. Later they notice that one value in **dem.share** was typed incorrectly and fix it. Without running anything else, they report the Republican shares from **rep.share**. Why are the reported values wrong?",
  correct = "Right. **rep.share** holds values calculated at the moment it was created.",
  answer("**rep.share** stores the values that existed when it was created, so it must be recalculated.", correct = TRUE,
         message = "Assignment stores a result, not a relationship. When `1 - dem.share` was evaluated, R computed 12 numbers and put those numbers in **rep.share**. It kept no record of where they came from. Nothing about **rep.share** changes when **dem.share** changes, which is why the corrected value never reached it. This is worth remembering whenever you build one object out of another: if the input changes, every object derived from it has to be rebuilt."),
  answer("R keeps **rep.share** linked to **dem.share** and updates it automatically, so the student must have made a different error.",
         message = "R does not maintain links between objects. This is a reasonable expectation if you are used to spreadsheets, where a cell containing a formula recalculates whenever its inputs change. R works differently: `<-` evaluates the right-hand side once and stores the result."),
  answer("Fixing one element of **dem.share** changed the vector's length, so the subtraction no longer lines up.",
         message = "Replacing a value at one position does not change how many elements a vector has. Even if the length had changed, that would not affect **rep.share**, which was already calculated and stored."),
  answer("The correction was made incorrectly --- assigning to a single position replaces the entire vector.",
         message = "Indexed assignment replaces only the position named inside the square brackets and leaves everything else intact. That is exactly what you did when you replaced the Clinton entries earlier in this tutorial."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Think about whether assignment stores a continuing relationship or the result calculated at that moment."
)
```

Now think about reading output carefully --- specifically, what a function
actually hands back to you.

**Apply 40: Read the output of a function**

```{r question-range-output, echo=FALSE}
question("A student wants a single number describing how spread out Republican vote shares are. They run `range(rep.pres.vote.share)`, see `0.453 0.592` in the output, and report that the spread is 0.453. What went wrong?",
  correct = "Right. `range()` returned two values, not one.",
  answer("`range()` returns the minimum and the maximum; the student reported the minimum as the spread.", correct = TRUE,
         message = "This is why it is worth checking what a function hands back rather than assuming. `range()` returns a two-element vector --- the smallest value and the largest value --- so `0.453` is the lowest Republican share in the data and `0.592` is the highest. Subtracting one from the other gives 0.139, which is a single number describing spread. Reading only the first value reports where the distribution starts, not how wide it is."),
  answer("The student should have reported 0.592, because the second value returned by `range()` is the spread.",
         message = "The second value is the maximum, not the spread. Both numbers `range()` returns are values that actually appear in the data --- the smallest and the largest. A single number describing spread would be the difference between them."),
  answer("The first number is the mean and the second is the spread.",
         message = "`range()` does not return a mean. Both numbers it returns are values that actually appear in the data --- the smallest and the largest."),
  answer("`range()` needs `na.rm = TRUE` to return a single value.",
         message = "`na.rm` controls how missing values are handled and has no effect on how many values `range()` returns. With or without it, `range()` returns two numbers."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Identify what each of the two numbers returned by `range()` represents."
)
```

Finally, think about what R is telling you when a summary function hands back
something other than a number.

**Apply 41: Interpret an unexpected result**

```{r question-na-diagnosis, echo=FALSE}
question("You create a vector called **income** holding the annual income reported by 1,500 survey respondents, and `class(income)` confirms it is numeric. When you run `mean(income)`, R returns `NA`. What does this suggest?",
  correct = "Right. At least one income value is missing.",
  answer("At least one respondent's income is missing.", correct = TRUE,
         message = "`NA` here is R declining to compute a mean from incomplete data rather than quietly working around the gap. Adding `na.rm = TRUE` will produce a number, but it is worth being clear about what that number is: the mean among respondents who answered the income question, not the mean among all 1,500 respondents. With income in particular that difference matters, because people who decline to report income are often not a random subset of respondents."),
  answer("Every value in the vector is missing.",
         message = "A single missing value is enough to produce this result. `mean()` does not distinguish between one missing value and a thousand --- either way it returns `NA` unless told to remove them."),
  answer("The incomes were entered as text rather than as numbers.",
         message = "Applying `mean()` to text does also return `NA`, but with a warning that the argument is not numeric --- and you have already confirmed that **income** is numeric, so that is not what happened here. Getting `NA` back means the calculation was attempted and could not be completed with the values present."),
  answer("`mean()` always returns `NA` unless `na.rm = TRUE` is specified.",
         message = "`mean()` returns a number perfectly well without `na.rm` when nothing is missing --- you saw that earlier in this tutorial with the vote share data. The argument matters only when missing values are present."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "You already know the vector is numeric. What else makes `mean()` return `NA`?"
)
```

## The Takeaways

This tutorial introduced you to the fundamental building blocks of working with data in R. These concepts will underlie everything you do in the rest of the series.

R serves two roles simultaneously: it is a calculator that performs mathematical operations, and a programming environment that stores, manipulates, and analyzes data. The bridge between these two roles is the [**object**]{.important-text} --- a named container that holds any kind of value. You assign values to objects with `<-`, and R keeps track of each object's [**class**]{.important-text} so it knows what operations are valid. The three classes you encountered here are numeric, character, and function.

Most real data analysis involves working with many values at once, which is why [**vectors**]{.important-text} are so central to R. You create them with `c()`, access individual elements with square bracket indexing `[]`, and apply arithmetic operations that automatically repeat across every element. These simple building blocks --- creating, indexing, and transforming vectors --- will appear in nearly every piece of code you write from here on.

[**Functions**]{.important-text} are what make R powerful. A function carries out a task and returns a result; most take one or more arguments. You learned to use `args()` to see what arguments a function accepts, `?` to read what a function does, and the distinction between required arguments and optional ones with default values --- along with the habit of supplying arguments by name, as you did with `seq()`, so that argument order cannot trip you up. When a function lives in a package rather than base R, you install it once with `install.packages()` and load it each session with `library()`.

Once you have a vector, a small set of functions will tell you most of what you need to know about it: `length()`, `min()`, `max()`, `range()`, `mean()`, and `sum()`. Two habits are worth carrying forward from them. The first is to check what a function actually returns rather than assuming --- `range()` hands back two numbers, the minimum and the maximum, not a single measure of spread. The second is that real data contain [**missing values**]{.important-text}, which R represents as `NA`. Most summary functions return `NA` when any value is missing, and `na.rm = TRUE` tells them to set those cases aside. That produces a number, but it also changes which cases the number describes --- a distinction worth keeping in view every time you use it.

One theme runs through all of this: R requires precision. Mismatched parentheses, a missing comma, a misspelled object name --- any of these will produce an error. This is not a sign that you are bad at R. It is just how programming works, and you will get faster at spotting these issues with practice. The fact that you have worked through this tutorial means you already have the foundation you need to keep building.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-01, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-01, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-01, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Welcome to R, you are off to a great start!">
  <span aria-hidden="true">🎉</span>
  Welcome to R, you are off to a great start!
</p>
')
```
