---
title: "Hypothesis Testing and Confidence Intervals"
tutorial:
  id: "11-Hypothesis-Testing-CI"
output:
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Null and alternative hypotheses, p-values, the decision rule, and confidence intervals.
---

## Overview {#main-content}

Every empirical claim eventually meets the same challenge: could the pattern have arisen even if nothing systematic were producing it? A difference between two groups, a correlation, a shift between two elections --- any of these can turn up even when there is no systematic relationship in whatever you are trying to learn about. The question is not settled by how large the pattern looks. It is settled by asking how often something at least that large would turn up if nothing systematic were going on, and deciding what answer would be surprising enough to act on.

This tutorial covers the two standard ways of answering that. A hypothesis test states a null claim of no relationship, works out how likely the observed result would be if that claim were true, and applies a decision rule. A confidence interval runs the same machinery in the other direction, reporting a range of plausible values for an unknown quantity, given what you observed and how much it could have varied. The two are more closely connected than they first appear, and seeing why is much of the point. Tutorial 10 supplies what both depend on --- sampling variation, the sampling distribution, and the standard error. That machinery was built for data drawn as a sample, but the reasoning is not confined to that case. The last section of this tutorial takes up what changes, and what does not, when your data cover every case there is.

In this tutorial you will learn to:

* State null and alternative hypotheses for a research question
* Explain why we test the null hypothesis rather than the alternative
* Interpret a p-value correctly and identify common misconceptions
* Apply the decision rule to reject or fail to reject the null hypothesis
* Construct and interpret a confidence interval
* Explain the connection between confidence intervals and hypothesis tests
* Apply inferential reasoning appropriately to state-level data

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(moderndive)
library(purrr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)

data("bowl", package = "moderndive")
```

## The Logic of Statistical Hypothesis Testing

Most research questions demand something more than a precise estimate: is the pattern I observe in my data systematic, or could it have arisen even if nothing systematic were producing it? A hypothesis test answers that question by evaluating two competing explanations for an observed result.

A [hypothesis test]{.important-text} evaluates two competing explanations for an observed result. The [null hypothesis]{.important-text} ($H_0$) is the specific benchmark claim the test evaluates. Most often it states that nothing is going on --- no relationship, no difference, no effect at all --- but it can name any particular value for the quantity being tested, as it will later in this tutorial when the benchmark is a mean of 50. The [alternative hypothesis]{.important-text} ($H_A$) states the competing possibility the research question asks you to assess.

### Why We Test the Null and Not the Alternative

You might wonder: why frame the test around the null hypothesis rather than directly assessing the alternative you actually care about? The standard hypothesis test is framed around the null because the null gives us a specific benchmark against which to evaluate the observed result. When the null is that a subject is guessing, each trial has exactly a 50% chance of success --- that lets us construct the distribution of results we would expect under the null and ask how surprising the actual result would be. The alternative --- that the subject performs better than chance --- includes many possible levels of ability and does not provide one single benchmark distribution. So we ask the question the null makes tractable: how surprising is this result if the null were true?

This also explains the asymmetry in how we state conclusions. Rejecting the null means the data are inconsistent with the precise null claim --- the result would be very rare if nothing were going on. Failing to reject does not prove the null true or confirm the alternative; it simply means the data are not surprising enough under the null to rule it out.

### A Motivating Example: Can Sarah Solve Problems?

In the 1970s, psychologist David Premack conducted a series of experiments with a chimpanzee named Sarah to investigate whether chimpanzees could reason about cause and effect. In one experiment, Sarah was shown a series of short videos depicting a human actor struggling with a problem --- a locked box, a disconnected hose, a shivering figure near a cold stove. After each video, she was shown two photographs: one depicting the correct solution to the problem and one depicting an incorrect or irrelevant action. Sarah chose the photograph she thought showed the solution.

Sarah got 7 out of 8 problems correct.

This result is striking, but it raises an immediate question: could Sarah have gotten 7 out of 8 simply by guessing? After all, with two choices on each trial, a chimp choosing randomly would be expected to get about 4 right by chance. Seven out of eight is better than expected --- but is it surprisingly better, or the kind of result a lucky guesser might produce?

The stakes of the answer matter. If Sarah was just guessing, her performance tells us nothing about chimpanzee cognition. If she was genuinely reasoning about cause and effect, this is evidence of a remarkable cognitive ability. We need a principled way to decide which explanation the data support.

Formally:

- **Null hypothesis** ($H_0$): Sarah was guessing. The probability of a correct answer on each trial is 0.5, like a fair coin flip. Expected correct answers: 4 out of 8.
- **Alternative hypothesis** ($H_A$): Sarah can solve problems and does better than chance. The probability of a correct answer is greater than 0.5.

Notice where the probabilities in this test come from. Nobody drew Sarah at random from a population of chimpanzees, and the test says nothing about chimpanzees in general. The 0.5 comes from the null hypothesis itself --- from what guessing would look like on a two-choice trial. That is the general pattern: a hypothesis test needs a specific claim precise enough to say what results it would produce, and the probabilities follow from that claim. Sometimes the claim is about a population you sampled from. Here it is about a process, and the reasoning works the same way.

**Practice 1: Reasoning before the simulation**

```{r q-sarah-prior, echo=FALSE}
question(
  "Before doing any calculation, 7 out of 8 correct answers
  seems like strong performance. But a skeptic says: 'You
  could get 7 out of 8 by luck --- it happens.' How should
  you respond to this argument?",
  correct = "Right. Before you can read a result you need to know how often chance alone would produce it.",
  answer(
    "The skeptic is right. Any result is possible by chance,
    so we can never conclude Sarah was doing anything other
    than guessing.",
    message = "It is true that any result is technically
    possible by chance, but some results are far more likely
    than others under the null. The question is not whether
    7 out of 8 is possible by guessing but how probable it
    is. If it is very improbable, that is evidence against
    the null --- even if it cannot be ruled out entirely."
  ),
  answer(
    "The skeptic has a point. We need to know how often chance
    alone gives 7 or more out of 8 before we can read Sarah's
    result.",
    correct = TRUE,
    message = "This is exactly the logic of hypothesis testing: acknowledge that chance is possible, then quantify how likely it is. We cannot dismiss the result as obviously real without checking, nor can we dismiss it as obviously luck without checking. We need to place Sarah's result in the full distribution of outcomes we would expect from a guessing chimp."
  ),
  answer(
    "The skeptic is wrong. Seven out of eight is obviously
    too good to be luck, so no further analysis is needed.",
    message = "Intuition can mislead us, especially with
    small samples. Eight trials is not many, and surprising
    results can occur by chance more often than we expect.
    We need a formal procedure to quantify how surprising
    Sarah's result actually is."
  ),
  answer(
    "The skeptic is technically right but it does not matter.
    Sarah got 7 of 8, and the chimp comparison is what the
    experiment was designed to settle, so the result speaks
    for itself.",
    message = "A result never speaks for itself. Without
    knowing how often 7 or more out of 8 comes up by chance,
    there is no way to say whether Sarah's score is
    remarkable --- that comparison is the whole point of the
    calculation, not a formality laid on top of it."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Can we simply decide 7 out of 8 is too good to be luck without any calculation? Can we decide it must be luck without any calculation? What do we actually need to know?"
)
```

### Simulating the Null Distribution

To determine how surprising Sarah's result is, we simulate what would happen if the null hypothesis were true --- if Sarah were guessing on every trial. We use `rbinom()` to do this. `rbinom(n = 1000, size = 8, prob = 0.5)` generates 1,000 random draws from a binomial distribution. Think of each draw as one simulated set of 8 trials for a guessing chimp: `size = 8` sets the number of trials, and `prob = 0.5` sets the probability of a correct answer under the null. Running this 1,000 times gives us a picture of the full range of outcomes we would expect from a chimp that is purely guessing. We divide by 8 to express each result as a proportion and mark Sarah's actual result of 7/8 with a vertical reference line.

**Run and observe 2: Simulate the null distribution**

```{r ex-sarah-sim, exercise=TRUE, exercise.cap="Run and observe 2", fig.alt="Histogram of the proportion correct across 1,000 simulated sets of 8 coin-flip guesses, with proportion correct on the x-axis and frequency on the y-axis. The bars rise to a peak of about 280 just below 0.5 and fall away on both sides, running from about 35 at the lowest proportions to about 30 in the 0.75 to 0.875 bin and almost nothing beyond. A vertical reference line marks Sarah's result at 0.875, out at the right-hand edge with very little of the distribution past it."}
set.seed(42)
sarah_null <- tibble(
  replicate    = 1:1000,
  correct      = rbinom(n = 1000, size = 8, prob = 0.5),
  prop_correct = correct / 8
)

ggplot(sarah_null, aes(x = prop_correct)) +
  geom_histogram(binwidth = 0.125, boundary = 0,
                 color = "white", fill = "steelblue") +
  geom_vline(xintercept = 7/8, color = "red",
             linewidth = 1.2) +
  labs(title = "Null Distribution: Outcomes for a Guessing Chimp",
       subtitle = "1,000 simulated sets of 8 trials; the reference line marks Sarah's result",
       x = "Proportion Correct",
       y = "Frequency")
```

The histogram shows the full range of outcomes a guessing chimp would produce across 1,000 experiments. The center of the distribution reflects the expected result under the null. Use this picture to decide where Sarah's result, marked by the reference line, falls relative to what we would expect from a guesser.

**Practice 3: Where does Sarah's result fall?**

```{r q-sarah-location, echo=FALSE}
question(
  "Looking at the histogram, how would you describe where
  Sarah's result of 7 out of 8 falls relative to outcomes
  we would expect from a guessing chimp?",
  correct = "Right. Sarah's result sits out in the tail, where guessing rarely lands.",
  answer(
    "Near the center --- a guessing chimp gets around 7
    out of 8 fairly often.",
    message = "The center of the distribution is near 4
    out of 8, which is the expected result under the null.
    Sarah's result of 7 out of 8 is well to the right of
    the center, not near it."
  ),
  answer(
    "In the right tail --- a guessing chimp rarely gets
    7 or more out of 8 correct.",
    correct = TRUE,
    message = "The histogram shows that 7 or 8 correct answers is rare for a guesser. Most simulated outcomes cluster between 3 and 6. Sarah's result falls in the thin right tail of the distribution --- where outcomes are uncommon under the null hypothesis. The rarer that location, the more the result challenges the null. Count the simulated outcomes that were 7 or more out of 8 --- the ones at least as extreme as Sarah's --- and you have the proportion that measures exactly how rarely a guessing chimp does this well."
  ),
  answer(
    "In the left tail --- 7 out of 8 is far below what a
    guessing chimp would usually manage.",
    message = "Check which way the axis runs. A guessing chimp
    averages 4 of 8, so 7 is well above the typical outcome, not
    below it. Being unusual is not the same as being low --- what
    matters is that the result sits far out on the side the
    alternative predicts."
  ),
  answer(
    "Outside the distribution entirely --- it is impossible
    for a guessing chimp to get 7 out of 8.",
    message = "Look at the rightmost bar of the histogram.
    A small number of simulations did produce 7 or 8 correct
    answers by chance. Sarah's result is rare under the null,
    but not impossible."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Look at the histogram. Where is the center of the distribution? Where is the red line marking Sarah's result relative to most of the bars?"
)
```

### The p-value

Sarah's result falls in the tail of the null distribution --- outcomes a guesser would rarely produce. But "rarely" needs to be made precise. The **p-value** is how we do that.

::: {.important-note}
**The p-value**

A [p-value]{.important-text} is the probability of observing a result as extreme as the one obtained --- or more extreme --- assuming the null hypothesis is true.

- A **small** p-value: the observed result would be rare under the null. The data are surprising if nothing is going on.
- A **large** p-value: the observed result would be common under the null. The data are not surprising if nothing is going on.
:::

Sarah's simulation answers the question that motivated it. Of the 1,000 guessing chimps, 26 got 7 or 8 correct --- a proportion of 0.026. That proportion is the simulated p-value: the estimated probability that a chimp with no ability at all would do at least as well as Sarah did. It is an estimate because it comes from 1,000 simulated repetitions rather than from every possible outcome; a different seed would give a slightly different number.

The figure below shows how the p-value fits into the bigger picture of the sampling distribution. Walk through it piece by piece:

- The curve itself **is** the sampling distribution under the null hypothesis --- the distribution of results you would see across many repeated samples if the null hypothesis were exactly true. A distribution of results expected under a null claim is called a [null distribution]{.important-text}. Here it comes from repeated sampling, but a null distribution can come from any process the null describes precisely enough --- Sarah's coin-flip guessing produced one without any sampling at all.
- The peak, labeled *"True value under the null hypothesis,"* is the most likely result under that assumption --- for Sarah, this is guessing correctly on half the problems.
- The two tails, labeled *"very unlikely observations,"* contain results that would be rare in either direction if the null were true --- doing much better, or much worse, than chance would predict.
- The dashed line marks the critical value for a test at the 5% significance level ($\alpha = 0.05$): results beyond this cutoff are rare enough under the null --- occurring less than 5% of the time --- to count as evidence against it.
- The dot, labeled *"Observed result,"* is where your actual data falls on this distribution --- for Sarah, the proportion of problems she actually solved correctly.
- The shaded region labeled "Area representing the p-value" shows the probability of getting a result as extreme as the one observed, or more extreme, if the null hypothesis were true. For a one-sided test, that area is in one tail. For a two-sided test, it includes results equally extreme in either direction.

Because the shaded region falls beyond the critical value, this observed result would lead to rejecting the null hypothesis: a result this far out is rare enough under the assumption of pure guessing that we reject the null hypothesis.

```{r inference-graph, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.alt="Bell curve showing the null distribution --- the distribution of results expected if the null hypothesis were true. The peak is labeled 'True value under the null hypothesis and most likely observation.' A dot to the right of center is labeled 'Observed result (value).' A vertical dashed line to the right of the dot is labeled 'Critical value, alpha equals 0.05.' The shaded tail area beyond the observed result is labeled 'Area representing the p-value.' Both tails are labeled 'very unlikely observations.'"}
navy <- "#1a3a6b"
green <- "#2d6a2d"

mu    <- 0
sigma <- 1
crit  <- 1.65
obs   <- 2.2

x  <- seq(-4, 4, length.out = 1000)
y  <- dnorm(x, mu, sigma)
df <- data.frame(x = x, y = y)
shade_df <- subset(df, x >= obs)

ggplot(df, aes(x, y)) +
  geom_line(linewidth = 2.0, color = navy) +
  geom_area(data = shade_df, fill = navy, alpha = 1.0) +
  annotate("segment", x = crit, xend = crit,
           y = 0, yend = dnorm(crit),
           linetype = "dashed", color = navy, linewidth = 1.0) +
  annotate("point", x = obs, y = 0, color = green, size = 7) +
  # x-axis
  annotate("segment", x = -3.9, xend = 4.1,
           y = 0, yend = 0,
           arrow = arrow(ends = "both", length = unit(0.2, "cm")),
           color = navy, linewidth = 0.9) +
  annotate("text", x = 0, y = -0.025,
           label = "set of possible results",
           size = 4.5, color = navy) +
  # y-axis
  annotate("segment", x = -3.9, xend = -3.9,
           y = 0, yend = 0.43,
           arrow = arrow(ends = "last", length = unit(0.2, "cm")),
           color = navy, linewidth = 0.9) +
  annotate("text", x = -4.35, y = 0.21,
           label = "probability of\nobservation",
           size = 4.0, color = navy, angle = 90, hjust = 0.5) +
  # Peak label and arrow
  annotate("text", x = -0.2, y = 0.475,
           label = "True value under the null hypothesis\nand most likely observation",
           size = 4.0, fontface = "bold", hjust = 0.5, color = navy) +
  annotate("segment", x = 0, xend = 0,
           y = 0.445, yend = dnorm(0) + 0.025,
           arrow = arrow(length = unit(0.28, "cm")),
           color = navy, linewidth = 0.9) +
  # Critical value label and arrow
  annotate("text", x = crit + 1.05, y = 0.33,
           label = "Critical value\n(\u03b1 = 0.05)",
           size = 4.0, fontface = "bold", hjust = 0.5, color = navy) +
  annotate("segment", x = crit + 0.55, xend = crit + 0.04,
           y = 0.27, yend = 0.13,
           arrow = arrow(length = unit(0.22, "cm")),
           color = navy, linewidth = 0.7) +
  # Observed result label and arrow
  annotate("text", x = obs - 1.4, y = 0.08,
           label = "Observed result\n(value)",
           size = 4.0, fontface = "bold", color = green, hjust = 0.5) +
  annotate("segment", x = obs - 1.0, xend = obs - 0.12,
           y = 0.055, yend = 0.012,
           arrow = arrow(length = unit(0.22, "cm")),
           color = green, linewidth = 0.7) +
  # Observed p-value label and arrow
  annotate("text", x = obs + 1.1, y = 0.26,
           label = "Area representing\n the p-value",
           size = 4.0, fontface = "bold", color = green, hjust = 0.5) +
  annotate("segment", x = obs + 0.75, xend = obs + 0.4,
           y = 0.21, yend = 0.055,
           arrow = arrow(length = unit(0.22, "cm")),
           color = green, linewidth = 0.7) +
  # Left bracket
  annotate("segment", x = -3.6, xend = -2.3,
           y = 0.018, yend = 0.018,
           color = navy, linewidth = 0.8) +
  annotate("segment", x = -3.6, xend = -3.6,
           y = 0.008, yend = 0.028,
           color = navy, linewidth = 0.8) +
  annotate("segment", x = -2.3, xend = -2.3,
           y = 0.008, yend = 0.028,
           color = navy, linewidth = 0.8) +
  annotate("text", x = -2.95, y = 0.046,
           label = "very unlikely\nobservations",
           size = 3.8, color = navy, hjust = 0.5) +
  # Right bracket
  annotate("segment", x = 2.9, xend = 3.7,
           y = 0.018, yend = 0.018,
           color = navy, linewidth = 0.8) +
  annotate("segment", x = 2.9, xend = 2.9,
           y = 0.008, yend = 0.028,
           color = navy, linewidth = 0.8) +
  annotate("segment", x = 3.7, xend = 3.7,
           y = 0.008, yend = 0.028,
           color = navy, linewidth = 0.8) +
  annotate("text", x = 3.3, y = 0.046,
           label = "very unlikely\nobservations",
           size = 3.8, color = navy, hjust = 0.5) +
  theme_void() +
  theme(plot.margin = margin(10, 60, 10, 70)) +
  coord_cartesian(xlim = c(-4.5, 4.5),
                  ylim = c(-0.045, 0.51),
                  clip = "off")
```

::: {.caution}
**The most common p-value misconception:** A p-value of 0.03 does NOT mean there is a 3% chance the null hypothesis is true. It means that IF the null were true, there would be only a 3% chance of observing a result as extreme as yours. These statements sound similar but are logically very different. The p-value is a probability about the data given the null --- not a probability about the null given the data.
:::

### The Decision Rule

To turn a p-value into a conclusion, we compare it to a pre-chosen threshold called the [significance level]{.important-text}, written $\alpha$. Throughout this course we use $\alpha = .05$, and it is the most common choice in political science --- but it is a convention, not a rule. Nothing about .05 makes it the boundary between a real relationship and a spurious one; a p-value of .049 and one of .051 are telling you nearly the same thing. Researchers sometimes use $\alpha = .10$ when a test has little power to detect an effect --- a small sample, or an effect expected to be modest --- and sometimes use $\alpha = .01$ when a false positive would be costly. What matters is that the threshold is chosen before the test is run, not after the p-value is known.

- If the p-value is **less than** $\alpha$: **reject the null hypothesis**. The result is [statistically significant]{.important-text} --- the data provide sufficient evidence to reject the null hypothesis at the chosen significance level.
- If the p-value is **greater than or equal to** $\alpha$: **fail to reject the null hypothesis**. The data are not surprising enough under the null to rule it out.

::: {.caution}
**Failing to reject is not the same as accepting the null.** You never prove the null hypothesis true. You only decide whether the evidence against it is strong enough to reject it. The correct phrasing is always "fail to reject" --- not "accept."
:::

**Apply 4: Interpreting a p-value correctly**

```{r q-pvalue-interpret, echo=FALSE}
question(
  "A researcher tests whether states with higher income
  inequality have higher rates of political polarization.
  She finds a correlation of r = 0.41 in her sample of
  48 states, with a p-value of 0.004. Which of the
  following is the correct interpretation?",
  correct = "Right. A p-value is a probability about the data given the null, not about the null given the data.",
  answer(
    "There is a 0.4% chance the null hypothesis is true,
    so the researcher can be 99.6% confident that
    inequality causes polarization.",
    message = "This is the most common p-value misconception.
    The p-value does not give the probability that the null
    is true, and it says nothing about causation. It only
    tells you how surprising the observed result would be
    if the null were true."
  ),
  answer(
    "The probability that the true population correlation
    is exactly 0.41 is 99.6%.",
    message = "A p-value is not a probability attached to
    the size of a parameter. It is the probability of
    observing the data, or something more extreme, given
    that the null hypothesis is true."
  ),
  answer(
    "If the null were true, a correlation this far from zero
    would occur only 0.4% of the time.",
    correct = TRUE,
    message = "The p-value is a conditional probability: the probability of observing the data, or something more extreme, given that the null hypothesis is true. A p-value of 0.004 means a correlation this far from zero would occur only 0.4% of the time if there were truly no relationship --- strong evidence against the null. Notice it says nothing about causation."
  ),
  answer(
    "A correlation of 0.41 with p = 0.004 means the
    relationship between inequality and polarization is
    strong, since the p-value is very small.",
    message = "A p-value measures how surprising the result
    would be if the null were true, not how large the
    relationship is --- that is what r = 0.41 reports. A tiny
    correlation can carry a very small p-value in a large
    sample, and a substantial one can miss significance in a
    small sample."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "A p-value is a probability about the data given the null, not a probability about the null given the data. Which answer correctly states what is conditional on what?"
)
```

**Apply 5: Significance versus importance**

```{r q-sig-vs-importance, echo=FALSE}
question(
  "Study 1 finds a mean difference of 5.2 points with
  p = 0.03. Study 2 finds a mean difference of 12.7 points
  with p = 0.21. Both use $\\alpha = 0.05$. A student says
  Study 2 is more convincing because it found a larger
  effect. What is wrong with this reasoning?",
  correct = "Right. A small p-value and a large effect are separate questions, and sample size drives the first.",
  answer(
    "The student is correct. A larger result is always
    stronger evidence against the null hypothesis.",
    message = "A larger result is not always stronger
    evidence. Statistical significance depends on both
    the size of the result and the precision of the
    estimate. A large result estimated imprecisely ---
    reflected in a high p-value --- provides weaker
    evidence than a smaller result estimated precisely."
  ),
  answer(
    "Study 1 matters more, because statistical significance
    means its effect is the larger of the two.",
    message = "Significance and size are separate. A small
    difference can reach significance with a large enough sample,
    and a large one can miss it with a small sample. The p-value
    says how surprising the result would be under the null, not
    how big the effect is."
  ),
  answer(
    "Nothing is wrong. Both studies should reject the
    null because both found positive differences.",
    message = "The direction of a difference does not
    determine whether to reject the null. The decision
    is based on comparing the p-value to alpha. Study 1:
    p = 0.03 < 0.05, reject. Study 2: p = 0.21 > 0.05,
    fail to reject."
  ),
  answer(
    "Statistical significance and practical importance are
    different things, and Study 2's larger effect did not
    reach significance.",
    correct = TRUE,
    message = "Statistical significance and practical importance are not the same thing. Study 1 is statistically significant despite the smaller effect because the result provides stronger evidence against the null --- a result this far from zero would be unlikely if the null were true. Study 2 found a larger effect but the evidence against the null was not strong enough --- possibly due to a smaller or noisier sample. Always consider both the p-value and the magnitude of the effect."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Statistical significance depends on both the size of the effect and the precision of the estimate. Which study has a p-value below the threshold? Does that make it more convincing than Study 2?"
)
```


## Confidence Intervals

Hypothesis testing gives you a yes/no decision: reject the null or fail to reject. But it does not tell you the magnitude of the association or how precisely your sample estimates the population parameter. A [confidence interval]{.important-text} provides that complementary information. Rather than asking "is the parameter different from zero?", a confidence interval asks "what range of values for the population parameter is consistent with my data?"

The key to constructing a confidence interval is knowing how far to stretch the interval in each direction from your sample statistic. That distance is measured in standard errors, and the number of standard errors you go out depends on two things: how confident you want to be, and the shape of the sampling distribution. The number that captures this is called the [critical value]{.important-text}. It is defined precisely below, but keep it in mind as we build the interval.

### Building a Confidence Interval from the Sampling Distribution

Recall from the Empirical Rule that in a normal distribution, about 95% of values fall within 1.96 standard deviations of the mean. Applied to the sampling distribution, this means that about 95% of all sample means fall within 1.96 standard errors of the true population mean $\mu$.

Now turn that logic around. If 95% of sample means land within 1.96 standard errors of $\mu$, then an interval stretching 1.96 standard errors in each direction from your sample mean will capture $\mu$ in 95% of repeated samples. That is the confidence interval, and 1.96 is the critical value for 95% confidence.

::: {.important-note}
**The 95% confidence interval**

A [95% confidence interval]{.important-text} for the population mean consists of a lower bound and an upper bound:

$$\text{Lower bound} = \bar{Y} - 1.96 \times s.e. = \bar{Y} - 1.96 \times \frac{s}{\sqrt{n}}$$

$$\text{Upper bound} = \bar{Y} + 1.96 \times s.e. = \bar{Y} + 1.96 \times \frac{s}{\sqrt{n}}$$

where $\bar{Y}$ is the sample mean, $s$ is the sample standard deviation, $n$ is the sample size, and 1.96 is the critical value corresponding to 95% confidence. The interval is written as [lower bound, upper bound].
:::

### Critical Values and When They Apply

The critical value of 1.96 is not arbitrary. It comes directly from the standard normal distribution --- the specific value that puts 95% of the distribution in the middle and 2.5% in each tail. This is the precise version of the Empirical Rule's "approximately 2 standard deviations." Different confidence levels require different critical values, because capturing a larger percentage of the distribution requires going further out into the tails.

| Confidence level | Critical value |
|:-----------------|:--------------:|
| 90% | 1.645 |
| 95% | 1.96 |
| 99% | 2.576 |

A higher [confidence level]{.important-text} uses a larger critical value, which produces a wider interval. This is the tradeoff at the heart of confidence intervals: more confidence requires more width.

::: {.caution}
**These critical values assume the relevant sampling distribution is approximately normal.** For sample means and proportions, the CLT often justifies this approximation when samples are large enough. For small samples, the [t-distribution]{.important-text} is used instead of the normal distribution, and the critical value is larger than 1.96 --- how much larger depends on the sample size. For other types of tests entirely --- chi-square tests for categorical variables, F-tests in regression --- different distributions apply and different critical values are used. What does not change is the basic logic: use the appropriate reference distribution to decide which values are plausible, given the sample statistic and its uncertainty. In practice, R functions like `t.test()`, `chisq.test()`, and `lm()` determine the correct critical value automatically. The interpretation of the resulting interval is the same regardless of which distribution is used.
:::

### A Worked Example

Back in the Introduction, Gallup reported that 30% of a sample of 1,200 American adults approved of the way the president was handling the job. That 30% is the point estimate. But how much uncertainty surrounds it?

Suppose the standard deviation of approval ratings in the sample is 45.8 percentage points (treating approval as a 0/1 variable --- 1 for approve, 0 for disapprove --- the standard deviation of a proportion $p$ is $\sqrt{p(1-p)}$, and at p = 0.30 the square root of 0.30 × 0.70, times 100, gives 45.8 percentage points). Using the 95% critical value of 1.96, we construct the interval in four steps.

**Step 1: Compute the standard error.**

$$s.e. = \frac{45.8}{\sqrt{1200}} = \frac{45.8}{34.6} = 1.32 \text{ points}$$

**Step 2: Compute the margin of error.**

$$1.96 \times 1.32 = 2.59 \text{ points}$$

**Step 3: Compute the lower and upper bounds.**

$$\text{Lower bound} = 30.0 - 2.59 = 27.41\%$$

$$\text{Upper bound} = 30.0 + 2.59 = 32.59\%$$

**Step 4: State the interval.**

The 95% confidence interval is [27.41%, 32.59%]. The interval says that between 27% and 33% of all American adults approve of the president, with 95% confidence.

```{r gallup-ci-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="The 95% confidence interval for presidential approval: a point estimate of 30.0% with a margin of error of 2.59 points, giving an interval of [27.41%, 32.59%].", fig.alt="Horizontal dot-and-whisker plot showing a single confidence interval. A point marks the estimate of 30 percent approval. A horizontal line through the point extends from 27.41 percent to 32.59 percent, representing the 95 percent confidence interval."}
gallup_data <- tibble(
  label    = "Approval estimate",
  estimate = 30.0,
  lower    = 27.41,
  upper    = 32.59
)

ggplot(gallup_data, aes(x = estimate, y = label)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                  height = 0.3, linewidth = 1, color = "#1565C0") +
  geom_point(size = 3, color = "#1565C0") +
  annotate("text", x = 30.0, y = 1.35,
           label = "Point estimate: 30.0%",
           color = "#1565C0", size = 3.2, fontface = "bold") +
  annotate("text", x = 30.0, y = 0.65,
           label = "95% CI: [27.41%, 32.59%]",
           color = "#1565C0", size = 3.2) +
  scale_y_discrete(expand = expansion(mult = c(0.6, 0.6))) +
  labs(
    title = "A 95% Confidence Interval Has a Center and a Width",
    subtitle = "Presidential approval, Gallup sample of 1,200 adults",
    x = "Percent Approving",
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The dot marks the point estimate --- the single best guess at the population value. The horizontal line extending from it shows the [margin of error]{.important-text} in both directions: the full range of values the data are consistent with. Every confidence interval you build from here on will have this same basic shape --- a center and a width --- whether you compute it by hand or with R.

### What "95% Confidence" Actually Means

The phrase "95% confidence" has a precise meaning that is easy to misstate.

::: {.important-note}
**Correct interpretation**

In repeated sampling, 95% of confidence intervals constructed this way will contain the true population parameter. The 95% refers to the long-run performance of the procedure across many samples --- not to the probability that this particular interval contains the truth.
:::

```{r ci-repeated-intervals, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=7, fig.cap="Twenty-five confidence intervals, each constructed from a different random sample of size 100 from the bowl. The dashed reference line marks the true population proportion of 0.375. Most intervals cross that line and contain the true value; the few drawn as dashed lines fall entirely to one side and miss it. In the long run, 95% of intervals constructed this way will contain the true proportion.", fig.alt="Horizontal dot-and-whisker plot showing twenty-five 95 percent confidence intervals stacked vertically, each from a different random sample of size 100. A dashed vertical reference line marks the true population proportion of 0.375. Most intervals are drawn as solid lines that cross the dashed line, meaning they contain the true proportion. A small number are drawn as dashed lines that fall entirely to one side of it, meaning they miss the true proportion."}
set.seed(42)
n_intervals <- 25
sample_size <- 100
true_pi     <- mean(bowl$color == "red")

ci_data <- map_dfr(1:n_intervals, function(i) {
  samp  <- bowl %>% rep_sample_n(size = sample_size, reps = 1)
  p_hat <- mean(samp$color == "red")
  se    <- sqrt(p_hat * (1 - p_hat) / sample_size)
  tibble(
    sample_id = i,
    estimate  = p_hat,
    lower     = p_hat - 1.96 * se,
    upper     = p_hat + 1.96 * se,
    contains  = (lower <= true_pi & upper >= true_pi)
  )
})

ggplot(ci_data,
       aes(x = estimate, y = sample_id,
           color = contains, linetype = contains,
           xmin = lower, xmax = upper)) +
  geom_vline(xintercept = true_pi, linetype = "dashed",
             color = "#C62828", linewidth = 0.8) +
  geom_errorbarh(height = 0.4, linewidth = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("FALSE" = "#B8860B", "TRUE" = "#00897B"),
    labels = c("FALSE" = "Misses true proportion",
               "TRUE"  = "Contains true proportion"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "dashed", "TRUE" = "solid"),
    labels = c("FALSE" = "Misses true proportion",
               "TRUE"  = "Contains true proportion"),
    name = NULL
  ) +
  labs(
    title    = "Twenty-Five 95% Confidence Intervals",
    subtitle = paste0("Each from a different random sample of size 100; dashed line marks the true proportion of ", round(true_pi, 3)),
    x        = "Sample Proportion (Red Balls)",
    y        = "Sample number"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


::: {.caution}
**A common misinterpretation:** It is tempting to say "there is a 95% probability that the true approval rate falls between 27.41% and 32.59%." This is incorrect. The true approval rate is a fixed value --- it either is or is not in this interval. The 95% describes the procedure: if Gallup ran this survey many times and constructed a confidence interval each time, about 95% of those intervals would contain the true approval rate.
:::

**Apply 6: Computing a confidence interval**

```{r q-ci-compute, echo=FALSE}
question(
  "A researcher surveys 100 countries and finds that the mean
  democracy score is 54 points with a standard deviation of
  20 points. Using the 95% critical value of 1.96, what are
  the lower and upper bounds of the confidence interval for
  the population mean democracy score?",
  correct = "Right. The interval is the mean plus and minus 1.96 standard errors.",
  answer(
    "Lower bound: 34, upper bound: 74. Computed as
    54 - 20 and 54 + 20.",
    message = "This uses the standard deviation directly rather
    than the standard error. The standard error accounts for
    sample size: $20 / \\sqrt{100} = 20 / 10 = 2.0$. The margin
    of error is $1.96 \\times 2.0 = 3.92$, not 20."
  ),
  answer(
    "Lower bound: 50.1, upper bound: 57.9. Computed as
    $54 - 1.96 \\times (20/\\sqrt{100})$ and
    $54 + 1.96 \\times (20/\\sqrt{100})$.",
    correct = TRUE,
    message = "Step by step: the standard error is $20 / \\sqrt{100} = 20 / 10 = 2.0$. The margin of error is $1.96 \\times 2.0 = 3.92$. The lower bound is 54 - 3.92 = 50.08, which rounds to 50.1; the upper bound is 54 + 3.92 = 57.92, which rounds to 57.9. The interval says that the true mean democracy score across all countries falls between 50.1 and 57.9 with 95% confidence."
  ),
  answer(
    "Lower bound: 53.6, upper bound: 54.4. Computed as
    $54 - 1.96 \\times (20/100)$ and
    $54 + 1.96 \\times (20/100)$.",
    message = "The standard error formula divides by
    $\\sqrt{n}$, not $n$. The standard error is
    $20 / \\sqrt{100} = 2.0$, not $20 / 100 = 0.20$, so this
    interval comes out far too narrow."
  ),
  answer(
    "Lower bound: 54.0, upper bound: 54.0. Computed as
    $54 - 1.96 \\times (20/\\sqrt{100}/100)$ and
    $54 + 1.96 \\times (20/\\sqrt{100}/100)$.",
    message = "This divides the standard error by an extra
    factor of 100, leaving an interval so narrow it collapses
    onto the sample mean. The standard error is simply
    $s / \\sqrt{n} = 20 / \\sqrt{100} = 2.0$."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Work through the four steps: compute the standard error ($s / \\sqrt{n}$), compute the margin of error (1.96 times the standard error), then add and subtract from the sample mean."
)
```

**Apply 7: Interpreting confidence level and width**

```{r q-ci-width, echo=FALSE}
question(
  "Using the same democracy score data (mean = 54, s = 20,
  n = 100), a colleague computes a 99% confidence interval
  using the critical value of 2.576 instead of 1.96. She
  finds the interval is wider and says this makes it less
  useful. What is the flaw in her reasoning?",
  correct = "Right. Width and usefulness are not the same thing --- a wider interval buys stronger coverage.",
  answer(
    "Her reasoning is incomplete. A wider interval is less
    precise, but the wider one buys stronger long-run
    coverage, and which you want depends on the stakes.",
    correct = TRUE,
    message = "There is always a tradeoff between confidence and precision. A 99% interval is wider --- [48.9, 59.1] versus [50.1, 57.9] --- but it is less likely to miss the true parameter across repeated samples. In high-stakes decisions where the cost of missing the truth is high, that additional coverage may be exactly what is needed. Width is a cost; coverage is a benefit. Whether the tradeoff is worth it depends on context."
  ),
  answer(
    "She is correct. A wider interval is always less
    informative because it is less precise.",
    message = "Width and usefulness are not the same thing.
    A 99% interval is wider precisely because it provides
    stronger coverage --- it will contain the true parameter
    in 99% of repeated samples rather than 95%. Whether
    that additional coverage is worth the wider range
    depends on the context."
  ),
  answer(
    "The wider interval means her data became less precise when she moved from 95% to 99% confidence.",
    message = "The data did not change, and neither did the standard error. The interval widened because a higher confidence level requires a larger critical value --- 2.576 in place of 1.96. Precision is a property of the estimate; the interval width reflects both that precision and the coverage you asked for."
  ),
  answer(
    "Her reasoning confuses the confidence level with the probability that this particular interval contains the true value. A 99% interval does not mean there is a 99% probability the true mean is inside it.",
    message = "This is a real misconception worth knowing, but it does not address the flaw in her reasoning about width and usefulness. Her error is treating width as purely a cost without recognizing that higher confidence level also means stronger long-run coverage --- a benefit that may be worth the wider range in some contexts."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "What does a 99% interval guarantee that a 95% interval does not? Is wider always worse, or does it depend on what you need?"
)
```

### The Connection Between Confidence Intervals and Hypothesis Tests

Confidence intervals and hypothesis tests are two sides of the same logic. Both are built from the sampling distribution. Both use the standard error and the critical value. And when the confidence level and test significance level match, a two-sided hypothesis test and its corresponding confidence interval reach the same reject/fail-to-reject conclusion about the null value.

Think of it this way: a 95% confidence interval marks the range of parameter values that are consistent with your data at the 0.05 level. A two-sided hypothesis test at $\alpha = 0.05$ asks whether the null value is consistent with your data at the same level. If the null value falls outside the interval --- outside the range of values consistent with the data at the 0.05 level --- the p-value is less than 0.05 and you reject the null. If it falls inside the interval, the p-value is greater than 0.05 and you fail to reject.

This equivalence holds specifically for two-sided tests at the matching significance level: a 95% CI corresponds to a two-sided test at $\alpha = 0.05$, a 99% CI to a two-sided test at $\alpha = 0.01$, and so on.

This equivalence is useful in practice: once you have a confidence interval, you can evaluate the corresponding two-sided hypothesis test about the mean at the matching significance level without computing a separate p-value --- simply check whether the null value falls inside or outside the interval.

The figure below uses the same kind of plot you saw earlier for the Gallup approval estimate --- a point and a horizontal interval --- but adds one new feature: the null hypothesis value, marked separately, so you can see directly whether it falls inside or outside the interval.

```{r ci-ht-connection, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="A 95% confidence interval of [50.1, 57.9] for the mean democracy score across 100 countries, with the null hypothesis value of 50 marked. Because the null value falls outside the interval, the corresponding two-sided hypothesis test at $\\alpha = 0.05$ rejects the null.", fig.alt="Horizontal dot-and-whisker plot showing a 95 percent confidence interval from 50.1 to 57.9 for the mean democracy score, drawn as a horizontal bar with a point at the center estimate of 54. An X to the left of the interval, at 50, marks the null hypothesis value. The X falls outside the bar, illustrating that the null value is excluded by the data."}
ci_lower    <- 50.1
ci_upper    <- 57.9
ci_estimate <- 54.0
null_value  <- 50

connection_data <- tibble(
  label    = "Sample estimate",
  estimate = ci_estimate,
  lower    = ci_lower,
  upper    = ci_upper
)

ggplot(connection_data, aes(x = estimate, y = label)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                  height = 0.3, linewidth = 1, color = "#00897B") +
  geom_point(size = 3, color = "#00897B") +
  geom_point(aes(x = null_value, y = label),
             shape = 4, size = 4, stroke = 1.3, color = "#C62828") +
  annotate("text", x = null_value, y = 1.35,
           label = paste0("Null value\nH0: mean = ", null_value),
           color = "#C62828", size = 3.2, fontface = "bold", hjust = 0) +
  annotate("text", x = ci_estimate, y = 0.65,
           label = paste0("95% CI: [", ci_lower, ", ", ci_upper, "]"),
           color = "#00897B", size = 3.2, fontface = "bold") +
  scale_y_discrete(expand = expansion(mult = c(0.6, 0.6))) +
  labs(
    title = "The Null Value Falls Outside the Confidence Interval",
    subtitle = "So the equivalent hypothesis test rejects the null at alpha = 0.05",
    x = "Mean Democracy Score (0-100 scale)",
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The null value (50, marked with the X) sits to the left of the entire interval --- it is not among the values consistent with the data at the 0.05 level. That is exactly the connection: when the null value falls outside the corresponding 95% confidence interval, the two-sided p-value is below 0.05. If the null value had fallen between the two ends of the bar instead, the test would fail to reject, because the null would be one of the plausible values given the data.

**Apply 8: Using a CI to evaluate a hypothesis test**

```{r q-ci-ht, echo=FALSE}
question(
  "A researcher surveys 100 countries and tests whether the mean
  democracy score differs from 50 --- the midpoint of the 0-to-100
  scale, representing neither autocracy nor democracy. Her sample
  yields a 95% confidence interval of [50.1, 57.9]. At $\\alpha = 0.05$,
  what should she conclude, and what does the confidence interval
  tell her beyond the hypothesis test?",
  correct = "Right. A null value outside the interval and a rejected null are the same finding twice.",
  answer(
    "Fail to reject the null. The interval [50.1, 57.9]
    is close to 50, so the difference is probably not
    meaningful.",
    message = "Statistical significance is not determined
    by whether values seem close to the null. It is
    determined by whether the null value falls inside or
    outside the confidence interval. Here 50 is entirely
    below the lower bound of 50.1, so she rejects the null."
  ),
  answer(
    "Cannot determine the conclusion from a confidence
    interval alone --- she would need to compute the
    p-value separately.",
    message = "A 95% confidence interval and the corresponding
    two-tailed hypothesis test at $\\alpha = 0.05$ reach the same
    conclusion when they are based on the same model and
    assumptions. If the null value falls outside the interval,
    reject. No separate p-value calculation is needed here."
  ),
  answer(
    "Reject the null hypothesis. The null value of 50 falls
    below the lower bound of the interval.",
    correct = TRUE,
    message = "Because the null value of 50 falls outside the interval [50.1, 57.9], the p-value must be below 0.05 --- reject the null. But the confidence interval adds something a p-value alone cannot provide: the range of plausible values for the mean. The population mean democracy score is estimated to lie between 50.1 and 57.9, so the entire 95% confidence interval lies above the scale midpoint. That gives substantively useful information beyond the yes/no hypothesis-test decision."
  ),
  answer(
    "Reject the null hypothesis. The null value of 50 falls
    above the upper bound of the interval, so the true mean
    is below the scale midpoint.",
    message = "Check which side. The interval runs from 50.1
    to 57.9, so 50 sits below its lower bound, not above its
    upper bound. The decision is right but the direction is
    backwards: the estimated mean is above the midpoint of
    the scale, not below it."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Check whether the null value of 50 falls inside or outside the confidence interval [50.1, 57.9]. If it falls outside, what does that tell you about the p-value relative to 0.05?"
)
```

## A Note on Inference with Complete Data

Throughout this tutorial the examples have involved clear sampling situations: surveys of voters, samples of countries, bowls of colored balls. You have a subset of a larger population and you want to say something about the whole.

But research in political science often uses data that cover every case in a defined population --- all 50 states and DC, all members of Congress in a given session, all countries in the world for a given year. Nothing was left out. Is inference even appropriate?

It depends on what you are trying to do. If your question is purely descriptive --- "what is the mean voter turnout rate across U.S. states this year?" --- then you have the full population of interest and no inference is needed. The statistic you compute is the population parameter, not an estimate of it.

But that is almost never the research question in political science. When you ask whether income inequality predicts polarization, whether stricter gun laws reduce crime rates, or whether states with higher education spending have better health outcomes, you are asking about a [relationship]{.important-text} --- a systematic pattern that could have looked different under different historical conditions. Inference is the tool for asking whether the observed pattern is strong enough to be unlikely if no systematic relationship existed.

When you run a hypothesis test on such data, the p-value answers: **if there were truly no relationship between these variables, how surprising would it be to observe a pattern this strong in these data?** A small p-value means the pattern would be unlikely under the null. It does not prove causation, but it is evidence the relationship is not simply noise. Researchers may therefore apply inferential tools to complete data when they treat the observed cases as realizations of a broader process that could have produced different outcomes. That broader inferential target must be made explicit; asking about a relationship does not by itself create sampling uncertainty.

::: {.important-note}
**When to use inference with complete data**

Use inference when your research question concerns a broader relationship or process and the observed cases can reasonably be treated as realizations of that process --- even if your data cover every case in a narrowly defined population. The p-value then answers: is this pattern strong enough to be unlikely under the null?

Skip inference when your question is purely **descriptive** of a fixed, enumerable population at a specific point in time. If you want to know the mean turnout rate across all U.S. states last year, and you have data on all of them, the mean you compute is the answer --- no uncertainty to quantify.
:::

**Apply 9: Inference or no inference?**

```{r q-states-inference, echo=FALSE}
question(
  "A researcher has data on all 50 U.S. states and DC for
  a given year --- every state is included, nothing is left
  out. She conducts two analyses. Analysis 1: She computes
  the mean percentage of the population living in poverty
  across all 51 observations and reports it as a descriptive
  fact. Analysis 2: She runs a regression testing whether
  states with higher poverty rates have lower voter turnout,
  and reports a p-value. For which analysis is statistical
  inference appropriate, and why?",
  correct = "Right. Describing the cases you have needs no inference; claiming a relationship holds beyond them does.",
  answer(
    "Inference is appropriate for both. Any time you
    compute a statistic from data you need a p-value
    to know if it is meaningful.",
    message = "Inference is not needed for every statistic.
    When your question is purely descriptive and your data
    cover the full population of interest, the statistic
    you compute is the population parameter --- there is
    no sampling uncertainty to quantify. Inference is
    needed when you are drawing conclusions about a
    relationship or process beyond the specific cases
    in your data."
  ),
  answer(
    "Inference is appropriate for neither. Because all
    50 states and DC are in the data, both analyses
    describe a complete population and no inference
    is needed.",
    message = "This is true for Analysis 1, but not
    Analysis 2. When the question is about a relationship
    between variables --- whether poverty predicts turnout
    --- the 51 observations are treated as one realization
    of an underlying process. Inference quantifies whether
    the observed relationship is strong enough to be
    unlikely under the null hypothesis that no systematic
    relationship exists."
  ),
  answer(
    "Inference is appropriate for Analysis 2 but not Analysis
    1. Describing all 51 observations needs no inference;
    claiming a relationship does.",
    correct = TRUE,
    message = "Analysis 1 is a descriptive claim about a fixed, enumerable population --- all U.S. states and DC --- and the mean computed from all 51 observations is the population parameter, not an estimate. Analysis 2 is a claim about a systematic relationship between two variables that could generalize beyond this particular historical moment, and inference is the right tool for assessing whether the observed pattern would be surprising under a null hypothesis of no systematic relationship."
  ),
  answer(
    "Inference is appropriate for Analysis 1 but not
    Analysis 2. Means require inference; regression
    results describe the data directly.",
    message = "This is backwards. A mean computed from
    all 51 observations is a population parameter ---
    no inference needed. A regression coefficient
    testing whether poverty predicts turnout makes a
    claim about a relationship that could have been
    different under different historical conditions,
    and inference is appropriate for assessing that
    claim."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "For each analysis, ask: is this a descriptive claim about a known population, or a claim about a relationship or process that extends beyond the data?"
)
```

## Check Your Understanding

These last questions step back from the calculations and ask about the reasoning behind them. Focus on what a test licenses you to say and what it does not, on where the probability in a p-value comes from, and on what a confidence interval adds once the test has given its verdict.

**Apply 10: What failing to reject does not mean**

```{r q-fail-to-reject, echo=FALSE}
question(
  "A researcher tests whether states with higher education
  spending have better health outcomes and obtains a p-value
  of 0.21. A colleague concludes that the study has shown
  there is no relationship between the two. What is wrong
  with that conclusion?",
  correct = "Right. A test can give you grounds to rule the null out, never grounds to rule it in.",
  answer(
    "Failing to reject the null is not the same as showing
    the null is true.",
    correct = TRUE,
    message = "Failing to reject means the data are not
    surprising if the null were true. They are consistent with
    no relationship, but also with a range of relationships too
    small or too uncertain to detect with this evidence."
  ),
  answer(
    "Nothing is wrong. A p-value above 0.05 is how a study
    shows that no relationship exists.",
    message = "This is the most common misreading of a
    hypothesis test. The test asks only whether the evidence
    is surprising under the null; it has no way of confirming
    the null is true."
  ),
  answer(
    "The colleague should have used a larger alpha, which
    would have made the result significant.",
    message = "Alpha is chosen before the test as a standard
    of evidence, not adjusted afterward to reach a preferred
    conclusion. Choosing it to change the verdict defeats its
    purpose."
  ),
  answer(
    "The colleague is wrong because 0.21 is close enough to
    0.05 that more data would certainly produce a significant
    result.",
    message = "More data narrows the interval and can change
    the verdict, but nothing guarantees which way. If there is
    no relationship, more data makes that clearer rather than
    reversing it."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Think about what the test actually compares your data against, and what a result that is unsurprising under the null does and does not establish."
)
```

**Apply 11: Where the probability comes from**

```{r q-null-probability, echo=FALSE}
question(
  "A researcher examines whether Supreme Court justices vote
  with the president who appointed them more often than they
  vote against. She has every vote by every justice in a full
  term --- nothing was sampled. She tests against the null
  that a justice is equally likely to vote either way. Where
  does the probability behind her p-value come from?",
  correct = "Right. The null describes a process, and the probabilities follow from that description.",
  answer(
    "From the null hypothesis itself, which describes a
    process specific enough to say what results it would
    produce.",
    correct = TRUE,
    message = "The null says each vote is like a coin flip,
    and that claim is what makes it possible to work out how
    often a term's worth of votes would come out as lopsided
    as hers. No sampling is involved anywhere in the
    calculation."
  ),
  answer(
    "From treating the justices as a random sample drawn from
    all the justices who might have served.",
    message = "Nothing was sampled, and the test does not
    require it. Inventing a population to have sampled from is
    exactly the move that makes inference on complete data
    look contrived."
  ),
  answer(
    "Nowhere. Because she observed every vote, there is no
    uncertainty and the p-value has no meaning.",
    message = "She has every vote in the term, so a purely
    descriptive question would need no inference. But her
    question is about how justices vote, and the term she
    observed is one realization of that."
  ),
  answer(
    "From the Central Limit Theorem, which supplies
    probabilities whatever the data are.",
    message = "The CLT describes the shape of a sampling
    distribution. It does not tell you what the null predicts,
    which is what a p-value is computed against."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The null says each vote is like a coin flip. What does that claim let you calculate, and does it require anyone to have sampled anything?"
)
```

**Apply 12: A confidence interval that contains the null**

```{r q-ci-contains-null, echo=FALSE}
question(
  "Another researcher, using the same measure, finds a mean
  democracy score of 52 with a standard deviation of 20 in
  100 countries, giving a 95% confidence interval of
  [48.1, 55.9]. She is testing whether the mean differs from
  50 at $\\alpha = 0.05$. What should she conclude?",
  correct = "Right. The null value is inside the interval, and the interval names the values the data are consistent with.",
  answer(
    "Fail to reject the null: 50 lies inside the interval,
    and the interval names the plausible values.",
    correct = TRUE,
    message = "Every value from 48.1 to 55.9 is plausible
    given this evidence, and 50 is one of them. The interval
    says more than the test does --- it names the range ---
    but on the reject-or-not question the two always agree."
  ),
  answer(
    "Reject the null, because the observed mean of 52 is
    above 50.",
    message = "The observed mean is always somewhere; the
    question is whether it is far enough from 50 to be
    surprising. Here it is not."
  ),
  answer(
    "Fail to reject, and conclude that the mean democracy
    score is 50.",
    message = "Failing to reject leaves 50 plausible; it does
    not single it out. Every other value in the interval is
    equally uneliminated."
  ),
  answer(
    "Fail to reject, but since most of the interval lies
    above 50, the data still show the mean is greater
    than 50.",
    message = "Where the interval sits is not a second test.
    If 50 is inside it, the evidence does not rule 50 out, and
    no part of the interval's position changes that."
  ),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Where does 50 fall relative to the interval, and what does that tell you about whether the evidence rules 50 out?"
)
```

## The Takeaways

This tutorial covered the two main tools of statistical inference. Here is what to carry forward.

**Two tools, one foundation.** Both hypothesis tests and confidence intervals are built on the same machinery from Tutorial 10: the sampling distribution, the standard error, and the CLT. They simply answer different questions. A [hypothesis test]{.important-text} asks: could this result have arisen if the null were true? The [p-value]{.important-text} quantifies how surprising your result would be if the null hypothesis were true. A [confidence interval]{.important-text} asks: what range of values for the population parameter is consistent with my data? For a two-sided test at a [significance level]{.important-text} of $\alpha = 0.05$, using the same model and assumptions, the hypothesis test and the [95% confidence interval]{.important-text} reach the same reject/fail-to-reject conclusion about the null value.

**How a confidence interval is built.** Start at the point estimate and go out in both directions by the [margin of error]{.important-text}, which is the [critical value]{.important-text} times the standard error. The critical value comes from the [confidence level]{.important-text} you choose --- 1.96 for 95%, 2.576 for 99% --- so raising the confidence level widens the interval. That is the trade-off at the heart of confidence intervals: higher confidence requires a wider interval --- you ask the procedure to capture the true parameter more often, so it must extend farther from the estimate, and which you want depends on what it costs to be wrong.

**Four things to remember when interpreting results.**

- A p-value is not the probability the null hypothesis is true --- it is the probability of observing your result, or something more extreme, if the null were true.
- Failing to reject the null is not the same as accepting it; it means the data are not sufficiently surprising under the null, not that the null is proven correct.
- A 95% confidence interval does not mean there is a 95% probability the true value is in this particular interval; it means 95% of intervals constructed this way will contain the true value across repeated samples.
- Statistical significance and practical importance are not the same thing --- a result can be [statistically significant]{.important-text} but substantively trivial, or practically important but not yet significant with the sample size available.


**When complete data still needs inference.** Political science data often cover every case in a defined population --- all 50 states, all members of a Congress, all countries in a year --- and nothing was sampled. Whether inference applies depends on the question, not on how the data were collected. A purely descriptive question about those cases at that moment needs none: you have the answer. But when you ask whether one thing predicts another, you are asking about a [relationship]{.important-text} --- a systematic pattern that could have come out differently under different conditions --- and inference is what lets you ask how compatible the observed pattern is with a specified null model for that broader process.

**Organizing what you've learned into five steps.** Every hypothesis test you have just learned about --- and every one you will see in the tutorials that follow --- uses the same five pieces: a [null hypothesis]{.important-text}, an [alternative hypothesis]{.important-text}, a test statistic, a p-value, and a decision based on comparing that p-value to a threshold. Starting in the next tutorial, those pieces are organized into five explicit steps: state $H_0$, state $H_A$, run the test, find the p-value, and decide. The ideas are not new --- you already have them --- but the five-step structure will make the same logic easier to apply consistently, test after test.


The logic developed across Tutorials 10 and 11 --- construct a [null distribution]{.important-text}, locate your observed result, compute the probability --- underlies every hypothesis test in the remaining tutorials. What changes is the specific distribution used. For means with large samples, the normal distribution applies. For means with small samples, the [t-distribution]{.important-text} is used instead, with slightly wider critical values to account for additional uncertainty. Tutorial 12 introduces the **chi-square distribution** for relationships between two categorical variables, Tutorial 13 brings in both the t-distribution and F for comparing group means, and Tutorial 14 extends the same logic to correlation. In every case R handles the distribution automatically. Your job is to interpret the result correctly --- and you now have the tools to do that.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-11, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-11, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-11, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Terrific work! You can now apply hypothesis tests and confidence intervals, the core tools of statistical inference.">
  <span aria-hidden="true">🎉</span>
  Terrific work! You can now apply hypothesis tests and confidence intervals, the core tools of statistical inference.
</p>
')
```
