---
title: "What Does Your Model Predict? Regression Predictions and Interpretation"
tutorial:
  id: "18-Regression-Predictions"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn to extract substantive meaning from regression models using predicted values and prediction plots.
---

## Overview {#main-content}

Regression output can feel both precise and opaque at the same time. The numbers are right there --- coefficients, standard errors, significance stars --- but a slope of 0.876 does not feel like an answer to a political question. Understanding comes when you can ask: if two states differed only in the proportion of their residents who identify as liberal, how different would their state legislatures look? If a country with few political rights opened up its economy to trade, how much better would its environmental performance be predicted to get? These are *what-if* questions --- and answering them is exactly what regression is designed for.

A predicted value is not a forecast about the future. It is the model's answer to a specific what-if: what would the outcome be for a case with *these particular values* of the predictors? The difference between two such predictions --- holding everything else constant except the variable you care about --- is the model's estimate of that variable's effect, stated as a concrete comparison rather than an abstract number.

Once you learn to make these comparisons well, regression output stops being a table of coefficients and starts being a tool for understanding the world. That is what this tutorial teaches --- and it is a skill that transfers to every model you will ever estimate.

In this tutorial you will learn:

* What a predicted value from a regression model means --- not a forecast, but a what-if comparison: what the model says the outcome would be for a specific combination of predictor values, and what comparing two such predictions reveals about an estimated effect
* How to obtain specific predicted values using `predict()` and how to visualize predicted values across a range using `plot_model(type = "pred")`, including how to vary two predictors at once and how to choose which variable goes on the x-axis
* How to interpret any predicted value comparison --- whether a pair of numbers or a full prediction plot --- using a systematic five-component framework

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
library(stargazer)
library(sjPlot)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(error = TRUE)
data("states", package = "QPATutorialsCourse")
legdata <- states
legdata$p.liberal <- legdata$liberal / 100
legdata$region_censusf <- factor(legdata$region_census,
                                 labels = c("South", "Northeast", "Midwest", "West"))
data("qog", package = "QPATutorialsCourse")
qog$rights[qog$fh_pr <= 2] <- 0
qog$rights[qog$fh_pr >  2] <- 1
qog$rights_f <- factor(qog$rights, labels = c("Few Rights", "Many Rights"))
fit1 <- lm(hs_dem_prop_all ~ nonwhite + p.liberal + povrate + region_censusf,
           data = legdata)
model_env <- lm(epi_epi ~ rights_f * ipi_tradeopen + hf_efiscore, data = qog)
```

## From Model to Meaning: Choosing and Interpreting Predictions

You have spent the last three tutorials learning to estimate models, interpret their coefficients, and test hypotheses. That work produces a table of numbers. This section is about what comes next: turning those numbers into answers to the questions that motivated your analysis in the first place. A coefficient tells you the direction and size of an effect in abstract units. A predicted value tells you what the model says the outcome would actually be for a specific case --- and how different that outcome would be if one thing changed.

The hardest part of using predicted values is not the mechanics. It is knowing what question to ask.

When you have a model with several predictors, which one should you vary? The answer comes from the research question, not the software. You vary the predictor whose effect you are trying to communicate --- usually the one your hypothesis is about. If the hypothesis concerns ideology, you produce predictions across the range of ideology. If an interaction is the central finding, you vary both interacting variables so the reader can see how one variable's effect changes across levels of the other. Everything else gets held constant, and you tell your reader what values you used.

In this tutorial you will use two complementary tools. `predict()` answers one what-if question at a time precisely: you specify a complete set of predictor values, and it returns the model's predicted outcome for that case. Running it for two carefully chosen cases gives you concrete, legible numbers in the units of the outcome. These numbers anchor your understanding of what the model is saying. `plot_model(type = "pred")` then shows many such predictions at once, across the full observed range of a predictor, with the uncertainty visible as a shaded band around the line or points. If you have already computed two predicted values for a specific comparison, you know exactly what the y-axis means when you see those values on a plot. The two tools work together --- the numbers anchor your interpretation, and the plot fills in the full picture.

Once you have made a comparison --- whether two numbers from `predict()` or a full prediction plot --- a complete interpretation covers five things:

1. **The pattern.** What does the comparison show overall? Is the relationship positive or negative, steep or shallow, consistent across the range or concentrated at one end?

2. **The held values.** Every predicted value assumes specific values for all predictors not being varied. State them. In an additive linear regression model (one without interactions involving the focal predictor), changing the held values shifts the predicted line up or down but does not change its slope --- the effect of the predictor you are varying is the same regardless of the baseline you choose. When the model includes an interaction involving the focal predictor, that invariance no longer holds, and the choice of held values changes the slope as well as the level. The conventional defaults --- means for interval variables, reference category for factor variables --- describe a typical case.

3. **Specific predicted values.** At least two. "The predicted proportion increases" is weaker than "the predicted proportion increases from 0.58 to 0.67 as the liberal proportion rises from 0.25 to 0.35." Magnitude is part of the finding.

4. **The uncertainty.** For a pair of specific predictions, adding `interval = "confidence"` to your `predict()` call returns lower and upper bounds alongside each point estimate. In a prediction plot, the same uncertainty appears as a shaded band around the line. A wide band --- especially one wide enough for the bands of two groups to overlap --- means the data cannot distinguish the outcomes as clearly as the point estimates suggest.

5. **The connection to the hypothesis.** A predicted value comparison is evidence. Does the pattern support the hypothesis, fail to support it, or complicate it?

Here is what a complete interpretation looks like written as a single paragraph. Suppose a researcher models county voter turnout as a function of median household income, unemployment rate, and urban/rural status, with urban counties as the reference category. She produces a prediction plot showing predicted turnout across the range of income for urban counties at average unemployment. A complete interpretation would read: *As median household income increases, predicted voter turnout rises --- the relationship is positive and linear across the observed range. Predictions hold the unemployment rate at its mean and assume urban counties; a rural county at the same income level would be predicted to have lower turnout by about 4 percentage points, but the income slope would be the same. At $30,000 median income, the model predicts about 54% turnout; at $60,000, about 67% --- a 13-point difference across a realistic range. The confidence band is moderately narrow throughout, reflecting a reasonably large sample. This positive relationship is consistent with the hypothesis that wealthier communities have more resources for civic participation.*

Notice that all five components appear naturally in the prose without being labeled. That is the standard this tutorial asks you to meet.

### The Workflow

Everything in Tutorials 12 through 17 has been building toward this moment: you have a model, you have tested it, and now you need to make it speak.

The workflow this tutorial follows has three steps, and the order matters. We begin with a **regression table** --- not because tables are exciting, but because they are honest. Every coefficient is there, with its standard error and its significance. Any reader who wants to verify the analysis or build on it starts here. More importantly, you start here, because the table tells you which coefficients are worth visualizing.

We then ask the model for **specific predicted values** --- a what-if comparison for two carefully chosen cases. A coefficient of 0.876 says the slope is positive; a predicted value of 0.58 at p.liberal = 0.25 and 0.67 at p.liberal = 0.35 says what that slope actually means for the gap between a moderately liberal and a more liberal state. Before you draw a line, you know two of its points.

Finally, we **visualize**. The prediction plot sweeps across the full range of the predictor, showing every prediction at once, with the uncertainty visible as a shaded band. By the time you see the plot, you already understand what you are looking at --- because you just computed it by hand for two specific cases.

Follow this sequence for both examples in this tutorial. It will become second nature.

## What Predicts Democratic Representation in State Legislatures?

State legislatures vary dramatically in their partisan composition. One obvious driver is the ideological makeup of the state's voters --- states with more liberal public opinion should elect more Democratic legislators. But ideology is not the only force: racial composition, poverty, and regional political culture all shape legislative outcomes too. The model we estimate below is new to this tutorial; our focus is not on interpreting every coefficient but on what the model predicts for specific comparisons and across the full range of the key predictors.

::: {.hypothesis}
In a comparison of states, those with more liberal public opinion will have a higher proportion of Democratic state house seats than those with less liberal public opinion, after controlling for racial composition, poverty rate, and census region.
:::

The data frame **legdata** contains 49 state observations with the following variables:

- **hs_dem_prop_all**: proportion of state house seats held by Democrats
- **p.liberal**: proportion of state residents identifying as liberal
- **nonwhite**: proportion of state residents who are not white
- **povrate**: state poverty rate (percent)
- **region_censusf**: a four-category factor --- South (reference), Northeast, Midwest, West

### Estimate the Model

`lm()`, `summary()`, and `predict()` are part of base R --- no `library()` call is needed for those functions.

**Practice: Estimate the regression model**

Save your model as **fit1** and present the results. Three hints are available. Click Submit Answer to check your work.

```{r exercise-fit1-estimate, exercise=TRUE, exercise.cap="Practice: Estimate the state legislature model and save as fit1"}

```

```{r exercise-fit1-estimate-hint-1}
# Hint 1 of 3
# You need to do three things: specify the regression formula
# (dependent variable and predictors), save the model to an
# object named fit1, and display the results. Think about which
# variable is the outcome we are trying to explain and which
# four variables are the predictors listed in the variable
# descriptions above.
```

```{r exercise-fit1-estimate-hint-2}
# Hint 2 of 3
# Use lm() with hs_dem_prop_all as the dependent variable and
# nonwhite, p.liberal, povrate, and region_censusf as predictors.
# Save to fit1 and call summary(fit1).
```

```{r exercise-fit1-estimate-hint-3}
# Hint 3 of 3: fill in the blanks
fit1 <- lm(___ ~ ___ + ___ + ___ + ___, data = legdata)
summary(fit1)
```

```{r exercise-fit1-estimate-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("lm\\s*\\(", code_nc))
    fail("Use lm() to estimate the model.")
  if (!grepl("fit1\\s*<-", code_nc))
    fail("Save your model as fit1.")
  if (!grepl("hs_dem_prop_all", code_nc))
    fail("hs_dem_prop_all should be the dependent variable.")
  if (!grepl("p\\.liberal", code_nc))
    fail("Include p.liberal as a predictor.")
  if (!grepl("nonwhite", code_nc))
    fail("Include nonwhite as a predictor.")
  if (!grepl("povrate", code_nc))
    fail("Include povrate as a predictor.")
  if (!grepl("region_censusf", code_nc))
    fail("Include region_censusf as a predictor.")
  if (!grepl("summary\\s*\\(", code_nc))
    fail("Call summary() on fit1 to display the results.")
  expected <- lm(hs_dem_prop_all ~ nonwhite + p.liberal + povrate + region_censusf,
                 data = legdata)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef))
    fail("Make sure you display the results by calling summary() on fit1.")
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6)))
    fail("Check your formula. The model should regress hs_dem_prop_all on nonwhite, p.liberal, povrate, and region_censusf using the legdata data frame.")
  pass("Right. Only two predictors are significant at .05: p.liberal and the West dummy. Keep that in mind as you interpret the predictions that follow.")
})
```

### Present the Model in a Table

Let's begin with the table. You have used `stargazer()` before --- the same arguments apply here; see Tutorial 16 if you need a refresher on the syntax.

**Practice: Produce a stargazer table for fit1**

The prescribed title and dependent variable label are:
- `title = "Predictors of Democratic State House Seat Share"`
- `dep.var.labels = "Proportion of Democratic House Seats"`

The six covariate labels for this model are: **Proportion Non-White**, **Proportion Liberal**, **Poverty Rate**, **Northeast**, **Midwest**, and **West**. Determine the correct order from your model output above. Three hints are available. Click Submit Answer to check your work.

```{r exercise-stargazer-fit1, exercise=TRUE, exercise.cap="Practice: Produce a labeled stargazer table for fit1"}
library(stargazer)
stargazer(fit1,
          ___ = "text",
          title = "___",
          dep.var.labels = "___",
          single.row = ___,
          ___ = c("Proportion Non-White",
                  "___",
                  "___",
                  "___",
                  "___",
                  "___"))
```

```{r exercise-stargazer-fit1-hint-1}
# Hint 1 of 3
# stargazer() needs to know the format of its output, a title,
# a label for the dependent variable column, and labels for each
# predictor row. Two of the blanks are argument names --- think
# about which argument controls what format the output appears in
# and which argument supplies the row labels for predictors.
```

```{r exercise-stargazer-fit1-hint-2}
# Hint 2 of 3
# The two argument name blanks are: type and covariate.labels.
# single.row takes TRUE.
# Covariate labels follow formula order; South is the reference
# category and is omitted, so region labels are Northeast, Midwest, West.
```

```{r exercise-stargazer-fit1-hint-3}
# Hint 3 of 3: fill in the blanks
stargazer(fit1,
          type = "text",
          title = "Predictors of Democratic State House Seat Share",
          dep.var.labels = "Proportion of Democratic House Seats",
          single.row = TRUE,
          covariate.labels = c("Proportion Non-White",
                               "Proportion Liberal",
                               "Poverty Rate",
                               "___",
                               "___",
                               "___"))
```

```{r exercise-stargazer-fit1-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("stargazer\\s*\\(", code_nc))
    fail("Use stargazer() to produce the table.")
  if (!grepl("type\\s*=", code_nc))
    fail("Include the type = argument.")
  if (!grepl("covariate\\.labels\\s*=", code_nc))
    fail("Include the covariate.labels = argument.")
  if (!grepl("Predictors of Democratic State House Seat Share", code_nc, fixed = TRUE))
    fail("Use the prescribed title.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use the prescribed dep.var.labels.")
  if (!grepl("single\\.row\\s*=\\s*TRUE", code_nc))
    fail("Set single.row = TRUE.")
  labels <- c("Proportion Non-White","Proportion Liberal","Poverty Rate",
              "Northeast","Midwest","West")
  for (lbl in labels)
    if (!grepl(lbl, code_nc, fixed = TRUE))
      fail(paste0("The label \"", lbl, "\" is missing."))
  pos <- sapply(labels, function(l) regexpr(l, code_nc, fixed=TRUE)[1])
  if (!all(diff(pos) > 0))
    fail("The labels are out of order. Follow formula order: Proportion Non-White, Proportion Liberal, Poverty Rate, Northeast, Midwest, West.")
  pass("Well done. The table gives your reader the complete picture; the predictions and plots that follow translate the significant coefficients into what-if comparisons.")
})
```

Once a package is loaded with `library()`, all of its functions are available for the rest of your R session --- you do not need to load it again each time you use a function from it. Since [stargazer]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial.

### Obtain Specific Predicted Values

The table shows that the West coefficient is positive and significant. But what does that mean concretely --- how much more Democratic representation does being a Western state predict, all else equal? `predict()` answers that question directly. It is doing exactly what you have done by hand in earlier tutorials: substituting specific values of the predictors into the fitted regression equation and computing the result. The difference is that R does the arithmetic for you and returns the answer in the units of the dependent variable, ready to compare and interpret.

The code below asks that question precisely, computing predictions for two otherwise identical states that differ only in region. Run it first, then we will unpack how it works.

**Run and observe: Compute predicted Democratic seat share for two specific states**

```{r exercise-predict-fit1, exercise=TRUE, exercise.cap="Run and observe: Compute two specific predicted values from fit1"}
# A Southern state at average nonwhite and poverty, with p.liberal = 0.25
predict(fit1,
        newdata = data.frame(
          p.liberal      = 0.25,
          nonwhite       = mean(legdata$nonwhite, na.rm = TRUE),
          povrate        = mean(legdata$povrate,  na.rm = TRUE),
          region_censusf = factor("South", levels = levels(legdata$region_censusf))
        ))

# Same characteristics, Western state
predict(fit1,
        newdata = data.frame(
          p.liberal      = 0.25,
          nonwhite       = mean(legdata$nonwhite, na.rm = TRUE),
          povrate        = mean(legdata$povrate,  na.rm = TRUE),
          region_censusf = factor("West", levels = levels(legdata$region_censusf))
        ))
```

**How this code works.** `predict()` takes the model object and a `newdata` argument --- a data frame describing the exact case you want to predict. Every predictor in the model must appear in `newdata`, named and typed exactly as in the formula. Here that means four variables: `p.liberal`, `nonwhite`, and `povrate` set to specific values, and `region_censusf`. For a factor variable, writing `"South"` as a bare string is not enough --- R needs to know that "South" is a level of a specific factor with specific other levels, or it cannot match the variable to what the model expects. The construction `factor("South", levels = levels(legdata$region_censusf))` creates a one-element factor with the correct levels attached. The output is in the same units as the dependent variable: the proportion of Democratic house seats, on a 0-to-1 scale.

**Practice: Interpret the difference**

The two outputs differ in exactly one way --- the region. Everything else is held constant. That difference is the West coefficient made concrete.

```{r quiz-predict-fit1, echo=FALSE}
question("Look at both outputs and identify which type of state is predicted to have more Democratic seats, and by approximately how much. What does that gap tell you?",
  correct = "Right. Western states are predicted to have substantially more Democratic house seats than otherwise identical Southern states --- the region coefficient made concrete.",
  answer("All else equal, Western states are predicted to have about 13 percentage points more Democratic house seats than Southern states with the same liberalism, racial composition, and poverty rate",
         correct = TRUE,
         message = "The West coefficient is 0.128, expressed here as a concrete what-if comparison. Rather than saying 'the West dummy is 0.128,' you can say 'a Western state at these values is predicted to have about 13 more percentage points of Democratic seats than an otherwise identical Southern state.' That is a claim your reader can evaluate without knowing what a regression coefficient is."),
  answer("Western states are more liberal than Southern states, which is why they have higher predicted Democratic seat shares",
         message = "Both predictions hold p.liberal at the same value (0.25). Ideological differences are not driving the gap --- this is the region coefficient, net of ideology, racial composition, and poverty."),
  answer("The predictions tell us that 58% of voters in Southern states and 71% of voters in Western states support Democrats",
         message = "The outcome variable is the proportion of state house seats held by Democrats, not the proportion of voters. And these are model predictions, not observed proportions."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at both predicted values and compute the difference. Then think about what held constant in both calls."
)
```

### Visualize the Effect of State Liberalism

The predict() calls answered one specific what-if at one value of p.liberal. The liberalism hypothesis calls for something richer: how does Democratic seat share change across the full observed range of p.liberal? `plot_model(type = "pred")` from the [sjPlot]{.package-name} package answers that. The key arguments:

- **`type = "pred"`** requests predicted values across a range, as opposed to `type = "est"` for a coefficient plot
- **`terms = "variable"`** names the predictor to vary on the x-axis; all others are held at their means (interval) or reference category (factor)
- **`axis.title = c("x label", "y label")`** labels the axes --- x-axis first, then y-axis (the reverse of what ggplot users sometimes expect)
- **`show.data = TRUE/FALSE`** overlays the raw observations; useful when N is small enough for individual points to be readable
- **`labs(caption = "...")`** adds a caption stating the held values; since `plot_model()` returns a ggplot object, any ggplot layer works with `+`

Run the code below and examine the output --- the direction of the line, the shaded band, the individual points, and the caption --- before reading the explanation that follows.

**Run and observe: Produce a prediction plot for state liberalism**

```{r exercise-predplot-liberal-tryit, exercise=TRUE, exercise.cap="Run and observe: Produce a prediction plot for p.liberal"}
library(sjPlot)
plot_model(fit1,
           type = "pred",
           terms = "p.liberal",
           title = "Predicted Democratic House Seat Share by State Liberalism",
           axis.title = c("Proportion Liberal",
                          "Proportion of Democratic House Seats"),
           show.data = TRUE) +
  labs(caption = "Predictions are for Southern states with average poverty rate and proportion non-white.")
```

Since [sjPlot]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial. In your own scripts, call `library(sjPlot)` once at the top before using `plot_model()`.

Notice what this plot is doing: it is sweeping across the full range of p.liberal, computing a predicted value at every point --- exactly what `predict()` did for one specific value, but repeated across the entire range and displayed as a continuous line. The two specific values you computed earlier (about 0.58 and 0.71) are points on this line. The shaded band is the 95% confidence region around those predictions. Its width is not constant: it is narrower in the middle of the p.liberal range, where more states are concentrated and the model has more information, and wider at the tails, where fewer states have very high or very low liberalism levels. This is a general property --- predictions are always more precise where data are dense and less precise where data are sparse.

### Interpret the Liberalism Plot

The five questions below walk through each component of the interpretation framework. The prediction plot is reproduced here for reference.

```{r fit1-predplot-ref, echo=FALSE, message=FALSE}
plot_model(fit1,
           type = "pred",
           terms = "p.liberal",
           title = "Predicted Democratic House Seat Share by State Liberalism",
           axis.title = c("Proportion Liberal",
                          "Proportion of Democratic House Seats"),
           show.data = FALSE) +
  labs(caption = "Predictions are for Southern states with average poverty rate and proportion non-white.")
```

**Practice: Component 1 --- The pattern**

Look at the direction and shape of the line before answering.

```{r quiz-liberal-pattern, echo=FALSE}
question("Look at the prediction plot above. Which statement best describes the overall pattern?",
  correct = "Right. The relationship is positive and exactly linear --- the line slopes upward from left to right.",
  answer("As the proportion of liberal residents increases, the predicted proportion of Democratic house seats increases --- the relationship is positive and exactly linear",
         correct = TRUE,
         message = "The upward slope matches the hypothesis. In a linear regression model without polynomial terms, the prediction line is a straight line by construction --- the relationship is exactly, not approximately, linear."),
  answer("As the proportion of liberal residents increases, the predicted proportion of Democratic house seats decreases",
         message = "The line slopes upward from left to right --- a positive relationship."),
  answer("There is no clear pattern --- the line is essentially flat",
         message = "The line has a clearly positive slope from left to right."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: trace the line from left to right --- does the y-axis value increase, decrease, or stay flat?"
)
```

**Practice: Component 2 --- The held values**

The caption tells you what was held constant. Think about what that means for how the slope generalizes.

```{r quiz-liberal-held, echo=FALSE}
question("The caption states that predictions are for Southern states with average poverty rate and proportion non-white. What does this mean for interpreting the slope?",
  correct = "Right. The slope is invariant to the baseline --- only the level of the line changes.",
  answer("The slope is the same regardless of which region or values of poverty and nonwhite we hold constant --- changing the baseline shifts the line up or down but does not change its steepness",
         correct = TRUE,
         message = "This is a key property of linear regression without interactions: the slope on p.liberal is the same at any baseline. The caption tells you which level of the line you are looking at, not which slope."),
  answer("The slope only applies to Southern states --- the liberalism effect is completely different in other regions",
         message = "Because this model does not include an interaction between p.liberal and region_censusf, the slope is identical in every region. Regional differences shift the line up or down but do not change its slope."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what an interaction term would be needed to make the slope differ across groups."
)
```

**Practice: Component 3 --- Specific predicted values**

Read two values from the plot --- one at each end of a realistic range --- and compute the difference.

```{r quiz-liberal-values, echo=FALSE}
question("Using the prediction plot above, approximately what is the predicted Democratic seat share at p.liberal = 0.25 and at p.liberal = 0.35 for a Southern state at average poverty and nonwhite values?",
  correct = "Right. Reading from the plot: about 0.58 at p.liberal = 0.25 and about 0.67 at p.liberal = 0.35 --- a 9-percentage-point difference.",
  answer("About 0.58 at p.liberal = 0.25 and about 0.67 at p.liberal = 0.35 --- a 9-percentage-point increase for a 10-point rise in the liberal proportion",
         correct = TRUE,
         message = "The slope on p.liberal is 0.876, so a 0.10 increase predicts about 0.088 more Democratic seats --- roughly 9 percentage points. You could verify this by plugging these values into the regression equation by hand; the plot makes it visible at a glance."),
  answer("About 0.25 at p.liberal = 0.25 and about 0.35 at p.liberal = 0.35",
         message = "Read the y-axis value at each x position, not the x-axis value itself."),
  answer("About 0.50 at both values --- the line is flat",
         message = "The line is not flat. Trace from x = 0.25 up to the line and read the y-axis value; repeat at x = 0.35."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find x = 0.25 on the horizontal axis, trace up to the line, and read the y-axis value. Repeat at x = 0.35."
)
```

**Practice: Component 4 --- The uncertainty**

Look at the width of the shaded band and whether it is consistent across the range or varies.

```{r quiz-liberal-band, echo=FALSE}
question("Look at the shaded confidence band in the plot above. Which statement best describes what it tells you?",
  correct = "Right. The band is moderately wide and widens at the extremes --- honest about the limits of a 49-state sample.",
  answer("The band is moderately wide and widens at the extremes, reflecting fewer states at very high or very low liberalism levels --- but it is consistent with a positive slope throughout",
         correct = TRUE,
         message = "With only 49 states, there is meaningful uncertainty around each prediction. The widening at the tails is expected. The band never crosses in a way that would reverse the directional conclusion."),
  answer("The band is so narrow that predictions can be read to three decimal places",
         message = "With 49 states, the band is visible and meaningful."),
  answer("The band is so wide that no conclusions can be drawn about direction",
         message = "The band limits precision but not directional conclusions. It is consistent with a positive slope throughout."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the width of the shaded band and whether it is wider at the tails than in the middle."
)
```

**Practice: Component 5 --- The hypothesis connection**

A coefficient and p-value tell you the relationship is statistically significant. The plot tells you something the table cannot.

```{r quiz-liberal-hypothesis, echo=FALSE}
question("Look at the p.liberal coefficient and its p-value in your output. What does the prediction plot add to that conclusion?",
  correct = "Right. The plot translates the significant coefficient into a visible, concrete comparison.",
  answer("The plot makes the substantive meaning visible --- showing that the predicted difference between a less liberal and more liberal state is about 9 percentage points across a realistic range, not just that the slope is positive and significant",
         correct = TRUE,
         message = "A significant p-value tells you the relationship is statistically distinguishable from zero. The plot tells you what it looks like in the units and range that matter for your research question."),
  answer("The plot adds nothing beyond the p-value",
         message = "The p-value tells you the relationship is unlikely to be chance. The plot tells you its magnitude across a realistic range --- substantively different information."),
  answer("The plot proves liberal voters cause Democratic victories",
         message = "A prediction plot is evidence of an association, not proof of causation."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a p-value tells you and what it cannot tell you about the size or practical meaning of an effect."
)
```

### Visualize the Effect of Census Region

The predict() calls showed a regional gap. A factor prediction plot extends that comparison to all four regions simultaneously, using the same held values.

**Practice: Produce a prediction plot for census region**

Produce the plot and examine it carefully before answering the interpretation question that follows.

Prescribed axis labels and caption:
- x-axis: `"Census Region"`
- y-axis: `"Proportion of Democratic House Seats"`
- caption: `"Predictions are for states with average poverty rate, proportion liberal, and proportion non-white."`

Three hints are available. Click Submit Answer to check your work.

```{r exercise-predplot-region, exercise=TRUE, exercise.cap="Practice: Produce a prediction plot for region_censusf"}
___(fit1,
    ___ = "pred",
    ___ = "region_censusf",
    title = "Predicted Democratic House Seat Share by Census Region",
    axis.title = c("___", "___")) +
  labs(caption = "___")
```

```{r exercise-predplot-region-hint-1}
# Hint 1 of 3
# plot_model() can produce different kinds of output depending on
# one argument. Another argument controls which variable appears
# on the x-axis. Both were filled in for you in the liberalism
# Try it above --- look at that code and think about what values
# you need here for a factor variable instead of an interval one.
```

```{r exercise-predplot-region-hint-2}
# Hint 2 of 3
# The argument that controls what kind of plot you get is type =.
# For a prediction plot, its value is "pred".
# The argument that controls the x-axis variable is terms =.
# Set it to the name of the factor variable, in quotes.
# Prescribed axis labels: x = "Census Region", y = "Proportion of Democratic House Seats"
# Prescribed caption: "Predictions are for states with average poverty rate,
#   proportion liberal, and proportion non-white."
```

```{r exercise-predplot-region-hint-3}
# Hint 3 of 3: fill in the blanks
plot_model(fit1,
           type = "pred",
           terms = "region_censusf",
           title = "Predicted Democratic House Seat Share by Census Region",
           axis.title = c("Census Region",
                          "Proportion of Democratic House Seats")) +
  labs(caption = "___")
```

```{r exercise-predplot-region-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("region_censusf", code_nc))
    fail("Set terms = \"region_censusf\".")
  if (!grepl("Census Region", code_nc, fixed = TRUE))
    fail("Use \"Census Region\" as the x-axis label.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use \"Proportion of Democratic House Seats\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = \"...\") layer stating the held values.")
  if (!grepl("Predictions are for states with average poverty rate, proportion liberal, and proportion non-white.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with average poverty rate, proportion liberal, and proportion non-white.\"")
  pass("Well done. When terms is a factor, plot_model() produces dots with confidence intervals rather than a line.")
})
```

When `terms` is a factor, `plot_model()` produces dots with confidence intervals rather than a continuous line --- there is no meaningful range to sweep across for a categorical variable, only discrete levels. Overlapping confidence intervals are a useful visual cue that two groups may not be statistically distinguishable, though overlap is a rough heuristic rather than a formal test of the difference.

**Practice: Interpret the regional plot**

Look at which dots stand apart and whether their confidence intervals overlap with South before answering.

```{r quiz-region-interpret, echo=FALSE}
question("Look at the regional prediction plot you just produced. Which statement best describes the key finding?",
  correct = "Right. The West stands out; the other three regions are statistically indistinguishable from South.",
  answer("Western states are predicted to have substantially higher Democratic seat shares than Southern states, and this is the only regional difference that is statistically significant --- the Northeast and Midwest confidence intervals overlap substantially with South",
         correct = TRUE,
         message = "The West coefficient (0.128, p = .047) is the only significant regional effect. Northeast (p = .797) and Midwest (p = .945) are not reliably different from South once ideology, poverty, and racial composition are held constant."),
  answer("All four regions are predicted to have essentially equal Democratic seat shares",
         message = "The West dot is clearly above the other three. Look at its confidence interval relative to South."),
  answer("The Northeast has the highest predicted Democratic seat share because it is the most liberal region",
         message = "The model controls for p.liberal, so the regional differences shown here are net of ideology. The Northeast is not significantly different from South in this model."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at which dot is highest and whether its confidence interval overlaps with the South interval."
)
```

### Visualize Two Predictors at Once

The liberalism and region plots each varied one predictor. A two-variable plot extends this: it shows whether predicted Democratic seat share is systematically higher or lower in some regions than others after accounting for liberalism, and whether the predicted effect of liberalism on seat share is the same size across regions.

Setting `terms` to a vector of two variable names produces this plot: the first variable goes on the x-axis, and the second variable's levels appear as separate colored lines.

**Practice: Produce a two-variable prediction plot**

Produce the plot and notice whether the four lines are parallel --- that parallelism is a finding about the model, not just a visual feature.

Prescribed values:
- `title = "Predicted Democratic House Seat Share by Liberalism and Region"`
- x-axis: `"Proportion Liberal"`, y-axis: `"Proportion of Democratic House Seats"`
- caption: `"Predictions are for states with average poverty rate and proportion non-white."`

Four hints are available. Click Submit Answer to check your work.

```{r exercise-predplot-twovar, exercise=TRUE, exercise.cap="Practice: Produce a two-variable prediction plot for p.liberal and region_censusf"}
___(fit1,
    ___ = "pred",
    ___ = c("p.liberal", "region_censusf"),
    title = "___",
    ___ = c("___", "___")) +
  labs(caption = "___")
```

```{r exercise-predplot-twovar-hint-1}
# Hint 1 of 4
# You have produced prediction plots for one variable at a time.
# This plot varies two variables simultaneously. Think about which
# variable should be on the x-axis (the continuous predictor your
# hypothesis is about) and which should produce separate colored
# lines (the grouping variable). The only change in the code
# structure is how the terms argument is specified.
```

```{r exercise-predplot-twovar-hint-2}
# Hint 2 of 4
# The blank argument names are: the function name (plot_model),
# type =, terms =, and axis.title =.
# terms takes a character vector of two variable names.
```

```{r exercise-predplot-twovar-hint-3}
# Hint 3 of 4
# Prescribed title: "Predicted Democratic House Seat Share by Liberalism and Region"
# Prescribed axis labels: x = "Proportion Liberal", y = "Proportion of Democratic House Seats"
# Prescribed caption: "Predictions are for states with average poverty rate and proportion non-white."
```

```{r exercise-predplot-twovar-hint-4}
# Hint 4 of 4: fill in the blanks
plot_model(fit1,
           type = "pred",
           terms = c("p.liberal", "region_censusf"),
           title = "Predicted Democratic House Seat Share by Liberalism and Region",
           axis.title = c("Proportion Liberal",
                          "Proportion of Democratic House Seats")) +
  labs(caption = "___")
```

```{r exercise-predplot-twovar-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("p\\.liberal", code_nc))
    fail("Include p.liberal as the first term (x-axis).")
  if (!grepl("region_censusf", code_nc))
    fail("Include region_censusf as the second term (separate colored lines).")
  if (regexpr("p\\.liberal", code_nc) > regexpr("region_censusf", code_nc))
    fail("p.liberal should come first in terms (x-axis), and region_censusf second (separate colored lines). Check the order inside c(...).")
  if (!grepl("Predicted Democratic House Seat Share by Liberalism and Region", code_nc, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Democratic House Seat Share by Liberalism and Region\".")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Proportion Liberal", code_nc, fixed = TRUE))
    fail("Use \"Proportion Liberal\" as the x-axis label.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use \"Proportion of Democratic House Seats\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = \"...\") layer.")
  if (!grepl("Predictions are for states with average poverty rate and proportion non-white.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with average poverty rate and proportion non-white.\"")
  pass("Well done. Look at whether the lines are parallel and what that tells you about the interaction between liberalism and region.")
})
```

**Practice: Interpret the two-variable plot**

The lines are parallel. Think about what that geometric fact tells you about the model before answering.

```{r quiz-twovar-interpret, echo=FALSE}
question("The four lines in the two-variable plot are parallel. What does this tell you about the relationship between liberalism and Democratic seats?",
  correct = "Right. Parallel lines mean the liberalism effect is the same in every region --- the visual signature of an additive model.",
  answer("The liberalism effect is the same in every region --- the parallel lines confirm there is no interaction between p.liberal and region_censusf in this model",
         correct = TRUE,
         message = "Parallel lines are the visual signature of an additive model. Each region shifts the line up or down by a fixed amount (the region coefficient), but the slope of liberalism is constant. If we wanted to test whether the liberalism effect differs by region, we would need to add an interaction term."),
  answer("The parallel lines mean liberalism has no effect on Democratic seat share",
         message = "Parallel lines mean the effect is the same across groups, not that there is no effect. The lines all slope upward --- liberalism has a clear positive effect in every region."),
  answer("The parallel lines mean the regional differences are not statistically significant",
         message = "Parallelism tells you about interactions, not about whether the group differences are significant. The vertical gaps between lines are the region coefficients, which may or may not be significant separately."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: parallel lines mean the x-axis variable has the same slope in every group. Non-parallel lines would indicate an interaction."
)
```

The four lines are parallel. That is not a coincidence --- it is what a model without an interaction produces. Because there is no interaction term between p.liberal and region_censusf in this model, the slope of liberalism is mathematically identical in every region. The vertical distances between lines are the region coefficients: the West line runs consistently above the others, and South, Northeast, and Midwest cluster closely together. Parallel lines tell you the effect of the x-axis variable does not depend on which group you are in.

### What We Learned About Ideology, Region, and Democratic Representation

The model confirms the hypothesis: states with more liberal public opinion do have higher proportions of Democratic state house seats, even after controlling for racial composition, poverty, and region. A 10-percentage-point increase in the liberal share of the electorate predicts about 9 more percentage points of Democratic seats --- a substantive effect, visible in the slope of the prediction plot, and significant (p = .013).

The regional finding is perhaps more surprising. The West stands apart from the other three regions even after controlling for ideology. The Northeast and Midwest, despite their reputations for Democratic strength, are not significantly different from the South once ideology and demographics are controlled. What makes the West distinct in this model is not just a more liberal electorate --- the regional advantage persists after liberalism is held constant.

Here is what a complete interpretation of the liberalism prediction plot looks like written as prose: *As the proportion of liberal state residents increases, predicted Democratic house seat share rises in a straight line across the full observed range of the variable. Predictions are for Southern states with average poverty rate and proportion non-white; a Western state at the same liberalism level would be predicted to have about 13 more percentage points of Democratic seats, but the liberalism slope would be identical. At a liberal proportion of 0.25, the model predicts about 58% Democratic seats; at 0.35, about 67% --- a 9-point difference for a 10-point shift in the liberal electorate. The confidence band is moderately wide, narrower in the middle of the range where most states are concentrated, and widens at the extremes. This is consistent with the hypothesis that public ideology shapes legislative representation, even after controlling for racial composition, poverty, and region.*

## Does Political Rights Status Condition the Effect of Trade Openness on Environmental Performance?

In Tutorial 17 you estimated this model and interpreted its interaction coefficients. It is available here as **model_env**.

```{r model-env-table, echo=TRUE}
stargazer(model_env,
          type = "text",
          title = "Trade Openness, Political Rights, and Environmental Performance",
          dep.var.labels = "Environmental Performance Index",
          single.row = TRUE,
          covariate.labels = c("Many Rights (vs. Few Rights)",
                               "Trade Openness",
                               "Economic Freedom Score",
                               "Many Rights x Trade Openness"))
```

In Tutorial 17 you found that **ipi_tradeopen** has a large positive effect on **epi_epi** in countries with few political rights (slope ≈ 4.75, p < .001), but near-zero effect in countries with many political rights (combined slope ≈ 0.12, not significant). The interaction coefficient is negative and significant (−4.63, p < .001), confirming that the **ipi_tradeopen** effect depends on **rights_f**. The economic freedom score (**hf_efiscore**) is also a significant positive predictor. The model explains about 60% of the variation in environmental performance across 116 countries.

Our focus here is not on revisiting those coefficients but on what the model predicts for specific cases and how those predictions look when visualized across the full range of trade openness.

::: {.hypothesis}
The effect of trade openness on environmental performance is stronger in countries with few political rights than in countries with many political rights --- because in rights-restricted countries, international economic integration may be one of the few channels through which global environmental norms reach domestic policy.
:::

The data frame **qog** contains the following variables used in this model:

- **epi_epi**: Environmental Performance Index (27--87, higher = better)
- **rights_f**: two-category factor --- Few Rights (reference) and Many Rights
- **ipi_tradeopen**: trade openness index (1--10, higher = more open)
- **hf_efiscore**: economic freedom score (control variable)

### Obtain Specific Predicted Values

Two comparisons are worth making here, and each requires two `predict()` calls --- four in total.

**Comparison 1** asks: how much does **epi_epi** differ between Few Rights and Many Rights countries at average trade openness? This is the most direct test of whether **rights_f** matters, holding the level of economic integration constant.

**Comparison 2** asks: within Few Rights countries specifically, how much does **ipi_tradeopen** matter? Moving from low to high trade openness within the rights-restricted group captures the steepness of the Few Rights slope --- the core of the interaction hypothesis.

Calls 1 and 2 below are scaffolded with blanks for you to fill in. Calls 3 and 4 are yours to write from scratch.

**Practice: Compute four predicted values**

These four predictions anchor everything that follows --- once you have them, you know exactly what the interaction plots are showing before you see them.

Three hints are available. Click Submit Answer to check your work.

```{r exercise-predict-env, exercise=TRUE, exercise.cap="Practice: Compute four predicted values from model_env"}
# Comparison 1: Few Rights vs. Many Rights at average trade openness
# Call 1: Few Rights
predict(model_env, newdata = data.frame(
  rights_f      = factor(___, levels = levels(qog$rights_f)),
  ipi_tradeopen = ___,
  hf_efiscore   = ___
))
# Call 2: Many Rights
predict(model_env, newdata = data.frame(
  rights_f      = factor(___, levels = levels(qog$rights_f)),
  ipi_tradeopen = ___,
  hf_efiscore   = ___
))

# Comparison 2: Few Rights at low vs. high trade openness
# Mean economic freedom score ≈ 60.26
# Write Calls 3 and 4 from scratch here

```

```{r exercise-predict-env-hint-1}
# Hint 1 of 3
# For Comparison 1, exactly one thing changes between Call 1 and Call 2:
# the rights category. Trade openness and economic freedom stay at the
# same values for both calls. Think about which factor level goes in
# each call and what "average" means for the two interval variables.
# For Comparison 2, the rights category stays the same (Few Rights)
# while trade openness changes between the two calls.
```

```{r exercise-predict-env-hint-2}
# Hint 2 of 3
# The factor level names are "Few Rights" and "Many Rights".
# Use: factor("Few Rights", levels = levels(qog$rights_f))
# Mean ipi_tradeopen ≈ 7.77; mean hf_efiscore ≈ 60.26
# For Comparison 2: use ipi_tradeopen = 6.9 (low, 25th percentile)
#   and ipi_tradeopen = 9.4 (high, 75th percentile), both Few Rights.
```

```{r exercise-predict-env-hint-3}
# Hint 3 of 3: skeleton for Calls 3 and 4
predict(model_env, newdata = data.frame(
  rights_f      = factor(___, levels = levels(qog$rights_f)),
  ipi_tradeopen = ___,
  hf_efiscore   = ___
))
predict(model_env, newdata = data.frame(
  rights_f      = factor(___, levels = levels(qog$rights_f)),
  ipi_tradeopen = ___,
  hf_efiscore   = ___
))
```

```{r exercise-predict-env-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("predict\\s*\\(", code_nc))
    fail("Use predict() to obtain predicted values.")
  if (!grepl("model_env", code_nc))
    fail("Pass model_env as the model object.")
  if (!grepl("rights_f", code_nc))
    fail("Include rights_f in your newdata data frame.")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Include ipi_tradeopen in your newdata data frame.")
  if (!grepl("hf_efiscore", code_nc))
    fail("Include hf_efiscore in your newdata data frame.")
  if (!grepl("Few Rights", code_nc, fixed = TRUE))
    fail("Include a prediction for Few Rights countries.")
  if (!grepl("Many Rights", code_nc, fixed = TRUE))
    fail("Include a prediction for Many Rights countries.")
  n_predict_calls <- lengths(regmatches(code_nc, gregexpr("predict\\s*\\(", code_nc)))
  if (n_predict_calls < 4)
    fail("You need four predict() calls: two for Comparison 1 (Few Rights vs. Many Rights at mean trade openness) and two for Comparison 2 (Few Rights at low vs. high trade openness).")
  if (!grepl("7\\.77|mean\\s*\\(\\s*qog\\s*\\$\\s*ipi_tradeopen", code_nc))
    fail("Comparison 1 should use mean trade openness (≈ 7.77) for both calls. Check the value you supplied for ipi_tradeopen in Calls 1 and 2.")
  if (!grepl("6\\.9", code_nc))
    fail("Comparison 2 should include a low trade openness value (6.9, the 25th percentile). Check Calls 3 and 4.")
  if (!grepl("9\\.4", code_nc))
    fail("Comparison 2 should include a high trade openness value (9.4, the 75th percentile). Check Calls 3 and 4.")
  pass("Right. At mean trade openness, Few Rights countries (≈59.5) are predicted to slightly outperform Many Rights countries (≈52.7) --- because at this level of trade openness, the large positive trade slope for Few Rights countries overcomes the rights disadvantage. Within Few Rights countries, moving from low (6.9) to high (9.4) trade openness predicts a jump of about 12 points (55.3 to 67.2) --- a substantial effect of trade in the rights-restricted context.")
})
```

**Practice: Interpret the what-if comparisons**

You now have four numbers. Read them together --- Comparison 1 tells you who leads at average trade, Comparison 2 tells you how much the Few Rights slope matters.

```{r quiz-predict-env, echo=FALSE}
question("Look at your four predicted values. Comparison 1 shows how Few Rights and Many Rights countries differ at average trade openness. Comparison 2 shows how much trade openness matters within Few Rights countries. What do these two comparisons together suggest about the role of trade openness in rights-restricted countries?",
  correct = "Right. Trade openness is doing most of the work for Few Rights countries at mean and above levels of trade.",
  answer("Trade openness is doing most of the work for Few Rights countries --- the steep positive slope means that at average or above-average trade openness, the trade effect is large enough to offset the rights disadvantage entirely",
         correct = TRUE,
         message = "This is the interaction story made concrete. At low trade openness, Many Rights countries outperform Few Rights countries. But as trade openness rises, Few Rights countries close the gap and eventually overtake them --- because trade is one of the few forces pushing environmental standards in rights-restricted contexts."),
  answer("Many Rights countries always perform better environmentally because political accountability matters more than trade",
         message = "At mean and high trade openness, the model actually predicts Few Rights countries to outperform Many Rights countries. The advantage of political rights is largest when trade openness is low."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: compare Comparison 1 (who leads at average trade?) to Comparison 2 (how steep is the Few Rights slope?). What happens as trade openness rises?"
)
```

### Visualize the Effect of Trade Openness

Before visualizing the interaction, let's look at the trade openness effect alone --- what does the model predict for Few Rights countries (the reference category) across the full range of trade openness?

**Practice: Produce a prediction plot for trade openness**

This plot shows the Few Rights slope in full --- the baseline you will need to compare against when the interaction plot appears.

Three hints are available. Click Submit Answer to check your work.

```{r exercise-predplot-trade, exercise=TRUE, exercise.cap="Practice: Produce a prediction plot for ipi_tradeopen"}
___(model_env,
    ___ = "pred",
    ___ = "ipi_tradeopen",
    ___ = "Predicted Environmental Performance by Trade Openness",
    ___ = c("Trade Openness (1-10)",
            "Environmental Performance Index"),
    show.data = FALSE) +
  ___(caption = "Predictions are for Few Rights countries with average economic freedom.")
```

```{r exercise-predplot-trade-hint-1}
# Hint 1 of 3
# You have produced prediction plots in Example 1. This exercise
# follows the same structure with a different model and variable.
# Think about which argument controls the type of plot, which
# controls the x-axis variable, and which labels the axes.
# All argument names are blank --- you need to supply them.
```

```{r exercise-predplot-trade-hint-2}
# Hint 2 of 3
# The blank argument names are: the function name (plot_model),
# type =, terms =, title =, axis.title =, and the labs argument name.
# For type =, a prediction plot uses the value "pred".
# For terms =, supply the variable name in quotes.
```

```{r exercise-predplot-trade-hint-3}
# Hint 3 of 3: fill in the blanks
plot_model(model_env,
           type = "pred",
           terms = "ipi_tradeopen",
           title = "Predicted Environmental Performance by Trade Openness",
           axis.title = c("Trade Openness (1-10)",
                          "Environmental Performance Index"),
           show.data = FALSE) +
  labs(caption = "___")
```

```{r exercise-predplot-trade-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Set terms = \"ipi_tradeopen\".")
  if (!grepl("title\\s*=", code_nc))
    fail("Include the title = argument.")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = ...) layer.")
  pass("Well done. This plot shows the steep positive trade openness slope for Few Rights countries --- the baseline from which the interaction plot will show how things change for Many Rights countries.")
})
```

**Practice: Interpret the trade openness plot**

This is the Few Rights slope in full. Read the y-axis at the low and high ends of trade openness before answering.

```{r quiz-trade-interpret, echo=FALSE}
question("The prediction plot shows a steep upward slope for Few Rights countries. Which statement best completes the interpretation?",
  correct = "Right. A very large positive effect of trade openness within the Few Rights baseline.",
  answer("For Few Rights countries at average economic freedom, moving from the lowest to highest trade openness predicts a very large improvement on the Environmental Performance Index",
         correct = TRUE,
         message = "The trade openness coefficient for Few Rights countries is 4.75 per unit, and the scale runs from 1 to 10 --- a 9-unit range predicts 4.75 × 9 ≈ 43 points of improvement. This is exactly the steep positive slope the interaction hypothesis predicted for rights-restricted countries."),
  answer("The upward slope means trade openness causes better environmental outcomes in Few Rights countries",
         message = "A significant positive coefficient is evidence of an association, not proof of causation. Trade openness may be correlated with other factors that also improve environmental performance."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: read the predicted values at the low and high ends of the trade openness scale and compute the difference."
)
```

### Visualize the Interaction: Rights Status Across the Range of Trade Openness

Now we look at the interaction directly. The first plot puts rights status on the x-axis, with separate lines for values of trade openness. Run the code and examine the output before reading the explanation.

**Practice: Produce the interaction plot with rights_f on the x-axis**

This orientation lets you compare the two groups at specific trade openness values --- useful, but as you will see, not the whole story.

Three hints are available. Click Submit Answer to check your work.

```{r exercise-predplot-int1, exercise=TRUE, exercise.cap="Practice: Produce an interaction plot with rights_f on the x-axis"}
___(model_env,
    ___ = "pred",
    ___ = c("rights_f", "ipi_tradeopen"),
    ___ = "Environmental Performance by Rights Status",
    ___ = c("Political Rights Status",
            "Environmental Performance Index")) +
  ___(caption = "Predictions are for countries with average economic freedom.")
```

```{r exercise-predplot-int1-hint-1}
# Hint 1 of 3
# This is a two-variable prediction plot --- the same structure as the
# two-variable plot you produced in Example 1, but now the first variable
# in terms is a factor. Think about which argument controls what kind of
# plot you get, which argument takes a vector of two variable names,
# and which argument labels the two axes.
```

```{r exercise-predplot-int1-hint-2}
# Hint 2 of 3
# The blank argument names are: the function name (plot_model),
# type =, terms =, title =, axis.title =, and the labs argument name.
# The first variable in terms goes on the x-axis;
# the second produces separate colored lines or points.
```

```{r exercise-predplot-int1-hint-3}
# Hint 3 of 3: fill in the blanks
plot_model(model_env,
           type = "pred",
           terms = c("rights_f", "ipi_tradeopen"),
           title = "Environmental Performance by Rights Status",
           axis.title = c("Political Rights Status",
                          "Environmental Performance Index")) +
  labs(caption = "___")
```

```{r exercise-predplot-int1-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model().")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("rights_f", code_nc))
    fail("Include rights_f in terms.")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Include ipi_tradeopen in terms.")
  if (regexpr("rights_f", code_nc) > regexpr("ipi_tradeopen", code_nc))
    fail("rights_f should come first in terms (x-axis), and ipi_tradeopen second (separate lines). Check the order inside c(...).")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = ...) layer.")
  pass("Well done. Now look at what this plot can and cannot show before answering the question.")
})
```

**Practice: Interpret this orientation of the interaction plot**

Think about what you can read from three discrete points and what you cannot see between them.

```{r quiz-int1-interpret, echo=FALSE}
question("The plot shows two dots (Few Rights and Many Rights) with confidence intervals, with separate lines for different trade openness values. What can you see from this orientation and what is harder to see?",
  correct = "Right. You can see the gap at each trade level, but not how the gap changes continuously across the full range.",
  answer("You can see that at low trade openness Many Rights countries outperform Few Rights countries, and at high trade openness the gap reverses --- but the plot makes it hard to see how the gap changes continuously across the full range of trade openness",
         correct = TRUE,
         message = "This orientation answers the question 'which rights category performs better at a given trade level?' but it cannot show the smooth evolution of the trade openness effect across its full range. Three points give you snapshots; a continuous line gives you the whole story."),
  answer("This is the most informative orientation because it shows both variables at once",
         message = "It shows both variables, but only at three specific trade openness values. The next plot will show the full continuous range of trade openness, which is more informative for understanding how the effect changes."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what you can see at the three specific trade openness values shown, and what you cannot see between them."
)
```

When **rights_f** is on the x-axis, you see two dots at three selected trade openness levels. This is useful for comparing Few Rights to Many Rights at specific values, but it makes it hard to see how the relationship between trade openness and environmental performance evolves continuously. For that, we need trade openness on the x-axis.

### Visualize the Interaction: Trade Openness Across the Full Range

With **ipi_tradeopen** on the x-axis and **rights_f** producing separate lines, we can see the entire conditional relationship: one line for Few Rights countries, one for Many Rights. Where the lines are not parallel, the interaction is visible directly.

**Practice: Produce the interaction plot with ipi_tradeopen on the x-axis**

Prescribed values:
- `title = "Predicted Environmental Performance by Trade Openness and Rights Status"`
- x-axis: `"Trade Openness (1-10)"`, y-axis: `"Environmental Performance Index"`
- caption: `"Predictions are for countries with average economic freedom."`

Four hints are available. Click Submit Answer to check your work.

```{r exercise-predplot-int2, exercise=TRUE, exercise.cap="Practice: Produce an interaction plot with ipi_tradeopen on the x-axis"}

```

```{r exercise-predplot-int2-hint-1}
# Hint 1 of 4
# This is the same two-variable plot as the previous exercise,
# but with the variables swapped: the interval variable goes on
# the x-axis and the factor variable produces separate colored lines.
# Think about why this orientation tells a richer story than the
# previous one, and which variable belongs in which position in terms.
```

```{r exercise-predplot-int2-hint-2}
# Hint 2 of 4
# Use plot_model() with type = "pred".
# terms takes a vector with ipi_tradeopen first (x-axis)
# and rights_f second (separate colored lines).
```

```{r exercise-predplot-int2-hint-3}
# Hint 3 of 4
# Prescribed title: "Predicted Environmental Performance by Trade Openness and Rights Status"
# Prescribed axis labels: x = "Trade Openness (1-10)", y = "Environmental Performance Index"
# Prescribed caption: "Predictions are for countries with average economic freedom."
# Add the caption with + labs(caption = "...")
```

```{r exercise-predplot-int2-hint-4}
# Hint 4 of 4: fill in the blanks
plot_model(model_env,
           type = "pred",
           terms = c("___", "___"),
           title = "___",
           axis.title = c("___", "___")) +
  labs(caption = "___")
```

```{r exercise-predplot-int2-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#.*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model().")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Include ipi_tradeopen in terms (as the first variable --- x-axis).")
  if (!grepl("rights_f", code_nc))
    fail("Include rights_f in terms (as the second variable --- separate lines).")
  if (regexpr("ipi_tradeopen", code_nc) > regexpr("rights_f", code_nc))
    fail("ipi_tradeopen should come first in terms (x-axis), and rights_f second (separate lines). Check the order inside c(...).")
  if (!grepl("title\\s*=", code_nc))
    fail("Include a title = argument.")
  if (!grepl("Predicted Environmental Performance by Trade Openness and Rights Status", code_nc, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Environmental Performance by Trade Openness and Rights Status\".")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include an axis.title = argument.")
  if (!grepl("Trade Openness \\(1-10\\)", code_nc))
    fail("Use \"Trade Openness (1-10)\" as the x-axis label.")
  if (!grepl("Environmental Performance Index", code_nc, fixed = TRUE))
    fail("Use \"Environmental Performance Index\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = ...) layer.")
  if (!grepl("Predictions are for countries with average economic freedom.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for countries with average economic freedom.\"")
  pass("Well done. Now look carefully at whether the two lines are parallel and where they cross.")
})
```

**Practice: Interpret the interaction plot. Select all that apply.**

Look at the two lines carefully --- their slopes, whether they are parallel, and where they cross --- before answering.

```{r quiz-int2-interpret, echo=FALSE}
question("Look at the interaction plot with trade openness on the x-axis. Which statements correctly describe what it shows? Select all that apply.",
  correct = "Right. Non-parallel lines, a crossover, and a near-flat slope for Many Rights countries.",
  answer("The two lines are not parallel --- this is the visual signature of the interaction: the effect of trade openness on environmental performance differs between Few Rights and Many Rights countries",
         correct = TRUE,
         message = "Non-parallel lines are exactly what an interaction looks like in a prediction plot. If the lines were parallel, the interaction term would be zero."),
  answer("The Few Rights line has a steep positive slope, while the Many Rights line is nearly flat --- confirming that trade openness predicts large improvements in Few Rights countries but almost no change in Many Rights countries",
         correct = TRUE,
         message = "The Few Rights slope is 4.75 per unit of trade openness; the Many Rights slope is approximately 0.12. This is the interaction finding made visible: trade openness matters enormously in rights-restricted contexts and barely at all in rights-protected ones."),
  answer("The lines cross at around a trade openness value of 6, meaning that below this value Many Rights countries are predicted to perform better, but above it Few Rights countries take the lead",
         correct = TRUE,
         message = "The crossover at approximately 6.3 on the trade openness scale is where the two predicted values are equal. Below it, political rights provide the advantage; above it, the steep trade slope in Few Rights countries closes and then reverses the gap."),
  answer("Because adjusted R-squared is below 0.65, the model's predictions are not reliable",
         message = "R-squared tells you how much variance the model explains; the value here is actually strong for cross-national data. It has nothing to do with whether the predictions or the interaction are reliable."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements. Look at the slope of each line, whether they are parallel, and whether they cross."
)
```

### What We Learned About Political Rights, Trade, and the Environment

The interaction plot reveals something that no table of coefficients can show directly: the relationship between **ipi_tradeopen** and **epi_epi** is not one relationship --- it is two, depending on the political context. In countries with few political rights, trade openness is a powerful predictor of better environmental outcomes, with a steep positive slope across the full range of the variable. In countries with many political rights, trade openness barely moves the needle --- the slope is near zero, because political accountability through elections and civil society provides environmental pressure regardless of economic integration.

The crossover at approximately **ipi_tradeopen** = 6.3 is a substantive finding in its own right: it tells you the tipping point where the trade advantage in Few Rights countries overtakes the institutional advantage of Many Rights countries. Below that threshold, political rights matter more; above it, trade integration does the work that political accountability would otherwise provide.

Here is what a complete interpretation of the interaction prediction plot looks like as prose: *As trade openness increases, predicted environmental performance rises steeply in Few Rights countries and barely changes in Many Rights countries. Predictions hold the economic freedom score at its mean throughout. At a trade openness score of 5 --- below the crossover --- Few Rights countries are predicted to score about 46 on the Environmental Performance Index while Many Rights countries are predicted to score about 52; Few Rights are behind. At a score of 9, Few Rights countries are predicted to reach about 65 while Many Rights countries remain near 53 --- Few Rights have overtaken them by a wide margin. The confidence bands for the two groups overlap near the middle of the trade openness range, where distinguishing between them is hardest, and diverge toward the extremes. This is consistent with the hypothesis: trade openness matters for environmental performance primarily in countries where political rights alone provide fewer accountability mechanisms.*

This is the kind of finding that prediction plots were designed to reveal. The interaction coefficient (−4.63) told you the sign and size; the plot tells you what that means for the world.

## Check Your Understanding

The exercises below test whether you can apply the five-component framework to output you have not seen before. A researcher models life expectancy (in years) across countries as a function of GDP per capita (in $10,000s), proportion of the population with access to basic healthcare (0 to 1), and world region (a four-category factor with Africa as the reference category: Africa, Americas, Asia, Europe). The model is significant, adjusted R-squared = 0.72. A prediction plot shows predicted life expectancy (y-axis, range 50--80 years) as a function of GDP per capita (x-axis, range \$2,000--\$60,000), holding healthcare access at its mean and assuming African countries. The prediction line rises from about 56 years at GDP = \$2,000 to about 72 years at GDP = \$30,000. The confidence band is moderately wide throughout.

**Apply: Apply the five-component framework**

Read the scenario description above carefully, then evaluate each response against all five components of a complete interpretation.

```{r quiz-cyu-fivecomponent, echo=FALSE}
question("Which response constitutes the most complete five-component interpretation of this prediction plot?",
  correct = "Right. All five components are addressed: pattern, held values, specific values, uncertainty, and hypothesis connection.",
  answer("Predicted life expectancy increases from about 56 years at low GDP to about 72 years at higher GDP --- a 16-year difference. Predictions hold healthcare access at its mean and assume African countries. The confidence band is moderately wide. This is consistent with the hypothesis that wealthier countries have higher life expectancy.",
         correct = TRUE,
         message = "This covers all five: (1) positive pattern, (2) held values stated with the specific baseline named, (3) two specific values with magnitude, (4) the confidence band assessed, (5) connection to hypothesis. This is the standard for a complete interpretation."),
  answer("As GDP increases, predicted life expectancy increases. The effect is statistically significant.",
         message = "This names the pattern and notes significance, but is missing three components: the held values, at least two specific predicted values to convey magnitude, and an assessment of the confidence band."),
  answer("Predicted life expectancy rises from about 56 to 72 years as GDP increases. The confidence band is moderately wide.",
         message = "This gives specific values and notes the band, but is missing the held values (healthcare at mean, African countries) and the connection to the research hypothesis."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: check each response against all five components: (1) pattern, (2) held values, (3) specific predicted values, (4) uncertainty, (5) hypothesis connection."
)
```

**Apply: Interpret the held values**

This question tests Component 2 --- whether you understand what the caption means for the slope and how it generalizes.

```{r quiz-cyu-heldvalues, echo=FALSE}
question("The prediction plot holds healthcare access at its mean and assumes African countries. A student says this means the plot only applies to Africa. How would you correct this?",
  correct = "Right. The slope applies everywhere in an additive model --- only the level of the line changes by baseline.",
  answer("In a linear regression without an interaction between GDP and region, the slope of GDP on life expectancy is the same for every region --- changing the baseline from Africa to Europe would shift the line up by the Europe coefficient but would not change its slope",
         correct = TRUE,
         message = "This is the key invariance property of additive linear regression. The caption tells you which level of the line you are looking at. The slope --- the what-if comparison the plot communicates --- applies equally at any baseline you might have chosen."),
  answer("The student is right --- the plot only describes African countries, and a separate plot would be needed for each region",
         message = "A separate plot for each region would have the same slope, just shifted up or down. The slope on GDP is the same everywhere in a model without a GDP-by-region interaction."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what an interaction between GDP and region would be needed to make the slope differ across regions."
)
```

**Apply: Interpret non-parallel lines**

The researcher has now added an interaction. Think about what non-parallel lines in a prediction plot are telling you before choosing.

```{r quiz-cyu-nonparallel, echo=FALSE}
question("The researcher adds an interaction between GDP and region and produces a new two-variable prediction plot with GDP on the x-axis and separate lines for each region. The lines are not parallel --- the Europe line is steep, the Africa line is shallow. What does this mean?",
  correct = "Right. The GDP effect is conditional on region --- non-parallel lines are the visual signature of an interaction.",
  answer("The effect of GDP on life expectancy differs by region --- the return to economic growth for life expectancy is larger in Europe than in Africa, and a model without the interaction would have obscured this difference",
         correct = TRUE,
         message = "Non-parallel lines are the visual signature of an interaction. The steeper Europe line means each additional unit of GDP predicts a larger increase in life expectancy in Europe than in Africa. This is exactly the kind of finding that prediction plots reveal that coefficient tables can only hint at."),
  answer("Non-parallel lines mean the model is misspecified and should be reestimated without the interaction",
         message = "Non-parallel lines are the expected output when an interaction term is present and significant. They are a finding, not a problem."),
  answer("Non-parallel lines mean that Europe has higher life expectancy than Africa, which is already known",
         message = "The vertical gap between lines at any given GDP value reflects the region coefficient --- the level difference. Non-parallel lines tell you something additional: the slope of GDP itself differs by region. That is a different claim from saying the levels differ."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: non-parallel lines mean the x-axis variable has a different slope in different groups. What does that tell you about the interaction?"
)
```

## The Takeaways

Prediction plots are the most powerful presentation tool in a regression analyst's toolkit because they do something tables cannot: they translate abstract coefficients into concrete predictions about what the model says will happen for specific, meaningful cases. A table tells you the slope is 0.876; a prediction plot shows you that moving from a 25% to a 45% liberal electorate predicts about 18 more percentage points of Democratic seats. Those are the same mathematical fact, but the second one is the one that sticks.

The key things to carry forward:

1. **The workflow is: table → specific predictions → plot.** The table shows the full model for readers who want to verify or build on your analysis. The `predict()` calls give you two specific numbers to anchor your interpretation before you see the full picture. The plot shows how predicted outcomes change across the full range of the predictor, with uncertainty visible.

2. **`predict()` requires a `newdata` data frame with every predictor named and typed exactly as in the model.** For factor variables, use `factor("level", levels = levels(data$variable))` --- a bare string will not work. The output is in the units of the dependent variable.

3. **`plot_model(type = "pred", terms = "variable")` sweeps across the full range of the specified predictor**, holding all others at their means or reference category. `axis.title` takes x first, then y. Always add `labs(caption = "...")` to state the held values.

4. **`terms = c("x", "grouping_variable")` shows conditional relationships.** When there is no interaction, the lines are parallel --- the x-variable has the same slope in every group. When there is an interaction, the lines are not parallel, and the plot makes visible exactly what the interaction coefficient only hints at.

5. **A complete interpretation covers five things:** the pattern, the held values, specific predicted values, the confidence band, and the connection to the hypothesis. Each component adds something the others do not.

The choice of which variable to put on the x-axis matters. When one of your two variables is categorical and the other is interval, put the interval variable on the x-axis. A continuous line across the full range of the interval variable is more informative than a few dots at selected values of the other variable.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below.

```{r encode-logic-18, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-18, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-18, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now translate regression coefficients into concrete what-if comparisons using predicted values and prediction plots.">
  <span aria-hidden="true">📊</span>
  Excellent work! You can now translate regression coefficients into concrete what-if comparisons --- and that changes what it means to understand a model.
</p>
')
```
