---
title: "Logistic Regression: Visualizing Predicted Probabilities"
tutorial:
  id: "21-Logistic-Regression-Predictions"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Prediction plots on the probability scale, where what you hold constant changes the shape.
---

## Overview {#main-content}

How much more likely is a country to ratify a treaty once its neighbors have? Does another year of schooling raise the chance that someone votes as much for a person already likely to turn out as for a person who rarely does? Questions like these ask what the probability of an outcome is and how far it moves --- and a logistic regression coefficient, reported in log-odds, answers neither. The size of the change has to be computed, and it is not one number: the same coefficient shifts the probability a great deal for some cases and hardly at all for others.

This tutorial shows how to turn a fitted logistic model into prediction plots that trace the predicted probability across the full range of a predictor, and how the values you hold the other predictors at decide not only where that curve sits but how steep it is --- a choice that carries far more weight here than it did for linear models. It uses the models fitted in Tutorial 20 and the prediction plots introduced in Tutorial 17.

In this tutorial you will learn:

* How to use `plot_model()` from the [sjPlot]{.package-name} package to produce prediction plots that show the S-shaped relationship between a predictor and the probability of the outcome
* How to read the message `plot_model()` prints when it simplifies a curve, and how bracket notation makes it draw the curve at every observed value
* How to use the `condition` argument to specify what values other predictors are held at when producing a prediction plot, and why that choice matters more in logistic regression than in linear regression
* How to produce prediction plots for factor predictors, showing predicted probabilities for each category with confidence intervals
* How to produce two-variable prediction plots that compare the predicted relationship across categories of another predictor, even when the model contains no interaction
* How to interpret prediction plots using the five-component framework introduced in Tutorial 17

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(gradethis)
library(sjPlot)
tutorial_options(exercise.checker = gradethis::grade_learnr, exercise.completion = FALSE)
options(lifecycle_verbosity = "quiet")
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "QPATutorialsCourse")
counties$rural_cat <- factor(counties$rural,
                             labels = c("Urban", "Intermediate", "Rural"))
model_trump <- glm(TrumpMajority ~ wage_growth + prop_college_grad +
                     prop_white + mortality_risk_25_45 + rural_cat,
                   family = "binomial", data = counties)
data("statesPolicy", package = "QPATutorialsCourse")
statesPolicy$gov_f <- factor(statesPolicy$party_governor2024)
statesPolicy$gov_f <- relevel(statesPolicy$gov_f, ref = "Democrat")
model_ghg <- glm(policy_ghgcap2024 ~ ideology_conservative2024 +
                   gov_f + econ_poverty2022,
                 family = "binomial", data = statesPolicy)
```

## From Coefficients to Probabilities: Why Visualization Matters More Here

In logistic regression, holding other variables constant is not enough to pin down an effect. The change in the predicted probability when a predictor shifts by one unit depends on which specific values you are holding those other variables at --- and on where those values place you on the S-curve.

The open carry example from Tutorial 20 makes this concrete. Moving from 15% to 20% of residents identifying as Republican --- a five-point increase --- raises the predicted probability of an open carry law by about 2 percentage points. Moving from 25% to 30% --- the same five-point increase --- raises it by about 62 percentage points. Moving from 35% to 40% --- again the same shift --- produces almost nothing, because the probability is already near certainty. The coefficient on Republican share cannot tell you any of those three numbers. It can only tell you the direction is positive and the relationship is statistically significant. The size of the effect depends entirely on where on the S-curve the comparison falls.

This is why prediction plots are particularly valuable for logistic regression in a way they are not for linear regression. To communicate what a coefficient means in substantive terms, you choose specific values for all the other explanatory variables, compute the predicted probability as the variable of interest moves across its range, and show the resulting curve. And because the shape of that curve --- how steep it is, where its steep middle falls, how quickly it flattens --- depends on what you hold constant in the background, those choices are not arbitrary technical details. They determine what the plot is actually showing. A prediction plot always represents the relationship between one explanatory variable and the outcome in a specific context, and naming that context is part of interpreting the result.

Both models come from Tutorial 20 and are used here as they were fitted there. Tutorial 20 gives the substantive rationale for each predictor; what follows is enough to read the plots.

The **county model** predicts **TrumpMajority**, whether a county gave Trump a majority of the two-party vote, from **wage_growth** (percent change in average weekly wages), **prop_college_grad** (proportion of adults 25 and older with a bachelor's degree, on a 0-to-1 scale), **prop_white** (proportion of residents who are white), **mortality_risk_25_45** (percentage risk of dying between 25 and 45) and **rural_cat** (Urban, Intermediate or Rural). It describes counties as of 2016, apart from **wage_growth**, which measures change over the twelve months ending in the third quarter of that year. **TrumpMajority** and **rural_cat** have no missing values; **prop_college_grad**, **prop_white** and **mortality_risk_25_45** are each missing for one county and **wage_growth** for two, so the model is estimated on 3,109 of the 3,112 counties.

The **state model** predicts **policy_ghgcap2024**, whether a state has adopted a greenhouse gas emissions cap, from **ideology_conservative2024** (percentage of residents identifying as conservative), **gov_f** (the governor's party, with Democrat as the reference) and **econ_poverty2022** (the state poverty rate). Its variables come from different years --- the emissions cap and the governor's party as of 2024, conservative identification from the 2024 PRRI American Values Atlas, the poverty rate from 2022. Only **ideology_conservative2024** has a missing value, for the District of Columbia, so the model is estimated on the 50 states.

## The Five-Component Framework

Knowing why prediction plots are particularly valuable here is half the work. Reading them well requires a systematic approach, and choosing what to put in them requires judgment. Tutorial 17 introduced a five-component framework that applies to every prediction plot regardless of what model produced it; we use it throughout this tutorial.

1. **The pattern** --- the direction and shape of the relationship: does the predicted outcome rise, fall, or curve as the explanatory variable increases?
2. **The held values** --- what the other explanatory variables were set to when the predictions were computed, and what that means for how the result generalizes
3. **Specific predicted values** --- what the model predicts at substantively meaningful values of the variable on the x-axis
4. **The uncertainty** --- how wide the confidence band is and whether it is consistent or varies across the range
5. **The hypothesis connection** --- whether the pattern is consistent with the hypothesis that motivated the model, and what the plot adds to the significance test

## The Three Tools: `predict()`, `plot_model()` and `condition`

Tutorial 20 introduced two tools for working with predicted probabilities. `predict()` with `type = "response"` answers a single what-if question: what is the predicted probability for a case with this exact combination of predictor values? `plot_model()` with `type = "pred"` sweeps across the full range of one predictor and shows how the predicted probability changes --- it is `predict()` applied at every point along a variable's range, turned into a plot.

Both tools have to hold the other explanatory variables at some specific values while the variable of interest varies. For `predict()`, you supply those values explicitly in `newdata` --- Tutorial 20 shows that every explanatory variable must appear in the data frame, with a specific value you chose. For `plot_model()`, if you do not specify otherwise, the other predictors are held at their means (for interval variables) or their reference category (for factor variables).

The values the other predictors are fixed at while the x-axis variable varies are the plot's [held values]{.important-text}, and the `condition` argument to `plot_model()` lets you override the defaults and specify exactly what context the plot represents. It works the same way `newdata` did with `predict()` in Tutorial 20: you bundle one or more named values together, where each name matches an explanatory variable in the model and each value is what you want it held at. The bundling tool here is `list()` rather than `data.frame()` --- `condition = list(prop_white = 0.85)` holds the proportion white at 0.85 rather than its mean while the x-axis variable varies across its range. You can set multiple variables at once: `condition = list(prop_white = 0.85, rural_cat = "Rural")`.

In an additive linear regression model (one without interactions involving the predictor being displayed), changing the held values shifts the predicted line up or down without changing its slope --- so the choice of held values affects the level of the prediction but not its shape. This is why Tutorial 17 rarely needed `condition`: any reasonable set of held values told essentially the same story. In logistic regression, the held values determine which part of the S-curve the plot is showing, which can change both the level and the steepness of the relationship. Choosing those values thoughtfully is therefore part of the analysis, not a technical detail.

When using `condition`, the most useful held values are ones that are **substantively meaningful** --- corresponding to real, recognizable cases rather than arbitrary numbers --- and that your research question is actually about. Values placing the predictions in a different region of the S-curve are often informative, but choose them because they represent theoretically relevant or commonly occurring cases, not because they make the plots look different. A curve that barely moves when you change the held values is a finding in its own right --- it says the estimated relationship holds across those contexts. Set `condition` to reflect the context your claim concerns rather than accepting whatever the software chooses by default.

## What Was Associated with Trump Majorities in 2016?

The 2016 election produced one of the sharpest educational divides in modern presidential voting history. Counties with lower college graduation rates shifted dramatically toward Trump relative to 2012, while highly educated counties moved toward Clinton. Scholars debate whether this reflects economic anxiety, cultural change, or the declining salience of traditional class politics --- but the empirical pattern is among the most robust findings about that election. The model from Tutorial 20 captures this: after controlling for wage growth, mortality risk, racial composition, and rurality, college graduation rates are associated with a large difference in the predicted probability that a county gave Trump a majority. Visualizing that relationship does not sidestep the retrospective voting hypothesis --- it shows what the model implies across the range of counties actually observed.

The model output below recalls the coefficients for reference. Check the direction and significance of each predictor before producing any plots --- the plots should confirm what the table already tells you about direction, and add what the table cannot tell you about magnitude.

```{r model-trump-recall, echo=TRUE, comment=NA}
summary(model_trump)
```

The workflow here parallels Tutorial 17: we anchor the interpretation with specific predicted probabilities before producing plots, so that when the S-curve appears you already know what two of its points mean. The plots then show you everything in between.

### Anchor with Specific Predicted Values

Tutorial 20 obtains predicted probabilities for two county profiles that isolate the effect of mortality risk: an urban, highly educated, moderately diverse county with low mortality risk (Profile 1), and the same county with higher mortality risk (Profile 2). Reproducing those predictions here anchors the mortality risk relationship before you see the full curve --- the gap between them will tell you something about where these profiles fall on the S-curve.

The shape is worth recalling before you write it. Every explanatory variable in the model must appear in `newdata`, each with a value you choose --- there are no defaults. For a factor we supply the level with `factor("level", levels = levels(data$variable))`, so the value explicitly carries **all** of the levels the model was fitted on. R can often work them out from a bare string, but writing them out keeps the factor structure visible and guards against a level the model never saw. And `type = "response"` is what puts the answer on the probability scale; without it you get log-odds.

**Example code:**

```{r predict-response-template, exercise=FALSE, eval=FALSE}
predict(yourmodel,
        newdata = data.frame(
          interval_predictor = value_you_choose,
          factor_predictor   = factor("level you want",
                                      levels = levels(yourdata$factor_predictor))),
        type = "response")
```

The `levels =` argument is what makes the factor complete. Using `levels = levels(yourdata$factor_predictor)` gives the new factor the same set of levels as the variable the model was fitted on, which is the convention used throughout these tutorials.

**Practice 1: Compute two predicted probabilities**

Write both `predict()` calls from scratch. Profile 1 is an Urban county with **wage_growth** = 4.3, **prop_college_grad** = 0.35, **prop_white** = 0.65, and **mortality_risk_25_45** = 2.5. Profile 2 is the same county with **mortality_risk_25_45** = 5.0. Four hints are available. Click "Submit Answer" to check your work.

```{r exercise-recall-trump, exercise=TRUE, exercise.cap="Practice 1"}
# Profile 1: Urban county, low mortality risk

# Profile 2: Same county, higher mortality risk

```

```{r exercise-recall-trump-hint-1}
# Hint 1 of 4
# You need to ask the model what probability it assigns to a
# specific county. Think about what function produces a predicted
# probability from a fitted logistic regression model, what
# information you need to supply about the county, and why you
# need to specify the type of prediction you want.
```

```{r exercise-recall-trump-hint-2}
# Hint 2 of 4
# predict() takes the fitted model as its first argument.
# The newdata argument takes a data.frame() with one row
# containing a value for every explanatory variable in the model.
# The type argument controls what scale the prediction is on ---
# use "response" to get a probability.
```

```{r exercise-recall-trump-hint-3}
# Hint 3 of 4
# Your newdata data frame needs all five predictors from the model.
# The values for each are given in the prose above the exercise.
# rural_cat is a factor, so it has to be supplied as a factor
# carrying all of the original levels rather than as a bare
# string --- the Example code above shows the construction.
# One further argument puts the answer on the probability scale.
```

```{r exercise-recall-trump-hint-4}
# Hint 4 of 4
# Profile 2 uses exactly the same predict() structure as Profile 1.
# The only difference is the value of mortality_risk_25_45, which
# is 5.0. Everything else --- the function, the arguments, the
# other values, and the factor() construction for rural_cat ---
# is identical to Profile 1.
```

```{r exercise-recall-trump-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("predict\\s*\\(", code_no_comments))
    fail("Use `predict()` to obtain the predicted probabilities.")
  n_predict <- lengths(regmatches(code_no_comments, gregexpr("predict\\s*\\(", code_no_comments)))
  if (n_predict < 2)
    fail("You need two separate `predict()` calls --- one for Profile 1 and one for Profile 2.")
  if (!grepl("type\\s*=\\s*['\"]response['\"]", code_no_comments))
    fail("Set `type = \"response\"` to obtain predictions on the probability scale.")
  if (!grepl("mortality_risk_25_45\\s*=\\s*2\\.5", code_no_comments))
    fail("Set `mortality_risk_25_45 = 2.5` for Profile 1.")
  if (!grepl("mortality_risk_25_45\\s*=\\s*5\\.?0?", code_no_comments))
    fail("Set `mortality_risk_25_45 = 5.0` for Profile 2.")
  if (!grepl("prop_college_grad\\s*=\\s*0\\.35", code_no_comments))
    fail("Set `prop_college_grad = 0.35` for both profiles.")
  if (!grepl("prop_white\\s*=\\s*0\\.65", code_no_comments))
    fail("Set `prop_white = 0.65` for both profiles.")

  # Parse the submission so each call is graded on its own; `.result` holds only
  # the last one, and both profiles are being asked for.
  exprs <- tryCatch(as.list(parse(text = code_no_comments)), error = function(e) NULL)
  if (is.null(exprs))
    fail("Your code could not be parsed. Check for a missing comma or bracket.")
  strip_assign <- function(x)
    if (is.call(x) && identical(deparse(x[[1]]), "<-") && length(x) == 3L) x[[3]] else x
  pcalls <- lapply(Filter(function(x) {
    y <- strip_assign(x); is.call(y) && identical(deparse(y[[1]]), "predict")
  }, exprs), strip_assign)
  if (length(pcalls) != 2)
    fail(paste0("Write two `predict()` calls, one per profile. Your code has ", length(pcalls), "."))
  got <- suppressWarnings(vapply(pcalls, function(cl) {
    v <- tryCatch(as.numeric(eval(cl))[1], error = function(e) NA_real_)
    if (length(v) == 0L) NA_real_ else v
  }, numeric(1)))
  want <- as.numeric(predict(model_trump, type = "response", newdata = data.frame(
    wage_growth = 4.3, prop_college_grad = 0.35, prop_white = 0.65,
    mortality_risk_25_45 = c(2.5, 5.0),
    rural_cat = factor("Urban", levels = levels(counties$rural_cat)))))
  for (i in 1:2)
    if (is.na(got[i]) || abs(got[i] - want[i]) > 0.005)
      fail(paste0("Profile ", i, " does not give the expected probability. Check that every value in that call matches the profile described, and that only **mortality_risk_25_45** differs between the two."))

  pass("Right. Both predictions computed: about 18% for Profile 1 and about 41% for Profile 2, the two profiles differing only in mortality risk.")
})
```

You now have two specific points on the mortality risk curve. Before moving to the full plot, think about what the gap between them implies about where these profiles fall on the S-curve.

**Practice 2: What do these two values tell you about the S-curve?**

Look at both predicted probabilities in your output before answering.

```{r quiz-recall-trump, echo=FALSE}
question("Look at the two predicted probabilities in your output. What do they tell you about where these profiles fall on the S-curve?",
  correct = "Right. A 23-point jump means we are in a region of meaningful slope, not near the flat extremes.",
  answer("We are somewhere in the middle of the S-curve, where the mortality risk relationship is changing meaningfully",
         correct = TRUE,
         message = "A 23-percentage-point change for a 2.5-unit increase tells you the comparison falls in a region of genuine slope. Two points can only describe the relationship between them: they say nothing about whether the same increase would matter as much starting somewhere else on the curve."),
  answer("We are near the flat bottom of the S-curve, because both probabilities are below 50%",
         message = "Profile 2 is actually close to 41% --- not near the bottom. More importantly, a 23-point jump is not consistent with a flat region. The flat regions of an S-curve are where large changes in the predictor produce almost no change in probability."),
  answer("The two values fully describe the mortality risk relationship --- a plot would not add anything",
         message = "Two points describe the relationship between those two specific values, but not the full shape of the curve: how it flattens at the extremes, where the steepest section falls, or how the curve shifts when other predictors change. The plot shows all of that."),
  answer("We are near the flat top of the S-curve, because the probabilities are relatively high",
         message = "Profile 1 is at 18% --- not near the top. The flat top of an S-curve is where probabilities are already near 1 and barely move. A 23-point jump from 18% to 41% is a region of genuine slope."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a 23-point change means for how steep the curve is at these values."
)
```

## How Much Does Education Change the Probability of a Trump Majority?

The hypothesis about retrospective voting focused on wage growth and mortality risk, but the coefficients table showed that college graduation rates have a strong negative relationship with Trump support --- one of the most discussed structural features of the 2016 result. Counties where fewer adults hold college degrees were far more likely to give Trump a majority, even after controlling for economic distress, racial composition, and rurality. Here we visualize that relationship across its full range.

`plot_model()` is from the [sjPlot]{.package-name} package, loaded above. Its key arguments are: `type = "pred"` for a prediction plot, `terms =` for the variable on the x-axis, `axis.title =` for the axis labels (x first, then y), and `condition =` to specify held values for other variables. Add the caption with `+ labs(caption = "...")`.

**Run and observe 3: The default prediction plot, and the message it produces**

Run the code below and look at two things: the message that appears above the plot, and the shape of the line.

```{r exercise-predplot-college-default, exercise=TRUE, exercise.cap="Run and observe 3", fig.alt="Predicted probability of a Trump majority against the proportion of adults with a college degree. The horizontal axis runs from 0 to 1. The line consists of three points joined by two straight segments meeting at a corner near the middle, falling from close to 100 percent at the left to near 0 at the right."}
library(sjPlot)
plot_model(model_trump,
           type       = "pred",
           terms      = "prop_college_grad",
           title      = "Predicted Probability of a Trump Majority by County College Graduation Rate",
           axis.title = c("Proportion of Adults with a College Degree",
                          "Predicted Probability of Trump Majority")) +
  labs(caption = "Predictions are for Urban counties with average wage growth, mortality risk, and proportion white (0.862).")
```

That line is not an S-curve. It is two straight segments meeting at a corner, and the message above the plot explains why.

`plot_model()` does not compute a prediction at every value a variable takes. It picks a small set of round, evenly spaced values, computes predictions there, and joins them. The message calls this prettifying. For **prop_college_grad** it chose three values --- 0, 0.5 and 1 --- so the plot has three points and two straight lines between them. The S-shape the model actually implies is invisible at that resolution.

There is a second problem, and it is the more serious one. No county in the data has a college graduation rate above 0.80, and none is at 0. Two of the three points fall outside the range of the data entirely, so the plot is showing you predictions for counties that do not exist.

The message also tells you the fix. Adding `[all]` inside the quoted variable name --- `terms = "prop_college_grad [all]"` --- tells `plot_model()` to use every value the variable actually takes instead of a handful of round numbers. The plot is then drawn at 428 points running from 0.03 to 0.80, which is both a real curve and a curve over real counties.

This message will appear whenever the x-axis variable is interval and you have not asked for `[all]`, in this tutorial and in your own work. Treat it as a prompt to add the brackets. When the x-axis variable is a factor there is nothing to smooth --- there are only the categories --- and no message appears.

**Practice 4: Visualize the association between college graduation rates and the probability of a Trump majority**

The call is repeated below. Add the bracket notation to `terms =` so the plot is drawn at every observed value, and change nothing else. The `condition` argument is not used here --- `plot_model()` will hold all other predictors at their means and Urban reference category by default.

Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-college, exercise=TRUE, exercise.cap="Practice 4", fig.alt="The same relationship drawn as a smooth curve. The horizontal axis runs from 0 to about 0.8. The curve is close to 100 percent at the left, falls fastest through the middle of the range, and approaches 0 at the right, with a shaded band following it."}
plot_model(model_trump,
           type       = "pred",
           terms      = "prop_college_grad",
           title      = "Predicted Probability of a Trump Majority by County College Graduation Rate",
           axis.title = c("Proportion of Adults with a College Degree",
                          "Predicted Probability of Trump Majority")) +
  labs(caption = "Predictions are for Urban counties with average wage growth, mortality risk, and proportion white (0.862).")
```

```{r exercise-predplot-college-hint-1}
# Hint 1 of 2
# The line you just saw runs straight between a few widely spaced
# points, and two of those points describe counties that do not
# exist. What you want is a predicted probability calculated at
# every value the variable actually takes, so the plot traces the
# shape the model implies across the range of real counties rather
# than a sketch drawn through a handful of round numbers.
```

```{r exercise-predplot-college-hint-2}
# Hint 2 of 2
# The message printed above the previous plot named the change you
# need. It belongs inside the quotation marks in terms =, after the
# variable name and separated from it by a space, so that the
# variable name and the instruction together form one quoted
# string. Nothing else in the call changes.
```

```{r exercise-predplot-college-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_trump\\b", code_no_comments))
    fail("Pass **model_trump** as the model object.")
  if (!grepl("Predicted Probability of a Trump Majority by County College Graduation Rate", code_no_comments, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Probability of a Trump Majority by County College Graduation Rate\".")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"` to produce a prediction plot.")
  if (!grepl("prop_college_grad", code_no_comments))
    fail("Set `terms = \"prop_college_grad\"`.")
  if (!grepl("prop_college_grad\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Add the bracket notation inside the quotes: `terms = \"prop_college_grad [all]\"`.")
  if (!grepl("axis\\.title\\s*=", code_no_comments))
    fail("Include the `axis.title =` argument with the prescribed labels.")
  if (!grepl("Proportion of Adults with a College Degree", code_no_comments, fixed = TRUE))
    fail("Use the prescribed x-axis label: \"Proportion of Adults with a College Degree\".")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer stating the held values.")
  if (!grepl("Predictions are for Urban counties", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption.")

  pass("Well done. The message is gone, the curve is drawn at every observed value, and the x-axis now covers only the range counties actually occupy. The shape of the curve, the caption and the confidence band are now all describing the same set of counties.")
})
```

Once a package is loaded with `library()`, all of its functions are available for the rest of your R session --- you do not need to load it again each time. [sjPlot]{.package-name} was loaded at the start of the exercise above, so it is available for every exercise that follows in this tutorial. In your own work, load a package the first time you need one of its functions.

**Practice 5: Component 1 --- The pattern**

Look at the direction and shape of the line before answering.

```{r quiz-college-pattern, echo=FALSE}
question("Look at the prediction plot above. Which statement best describes the overall pattern?",
  correct = "Right. A strong negative S-curve: highly educated counties are much less likely to be Trump territory.",
  answer("As the proportion of college graduates increases, the predicted probability of a Trump majority falls steeply along an S-curve",
         correct = TRUE,
         message = "The curve drops from near certainty at low graduation rates to well below 50% at high graduation rates, tracing the characteristic shape of a logistic regression prediction. In a linear regression model, this line would be straight; in a logistic model, it accelerates through the steep middle and flattens at the extremes."),
  answer("As the proportion of college graduates increases, the predicted probability of a Trump majority rises",
         message = "The line slopes downward from left to right --- a negative relationship. More educated counties were less likely to give Trump a majority."),
  answer("The line is essentially flat, indicating that college graduation rates have no relationship with Trump support after controlling for other variables",
         message = "The line has a clearly negative slope, dropping substantially from low to high graduation rates. Across the range plotted, the predicted probability changes by a large amount."),
  answer("The relationship is negative and the line is straight, so each increase of 0.10 in the graduation rate lowers the predicted probability by the same amount",
         message = "The direction is right and the shape is not. Between 0.10 and 0.20 the predicted probability barely moves; between 0.30 and 0.40 it falls by tens of points. Equal steps in the predictor produce very unequal steps in the probability, which is what makes the curve an S."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: trace the line from left to right. Does the y-axis value rise, fall, or stay flat as the proportion of college graduates increases?"
)
```

The caption at the bottom of the plot tells you exactly what the other predictors were set to. Before answering, read it and think about why those choices matter more here than they did in Tutorial 17.

**Practice 6: Component 2 --- The held values**

The caption tells you what was held constant. This choice matters more in logistic regression than it does in linear regression.

```{r quiz-college-held, echo=FALSE}
question("The caption states that predictions hold other predictors at their means and assume Urban counties. What does this mean for interpreting the curve?",
  correct = "Right. The held values place these predictions in the context of an average, Urban county --- and changing them would shift the curve's position.",
  answer("The curve describes one specific profile --- an average Urban county --- and different held values would move it",
         correct = TRUE,
         message = "This is the key difference from linear regression. In an additive linear model without interactions involving the focal predictor, changing the held values shifts the line up or down without changing its slope. In a logistic model, different held values place you in a different region of the S-curve --- potentially changing both the level and the steepness of the relationship you see. The condition argument is what lets you set those held values yourself rather than accepting the defaults."),
  answer("The curve applies to all counties equally --- the held values are just a technical detail that does not affect interpretation",
         message = "The held values are not a technical detail here. They determine which part of the S-curve the plot is showing. A county with very different racial composition or in a different rurality category would be positioned differently on the S-curve, potentially producing a steeper or flatter curve at a different level."),
  answer("Because Urban is the reference category, the curve only applies to cities and should not be used to draw conclusions about rural counties",
         message = "The caption tells you the baseline for these predictions, not the population the finding applies to. Urban is the reference category, so it is where the curve is drawn; the education relationship the model estimates is the same in every rurality category."),
  answer("The held values are the averages of the data, so the curve describes the average relationship across all counties",
         message = "A profile built from means is one specific combination, not an average over counties --- and with rurality held at Urban it may describe no county that exists. The curve is the model's expectation for that single profile, not an average of the curves you would get for each county separately."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what an interaction term would be needed to make the slope itself differ by rurality category, and what the held values tell you about which version of the relationship you are seeing."
)
```

Direction and shape are not enough --- the size of the effect is what makes a result substantively meaningful. Read two specific values from the plot before answering.

**Practice 7: Component 3 --- Specific predicted values**

Read two values from the plot before answering --- one near the low end of the observed graduation rate range and one near the high end.

```{r quiz-college-values, echo=FALSE}
question("Using the prediction plot, approximately what is the predicted probability of a Trump majority at prop_college_grad = 0.10 and at prop_college_grad = 0.45 for an Urban county at average other values?",
  correct = "Right. About 98% at low graduation rates, about 43% at high --- a 55-point drop across a realistic range.",
  answer("About 98% at `prop_college_grad = 0.10` and about 43% at `prop_college_grad = 0.45` --- a 55-percentage-point difference",
         correct = TRUE,
         message = "A 55-point drop is a large substantive effect, visible in the steep descent of the S-curve through the middle of the graduation rate range. This is what the college graduation coefficient in the table could not tell you directly: not just that the relationship is negative and significant, but that moving from a low-graduation county to a high-graduation county predicts the probability of a Trump majority nearly in half, passing through the steep middle of the curve."),
  answer("About 50% at both values --- the curve is symmetric around 0.5",
         message = "Read the y-axis value at `prop_college_grad = 0.10` and at 0.45 separately. The curve is not symmetric around 0.5 at these values."),
  answer("About 80% at `prop_college_grad = 0.10` and about 80% at `prop_college_grad = 0.45` --- the curve is nearly flat",
         message = "The curve is not flat. It drops substantially from low to high graduation rates. Trace from x = 0.10 up to the line and read the y-axis; repeat at x = 0.45."),
  answer("About 98% and about 43%, so the predicted probability falls by 55% between the two graduation rates",
         message = "The two readings are right and the arithmetic describing them is not. A fall from 98% to 43% is 55 percentage points; as a share of where it started it is a fall of about 56%. Percentage points and percent are different quantities, and reporting one as the other misstates the size of the finding."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find `prop_college_grad = 0.10` on the horizontal axis, trace up to the line, and read the y-axis value. Repeat at `prop_college_grad = 0.45.`"
)
```

Every prediction comes with uncertainty. The shaded [confidence band]{.important-text} around the curve makes that uncertainty visible in a way the p-value cannot.

::: {.important-note}
**What the shaded band is.** It shows how precisely the model estimates the predicted probability at each value of the x-axis variable --- wide where the estimate is uncertain, narrow where it is well determined. It is **not** a range in which individual outcomes fall. Every case in the data is a 0 or a 1; the band is about the model's estimate of the probability, not about where any particular county or state will land.
:::

**Practice 8: Component 4 --- The uncertainty**

Look at the width of the confidence band and whether it is consistent across the range before answering.

```{r quiz-college-band, echo=FALSE}
question("Look at the shaded confidence band in the plot. Which statement best describes what it tells you?",
  correct = "Right. The band is narrower in the middle of the range where most counties are, and wider at the extremes.",
  answer("The band is narrowest in the middle of the range and widens at both extremes",
         correct = TRUE,
         message = "With data from over 3,000 counties, the band is fairly narrow throughout, but the tightening in the middle reflects where most counties fall: between 0.10 and 0.35. The widening at very low or very high graduation rates (fewer counties) is expected and does not undermine the directional conclusion."),
  answer("The band is so wide throughout that no conclusions can be drawn",
         message = "With over 3,000 counties, the band is actually relatively narrow. The relationship is estimated with reasonable precision across most of the observed range."),
  answer("The band is perfectly uniform, indicating equal precision at all graduation rate levels",
         message = "Prediction bands are almost never perfectly uniform. With more counties concentrated in the middle of the graduation rate range, predictions there are more precise than at the extremes."),
  answer("The band is narrow where the model fits the data well and wide where it fits poorly",
         message = "Band width measures how precisely the curve is located, not how well the model describes the cases. A model can fit badly and still have a narrow band if its coefficients are precisely estimated. Width here reflects where the data are dense and how the probability scale compresses near 0 and 1."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the band width at the far left, the middle, and the far right of the plot."
)
```

The plot is not just an illustration of the coefficient --- it adds information the table cannot provide. Think about what that new information is before answering.

**Practice 9: Component 5 --- The hypothesis connection**

Think about what the plot adds to what the coefficients table already told you.

```{r quiz-college-hypothesis, echo=FALSE}
question("The coefficients table showed that prop_college_grad is a significant negative predictor. What does the prediction plot add to that conclusion?",
  correct = "Right. The plot translates the significant negative coefficient into a concrete, visible magnitude.",
  answer("The plot shows the substantive size of the educational divide, which the coefficient alone could not communicate",
         correct = TRUE,
         message = "A statistically significant negative coefficient provides evidence that the population association is negative. The prediction plot tells you the magnitude, and it also tells you where on the curve the counties sit. Across the middle half of counties --- from the 25th percentile of graduation rates (about 0.14) to the 75th (about 0.25) --- the predicted probability falls only from about 97% to about 89%, because that whole range sits on the flat upper shoulder of the S-curve. The 55-point drop appears only once the range extends past the 75th percentile into the steep middle. That is the educational divide made visible, and the plot shows both how large it is and which counties it separates."),
  answer("The plot adds nothing beyond the p-value --- a significant coefficient and a downward-sloping line convey the same information",
         message = "The p-value tells you that a coefficient at least this extreme would be unlikely if the population coefficient were zero. The plot tells you the magnitude and shape across the observed range --- how large the gap is between high- and low-graduation counties, and where on the range the steepest change occurs. That is substantively different information."),
  answer("The plot proves that education causes counties to vote against Trump",
         message = "A prediction plot shows an association between college graduation rates and the predicted probability of a Trump majority. It does not establish causation. Many factors correlated with education may be driving the relationship."),
  answer("The plot confirms the coefficient is statistically significant, because the confidence band does not contain a flat line",
         message = "Significance is established by the coefficient's test, not by inspecting a band. Reading significance off a plot is the error the coefficient table exists to prevent --- and what the plot contributes is magnitude, which the table cannot give."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a p-value tells you and what it cannot tell you about the size or practical meaning of an effect."
)
```

Now consider what the plot looks like for counties with a lower proportion of white residents --- say 0.70, compared to the mean of 0.862. Counties at 0.70 are more racially diverse than average; they represent about 14% of counties. The `condition` argument lets us ask: is the educational divide in Trump support different in more diverse counties than in majority-white counties? This is the kind of contextual question that the default plot cannot answer.

**Practice 10: The same plot, conditioned on prop_white = 0.70**

Fill in the blanks. The `condition` argument and the caption are the only changes from the exercise above --- the caption has to tell a reader what this plot holds constant, so the one from the previous exercise will not do.

- caption: `"Predictions are for Urban counties with `prop_white = 0.70.`"`

Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-college-condition, exercise=TRUE, exercise.cap="Practice 10", fig.alt="Predicted probability of a Trump majority against the proportion of adults with a college degree, for counties with the proportion white held at 0.70. The curve falls from near 100 percent at the left of the range to near 0 at the right, with a shaded band around it."}
plot_model(model_trump,
           type       = "pred",
           terms      = "prop_college_grad [all]",
           condition  = list(prop_white = ___),
           title      = "Predicted Probability of a Trump Majority by College Graduation Rate",
           axis.title = c("Proportion of Adults with a College Degree",
                          "Predicted Probability of Trump Majority")) +
  labs(caption = "___")
```

```{r exercise-predplot-college-condition-hint-1}
# Hint 1 of 2
# The task is the same as the previous plot, but you need to tell
# the function to hold prop_white at 0.70 rather than its mean.
# Think about which argument controls the values of explanatory
# variables that are not on the x-axis, and what structure that
# argument uses to pair a variable name with a specific value.
```

```{r exercise-predplot-college-condition-hint-2}
# Hint 2 of 2
# The condition argument takes a named list where each name is
# an explanatory variable and each value is what you want it
# held at: condition = list(variable_name = value).
# Fill in the argument name and the value for prop_white.
# Update the caption to reflect that prop_white = 0.70.
```

```{r exercise-predplot-college-condition-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_trump\\b", code_no_comments))
    fail("Pass **model_trump** as the model object.")
  if (!grepl("Predicted Probability of a Trump Majority by College Graduation Rate", code_no_comments, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Probability of a Trump Majority by College Graduation Rate\".")
  if (!grepl("condition\\s*=", code_no_comments))
    fail("Include the `condition =` argument to specify the held value for **prop_white**.")
  if (!grepl("prop_white\\s*=\\s*0\\.70", code_no_comments))
    fail("Set `prop_white = 0.70` inside condition = list(...).")
  if (!grepl("prop_college_grad", code_no_comments))
    fail("Keep `terms = \"prop_college_grad\"`.")
  if (!grepl("prop_college_grad\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Keep the bracket notation inside the quotes: `terms = \"prop_college_grad [all]\"`.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer with an updated caption reflecting `prop_white = 0.70.`")
  if (!grepl("counties with prop_white = 0.70", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption, which tells a reader that **prop_white** is held at 0.70 --- the caption from the previous exercise is not correct here.")
  if (!grepl("Predictions are for Urban counties with prop_white = 0.70.", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for Urban counties with `prop_white = 0.70.`\".")

  pass("Correct. The plot now holds **prop_white** at 0.70 rather than at its mean, and the caption says so.")
})
```

You now have two plots showing the same relationship in two different demographic contexts. The question is not just whether they look different, but why --- and what that difference tells you about how to read logistic regression prediction plots.

**Practice 11: Compare the two college graduation plots**

Look at both plots --- the default (`prop_white = 0.862`) and the conditioned (`prop_white = 0.70`) --- before answering.

```{r quiz-college-condition, echo=FALSE}
question("Compare the two college graduation plots. Which statement best describes what changed and why?",
  correct = "Right. The curve shifted downward because lower **prop_white** places these counties in a different region of the S-curve.",
  answer("The curve shifted downward: at any given graduation rate the predicted probability is lower for the more diverse counties",
         correct = TRUE,
         message = "At `prop_college_grad = 0.10`, the default curve gives about 98% and the conditioned curve gives about 90% --- an 8-point drop. At `prop_college_grad = 0.35`, the default gives about 71% and the conditioned gives about 32% --- a 39-point drop. The gap is larger in the steep middle than at the extremes, because conditioning on lower **prop_white** moves us into a steeper region of the S-curve there. This is the condition argument in action: the same educational divide looks different depending on the racial composition context."),
  answer("The curves are essentially identical --- changing prop_white has no effect on the college graduation relationship",
         message = "The curves are not identical. Compare the predicted probabilities at the same graduation rate values across the two plots --- they differ substantially, especially in the middle of the range."),
  answer("The curve shifted upward --- more diverse counties are more likely to give Trump a majority at any graduation rate level",
         message = "The curve shifted downward, not upward. Counties with lower **prop_white** have lower predicted probabilities of a Trump majority throughout, consistent with the positive **prop_white** coefficient in the model."),
  answer("The curve shifted downward and became steeper, which shows the education coefficient is larger in more diverse counties",
         message = "The education coefficient is identical in both plots --- the model was fitted once and nothing about it changed. Conditioning moved the predictions to a different part of the S-curve, where the same log-odds coefficient converts into a steeper change in probability. Changing steepness without changing a coefficient is what distinguishes logistic from linear prediction plots."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: compare the y-axis values at the same x-axis position in each plot. Did the curve move up or down, and does the gap change across the graduation rate range?"
)
```

## Does Mortality Risk Matter Once Education Is Accounted For?

Tutorial 20 found that mortality risk is a significant positive predictor of Trump support (p = .001) --- higher working-age mortality is associated with a higher probability of a Trump majority, consistent with the retrospective voting hypothesis. But the default prediction plot at mean county characteristics barely moves: because the average county has high **prop_white** (0.862), most counties are already near the flat top of the S-curve, where additional mortality risk adds little to an already-high probability. To see the mortality risk relationship in full, we need to ask what it looks like in counties where the demographic baseline does not already place the probability near certainty.

Conditioning on `prop_white = 0.60` describes a racially diverse urban county --- the kind of place where the deaths-of-despair argument is contested, since that literature developed around predominantly white communities. Whether working-age mortality is associated with Trump support in more diverse counties is a question worth asking on its own terms, and it is the reason for choosing this value. Counties at this level of racial diversity are uncommon --- about 8.5% of all counties --- so the plot describes a real but small part of the country, which the caption has to say.

**Run and observe 12: Mortality risk at mean values vs. conditioned on prop_white = 0.60**

The code below produces both plots, the default first and the conditioned one below it, so you can see the difference directly.

```{r exercise-mortality-condition, exercise=TRUE, exercise.cap="Run and observe 12", fig.alt="Two plots, one above the other, of the predicted probability of a Trump majority against mortality risk for ages 25 to 45. The horizontal axis of each runs from about 1 to about 8. In the upper plot, at the mean proportion white, the curve rises from left to right and flattens toward the top of the vertical axis. In the lower plot, with the proportion white held at 0.60, the curve rises from left to right across a much taller span of the vertical axis."}
# Default: prop_white held at its mean (0.862)
plot_model(model_trump,
           type       = "pred",
           terms      = "mortality_risk_25_45 [all]",
           title      = "Mortality Risk at Mean prop_white (0.862)",
           axis.title = c("Mortality Risk (Ages 25-45)",
                          "Predicted Probability of Trump Majority")) +
  labs(caption = "Predictions are for Urban counties with average wage growth, college graduation rate, and prop_white (0.862).")

# Conditioned: prop_white held at 0.60
plot_model(model_trump,
           type       = "pred",
           terms      = "mortality_risk_25_45 [all]",
           condition  = list(prop_white = 0.60),
           title      = "Mortality Risk Conditioned on prop_white = 0.60",
           axis.title = c("Mortality Risk (Ages 25-45)",
                          "Predicted Probability of Trump Majority")) +
  labs(caption = "Predictions are for Urban counties with average wage growth and college graduation rate, and prop_white = 0.60.")
```

The two plots show the same predictor --- mortality risk --- at two different held values of **prop_white**. The difference between them is the argument for why held values matter in logistic regression, visible directly.

**Practice 13: What does the comparison show?**

Look at both plots before answering.

```{r quiz-mortality-condition, echo=FALSE}
question("Compare the two mortality risk plots. Which statement best describes what the comparison reveals?",
  correct = "Right. The same predictor tells very different stories depending on what is held constant.",
  answer("At mean **prop_white** the curve barely moves, but conditioned on `prop_white = 0.60` the same predictor produces a large rise",
         correct = TRUE,
         message = "This is the core lesson about condition in logistic regression. The mortality risk coefficient is the same in both plots --- the model has not changed. What changed is where on the S-curve the plot is showing the relationship. At mean **prop_white**, most counties are already near the probability ceiling, so mortality risk has little room to push the probability higher. At `prop_white = 0.60`, counties start in the steep middle, where mortality risk produces a large, visible effect. Reading the caption is not a formality --- it tells you which story the plot is telling."),
  answer("The two plots show the same curve at different vertical positions --- the shape is identical and only the level changes",
         message = "The shape is not identical. At mean prop_white, the curve is nearly flat because it is compressed near the top of the S-curve. At prop_white = 0.60, the curve has a clear positive slope through the steep middle. Both the level and the steepness change when the held value changes --- this is the key difference from additive linear regression, where changing held values shifts the line without changing its slope."),
  answer("The conditioned plot shows that mortality risk has no effect on Trump support in diverse counties",
         message = "The conditioned plot shows the opposite: a large positive effect of mortality risk in more diverse counties, rising from about 27% to about 91%. The effect is actually more visible there than in the default plot."),
  answer("The default plot is the more trustworthy of the two, because it holds **prop_white** at its mean rather than at a value the analyst picked",
         message = "Neither is more trustworthy; they answer different questions. The mean is a default, not a neutral choice --- it describes counties near the top of the S-curve, where mortality risk has little room to move the probability. Choosing a held value is part of the analysis, and the caption is what makes it honest."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: compare the range of the y-axis values in each plot. Which curve spans a larger range of predicted probabilities, and what does that tell you about where on the S-curve each plot is positioned?"
)
```

## How Much Do Rural and Urban Counties Differ?

**rural_cat** is a three-category factor. When `terms` is a factor variable, `plot_model()` plots the predicted probability for each category with a confidence interval rather than a continuous curve --- there is no meaningful range to sweep across for a categorical variable, only discrete levels to compare. Each predicted probability is for one rurality category at the held values stated in the caption.

**Practice 14: Produce a prediction plot for rurality**

Prescribed values:
- `title = "Predicted Probability of a Trump Majority by County Rurality"`
- x-axis: \"Rurality Category\", y-axis: \"Predicted Probability of Trump Majority\"
- caption: \"Predictions are for counties with average wage growth, college graduation rate, mortality risk, and proportion white.\"

Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-rural, exercise=TRUE, exercise.cap="Practice 14", fig.alt="Three predicted probabilities, one for each rurality category, each drawn with a vertical confidence interval. The categories are labeled Urban, Intermediate and Rural, and the vertical axis spans roughly 91 to 97 percent."}
___(model_trump,
    ___ = "pred",
    ___ = "rural_cat",
    title = "Predicted Probability of a Trump Majority by County Rurality",
    ___ = c("Rurality Category",
             "Predicted Probability of Trump Majority")) +
  ___(caption = "Predictions are for counties with average wage growth, college graduation rate, mortality risk, and proportion white.")
```

```{r exercise-predplot-rural-hint-1}
# Hint 1 of 3
# The task is to compare the predicted probability of a Trump
# majority across the three rurality categories, holding the
# other explanatory variables at their means. Think about what
# kind of plot shows discrete predictions for each category of
# a factor variable, and which argument tells the plot which
# variable to use for the categories on the x-axis.
```

```{r exercise-predplot-rural-hint-2}
# Hint 2 of 3
# The function and argument names are the same as in the college
# graduation plot you just produced: plot_model(), type =, terms =,
# axis.title =, and + labs(caption = "..."). The only difference
# is that when terms = names a factor variable, plot_model()
# automatically plots one predicted probability with a confidence
# than a continuous line --- no other change is needed.
```

```{r exercise-predplot-rural-hint-3}
# Hint 3 of 3
# Your completed call uses plot_model() with model_trump as the
# first argument, type = "pred", and terms = "rural_cat" ---
# both of which are already in the starter code. The blanks
# are the function name itself, axis.title = with the two
# prescribed labels in order, and + labs(caption = "...") with
# the prescribed caption. The title is already provided.
```

```{r exercise-predplot-rural-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_trump\\b", code_no_comments))
    fail("Pass **model_trump** as the model object.")
  if (!grepl("Predicted Probability of a Trump Majority by County Rurality", code_no_comments, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Probability of a Trump Majority by County Rurality\".")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"`.")
  if (!grepl("rural_cat", code_no_comments))
    fail("Set `terms = \"rural_cat\"`.")
  if (!grepl("Rurality Category", code_no_comments, fixed = TRUE))
    fail("Use \"Rurality Category\" as the x-axis label.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer.")
  if (!grepl("Predictions are for counties with average", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption.")
  if (!grepl("Predicted Probability of Trump Majority", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of Trump Majority\".")

  pass("Exactly. A factor on the x-axis gives one predicted probability and its confidence interval per category rather than a continuous curve.")
})
```

The rurality plot will look very different from the education plot. Before answering, check the y-axis scale and connect what you see to the rurality p-values from Tutorial 20.

**Practice 15: Interpret the rurality plot**

Look at which predicted probabilities stand apart before answering, and remember that the coefficient tests, not the intervals, settle whether categories differ.

```{r quiz-rural-interpret, echo=FALSE}
question("Look at the rurality prediction plot. Which statement best describes the key finding?",
  correct = "Right. Rural counties stand apart; the intermediate category does not.",
  answer("All three categories sit above 90%, and Rural has the highest predicted probability of the three",
         correct = TRUE,
         message = "The predicted probabilities are about 93% for Urban, 92% for Intermediate, and 96% for Rural at average values of the other predictors. All three are high because the mean county has a high proportion of white residents and a low graduation rate, which places the baseline near the top of the S-curve. The plot gives you the estimates and their uncertainty; whether a category differs from the reference is settled by the coefficient tests in Tutorial 20, which distinguish Rural from Urban (p = .032) but not Intermediate (p = .599). The estimated rural-urban pattern is concentrated at the rural extreme rather than forming a clear stepwise gradient from Urban to Intermediate to Rural."),
  answer("Rural counties are predicted to have dramatically higher Trump support than Urban counties, confirming that the rural-urban divide drove the 2016 election",
         message = "The Rural predicted probability is higher, and significantly so, but the gap is about three percentage points on a scale where every category already exceeds 90%. That is a detectable difference, not a dramatic one --- and the intermediate counties, which are also nonmetropolitan, are indistinguishable from urban ones."),
  answer("The confidence intervals are wide enough that the model cannot be relied on to predict Trump support in any rurality category",
         message = "The intervals are not wide relative to the scale --- all three predicted probabilities sit between 91% and 97%. They show how precisely each category's probability is estimated; they are not a test of whether the categories differ."),
  answer("All three predicted probabilities sit above 90%, which shows the model predicts a Trump majority for almost every county",
         message = "All three describe one profile: a county at the mean on every other predictor, which has a high proportion of white residents and a low graduation rate. That profile sits near the top of the S-curve. Counties with other characteristics get very different predictions, as the education plot shows."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: read the predicted probability for each category off the y-axis. Whether a difference is statistically distinguishable is settled by the coefficient tests, not by the plot."
)
```

## How Do Education and Rurality Work Together?

A two-variable prediction plot puts **prop_college_grad** on the x-axis and produces separate lines for each level of **rural_cat**.

::: {.important-note}
**Separate curves do not mean an interaction.** This model is additive, so the same educational coefficient applies in Urban, Intermediate and Rural counties on the log-odds scale --- each category is simply shifted by its rurality coefficient. On the probability scale the curves can still look differently steep, because each group sits on a different part of the S-curve, which flattens as it approaches 0 and 1. That flattening is [S-curve compression]{.important-text}: a constant difference in log-odds turns into fewer and fewer percentage points as the predicted probability nears either bound. So do not read differently shaped probability curves as evidence that the predictor works differently across groups. Curves that actually crossed, or group-specific coefficients, would require an interaction term.
:::

**Practice 16: Produce a two-variable prediction plot**

Prescribed values:
- `title = "Predicted Probability of a Trump Majority by Education and Rurality"`
- x-axis: \"Proportion of Adults with a College Degree\", y-axis: \"Predicted Probability of Trump Majority\"
- legend: \"Rurality Category\"
- caption: \"Predictions are for counties with average wage growth, mortality risk, and proportion white.\"

Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-twovar-trump, exercise=TRUE, exercise.cap="Practice 16", fig.alt="Three curves, one per rurality category, showing predicted probability of a Trump majority against the proportion of adults with a college degree. Each falls from near 100 percent at the left to near 0 at the right, each with its own shaded band, and a legend names the three categories."}
___(model_trump,
    ___ = "pred",
    ___ = c("prop_college_grad [all]", "rural_cat"),
    title = "Predicted Probability of a Trump Majority by Education and Rurality",
    ___ = c("Proportion of Adults with a College Degree",
             "Predicted Probability of Trump Majority")) +
  ___(color   = "Rurality Category",
      caption = "Predictions are for counties with average wage growth, mortality risk, and proportion white.")
```

```{r exercise-predplot-twovar-trump-hint-1}
# Hint 1 of 3
# The task is to show how the education relationship varies across
# rurality categories simultaneously --- combining the continuous
# plot you made for college graduation rates with the factor
# categories from the rurality plot into a single visualization.
# Think about which variable sweeps along the x-axis and which
# produces separate colored lines for each of its levels.
```

```{r exercise-predplot-twovar-trump-hint-2}
# Hint 2 of 3
# The function and argument names are the same as in the
# single-variable plots: plot_model(), type =, terms =,
# axis.title =, and labs(). The only structural change is that
# terms = now takes a character vector: c("x_variable", "group_variable").
# The first element goes on the x-axis; the second produces lines.
```

```{r exercise-predplot-twovar-trump-hint-3}
# Hint 3 of 3
# Your completed call uses plot_model() with model_trump,
# type = "pred", and terms = c("prop_college_grad [all]", "rural_cat")
# --- all already in the starter code. The blanks are the
# function name itself, axis.title = with the two prescribed
# labels in order (x-axis first), and the labs() layer carrying
# both the prescribed legend title and the prescribed caption.
# The title is already provided.
```

```{r exercise-predplot-twovar-trump-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_trump\\b", code_no_comments))
    fail("Use **model_trump** as the model in `plot_model()`.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"`.")
  if (!grepl("prop_college_grad", code_no_comments))
    fail("Include **prop_college_grad** as the first term (x-axis).")
  if (!grepl("prop_college_grad\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Keep the bracket notation on the x-axis variable: c(\"prop_college_grad [all]\", \"rural_cat\").")
  if (!grepl("rural_cat", code_no_comments))
    fail("Include **rural_cat** as the second term (separate lines).")
  if (regexpr("prop_college_grad", code_no_comments) > regexpr("rural_cat", code_no_comments))
    fail("prop_college_grad should come first in terms (x-axis), and **rural_cat** second (separate colored lines). Check the order inside c(...).")
  if (!grepl("axis\\.title\\s*=", code_no_comments))
    fail("Include the `axis.title =` argument.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer.")
  if (!grepl("Rurality Category", code_no_comments, fixed = TRUE))
    fail("Name the legend with the prescribed label: labs(`color = \"Rurality Category\"`, ...).")
  if (!grepl("Proportion of Adults with a College Degree", code_no_comments, fixed = TRUE))
    fail("Use the prescribed label: \"Proportion of Adults with a College Degree\".")
  if (!grepl("Predicted Probability of Trump Majority", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of Trump Majority\".")

  # The checks above find the required pieces somewhere in the submission; this
  # one establishes they belong to the plot actually produced.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument sits inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one --- check it is set inside this plot rather than elsewhere in your code."))
  }
  chk("x", "Proportion of Adults with a College Degree", "x-axis label")
  chk("y", "Predicted Probability of Trump Majority", "y-axis label")
  chk("colour", "Rurality Category", "legend title")
  if (length(unique(bl$data[[1]]$group)) != nlevels(counties$rural_cat))
    fail("The plot should draw one line per category. Check the second variable in `terms`.")

  pass("Well done. Each rurality category has its own line, and the vertical distance between them is what the rurality coefficients buy.")
})
```

The visual structure of the two-variable plot carries a specific message about the model. Before answering, identify the geometric feature that distinguishes an additive model from an interactive one.

**Practice 17: Interpret the two-variable plot**

The lines do not cross. Before answering, think about what that tells you about the model.

```{r quiz-twovar-trump, echo=FALSE}
question("The three lines in the two-variable plot do not cross. Given that this model has no interaction term, what does that tell you about the relationship between education, rurality, and Trump support?",
  correct = "Right. The same log-odds coefficient applies in every group, which is why the lines cannot cross.",
  answer("On the log-odds scale the educational coefficient is the same in every rurality category, and the lines do not cross",
         correct = TRUE,
         message = "Each rurality category shifts the predicted curve up or down by its coefficient on the log-odds scale, but the educational coefficient is mathematically the same in every category, so the curves cannot cross. Note the direction of that reasoning: an additive model cannot produce crossing curves, so a crossing would require an interaction --- but non-crossing curves do not by themselves establish that a model is additive, since an interaction can give two groups different slopes without their curves meeting anywhere in the observed range. If we wanted to test whether the educational divide differs by rurality on the log-odds scale, we would need to add an interaction term between **prop_college_grad** and **rural_cat**."),
  answer("The lines not crossing means education has no effect on Trump support in any rurality category",
         message = "In this fitted additive model the log-odds effect is the same across rurality categories, and the lines not crossing is consistent with that --- though non-crossing lines alone would not establish it. What they certainly do not show is an absence of effect: all three slope downward steeply, so education is strongly associated with the outcome in every category."),
  answer("The vertical gaps between lines are constant at all education levels, which means the rurality coefficients are large and significant",
         message = "The gaps between lines are determined by the rurality coefficients, but they may appear to narrow near the extremes of the probability scale due to S-curve compression. Whether those gaps are statistically significant is a separate question from whether the lines cross. In this model the Rural line sits visibly above the other two through the middle of the range, matching its significant coefficient in Tutorial 20, while the Intermediate line tracks Urban closely."),
  answer("The lines do not cross because rurality entered the model as a control rather than as a predictor of interest",
         message = "A model does not distinguish controls from predictors of interest --- that distinction is the analyst's, not the model's. The lines do not cross because the model has no interaction term between education and rurality, which would be equally true if rurality were the variable the research question was about."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the model has no interaction term. What does that force the educational coefficient to be in each rurality category, and what would it take for the curves to cross?"
)
```

### What We Learned About Education, Mortality Risk, and Trump Support

The prediction plots complete the picture from Tutorial 20. The educational divide emerges as the central visual finding: the predicted probability of a Trump majority drops from near certainty in low-graduation counties to below 50% in high-graduation counties, tracing a clear S-curve that the coefficient table could only summarize as "negative and significant." The condition exercise shows the same additive log-odds relationship looking quite different on the probability scale in different demographic contexts --- in more diverse counties (`prop_white = 0.70`), the curve shifts down substantially and passes through the steep middle at lower graduation rates, so the educational signal is visible at a lower baseline. The educational coefficient has not changed; the curvature of the logistic function means an unchanged log-odds slope produces different probability differences at different baselines. The mortality risk exercise demonstrates the core lesson about held values: the same predictor looks nearly flat at mean county characteristics (because most counties are already near the top of the S-curve) and dramatically steep in more diverse counties --- the same model, the same coefficient, two very different pictures depending on context. The rurality predictions separate the completely rural counties from the rest: Urban and Intermediate sit together near 92-93% while Rural sits at about 96% --- matching the coefficient tests in Tutorial 20, which distinguish `rural_catRural` from the Urban reference but not `rural_catIntermediate`. The non-crossing lines in the two-variable plot are what an additive structure forces: the log-odds educational coefficient is the same in Urban, Intermediate, and Rural counties, so the curves do not cross.

Here is what a complete interpretation of the college graduation prediction plot looks like written as prose: *As the proportion of adults with college degrees increases, the predicted probability of a Trump majority falls steeply, tracing an S-curve from near certainty at low graduation rates to well below 50% at high graduation rates. Predictions hold other predictors at their means and assume Urban counties; conditioning on a more racially diverse county would shift the entire curve downward, placing the comparison in a steeper region of the S-curve. At a graduation rate of 10%, the model predicts about 98% probability of a Trump majority; at 45%, about 43% --- a 55-point drop across a realistic range of values. The confidence band is moderately narrow throughout and widens slightly at the extremes, where fewer counties are observed. This pattern is strongly consistent with the diploma divide documented in the 2016 election literature: after controlling for economic distress, racial composition, and geography, the model implies a large difference in the probability of a Trump majority across the observed range of county educational attainment.*


## Is State Ideology Associated with Greenhouse Gas Cap Adoption?

This example is where the opening section's argument pays off most concretely. The coefficients table told you the direction and significance of each predictor. The prediction plots will show you what those coefficients actually mean for the probability of emissions cap adoption across the range of state ideology --- and how dramatically that picture changes depending on a state's economic context.

The model output below recalls the coefficients for reference. Pay particular attention to the ideology coefficient and its significance before producing any plots.

```{r model-ghg-recall, echo=TRUE, comment=NA}
summary(model_ghg)
```

### Anchor with Specific Predicted Values

Tutorial 20 obtains predicted probabilities for two state profiles that isolate the effect of ideology: a state with low conservative identification and a Democratic governor (Profile 1), and the same state with moderately higher conservative identification (Profile 2). Reproducing those predictions here anchors the ideology relationship before you see the full curve --- the gap between them will place the comparison on the S-curve and set up the full plot that follows.

**Practice 18: Compute two predicted probabilities**

Write both `predict()` calls from scratch. Profile 1 is a Democrat-governed state with **ideology_conservative2024** = 20 and **econ_poverty2022** = 12.5. Profile 2 is the same state with **ideology_conservative2024** = 32. Four hints are available. Click "Submit Answer" to check your work.

```{r exercise-recall-ghg, exercise=TRUE, exercise.cap="Practice 18"}
# Profile 1: Low conservative identification, Democrat governor

# Profile 2: Same state, higher conservative identification

```

```{r exercise-recall-ghg-hint-1}
# Hint 1 of 4
# You need to ask the model what probability it assigns to a
# specific state. Think about what function produces a predicted
# probability from a fitted logistic regression model, what
# information you need to supply about the state, and why you
# need to specify the type of prediction you want.
```

```{r exercise-recall-ghg-hint-2}
# Hint 2 of 4
# predict() takes the fitted model as its first argument.
# The newdata argument takes a data.frame() with one row
# containing a value for every explanatory variable in the model.
# The type argument controls what scale the prediction is on ---
# use "response" to get a probability.
```

```{r exercise-recall-ghg-hint-3}
# Hint 3 of 4
# Your newdata data frame needs all three predictors from the model.
# The values for each are given in the prose above the exercise.
# gov_f is a factor, so it has to be supplied as a factor
# carrying all of the original levels rather than as a bare
# string --- the same construction you used for the county
# profiles. One further argument puts the answer on the
# probability scale.
```

```{r exercise-recall-ghg-hint-4}
# Hint 4 of 4
# Profile 2 uses exactly the same predict() structure as Profile 1.
# The only difference is the value of ideology_conservative2024,
# which is 32. Everything else --- the function, the arguments,
# the other values, and the factor() construction for gov_f ---
# is identical to Profile 1.
```

```{r exercise-recall-ghg-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("predict\\s*\\(", code_no_comments))
    fail("Use `predict()` to obtain the predicted probabilities.")
  n_predict <- lengths(regmatches(code_no_comments, gregexpr("predict\\s*\\(", code_no_comments)))
  if (n_predict < 2)
    fail("You need two separate `predict()` calls --- one for Profile 1 and one for Profile 2.")
  if (!grepl("type\\s*=\\s*['\"]response['\"]", code_no_comments))
    fail("Set `type = \"response\"` to obtain predictions on the probability scale.")
  if (!grepl("ideology_conservative2024\\s*=\\s*20", code_no_comments))
    fail("Set `ideology_conservative2024 = 20` for Profile 1.")
  if (!grepl("ideology_conservative2024\\s*=\\s*32", code_no_comments))
    fail("Set `ideology_conservative2024 = 32` for Profile 2.")
  if (!grepl("econ_poverty2022\\s*=\\s*12\\.5", code_no_comments))
    fail("Set `econ_poverty2022 = 12.5` for both profiles.")
  if (!grepl("Democrat", code_no_comments))
    fail("Set **gov_f** to Democrat for both profiles, using `factor(\"Democrat\", levels = levels(statesPolicy$gov_f))`.")

  # Parse the submission so each call is graded on its own; `.result` holds only
  # the last one, and both profiles are being asked for.
  exprs <- tryCatch(as.list(parse(text = code_no_comments)), error = function(e) NULL)
  if (is.null(exprs))
    fail("Your code could not be parsed. Check for a missing comma or bracket.")
  strip_assign <- function(x)
    if (is.call(x) && identical(deparse(x[[1]]), "<-") && length(x) == 3L) x[[3]] else x
  pcalls <- lapply(Filter(function(x) {
    y <- strip_assign(x); is.call(y) && identical(deparse(y[[1]]), "predict")
  }, exprs), strip_assign)
  if (length(pcalls) != 2)
    fail(paste0("Write two `predict()` calls, one per profile. Your code has ", length(pcalls), "."))
  got <- suppressWarnings(vapply(pcalls, function(cl) {
    v <- tryCatch(as.numeric(eval(cl))[1], error = function(e) NA_real_)
    if (length(v) == 0L) NA_real_ else v
  }, numeric(1)))
  want <- as.numeric(predict(model_ghg, type = "response", newdata = data.frame(
    ideology_conservative2024 = c(20, 32), econ_poverty2022 = 12.5,
    gov_f = factor("Democrat", levels = levels(statesPolicy$gov_f)))))
  for (i in 1:2)
    if (is.na(got[i]) || abs(got[i] - want[i]) > 0.005)
      fail(paste0("Profile ", i, " does not give the expected probability. Check that every value in that call matches the profile described, and that only **ideology_conservative2024** differs between the two."))

  pass("Right. Both predictions computed: about 94% for Profile 1 and about 31% for Profile 2, the two profiles differing only in conservative identification.")
})
```

You now have two points anchoring the ideology relationship. Before moving to the full curve, think about what the gap between them implies about where these values fall on the S-curve.

**Practice 19: What do these two values tell you about the S-curve?**

Look at both predicted probabilities in your output before answering.

```{r quiz-recall-ghg, echo=FALSE}
question("Look at the two predicted probabilities in your output. What do they tell you about where these profiles fall on the S-curve?",
  correct = "Right. A 63-point drop in the steep middle of the curve.",
  answer("We are in the steep middle of the S-curve, where the ideology relationship is most sensitive to change",
         correct = TRUE,
         message = "A 63-percentage-point drop for a 12-point shift is dramatic --- it tells you this comparison passes through the steepest section of the S-curve. Two points describe only the stretch between them, not what a 12-point shift would do elsewhere in the ideology range."),
  answer("We are near the flat top of the S-curve because Profile 1 is close to 100%",
         message = "Profile 1 is near the top, but a 63-point drop to Profile 2 means we passed through the steep middle between them. The comparison spans both a high-probability region and the steep descending section."),
  answer("The two values fully describe the ideology relationship --- a plot would add nothing beyond what these two numbers show",
         message = "Two points describe the relationship between those specific values but not the full curve: where it flattens at the extremes, exactly where the steepest section falls, or how the curve shifts for different poverty levels. The plot shows all of that."),
  answer("The large drop means we must be near the flat bottom of the S-curve",
         message = "A 63-point drop is the opposite of a flat region. Flat regions of the S-curve are where even large predictor changes produce almost no probability change. We are in the steep middle."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a 63-point change implies about the steepness of the curve at these values."
)
```

## How Much Does Conservative Identification Change the Probability?

States with more conservative public opinion are far less likely to have adopted emissions caps, even after controlling for governor party and poverty rate. This relationship --- large, significant, and central to the hypothesis --- is the one to visualize first. The prediction plot shows the full S-curve across the observed range of conservative identification (14% to 50%), making visible what the coefficient of −0.314 in log-odds units could not communicate on its own.

The call has the same four parts you used in Example 1, and two optional arguments that matter here. The model comes first, `type = "pred"` asks for predicted values, `terms` names the variable that sweeps across the x-axis --- with `[all]` inside the quotation marks when it is interval --- and `condition` overrides what the other predictors are held at. Everything else is labeling: `title` and `axis.title` go inside the call, while the legend title and the caption are added afterwards in a `labs()` layer joined by `+`.

**Example code:**

```{r plot-model-logistic-template, exercise=FALSE, eval=FALSE}
plot_model(yourmodel,
           type       = "pred",
           terms      = "predictor_for_the_x_axis [all]",
           condition  = list(other_predictor = value_to_hold_it_at),
           title      = "What the plot shows",
           axis.title = c("x-axis label", "y-axis label")) +
  labs(color   = "Legend title",
       caption = "What the other predictors were held at.")
```

`condition` is only needed when you want something other than the defaults, and `color` only when a second variable in `terms` produces separate curves. The caption is not decoration: it is the only place a reader learns what everything else was held at.

**Practice 20: Produce a prediction plot for conservative identification**

The prescribed labels and caption are below.

- `title`: \"Predicted Probability of Greenhouse Gas Cap Adoption by State Ideology\"
- x-axis: \"Conservative Identification (%)\"
- y-axis: \"Predicted Probability of GHG Cap Adoption\"
- caption: \"Predictions are for Democrat-governed states with average poverty rate (12.4%).\"

**ideology_conservative2024** is an interval variable, so name it with `[all]` inside the quotation marks.

Write the complete code from scratch. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-ideology, exercise=TRUE, exercise.cap="Practice 20", fig.alt="Predicted probability of greenhouse gas cap adoption against conservative identification, with the horizontal axis running from about 14 to 50 percent. The curve is high at the left, falls fastest through the middle, and is low at the right, with a shaded band following it."}

```

```{r exercise-predplot-ideology-hint-1}
# Hint 1 of 3
# The task is to show how the probability of emissions cap adoption
# changes as state ideology varies across its full range. This is
# the same kind of plot you produced for college graduation rates
# in Example 1 --- a continuous relationship across a range ---
# but now for a different model and a different x-axis variable.
# Think about what the caption should say about what is held constant.
```

```{r exercise-predplot-ideology-hint-2}
# Hint 2 of 3
# You need plot_model() with the GHG model as the first argument.
# The argument that controls what type of plot you get takes
# "pred" for a prediction plot. The argument that names the
# x-axis variable is terms =. The argument that labels both axes
# is axis.title = (x-axis label first, then y-axis). Add the
# caption with + labs(caption = "..."). No condition argument
# is needed --- the defaults match the caption.
```

```{r exercise-predplot-ideology-hint-3}
# Hint 3 of 3
# Your completed call should be: plot_model() with model_ghg,
# type = "pred", terms = "ideology_conservative2024 [all]", the
# prescribed title, axis.title = with the two prescribed labels,
# and + labs(caption = "...") with the prescribed caption.
# The title and labels are all given in the prose above.
```

```{r exercise-predplot-ideology-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_ghg\\b", code_no_comments))
    fail("Use **model_ghg** as the model in `plot_model()`.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"` to produce a prediction plot.")
  if (!grepl("ideology_conservative2024", code_no_comments))
    fail("Set `terms = \"ideology_conservative2024\"`.")
  if (!grepl("ideology_conservative2024\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Add the bracket notation inside the quotes: `terms = \"ideology_conservative2024 [all]\"`.")
  if (!grepl("axis\\.title\\s*=", code_no_comments))
    fail("Include the `axis.title =` argument with the prescribed labels.")
  if (!grepl("Conservative Identification (%)", code_no_comments, fixed = TRUE))
    fail("Use the prescribed x-axis label: \"Conservative Identification (%)\".")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer stating the held values.")
  if (!grepl("Democrat-governed", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption specifying Democrat-governed states and the average poverty rate.")
  if (!grepl("Predicted Probability of Greenhouse Gas Cap Adoption by State Ideology", code_no_comments, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Probability of Greenhouse Gas Cap Adoption by State Ideology\".")
  if (!grepl("Predicted Probability of GHG Cap Adoption", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of GHG Cap Adoption\".")
  if (!grepl("Predictions are for Democrat-governed states with average poverty rate (12.4%).", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for Democrat-governed states with average poverty rate (12.4%).\".")

  # The checks above find the required pieces somewhere in the submission; this
  # one establishes they belong to the plot actually produced.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument sits inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one --- check it is set inside this plot rather than elsewhere in your code."))
  }
  chk("x", "Conservative Identification (%)", "x-axis label")
  chk("y", "Predicted Probability of GHG Cap Adoption", "y-axis label")

  pass("Good work. The curve is steepest through the middle of the ideology range and flattens at both ends --- the S-shape the coefficient alone could not show.")
})
```

This is the most important plot in the tutorial --- the one that shows what the ideology coefficient actually means. Work through all five components before answering: trace the shape of the curve, read the caption, identify two specific values, assess the band, and connect the pattern to the hypothesis.

**Practice 21: Use the five-component framework**

Look at the shape of the curve, the caption, the specific predicted values at key points, the confidence band, and the connection to the hypothesis before answering. Select all that apply.

```{r quiz-ideology-framework, echo=FALSE}
question("Which of the following statements correctly apply the five-component framework to this plot? Select all that apply.",
  correct = "Right: the pattern, held values, specific values, uncertainty, and hypothesis connection are all addressed.",
  answer("The curve drops steeply from near certainty at low conservative identification to near zero at high conservative identification, tracing a clear S-shape --- the pattern is strongly negative and nonlinear",
         correct = TRUE,
         message = "The S-curve drops from about 99% at ideology = 15 to less than 1% at ideology = 50, passing through the steep middle between roughly ideology = 20 and ideology = 35. This is the nonlinear pattern that makes logistic regression prediction plots more informative than the coefficient alone."),
  answer("The predictions hold governor party at Democrat and poverty rate at its mean (12.4%) --- conditioning on a Republican governor or a higher poverty rate would shift the curve downward, and the steep portion would occur at lower ideology values",
         correct = TRUE,
         message = "The held values are not a technicality --- they define which version of the ideology relationship the plot is showing. A state with a Republican governor and high poverty is already at a lower baseline probability, so the same shift in ideology produces a different-looking curve. The condition exercise coming next demonstrates this directly."),
  answer("At conservative identification = 20%, the model predicts about 94% probability of adoption; at 35%, about 16% --- a 78-point drop across a 15-point ideology shift that falls in the steep middle of the curve",
         correct = TRUE,
         message = "These two specific values anchor the pattern in concrete numbers. The gap between them (78 points) is striking --- it makes visible what the coefficient of −0.314 in log-odds units could not. Moving from a liberal state to an average-ideology state in this model is associated with an enormous decline in the predicted probability of emissions cap adoption."),
  answer("In this plot, the confidence band is widest through much of the steep middle portion, where the model has the most uncertainty about exactly how fast the probability drops",
         correct = TRUE,
         message = "Prediction uncertainty in logistic regression is typically largest in the steep middle of the S-curve, where the predicted probabilities are changing most rapidly and small changes in the underlying parameters produce large changes in the predicted value. The narrowing at the extremes reflects the compression of the probability scale near 0 and 1."),
  answer("Because the coefficient is −0.314 and significant, we can conclude that a one-unit increase in conservative identification decreases the probability of adoption by 0.314 percentage points",
         message = "This is the most common misinterpretation of a logistic regression coefficient. The coefficient is in log-odds units --- it does not translate directly to a probability change. The size of the probability change depends entirely on where you are on the S-curve, as the plot makes clear: the same one-unit shift produces a very large probability change in the steep middle and almost no change near the extremes."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements. Check the pattern, held values, specific predicted values, uncertainty band, and hypothesis connection against the plot."
)
```

Poverty rate may shape the feasibility of emissions cap adoption independently of ideology. Wealthier states --- those with lower poverty rates --- may be better positioned to absorb the economic costs of climate regulation, making adoption more likely regardless of ideological leaning. Conditioning on a low-poverty state (poverty = 8%, near the minimum observed) vs. a high-poverty state (poverty = 18%, near the maximum) lets us see whether the ideology relationship looks different in these two economic contexts.

**Practice 22: The same plot, conditioned on poverty = 8%**

Fill in the blanks. The `condition` argument and the caption are the only changes --- the caption has to tell a reader what this plot holds constant, so the one from the previous exercise will not do.

- caption: \"Predictions are for Democrat-governed states with poverty rate = 8%.\"

Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-ideology-poverty, exercise=TRUE, exercise.cap="Practice 22", fig.alt="The same relationship with the poverty rate held at 8 percent. The curve is high at the left, falls through the middle of the range, and is low at the right, with a shaded band around it."}
plot_model(model_ghg,
           type      = "pred",
           terms     = "ideology_conservative2024 [all]",
           ___       = list(econ_poverty2022 = ___),
           title     = "Predicted Probability of GHG Cap Adoption by State Ideology",
           axis.title = c("Conservative Identification (%)",
                          "Predicted Probability of GHG Cap Adoption")) +
  labs(caption = "___")
```

```{r exercise-predplot-ideology-poverty-hint-1}
# Hint 1 of 2
# The task is the same as the previous ideology plot, but you need
# to hold the poverty rate at 8 rather than its mean. Think about
# which argument controls the values of explanatory variables that
# are not on the x-axis, what structure that argument uses, and
# what the name of the poverty variable is in the model formula.
# Remember to update the caption to reflect the new held value.
```

```{r exercise-predplot-ideology-poverty-hint-2}
# Hint 2 of 2
# The condition argument takes a named list where each name is
# an explanatory variable and each value is what you want it
# held at: condition = list(variable_name = value).
# Fill in the argument name and the value for econ_poverty2022.
# Prescribed caption:
# "Predictions are for Democrat-governed states with poverty rate = 8%."
```

```{r exercise-predplot-ideology-poverty-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_ghg\\b", code_no_comments))
    fail("Pass **model_ghg** as the model object.")
  if (!grepl("condition\\s*=", code_no_comments))
    fail("Include the `condition =` argument to specify the held value for **econ_poverty2022**.")
  if (!grepl("econ_poverty2022\\s*=\\s*8", code_no_comments))
    fail("Set `econ_poverty2022 = 8` inside condition = list(...).")
  if (!grepl("ideology_conservative2024", code_no_comments))
    fail("Keep `terms = \"ideology_conservative2024\"`.")
  if (!grepl("ideology_conservative2024\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Keep the bracket notation inside the quotes: `terms = \"ideology_conservative2024 [all]\"`.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer with an updated caption reflecting poverty = 8%.")
  if (!grepl("poverty rate = 8%", code_no_comments, fixed = TRUE))
    fail("Update your caption to state that the poverty rate is held at 8% --- the default caption is not correct here.")
  if (!grepl("Predictions are for Democrat-governed states with poverty rate = 8%.", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for Democrat-governed states with poverty rate = 8%.\".")

  pass("Correct. The plot now holds **econ_poverty2022** at 8 rather than at its mean, and the caption says so.")
})
```

You have now seen the ideology relationship in two economic contexts. The question is not just whether the curves differ but how --- and what that difference reveals about how ideology and poverty jointly shape the predicted probability of climate policy adoption.

**Practice 23: Compare the two ideology plots**

Look at both plots --- the default (mean poverty = 12.4%) and the conditioned (poverty = 8%) --- before answering.

```{r quiz-ideology-poverty, echo=FALSE}
question("Compare the two ideology plots. Which statement best describes what changed and why?",
  correct = "Right. Lower poverty shifts the curve upward, keeping high-probability adoption visible further into the ideology range.",
  answer("The curve shifted upward at lower poverty: at any ideology level a low-poverty state is likelier to have adopted a cap",
         correct = TRUE,
         message = "At ideology = 35 (the mean), the default curve gives about 16% while the poverty = 8% curve gives about 67% --- a 51-point difference. At ideology = 25, the gap is about 21 points (79% vs. 98%). The shift is largest in the steep middle, where conditioning on lower poverty moves us higher on the S-curve. This is the condition argument in action: poverty and ideology together shape the predicted probability, and holding poverty constant at an unusually low value reveals a version of the ideology relationship that the default plot, averaging across all poverty levels, cannot show."),
  answer("The curves are essentially identical --- poverty rate has no effect on the ideology relationship",
         message = "The curves differ substantially, especially in the middle of the ideology range. Compare the predicted probability at ideology = 35 in each plot --- the difference is about 51 percentage points."),
  answer("The curve shifted downward --- wealthier states are less likely to adopt emissions caps because they have more to lose economically",
         message = "The curve shifted upward, not downward. Lower poverty (wealthier states) is associated with a higher predicted probability of adoption at any ideology level, consistent with the negative poverty coefficient in the model."),
  answer("The curve shifted upward, which shows that reducing poverty would raise the probability that a state adopts an emissions cap",
         message = "The two curves describe two kinds of state, not the same state before and after a change. Conditioning selects which profile the plot represents; it does not simulate an intervention. Whether reducing poverty would change adoption is a causal question this observational model cannot answer."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: compare the y-axis values at ideology = 35 in the two plots. Did the curve shift up or down, and does the size of the shift vary across the ideology range?"
)
```

## Does the Governor's Party Matter?

**gov_f** is a two-category factor with Democrat as the reference. `plot_model()` plots the predicted probability for each level with a confidence interval. Both predicted probabilities will be low because the plot holds ideology at its mean (35.78%) --- the average state is quite conservative, which suppresses the baseline probability for both parties. Their positions tell you the governor party effect in a typical state; the confidence intervals show the uncertainty around each predicted probability. Whether the two parties differ is settled by the coefficient test in Tutorial 20, not by comparing the intervals.

**Practice 24: Produce a prediction plot for governor party**

Prescribed values:
- `title = "Predicted Probability of GHG Cap Adoption by Governor Party"`
- x-axis: \"Governor Party\", y-axis: \"Predicted Probability of GHG Cap Adoption\"
- caption: \"Predictions are for states with average conservative identification (35.8%) and poverty rate (12.4%).\"

Write the complete code from scratch. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-govparty, exercise=TRUE, exercise.cap="Practice 24", fig.alt="Two predicted probabilities, one for each governor party, each drawn with a vertical confidence interval, on a vertical axis spanning roughly 0 to 45 percent."}

```

```{r exercise-predplot-govparty-hint-1}
# Hint 1 of 3
# The task is to compare the predicted probability of emissions cap
# adoption between Democrat-governed and Republican-governed states,
# holding ideology and poverty at their mean values. You produced
# a very similar plot for the rurality variable in Example 1 ---
# the structure is the same, just with a different model and a
# different factor variable on the x-axis.
```

```{r exercise-predplot-govparty-hint-2}
# Hint 2 of 3
# You need plot_model() with the GHG model as the first argument,
# type = "pred", and terms = naming the governor party variable.
# When terms names a two-level factor, plot_model() plots one
# predicted probability per level, each with a confidence interval. Add axis.title = with the prescribed
# labels (x first, y second) and + labs(caption = "...").
# No condition argument is needed.
```

```{r exercise-predplot-govparty-hint-3}
# Hint 3 of 3
# Your completed call should be: plot_model() with model_ghg,
# type = "pred", terms = "gov_f", the prescribed title,
# axis.title = with "Governor Party" first and the y-axis label
# second, and + labs(caption = "...") with the prescribed caption.
# All labels and the caption are given in the prose above.
```

```{r exercise-predplot-govparty-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_ghg\\b", code_no_comments))
    fail("Use **model_ghg** as the model in `plot_model()`.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"`.")
  if (!grepl("gov_f", code_no_comments))
    fail("Set `terms = \"gov_f\"`.")
  if (!grepl("Governor Party", code_no_comments, fixed = TRUE))
    fail("Use \"Governor Party\" as the x-axis label.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer.")
  if (!grepl("average conservative identification", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption specifying average conservative identification and poverty rate.")
  if (!grepl("Predicted Probability of GHG Cap Adoption", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of GHG Cap Adoption\".")

  # The checks above find the required pieces somewhere in the submission; this
  # one establishes they belong to the plot actually produced.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument sits inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one --- check it is set inside this plot rather than elsewhere in your code."))
  }
  chk("x", "Governor Party", "x-axis label")
  chk("y", "Predicted Probability of GHG Cap Adoption", "y-axis label")

  pass("Right. A factor on the x-axis gives one predicted probability per governor party, each with its own confidence interval.")
})
```

The governor party plot will look surprising if you do not read the caption first. Before answering, check where the predicted probabilities fall on the y-axis and connect that position to the held values stated in the caption.

**Practice 25: Interpret the governor party plot**

Look at both predicted probabilities and their confidence intervals before answering.

```{r quiz-govparty-interpret, echo=FALSE}
question("Look at the governor party prediction plot. Which statement best describes the key finding?",
  correct = "Right. Both predicted probabilities are low and close together at these held values, and the coefficient test is a separate matter.",
  answer("Both predicted probabilities are low because mean ideology places the average state near the bottom of the S-curve",
         correct = TRUE,
         message = "Democrat-governed states at mean ideology and poverty have a predicted probability of about 14%; Republican-governed states about 4%. Both are low because ideology = 35.78 places these states in a conservative context where cap adoption is unlikely regardless of who governs. Tutorial 20 separately found the governor-party coefficient not statistically distinguishable from zero, but that is not what the small gap here shows. A gap this size is what the S-curve produces wherever both predictions sit near one of its flat ends --- a substantial log-odds coefficient would still yield a small probability difference at this held value. The coefficient test, not the visual gap, determines statistical significance."),
  answer("The Democrat predicted probability is so much higher than the Republican one that governor party is clearly the most important predictor in the model",
         message = "The Democrat predicted probability is higher, but by a small absolute amount (about 10 percentage points), and the governor party coefficient was not distinguishable from zero in Tutorial 20. The governor-party coefficient is estimated imprecisely enough that the data do not distinguish it from zero, while ideology produces a large difference in predicted probability across the range shown here. The small gap here partly reflects the fact that mean ideology (35.8%) is conservative, suppressing both probabilities."),
  answer("The wide confidence intervals mean the model cannot predict emissions cap adoption at all",
         message = "Wide confidence intervals for the factor plot at mean ideology reflect two things: the small number of states (50) and the fact that mean ideology places states in a region of moderate uncertainty. The ideology S-curve shows the estimated probabilities becoming very high at one end and very low at the other; that is separate from how accurately individual states are classified."),
  answer("The two predicted probabilities are close together, which shows governor party has no relationship with adoption once ideology and poverty are controlled",
         message = "How far apart two predictions look is not a test. The governor party coefficient is estimated, and its test is what settles whether it is distinguishable from zero. A small visible gap at one held value is also compatible with a substantial coefficient, because both predictions sit low on the S-curve where differences in log-odds compress."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at both predicted probabilities and how close together they are. Connect what you see to the p-value for the **gov_f** coefficient from Tutorial 20."
)
```

## How Do Ideology and Governor Party Work Together?

A two-variable prediction plot puts **ideology_conservative2024** on the x-axis and produces separate S-curves for Democrat- and Republican-governed states. Because the model has no interaction between ideology and governor party, the curves will not cross --- on the log-odds scale, the ideology coefficient is identical in both types of states, with the Republican curve shifted downward by the governor party coefficient. On the probability scale, the apparent gap between curves may narrow near the extremes due to S-curve compression. The visual gap between the two curves shows the governor party effect across the full ideology range.

**Practice 26: Produce a two-variable prediction plot**

Prescribed values:
- `title = "Predicted Probability of GHG Cap Adoption by Ideology and Governor Party"`
- x-axis: \"Conservative Identification (%)\", y-axis: \"Predicted Probability of GHG Cap Adoption\"
- legend: \"Governor Party\"
- caption: \"Predictions are for states with average poverty rate (12.4%).\"

**ideology_conservative2024** goes on the x-axis and is interval, so it takes `[all]`; **gov_f** is a factor and does not.

Write the complete code from scratch. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-twovar-ghg, exercise=TRUE, exercise.cap="Practice 26", fig.alt="Two curves, one per governor party, showing predicted probability of greenhouse gas cap adoption against conservative identification. Each falls from high at the left to low at the right, each with a shaded band, and a legend names the two parties."}

```

```{r exercise-predplot-twovar-ghg-hint-1}
# Hint 1 of 3
# The task is to show the ideology relationship separately for
# Democrat-governed and Republican-governed states on the same
# plot. Think about which variable should vary continuously along
# the x-axis and which should produce separate S-curves for each
# of its levels. Because there is no interaction, predict what
# the relationship between the two curves will look like.
```

```{r exercise-predplot-twovar-ghg-hint-2}
# Hint 2 of 3
# The structure is the same as the two-variable plot you produced
# in Example 1 for education and rurality: plot_model() with the
# model, type = "pred", terms = a character vector with the x-axis
# variable first and the grouping variable second, axis.title =
# with the prescribed labels, and + labs(caption = "...").
# The only changes are the model and the variable names.
```

```{r exercise-predplot-twovar-ghg-hint-3}
# Hint 3 of 3
# terms takes both variables in one character vector, x-axis
# variable first. Only the x-axis variable is interval, so only
# that one carries the bracket notation. The title, both axis
# labels, the legend title and the caption are all prescribed
# in the prose above.
```

```{r exercise-predplot-twovar-ghg-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_ghg\\b", code_no_comments))
    fail("Use **model_ghg** as the model in `plot_model()`.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_no_comments))
    fail("Set `type = \"pred\"`.")
  if (!grepl("ideology_conservative2024", code_no_comments))
    fail("Include **ideology_conservative2024** as the first term (x-axis).")
  if (!grepl("ideology_conservative2024\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Add the bracket notation on the x-axis variable: c(\"ideology_conservative2024 [all]\", \"gov_f\").")
  if (!grepl("gov_f", code_no_comments))
    fail("Include **gov_f** as the second term (separate curves).")
  if (regexpr("ideology_conservative2024", code_no_comments) > regexpr("gov_f", code_no_comments))
    fail("ideology_conservative2024 should come first in terms (x-axis), and **gov_f** second (separate curves). Check the order inside c(...).")
  if (!grepl("axis\\.title\\s*=", code_no_comments))
    fail("Include the `axis.title =` argument.")
  if (!grepl("labs\\s*\\(", code_no_comments))
    fail("Add a labs(`caption = \"...\"`) layer.")
  if (!grepl("Governor Party", code_no_comments, fixed = TRUE))
    fail("Name the legend with the prescribed label: labs(`color = \"Governor Party\"`, ...).")
  if (!grepl("average poverty rate", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption specifying the average poverty rate.")
  if (!grepl("Conservative Identification (%)", code_no_comments, fixed = TRUE))
    fail("Use the prescribed label: \"Conservative Identification (%)\".")
  if (!grepl("Predicted Probability of GHG Cap Adoption", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of GHG Cap Adoption\".")
  if (!grepl("Predictions are for states with average poverty rate (12.4%).", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with average poverty rate (12.4%).\".")

  # The checks above find the required pieces somewhere in the submission; this
  # one establishes they belong to the plot actually produced.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument sits inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one --- check it is set inside this plot rather than elsewhere in your code."))
  }
  chk("x", "Conservative Identification (%)", "x-axis label")
  chk("y", "Predicted Probability of GHG Cap Adoption", "y-axis label")
  chk("colour", "Governor Party", "legend title")
  if (length(unique(bl$data[[1]]$group)) != nlevels(statesPolicy$gov_f))
    fail("The plot should draw one line per category. Check the second variable in `terms`.")

  pass("Exactly. Two curves, one per governor party, drawn from a model with no interaction between ideology and party.")
})
```

The two-variable plot shows the same additive structure as the governor party factor plot, but now across the full ideology range. Before answering, observe whether the curves cross and what the gap between them represents.

**Practice 27: Interpret the two-variable plot**

Look carefully at the two S-curves before answering.

```{r quiz-twovar-ghg, echo=FALSE}
question("Look at the two-variable prediction plot. Which statement best describes what it shows?",
  correct = "Right. Non-crossing S-curves in an additive model, with the Republican line consistently below the Democrat line.",
  answer("The two S-curves do not cross, and the Republican curve runs consistently below the Democrat curve",
         correct = TRUE,
         message = "This fitted model is additive, so its curves cannot cross on the probability scale --- though non-crossing curves alone would not establish that a model is additive. At ideology = 25, Democrat-governed states have about 79% predicted probability while Republican-governed states have about 50% --- a 29-point gap. At ideology = 35, the gap narrows to about 11 points (16% vs. 5%). The narrowing gap near the extremes is not because the model treats them differently --- it is S-curve compression. On the log-odds scale, the ideology coefficient is identical in both types of states, and governor party shifts the baseline."),
  answer("The two curves cross somewhere in the middle, indicating that Republican governors are actually better for climate policy at high ideology levels",
         message = "The curves do not cross --- the Democrat curve is above the Republican curve throughout. Both approach zero at high ideology, but the Democrat curve remains higher at every point."),
  answer("Non-crossing curves would indicate a stronger ideology effect in one party than the other --- but because the model has no interaction, we cannot test this",
         message = "Non-crossing curves do not indicate a stronger ideology effect in one party --- in this additive model the log-odds slope is the same for both, which is why the curves cannot cross. We can test whether that slope differs by governor party by adding an interaction term: the coefficient test addresses whether the slopes are statistically distinguishable, and whether the resulting probability curves cross is a separate feature of the fitted predictions. The curves here are the model's claim about the data, not a limitation of the visualization."),
  answer("The gap between the curves is widest in the middle of the ideology range, showing the governor party effect is strongest for moderately conservative states",
         message = "The gap is widest in the middle because that is where the probability scale is least compressed, not because the effect is larger there. On the log-odds scale the party difference is identical at every ideology level --- the model contains no interaction, so it could not vary."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the model is additive. Given that structure, should the curves cross? Where is the gap between them largest, and why does it narrow near the extremes?"
)
```

## Does the Picture Change in Low-Poverty States?

The plot you just produced holds poverty rate at its mean, 12.4%. That is one context, and the caption says so --- but it is not the only one worth seeing. Adding `condition` to a two-variable plot lets you ask the same question in a different setting: what do the two governor-party curves look like in a high-poverty state?

This is the combination the tutorial has been building toward. `terms` decides what varies and what separates the curves; `condition` decides the context they sit in. Both go in the same `plot_model()` call.

**Practice 28: Condition a two-variable prediction plot**

Produce the same plot from **model_ghg** again, this time conditioning on **econ_poverty2022** = 18. Prescribed labels:

- `title = "Predicted Probability of GHG Cap Adoption by Ideology and Governor Party"`
- x-axis: \"Conservative Identification (%)\", y-axis: \"Predicted Probability of GHG Cap Adoption\"
- legend: \"Governor Party\"
- caption: \"Predictions are for states with a poverty rate of 18%.\"

**ideology_conservative2024** goes on the x-axis and is interval, so it takes `[all]`; **gov_f** is a factor and does not.

Write the complete code from scratch. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-twovar-ghg-cond, exercise=TRUE, exercise.cap="Practice 28", fig.alt="Two curves, one per governor party, showing predicted probability of greenhouse gas cap adoption against conservative identification, with the poverty rate held at 18 percent. Each falls from high at the left to low at the right, each with a shaded band, and a legend names the two parties."}

```

```{r exercise-predplot-twovar-ghg-cond-hint-1}
# Hint 1 of 3
# This combines two things you have done separately: a plot with
# two variables in it, and a plot where you chose the context
# rather than accepting the default. Think about which argument
# does each job, and what the caption now needs to tell a reader.
```

```{r exercise-predplot-twovar-ghg-cond-hint-2}
# Hint 2 of 3
# Start from the call you just wrote for the unconditioned plot.
# One argument is added: the one that overrides a default held
# value. It takes a named list, naming the variable and the value
# to hold it at. The caption changes to match.
```

```{r exercise-predplot-twovar-ghg-cond-hint-3}
# Hint 3 of 3
# The added argument is condition, and it sits alongside terms
# and axis.title inside the same plot_model() call. It takes a
# named list pairing the poverty variable with the value the
# prompt gives. Everything else is unchanged except the caption
# text.
```

```{r exercise-predplot-twovar-ghg-cond-check}
grade_this({
  code             <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  if (grepl("=\\s*[TF](?![A-Za-z._0-9])", code_no_comments, perl = TRUE)) {
    fail("Write `TRUE` and `FALSE` in full. R accepts `T` and `F` as shorthand, but these tutorials do not: `TRUE` and `FALSE` are reserved words and always mean what you expect, while `T` and `F` are ordinary object names that other code can overwrite --- after which an argument set to `T` or `F` quietly means something else. Replace them with `TRUE` and `FALSE` and submit again.")
  }

  if (!grepl("plot_model\\s*\\(", code_no_comments))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("plot_model\\s*\\(\\s*model_ghg\\b", code_no_comments))
    fail("Pass **model_ghg** as the model object.")
  if (!grepl("condition\\s*=", code_no_comments))
    fail("Add the argument that overrides a default held value.")
  if (!grepl("econ_poverty2022", code_no_comments))
    fail("The variable to hold at a chosen value is **econ_poverty2022**.")
  if (!grepl("18", code_no_comments))
    fail("Hold **econ_poverty2022** at 18.")
  if (!grepl("terms\\s*=", code_no_comments))
    fail("Keep the terms argument that puts ideology on the x-axis and separates the curves by governor party.")
  if (!grepl("gov_f", code_no_comments))
    fail("Include **gov_f** as the second term, so the plot still shows separate curves by governor party.")
  if (!grepl("ideology_conservative2024\\s*\\[\\s*all\\s*\\]", code_no_comments))
    fail("Add the bracket notation on the x-axis variable: c(\"ideology_conservative2024 [all]\", \"gov_f\").")
  if (!grepl("Governor Party", code_no_comments, fixed = TRUE))
    fail("Name the legend with the prescribed label: labs(`color = \"Governor Party\"`, ...).")
  if (!grepl("poverty rate of 18%", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption, which tells a reader the new held value.")
  if (!grepl("Conservative Identification (%)", code_no_comments, fixed = TRUE))
    fail("Use the prescribed label: \"Conservative Identification (%)\".")
  if (!grepl("Predicted Probability of GHG Cap Adoption", code_no_comments, fixed = TRUE))
    fail("Use the prescribed y-axis label: \"Predicted Probability of GHG Cap Adoption\".")
  if (!grepl("Predictions are for states with a poverty rate of 18%.", code_no_comments, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with a poverty rate of 18%.\".")

  # The checks above find the required pieces somewhere in the submission; this
  # one establishes they belong to the plot actually produced.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument sits inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one --- check it is set inside this plot rather than elsewhere in your code."))
  }
  chk("x", "Conservative Identification (%)", "x-axis label")
  chk("y", "Predicted Probability of GHG Cap Adoption", "y-axis label")
  chk("colour", "Governor Party", "legend title")
  if (length(unique(bl$data[[1]]$group)) != nlevels(statesPolicy$gov_f))
    fail("The plot should draw one line per category. Check the second variable in `terms`.")

  pass("Well done. The conditioned two-variable plot is the last of the four shapes: two curves, both at a poverty rate you chose rather than the default.")
})
```

Compare the two plots. Every prediction is lower in the high-poverty context, because poverty carries a negative coefficient and shifts the whole linear combination down --- which moves both curves into a different part of the S-curve. The ideology coefficient has not changed and neither has the governor-party coefficient; what changes is where on the curve those fixed coefficients are being evaluated. That is the difference between logistic and linear regression in one picture: in a linear model the two lines would simply drop by a constant, keeping their shape and their separation.

### What We Learned About Ideology, Poverty, and Climate Policy

The prediction plots translate Tutorial 20's coefficients into a complete picture of how state ideology, governor party, and economic conditions jointly shape the predicted probability of emissions cap adoption. The ideology S-curve is the central finding: a coefficient of −0.314 in log-odds units becomes a 78-point drop in predicted probability across the steep middle of the ideology range --- from near certainty in liberal states to near zero in conservative ones. The poverty condition exercise shows the same additive log-odds relationship looking different on the probability scale in different economic contexts: a low-poverty state (poverty = 8%) at ideology = 35 has a predicted probability of about 67% --- four times higher than the same state at mean poverty. The governor party factor plot shows both parties clustered at low probability at mean ideology, with a small gap between them --- a gap that size is what the flat end of the S-curve produces, whatever the coefficient, so the coefficient test rather than the visual gap determines statistical significance. The non-crossing S-curves in the two-variable plot display the fitted additive structure: the log-odds ideology coefficient is the same in Democrat- and Republican-governed states, with governor party shifting the baseline but not the log-odds slope.

Here is what a complete interpretation of the ideology prediction plot looks like written as prose: *The predicted probability of greenhouse gas cap adoption drops steeply as conservative identification increases, tracing a clear S-curve from near certainty at the low end of the ideology range to near zero at the high end. Predictions hold governor party at Democrat and poverty rate at its mean (12.4%); conditioning on a high-poverty state would shift the entire curve downward substantially, since poverty rate carries its own negative coefficient in the model. At conservative identification = 20%, the model predicts about 94% probability of adoption; at 35%, about 16% --- a 78-point drop across a 15-point ideology shift that falls in the steep middle of the curve. In this plot, the confidence band is widest through much of the steep middle region. This partly reflects compression near probabilities of 0 and 1, although confidence-band width also depends on where the observations are concentrated and how precisely the model parameters are estimated. This is strongly consistent with the hypothesis that ideology is associated with climate policy adoption, and the prediction plot makes visible what the significant coefficient of −0.314 could only hint at: across the steep middle of the range, the predicted difference is very large.*


## Check Your Understanding

The two examples showed that held values decide which part of the S-curve a prediction plot displays, that the shape of the curve carries information the coefficient table cannot, and that reading such a plot means reading its caption as carefully as its line. The questions below ask you to apply these ideas to new scenarios without looking back at the examples.

**Apply 29: Defend a change in slope**

A researcher fits a logistic model of whether a city has adopted a climate action plan, using population density, median income, and the share of residents with a college degree. She produces two prediction plots with population density on the x-axis: one at the means of the other predictors, and one conditioned on a high median income. The curve in the second is noticeably steeper. A reviewer objects that one plot must be wrong, because changing what you hold constant can only move a line up or down, not change its shape.

```{r quiz-cyu-steepness, echo=FALSE}
question("Who is right, and why?",
  correct = "Right. Held values decide which part of the S-curve you are looking at.",
  answer("The researcher. Held values decide where on the S-curve the predictions fall, and the curve is steeper in the middle than at its ends",
         correct = TRUE,
         message = "This is the difference between a logistic model and a linear one. In a linear model the reviewer would be correct: changing a held value shifts the line by that variable's coefficient and leaves the slope alone. In a logistic model the held values move the whole linear combination, which relocates the predictions along the S-curve --- and because the curve is steep in the middle and flat at both ends, the same coefficient produces a visibly different slope depending on where it lands."),
  answer("The reviewer. Holding a variable at a different value shifts the entire curve vertically by that variable's coefficient, so the two curves must have the same shape",
         message = "That is exactly right for linear regression, and exactly why this is the point at which logistic prediction plots stop behaving like the ones in Tutorial 17. On the log-odds scale the shift is constant; on the probability scale it is not, because the transformation from log-odds to probability is not a straight line."),
  answer("The reviewer, unless the model contains an interaction between population density and median income --- only an interaction can make the relationship steeper in one context than another",
         message = "An interaction would make the log-odds slope differ by context. Here the log-odds slope is identical in both plots; what differs is where on the S-curve those identical log-odds are being converted into probabilities. Different steepness on the probability scale is expected in an additive logistic model and does not require an interaction term."),
  answer("The researcher, because conditioning on a high median income re-estimates the model using only high-income cities, which produces different coefficients and so a different slope",
         message = "Conditioning does not re-estimate anything. The model is fitted once, on all the data; the condition argument only chooses the values at which the fitted coefficients are evaluated when the predictions are computed. Every coefficient is the same in both plots."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: ask whether the two curves come from different coefficients, or from the same coefficients evaluated somewhere else on the curve."
)
```

**Apply 30: Interpret converging curves**

In an additive logistic model of whether a bill received a committee hearing, a researcher plots the number of cosponsors on the x-axis with separate curves for majority-party and minority-party sponsors. The curves are far apart at low cosponsor counts and nearly meet at high ones. They do not cross.

```{r quiz-cyu-converging, echo=FALSE}
question("What does the narrowing gap tell you about the relationship between sponsor party and the probability of a hearing?",
  correct = "Right. The curves are running out of room, not changing their relationship.",
  answer("Nothing about a changing relationship --- both curves are approaching the ceiling of the probability scale, where little room for a gap remains",
         correct = TRUE,
         message = "On the log-odds scale the party difference is the same at every cosponsor count, because the model has no interaction term. Converting to probabilities compresses that constant difference into fewer and fewer percentage points as both curves approach 1. A gap that narrows near the top is what an additive logistic model looks like, and reading it as a substantive finding is the most common error in these plots."),
  answer("That sponsor party matters less for heavily cosponsored bills, since the two groups become nearly indistinguishable at high cosponsor counts",
         message = "This is the reading the plot invites and the model does not support. The estimated party difference in log-odds is unchanged across the range; only its expression in percentage points shrinks. To claim that party matters less for heavily cosponsored bills, the model would need an interaction term, and it has none."),
  answer("That the model contains an interaction between cosponsors and sponsor party, which is what allows the gap between the curves to change",
         message = "The question states the model is additive, and the curves do not cross. An interaction produces different log-odds slopes by group. Whether the resulting probability curves cross, converge or diverge depends on the fitted coefficients and the range being plotted."),
  answer("That sponsor party is not a significant predictor once the number of cosponsors is accounted for, since the curves end up in nearly the same place",
         message = "Significance is settled by the coefficient's test, not by how far apart two curves look at one end of the range. The curves are far apart through the middle, which is where the party coefficient is most visible on the probability scale."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a constant difference in log-odds looks like once it is converted to probabilities near the top of the scale."
)
```

**Apply 31: Identify what the band shows**

A prediction plot shows the probability that a country holds a competitive election as a function of GDP per capita, with a shaded band around the curve. The band is narrow at low GDP and much wider through the middle of the range.

```{r quiz-cyu-band-meaning, echo=FALSE}
question("What does the shaded band represent, and what does its changing width reflect?",
  correct = "Right. Uncertainty about the estimated probability, compressed near the bounds of the scale.",
  answer("Uncertainty in the estimated probability, whose width in percentage points shrinks near 0 and 1 because the probability scale is compressed there",
         correct = TRUE,
         message = "The band answers the question: given that these coefficients were estimated from a sample, how precisely does the model locate this curve? Its width in percentage points is not a constant, because the same uncertainty in the coefficients spans fewer points of probability near the bounds than it does through the steep middle. Band width is also affected by where the observations are concentrated and by how precisely each coefficient is estimated."),
  answer("The range within which the outcomes of individual countries fall, so it is narrow where nearly all countries share the same outcome",
         message = "The outcomes here are 0 or 1 --- a country either holds a competitive election or it does not --- so no interval strictly between them could contain an individual outcome. The band describes the estimated probability, which is a property of the model, not the outcomes themselves."),
  answer("The spread of the observed data around the fitted curve, so it is wider where countries vary more in whether they hold competitive elections",
         message = "That describes something closer to residual scatter, which is not what is drawn here. The band comes from the uncertainty in the estimated coefficients carried through to each predicted probability, and it would exist even if the model fitted the data perfectly."),
  answer("How well the model fits at each value, so a wide band means the model describes those countries poorly",
         message = "Band width is not a measure of fit. A model can describe its data badly and still produce a narrow band if its coefficients are precisely estimated, and the compression near 0 and 1 narrows the band regardless of how well the model performs in that region."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: ask what quantity the band is uncertain about --- the probability the model estimates, or the outcome of any particular country."
)
```

**Apply 32: Diagnose an incomplete interpretation**

A student writes the following interpretation of a prediction plot showing the relationship between a country's unemployment rate and the probability that the incumbent government loses re-election. Unemployment in the data ranges from about 2% to about 20%.

*"As the unemployment rate rises, the predicted probability that the incumbent loses increases. Each additional point of unemployment raises the predicted probability by about 4 percentage points. At an unemployment rate of 30%, defeat is essentially certain."*

```{r quiz-cyu-diagnose-logistic, echo=FALSE}
question("Which of the following best identifies what is wrong with this interpretation?",
  correct = "Right: a constant per-unit change, no held values, and a prediction beyond the data.",
  answer("It reports a constant change per unit, omits the held values, and reads a prediction outside the range of the data",
         correct = TRUE,
         message = "Three separate faults. A logistic curve has no single per-unit change --- the same point of unemployment moves the probability a great deal in the steep middle and almost not at all near either end, so \"about 4 percentage points\" can only be true somewhere, not everywhere. Without the held values a reader cannot tell which part of the curve is being described. And 30% lies beyond the observed range, so the model will return a number that nothing in the data supports."),
  answer("Nothing is wrong --- the direction is right, the magnitude is specific, and the interpretation ends with a concrete prediction",
         message = "Specificity is not the same as accuracy. A single per-unit figure misdescribes a curve whose slope changes across its range, and the concrete prediction at the end is at a value no country in the data reaches."),
  answer("The only problem is the missing held values; stating what the other predictors were fixed at would complete the interpretation",
         message = "Adding the held values is necessary and not sufficient. The constant per-unit claim would still misdescribe the shape of the curve, and the prediction at 30% would still be an extrapolation past the data."),
  answer("The only problem is the causal implication --- rewriting the claim as an association rather than an effect would complete the interpretation",
         message = "This interpretation is already phrased in terms of what the model predicts rather than what unemployment does, so causal language is not the fault here. The problems are the constant per-unit change, the missing held values, and the prediction beyond the observed range."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: check the claim about magnitude against the shape of an S-curve, and check the final prediction against the range the data cover."
)
```

## The Takeaways

Tutorial 17 introduced prediction plots for additive linear-regression models, while Tutorial 19 extended them to interaction models. This tutorial applies the same tools to logistic regression --- with one fundamental difference. In an additive linear model without interactions involving the predictor being displayed, changing the values held constant for the other explanatory variables shifts the predicted line up or down but leaves its slope unchanged. In a logistic model, the held values determine which part of the S-curve the plot is showing, which can change both the level and the steepness of the relationship. This is why choosing held values thoughtfully is part of the analysis in logistic regression, not a technicality.

**Why prediction plots matter more here.** In linear regression, the slope directly gives the expected difference in the outcome associated with a one-unit difference in the predictor --- the same everywhere along the range. The coefficients table gives you that quantity once, and it applies everywhere. In logistic regression, the coefficient gives you direction and significance but not size. The change in the predicted probability for a given shift in an explanatory variable depends on where on the S-curve the comparison falls, which depends on the values of every other variable in the model. The prediction plot is the primary tool for communicating the substantive size of model-implied probability differences --- not an illustration of something you already know from the table, but the place where interpretation actually happens.

**`plot_model()` from [sjPlot]{.package-name}.** The `plot_model()` function from the [sjPlot]{.package-name} package produces prediction plots from fitted models. For logistic regression, the key arguments are:

- `type = "pred"` to produce a prediction plot on the probability scale
- `terms =` to name the variable on the x-axis (a single string for a single-variable plot, or `c("x_variable", "group_variable")` for a two-variable plot)
- `axis.title =` to label the x-axis and y-axis, in that order
- `+ labs(caption = "...")` to add a caption describing the held values

When `terms =` names an interval variable, `plot_model()` produces a continuous S-curve. When `terms =` names a factor variable, it produces a predicted probability with a confidence interval for each level --- there is no continuous range to sweep across for a categorical variable, only discrete levels to compare.

**Bracket notation.** When `terms =` names an interval variable, `plot_model()` computes predictions at a handful of round values rather than at every observed value, and prints a message saying so. Adding `[all]` inside the quotation marks --- `"variable_name [all]"` --- computes a prediction at every value the variable takes, which both smooths the curve and confines it to the range the data actually cover. A factor has only its categories, so it needs no bracket notation and produces no message.

**The `condition` argument.** The values the other explanatory variables are fixed at are the plot's [held values]{.important-text}. By default `plot_model()` uses their means (for interval variables) or their reference category (for factor variables). The `condition` argument overrides those defaults: `condition = list(variable_name = value)` holds the named variable at the specified value while the x-axis variable varies. You can set multiple variables at once: `condition = list(variable1 = value1, variable2 = value2)`. The most useful condition values are **substantively meaningful** --- corresponding to real, recognizable cases rather than arbitrary numbers. When substantively justified values sit in different regions of the S-curve, comparing them shows how the same fitted relationship appears in different contexts. The caption is not a formality: it is essential information about what the plot is actually showing, and reading it carefully is part of applying the five-component framework.

**Reading non-crossing curves.** When `terms =` names two variables, `plot_model()` produces one curve per level of the grouping variable. In an additive logistic model --- one without an interaction term --- these curves will not cross: on the log-odds scale, the same coefficient for the x-axis variable applies in every group, and the vertical gap between the curves on the probability scale is determined by the group coefficient. The gap may appear to narrow near the extremes of the probability scale due to [S-curve compression]{.important-text}, but the curves themselves do not cross. Non-crossing curves are not a limitation; they are the model's claim on the probability scale. Curves that cross --- where one group starts below and ends above another --- require an interaction term in the model. If your theory predicts that the underlying log-odds relationship differs across groups --- not merely that the probability curves look steeper in one place than another, which an additive model already produces --- that prediction calls for an interaction, not just a different held value.

**The five-component framework.** Every prediction plot, regardless of what model produced it, can be read using the same five components introduced in Tutorial 17:

1. **The pattern** --- the direction and shape of the relationship as the x-axis variable increases
2. **The held values** --- what the other explanatory variables were set to, and what that means for how the result generalizes
3. **Specific predicted values** --- what the model predicts at substantively meaningful points along the x-axis
4. **The uncertainty** --- the width of the [confidence band]{.important-text} and whether it varies across the range
5. **The hypothesis connection** --- whether the pattern is consistent with the hypothesis that motivated the model, and what the plot adds beyond the significance test

In logistic regression, component 2 deserves particular attention: the held values determine which part of the S-curve you are seeing, which affects every other component.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-21, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-21, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-21, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now visualize predicted probabilities from logistic regression models — and you understand why those plots make the size and shape of model-implied probability differences visible, not just their direction.">
  <span aria-hidden="true">📈</span>
  Excellent work! You can now visualize predicted probabilities from logistic regression models &mdash; and you understand why those plots make the size and shape of model-implied probability differences visible, not just their direction.
</p>
')
```
