---
title: "Univariate Description: Interval Variables"
tutorial:
  id: "06-Interval-Description"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Mean, standard deviation, percentiles, histograms, and box plots.
---

## Overview {#main-content}

Most of the quantities political argument turns on are measured as numbers whose differences mean something: a candidate's share of the vote, a county's wage growth, a country's level of inequality, a state's unemployment rate. Because the distance between two values carries meaning, you can add such values, average them, and measure how far apart they lie. That is what makes an average possible in the first place, and it is also what makes an average easy to misread, since a single extreme value can pull it away from anything typical.

This tutorial builds the fuller account that arithmetic makes available. You will locate the center of a distribution with the median and the mean and learn to read the gap between them, measure spread with the range, percentiles, the interquartile range, and the standard deviation, and see the shape directly in a histogram and a box plot. The questions are the ones you have been asking all along --- what is typical, how spread out are the values, what pattern do they form --- but each new number carries its own question about what it actually describes and which cases it covers.

This tutorial completes the three-part sequence on **univariate description**, and it covers **interval variables**, whose values are ordered and evenly spaced --- which is what makes the mean, the standard deviation, and the histogram available. If you need to review what changes when categories can be ranked but their spacing is not meaningful, Tutorial 5 covers the ordinal case.

In this tutorial you will learn:

* How to determine and describe the central tendency of an interval level variable
* How to use the `mean()` function
* How to summarize the dispersion of an interval variable using `range()`, `quantile()`, and `sd()`
* How to describe the shape of an interval variable's distribution
* How to visualize an interval variable's distribution using `ggplot()` from the [ggplot2]{.package-name} package with `geom_histogram()` and `geom_boxplot()`, controlling the level of detail with `bins` and `binwidth`

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)
data("counties", package = "QPATutorialsCourse")
data("world", package = "QPATutorialsCourse")
```

## Univariate Description

The same three dimensions organize univariate description here: central tendency, dispersion, and the shape of the distribution. What changes is the toolkit. Because the values of an [**interval variable**]{.important-text} are ordered *and* separated by meaningful distances, you can do arithmetic on them --- add them, average them, measure how far apart they are --- and every new statistic in this tutorial comes from that one property.

**Practice 1: Identify interval-level variables**

```{r concepts-quiz-interval-level, echo=FALSE}
question("Which of the following variables are interval level measures? Select all that apply.",
  correct = "Right. All three are interval-level variables.",
  answer("The number of terrorist acts in the Middle East in a given year", correct = TRUE,
         message = "A count of terrorist acts is interval-level. The values are ordered and the differences between them are equal --- five more acts is five more acts anywhere on the scale. Neither ideology scales nor ordinal rankings have this property."),
  answer("The poverty rate in a country", correct = TRUE,
         message = "A rate measured as a percentage is interval-level. The values are ordered and a one-point difference means the same thing anywhere on the scale."),
  answer("The number of COVID-19 cases daily worldwide", correct = TRUE,
         message = "A daily count is interval-level. The values are ordered and equal differences carry equal meaning --- 1,000 more cases is 1,000 more cases regardless of the baseline."),
  answer("Individual ideology measured on a scale of 1 to 7 from very liberal to very conservative",
         message = "Ideology on a 1--7 scale is ordinal. The values are rank-ordered from liberal to conservative, but the differences between adjacent points are not necessarily equal --- a move from 3 to 4 may not represent the same ideological shift as a move from 6 to 7."),
  answer("The Polity democracy score measured on a scale from -10 (autocracy) to +10 (democracy)",
         message = "The Polity score is technically ordinal because the differences between values are not guaranteed to be equal. However, as discussed in Tutorial 3, analysts frequently treat it as interval-level in practice --- a convention worth knowing."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Three of these are interval. For each, ask whether the distance between two adjacent values means the same thing everywhere on the scale."
)
```

## Central Tendency

For interval variables, you can describe [**central tendency**]{.important-text} using the mode, median, *and* mean. That does not mean all three are equally useful. As you will see, the mode is usually not very informative for interval data, and the mean can be strongly affected by outliers, or extreme values. The median and mean will be the primary tools.

The median is the value in the middle when you list all values in rank order for all cases. Thus it is the value at the 50th percentile: half the cases have values at or below the median, and half have values at or above the median.


You are likely very familiar with the mean, or average, value. The arithmetic [**mean**]{.important-text} is computed by adding up the values for all cases and dividing by the number of cases:

$$\bar{Y} = \frac{Y_1 + Y_2  + ... + Y_n}{n} = \frac{1}{n} \Sigma_{i=1}^n Y_i$$

Can you tell why the mean is not an appropriate measure of central tendency for nominal and ordinal variables? You cannot add up categories like married, single, divorced, and widowed. Even if they are coded numerically, what would the mean of marital status *mean*? It is also an inappropriate measure of central tendency for an ordinal variable. Although you may be able to add up the values of an ordinal variable, because the differences between them are not equal, the mean is misleading.
 
 
The mean is the most commonly reported measure of central tendency, but it has one important weakness: it is sensitive to outliers. Suppose you wanted to describe the income of patrons at a bar in State College. On most nights, the median and mean income would probably be fairly close. But what happens if Elon Musk walks in? The median income barely moves, because one person cannot change the middle case very much. The mean, however, would increase dramatically. For this reason, the median is said to be **robust** to outliers, but the mean is not.

### The Mode

For nominal and ordinal variables, the mode is identified from a frequency table. Try the same approach with the interval variable **wage_growth** in the **counties** data frame, which records the percentage change in a county's average weekly wage from the third quarter of 2015 to the third quarter of 2016. It has 2 missing values, so any table or summary you compute from it describes the 3,110 counties with a recorded value.

**Run and observe 2: Find the mode of wage_growth**

```{r Freq_Prop_table_for_wage_growth, exercise=TRUE, exercise.cap="Run and observe 2"}
library(dplyr)
counties %>%
  filter(!is.na(wage_growth)) %>%
  group_by(wage_growth) %>%
  summarise(freq = n()) %>%
  mutate(prop = freq / sum(freq))
```

Can you see the problem? The values of **wage_growth** are so precisely measured that almost no two counties have the exact same wage growth rate, so no value occurs often enough to give a meaningful mode. The mode tells you nothing useful here. This is the typical situation with interval data. **For this reason, the mode is almost never reported for an interval variable.** Instead, focus on the median and mean.

Once a package is loaded with `library()`, all of its functions are available for the rest of your R session --- you do not need to load it again each time you use a function from it. Since [dplyr]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial.

### The Median

The frequency table showed why the mode is not helpful for wage growth. A better way to locate the center of the county distribution is the median, which finds the wage-growth value at the middle of the ordered distribution. As noted, **wage_growth** has 2 missing values, so any summary you compute from it describes the 3,110 counties with a recorded value rather than all 3,112 in the data. The difference is small here, but the habit of knowing how many cases you set aside is worth keeping.

**Practice 3: Compute the median of wage_growth**

Find the value at the 50th percentile of **wage_growth** in the **counties** data frame. Two hints are available. Click "Submit Answer" to check your work.

```{r compute_median_wage_growth, exercise=TRUE, exercise.cap="Practice 3"}

```

```{r compute_median_wage_growth-hint-1}
# Hint 1 of 2
# To describe the typical county's wage growth, you need
# the value that divides the distribution in half ---
# the value at the 50th percentile, with half the counties
# above it and half below. What function computes
# this for a numeric variable?
```

```{r compute_median_wage_growth-hint-2}
# Hint 2 of 2
# The function is median(). Pass it the variable using
# the $ operator --- counties$wage_growth --- and include
# na.rm = TRUE as a second argument so missing values are
# excluded rather than returning NA.
```

```{r compute_median_wage_growth-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("median\\s*\\(", code_no_comments)) {
    fail("Use the `median()` function.")
  }
  if (!grepl("wage_growth", code_no_comments)) {
    fail("Apply `median()` to the **wage_growth** variable.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number: the median of **wage_growth**.")
  }
  if (is.na(.result)) {
    fail("Your result is NA. Make sure missing values are removed before computing the median.")
  }
  if (!isTRUE(all.equal(as.numeric(.result),
                        as.numeric(median(counties$wage_growth, na.rm = TRUE))))) {
    fail("Check that you are computing the median of **wage_growth** in the **counties** data frame.")
  }
  pass(paste0("Right! The median of **wage_growth** is ",
              round(median(counties$wage_growth, na.rm = TRUE), 2),
              "%."))
})
```

The median value of wage growth is 4.70%, which means that after ranking all counties' wage growth rates, the value at the 50th percentile was just under 5% growth. Half the counties experienced wage growth at or below 4.7%.

Where does this variable come from? **wage_growth** is built from the Bureau of Labor Statistics Quarterly Census of Employment and Wages: it is the change in a county's average weekly wage from the third quarter of 2015 to the third quarter of 2016. Nationally that figure rose 5.4% over those twelve months, above the county median of 4.70%. The gap is not a contradiction. The national number pools all covered employment, so a worker in Los Angeles County counts the same as a worker in Loving County, Texas. The median gives every county one vote regardless of size. When the two disagree, the larger places are moving differently from the typical place. Call this the **weighting effect**: a national figure weights by population, a county median weights every county the same, and the two answer different questions. It comes up twice more in this tutorial.

### The Mean

We can calculate the mean using the `mean()` function. Now calculate it so you can compare it to the median. When they are far apart, that difference is often a clue that outliers or a lopsided distribution are pulling the mean away from the middle.

**Practice 4: Calculate the mean of wage_growth**

Compute the arithmetic average of **wage_growth**. Two hints are available. Click "Submit Answer" to check your work.

```{r compute_mean_wage_growth, exercise=TRUE, exercise.cap="Practice 4"}

```

```{r compute_mean_wage_growth-hint-1}
# Hint 1 of 2
# You need the arithmetic average of wage growth ---
# the sum of all county values divided by the count
# of counties. This is one of the two main measures
# of central tendency for interval variables.
# What function in R computes this?
```

```{r compute_mean_wage_growth-hint-2}
# Hint 2 of 2
# The function is mean(). Pass it the variable using
# the $ operator --- counties$wage_growth --- and include
# na.rm = TRUE as a second argument.
```

```{r compute_mean_wage_growth-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("mean\\s*\\(", code_no_comments)) {
    fail("Use the `mean()` function.")
  }
  if (!grepl("wage_growth", code_no_comments)) {
    fail("Apply `mean()` to the **wage_growth** variable.")
  }
  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number: the mean of **wage_growth**.")
  }
  if (is.na(.result)) {
    fail("Your result is NA. Make sure missing values are removed before computing the mean.")
  }
  if (!isTRUE(all.equal(as.numeric(.result),
                        as.numeric(mean(counties$wage_growth, na.rm = TRUE))))) {
    fail("Check that you are computing the mean of **wage_growth** in the **counties** data frame.")
  }
  pass(paste0("Good work! The mean of **wage_growth** is ",
              round(mean(counties$wage_growth, na.rm = TRUE), 2),
              "%."))
})
```

When the mean and median are close, the distribution may be fairly balanced around the center --- though you should always check a histogram to confirm.

Now try the same comparison with a more politically familiar variable. **dem2p_percent** is the percent of the two-party presidential vote cast for the Democrat in a county in 2016, and unlike **wage_growth** it has 0 missing values, so both statistics describe all 3,112 counties. The gap between its median and mean will give you an early clue about the shape of the county-level distribution of Democratic vote percentages.

**Practice 5: Compute the median and mean of dem2p_percent**

Compute both statistics for **dem2p_percent** in the **counties** data frame on separate lines, then compare them. Two hints are available. Click "Submit Answer" to check your work.

```{r compute_mean_median_dem2p_percent, exercise=TRUE, exercise.lines=4, exercise.cap="Practice 5"}

```

```{r compute_mean_median_dem2p_percent-hint-1}
# Hint 1 of 2
# You are computing two different summaries of the same
# variable: the value at the midpoint of the distribution,
# and the arithmetic average. Comparing them is the point
# --- a gap between them can signal that a few
# extreme values are pulling the mean around.
```

```{r compute_mean_median_dem2p_percent-hint-2}
# Hint 2 of 2
# Use median() and mean(), each applied to
# counties$dem2p_percent. Put each on its own line.
```

```{r compute_mean_median_dem2p_percent-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (!grepl("median\\s*\\(", code_no_comments)) {
    fail("Include a call to `median()` for **dem2p_percent**.")
  }
  if (!grepl("mean\\s*\\(", code_no_comments)) {
    fail("Include a call to `mean()` for **dem2p_percent**.")
  }
  if (!grepl("median\\s*\\([^)]*dem2p_percent", code_no_comments)) {
    fail("Apply `median()` to **dem2p_percent**.")
  }
  if (!grepl("mean\\s*\\([^)]*dem2p_percent", code_no_comments)) {
    fail("Apply `mean()` to **dem2p_percent**.")
  }

  expected_median <- median(counties$dem2p_percent, na.rm = TRUE)
  expected_mean <- mean(counties$dem2p_percent, na.rm = TRUE)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  result_val <- as.numeric(.result)
  if (!is.numeric(.result) || length(result_val) != 1 || is.na(result_val)) {
    fail("Your final line should return either the median or the mean of **dem2p_percent**.")
  }
  if (!isTRUE(all.equal(result_val, expected_mean)) &&
      !isTRUE(all.equal(result_val, expected_median))) {
    fail("Put `median()` on one line and `mean()` on the other. Make sure both are computed for **dem2p_percent**.")
  }

  pass("Correct! You computed both the median and the mean of **dem2p_percent**.")})
```

The median percent of the two-party vote cast for Clinton was just under 30% --- half the counties cast 30% or fewer of their two-party votes for her. The mean of about 33% is higher than the median, suggesting that some counties with very high Democratic vote percentages are pulling the mean upward. The median is much less affected by those extreme values because it depends on the ordering of observations rather than how far an extreme value lies from the rest --- but the mean moves whenever a value changes. When the mean exceeds the median, it usually means a long tail of high values is pulling the mean up while the median stays closer to where most observations are.

The weighting effect you met with wage growth is at its most dramatic here. Clinton received 51.1% of the national two-party vote, but the median county gave her just under 30%. Both numbers are correct, and the distance between them --- more than twenty percentage points --- is the whole point: Democratic votes were concentrated in a relatively small number of very populous counties, so a summary that gives every county one vote lands far below a summary that gives every voter one vote. What the size of that gap does tell you is that Democratic vote share and county population were related: the populous counties contributed disproportionately to the national total. Separately from the weighting, the distance between the county mean and the county median suggests the county distribution has a longer upper tail --- something you will see directly when you plot this variable later in the tutorial. One thing these county figures cannot tell you is who won: a median across counties is not a vote total, and nothing in this distribution reports the national result.

## Dispersion

Once you know where the center of a distribution is, the next question is how spread out the values are. For interval variables, you have several ways to describe [**dispersion**]{.important-text}. You will consider:

1. **Range**
2. **Percentiles**, especially the **interquartile range**
3. **Variance** and **standard deviation**


### Range

The [**range**]{.important-text} of values a variable takes is given by the smallest (minimum) value and the largest (maximum) value. As such, it provides bookends on the spread of the values of the variable across cases. It is useful to report the range of the data in the sample because the sample range may be different from the theoretical range of the data. It can also help identify the presence of outliers.

The R function `range()` directly calculates both the minimum and maximum values in the data. The function `min()` calculates the minimum value of a variable, and the `max()` function calculates the maximum value. Each function is passed the name of the variable being analyzed. If there are missing values, include the `na.rm` argument and set it equal to `TRUE` so that R ignores those values when computing the statistic.

Let's calculate the range of **wage_growth** to see just how much variation there was across counties. Since wages can fall, wage growth need not be positive at all.

**Practice 6: Compute the range, min, and max of wage_growth**

See how far the distribution of **wage_growth** in the **counties** data frame stretches from its most extreme values. Put each function call on its own line. Two hints are available. Click "Submit Answer" to check your work.

```{r compute_range_min_max_wage_growth, exercise=TRUE, exercise.cap="Practice 6"}

```

```{r compute_range_min_max_wage_growth-hint-1}
# Hint 1 of 2
# You need three separate function calls on the same
# variable: one that returns both endpoints together,
# and two that return each endpoint separately.
# Think about what three functions from Tutorials 1
# and 2 give you the minimum, the maximum, and
# both endpoints at once.
```

```{r compute_range_min_max_wage_growth-hint-2}
# Hint 2 of 2
# The three functions are range(), min(), and max().
# Each takes counties$wage_growth as its first argument
# and na.rm = TRUE as its second. Put each call on its
# own line.
```

```{r compute_range_min_max_wage_growth-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (!grepl("range\\s*\\(", code_no_comments)) {
    fail("Include a call to `range()` for **wage_growth**.")
  }
  if (!grepl("min\\s*\\(", code_no_comments)) {
    fail("Include a call to `min()` for **wage_growth**.")
  }
  if (!grepl("max\\s*\\(", code_no_comments)) {
    fail("Include a call to `max()` for **wage_growth**.")
  }
  if (!grepl("range\\s*\\([^)]*wage_growth", code_no_comments)) {
    fail("Your `range()` call should use **wage_growth** and handle its missing values.")
  }
  if (!grepl("min\\s*\\([^)]*wage_growth", code_no_comments)) {
    fail("Your `min()` call should use **wage_growth** and handle its missing values.")
  }
  if (!grepl("max\\s*\\([^)]*wage_growth", code_no_comments)) {
    fail("Your `max()` call should use **wage_growth** and handle its missing values.")
  }

  if (!grepl("range\\s*\\([^)]*na\\.rm\\s*=\\s*TRUE", code_no_comments)) {
    fail("Your `range()` call returns NA. **wage_growth** has missing values, and they have to be excluded before the function can return a number.")
  }
  if (!grepl("min\\s*\\([^)]*na\\.rm\\s*=\\s*TRUE", code_no_comments)) {
    fail("Your `min()` call returns NA. **wage_growth** has missing values, and they have to be excluded before the function can return a number.")
  }
  if (!grepl("max\\s*\\([^)]*na\\.rm\\s*=\\s*TRUE", code_no_comments)) {
    fail("Your `max()` call returns NA. **wage_growth** has missing values, and they have to be excluded before the function can return a number.")
  }

  expected_range <- range(counties$wage_growth, na.rm = TRUE)
  expected_min <- min(counties$wage_growth, na.rm = TRUE)
  expected_max <- max(counties$wage_growth, na.rm = TRUE)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  result_val <- as.numeric(.result)

  if (any(is.na(result_val)) ||
      !(isTRUE(all.equal(result_val, expected_range)) ||
        isTRUE(all.equal(result_val, expected_min)) ||
        isTRUE(all.equal(result_val, expected_max)))) {
    fail("Put `range()`, `min()`, and `max()` on separate lines, all applied to **wage_growth** with missing values removed.")
  }

  pass("Exactly! All three calls are correct --- `range()` returned both endpoints together, and `min()` and `max()` returned each one separately.")
})
```

Before deciding those extremes are errors, think about what the measure counts. A county's average weekly wage is an average over everyone employed there, so it moves when the mix of jobs changes, not only when anyone's pay changes. In a county with a few thousand workers, a single plant closing removes hundreds of similar wages at once, and a new hospital or distribution center adds a block of higher ones. Either event can shift the county average by tens of percent in a year while no individual worker's wage moved at all. The extremes are real --- they are reporting employment composition in small counties rather than raises.

Those extreme values are worth understanding. You might expect outlier counties to be visible in a histogram, but you probably will not see them. With over 3,000 counties in the data, a handful of outlier counties contribute bars so short --- perhaps one or two counties per bin --- that they are nearly invisible against the bars near the center, which each represent hundreds of counties. The x-axis stretches to include the extremes, but the bars themselves disappear into the baseline. This is one reason why computing the range is good practice even when you have a histogram: the two tools can tell different stories about the same data.

### Percentiles

A [**percentile**]{.important-text} is the value of a variable that has a certain percentage of observations below it. In other words, the nth percentile is the value that cuts off the first n percent of the data when all values are sorted from least to greatest. You have already worked with the 50th percentile: it is the median. The 90th percentile is the value that separates the bottom 90% of values from the top 10%.

Percentiles can be used to answer questions such as:

::: {.research-example}
Among counties in the bottom quarter of the life-expectancy distribution, what is the highest life expectancy?
:::

To answer this question, we'd find the 25th percentile of life expectancy across counties. This is the life expectancy value that separates the bottom 25% of observations from the top 75%.

::: {.research-example}
What range of jobs per 100 working-age residents covers the middle half of all counties?
:::

To answer this question, find the 75th percentile and the 25th percentile of the number of jobs located in the county per 100 residents aged 25 to 64. These are the two values that determine the upper and lower bounds for the middle 50% of the data.


The distance between the 25th and 75th percentiles is known as the [**interquartile range**]{.important-text} or [**IQR**]{.important-text}. The quartiles divide the ordered observations into four groups, each containing approximately 25% of the cases. The IQR spans the middle 50% of the observations. Its advantage over the full range as a measure of spread is that extreme values do not affect it.


You can calculate percentiles with the `quantile()` function. By default, passing the variable name to the function will return the 0th, 25th, 50th,  75th, and 100th percentiles.

**Run and observe 7: Calculate the percentiles for wage_growth**
```{r calculate_percentile_wage_growth, exercise=TRUE,exercise.cap="Run and observe 7"}
quantile(counties$wage_growth, na.rm = TRUE)
```

The 0% and 100% entries are the minimum and maximum, so this one call gives you the range as well. Focus on the interquartile range, the values at the 25th and 75th percentile, which demarcate the middle half of the data. The IQR tells you that the growth rate of wages for half the counties was between 2.5% and 6.5% --- a span of about four percentage points. The full range runs from about -69% to 63%. The extremes sit enormously far outside the band where most counties fall, which is the first hard evidence that this distribution has a long reach in both directions.

You can use the `quantile()` function to list additional percentiles by adding the `probs` argument inside the function. To list a specific set of percentiles set `probs = c(0.05, .1, .25, .5, .75, .9, .95)`.

**Run and observe 8: Calculate the additional percentiles for wage_growth**
```{r calculate_additional_percentile_wage_growth, exercise=TRUE,exercise.cap="Run and observe 8"}
quantile(counties$wage_growth, na.rm = TRUE, 
         probs = c(0.05, .1, .25, .5, .75, .9, .95))
```

The result shows that 90% of the cases fall between -2.60% and 9.92%. In other words, the bottom 5% of counties had wage growth of -2.60% or lower, meaning wages declined by at least about 2.6% in those counties, and the top 5% of counties experienced wage increases approaching 10% or higher.


Now use percentiles to describe a different kind of county economy measure: **employ_pop_ratio_25_64**, the number of jobs located in a county for every 100 residents aged 25 to 64. It has 2 missing values. The default quantiles will show the minimum, the maximum, the median, and the middle half of counties, giving you a compact picture of both center and spread.

**Practice 9: Calculate the percentiles for employ_pop_ratio_25_64**

One call to `quantile()` gives you all five values for **employ_pop_ratio_25_64** in the **counties** data frame. Two hints are available. Click "Submit Answer" to check your work.

```{r calculate_percentile_employ_pop_ratio_25_64, exercise=TRUE, exercise.cap="Practice 9"}

```

```{r calculate_percentile_employ_pop_ratio_25_64-hint-1}
# Hint 1 of 2
# You want to see the minimum, the 25th percentile,
# the median, the 75th percentile, and the maximum ---
# the five values that show where the distribution
# falls at even intervals. The function you need
# returns all five by default in a single call.
# What function produces this output?
```

```{r calculate_percentile_employ_pop_ratio_25_64-hint-2}
# Hint 2 of 2
# The function is quantile(). Pass it
# counties$employ_pop_ratio_25_64 with na.rm = TRUE.
# No probs argument is needed --- the default already
# returns the five percentiles you want.
```

```{r calculate_percentile_employ_pop_ratio_25_64-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("quantile\\s*\\(", code_no_comments)) {
    fail("Use the `quantile()` function.")
  }
  if (!grepl("employ_pop_ratio_25_64", code_no_comments)) {
    fail("Apply `quantile()` to **employ_pop_ratio_25_64**.")
  }

  expected <- quantile(counties$employ_pop_ratio_25_64, na.rm = TRUE)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != length(expected)) {
    fail("Your result should show the default quantiles: minimum, 25th percentile, median, 75th percentile, and maximum.")
  }
  if (any(is.na(.result))) {
    fail("Your result includes NA. Make sure missing values are removed before computing quantiles.")
  }
  if (!isTRUE(all.equal(as.numeric(.result), as.numeric(expected), tolerance = 1e-8))) {
    fail("Check that you are computing the default quantiles for **employ_pop_ratio_25_64**.")
  }

  pass("Good work! Compare that with the three separate calls it took to get the endpoints of **wage_growth** earlier --- `quantile()` returns the endpoints, the quartiles, and the median together.")
})
```

What is the range of the variable? 9.21 to 668.07. A value above 100 is not an error here, and 331 counties --- better than one in nine --- have one. The numerator counts jobs located in the county, while the denominator counts residents of the county, and those are not the same people: a county can hold far more jobs than working-age residents when workers commute in. Butte County, Idaho tops the list at 668 jobs per 100 residents, because the Idaho National Laboratory sits in a county of about 2,500 people. New York County --- Manhattan --- reaches 239 for the same reason on a much larger scale. A second, smaller reason is that the numerator counts jobs rather than workers, so someone holding two covered jobs is counted twice.

The IQR runs from 51.23 to 83.53. The middle half of counties therefore have somewhere between about 51 and 84 jobs located in them for every 100 residents aged 25 to 64.

You can also extract the median from the output. The middle value in the data is 66.81. Half of counties have 66.81 jobs or fewer for every 100 residents aged 25 to 64.

What does a median of 66.81 actually say? It says that the middle county had about 67 jobs located within it for every 100 residents aged 25 to 64. It does *not* say that a third of those residents were out of work. As the range discussion above showed, the people counted in the numerator need not live in the county: some residents commute out to jobs elsewhere, and some of the jobs counted here are filled by people commuting in.

The weighting effect is at work here for the third time. Nationally in 2016 there were about 84 covered jobs for every 100 residents aged 25 to 64, computed by dividing all covered employment by the whole population in that age range; the county median is 66.81. The two answer different questions --- the national ratio weights people equally, the county median weights counties equally --- so the gap is a reminder that the unit of analysis shapes the answer. Explaining the size of that gap would mean examining how county ratios vary with county population, which is a bivariate question and beyond what a single median can tell you.

### Standard Deviation and Variance

The [**standard deviation**]{.important-text} describes the typical distance of values from the mean of the variable. The [**variance**]{.important-text} is the squared standard deviation.
 
The formula looks a little intimidating at first, but the idea is simple: measure how far each case is from the mean, average those distances in a way that keeps them from canceling out, and then put the result back into the original units.

The formula for the standard deviation, denoted by $s$, is
 
$$s=\sqrt{\frac{\Sigma_{i=1}^n({Y_i-\bar{Y}})^2}{n-1}}$$
 
Let's pick this apart by considering an example where the variable of interest is the height of presidents from Johnson through Biden. Below is a table listing the heights (in centimeters) of 10 selected presidents. Note that Bush II and Biden are the same height (182 cm) --- a useful reminder that tied values are perfectly possible and handled naturally by the formula. The table also presents the deviations in column three and deviations squared in column four. The last row sums the values in each column, giving you the building blocks to calculate the standard deviation.
 

<table id="pres-stats-table" aria-labelledby="pres-stats-caption"> <caption id="pres-stats-caption">Presidential heights and deviations from the mean height, used to illustrate variance and standard deviation.</caption> <thead> <tr> <th scope="col">President</th> <th scope="col">\(Y_i\)</th> <th scope="col">Deviation \((Y_i - \bar{Y})\)</th> <th scope="col">Deviation\(^2\) \((Y_i - \bar{Y})^2\)</th> </tr> </thead> <tbody> <tr><th scope="row">Trump</th><td>188</td><td>188 − 184.9 = 3.1</td><td>9.61</td></tr> <tr><th scope="row">Biden</th><td>182</td><td>182 − 184.9 = −2.9</td><td>8.41</td></tr> <tr><th scope="row">Obama</th><td>185</td><td>185 − 184.9 = 0.1</td><td>0.01</td></tr> <tr><th scope="row">Bush II</th><td>182</td><td>182 − 184.9 = −2.9</td><td>8.41</td></tr> <tr><th scope="row">Clinton</th><td>188</td><td>188 − 184.9 = 3.1</td><td>9.61</td></tr> <tr><th scope="row">Bush I</th><td>188</td><td>188 − 184.9 = 3.1</td><td>9.61</td></tr> <tr><th scope="row">Reagan</th><td>185</td><td>185 − 184.9 = 0.1</td><td>0.01</td></tr> <tr><th scope="row">Carter</th><td>177</td><td>177 − 184.9 = −7.9</td><td>62.41</td></tr> <tr><th scope="row">Ford</th><td>183</td><td>183 − 184.9 = −1.9</td><td>3.61</td></tr>  <tr><th scope="row">Johnson</th><td>191</td><td>191 − 184.9 = 6.1</td><td>37.21</td></tr> </tbody> <tfoot> <tr> <th scope="row">Summary</th> <td>\(\bar{Y}=\frac{1}{n}\sum_{i=1}^n Y_i=184.9\)</td> <td>\(\sum_{i=1}^n (Y_i-\bar{Y})=0\)</td> <td>\(\sum_{i=1}^n (Y_i-\bar{Y})^2=148.9\)</td> </tr> </tfoot> </table>

<br>

Steps to calculate the standard deviation:

1. Calculate the deviation --- how far above or below the mean each observation is: $(Y_i - \bar{Y})$ --- for each observation (column three). Observations below the mean have negative deviations, and observations above the mean have positive deviations. If the variable is not very dispersed, the deviations will be close to zero. Bigger deviations indicate more dispersion.
2. By definition, the average deviation will always be zero: deviations above the mean offset deviations below the mean. To avoid this problem, we square the deviations (column four).
3. Now we can calculate the variance by summing the squared deviations and dividing by $n-1$ rather than $n$. We divide by $n-1$ rather than $n$ because we are estimating the population standard deviation from a sample; dividing by $n-1$ corrects for a small but systematic bias that would otherwise make the estimate too small. Here, $n=10$, so we divide by 9: 148.9/9 = 16.54.
4. If we stop here, we have the variance. Taking the square root gets us back to the original units of the variable. This is called the standard deviation:

$$s=\sqrt{\frac{\Sigma_{i=1}^n({Y_i-\bar{Y}})^2}{n-1}}= \sqrt{\frac{148.9}{9}}=4.07$$

Note that $s$ is always positive.

A larger standard deviation means values are more spread out from the mean; a smaller one means they cluster more closely around it. One important caution: **Raw standard deviations generally should not be compared across variables measured in different units or on different scales. They are most directly comparable across samples or groups measured using the same variable.** A standard deviation of 4.68 for **wage_growth** (measured in percentage points) and a standard deviation of 15.3 for **dem2p_percent** (also percentage points but on a completely different scale with a different mean) do not tell you which variable is "more dispersed" in any general sense. Standard deviations are only comparable within a variable or across samples of the same variable.

To calculate the standard deviation, $s$, of a variable we use the `sd()` function.

**Run and observe 10: Calculate the standard deviation of wage_growth**

```{r calculate_sd, exercise=TRUE, exercise.cap="Run and observe 10"}
sd(counties$wage_growth, na.rm = TRUE)
```

A standard deviation of 4.68 percentage points tells you that county wage-growth rates typically vary by about 4.68 percentage points around the mean of 4.33%. The standard deviation becomes especially important in statistical inference, where the Empirical Rule connects it to the percentage of observations falling within specific ranges.


## The Shape of the Distribution

Assessing the [**shape of the distribution**]{.important-text} of an interval variable is most easily done after visualizing the distribution in a plot. The key questions are:

- Is the distribution [**symmetric**]{.important-text} --- does it look roughly the same on both sides of the center --- or [**skewed**]{.important-text} toward one tail?
- Is it [**unimodal**]{.important-text} (one peak), [**bimodal**]{.important-text} (two peaks), or something else?
- Are there [**outliers**]{.important-text} --- observations far from the rest of the distribution that may warrant investigation?

Not all variables will have symmetric distributions. Instead, they may be skewed. Histograms are said to be skewed to the side of the long tail. The two histograms below are clearly not symmetric. The image on the left is skewed to the right. The image on the right is skewed to the left.

```{r skew-illustration, echo=FALSE, fig.height=3, fig.width=7, fig.alt="Two histograms side by side. The left panel, labeled positively skewed, has most observations bunched at low values with a long tail extending to the right. The right panel, labeled negatively skewed, is the mirror image, with most observations at high values and a long tail extending to the left."}
set.seed(1234)
skew_demo <- data.frame(
  value = c(rgamma(300, shape = 2, rate = 1), 10 - rgamma(300, shape = 2, rate = 1)),
  shape = factor(rep(c("Positively skewed (long tail on the right)",
                       "Negatively skewed (long tail on the left)"), each = 300),
                 levels = c("Positively skewed (long tail on the right)",
                            "Negatively skewed (long tail on the left)"))
)

ggplot(skew_demo, aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~ shape, scales = "free_x") +
  labs(x = "Value", y = "Number of observations") +
  theme_minimal()
```

## Visualizing an Interval Variable: Histograms

The histogram and the box plot are the primary tools for assessing shape, and the histogram comes first. For the same reason a frequency table is not very helpful for interval variables, a bar plot is not the right visualization either. Instead, we use a [**histogram**]{.important-text}. A histogram groups observed values into **bins** that cover **ranges** of values, then uses the height of each bar to show how many observations fall in that range. Histograms can also be constructed to show proportions rather than counts. We can control the number of bins, control their width, or let R choose a default. A histogram shows the central tendency of the variable and its dispersion, as well as the shape of the distribution.

Generating a histogram with the `ggplot()` function requires a simple change to the basic code for a bar plot. Because we want a visual display for interval data, we replace `geom_bar()` with `geom_histogram()`. The basic structure is familiar, but the interpretation changes because each bar now represents a range of values rather than one category. Note that we do not wrap **wage_growth** in `factor()`. That function is needed only when a variable stored as numbers should be treated as categorical. Interval variables are genuinely numeric, so no wrapping is needed.

As with proportion bar charts, filter out missing values before passing data to `ggplot()`. Without the filter ggplot still draws the plot, but it reports how many rows it dropped, and that report follows the figure into any document you knit. Always note in your text how many cases were excluded.

Inside `geom_histogram()`, two arguments control the appearance of the bars. `fill` sets the interior color of the bars, just as it did in the bar charts in Tutorials 4 and 5. `color` sets the border color --- adding white borders between bars makes individual bins visually separable, which is especially helpful when bins are narrow.

**Run and observe 11: Generate the histogram for wage_growth**

Like [dplyr]{.package-name}, [ggplot2]{.package-name} only needs to be loaded once per session --- all subsequent exercises in this tutorial can use it without calling `library(ggplot2)` again.

```{r generate_histogram_wage_growth, exercise=TRUE, warning=TRUE, message=TRUE, exercise.cap="Run and observe 11", fig.alt="Histogram of wage growth across US counties. The horizontal axis runs from about -60 to 60 percent. Almost all counties fall between roughly -10 and 20 percent: one tall bar just above zero reaches about 1,500 counties and the bar beside it about 950. Anything outside that narrow range is too short to see at this scale."}
library(ggplot2)
counties %>%
  filter(!is.na(wage_growth)) %>%
  ggplot(mapping = aes(x = wage_growth)) +
         geom_histogram(fill = "steelblue", color = "white") +
         labs(title = "Wage Growth in US Counties in 2016", 
              x = NULL, 
              y = "Frequency", 
              caption = "Source: Linn, Nagler, Zilinsky")
```



Almost every county falls between about -10 and 20 percent, and the tallest bar sits just above zero. A handful of counties lie far outside that range --- far enough to stretch the axis from roughly -60 to 60 --- but they are too few to be visible at this scale, which is why the bulk of the distribution is crammed into a narrow strip of the plot.

R also prints a note above the plot saying that `stat_bin()` used 30 bins and suggesting you pick a better value. That is not an error and you have done nothing wrong. `geom_histogram()` has to group the values into bars somehow, and 30 is its default guess; the note is R telling you that the choice is really yours to make.

Before changing anything, it is worth knowing what the choice costs in each direction. With too few bins, each bar covers a wide range of values and real features of the distribution disappear inside them. With too many, the plot becomes noisy and the overall pattern is hard to see through the clutter. What you are looking for is the smallest number of bins that still shows the features you care about, and the only reliable way to find it is to try several and compare.

The right answer depends entirely on the variable. Thirty bins is too coarse for **wage_growth** because nearly every county sits inside a band a few percentage points wide, so the default spreads those bins across a range dominated by a handful of distant counties. A variable spread more evenly across its range might look fine at 30, or even at 10.

There are two options for specifying a different number of bins. You can set the number of bins by adding the `bins` argument to `geom_histogram()` and setting it equal to some number.


**Practice 12: Generate the histogram for wage_growth with a different number of bins**

Change the number of bins in the plot to 50. Two hints are available. Click "Submit Answer" to check your work.

```{r generate_histogram_wage_growth_custom_bin, exercise=TRUE, exercise.cap="Practice 12", fig.alt="The same wage growth histogram drawn with 50 bins. The bars are narrower, so the single tall peak now splits into two adjacent bars of roughly 900 and 1,050 counties, and several shorter bars on either side become visible that the default binning had merged together."}
counties %>%
  filter(!is.na(wage_growth)) %>%
  ggplot(mapping = aes(x = wage_growth)) +
         geom_histogram(fill = "steelblue", color = "white", ___ = ___) +
         labs(title = "Wage Growth in US Counties in 2016", 
              x = NULL, 
              y = "Frequency", 
              caption = "Source: Linn, Nagler, Zilinsky")
```


```{r generate_histogram_wage_growth_custom_bin-hint-1}
# Hint 1 of 2
# A histogram divides the range of the variable into
# intervals and counts how many cases fall in each. What
# you need to change is the number of intervals. That
# setting goes inside the same function that draws the
# bars, alongside the fill and color you already have.
```

```{r generate_histogram_wage_growth_custom_bin-hint-2}
# Hint 2 of 2
# The argument is bins, and the prompt gives the value.
# It goes inside geom_histogram() after color = "white",
# separated by a comma.
```



```{r generate_histogram_wage_growth_custom_bin-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  expr_text <- function(x) {
    if (is.null(x)) return("")
    paste(deparse(rlang::get_expr(x)), collapse = "")
  }

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }
  if (!is.data.frame(.result$data) || !"wage_growth" %in% names(.result$data)) {
    fail("Use **counties** as the data frame in `ggplot()`.")
  }
  x_expr <- expr_text(.result$mapping$x)
  if (!grepl("wage_growth", x_expr)) {
    fail("Map **wage_growth** to the x-axis.")
  }
  hist_layers <- Filter(function(layer) inherits(layer$stat, "StatBin"), .result$layers)
  if (length(hist_layers) == 0) {
    fail("Use `geom_histogram()` to create the histogram.")
  }
  has_50_bins <- any(vapply(hist_layers, function(layer) {
    identical(layer$stat_params$bins, 50)
  }, logical(1)))
  if (!has_50_bins && !grepl("bins\\s*=\\s*50", code_no_comments)) {
    fail("Set `bins = 50` inside `geom_histogram()`.")
  }

  pass("Good work! Your histogram uses 50 bins.")
})
```


Alternatively, you can control the width of each bin directly. If a variable spans about 100 units and you set `binwidth = 10`, the plot will use roughly ten bins. The two arguments are two routes to the same kind of control: `bins` sets the approximate number of bins, and `binwidth` sets their width directly. They are not exactly interchangeable, because where the bin boundaries fall depends on the range of the data. The code below starts you at a binwidth of 3.

**Explore 13: Find the coarsest binwidth that still shows the shape**

The starter code uses `binwidth = 3`. Change that value, run the code, and look at what happens to the shape --- then change it again. Go far in both directions before narrowing in --- try values below 1 as well as values substantially above 3. You are looking for the largest binwidth that still lets you see the shape of the distribution, since anything finer than that adds clutter without adding information.

```{r generate_histogram_wage_growth_custom_bin-width, exercise=TRUE,exercise.cap="Explore 13", fig.alt="The same wage growth histogram drawn with a binwidth of 3. This is slightly coarser than the 50-bin version, so the two adjacent bars that appeared there merge back into a single peak of about 1,180 counties, with much shorter bars on either side."}
counties %>%
  filter(!is.na(wage_growth)) %>%
  ggplot(mapping = aes(x = wage_growth)) +
         geom_histogram(fill = "steelblue", color = "white", binwidth = 3) +
         labs(title = "Wage Growth in US Counties in 2016", 
              x = NULL, 
              y = "Frequency", 
              caption = "Source: Linn, Nagler, Zilinsky")
```

**Practice 14: Choose a binwidth**

```{r q-binwidth-choice, echo=FALSE}
question("Based on what you saw, which of these binwidths is closest to the coarsest one that still shows the shape of the **wage_growth** distribution?",
  correct = "Right. Around 0.8 is the coarsest binwidth that still shows the shape.",
  answer("20",
         message = "Wage growth runs from about -69 to 63, so a binwidth of 20 gives you roughly seven bars across the whole range. Nearly every county lands in one or two of them, and the shape you are trying to see disappears inside a single tall bar."),
  answer("5",
         message = "This gives about 26 bars, close to the default 30 --- and it has the same problem the default had, with the counties packed into a few bars near zero."),
  answer("0.8", correct = TRUE,
         message = "At this width the central spike separates into a recognizable shape and you can see where the bulk of counties actually falls. Going finer than this adds bars without adding anything you could not already see."),
  answer("0.05",
         message = "This gives more than two thousand bins. Individual counties become their own bars and the plot turns into noise, with no pattern visible through the clutter."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "One of these is far too wide and one far too narrow. Of the two in between, which is the wider one that still lets you see the shape?"
)
```

At the binwidth you settled on, the shape of the distribution is much clearer. Note something important, though: even at that width, **you cannot see the outlier counties** in this histogram. Their bars are so short compared to the central spike that they are invisible at this scale. You know they exist because of the range computed earlier --- this is one limitation of histograms. Rare extreme values can disappear visually even though they strongly affect statistics like the mean and range.

**Practice 15: Describe the shape of the distribution**

```{r q-histogram-shape, echo=FALSE}
question("Using the binwidth you settled on in the last question, which of the following best describes the shape of the distribution you can see in the plot?",
  correct = "Right. The main body of the distribution is roughly symmetric.",
  answer("The main body is roughly symmetric and unimodal, with most counties clustered near the center",
         correct = TRUE,
         message = "The visible main body has a single peak near 4 to 5 percent and tapers fairly evenly on both sides, which is consistent with the mean and median being close to each other."),
  answer("Asymmetric, with most counties at low growth rates but a long tail of counties with very high wage growth",
         message = "The plot shows bars on both sides of the center that taper off fairly evenly in both directions. There is not a long tail extending only to one side --- the distribution is fairly balanced around the center."),
  answer("Uniform --- all values of wage growth are roughly equally common",
         message = "A uniform distribution would show bars of similar height across the full range. Here the bars are much taller near the center and shorter in the tails."),
  answer("Bimodal --- there are two distinct clusters of counties",
         message = "A bimodal distribution would show two separate peaks. The wage growth histogram has one central peak, not two."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Look at the plot rather than at what you expect. Are the two sides of the distribution roughly mirror images, or does one tail run much further out?"
)
```

### Adding Reference Lines

Often it is useful to add a vertical line to a histogram to mark where the median or mean falls. The `geom_vline()` function adds this layer, and its key argument is `xintercept`, which sets the x-axis value where the line appears.

When that value has to be computed from the data, you have two routes, and the difference is about where R looks for the variable. Inside `aes()`, ggplot resolves bare variable names against the data you piped in, so `aes(xintercept = median(wage_growth, na.rm = TRUE))` works. Outside `aes()` you are in ordinary R, which knows nothing about the pipeline, so you have to name the data frame: `xintercept = median(counties$wage_growth, na.rm = TRUE)`. Both produce the same line. That same rule explains why `annotate()` needs `counties$wage_growth` later in this section --- it does not inherit the data context either.

Two more arguments control how the line looks, and both are fixed settings rather than mappings, so both go outside `aes()`. `color` sets the line's color; red is a good choice here because it stands out against the steelblue bars, and a reference line that blends into the data is not doing its job. `linetype` sets the line's pattern, and takes values including `"solid"`, `"dashed"`, `"dotted"`, and `"dotdash"`. A broken line is conventional for a reference line: it signals that the line is an annotation you have added rather than part of the data itself. `linewidth` sets how thick the line is, and defaults to 0.5, which is thin enough to disappear against filled bars --- a reference line usually wants a little more weight than that.

**Run and observe 16: Add the median line**

```{r try_it_label_median_line, exercise=TRUE,exercise.cap="Run and observe 16", fig.alt="Histogram of wage growth with 60 bins, overlaid with a dashed vertical line marking the median at about 4.7 percent. The line passes through the tallest bar, near the center of the main cluster of counties."}
counties %>%
  filter(!is.na(wage_growth)) %>%
  ggplot(mapping = aes(x = wage_growth)) +
         geom_histogram(fill = "steelblue", color = "white", bins = 60) +
         geom_vline(aes(xintercept = median(wage_growth, na.rm = TRUE)),
                    color = "firebrick", linetype = "dashed", linewidth = 0.8) +
         labs(title = "Wage Growth in US Counties in 2016", 
              x = NULL, 
              y = "Frequency", 
              caption = "Source: Linn, Nagler, Zilinsky")
```

The dashed line marks the median of wage growth, sitting close to the center of the distribution. When the distribution is roughly symmetric, the median sits near the peak. In a right-skewed distribution, the median often lies to the right of the peak, while the mean is pulled farther toward the long right tail. Here, the median and mean are close to each other (4.70% and 4.33%), which is consistent with the roughly symmetric main body visible in the histogram.

The line is useful but unlabeled. What should a label say? A reference line is useless to a reader who cannot tell what it marks, so a good annotation names the statistic and reports its value --- "Mean = 4.33" rather than either "4.33" or "Mean" alone.

The `annotate()` function builds that label. Its first argument names what kind of annotation to draw --- `"text"` here, though `annotate()` can also draw rectangles, segments, and other shapes. Its `x` and `y` arguments place the center of the text on the plot axes, and `label` sets the text itself. Here the text has two parts: the fixed words "Mean = " and the value, which you compute rather than type. `paste0()` joins them, and `round()` controls how many decimals the number shows.

Why not just type `"Mean = 4.33"` rather than have R compute it? Because a typed number can go stale. Filter the data differently, correct an error, or rerun this on next year's file, and the line moves while a hand-typed label keeps insisting on the old value. Computing it inside the label means the two can never disagree.

One catch: `annotate()` does not inherit the data context from `ggplot()`, so inside it you write `counties$wage_growth` rather than a bare variable name.

::: {.tip}
**Choosing annotation coordinates:** The `x` and `y` values in `annotate()` are on the same scale as the plot axes --- `x` is a value of the variable being plotted and `y` is a frequency count. To choose good coordinates: (1) run the plot first to see the axis ranges; (2) pick an `x` value near the line but offset so the text doesn't overlap it; (3) pick a `y` value where that `x` has no bars --- and keep it inside the range the plot already covers, because naming a `y` far above the data stretches the axis and flattens every bar toward the baseline. If the annotation disappears or behaves strangely, your coordinates are probably outside the plot's visible range --- try values closer to the center. In your own analyses you will need to experiment to find coordinates that work.
:::

The exercise below builds the same histogram again, this time marking the mean rather than the median --- and labeling it.

**Practice 17: Add a mean line and label to the histogram**

Four blanks to fill. Two take the function that computes the arithmetic mean: one places the vertical line, the other supplies the value the label reports. One is the label's text, which should name what the line marks so a reader can interpret it without reading your prose. The last is `annotate()`'s first argument, naming what kind of annotation to draw. Use `x = 25, y = 900` for the coordinates. Two hints are available. Click "Submit Answer" to check your work.

```{r plot_vertical_and_text, exercise=TRUE, exercise.cap="Practice 17", fig.alt="Histogram of wage growth with 60 bins, overlaid with a dotted vertical line at the mean. A text label naming the mean and reporting its value sits to the right of the peak."}
counties %>%
  filter(!is.na(wage_growth)) %>%
  ggplot(mapping = aes(x = wage_growth)) +
         geom_histogram(fill = "steelblue", color = "white", bins = 60) +
         geom_vline(aes(xintercept = ___(wage_growth, na.rm = TRUE)),
                    color = "firebrick", linetype = "dotted", linewidth = 0.8) +
         annotate(___, x = 25, y = 900, 
                  label = paste0("___", 
                                 round(___(counties$wage_growth, na.rm = TRUE), 2))) +
         labs(title = "Wage Growth in US Counties in 2016", 
              x = NULL, 
              y = "Frequency", 
              caption = "Source: Linn, Nagler, Zilinsky")
```

```{r plot_vertical_and_text-hint-1}
# Hint 1 of 2
# You are marking the center of the distribution on the
# histogram, so a reader can see at a glance where the
# average county sits relative to the whole spread ---
# and then labeling that line with its value.
```

```{r plot_vertical_and_text-hint-2}
# Hint 2 of 2
# Two blanks take mean(), with na.rm = TRUE: the first
# sets the xintercept aesthetic inside geom_vline(), the
# second supplies the value that round() displays. One is
# the text that goes before that value --- name the
# statistic and leave a space and an equals sign, so the
# finished label reads like a sentence fragment. The last
# is annotate()'s first argument: the kind of annotation
# you want, given as a quoted string.
```


```{r plot_vertical_and_text-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  expr_text <- function(x) {
    if (is.null(x)) return("")
    paste(deparse(rlang::get_expr(x)), collapse = "")
  }

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }
  vline_layers <- Filter(function(layer) inherits(layer$geom, "GeomVline"), .result$layers)
  if (length(vline_layers) == 0) {
    fail("Add a `geom_vline()` layer to draw the mean line.")
  }
  expected_mean <- mean(counties$wage_growth, na.rm = TRUE)
  has_mean_line <- any(vapply(vline_layers, function(layer) {
    x_expr <- expr_text(layer$mapping$xintercept)
    fixed_x <- layer$geom_params$xintercept
    grepl("mean\\s*\\(", x_expr) && grepl("wage_growth", x_expr) ||
      (!is.null(fixed_x) && isTRUE(all.equal(as.numeric(fixed_x), expected_mean)))
  }, logical(1)))
  if (!has_mean_line) {
    fail("Set `xintercept` to `mean(wage_growth, na.rm = TRUE)` inside `geom_vline()`.")
  }
  text_layers <- Filter(function(layer) inherits(layer$geom, "GeomText"), .result$layers)
  if (length(text_layers) == 0 && !grepl("annotate\\s*\\(", code_no_comments)) {
    fail("Add an `annotate()` layer to label the mean line.")
  }
  if (length(text_layers) == 0) {
    fail('Fill `annotate()`\'s first blank with `"text"` --- that is the kind of annotation you want here.')
  }
  if (!grepl("round\\s*\\(\\s*mean\\s*\\(", code_no_comments)) {
    fail("The label should display the mean as well. Fill the second blank with the same function you used for the line.")
  }
  shown_label <- if (length(text_layers) > 0) {
    lab_param <- text_layers[[1]]$aes_params$label
    lab_data  <- if (is.data.frame(text_layers[[1]]$data)) text_layers[[1]]$data$label else NULL
    paste(as.character(c(lab_param, lab_data)), collapse = " ")
  } else {
    ""
  }
  if (!grepl("[Mm]ean", shown_label)) {
    fail('Fill the label blank with text naming the statistic. The finished label should read something like "Mean = 4.33", not a bare number.')
  }

  pass("Well done! The mean line and its label are both in place, each computed from the data rather than typed in as a number.")
})
```


Now return to **dem2p_percent** and use a histogram to check the clue you got from the mean and median. If the mean is higher than the median, right skew is one possibility to look for in the histogram.

**Practice 18: Histogram for dem2p_percent**

Every layer here is one you have already seen. And the plot is less work than it looks: the mean line and its label have exactly the same shape as the median line and its label.

Build it in four passes, running the code after each one so you catch mistakes while they are still easy to find:

1. the histogram alone
2. add the median line
3. add the median label
4. copy those two layers, change `median` to `mean`, and adjust the linetype, the label text, and the `y` coordinate

Build a histogram of **dem2p_percent** with median and mean reference lines, each labeled. Four hints are available. Click "Submit Answer" to check your work.

Your plot should have:

- 30 bins, `fill = "skyblue"`, `color = "white"`
- A median line: firebrick, dashed, `linewidth = 0.8`, labeled at `x = 55, y = 280`
- A mean line: firebrick, dotted, `linewidth = 0.8`, labeled at `x = 55, y = 250`
- Labels built with `paste0()`, each naming its statistic and reporting the value rounded to two digits
- Title \"Democratic Vote Share\", x-axis label \"Percent of Two-Party Vote\", y-axis label \"Frequency\", caption \"Source: Linn, Nagler, Zilinsky\", and no subtitle

```{r histogram_dem2p_percent, exercise=TRUE, exercise.cap="Practice 18", fig.alt="Right-skewed histogram of Democratic two-party vote percentages across US counties, with a peak near 25 percent and a long tail toward 100. Two vertical reference lines stand a few points apart just right of the peak: a dashed line at the median of about 30 and a dotted line at the mean of about 33. Two text labels to their right identify each statistic and report its value."}

```

```{r histogram_dem2p_percent-hint-1}
# Hint 1 of 4
# You are building a histogram that shows where both
# measures of center fall, so a reader can see the gap
# between them against the full distribution --- and that
# gap is the visual signature of a skewed distribution.
```

```{r histogram_dem2p_percent-hint-2}
# Hint 2 of 4
# Map dem2p_percent to the x-axis and add
# geom_histogram(), then geom_vline() and annotate() for
# the median --- then the same pair again for the mean.
```

```{r histogram_dem2p_percent-hint-3}
# Hint 3 of 4
# In geom_histogram(), set bins = 30, fill = "skyblue",
# and color = "white". In each geom_vline(), set
# xintercept inside aes() --- median() for one line and
# mean() for the other --- with color = "firebrick" and
# linewidth = 0.8 outside aes().
```

```{r histogram_dem2p_percent-hint-4}
# Hint 4 of 4
# Build each annotation label with paste0(), rounding to
# two digits. Inside annotate(), reach the variable through
# the data frame --- annotate() does not inherit the data
# context from ggplot(). The two labels differ only in the
# function, the text, and the y coordinate.
```


```{r histogram_dem2p_percent-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  expr_text <- function(x) {
    if (is.null(x)) return("")
    paste(deparse(rlang::get_expr(x)), collapse = "")
  }

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }
  if (!is.data.frame(.result$data) || !"dem2p_percent" %in% names(.result$data)) {
    fail("Use **dem2p_percent** from the **counties** data frame for this plot.")
  }

  expected_values <- sort(as.numeric(counties$dem2p_percent[!is.na(counties$dem2p_percent)]))
  actual_values   <- sort(as.numeric(.result$data$dem2p_percent))
  if (!isTRUE(all.equal(actual_values, expected_values, tolerance = 1e-8))) {
    fail("Use all observations of **dem2p_percent** from the **counties** data frame. Do not remove any valid observations.")
  }
  x_expr <- expr_text(.result$mapping$x)
  if (!grepl("dem2p_percent", x_expr)) {
    fail("Map **dem2p_percent** to the x-axis.")
  }
  hist_layers <- Filter(function(layer) inherits(layer$stat, "StatBin"), .result$layers)
  if (length(hist_layers) == 0) {
    fail("Use `geom_histogram()` to create the histogram.")
  }
  has_fill <- any(vapply(hist_layers, function(layer) {
    identical(layer$aes_params$fill, "skyblue")
  }, logical(1)))
  if (!has_fill) {
    fail('Set `fill = "skyblue"` inside `geom_histogram()`.')
  }
  has_white_outline <- any(vapply(hist_layers, function(layer) {
    identical(layer$aes_params$colour, "white") ||
      identical(layer$aes_params$color, "white")
  }, logical(1)))
  if (!has_white_outline) {
    fail('Set `color = "white"` inside `geom_histogram()` to outline the bars.')
  }
  vline_layers <- Filter(function(layer) inherits(layer$geom, "GeomVline"), .result$layers)
  if (length(vline_layers) == 0) {
    fail("Add a `geom_vline()` layer for the median line.")
  }
  expected_median <- median(counties$dem2p_percent, na.rm = TRUE)
  has_median_line <- any(vapply(vline_layers, function(layer) {
    xint_expr <- expr_text(layer$mapping$xintercept)
    fixed_x <- tryCatch(as.numeric(layer$geom_params$xintercept[1]), error = function(e) NA)
    in_aes <- grepl("median\\s*\\(", xint_expr) && grepl("dem2p_percent", xint_expr)
    in_code <- grepl("median\\s*\\(.*dem2p_percent", code_no_comments)
    in_fixed <- !is.na(fixed_x) && isTRUE(all.equal(round(fixed_x, 2), round(expected_median, 2)))
    in_aes || in_code || in_fixed
  }, logical(1)))
  if (!has_median_line) {
    fail("Inside `geom_vline()`, set `xintercept` to the median of **dem2p_percent**. You can compute it inside `aes()` or pass the value directly.")
  }
  expected_mean <- mean(counties$dem2p_percent, na.rm = TRUE)
  line_role <- function(layer) {
    xint_expr <- expr_text(layer$mapping$xintercept)
    fixed_x <- tryCatch(as.numeric(layer$geom_params$xintercept[1]), error = function(e) NA)
    if (grepl("median\\s*\\(", xint_expr) ||
        (!is.na(fixed_x) && isTRUE(all.equal(round(fixed_x, 2), round(expected_median, 2))))) {
      return("median")
    }
    if (grepl("mean\\s*\\(", xint_expr) ||
        (!is.na(fixed_x) && isTRUE(all.equal(round(fixed_x, 2), round(expected_mean, 2))))) {
      return("mean")
    }
    NA_character_
  }
  roles <- vapply(vline_layers, line_role, character(1))
  median_layers <- vline_layers[!is.na(roles) & roles == "median"]
  if (length(median_layers) > 0 &&
      !identical(median_layers[[1]]$aes_params$linetype, "dashed")) {
    fail('The dashed line should be the median. Set `linetype = "dashed"` on the `geom_vline()` whose `xintercept` is the median.')
  }
  has_dashed_line <- any(vapply(vline_layers, function(layer) {
    identical(layer$aes_params$linetype, "dashed")
  }, logical(1)))
  if (!has_dashed_line) {
    fail('Set `linetype = "dashed"` for the median line.')
  }
  all_firebrick <- all(vapply(vline_layers, function(layer) {
    identical(layer$aes_params$colour, "firebrick") ||
      identical(layer$aes_params$color, "firebrick")
  }, logical(1)))
  if (!all_firebrick) {
    fail('Set `color = "firebrick"` for both reference lines.')
  }
  all_weighted <- all(vapply(vline_layers, function(layer) {
    lw <- layer$aes_params$linewidth
    if (is.null(lw)) lw <- layer$aes_params$size
    !is.null(lw) && isTRUE(all.equal(as.numeric(lw), 0.8))
  }, logical(1)))
  if (!all_weighted) {
    fail("Set `linewidth = 0.8` on both reference lines so they are visible against the bars.")
  }
  text_layers <- Filter(function(layer) inherits(layer$geom, "GeomText"), .result$layers)
  if (length(text_layers) == 0 && !grepl("annotate\\s*\\(", code_no_comments)) {
    fail("Add an `annotate()` layer to label the median line.")
  }
  has_coords <- grepl("x\\s*=\\s*55", code_no_comments) && grepl("y\\s*=\\s*280", code_no_comments)
  if (!has_coords) {
    fail("Use `x = 55` and `y = 280` in `annotate()`.")
  }
  if (!grepl("bins\\s*=\\s*30", code_no_comments)) {
    fail("Set `bins = 30` inside `geom_histogram()` so the plot uses the requested level of detail.")
  }
  if (!grepl("paste0\\s*\\(", code_no_comments)) {
    fail("Build the annotation label with `paste0()` so the text combines the name of the statistic with its computed value.")
  }
  if (!grepl("round\\s*\\(", code_no_comments)) {
    fail("Round the median before displaying it, so the label reports a readable number rather than every decimal place.")
  }
  if (!grepl("round\\s*\\(\\s*median\\s*\\(", code_no_comments)) {
    fail("The label should display the computed median, not the word alone --- round the median inside the label.")
  }

  if (!grepl("Median", code_no_comments)) {
    fail('Use a label that identifies the line as the median, such as "Median = " followed by the rounded value using `paste0()`.')
  }

  if (length(vline_layers) < 2) {
    fail("Add a second `geom_vline()` for the mean, copied from the median line and edited.")
  }
  has_mean_line <- any(vapply(vline_layers, function(layer) {
    xint_expr <- expr_text(layer$mapping$xintercept)
    fixed_x <- tryCatch(as.numeric(layer$geom_params$xintercept[1]), error = function(e) NA)
    (grepl("mean\\s*\\(", xint_expr) && grepl("dem2p_percent", xint_expr)) ||
      grepl("mean\\s*\\(.*dem2p_percent", code_no_comments) ||
      (!is.na(fixed_x) && isTRUE(all.equal(round(fixed_x, 2), round(expected_mean, 2))))
  }, logical(1)))
  if (!has_mean_line) {
    fail("Inside the second `geom_vline()`, set `xintercept` to the mean of **dem2p_percent**.")
  }
  mean_layers <- vline_layers[!is.na(roles) & roles == "mean"]
  if (length(mean_layers) > 0 &&
      !identical(mean_layers[[1]]$aes_params$linetype, "dotted")) {
    fail('The dotted line should be the mean. Set `linetype = "dotted"` on the `geom_vline()` whose `xintercept` is the mean, so the two lines stay distinguishable.')
  }
  has_dotted_line <- any(vapply(vline_layers, function(layer) {
    identical(layer$aes_params$linetype, "dotted")
  }, logical(1)))
  if (!has_dotted_line) {
    fail('Set `linetype = "dotted"` for the mean line, so it is distinguishable from the dashed median line.')
  }
  if (!grepl("round\\s*\\(\\s*mean\\s*\\(", code_no_comments)) {
    fail("The second label should display the computed mean, rounded the same way the median was.")
  }
  if (!grepl("y\\s*=\\s*250", code_no_comments)) {
    fail("Place the mean label at `y = 250` so it sits below the median label.")
  }
  if (length(gregexpr("x\\s*=\\s*55", code_no_comments)[[1]]) < 2) {
    fail("Both labels sit at `x = 55` --- give the mean annotation the same `x` position as the median annotation.")
  }
  if (!grepl("Mean", code_no_comments)) {
    fail('Use a label that identifies the second line as the "Mean".')
  }

  if (!identical(.result$labels$title, "Democratic Vote Share")) {
    fail('Check the plot title. Use "Democratic Vote Share".')
  }
  if (!identical(.result$labels$x, "Percent of Two-Party Vote")) {
    fail('Set the x-axis label to "Percent of Two-Party Vote".')
  }
  if (!identical(.result$labels$y, "Frequency")) {
    fail('Set the y-axis label to "Frequency".')
  }
  if (!identical(.result$labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Check the caption. Use "Source: Linn, Nagler, Zilinsky".')
  }

  pass("Well done! Both reference lines are computed from the data, drawn with distinct line types, and labeled.")
})
```

The distribution is clearly **right-skewed**. A large concentration of counties have Democratic vote percentages well below the median, while fewer counties gave high percentages of their two-party vote to the Democratic candidate. The long right tail shows that only a small subset of counties cast very large percentages of their two-party votes for Clinton.

## Visualizing an Interval Variable: Box Plots

A [**box plot**]{.important-text} is a second way to visualize an interval variable. It brings the median, the IQR, and outlying observations together in a single compact summary. That compactness is what makes it the standard tool for comparing a distribution across categories of a second variable --- several box plots fit side by side where several histograms would not. Here you will build them one variable at a time.

To generate a box plot, change the basic histogram plot in two ways. First, map the variable to the y-axis rather than the x-axis. This is because a box plot displays the distribution vertically --- the median line runs horizontally across the box, the whiskers extend up and down, and outliers appear as dots above or below. If the variable were mapped to x instead, the plot would be rotated 90 degrees and harder to read in the standard orientation. Use `x = ""` to give the box a single position to sit at; without it the box stretches across the whole plot area. That is a separate step from removing the axis label, which is done with `x = NULL` inside `labs()` --- the first controls where the box goes, the second controls what is written beneath it. Second, use `geom_boxplot()` rather than `geom_histogram()`.

Let's plot the number of jobs per 100 working-age residents, which is **employ_pop_ratio_25_64**. 

**Run and observe 19: Plot the jobs-to-population ratio with a box plot**
```{r boxplot_employment_ratio, exercise=TRUE,exercise.cap="Run and observe 19", fig.alt="Box plot of the number of jobs per 100 residents aged 25 to 64 across US counties, plotted vertically. The box is short and sits low on the scale, with the median line near 67. The whiskers are short relative to the axis. Many individual points are plotted above the upper whisker, thinning out as they rise, with the highest near 660."}
counties %>%
  filter(!is.na(employ_pop_ratio_25_64)) %>%
  ggplot(aes(x = "", y = employ_pop_ratio_25_64)) +
  geom_boxplot() +
  labs(title = "Distribution of Jobs per 100 Working-Age Residents",
       subtitle = "US Counties, 2016",
       x = NULL,
       y = "Jobs per 100 residents aged 25-64",
       caption = "Source: Linn, Nagler, Zilinsky") 
```
A box plot has four parts:

- **The heavy line inside the box** --- the median.
- **The box** --- the middle 50% of the data, from the 25th to the 75th percentile. Its height is the interquartile range.
- **The whiskers** --- they reach to the most extreme value within 1.5 × IQR of the box.
- **The circles** --- observations beyond the whiskers, plotted individually.

All four are worth checking against what you already computed. The median line sits at 66.81, the value `quantile()` gave you. The box height is that same interquartile range, now shown as a distance rather than a pair of numbers. And the fence falls at 83.53 + 1.5 × 32.30, or about 132 --- so every circle is a county above that, the extreme cases of the commuting-in effect you met when you computed the percentiles.

Two things are worth saying before you build one yourself. The first is that 1.5 is a convention, not a result. Nothing about the data determines it --- it is a rule of thumb that has become standard because it flags a manageable number of cases in most distributions. A circle on a box plot means "this value is far enough from the middle half to be worth a look," not "this value is an error" and not "this value fails a test." You already know that some of the counties above the fence are recording something real.

The second is what a box plot gives up. It shows the median, the two quartiles, the whisker endpoints, and the observations beyond them --- and everything else about the distribution is gone. Two variables with identical quartiles can have completely different shapes --- one with a single peak in the middle, another with two peaks and a hollow between them --- and their box plots would be indistinguishable. The histogram you built for **dem2p_percent** showed a shape; a box plot of the same variable will not. That is the trade: you lose the shape and gain a summary compact enough to repeat across many groups.

**Practice 20: Create a box plot for dem2p_percent**

Produce the box plot. Then look at where the median line sits within the box and whether the whiskers extend evenly before reading the interpretation below. Three hints are available. Click "Submit Answer" to check your work.

Your plot should have:

- Title: \"Democratic Vote Share\"
- `x = ""` inside `aes()`, with **dem2p_percent** mapped to `y`
- x-axis label: `NULL`, so the plot expands into the space a label would occupy
- y-axis label: \"Percent of Two-Party Vote\"
- Caption: \"Source: Linn, Nagler, Zilinsky\"

```{r boxplot_dem2p, exercise=TRUE, exercise.cap="Practice 20", fig.alt="Vertical box plot of Democratic two-party vote percentages across US counties. The median line sits near 30 percent and the box runs from roughly 21 to 42. The lower whisker reaches close to 0 with no points beyond it, while a dense column of individually plotted counties rises above the upper whisker toward 95 percent --- the long upper tail of a right-skewed distribution."}

```

```{r boxplot_dem2p-hint-1}
# Hint 1 of 3
# A box plot summarizes a single interval variable through
# its percentiles: the box spans the middle half of the
# data, the line inside marks the median, and the whiskers
# reach out toward the extremes. Because there is only one
# variable, there is nothing to spread along the x-axis.
```

```{r boxplot_dem2p-hint-2}
# Hint 2 of 3
# Build the plot from counties with ggplot().
# Map the empty string to x and dem2p_percent to y,
# then add the box plot layer.
# Use labs() for the title, the y-axis label, and the caption.
```

```{r boxplot_dem2p-hint-3}
# Hint 3 of 3
# Two separate things control the x axis here.
# Mapping x to an empty string gives the box a single
# position to sit at, since there is no grouping variable.
# Setting the x-axis label to NULL in labs() removes the
# label entirely, so the plot expands into the space it
# would otherwise occupy.
```


```{r boxplot_dem2p-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)
  expr_text <- function(x) {
    if (is.null(x)) return("")
    paste(deparse(rlang::get_expr(x)), collapse = "")
  }

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. If you're seeing an error rather than a plot, the most common causes are: a missing `+` at the end of a line, unmatched parentheses, or a label string that is missing its quotation marks.")
  }
  if (!is.data.frame(.result$data) || !"dem2p_percent" %in% names(.result$data)) {
    fail("Use **dem2p_percent** from the **counties** data frame for this plot.")
  }
  expected_values <- sort(as.numeric(counties$dem2p_percent[!is.na(counties$dem2p_percent)]))
  actual_values   <- sort(as.numeric(.result$data$dem2p_percent))
  if (!isTRUE(all.equal(actual_values, expected_values, tolerance = 1e-8))) {
    fail("Use all observations of **dem2p_percent** from the **counties** data frame. Do not remove any valid observations.")
  }
  x_expr <- expr_text(.result$mapping$x)
  if (!identical(x_expr, "\"\"") && !identical(x_expr, "''")) {
    fail('Use `x = ""` inside `aes()` for a cleaner one-variable box plot.')
  }
  y_expr <- expr_text(.result$mapping$y)
  if (!grepl("dem2p_percent", y_expr)) {
    fail("Map **dem2p_percent** to the y-axis.")
  }
  has_boxplot <- any(vapply(.result$layers, function(layer) {
    inherits(layer$geom, "GeomBoxplot")
  }, logical(1)))
  if (!has_boxplot) {
    fail("Use `geom_boxplot()` to create the box plot.")
  }
  if (!identical(.result$labels$title, "Democratic Vote Share")) {
    fail('Use "Democratic Vote Share" as the title.')
  }
  if (!is.null(.result$labels$x)) {
    fail("Set the x-axis label to `NULL` with `labs(x = NULL)` so no unused label appears beneath the box plot.")
  }
  if (!identical(.result$labels$y, "Percent of Two-Party Vote")) {
    fail('Set the y-axis label to "Percent of Two-Party Vote".')
  }
  if (!identical(.result$labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Check the caption. Use "Source: Linn, Nagler, Zilinsky".')
  }

  pass("Good work! You built the whole plot from scratch. The fiddly part of a single-variable box plot is giving the box somewhere to sit on the x-axis and then clearing the label that would otherwise hang beneath it.")
})
```

The box plot for Democratic vote percentage shows a median around 30%, consistent with what you computed earlier. The IQR box spans roughly 21% to 42%, meaning the middle half of counties gave Clinton between about 21% and 42% of the two-party vote. The long upper whisker and outlier circles above it indicate that a small number of counties gave Clinton very high vote percentages --- consistent with the right-skewed distribution you identified from the mean exceeding the median.

## Check Your Understanding

Now let's step back and practice choosing among the tools from this tutorial. The questions below ask you to connect the level of measurement to the appropriate summary statistic or plot, and then to interpret what those summaries tell you about the center, spread, and shape of a distribution.

**Apply 21: Identify core distribution concepts**

```{r identify_distribution_concepts, echo=FALSE}
question("Which of the following statements are true? Select all that apply.",
  correct = "Right. Both statements are true.",
  answer("The mean is a robust summary statistic",
         message = "The mean is not robust --- it is sensitive to outliers. A single extreme value can pull the mean far from the center of the distribution. The median is the robust alternative."),
  answer("A distribution whose shape is skewed right has a long tail on the right", correct = TRUE,
         message = "The direction of skew is named for the tail, not the peak. A right-skewed distribution has most values clustered on the left and a long tail extending to the right. The terminology can be counterintuitive because the peak is on the left, not the right --- it is the tail that gives skew its direction."),
  answer("The standard deviation measures the typical distance of observations from the mean", correct = TRUE,
         message = "The standard deviation is the typical distance of observations from the mean in the original units of the variable. A larger standard deviation means observations are more dispersed; a smaller one means they cluster more tightly around the mean."),
  answer("The mode should be reported as a measure of central tendency regardless of a variable's level of measurement",
         message = "The mode is generally not useful for interval variables, where values are so precisely measured that no single value repeats often enough to be informative. For interval data, the median and mean are the appropriate measures of central tendency."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Two of these statements are true. Take each one separately and ask whether it holds for every distribution or only for some."
)
```

**Apply 22: Compare two standard deviations**

```{r cyu-sd-magnitude, echo=FALSE}
question("A researcher reports that county unemployment has a standard deviation of 4 and county median income has a standard deviation of 12,000. Which variable is more spread out?",
  correct = "Right. The two standard deviations are not comparable as they stand.",
  answer("You cannot tell, because the variables are measured in different units", correct = TRUE,
         message = "A standard deviation carries the units of its variable. Four percentage points and twelve thousand dollars are not on the same scale, so comparing the raw numbers says nothing about which variable varies more. To compare spread across variables measured differently, you need to relate each standard deviation to the scale of its own variable."),
  answer("Median income, because 12,000 is much larger than 4",
         message = "The larger number reflects the units, not more variation. Income measured in dollars will produce a large standard deviation simply because dollars are small units; the same incomes measured in thousands would give a standard deviation of 12."),
  answer("Unemployment, because percentages are bounded between 0 and 100",
         message = "Being bounded limits how large the standard deviation can get, but it does not make a variable more spread out. The comparison still fails because the two are measured in different units."),
  answer("They are equally spread out, because both standard deviations describe typical distance from the mean",
         message = "Both do describe typical distance from the mean --- in their own units. That shared definition is exactly why the two numbers cannot be compared directly."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "One standard deviation is in percentage points and the other is in dollars. Ask whether those two numbers are on the same scale."
)
```

Consider the following histogram and answer the question below. Public expenditure on education is recorded for 150 of the 167 countries, so the histogram and the standard deviation it reports describe those 150 rather than the full set.

```{r cyu-spendeduc-histogram, echo=FALSE, fig.alt="Histogram of public expenditure on education as a percentage of GDP across countries. Bars show the number of countries in each spending interval, and a text annotation reports a standard deviation of 1.92."}
world %>%
  filter(!is.na(spendeduc)) %>%
ggplot(aes(x = spendeduc)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 20) +
  labs(title = "Public Expenditure on Education Around the World",
       caption = "Country Data from the United Nations",
       x = "Percent of GDP",
       y = "Frequency") +
          annotate("text", x = 10, y = 18, 
                  label = paste0("Standard Deviation = ", 
                                 round(sd(world$spendeduc, na.rm = TRUE), 2))) 
```

**Apply 23: Interpret the education-spending histogram**

```{r plot_interpretation, echo=FALSE}
question("In this histogram, most countries fall between 2 and 7 percent of GDP, and two sit near 12 percent. Which of the following are true? Select all that apply.",
  correct = "Right. Both statements are true.",
  answer("The outliers near 12 percent of GDP are likely pulling the mean above the median", correct = TRUE,
         message = "Extreme values in the upper tail can pull the mean upward much more strongly than they affect the median. That is the defining consequence of right skew with high outliers: the mean is pulled away from the center while the median stays much closer to where most observations sit."),
  answer("The median is a more robust summary of the typical country's education spending than the mean", correct = TRUE,
         message = "The median is much less affected by the countries spending 10 percent or more, while the mean is pulled upward by them. When outliers are present and the distribution is skewed, the median better represents what is typical."),
  answer("The mean is likely a better measure of central tendency than the median",
         message = "The right tail contains a small number of countries spending much more than typical. Those values pull the mean upward much more strongly than they affect the median, making the median the more robust measure here."),
  answer("The distribution appears bimodal, with two distinct clusters of countries at different spending levels",
         message = "The histogram shows a single peak near 4 to 5 percent of GDP. The isolated bars near 12 percent are outliers, not a second peak of comparable size."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Two of these are true. The outliers sit well above the bulk of the distribution --- ask which statistic that pulls, and in which direction."
)
```

**Apply 24: Read a box plot**

Read the numbers in the question carefully before deciding what the whisker ends can be.

```{r cyu-boxplot-whiskers, echo=FALSE}
question("A box plot of a variable shows a box extending from 20 to 45 with a line at 30. The lower whisker ends at 8, the upper whisker ends at 78, and three points are drawn individually above 90. What do the ends of the whiskers represent?",
  correct = "Right. The whiskers stop at the most extreme values that are not plotted separately.",
  answer("The most extreme observations not far enough out to be plotted individually", correct = TRUE,
         message = "The whiskers extend to the furthest observations that are not far enough from the box to be singled out. That is why the upper whisker stops at 78 while three observations above 90 appear as separate points --- those three lie beyond the whisker's reach and are shown individually so you can see them."),
  answer("The minimum and maximum values of the variable, so nothing lies beyond them",
         message = "If the whiskers marked the minimum and maximum, nothing would ever be plotted beyond them. Here three points sit above 90, so the upper whisker at 78 cannot be the maximum."),
  answer("One standard deviation below and above the mean of the variable",
         message = "Box plots are built from percentiles, not from the mean and standard deviation. Nothing in a box plot is computed from either."),
  answer("The 10th and 90th percentiles, matching the percentile logic of the box",
         message = "The box edges are percentiles, but the whisker ends are not fixed percentiles --- they land on actual observations, and where they land depends on how spread out the extreme values happen to be."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Three points are drawn above 90, beyond the upper whisker at 78. What does that tell you about what the whisker end can and cannot be?"
)
```

## The Takeaways

::: {.qpa-infographic}
<table class="qpa-table">
  <caption>Univariate Description</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Aspect of the distribution</span></th>
      <th scope="col">Nominal Level Variable</th>
      <th scope="col">Ordinal Level Variable</th>
      <th scope="col">Interval Level Variable</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Central Tendency</th>
      <td>Mode</td>
      <td>Mode, Median</td>
      <td>Mode, Median, Mean</td>
    </tr>
    <tr>
      <th scope="row">Dispersion</th>
      <td>Frequency or Proportion Distribution</td>
      <td>Frequency or Proportion Distribution</td>
      <td>Range, Percentiles (IQR), Standard Deviation, Variance</td>
    </tr>
    <tr>
      <th scope="row">Distribution Shape</th>
      <td>Unimodal, Bimodal, Multimodal, Uniform</td>
      <td>Unimodal, Bimodal, Multimodal, Uniform; Direction Across the Scale</td>
      <td>Unimodal, Bimodal, Multimodal, Uniform, Symmetric, Skewed</td>
    </tr>
    <tr>
      <th scope="row">Visualization</th>
      <td>Bar Plot of Frequencies or Proportions</td>
      <td>Bar Plot of Frequencies or Proportions</td>
      <td>Histogram, Box Plot</td>
    </tr>
    <tr>
      <th scope="row">R Functions</th>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>mutate()</code>, <code>filter()</code>, <code>geom_bar()</code></td>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>mutate()</code>, <code>filter()</code>, <code>median()</code>, <code>factor()</code>, <code>geom_bar()</code></td>
      <td><code>mean()</code>, <code>median()</code>, <code>range()</code>, <code>quantile()</code>, <code>sd()</code>, <code>filter()</code>, <code>geom_histogram()</code>, <code>geom_boxplot()</code></td>
    </tr>
  </tbody>
</table>
:::

The table brings together the tools for describing **nominal**, **ordinal**, and [**interval variable**]{.important-text}s across all three tutorials. Everything that follows explains why the entries differ and how to produce them in R.

Interval variables give you a wider set of descriptive tools because their values are ordered and the distances between values are meaningful. The same three dimensions apply at every level of measurement --- central tendency, dispersion, and shape --- but the tools available depend on what the values actually mean. The median requires a meaningful rank order, so it is not appropriate for nominal variables. The mean assumes equal intervals between values, so it is not appropriate for ordinal variables. Interval variables permit the mean in addition to the median and mode, though the mode is rarely informative for precisely measured data.

- [**Central tendency**]{.important-text} for interval variables is described by the **median** and [**mean**]{.important-text}, computed with `median()` and `mean()`, adding `na.rm = TRUE` when the variable has missing values. The mode is almost never useful for interval data because values are often measured so precisely that no single value repeats very often. The median is the value at the 50th percentile and is robust to [**outliers**]{.important-text}. The mean is the arithmetic average and can be pulled away from the center by extreme values. When the distribution is roughly [**symmetric**]{.important-text}, the two are similar. When they diverge, inspect the distribution for skew or outliers. In a strongly [**skewed**]{.important-text} distribution the median often better represents a typical case, because it is not pulled by extreme values.

- [**Dispersion**]{.important-text} for interval variables is captured by the [**range**]{.important-text} (`range()`), [**percentile**]{.important-text}s (`quantile()`), and [**standard deviation**]{.important-text} (`sd()`). The [**interquartile range**]{.important-text}, or [**IQR**]{.important-text}, is the distance from the 25th to the 75th percentile; it describes the middle half of the data and is less affected by outliers than the full range. The standard deviation is the typical distance of observations from the mean, in the original units of the variable; the [**variance**]{.important-text} is its square, in squared units, which is why the standard deviation is the one you report. One important caution: standard deviations are only comparable within a variable or across samples of the same variable.

- The [**shape of the distribution**]{.important-text} of an interval variable is best assessed visually. A [**histogram**]{.important-text} (`geom_histogram()`) groups values into bins and shows the overall pattern of the distribution --- its center, spread, symmetry or skew, and whether it is [**unimodal**]{.important-text}, [**bimodal**]{.important-text}, or uniform. The `bins` and `binwidth` arguments affect how much detail you see. A [**box plot**]{.important-text} (`geom_boxplot()`) summarizes the median, IQR, and outliers in a compact form that becomes especially useful when comparing distributions across groups. When reading either plot, ask: where is the center, how spread out are the values, is the distribution symmetric or skewed, and are there outliers?

In R, load [dplyr]{.package-name} and [ggplot2]{.package-name} once at the top of your script or Quarto document with `library(dplyr)` and `library(ggplot2)` before any code that uses functions from those packages. The analysis tools introduced in this tutorial are `median()`, `mean()`, `range()`, `min()`, `max()`, `quantile()`, and `sd()` --- all take the variable as their first argument, and `na.rm = TRUE` as a second when the variable has missing values.

For visualization, histograms and box plots follow the same layered ggplot structure as bar charts. When a plotted interval variable has missing values, filter those cases before passing the data to `ggplot()` --- otherwise ggplot prints a warning that would appear in a knitted document.

- **Histograms:** Use `geom_histogram()` in place of `geom_bar()`, with `fill` to set the bar color, `color = "white"` to add bar outlines, and `bins` or `binwidth` to control bin size.

- **Reference lines:** Use `geom_vline()` to mark a statistic. Inside `aes()`, bare variable names work: `aes(xintercept = median(wage_growth, na.rm = TRUE))`. Outside `aes()`, use the full `data$variable` syntax. Fixed settings like `color` and `linetype` always go outside `aes()`. To label the line, add `annotate("text", x = ..., y = ..., label = ...)` --- use `counties$variable` inside `annotate()` since it does not inherit the data context from `ggplot()`.

- **Box plots:** Map the variable to `y` rather than `x`, use `geom_boxplot()`, and set `x = ""` to suppress the unnecessary x-axis category label.

Which summaries should you present? Report the mean or median and the standard deviation or IQR, or show a histogram or box plot --- but do not bury your reader in every statistic and every graph just because you can. Tables and plots often communicate overlapping information, so choose the display that best serves your purpose rather than presenting both automatically. Whatever you choose, describe its important features in the text --- do not leave a table or figure to speak for itself. The goal is always to give readers enough to see the center, spread, and shape of the distribution.

Tutorial 7 turns from one variable to two. The question changes from what a variable looks like on its own to whether two variables are related, and the level of measurement continues to decide which tools apply --- now for both variables at once.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-06, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-06, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-06, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Well done! You can now describe and visualize interval variables using measures of central tendency, dispersion, and distribution shape.">
  <span aria-hidden="true">📊</span>
  Well done! You can now describe and visualize interval variables using measures of central tendency, dispersion, and distribution shape.
</p>
')
```
