---
title: "Hypothesis Tests: Two Categorical Variables"
tutorial:
  id: "12-Hypotheses-Categorical"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to conduct a hypothesis test of the relationship between two categorical variables.
---

## Overview {#main-content}

Tutorials 7, 8, and 9 described bivariate relationships --- cross-tabs, group means, scatter plots --- but every example ended with the same caution: a descriptive pattern consistent with a hypothesis is not the same as evidence that the pattern is real. In Tutorials 10 and 11, you built the tools a formal test requires: the sampling distribution, the p-value, and the decision rule for rejecting or failing to reject a null hypothesis. Tutorials 12, 13, and 14 put those tools to work. This tutorial takes the cross-tab relationships from Tutorial 7 and subjects them to formal [**hypothesis tests**]{.important-text} --- using sample data to assess whether an observed relationship is strong enough to rule out chance. Tutorial 13 does the same for the group-mean relationships from Tutorial 8, and Tutorial 14 for the correlations from Tutorial 9.

In this tutorial, you will learn how to test whether two categorical variables are related in the population. Specifically, we will cover:

* How to test whether two categorical variables are related using a chi-square test of independence, with the `CrossTable()` function in the [descr]{.package-name} package
* How to decide when a chi-square test is the right tool, when a difference-of-proportions test is, and how building a cross-tab lets you ask the general question before deciding whether to sharpen to a specific one
* How to organize a hypothesis test into five explicit steps, building on the null hypothesis, alternative hypothesis, p-value, and decision rule from Tutorial 11
* How to interpret a p-value and apply the decision rule to determine whether a relationship is statistically significant
* How to test whether a single sample proportion differs from a hypothesized population value


```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(descr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
counties$rural <- 1
counties$rural[counties$rural_urban <= 3] <- 0
counties$rural[counties$rural_urban >= 8] <- 2
```


## Testing the Relationship Between Two Categorical Variables

In Tutorial 7, you read a cross-tab by comparing column proportions across categories of an explanatory variable. If those proportions differed in the direction a hypothesis predicted, the pattern was consistent with the hypothesis. But a sample is just one sample --- if we drew a different sample from the same population, the proportions would differ at least a little, just by chance.

A word about the data frames in this tutorial. Neither **counties** nor **world** is a random sample from a larger population the way a survey is. As Tutorial 11 explained, inference is still appropriate when the research question is about a relationship --- a systematic pattern that could have looked different under different conditions. A p-value answers: if there were truly no relationship, how surprising would a pattern this strong be in these data?

We test whether two categorical variables are related using two complementary tools. The [**chi-square test of independence**]{.important-text} considers every category at once and asks whether the variables are related at all. A [**difference-of-proportions test**]{.important-text} asks a narrower but sharper question: comparing exactly two specific groups, it can test a specific directional prediction in a way the chi-square test cannot. Since building a cross-tab already gives us everything the chi-square test needs, we test the general claim first, then sharpen to a directional two-group test when the hypothesis supports one. We also look at a related but different question: whether a single sample proportion differs from some fixed hypothesized value.

The logic starts with a question: what would the cross-tab look like if the two variables had no relationship at all --- if they were completely independent? We can calculate exactly that. For any cell in the table, the [**expected count**]{.important-text} under independence is:

$$E=\frac{(\text{row total})(\text{column total})}{n}$$

where $n$ is the total number of cases in the table. This formula says: if the two variables are independent, each cell should contain its fair share of cases, in proportion to the row and column totals.

Let's see this with a small example. Suppose 100 people were each categorized by two variables: Group (A or B) and Outcome (Yes or No). Fifty people are in Group A and fifty are in Group B; 60 people said Yes and 40 said No, overall. If Group and Outcome were independent, here is what we would expect to see:

| | Yes | No | Total |
|:--|:--:|:--:|:--:|
| **Group A** | 30 | 20 | 50 |
| **Group B** | 30 | 20 | 50 |
| **Total** | 60 | 40 | 100 |

Now suppose this is what we actually observed:

| | Yes | No | Total |
|:--|:--:|:--:|:--:|
| **Group A** | 40 | 10 | 50 |
| **Group B** | 20 | 30 | 50 |
| **Total** | 60 | 40 | 100 |

Group A said "Yes" more often than independence would predict (40 observed vs. 30 expected), and Group B said "Yes" less often than independence would predict (20 observed vs. 30 expected). This is exactly the kind of pattern a chi-square test is built to detect.

### Calculating the Chi-Square Statistic

The [**chi-square statistic**]{.important-text}, denoted $\chi^2$, summarizes how far the observed counts are from the expected counts, across every cell in the table:

$$\chi^2=\Sigma \frac{(O-E)^2}{E}$$

where $O$ is the observed count in a cell and $E$ is the expected count in that same cell, calculated as above.

Think through what this formula tells us. For each cell, we square the gap between what we observed and what independence would predict, then divide by the expected count, so that cells with naturally larger expected counts do not dominate just because they are bigger. We add that result up across every cell in the table. When observed counts are close to expected counts everywhere, each term is small, and $\chi^2$ is small. When observed counts differ substantially from expected counts, $\chi^2$ is large.

For example, the Group A/Yes cell contributes $(40-30)^2/30=3.33$ to the total, and the Group B/Yes cell contributes $(20-30)^2/30=3.33$. Working through all four cells the same way and adding them up gives $\chi^2=16.67$ for this table.

In Tutorials 10 and 11, the sampling distribution you built intuition for was the normal curve --- symmetric, bell-shaped, centered on the population parameter. $\chi^2$ follows a different distribution entirely. Because it is built from squared terms, $\chi^2$ can never be negative, so its distribution starts at 0 and stretches out in a long tail to the right, rather than spreading symmetrically in both directions the way the normal curve does. The exact shape depends on the table's degrees of freedom, which is why R reports them alongside the test statistic --- but the precise calculation is one more thing we will not dwell on in this course.

What does not change is the underlying logic. Just as with the normal distribution, a hypothesis test locates your observed statistic on this distribution and asks how far into the tail it falls. A $\chi^2$ value far out in that long right tail is one that would be rare if the two variables were truly independent --- exactly the kind of evidence a hypothesis test is built to evaluate. R converts $\chi^2$ into a p-value this way automatically, and we apply the same decision rule as always: a small p-value (conventionally below .05) leads us to reject the null hypothesis of independence, while a large p-value means we fail to reject it.

### Running the Test with `CrossTable()`

We already know how to build a cross-tab with the `CrossTable()` function in the [descr]{.package-name} package. The `x` argument takes the **outcome variable** and `y` takes the **explanatory variable** --- `x` defines the rows and `y` defines the columns, so with `prop.c = TRUE` each column proportion shows how the outcome distribution varies across categories of the explanatory variable. To add the chi-square test, we set the argument `chisq = TRUE`. This adds the chi-square statistic, its degrees of freedom, and its p-value to the output. The degrees of freedom are not something we will focus on in this course --- R reports them, but the test statistic and the p-value are what we use to reach a conclusion.

We will also set `expected = TRUE`, which adds the expected count --- calculated with the formula above --- to every cell. Each cell will now print three numbers, in order: the observed count, the expected count, and the column proportion you already know how to read.

There is a second way to ask a related question. A [**difference-of-proportions test**]{.important-text}, run with `prop.test()`, compares the proportion of "successes" in one group to the proportion of "successes" in another. It compares exactly two groups at a time. Unlike the chi-square test, it can test a specific direction --- a one-tailed test, when a hypothesis predicts which group should have the higher proportion. Because building a cross-tab already gives us the counts a chi-square test needs, our usual approach will be to test the general claim first --- are these two variables related at all --- and then, when the hypothesis makes a specific, directional prediction about two groups, sharpen to a difference-of-proportions test as well.

### Running the Test with `prop.test()`

To run a difference-of-proportions test, we use the `prop.test()` function in base R. Unlike `CrossTable()`, this function does not work directly with the raw data --- instead, we supply the counts ourselves, using the cross-tab we already built.

`prop.test()` compares exactly two groups. Think of `x` and `n` as two parallel lists, one entry per group: `x = c(successes in Group 1, successes in Group 2)` and `n = c(total in Group 1, total in Group 2)`. Group 1's count must appear first in *both* vectors; Group 2's count must appear second in *both*. If you mismatch the order in one vector but not the other, R will not give an error --- it will just silently compare the wrong numbers.

In other words: the first entry in `x` is the numerator of the first group's proportion, and the first entry in `n` is that same group's denominator. The second entries in `x` and `n` are the numerator and denominator of the second group's proportion.

The `alternative` argument controls whether the test is one-tailed or two-tailed. The default, `alternative = "two.sided"`, tests whether the two proportions differ in either direction. When a hypothesis predicts a specific direction --- as ours does in the example below --- we instead use `alternative = "less"` or `alternative = "greater"`, matching the direction the hypothesis predicts. The key rule: `alternative = "less"` tests whether the *first* group's proportion is lower than the *second's*; `alternative = "greater"` tests whether the first group's proportion is higher. Which group you list first in `x` and `n` therefore determines which `alternative` setting matches your hypothesis.

Alongside the p-value, `prop.test()` also reports a confidence interval. For a two-group test, that interval is for the difference
between the two proportions. For a one-sample proportion test, it is
for the population proportion itself. When the confidence interval and hypothesis test are matched --- same significance level and same one-tailed or two-tailed specification --- they should point to the same conclusion. For the two-group tests in this tutorial, the hypothesized difference is usually 0. If the interval excludes 0, that matches a p-value small enough to reject the null. For a one-tailed test, one side of the interval may appear as the theoretical extreme (-1 or 1, since a difference in proportions cannot go beyond that range). That is expected, not an error. The bound that matters is the calculated one: does the interval exclude 0? Note that the fixed bound at ±1 does not guarantee this --- if the calculated bound were, say, 0.05, the interval would be (-1, 0.05) and would *include* 0, meaning the test would fail to reject the null despite one bound being fixed. We will rely on the p-value and decision rule throughout this tutorial, but the interval is worth a glance as a cross-check.

### Applying the Five-Step Hypothesis Testing Procedure

In Tutorial 11, you learned the building blocks of a hypothesis test: a null and alternative hypothesis, a test statistic, a p-value, and the decision rule for rejecting or failing to reject the null hypothesis. Here, we organize those same building blocks into five explicit steps, which we will follow for every hypothesis test in this tutorial and in those that follow:

1. State the null hypothesis, $H_0$.
2. State the alternative hypothesis, $H_A$.
3. Run the test and obtain a test statistic.
4. Find the p-value associated with that test statistic.
5. Decide whether to reject or fail to reject $H_0$, and report what that means substantively.

We will follow these same five steps to test relationships between two categorical variables, and we will follow them again in later tutorials for other kinds of hypothesis tests.

## Are Rural Counties More Likely to Support Trump?

The rural/Trump relationship was one of the cleaner patterns in Tutorial 7 --- the column proportions rose steadily with rurality, and the difference between urban and rural counties was nearly 20 percentage points. The argument behind the hypothesis is intuitive: rural and urban communities differ in their economic conditions, social composition, and the political messages that resonate locally, and those differences shaped how receptive voters were to Trump's 2016 campaign. The cross-tab already showed a pattern consistent with that argument. Now we ask whether the pattern is strong enough to rule out chance.

::: {.hypothesis}
In a comparison of [counties], those that are [more rural] will be more likely to [give a majority of their support to Trump] than will those that are [less rural (more urban)].
:::

The data frame **counties** contains **TrumpMajority**, which takes a value of 0 if Clinton won a majority of the vote and 1 if Trump won a majority, and **rural**, a 3-category variable where 0 is more urban, 1 is neither urban nor rural, and 2 is more rural.

The cross-tab you built in Tutorial 7 already contains everything needed for the chi-square test --- we just need to add two arguments to the call. We start with the general claim, then sharpen to the directional one the hypothesis actually makes.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$H_0:$ **rural** and **TrumpMajority** are not related in the population.

#### Step 2: State the Alternative Hypothesis

$H_A:$ **rural** and **TrumpMajority** are related in the population.

#### Step 3: Run the Test

**Run and observe: Run the chi-square test of independence**

The [descr]{.package-name} package must be loaded with `library(descr)` before you can use `CrossTable()`. Once a package is loaded, all of its functions are available for the rest of your R session --- you do not need to load it again each time you use a function from it. The exercise below loads it once; you can assume it is already loaded for all subsequent exercises in this tutorial.

**TrumpMajority** is assigned to `x` because it is our outcome variable --- the one whose distribution we are comparing across groups --- and **rural** is assigned to `y` because it is our explanatory variable, defining the columns across which we read column proportions.

```{r crosstab-trumpmajority-by-rural, exercise=TRUE, exercise.cap="Run and observe: Chi-square test of TrumpMajority and rural"}
library(descr)
CrossTable(x = counties$TrumpMajority,
           y = counties$rural,
           prop.r = FALSE,
           prop.c = TRUE,
           prop.t = FALSE,
           prop.chisq = FALSE,
           chisq = TRUE,
           expected = TRUE)
```

Look at the output carefully before moving on. Each cell now shows three numbers stacked vertically: the observed count, the expected count under independence (the number that would appear if the two variables were completely unrelated), and the column proportion you already know how to read. The chi-square statistic and its p-value appear at the bottom of the output, in the section labeled "Pearson's Chi-squared test." That is where you will look in Steps 4 and 5 to reach a decision.

**Practice: Connect the output to the formula**

```{r quiz-rural-expectedcounts, echo=FALSE}
question("Using the row total, column total, and grand total from your output, what is the expected count for the Clinton-majority/urban cell, and how does it compare to the observed count?",
  correct = "Right. The observed count exceeds the expected count --- urban counties were more Clinton-leaning than statistical independence would predict. The expected count formula gives (row total)(column total)/n. With a Clinton-majority row total of 490, an urban column total of 1162, and n = 3112: (490)(1162)/3112 = 183.0. The observed count of 303 is well above that --- this is exactly the kind of departure from independence that produces a large chi-square statistic.",
  answer("The expected count is below the observed count, meaning urban counties were more likely to be Clinton-majority than independence would predict",
         correct = TRUE),
  answer("(490)(1162)/3112 = 183.0, but this cannot be compared to the observed count directly since they measure different things",
         message = "They can be compared directly --- that comparison is the entire logic behind the chi-square test. Observed well above expected, as it is here, means urban counties were more likely to be Clinton-majority than independence would predict."),
  answer("The expected count cannot be found from the output alone and would require re-running the model with different settings",
         message = "The expected count is just (row total)(column total)/n --- the same formula from earlier in this tutorial --- and it is already printed in your output, as the second number in the cell stack."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The expected count formula is (row total)(column total)/n. Find the row total for Clinton-majority counties, the column total for urban counties, and the grand total. Then compare to the observed count the output prints."
)
```

Now that you can see how far the observed counts stray from what independence would predict, let's look at what the test statistic and p-value tell us about whether that gap is larger than chance would produce.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

```{r quiz-rural-general-result, echo=FALSE}
question("Find the p-value in your output. Which statement best summarizes what this test tells us?",
  correct = "Right. The chi-square establishes a relationship exists, but its p-value does not test the specific direction the hypothesis predicts.",
  answer("We reject the null hypothesis --- rural and TrumpMajority are related in the population --- but the chi-square p-value does not directly test the specific directional contrast our hypothesis predicts",
         correct = TRUE),
  answer("We reject the null hypothesis, which provides evidence consistent with the hypothesis that more rural counties are more likely to support Trump",
         message = "Rejecting this null hypothesis only provides evidence that some relationship exists between rural and TrumpMajority (p < .001). It does not, by itself, evaluate the specific direction our hypothesis predicts --- for that we need the proportions test that follows."),
  answer("We fail to reject the null hypothesis, so there is no relationship between rural and TrumpMajority",
         message = "A p-value this small (p < .001) leads us to reject the null hypothesis, not fail to reject it. Failing to reject would require a p-value above .05."),
  answer("We reject the null hypothesis, and since the relationship runs across all three rurality categories, the directional claim is also established",
         message = "A chi-square test treats all categories simultaneously and its p-value has no direction built into it. Rejecting the null tells us a relationship exists --- to evaluate the specific directional contrast the hypothesis predicts, we need the targeted proportions test that follows."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value at the bottom of your output. Is it above or below .05? Then ask: does rejecting the null here tell you anything about the direction of the relationship?"
)
```

### Testing the Directional Claim

Our hypothesis does not just predict that **rural** and **TrumpMajority** are related --- it predicts a specific direction: Trump support should be higher among rural counties than among urban counties specifically. Looking at the cross-tab, we can see that direction in the column proportions. But the chi-square p-value does not directly test that specific one-tailed contrast. To do that, we compare the two endpoint categories --- urban and rural --- using the counts from the table above.

The hypotheses below use $\pi$ for the population proportion --- the Greek-letter convention from Tutorial 10, where population parameters are written in Greek and the sample statistics that estimate them are written in Roman letters; for a proportion, that sample statistic is $p$.

#### Step 1: State the Null Hypothesis

$H_0: \pi_{urban}=\pi_{rural}$, where $\pi$ is the proportion of counties with a Trump majority in the population.

#### Step 2: State the Alternative Hypothesis

$H_A: \pi_{urban} < \pi_{rural}$

#### Step 3: Run the Test

Reading from the cross-tab you just produced above, Trump won a majority in 859 of the 1162 urban counties, and in 584 of the 627 rural counties.

**Run and observe: Run a one-tailed difference-of-proportions test**

`prop.test()` is part of base R --- it does not require loading any package. We set `alternative = "less"` because the hypothesis predicts that the urban proportion is *lower* than the rural proportion --- not just different from it.

```{r proptest-trumpmajority-urban-rural, exercise=TRUE, exercise.cap="Run and observe: One-tailed difference-of-proportions test"}
prop.test(x = c(859, 584), n = c(1162, 627), alternative = "less")
```

The output reports a 95% confidence interval of (-1.0000000, -0.1640333) for the difference in proportions (urban minus rural). For a one-tailed test, R fixes one bound at the theoretical extreme --- here, -1, since a difference in proportions cannot go below that --- rather than calculating it from the data. The bound that matters is the other one: -0.164. Because the entire interval falls below zero, excluding the null value of 0, this points to the same conclusion we will reach from the p-value. Both the p-value and the confidence interval will be in agreement throughout this tutorial.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

```{r quiz-rural-directional-result, echo=FALSE}
question("Find the p-value in your output. Does this result provide statistically significant evidence in support of the hypothesis?",
  correct = "Right. The p-value is far below .05 and the direction matches the hypothesis. Trump won a majority in 73.9% of urban counties compared to 93.1% of rural counties (p < .001). The confidence interval excludes 0, consistent with the p-value. Combined with the chi-square test, this is statistically significant evidence in support of the hypothesis.",
  answer("Yes --- the p-value is far below .05, and the proportion of Trump-majority counties is significantly lower among urban counties than rural counties, matching the predicted direction",
         correct = TRUE),
  answer("Yes, but only weakly --- the p-value is small because the sample is large, not because the effect is meaningful",
         message = "Sample size does affect the p-value, but that does not make this result meaningless. The proportions themselves --- 73.9% versus 93.1% --- represent a 19-percentage-point difference between urban and rural counties. Both the statistical significance and the magnitude of the difference are relevant to evaluating the hypothesis."),
  answer("We cannot conclude anything about direction from this test alone --- prop.test() only tells us whether the proportions differ, not which is higher",
         message = "prop.test() does tell you which proportion is higher --- the sample proportions appear in the output, and alternative = \"less\" specifically tests whether the first group's proportion is lower. The direction is right there in the output."),
  answer("No --- the p-value is below .05, so we fail to reject the null hypothesis",
         message = "This reverses the decision rule. A p-value *below* the significance threshold (.05) leads us to *reject* the null hypothesis, not fail to reject it. We fail to reject the null only when the p-value is *above* .05 --- meaning the observed pattern would not be surprising if the null were true."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value and check whether it is above or below .05. Then check whether the sample proportions in the output match the direction the hypothesis predicted."
)
```

## Are Majority-White Counties More Likely to Support Trump?

The racial majority example from Tutorial 7 produced one of the sharpest contrasts in the whole tutorial: White Majority counties gave Trump a majority at a dramatically higher rate than any other category. The argument was that racial composition connects to local political context, party coalitions, and the social environments in which campaigns are received.

::: {.hypothesis}
In a comparison of [counties], those that are [majority-white] will be more likely to [give a majority of their support to Trump] than will those that are [not majority-white].
:::

The data frame **counties** contains **TrumpMajority** and **racial_majority**, a nominal variable stored as a character variable that takes the values "Black Majority," "Hispanic Majority," "Other/No Majority," and "White Majority."

As in the previous example, we start with the general claim that the cross-tab can answer for free, then sharpen to the directional claim the hypothesis actually makes.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$H_0:$ **racial_majority** and **TrumpMajority** are not related in the population.

#### Step 2: State the Alternative Hypothesis

$H_A:$ **racial_majority** and **TrumpMajority** are related in the population.

#### Step 3: Run the Test

The cross-tab you built in Tutorial 7 already contains everything the chi-square test needs --- just add `chisq = TRUE` and `expected = TRUE` to the call you already know.

**Practice: Run the chi-square test of independence**

Fill in the outcome variable in `x` and the explanatory variable in `y`. One hint is available. Click Submit Answer to check your work.

```{r crosstab-trumpmajority-by-racialmajority, exercise=TRUE, exercise.cap="Practice: Chi-square test of TrumpMajority and racial_majority"}
library(descr)
CrossTable(x = counties$___,
           y = counties$___,
           prop.r = FALSE,
           prop.c = TRUE,
           prop.t = FALSE,
           prop.chisq = FALSE,
           chisq = TRUE,
           expected = TRUE)
```

```{r crosstab-trumpmajority-by-racialmajority-hint-1}
# Hint 1 of 1
# The structure mirrors the chi-square Run and observe above --- same function,
# same argument pattern, same chisq and expected settings. The only change
# is the variables: the outcome goes in x and the explanatory variable goes in y.
```

```{r crosstab-trumpmajority-by-racialmajority-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (is.null(.result)) {
    fail("Your code did not produce output. Check for unmatched parentheses, a missing comma between arguments, or a misspelled argument name.")
  }
  if (!grepl("CrossTable\\s*\\(", code_no_comments)) {
    fail("Use CrossTable() to create the cross-tab.")
  }
  if (!grepl("x\\s*=\\s*counties\\$TrumpMajority", code_no_comments)) {
    fail("TrumpMajority is our outcome variable, so it belongs in the x argument.")
  }
  if (!grepl("y\\s*=\\s*counties\\$racial_majority", code_no_comments)) {
    fail("racial_majority is our explanatory variable, so it belongs in the y argument.")
  }
  if (!grepl("prop\\.r\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.r = FALSE to suppress row proportions.")
  }
  if (!grepl("prop\\.c\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set prop.c = TRUE to show column proportions.")
  }
  if (!grepl("prop\\.t\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.t = FALSE to suppress total proportions.")
  }
  if (!grepl("prop\\.chisq\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.chisq = FALSE to suppress chi-square cell contributions.")
  }
  if (!grepl("chisq\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set chisq = TRUE to add the chi-square test of independence to the output.")
  }
  if (!grepl("expected\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set expected = TRUE to add the expected counts to the output.")
  }

  pass("Right! TrumpMajority is assigned to x and racial_majority is assigned to y, with the chi-square test and expected counts included.")
})
```

You will notice the output now has four columns instead of three, and each cell shows three numbers. The structure is the same as in the first example: observed count, expected count, column proportion. With four racial majority categories and two outcome categories, the table is larger, but the reading logic is identical.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

The p-value appears at the bottom of the CrossTable output, in the Pearson's Chi-squared section. Find it, compare it to .05, and then decide what this test --- which treats all four racial majority categories at once --- can and cannot tell you.

**Practice: Evaluate the result**

```{r quiz-racialmajority-general-result, echo=FALSE}
question("A student reads the chi-square output and says: 'p < .001, so we have strong evidence that majority-white counties are more likely to support Trump.' What is wrong with this conclusion?",
  correct = "Right. A significant chi-square only tells us some relationship exists across all four categories --- it does not identify which specific groups drive it.",
  answer("The chi-square test is significant, but because its p-value tests all four racial majority categories simultaneously, it does not directly test the specific claim that White Majority counties are more likely to support Trump",
         correct = TRUE),
  answer("Nothing is wrong --- a p-value below .001 is strong enough to support any specific claim about the variables",
         message = "A small p-value tells you that some relationship exists across the four categories, but it does not directly test which specific groups drive it. The proportions test is what evaluates the White Majority comparison specifically."),
  answer("The conclusion is wrong because the chi-square test has not yet been confirmed by a proportions test",
         message = "The chi-square result is valid on its own for the question it answers --- whether a relationship exists across all four categories. The issue is that its p-value does not directly test the more specific claim the hypothesis makes about White Majority counties."),
  answer("The conclusion is wrong because the p-value reflects sample size as much as effect size, so p < .001 could be trivial",
         message = "Sample size does affect p-values, but that is not the main problem here. The core issue is that the chi-square p-value tests all four categories simultaneously and does not directly evaluate the specific White Majority contrast."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The chi-square p-value tests all four racial majority categories at once. Does that directly test whether White Majority counties specifically stand out?"
)
```

### Testing the Directional Claim

Our hypothesis is specifically about majority-white counties compared to *all other* counties combined --- not about comparing White Majority counties to just one other category, and not about the four categories generally. This comparison group does not exist as a single column in the cross-tab --- we have to build it by adding across the three non-white-majority categories.

`prop.test()` needs four numbers: the Trump-majority count and the total for each of the two groups. Both come from your cross-tab, but from different rows:

- **x** takes the Trump-majority counts --- these come from the row where **TrumpMajority** = 1. For White Majority counties, read that cell directly: 2549. For the combined non-white-majority group, find the **TrumpMajority** = 1 cell for each of the three non-white columns and add them: 4 + 55 + 14.
- **n** takes the column totals --- these come from the Total row at the bottom. For White Majority counties, read the column total directly: 2866. For the combined group, find the Total cell for each of the three non-white columns and add them: 102 + 96 + 48.

Rather than computing the sums in your head, write the addition directly inside the `prop.test()` call --- R evaluates arithmetic expressions anywhere you can pass a number, so `4 + 55 + 14` inside `c()` works exactly like typing `73`. The calculation stays visible and checkable in your code. Pass both pairs of numbers as vectors using `c()`: `x = c(White_count, combined_count)` and `n = c(White_total, combined_total)`, with the White Majority group listed first in both.

#### Steps 1 and 2: State the Hypotheses

$H_0: \pi_{White}=\pi_{NotWhite}$, where $\pi$ is the proportion of counties with a Trump majority in the population.

$H_A: \pi_{White} > \pi_{NotWhite}$

Before building the test, identify the correct `alternative` argument.

```{r quiz-racialmajority-alternative, echo=FALSE}
question("White Majority counties are listed first in `x` and `n`. Which value of `alternative` correctly tests the directional claim?",
  correct = "Right. With White Majority listed first and the hypothesis predicting it has the higher proportion, `alternative = \"greater\"` tests that direction. `alternative = \"greater\"` tests whether the first group's proportion exceeds the second's --- exactly the direction the hypothesis predicts.",
  answer('`alternative = "greater"`',
         correct = TRUE),
  answer('`alternative = "less"`',
         message = 'This would test whether the White Majority proportion is lower than the combined-other proportion --- the opposite of what the hypothesis predicts. Check the direction again.'),
  answer('`alternative = "two.sided"`',
         message = 'A two-sided alternative tests whether the proportions differ in either direction. The hypothesis makes a specific directional prediction, so a one-tailed test is appropriate here.'),
  answer('`alternative = "greater"` would only be correct if the groups were listed in the opposite order',
         message = 'The choice of `alternative` depends on both the direction the hypothesis predicts and which group is listed first. With White Majority first and a hypothesis predicting it has the higher proportion, `alternative = "greater"` is correct as written --- it would need to change to `"less"` only if the groups were swapped.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The hypothesis predicts White Majority counties have a higher Trump-majority proportion. With that group listed first in x and n, does alternative = \"greater\" or \"less\" test that direction?"
)
```

#### Step 3: Run the Test

**Practice: Build the prop.test() call for racial majority**

Write the `prop.test()` call using inline addition for the three non-white-majority categories, with White Majority listed first. Two hints are available. Click "Submit Answer" to check your work.

```{r proptest-trumpmajority-white-other, exercise=TRUE, exercise.cap="Practice: Difference-of-proportions test for White Majority vs. other counties"}

```

```{r proptest-trumpmajority-white-other-hint-1}
# Hint 1 of 2
# prop.test() compares two proportions. Each group needs
# a success count (x) and a total (n). For the combined
# non-white-majority group, neither number appears as a
# single cell in the cross-tab --- you have to build both
# by adding across the three non-white columns.
# Which row gives you the success counts, and which gives
# you the totals?
```

```{r proptest-trumpmajority-white-other-hint-2}
# Hint 2 of 2
# x = c(White_success, non-white_success added across 3 columns)
# n = c(White_total, non-white_total added across 3 columns)
# The hypothesis predicts White Majority has the higher
# proportion --- with White Majority listed first, which
# value of alternative tests that direction?
```

```{r proptest-trumpmajority-white-other-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (is.null(.result)) {
    fail("Your code did not produce a result. Check for unmatched parentheses, values in the wrong order, or a missing argument.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the prop.test() output. Check for unmatched parentheses, values in the wrong order, or a missing argument.")
  }
  if (!grepl("prop\\.test\\s*\\(", code_no_comments)) {
    fail("Use prop.test() to run the difference-of-proportions test.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts the White Majority proportion is higher than the combined-other proportion. Set alternative = "greater" to test that specific direction.')
  }

  expected <- prop.test(x = c(2549, 73),
                        n = c(2866, 246),
                        alternative = "greater")

  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check your counts. x should give c(2549, 73) --- Trump-majority counties among White Majority, then among the other three categories combined (4 + 55 + 14). n should give c(2866, 246) --- totals for those same two groups (102 + 96 + 48 for the combined group).")
  }

  pass("Right! White Majority counties are compared to all other counties combined, with alternative = \"greater\" matching the direction the hypothesis predicts. Using inline addition inside c() keeps your calculation transparent and checkable.")
})
```

**Practice: Evaluate the result**

```{r quiz-racialmajority-directional-result, echo=FALSE}
question("Find the p-value in your output. Does this result provide statistically significant evidence in support of the hypothesis?",
  correct = "Right. The p-value is far below .05 and the direction matches the hypothesis. Trump won a majority in 88.9% of White Majority counties compared to 29.7% of all other counties combined (p < .001). The confidence interval excludes 0, consistent with the p-value. Combined with the chi-square test, this is statistically significant evidence in support of the hypothesis.",
  answer("Yes --- the p-value is far below .05, and the proportion of Trump-majority counties is higher among White Majority counties than among all other counties combined, matching the predicted direction",
         correct = TRUE),
  answer("Yes, but the result may be driven by the very small number of non-white-majority counties --- with only 246 counties in that group, the p-value could be unreliable",
         message = "246 counties is not a particularly small comparison group for this test, and both the Trump-majority and non-Trump-majority counts within that group are large enough for the large-sample approximation to be reasonable. The difference is both statistically significant and substantively large."),
  answer("No --- the p-value is below .05, so we fail to reject the null hypothesis",
         message = "This reverses the decision rule. A p-value *below* .05 leads us to *reject* the null hypothesis, not fail to reject it. We fail to reject the null only when the p-value is *above* .05."),
  answer("We cannot tell from the output which direction the difference runs --- prop.test() only reports whether the proportions differ",
         message = "The sample proportions appear in the output. The direction is visible there, and alternative = \"greater\" specifically tested whether the White Majority proportion was higher."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value. Then check whether the sample proportions in the output match the direction the hypothesis predicted: which group has the higher Trump-majority proportion?"
)
```

## Are Dictatorships Less Likely to Have High Gender Equality?

The regime type and gender equality example from Tutorial 7 involved a small dataset --- only 61 countries --- and a 3×3 table. The pattern was consistent with the hypothesis: dictatorships lagged behind democracies on high gender equality. Formally testing it introduces a practical complication that comes up often in political science: with few cases spread across many cells, some expected counts fall below 5, and the standard chi-square approximation becomes unreliable. This example shows what to do when that happens.

::: {.hypothesis}
In a comparison of [countries], those that are [dictatorships] will be less likely to [have high gender equality] than will those that are [democracies].
:::

The data frame **world** contains **gender_equal3**, a 3-category factor variable with values "Low," "Medium," and "High," and **regime_type3**, a 3-category factor variable with values "Dictatorship," "Parliamentary Democracy," and "Presidential Democracy." Note that in the actual data, the regime type labels are abbreviated --- "Parliamentary democ" and "Presidential democ" --- so those abbreviations will appear in your cross-tab output. This tutorial uses the fuller names for clarity.

As in the previous two examples, we start with the general claim the cross-tab can answer for free, then sharpen to the directional claim the hypothesis actually makes.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$H_0:$ **regime_type3** and **gender_equal3** are not related in the population.

#### Step 2: State the Alternative Hypothesis

$H_A:$ **regime_type3** and **gender_equal3** are related in the population.

#### Step 3: Run the Test

The structure is identical to the previous two examples --- only the variable names and data frame change. This time you will build the full `CrossTable()` call from scratch.

**Practice: Build the chi-square test of independence**

Use **gender_equal3** as the outcome (`x`), **regime_type3** as the explanatory variable (`y`), and the **world** data frame. Include `chisq = TRUE` and `expected = TRUE`. Two hints are available. Click Submit Answer to check your work.

```{r crosstab-genderequality-by-regimetype, exercise=TRUE, exercise.cap="Practice: Chi-square test of gender_equal3 and regime_type3", warning=TRUE}

```

```{r crosstab-genderequality-by-regimetype-hint-1}
# Hint 1 of 2
# The structure mirrors the two chi-square exercises earlier in this tutorial.
# Same function, same prop.* arguments, same chisq and expected settings ---
# adapt the data frame, x, and y to this hypothesis.
```

```{r crosstab-genderequality-by-regimetype-hint-2}
# Hint 2 of 2
# x = world$outcome.variable, y = world$explanatory.variable.
# gender_equal3 is the outcome; regime_type3 is the explanatory variable.
```

```{r crosstab-genderequality-by-regimetype-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (is.null(.result)) {
    fail("Your code did not produce output. Check for unmatched parentheses, a missing comma between arguments, or a misspelled argument name.")
  }
  if (!grepl("CrossTable\\s*\\(", code_no_comments)) {
    fail("Use CrossTable() to create the cross-tab.")
  }
  if (!grepl("x\\s*=\\s*world\\$gender_equal3", code_no_comments)) {
    fail("gender_equal3 is our outcome variable, so it belongs in the x argument.")
  }
  if (!grepl("y\\s*=\\s*world\\$regime_type3", code_no_comments)) {
    fail("regime_type3 is our explanatory variable, so it belongs in the y argument.")
  }
  if (!grepl("prop\\.r\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.r = FALSE to suppress row proportions.")
  }
  if (!grepl("prop\\.c\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set prop.c = TRUE to show column proportions.")
  }
  if (!grepl("prop\\.t\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.t = FALSE to suppress total proportions.")
  }
  if (!grepl("prop\\.chisq\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set prop.chisq = FALSE to suppress chi-square cell contributions.")
  }
  if (!grepl("chisq\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set chisq = TRUE to add the chi-square test of independence to the output.")
  }
  if (!grepl("expected\\s*=\\s*TRUE", code_no_comments)) {
    fail("Set expected = TRUE to add the expected counts to the output.")
  }

  pass("Right! gender_equal3 is assigned to x and regime_type3 is assigned to y, with chisq = TRUE and expected = TRUE added to the output.")
})
```

You should see a warning: "Chi-squared approximation may be incorrect." Two cells fall below the conventional threshold of 5 for expected counts --- **Dictatorship/Low** and **Presidential Democracy/Low**, both with an expected count of 4.7. This happens because both the Low gender equality row (only 18 countries) and the Dictatorship and Presidential Democracy columns (16 countries each) have small totals: (18)(16)/61 = 4.7. When row and column totals are both small, the expected count in that cell will be small too. With only 61 countries spread across nine cells, this is hard to avoid.

This is not a sign of a bad hypothesis or bad data --- it is just a situation where the standard tool needs a small adjustment. The fix is to ask R to calculate the p-value by simulation rather than by approximation, using `simulate.p.value = TRUE`.

**Run and observe: Re-run the test with a simulated p-value**

```{r crosstab-genderequality-by-regimetype-simulated, exercise=TRUE, exercise.cap="Run and observe: Chi-square test with simulated p-value"}
CrossTable(x = world$gender_equal3,
           y = world$regime_type3,
           prop.r = FALSE,
           prop.c = TRUE,
           prop.t = FALSE,
           prop.chisq = FALSE,
           chisq = TRUE,
           expected = TRUE,
           simulate.p.value = TRUE)
```

Notice that this version of the output does not report degrees of freedom at all --- simulation does not rely on the chi-square distribution the way the standard approximation does, so there is no df to report. The p-value is small enough to reject the null hypothesis: **regime_type3** and **gender_equal3** are related in the underlying process. As with the previous two examples, this chi-square result establishes that a relationship exists across all three categories of each variable --- it does not yet tell us whether dictatorships specifically lag behind democracies on high gender equality. We test that directional claim next.

### Testing the Directional Claim

Our hypothesis compares dictatorships to democracies generally --- not to one specific type of democracy. Just as we combined non-white-majority counties into a single group in the previous example, we combine both democratic regime types here to build a group that matches what the hypothesis actually predicts. Reading from the cross-tab above, 1 out of 16 dictatorships has high gender equality. Among the two democratic categories combined, 16 + 4 = 20 out of 29 + 16 = 45 countries have high gender equality.

#### Steps 1 and 2: State the Hypotheses

This time, identify the complete null and alternative hypothesis pair before building the test.

```{r quiz-genderequality-hypotheses, echo=FALSE}
question("Which pair of hypotheses correctly formalizes the directional claim, with dictatorships listed first?",
  correct = "Right. The null is always equal proportions stated in terms of the population parameter $\\pi$; the alternative encodes the directional prediction. $H_0: \\pi_{Dictatorship} = \\pi_{Democracy}$ and $H_A: \\pi_{Dictatorship} < \\pi_{Democracy}$, where $\\pi$ is the proportion of countries with high gender equality in the population.",
  answer("$H_0: \\pi_{Dictatorship} = \\pi_{Democracy}$ and $H_A: \\pi_{Dictatorship} < \\pi_{Democracy}$",
         correct = TRUE),
  answer("$H_0: p_{Dictatorship} = p_{Democracy}$ and $H_A: p_{Dictatorship} < p_{Democracy}$",
         message = "The direction is right, but hypothesis tests are always stated in terms of population parameters, denoted with Greek letters. $\\pi$ (pi) is the population proportion --- the quantity the hypothesis is actually about. $p$ is the sample proportion that estimates it. The null and alternative describe what we believe about the population, not what we observe in the sample."),
  answer("$H_0: \\pi_{Dictatorship} = \\pi_{Democracy}$ and $H_A: \\pi_{Dictatorship} \\neq \\pi_{Democracy}$",
         message = "The null is right, but a two-sided alternative does not capture the directional prediction the hypothesis makes. The hypothesis specifically predicts dictatorships lag behind --- that is a one-tailed claim."),
  answer("$H_0: \\pi_{Dictatorship} < \\pi_{Democracy}$ and $H_A: \\pi_{Dictatorship} = \\pi_{Democracy}$",
         message = "The null and alternative are reversed. The null hypothesis is always the no-difference baseline --- equal proportions --- and the alternative is the directional claim we are testing. A null hypothesis never contains an inequality."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: What does the null hypothesis always say about the two population proportions? Then check which direction the hypothesis predicts --- and which letter denotes a population parameter."
)
```

#### Step 3: Run the Test

The chi-square test established that a relationship exists across the three regime types. Now we test the specific directional claim: that dictatorships have a lower proportion of countries with high gender equality than democracies. Because the hypothesis compares dictatorships to all democracies, the second group requires adding parliamentary and presidential democracies together --- the same cell-aggregation logic you used in the racial majority example.

**Practice: Build the difference-of-proportions test**

Read the counts for each group from the cross-tab you just produced, combining the two democratic categories. Two hints are available. Click Submit Answer to check your work.

```{r proptest-genderequality-dictatorship-democracy, exercise=TRUE, exercise.cap="Practice: Difference-of-proportions test for dictatorships vs. all democracies"}

```

```{r proptest-genderequality-dictatorship-democracy-hint-1}
# Hint 1 of 2
# The hypothesis compares dictatorships to ALL democracies combined ---
# both parliamentary and presidential. Find the high gender equality count
# and the total for each democratic category in your cross-tab, then add
# them together to build the second group. The first group is dictatorships alone.
```

```{r proptest-genderequality-dictatorship-democracy-hint-2}
# Hint 2 of 2
# The hypothesis predicts dictatorships have a LOWER proportion
# of high gender equality than democracies combined. With
# Dictatorship listed first in x and n, alternative = "less"
# tests that direction. For the counts, x takes the high
# gender equality count for each group and n takes the total
# --- use the values from your cross-tab and the addition
# described in Hint 1.
```

```{r proptest-genderequality-dictatorship-democracy-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (is.null(.result)) {
    fail("Your code did not produce a result. Check for unmatched parentheses, values in the wrong order, or a missing argument.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the prop.test() output. Check for unmatched parentheses, values in the wrong order, or a missing argument.")
  }
  if (!grepl("prop\\.test\\s*\\(", code_no_comments)) {
    fail("Use prop.test() to run the difference-of-proportions test.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]less['\"]", code_no_comments)) {
    fail('The hypothesis predicts dictatorships have a lower proportion of high gender equality than democracies. Set alternative = "less" to test that specific direction.')
  }

  expected <- prop.test(x = c(1, 20),
                        n = c(16, 45),
                        alternative = "less")

  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check your counts. x should be c(1, 20) --- 1 dictatorship with high gender equality, and 16 + 4 = 20 democracies with high gender equality. n should be c(16, 45) --- 16 dictatorships total, and 29 + 16 = 45 democracies total.")
  }

  pass("Right! Dictatorships are compared to all democracies combined, with alternative = \"less\" matching the direction the hypothesis predicts. The sample proportions are 6.3% for dictatorships and 44.4% for all democracies combined.")
})
```

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

```{r quiz-genderequality-directional-result, echo=FALSE}
question("Find the p-value and the sample proportions in your output. Which statement best describes what this result tells us?",
  correct = "Right. The result is statistically significant, in the predicted direction, and the gap is substantively large. Only 6.3% of dictatorships have high gender equality compared to 44.4% of all democracies combined (p = .007). The confidence interval excludes 0, consistent with the p-value. Combined with the chi-square test, this is statistically significant evidence in support of the hypothesis.",
  answer("The p-value is below .05 and the proportion of countries with high gender equality is lower among dictatorships than among all democracies combined, matching the predicted direction --- and the gap between the two groups is substantively large",
         correct = TRUE),
  answer("The p-value is below .05 and the direction matches, but the gap between 6.3% and 44.4% is too small to be substantively meaningful",
         message = "A gap of over 38 percentage points between dictatorships and democracies on high gender equality is substantively large, not trivial. Statistical significance and substantive size point in the same direction here."),
  answer("No --- the p-value is below .05, so we fail to reject the null hypothesis",
         message = "This reverses the decision rule. A p-value *below* .05 leads us to *reject* the null hypothesis, not fail to reject it. We fail to reject the null only when the p-value is *above* .05."),
  answer("The p-value is below .05 but the direction is wrong --- dictatorships have a higher proportion of high gender equality than democracies",
         message = "Look at the sample proportions in your output: prop 1 (dictatorships) = 0.063 and prop 2 (democracies combined) = 0.444. Dictatorships have the lower proportion, which is exactly what the hypothesis predicted."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value and compare it to .05. Then look at the sample proportions --- which group has the higher proportion of high gender equality, and does that match the predicted direction?"
)
```

## Testing a Single Proportion

The three examples above all tested whether two variables are related --- the general-then-directional procedure we used throughout. This section introduces a separate use of `prop.test()` that is not part of that procedure: testing whether a single sample proportion differs from some fixed, hypothesized value, with no second variable and no relationship involved at all. For example, we might ask whether the proportion of Trump-majority counties in our data is consistent with a hypothesized value of 0.80.

We use the same `prop.test()` function, but with different arguments: `x` is the number of "successes," `n` is the total sample size, and `p` is the hypothesized population proportion we are comparing against --- R's name for this argument, unrelated to the p-value despite sharing a letter.

The numbers 2622 and 3112 come from the first example's cross-tab --- 2622 Trump-majority counties out of 3112 total counties. The hypothesized value of 0.80 is chosen as an example to illustrate the test.

**Run and observe: Run a one-sample proportion test**

```{r proptest-onesample-trumpmajority, exercise=TRUE, exercise.cap="Run and observe: One-sample proportion test"}
prop.test(x = 2622, n = 3112, p = 0.80)
```

The logic of the five steps still applies here: a null hypothesis ($H_0: \pi=0.80$), an alternative ($H_A: \pi \ne 0.80$), a test statistic, a p-value, and a decision --- but with only one sample proportion involved, not a comparison between two groups or a relationship between two variables.

The sample proportion, $p=0.843$ (2622/3112), is compared to the hypothesized value of 0.80. (This $p$ is the Roman-letter sample statistic from Tutorial 10's table, not the p-value reported just below --- the two are unrelated quantities that happen to share a letter. Similarly, the `p =` argument in `prop.test()` sets the hypothesized population proportion, which is also unrelated to the p-value in the output.) The test statistic is reported as a chi-squared value ($\chi^2 = 34.941$, df = 1) rather than a t- or z-statistic --- `prop.test()` uses a chi-square approximation by default, with a continuity correction applied automatically for a one-sample test --- but the conclusion follows the same logic as every other test in this tutorial. Because p < .001, we reject the null hypothesis: the proportion of Trump-majority counties in our data differs significantly from the hypothesized value of 0.80.

The output's 95% confidence interval, (0.829, 0.855), gives the same answer a different way: since it excludes the hypothesized value of 0.80 entirely, it points to the same conclusion as the p-value --- and unlike the directional tests earlier in this tutorial, this one is two-tailed, so both bounds here are calculated from the data rather than fixed at a theoretical extreme.

## Check Your Understanding

The exercises below test whether you have internalized the key ideas from this tutorial: when cell-adding is needed, what a significant chi-square result can and cannot tell you, and how to translate a hypothesis into the correct `alternative` argument.

**Apply: Choose and set up the right test**

```{r cyu-recognize-celladding, echo=FALSE}
question("A hypothesis predicts that people with a college degree are more likely to support a policy than people without one, and education is measured with a 5-category variable (less than high school, high school, some college, bachelor's degree, graduate degree). Which approach correctly tests this hypothesis?",
  correct = "Right. The hypothesis defines two groups that require adding cells --- bachelor's degree and graduate degree on one side, the remaining three categories on the other --- and predicts the college degree group has the higher proportion, so alternative = \"greater\" with the college degree group listed first.",
  answer("Add the counts for bachelor's degree and graduate degree into one group and the remaining three categories into another, then run a one-tailed difference-of-proportions test with the college degree group predicted to have the higher proportion",
         correct = TRUE),
  answer("Run a chi-square test of independence between education category and policy support",
         message = "A chi-square test would tell you whether education and policy support are related across all five categories, but its p-value does not directly test whether the college degree group specifically has a higher proportion of support than the non-college group. The hypothesis makes a specific directional comparison between two defined groups --- that requires a difference-of-proportions test."),
  answer("Run a one-tailed difference-of-proportions test comparing bachelor's degree holders to those with some college, since those are the two most comparable adjacent categories",
         message = "The hypothesis defines the comparison: college degree vs. no college degree. Choosing categories based on comparability rather than what the hypothesis predicts would change the question being asked. Both sides of the comparison require combining categories to match what the hypothesis actually predicts."),
  answer("Run a one-tailed difference-of-proportions test comparing the college degree group to the non-college group, with the non-college group predicted to have the higher proportion",
         message = "The direction is reversed. The hypothesis predicts that people with a college degree are more likely to support the policy --- so the college degree group should have the higher proportion, not the non-college group."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The hypothesis defines two groups that don't exist as single categories in the variable. It also predicts a specific direction. Both matter for choosing and setting up the right test."
)
```

**Apply: Distinguish a general result from a directional conclusion**

```{r cyu-general-vs-directional, echo=FALSE}
question("A researcher hypothesizes that military regimes are less likely than civilian regimes to hold competitive elections. A chi-square test comparing four regime categories and three election-competitiveness categories produces p < .001. Which conclusion is justified by that test alone?",
  correct = "Right. A significant chi-square establishes that some relationship exists, not which specific contrast drives it.",
  answer("That regime type and election competitiveness are related in the population, but the chi-square p-value does not directly test whether military regimes specifically have less competitive elections",
         correct = TRUE),
  answer("That military regimes have significantly less competitive elections than civilian regimes, since p < .001 is very strong evidence",
         message = "A p-value below .001 tells us the relationship is very unlikely to be due to chance, but it does not tell us which specific groups are driving it. The chi-square test considers all four regime categories at once --- the significant result could come from any combination of differences, not necessarily the military-versus-civilian contrast the hypothesis predicts."),
  answer("That the hypothesis is fully supported, since the test is highly statistically significant",
         message = "A significant chi-square result provides evidence that some relationship exists across the categories, but its p-value does not directly evaluate the specific directional contrast the hypothesis predicts. The hypothesis makes a claim about military regimes in particular --- establishing that takes a targeted test, not an omnibus one."),
  answer("That the researcher made an error, since a chi-square test cannot be used when a hypothesis predicts a specific direction",
         message = "Running a chi-square test here is not an error --- it is still useful for establishing that some relationship exists across regime types. The issue is not that the test is wrong to run, but that its p-value alone does not directly test the specific military-versus-civilian directional claim."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The chi-square test treats all four regime categories simultaneously. What does a significant result establish, and what does it leave unanswered about the specific groups the hypothesis names?"
)
```

**Apply: Predict the right direction for a new hypothesis**

```{r cyu-direction-prediction, echo=FALSE}
question('A hypothesis predicts that countries with lower levels of press freedom have a higher rate of high government corruption. You list the high-press-freedom group first in `x` and `n`. Which value of `alternative` tests the hypothesis directly?',
  correct = "Right. With the high-press-freedom group listed first and the hypothesis predicting it has less corruption, `alternative = \"less\"` is correct. The hypothesis predicts that low-press-freedom countries have a higher proportion of high corruption. With the high-press-freedom group listed first, the question becomes: is the first group's proportion lower than the second's? Yes --- the hypothesis predicts exactly that. `alternative = \"less\"` tests whether the first group's proportion is lower than the second's, which matches the hypothesis.",
  answer('`alternative = "less"`',
         correct = TRUE),
  answer('`alternative = "greater"`',
         message = 'This would test whether the high-press-freedom group\'s proportion of high corruption is greater than the low-press-freedom group\'s --- the opposite of what the hypothesis predicts. The hypothesis says low-press-freedom countries have more corruption, which means high-press-freedom countries have less. With the high-press-freedom group listed first, `alternative = "less"` is the correct direction.'),
  answer('`alternative = "two.sided"`, because the test should account for uncertainty about which direction the relationship runs',
         message = "A two-sided test is appropriate when we have no directional prediction. The hypothesis makes a specific directional prediction --- lower press freedom goes with higher corruption --- so a one-tailed test is more appropriate. The only question is which direction to specify given which group is listed first."),
  answer('`alternative = "greater"`, because the low-press-freedom countries are the ones with more corruption, and they should go first',
         message = 'The group ordering is already fixed --- the high-press-freedom group is listed first. You cannot change that; you have to choose `alternative` to match the direction given that order. With high-press-freedom first and the hypothesis predicting it has lower corruption, `alternative = "less"` is correct.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The high-press-freedom group is listed first. The hypothesis predicts that group has lower corruption than the second group. Which value of alternative tests whether the first group's proportion is lower?"
)
```

## The Takeaways

This tutorial turned the cross-tab comparisons from Tutorial 7 into formal [**hypothesis tests**]{.important-text}. Finding that column proportions differed in the direction a hypothesis predicted was a useful first look, but it could not tell us whether that difference was bigger than we would expect to see by chance alone. The two tests in this tutorial let us ask related but different questions: first whether the variables are related across all categories, and then whether a particular two-group contrast runs in the direction the hypothesis predicts.

This tutorial introduced three tests for categorical variables:

- The [**chi-square test of independence**]{.important-text} asks the general question: are these two categorical variables related at all in the population? Run it with `CrossTable()`, adding `chisq = TRUE` and `expected = TRUE` to the standard cross-tab call. The `expected = TRUE` argument adds the [**expected count**]{.important-text} --- (row total)(column total)/n --- to every cell, showing how far observed counts stray from what [**independence**]{.important-text} would predict. A chi-square test can tell you a relationship exists, but its p-value does not directly test the specific directional contrast a hypothesis predicts. You can see the direction of the observed pattern in the cross-tab column proportions; the proportions test evaluates that particular directional claim directly. When R warns that the chi-square approximation may be incorrect because some expected counts are below 5, add `simulate.p.value = TRUE` to calculate the p-value by simulation instead of the standard approximation.

- The [**difference-of-proportions test**]{.important-text} asks the sharper, directional question: is the proportion of successes higher (or lower) in one specific group than in another? Run it with `prop.test(x = c(successes1, successes2), n = c(total1, total2), alternative = ...)`. The first entries in `x` and `n` must describe the same group; so must the second entries. Use `alternative = "less"` or `"greater"` when a hypothesis predicts a specific direction. When a hypothesis compares a group to all other categories combined, read the counts from the cross-tab and add across the cells you need before passing them to `prop.test()`.

- The [**one-sample proportion test**]{.important-text} asks a different question entirely: whether a single sample proportion differs from a fixed hypothesized value, with no second variable and no relationship involved. Use the same `prop.test()` function with `p =` set to the hypothesized value instead of a second group's counts.

Our approach throughout was to begin with the general question and then sharpen the analysis to the specific directional comparison in the hypothesis. The chi-square test evaluates whether the variables are related across all categories; the proportions test directly evaluates the particular two-group directional contrast the hypothesis predicts.

The categorical variables in this tutorial were stored in three different ways: **TrumpMajority** as numeric (0 and 1), **racial_majority** as a character variable, and **gender_equal3** and **regime_type3** as factor variables. `CrossTable()` handles all three storage types without any special treatment. `prop.test()` works with counts you supply directly, so storage class does not affect it. Always check how a variable is stored before building a cross-tab or choosing an `alternative` --- factor variables sort by level order, character variables sort alphabetically, and numeric variables sort by value.

Every test follows the same [**five-step procedure**]{.important-text}: state $H_0$, state $H_A$, run the test, find the p-value, and decide whether to reject or fail to reject --- then report what that decision means substantively.

That decision is worth being precise about. Rejecting a null hypothesis --- even with a very small p-value --- never proves a hypothesis is true, and failing to reject it never proves it is false. A p-value tells us how surprising our result would be if the null were exactly correct; a small p-value is [**statistically significant evidence in support of a hypothesis**]{.important-text}, not certainty that it is right. That is not a small distinction. It is the most a hypothesis test can honestly tell us.

Tutorial 13 applies the same logic to a different combination of variable types: one categorical, one interval. The tool changes --- ANOVA for the general question, a t-test for the directional one --- but the procedure and the interpretive discipline stay exactly the same.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-12, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-12, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```


```{r tutorial-success-banner-12, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You have completed this tutorial and can now test relationships between categorical variables using chi-square tests and difference-of-proportions tests.">
  <span aria-hidden="true">🎉</span>
  Excellent work! You can now test relationships between categorical variables using chi-square tests and difference-of-proportions tests.
</p>
')
```
