---
title: "Multiple Regression"
tutorial:
  id: "16-Multiple-Regression"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to estimate and interpret multiple regression models.
---

## Overview {#main-content}

Tutorial 15 introduced linear regression in the simplest case: one explanatory variable and one outcome. But most outcomes in political and social life have many causes, not one, and a model with a single explanatory variable leaves most of that variation unexplained. Multiple regression extends the same `lm()` framework to models with more than one explanatory variable. This lets you do two things that simple regression cannot. First, you can ask whether a relationship holds up once you account for other factors that might be driving it. Second, you can model several explanatory variables at once and get a much better picture of what shapes the outcome. The mechanics are almost identical to Tutorial 15 --- the interpretation is where new thinking is required.

The data frames **counties** and **qog** each cover close to a complete population for their unit of analysis --- US counties and countries of the world, respectively. Neither is a random sample the way a public opinion survey is. What makes hypothesis tests appropriate is that our questions are not really about these specific counties and countries; they are about the political and social processes that generate outcomes like the ones we observe.

In this tutorial, you will extend your regression toolkit to models with multiple independent variables. Specifically, we will cover:

* How to estimate a multiple regression model using `lm()` by adding explanatory variables with `+` in the formula
* How to interpret coefficients in a multiple regression, including the "all else equal" qualifier that changes meaning compared to simple regression
* How to choose which variables to include and why, including the logic of controlling for competing explanations
* How to interpret the slope on a nominal variable with three or more categories, using `factor()` and `relevel()`
* How to assess model fit using two statistics: the Adjusted $R^2$ and the residual standard error, and when to prefer the Adjusted $R^2$ over the Multiple $R^2$

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
library(stargazer)
gradethis::grade_learnr,
                 exercise.completion = FALSE)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "QPATutorialsCourse")
counties$dem2p_percent <- counties$dem2p_vote_share * 100
counties$wage_growth <- counties$wage_growth * 100
counties$Black_percent <- counties$prop_black * 100
counties$white_percent <- counties$prop_white * 100
counties$hispanic_percent <- counties$prop_hispanic * 100
counties$college_grad_percent <- counties$prop_college_grad * 100
data("qog", package = "QPATutorialsCourse")
qog$democ <- dplyr::case_when(
  qog$fh_ipolity2 < 4  ~ 0,
  qog$fh_ipolity2 > 6  ~ 2,
  TRUE                  ~ 1
)
qog$democ_f <- factor(qog$democ, labels = c("Autocracy", "Partial Democracy", "Democracy"))
model_gov <- lm(iiag_gov ~ wdi_fertility + bti_eo + democ_f, data = qog)
model_gov_dem <- lm(iiag_gov ~ wdi_fertility + bti_eo + relevel(democ_f, ref = "Democracy"), data = qog)
model1 <- lm(dem2p_percent ~ mortality_risk_25_45 + wage_growth + college_grad_percent + hispanic_percent + white_percent + Black_percent, data = counties)
model_simple <- lm(dem2p_percent ~ wage_growth, data = counties)
```

## Why Control for Other Variables?

Before running a model, it is worth being clear about why multiple regression exists in the first place. There are three situations where controlling for other variables matters.

[**Spurious relationships.**]{.important-text} Sometimes a variable appears to predict an outcome, but only because a third variable drives both of them. If county-level health outcomes are poor in places where wages are low, and poor health also makes people more likely to vote against the incumbent, then part of what looks like a "wage growth effect" on the vote might really be a "health effect" in disguise. Including a health measure in the model lets us see the wage growth effect after the health effect has been taken into account --- and if the wage growth coefficient shrinks or disappears, we have learned something important about what was really going on. When a third variable drives both the explanatory variable and the outcome, making their relationship appear stronger or weaker than it really is, that variable is called a [**confounder**]{.important-text} --- and the distortion it produces is called [**confounding**]{.important-text}.

[**Multiple correlated causes.**]{.important-text} Most outcomes have several causes that are correlated with each other. Counties with higher college-educated populations tend to have higher wage growth and lower mortality risk. If we omit any one of these from the model, the coefficient on the others will be off --- each included variable will pick up some of the effect that really belongs to the omitted one. This is called [**omitted variable bias**]{.important-text}, and it is the main reason you will almost never see a published regression with only one independent variable.

[**Conditional relationships.**]{.important-text} Sometimes the effect of one variable depends on the level of another. We will address this case in Tutorial 17.

The remedy in the first two cases is the same: include the relevant variables in the model. Once you do, each coefficient is estimated holding the other variables constant. That is the meaning of the phrase you will see throughout this tutorial: [**controlling for all other variables in the model**]{.important-text}, or equivalently, [**all else equal**]{.important-text}.

::: {.caution}
**Controlling for variables reduces confounding but does not eliminate it.** Multiple regression estimates conditional associations --- the relationship between an explanatory variable and an outcome after holding other variables in the model constant. This is more informative than a simple bivariate relationship, but it is not the same as a causal effect. Variables not included in the model may still confound the relationships we estimate, and observational data cannot rule that out entirely. Establishing causation requires either a research design that rules out confounding by construction --- a randomized experiment, a natural experiment, an instrumental variable --- or a theoretical argument strong enough to defend the assumption that no important confounders remain. Regression alone cannot do that work. Throughout this tutorial, we describe our estimates as relationships and associations, not causes.
:::

## Does Wage Growth Predict Clinton's Vote, Controlling for Other County Characteristics?

A simple regression of Clinton's two-party vote percent on wage growth finds a positive relationship --- counties where wages grew faster gave her more of the vote, consistent with retrospective voting theory. But counties with higher wage growth also tend to have higher education levels, lower mortality risk, and different racial compositions. These are plausible confounders: variables correlated with both wage growth and the vote that could make the wage growth effect look larger than it really is. This example tests whether the relationship holds up once those confounders are controlled for.

::: {.hypothesis}
In a comparison of counties, those with higher wage growth will cast a higher percentage of the two-party vote for Hillary Clinton than those with lower wage growth.
:::

The data frame **counties** contains **wage_growth**, the percent by which wages grew or fell in a county, and **dem2p_percent**, the percent of the two-party vote cast for Hillary Clinton. Both variables are already measured in percentage points, so a one-unit increase in either is a one-percentage-point increase --- wage growth moving from 4% to 5%, for example --- not a one-percent relative increase on top of whatever the value already was. These sound similar but mean very different things, and the distinction matters when you interpret the slope.

The other variables we include as controls are **mortality_risk_25_45**, **college_grad_percent**, **hispanic_percent**, **white_percent**, and **Black_percent**. Each was chosen because it is plausibly correlated with both wage growth and the vote --- meeting the definition of a confounder introduced in the "Why Control for Other Variables?" section earlier in this tutorial.

### Estimate the Regression

Before estimating the full model, run the model below with wage growth alone. Note the slope on **wage_growth**, the intercept, and the $R^2$ --- if the confounders are doing real work, all three will change when we add them.

**Run and observe: The simple regression baseline**

```{r exercise-simple-baseline, exercise=TRUE, exercise.cap="Run and observe: The simple regression baseline"}
model_simple <- lm(dem2p_percent ~ wage_growth, data = counties)
summary(model_simple)
```

The output shows a slope of 0.456 on **wage_growth** --- but this estimate is almost certainly biased. Because wage growth is correlated with education, mortality risk, and racial composition, the slope is picking up the effects of those variables too, not just the effect of wage growth itself. The $R^2$ of about 0.017 confirms how little of the vote's variation this single variable captures. Adding the confounders will give us a cleaner estimate of wage growth's actual effect and should substantially increase the share of variation in Clinton's vote that the model explains.

Now add the controls. The `lm()` formula extends naturally to multiple explanatory variables: where you previously wrote `outcome ~ explanatory`, you now write `outcome ~ explanatory1 + explanatory2 + explanatory3`, listing each variable separated by a `+`. Everything else stays the same: the `data =` argument, the assignment to a named object, and the call to `summary()`. The formula is written across multiple lines --- R allows you to break a call anywhere after a comma or `+` operator, and continue on the next line. This keeps long formulas readable without horizontal scrolling, and is good practice whenever a formula has more than two or three variables.

**Run and observe: Regress Clinton's two-party vote percent on wage growth and county characteristics**

```{r exercise-multipleregression-clintonvote, exercise=TRUE, exercise.cap="Run and observe: Regress Clinton's two-party vote percent on wage growth and county characteristics"}
model1 <- lm(dem2p_percent ~ mortality_risk_25_45 + wage_growth +
               college_grad_percent + hispanic_percent +
               white_percent + Black_percent,
             data = counties)
summary(model1)
```

Take a moment to look at the output before moving on. Find the `Coefficients` table and locate each variable by name. Notice that the output structure is identical to simple regression --- there is one row per variable plus a row for the intercept, each with an estimate, standard error, t-statistic, and p-value. The only thing that has changed is the number of rows.


### Interpret the Coefficients

The key interpretive rule is the same as in simple regression: the slope is the expected change in the dependent variable for a one-unit increase in the independent variable. The crucial difference in multiple regression is what comes after: [**controlling for all other variables in the model**]{.important-text}. Every coefficient is now a partial effect --- the effect of that one variable with everything else held constant.

That qualifier matters more than it might seem. Without controls, the simple regression of Clinton's two-party vote percent on **wage_growth** alone yields a larger slope. In this model, controlling for mortality risk, education, and racial composition, the slope on **wage_growth** is smaller --- because some of what looked like a "wage growth effect" in the simple model was really a "college education effect" or "racial composition effect" in disguise. The Coefficients table now lets us untangle those contributions. Find the slope on **wage_growth** and see for yourself what controlling for the other variables does to it.

**Practice: Interpret the slope on wage_growth**

Look at your Coefficients table and note that each coefficient is now a partial slope before answering.

```{r quiz-interpret-wagegrowth, echo=FALSE}
question("Look at the slope on wage_growth in your output. Which statement correctly gives its value and what it means?",
  correct = "Right: 0.314 is the expected change in Clinton's percent of the two-party vote for a one-percentage-point increase in wage growth, holding all other variables constant.",
  answer("The slope is 0.314; for every one-percentage-point increase in wage growth, Clinton's percent of the two-party vote is expected to increase by about 0.31 percentage points, controlling for mortality risk, education, and racial composition",
         correct = TRUE,
         message = "The coefficient on wage growth is smaller than it appeared without controls. Some of what looks like a wage growth association is actually driven by correlated county characteristics --- education levels, racial composition, mortality risk --- that also shape the vote."),
  answer("The slope is 0.039; for every one-percentage-point increase in wage growth, Clinton's percent of the two-party vote is expected to increase by about 0.04 percentage points, controlling for other variables",
         message = "0.039 is the standard error of the slope, not the slope itself. The Estimate column gives the slope; the Std. Error column measures the uncertainty around that estimate. Always read the coefficient from the Estimate column."),
  answer("The slope is 0.314; for every one-percentage-point increase in Clinton's percent of the two-party vote, wage growth is expected to increase by about 0.31 percentage points, controlling for other variables",
         message = "This reverses the dependent and independent variables. The slope tells us the expected change in dem2p_percent for a one-unit change in wage_growth, not the other way around."),
  answer("The slope is 0.314; this is the expected percent of the two-party vote for Hillary Clinton in a county with zero wage growth, holding all other variables constant",
         message = "That would be the intercept's job, not the slope's. The slope gives the expected change in the outcome for a one-unit change in the predictor, not the expected level of the outcome at any particular value."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the slope gives the expected change in the outcome for a one-unit increase in the predictor, holding all other variables constant. Look at the wage_growth row in the Coefficients table."
) 
```

Now consider **mortality_risk_25_45**, the probability of dying between ages 25 and 45, expressed as a percent of the population in that age range.

**Practice: Interpret the slope on mortality_risk_25_45**

Look at your Coefficients table before answering.

```{r quiz-interpret-mortality, echo=FALSE}
question("Which statement correctly gives the value and meaning of the slope on mortality_risk_25_45?",
  correct = "Right: -3.117 is the expected change in Clinton's percent of the two-party vote for a one-percentage-point increase in mortality risk, holding all other variables constant.",
  answer("The slope is -3.117; for every one-percentage-point increase in mortality risk, Clinton's percent of the two-party vote is expected to decrease by about 3.12 percentage points, controlling for all other variables in the model",
         correct = TRUE,
         message = "The negative sign matches the hypothesis: counties where more young adults are dying show lower support for the incumbent party, even after controlling for wage growth, education, and racial composition."),
  answer("The slope is -3.117; for every one-percentage-point increase in Clinton's percent of the two-party vote, mortality risk is expected to decrease by about 3.12 percentage points, controlling for all other variables",
         message = "This reverses the dependent and independent variables. The slope tells us the expected change in dem2p_percent for a change in mortality_risk_25_45, not the reverse."),
  answer("The slope is -3.117; this is the expected percent of the two-party vote for Hillary Clinton in a county with zero mortality risk, holding all other variables constant",
         message = "That would be the intercept's job, not the slope's. The slope gives the expected change in the outcome for a one-unit change in the predictor, not the expected level of the outcome at any particular value."),
  answer("The slope is 0.302; for every one-percentage-point increase in mortality risk, Clinton's percent of the two-party vote is expected to decrease by about 0.30 percentage points, controlling for all other variables",
         message = "0.302 is the standard error of the slope, not the slope itself. The Estimate column gives the slope; the Std. Error column measures the uncertainty around that estimate."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the Estimate column in the mortality_risk_25_45 row. The slope gives the expected change in Clinton's percent of the two-party vote for a one-percentage-point increase in mortality risk, controlling for all other variables."
)
```

Now consider **college_grad_percent**, the percent of county residents with a college degree.

**Practice: Interpret the slope on college_grad_percent**

Look at your Coefficients table before answering.

```{r quiz-interpret-college, echo=FALSE}
question("Which statement correctly gives the value and meaning of the slope on college_grad_percent?",
  correct = "Right. 0.646 is the expected change in Clinton's two-party vote for each one-percentage-point increase in the county's college graduation rate, holding all other variables constant.",
  answer("The slope is 0.646; for every one-percentage-point increase in the county's college graduation rate, Clinton's percent of the two-party vote is expected to increase by about 0.65 percentage points, controlling for all other variables in the model",
         correct = TRUE,
         message = "Right. Each partial slope tells you the expected change in the outcome for a one-unit increase in that variable, holding everything else constant. With a positive slope, counties with higher college graduation rates gave Clinton a higher share of the two-party vote, controlling for mortality risk, wage growth, and racial composition."),
  answer("The slope is 0.646; for every one-percentage-point increase in Clinton's percent of the two-party vote, the county's college graduation rate is expected to increase by about 0.65 percentage points, controlling for all other variables",
         message = "This reverses the dependent and independent variables. The slope tells us the expected change in dem2p_percent for a change in college_grad_percent, not the reverse."),
  answer("The slope is 24.517; for every one-percentage-point increase in the county's college graduation rate, Clinton's percent of the two-party vote is expected to increase by about 24.5 percentage points, controlling for all other variables",
         message = "24.517 is the t-statistic, not the coefficient estimate. Look at the Estimate column, not the t value column."),
  answer("The slope is 0.646; for every one-percentage-point increase in the county's college graduation rate, Clinton's percent of the two-party vote is expected to increase by about 0.65 percentage points, regardless of other county characteristics",
         message = "The 'controlling for all other variables' qualifier is essential to the interpretation of a partial slope. Omitting it implies the slope describes a marginal effect without any controls, which is what a simple regression coefficient would give --- not a multiple regression coefficient."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the Estimate column in the college_grad_percent row. The slope gives the expected change in Clinton's percent of the two-party vote for a one-percentage-point increase in college graduates, controlling for all other variables."
)
```

The remaining variables --- **hispanic_percent**, **white_percent**, and **Black_percent** --- follow the same interpretive logic. Each slope gives the expected change in Clinton's percent of the two-party vote for a one-percentage-point increase in that group's share of the county population, holding all other variables constant.

### How Large Are the Effects?

Reading a slope tells you the expected change in the outcome for a one-unit increase in that variable. But whether that change is large or small in practice depends on the scale of the variable --- specifically, what a one-unit change actually represents relative to the variation we observe in the data. A slope of 0.314 could be enormous or trivial depending on whether typical variation in that variable spans 1 unit or 100 units.

The standard deviation is a natural anchor for gauging effect size because it summarizes what a typical county-to-county difference looks like. A one-standard-deviation shift is common but not extreme. For any change of size $c$, the expected change in the outcome is $c * \hat{\beta}$ --- the same multiplication you used to compute predicted values in Tutorial 15. This gives you a general recipe to follow whenever you want to assess the practical significance of regression coefficients: calculate the standard deviation of each explanatory variable, choose $c$ as one standard deviation (or a round number close to it), and compute $c * \hat{\beta}$ for each variable. Run the code below to get the standard deviations of the three focal variables.

**Run and observe: Standard deviations of the three focal variables**

```{r exercise-sds-clintonvote, exercise=TRUE, exercise.cap="Run and observe: Standard deviations of the three focal variables"}
sd(counties$wage_growth, na.rm = TRUE)
sd(counties$mortality_risk_25_45, na.rm = TRUE)
sd(counties$college_grad_percent, na.rm = TRUE)
```

The output shows standard deviations of about 4.7, 0.9, and 9.1 percentage points respectively. For **wage_growth**, one standard deviation is about 5 percentage points. With a slope of 0.314, the expected change in Clinton's vote for a one-sd shift in wage growth is $5 * 0.314 = 1.57$ percentage points --- a modest effect. Now apply the same recipe to the other two variables.

**Practice: Apply a meaningful unit change to mortality risk**

The standard deviation of **mortality_risk_25_45** is about 0.9 percentage points. Find the slope in your output and apply the same recipe before answering.

```{r quiz-predict-mortality, echo=FALSE}
question("What is the expected change in Clinton's percent of the two-party vote for a 0.9 percentage point increase in mortality risk --- one standard deviation --- controlling for all other variables?",
  correct = "Right: 0.9 * (-3.117) = -2.805.",
  answer("A decrease of about 2.81 percentage points",
         correct = TRUE,
         message = "A one-standard-deviation increase in mortality risk is associated with a decrease of about 2.8 percentage points in Clinton's two-party vote, controlling for wage growth, education, and racial composition --- a meaningfully larger effect than the wage growth result."),
  answer("A decrease of about 3.12 percentage points",
         message = "That is the slope itself --- the expected change for a full one-percentage-point increase, which exceeds one standard deviation. Multiply by the standard deviation: 0.9 * (-3.117) = -2.805."),
  answer("An increase of about 2.81 percentage points",
         message = "The slope on mortality_risk_25_45 is negative, so higher mortality risk is associated with lower Clinton two-party vote percent, not higher."),
  answer("A decrease of about 1.56 percentage points",
         message = "Check your value of c. The standard deviation is 0.9, not 0.5: 0.9 * (-3.117) = -2.805."),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

**Practice: Apply a meaningful unit change to college attainment**

The standard deviation of **college_grad_percent** is about 9 percentage points. A round increment of 10 is close to one standard deviation and easy to interpret. Find the slope in your output and apply the same recipe before answering.

```{r quiz-predict-college, echo=FALSE}
question("What is the expected change in Clinton's percent of the two-party vote for a 10 percentage point increase in the county's college graduation rate, controlling for all other variables?",
  correct = "Right: 10 * 0.646 = 6.46.",
  answer("An increase of about 6.46 percentage points",
         correct = TRUE,
         message = "A 10 percentage point increase in the college graduation rate --- close to one standard deviation --- is associated with an increase of about 6.5 percentage points in Clinton's two-party vote, controlling for everything else. That is roughly four times the effect of a one-standard-deviation shift in wage growth, and about twice the effect of a one-standard-deviation shift in mortality risk."),
  answer("An increase of about 0.65 percentage points",
         message = "That is the slope itself --- the expected change for a one-percentage-point increase. Multiply by c = 10: 10 * 0.646 = 6.46."),
  answer("An increase of about 64.6 percentage points",
         message = "Check your multiplication: 10 * 0.646 = 6.46, not 64.6. The slope is 0.646, not 6.46."),
  answer("A decrease of about 6.46 percentage points",
         message = "The slope on college_grad_percent is positive, so higher college graduation rates are associated with higher Clinton two-party vote percent, not lower."),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

Comparing across all three: a one-standard-deviation shift in college attainment predicts about four times the change in Clinton's vote as a one-standard-deviation shift in wage growth, and about twice the change as a one-standard-deviation shift in mortality risk. Education is doing most of the work in this model --- a finding that emerges clearly only when you scale each effect to a comparable unit of variation.

### Conduct Hypothesis Tests

The coefficients tell us the direction and size of each effect; hypothesis tests tell us whether those effects are statistically distinguishable from zero in the population. For each coefficient, we test $H_0: \beta = 0$ (the variable has no effect in the population, holding all other variables constant) against $H_A: \beta \ne 0$. The model output is reproduced here so you do not need to scroll back.

```{r model1-output-reference-tests, echo=FALSE, comment=NA}
summary(model1)
```

**Practice: Test and interpret the wage_growth coefficient**

Find the t-statistic and p-value for the slope on **wage_growth** in your output before answering.

```{r quiz-hypothesistesting-clintonvote, echo=FALSE}
question("Which statements are true? Select all that apply.",
  correct = "Right. The evidence is consistent: we reject the null hypothesis that wage growth has no relationship with Clinton's percent of the two-party vote, controlling for the other county characteristics.",
  answer("The p-value is less than .05, so we reject the null hypothesis --- wage growth has a significant positive relationship with Clinton's percent of the two-party vote, controlling for the other county characteristics",
         correct = TRUE,
         message = "The data provide strong evidence against a zero relationship between wage growth and Clinton's two-party vote percent, even after accounting for correlated county characteristics."),
  answer("The absolute value of the t-statistic exceeds 1.96, which leads to the same conclusion as the p-value",
         correct = TRUE,
         message = "The t-statistic and p-value always agree for the same two-tailed test, since both are built from the same estimate and standard error."),
  answer("The significant result proves that wage growth caused counties to shift toward Clinton in 2016",
         message = "A significant coefficient tells us the relationship is unlikely to be due to chance. It does not establish causation --- counties with higher wage growth may also differ in other unmeasured ways that drive the vote."),
  answer("A p-value below .05 means there is less than a 5% probability that the null hypothesis is true",
         message = "The p-value gives the probability of observing a result this extreme *if the null were true*, not the probability that the null is true. A small p-value means the data would be surprising under the null --- not that the null is probably false."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

**Practice: Test the slope on mortality_risk_25_45**

Find the t-statistic and p-value for mortality risk in your output before answering.

```{r quiz-hypothesistesting-mortality, echo=FALSE}
question("What can you conclude from the hypothesis test on mortality_risk_25_45?",
  correct = "Right. We reject the null hypothesis --- counties with higher mortality risk show significantly lower support for Clinton, controlling for the other variables.",
  answer("The p-value is far below .05, so we reject the null hypothesis that mortality risk has no relationship with Clinton's percent of the two-party vote, controlling for the other county characteristics",
         correct = TRUE,
         message = "The negative slope (-3.117) is large in practical terms and highly significant. Even after controlling for wage growth, education, and racial composition, counties where more young adults are dying show substantially lower support for Clinton."),
  answer("We fail to reject the null hypothesis because the slope is negative",
         message = "The sign of the coefficient has nothing to do with statistical significance. What matters is the coefficient relative to its standard error --- a large negative coefficient can be just as significant as a large positive one."),
  answer("The p-value is far below .05, so we fail to reject the null hypothesis",
         message = "It is the other way around: a p-value below .05 is evidence against the null, which leads us to reject it. We fail to reject the null only when the p-value is above .05 --- meaning the data would not be surprising if the null were true."),
  answer("A significant result for mortality risk means wage growth is no longer a statistically significant predictor in this model",
         message = "Each coefficient has its own separate hypothesis test. A significant result for mortality risk tells us that variable has a relationship with the outcome --- it says nothing about whether other predictors are significant. Look at the p-value for wage_growth separately."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find the t-statistic and p-value for mortality_risk_25_45. Apply the same decision rule as always: if p < .05, reject the null."
)
```

**Practice: Test the hypothesis for college_grad_percent**

Find the t-statistic and p-value for college_grad_percent in your output before answering.

```{r quiz-test-college, echo=FALSE}
question("What can you conclude from the hypothesis test on college_grad_percent?",
  correct = "Right: a large t-statistic and a very small p-value lead to rejecting the null hypothesis.",
  answer("The p-value is far below .05 and the t-statistic is large, so we reject the null hypothesis that college attainment has no relationship with Clinton's percent of the two-party vote, controlling for the other county characteristics",
         correct = TRUE,
         message = "With over 3,100 counties, we have a very precise estimate of this effect. College attainment is one of the strongest predictors in the model --- both statistically and in practical terms."),
  answer("We fail to reject the null hypothesis because the college_grad_percent coefficient is smaller than 1",
         message = "The size of the coefficient alone does not determine statistical significance. What matters is the coefficient relative to its standard error --- a small coefficient with a tiny standard error can be highly significant."),
  answer("The p-value is far below .05, but since the t-statistic exceeds 1.96, the two tests disagree and we cannot draw a conclusion",
         message = "The t-statistic and p-value do not disagree --- they are two ways of expressing the same result. A t-statistic exceeding 1.96 in absolute value corresponds to a p-value below .05 for a two-tailed test. They will always point to the same conclusion."),
  answer("We reject the null hypothesis, which proves that higher college attainment caused counties to vote for Clinton",
         message = "Rejecting the null tells us the relationship is unlikely to be due to chance. It does not establish causation --- counties with more college-educated residents may differ in many other ways that also predict the vote."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find the p-value for college_grad_percent and apply the decision rule: if p < .05, reject the null."
)
```

### Assess Model Fit

The individual coefficients tell us about each predictor's conditional relationship with the outcome. But how much of the county-to-county variation in Clinton's vote does this six-variable model actually account for --- and how does that compare to the near-zero fit we saw in Tutorial 15 with wage growth alone? The model output is reproduced here so you do not need to scroll back.

```{r model1-output-reference, echo=FALSE, comment=NA}
summary(model1)
```

The model output reports two versions of $R^2$. The [**Multiple R²**]{.important-text} has the cleanest interpretation: it is the proportion of the variation in the outcome that the model explains, expressed as a percentage. An $R^2$ of 0.607 means the model explains about 60.7% of the county-to-county variation in Clinton's two-party vote --- the remaining 39.3% is variation the model does not capture.

The [**Adjusted R²**]{.important-text} is a corrected version that penalizes for the number of predictors. Every variable you add to a regression model, even a useless one, will mechanically increase the Multiple $R^2$ at least a tiny bit, just because the model has more flexibility to fit the data. The Adjusted $R^2$ subtracts a penalty that grows with the number of predictors, making it less likely to increase unless an added variable genuinely improves fit. In this model the two values are nearly identical --- 0.607 and 0.606 --- because with over 3,100 counties and only six predictors, the penalty is tiny.

For describing how well a model fits, use the Multiple $R^2$ --- it gives the clean "percentage of variation explained" interpretation. Reserve the Adjusted $R^2$ for one specific purpose: comparing models that differ in the number of predictors, where the Multiple $R^2$ would mechanically favor the model with more variables regardless of whether they are genuinely useful. In this tutorial, we use Multiple $R^2$ when describing fit and Adjusted $R^2$ only when comparing models directly.

The [**residual standard error**]{.important-text} complements $R^2$ by giving the estimated standard deviation of the residuals --- the typical distance between observed outcomes and the fitted regression line. Under roughly normal errors, about two-thirds of cases have an observed value within one residual standard error of their fitted value.

**Practice: Assess the fit of the model**

Find the Adjusted R-squared and residual standard error in your output before answering.

```{r quiz-modelfit-clintonvote, echo=FALSE}
question("Which statement correctly describes what the model fit statistics tell us?",
  correct = "Right. The Multiple R-squared gives the percentage of variation explained, and the residual standard error gives the estimated standard deviation of the residuals.",
  answer("The Multiple R-squared of 0.6066 tells us the model explains about 60.66% of the county-to-county variation in Clinton's two-party vote, and under roughly normal errors, about two-thirds of counties have a predicted vote share within about 10.13 percentage points of their actual value",
         correct = TRUE,
         message = "Compare this to the simple regression of Clinton's vote on wage growth alone, which explained only about 1.7% of the variation. Adding the other county characteristics dramatically improves the model's fit."),
  answer("The model explains about 60.66% of the county-to-county variation in Clinton's two-party vote, and our predictions are off by exactly 10.13 percentage points for every county",
         message = "The residual standard error is not a fixed error for every prediction. It is the standard deviation of the residuals --- under roughly normal errors, it tells us that about two-thirds of predictions are within about 10 points, not all of them."),
  answer("The Adjusted R-squared of 0.6058 tells us the model explains about 60.58% of the variation in Clinton's two-party vote",
         message = "The Adjusted R-squared penalizes for the number of predictors and does not have the clean 'percentage of variation explained' interpretation. For that, use the Multiple R-squared of 0.6066."),
  answer("The Multiple R-squared of 0.6066 tells us that wage growth alone explains about 60.66% of the variation in Clinton's two-party vote",
         message = "This model has six explanatory variables, not just wage growth. The Multiple R-squared reflects the combined explanatory power of all six variables together."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: which statistic gives the clean 'percentage of variation explained' interpretation, and what does the residual standard error tell you?"
)
```

### What We Learned About Wage Growth and Clinton's Vote

So does wage growth predict Clinton's vote once we account for everything else? Yes --- but the full picture is more nuanced than the simple regression suggested.

This example produced four findings worth carrying forward. First, wage growth has a statistically significant positive relationship with Clinton's percent of the two-party vote --- counties where wages grew faster gave her more of the vote, consistent with retrospective voting theory. Second, that relationship is smaller than simple regression would suggest: controlling for education, racial composition, and mortality risk, the wage growth slope shrank from 0.456 to 0.314, indicating that part of the original bivariate relationship overlapped with differences in those county characteristics. Third, scaling each coefficient to one standard deviation reveals the relative importance of the predictors: a one-sd shift in college attainment predicts about 6.5 percentage points in Clinton's vote, compared to about 2.8 points for mortality risk and 1.6 points for wage growth --- education is doing most of the work. Fourth, the Multiple $R^2$ of 0.607 tells us the model accounts for about 60.7% of the county-to-county variation in Clinton's vote, a dramatic improvement over the near-zero fit of the simple wage growth model; the residual standard error of 10.13 percentage points gives the estimated standard deviation of the residuals, with about two-thirds of counties having observed vote shares within about 10 points of their fitted values. Together, these findings suggest that Clinton's 2016 performance was shaped as much by the long-run demographic and educational character of counties as by the economic conditions of the preceding years.

## Does Democracy Predict the Quality of Governance?

Let's consider a second example using a different unit of analysis: countries. Does the level of democracy in a country predict how well its government performs overall? The theoretical argument runs through accountability: democratic institutions create electoral incentives for governments to deliver public goods, civil society organizations that monitor performance, and independent courts that constrain abuses of power. All of these channels suggest that more democratic countries should, on average, be better governed --- though we should also consider whether the relationship is linear across the full spectrum from autocracy to full democracy, or whether the gains are concentrated at the top.

::: {.hypothesis}
In a comparison of countries, those with higher levels of democratic governance will have higher quality of governance than those with lower levels of democratic governance, after controlling for economic equality of opportunity and the fertility rate.
:::

The data frame **qog** contains country-level data from multiple sources. Our dependent variable is **iiag_gov**, an overall governance quality score from the Ibrahim Index extended to all countries --- it ranges from about 11 to 80, where higher values indicate better governance across dimensions including rule of law, human rights, sustainable economic opportunity, and human development. Our independent variables are:

- **wdi_fertility**: the total fertility rate (average number of children per woman)
- **bti_eo**: equal opportunity, scored from 1 (least equal) to 10 (most equal)
- **democ_f**: a three-category factor in which **Autocracy** = lowest democracy level, **Partial Democracy** = middle, and **Democracy** = highest

### Choosing a Reference Category

Before estimating a model with a categorical predictor, it is worth thinking about which category to use as the reference. The choice matters because every factor coefficient measures the difference between one category and the reference --- so the question you can answer directly depends on which category you omit. If Democracy is the reference, you can compare both Autocracy and Partial Democracy to Democracy, but you cannot directly test whether Autocracy and Partial Democracy differ from each other. If Autocracy is the reference, you get both comparisons to the lowest level --- which is usually what we want when asking whether more democracy produces better outcomes. The variable **democ_f** has already been created with three labels. Let's check which category R is treating as the reference.

**Run and observe: Check the levels of democ_f**

```{r exercise-levels-democ, exercise=TRUE, exercise.cap="Run and observe: Check the levels of democ_f"}
levels(qog$democ_f)
```

The first level listed is the reference --- the category that will be omitted from the regression output, against which the other categories are compared. Here, **Autocracy** comes first, so it will serve as the reference. That makes sense for this question: we want to ask how much better partial democracies and full democracies are governed compared to the baseline of autocracy.

But you should always check, and you should think about whether the default serves your question. To see why the choice matters, let's first run the model with **Democracy** as the reference instead --- so we can ask how much worse autocracies and partial democracies are compared to the democratic baseline.

To do this, we need to tell `lm()` to use Democracy as the reference category for this model only. Instead of writing just `democ_f` in the formula, we write `relevel(democ_f, ref = "Democracy")` in its place. The `relevel()` function takes two arguments: the factor variable and `ref =` naming the new reference category. The change is temporary --- it applies only to this model and does not permanently alter the variable.

**Run and observe: Estimate the governance model with Democracy as the reference category**

```{r exercise-multipleregression-governance-dem, exercise=TRUE, exercise.cap="Run and observe: Estimate the governance model with Democracy as the reference category"}
model_gov_dem <- lm(iiag_gov ~ wdi_fertility + bti_eo +
                      relevel(democ_f, ref = "Democracy"),
                    data = qog)
summary(model_gov_dem)
```

Look at the two factor coefficients. With **Democracy** as the reference, both **Partial Democracy** (−10.556, p < .001) and **Autocracy** (−11.905, p < .001) appear significantly worse --- and at first glance, they look almost equally bad. Now let's run the model with **Autocracy** as the reference and see if that impression holds.

### Estimate the Regression

Now let's estimate the model with **Autocracy** as the reference --- which is already the default, so we write `democ_f` in the formula without any `relevel()` call. This lets us read off directly how partial democracies and full democracies compare to the autocratic baseline.

**Practice: Estimate the governance model with Autocracy as the reference category**

Fill in the blanks to estimate the model, assigning it to **model_gov**, then display the results. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-multipleregression-governance, exercise=TRUE, exercise.cap="Practice: Estimate the governance model with Autocracy as the reference category"}
model_gov <- lm(___ ~ ___ + ___ + ___, data = ___)
summary(model_gov)
```

```{r exercise-multipleregression-governance-hint-1}
# Hint 1 of 2
# Look back at the hypothesis. Which variable is the outcome?
# Which variables do we think drive it? In a regression formula,
# the outcome goes on the left of ~ and the predictors go on the
# right, separated by +.
```

```{r exercise-multipleregression-governance-hint-2}
# Hint 2 of 2
# The outcome is iiag_gov and the data frame is qog.
# You have two continuous variables and one factor variable
# as explanatory variables. democ_f is already in the environment.
# Assign your model to model_gov, then call summary() on it.
```

```{r exercise-multipleregression-governance-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use lm() to estimate the regression.")
  }
  if (!grepl("model_gov\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model_gov.")
  }
  if (!grepl("iiag_gov\\s*~", code_no_comments)) {
    fail("The dependent variable should be iiag_gov --- governance quality is what we are trying to explain.")
  }
  if (!grepl("wdi_fertility", code_no_comments)) {
    fail("Include wdi_fertility as an independent variable.")
  }
  if (!grepl("bti_eo", code_no_comments)) {
    fail("Include bti_eo as an independent variable.")
  }
  if (!grepl("democ_f", code_no_comments)) {
    fail("Include democ_f as an independent variable.")
  }
  if (!grepl("data\\s*=\\s*qog", code_no_comments)) {
    fail("Set data = qog so R knows where to find the variables.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call summary() on your saved model object to display the results.")
  }
  expected <- lm(iiag_gov ~ wdi_fertility + bti_eo + democ_f, data = qog)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling summary() on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check that your model includes iiag_gov as the outcome and wdi_fertility, bti_eo, and democ_f as predictors, using the qog data.")
  }
  pass("Correct! A side-by-side comparison of both models appears in the next section --- look at the factor coefficient rows to see what changes and what stays the same.")
})
```

### Interpret the Coefficients

The table below places both models side by side so you can compare them directly. The key question is what changes when we switch the reference category from Democracy to Autocracy --- and what that change reveals about the relationship between regime type and governance quality. You will learn to produce tables like this with the `stargazer()` function later in this tutorial --- for now, use it as a reference for the questions that follow.

```{r governance-comparison, echo=FALSE}
stargazer(model_gov_dem, model_gov,
          type = "text",
          title = "Governance Quality Under Two Reference Categories",
          column.labels = c("(1) Demo. ref.", "(2) Auto. ref."),
          dep.var.labels = "Governance Quality",
          covariate.labels = c("Fertility Rate",
                               "Equal Opportunity",
                               "Partial Democracy",
                               "Autocracy",
                               "Partial Democracy",
                               "Democracy"),
          no.space = TRUE,
          omit.stat = c("f", "ser"))
```

Look carefully at both columns. You can verify directly from the table which rows are identical across the two models and which differ. Look at the factor coefficients and the constants before answering.

**Practice: Interpret the factor coefficients**

Look at the factor coefficient rows in both columns of the table above before answering.

```{r quiz-interpret-democ-governance, echo=FALSE}
question("Look at both columns in the table above. Which statements are true? Select all that apply.",
  correct = "Right: three statements correctly describe what the table shows.",
  answer("In column (2) (Autocracy as reference), full democracies score about 11.9 governance points higher than autocracies on average, controlling for fertility and equal opportunity",
         correct = TRUE,
         message = "Column (2) shows the Democracy vs. Autocracy coefficient as 11.905 (p < .001). This is the comparison we wanted: how much better governed are full democracies relative to the autocratic baseline?"),
  answer("In column (2) (Autocracy as reference), the Partial Democracy coefficient is about 1.35 points and is not statistically significant --- partial democracies are not reliably better governed than autocracies once we control for fertility and equal opportunity",
         correct = TRUE,
         message = "This finding was invisible in column (1). When Democracy was the reference, both other categories looked nearly equally worse. Column (2) reveals that partial democracies and autocracies are statistically indistinguishable on governance quality."),
  answer("The constant in column (2), 34.254, gives the predicted governance score for autocracies when fertility rate and equal opportunity are both zero",
         correct = TRUE,
         message = "Correct. In column (2) the reference category is Autocracy, so the constant gives the predicted value for autocracies at X = 0. Compare to column (1): there the reference is Democracy, so the constant (46.160) gives the predicted value for full democracies at X = 0 --- a different quantity entirely."),
  answer("The constant in column (1), 46.160, gives the predicted average governance quality score for full democracies",
         message = "The constant gives the predicted value of the outcome when all other variables are zero --- not the average for the reference category across the real data. Since full democracy is the reference in column (1), the constant is the predicted governance score for full democracies only when fertility rate = 0 and equal opportunity = 0, values no country actually has."),
  answer("The choice of reference category changes the coefficients on wdi_fertility and bti_eo",
         message = "Look at the Fertility Rate and Equal Opportunity rows in both columns --- the values are identical. The reference category choice only affects the intercept and the factor coefficient rows, not the interval variable estimates."),
  type = "learnr_checkbox",
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

### What the Reference Category Reveals

The non-significant partial democracy finding is substantively important. It suggests that the governance advantage of democracy is not a smooth gradient --- partial democracies, which have some competitive elections but lack full civil liberties and rule of law, are not reliably better governed than autocracies once we account for economic equality and fertility. Full democracies are. This finding was completely invisible when Democracy was the reference category.

### How Large Are the Effects?

The same recipe from Example 1 applies to the two continuous variables. Run the code below to get their standard deviations.

**Run and observe: Standard deviations of the continuous explanatory variables**

```{r exercise-sds-governance, exercise=TRUE, exercise.cap="Run and observe: Standard deviations of the continuous explanatory variables"}
sd(qog$wdi_fertility, na.rm = TRUE)
sd(qog$bti_eo, na.rm = TRUE)
```

The output shows standard deviations of about 1.3 children per woman for **wdi_fertility** and about 1.9 points for **bti_eo**. With a slope of -2.068 on fertility rate, a one-sd shift of 1.3 produces an expected change of $1.3 * (-2.068) = -2.7$ governance points --- a modest effect. With a slope of 4.613 on equal opportunity, a one-sd shift of 1.9 produces an expected change of $1.9 * 4.613 = 8.8$ governance points --- a larger effect. Equal opportunity is the more powerful of the two continuous predictors.

The factor variable **democ_f** works differently. Its coefficients describe a discrete jump from one named category to another --- the Democracy and Partial Democracy coefficients already express the most natural comparison: how much better governed are those countries than autocracies, controlling for fertility and equal opportunity. There is no standard deviation to scale by because the variable doesn't vary continuously.

For factor variables, the right question is how large the gap is relative to the outcome's distribution. Run the code below to see the distribution of **iiag_gov**.

**Run and observe: Distribution of the governance outcome**

```{r exercise-dist-governance, exercise=TRUE, exercise.cap="Run and observe: Distribution of the governance outcome"}
summary(qog$iiag_gov)
sd(qog$iiag_gov, na.rm = TRUE)
```

The outcome ranges from roughly 10 to 80, with a standard deviation of about 16 points. The Democracy vs. Autocracy gap of 11.9 points is roughly three-quarters of a standard deviation --- a substantively meaningful difference. The Partial Democracy vs. Autocracy gap of 1.35 points is less than a tenth of a standard deviation --- negligible. The governance gains from democratization are concentrated at the top of the democracy scale, not spread evenly across the spectrum.

### Conduct Hypothesis Tests

The table comparison showed us the estimates. Now let's check whether those estimates hold up to formal scrutiny --- particularly the democracy coefficient, which represents the core theoretical claim. The summary output for the Autocracy-reference model is reproduced here so you can find t-statistics and p-values without scrolling back.

```{r model-gov-output-reference, echo=FALSE, comment=NA}
summary(model_gov)
```

Look at both factor coefficient rows in this output before answering.

**Practice: Test and interpret the Democracy coefficient**

Find the t-statistic and p-value for the Democracy coefficient in your output before answering.

```{r quiz-hypothesistesting-governance, echo=FALSE}
question("Find the t-statistic and p-value for the Democracy vs. Autocracy coefficient in your model output. Which statements are true? Select all that apply.",
  correct = "Right: the p-value provides strong evidence against a zero democracy effect.",
  answer("The p-value is below .05, so we reject the null hypothesis --- full democracies have significantly higher governance quality than autocracies, controlling for fertility and equal opportunity",
         correct = TRUE,
         message = "The data provide strong evidence against a zero difference in governance quality between full democracies and autocracies."),
  answer("Because the Partial Democracy p-value is above .05, neither the Partial Democracy nor the Democracy coefficient is statistically significant",
         message = "These are two separate coefficients with separate tests. The Democracy coefficient is significant; the Partial Democracy coefficient is not. Look at them separately."),
  answer("The significant result proves that democracy causes better governance",
         message = "A significant coefficient is evidence of a relationship in the population. It does not establish causation --- more democratic countries may also differ in other ways (historical, geographic, economic) that contribute to governance quality."),
  answer("A p-value below .05 means there is less than a 5% probability that the null hypothesis is true",
         message = "This is a common misreading of the p-value. The p-value gives the probability of observing a t-statistic this large *if the null hypothesis were true* --- it does not give the probability that the null hypothesis is true. A small p-value means the data would be surprising under the null, not that the null is probably false."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the p-value for the Democracy coefficient. Is it above or below .05?"
)
```

### Assess Model Fit

The coefficients and tests tell us whether the relationships are statistically distinguishable from zero. How much of the variation in governance quality across countries does this three-variable model actually capture?

**Practice: Assess the fit of the governance model**

Find the Adjusted R-squared and residual standard error in your output before answering.


```{r quiz-modelfit-governance-check, echo=FALSE}
question("Which statement correctly describes what the model fit statistics tell us?",
  correct = "Right. The model explains a substantial share of the variation in governance quality, and the residual standard error is modest given the scale of the outcome.",
  answer("The Multiple R-squared of 0.7176 tells us the model explains about 71.76% of the variation in governance quality across the 50 countries with complete data, and under roughly normal errors, about two-thirds of countries fall within about 7.49 governance points of their predicted value",
         correct = TRUE,
         message = "That is strong explanatory power for a three-variable model. Note, however, that only 50 countries have complete data on all variables --- the large missingness means we should be cautious about how broadly these results generalize."),
  answer("The Adjusted R-squared of 0.6925 tells us the model explains about 69.25% of the variation in governance quality",
         message = "The Adjusted R-squared penalizes for the number of predictors and does not have the clean 'percentage of variation explained' interpretation. For that, use the Multiple R-squared of 0.7176."),
  answer("An Adjusted R-squared of 0.6925 and a residual standard error of 7.486 cannot both be correct for the same model, since R-squared must always equal the squared residual standard error",
         message = "These two statistics measure completely different things and there is no mathematical relationship between them. R-squared is a proportion of variance explained; the residual standard error is a standard deviation measured in the units of the outcome."),
  answer("The Multiple R-squared of 0.7176 should be preferred over the Adjusted R-squared of 0.6925 for comparing models with different numbers of predictors",
         message = "It is the other way around: when comparing models with different numbers of predictors, the Multiple R-squared mechanically favors the model with more variables. The Adjusted R-squared penalizes for additional predictors and is the right measure for model comparison."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: which statistic gives the clean 'percentage of variation explained' interpretation, and what does the residual standard error tell you?"
)
```

### What We Learned About Democracy and Governance Quality

So does democracy predict governance quality? The answer depends on what kind of democracy --- and that distinction only becomes visible once you choose the right reference category.

The most important finding from this example is not the significant full democracy coefficient --- it is the non-significant partial democracy finding. When Autocracy is the reference category, we can directly ask whether partial democracies are associated with better governance than autocracies, and in these data they are not distinguishable once we control for fertility and equal opportunity. The governance advantage appears concentrated among full democratic regimes. Partial democracies --- those with competitive elections but weak civil liberties and rule of law --- do not appear to score meaningfully higher on governance quality than their autocratic counterparts, controlling for these two variables.

## Does Democracy Predict the Rule of Law?

Governance quality is a composite measure that aggregates many dimensions --- rule of law, human rights, sustainable economic opportunity, human development. Composite scores are useful summaries, but they can obscure as much as they reveal. A country might score well overall while performing unevenly across specific dimensions. It is worth asking whether democracy's relationship to governance quality holds when we look at one of its most contested components in isolation.

Rule of law --- whether legal systems apply equally to all, whether courts are independent, whether due process is respected --- sits at the core of what democratic institutions are supposed to protect. Free elections remove governments that undermine legal accountability; independent judiciaries constrain arbitrary executive power; civil society monitors compliance. The theoretical case for democracy's effect on rule of law is at least as strong as its case for overall governance quality. But the empirical relationship may look different. Does the pattern we found in Example 2 hold when we zoom in on rule of law specifically?

::: {.hypothesis}
In a comparison of countries, those with higher levels of democratic governance will have stronger rule of law than those with lower levels of democratic governance, after controlling for the fertility rate.
:::

The data frame **qog** contains **fh_rol**, the Freedom House Rule of Law score. It ranges from 0 to 16, where higher values indicate stronger rule of law. The independent variables are **wdi_fertility** and **democ_f**. Use **democ_f** with its default reference category --- Autocracy, the first level, as you confirmed earlier in this tutorial.

### Estimate the Regression

**Practice: Regress rule of law on fertility rate and democracy level**

Write the code from scratch to estimate the model, assigning it to **model3**, then display the results. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-multipleregression-fhrol, exercise=TRUE, exercise.cap="Practice: Regress rule of law on fertility rate and democracy level"}

```

```{r exercise-multipleregression-fhrol-hint-1}
# Hint 1 of 2
# Look back at the hypothesis. Which variable is the outcome
# you are trying to explain? Which variables do we think
# predict it? The outcome goes on the left of ~; predictors
# go on the right separated by +.
```

```{r exercise-multipleregression-fhrol-hint-2}
# Hint 2 of 2
# The structure is the same as Examples 1 and 2: lm() with
# a formula and a data argument, assigned to a named object,
# then passed to summary(). Use the same data frame as
# Example 2. democ_f is already in the environment.
```

```{r exercise-multipleregression-fhrol-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)
  if (!grepl("lm\\s*\\(", code_no_comments)) {
    fail("Use lm() to estimate the regression.")
  }
  if (!grepl("model3\\s*(<-|=)", code_no_comments)) {
    fail("Assign your model to an object called model3.")
  }
  if (!grepl("fh_rol\\s*~", code_no_comments)) {
    fail("The dependent variable should be fh_rol --- rule of law is what we are trying to explain.")
  }
  if (!grepl("wdi_fertility", code_no_comments)) {
    fail("Include wdi_fertility as an independent variable.")
  }
  if (!grepl("democ_f", code_no_comments)) {
    fail("Include democ_f as an independent variable.")
  }
  if (!grepl("data\\s*=\\s*qog", code_no_comments)) {
    fail("Set data = qog so R knows where to find the variables.")
  }
  if (!grepl("summary\\s*\\(", code_no_comments)) {
    fail("Call summary() on your saved model object to display the results.")
  }
  expected <- lm(fh_rol ~ wdi_fertility + democ_f, data = qog)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef)) {
    fail("Make sure you display the results by calling summary() on your model object.")
  }
  if (!isTRUE(all.equal(
    as.numeric(result_coef[order(names(result_coef))]),
    as.numeric(coef(expected)[order(names(coef(expected)))]),
    tolerance = 1e-6))) {
    fail("Check that your model includes fh_rol as the outcome and wdi_fertility and democ_f as predictors, using the qog data.")
  }
  pass("Correct! Look carefully at both the sign and significance of each coefficient --- including the partial democracy one.")
})
```

### Interpret the Coefficients

With your model estimated, the Coefficients table now tells us how fertility rate and each democracy category relate to rule of law, controlling for the other variables. Pay particular attention to the partial democracy coefficient --- is the pattern the same as in Example 2, or different?

**Practice: Interpret the coefficients**

Look at your Coefficients table, paying particular attention to the Partial Democracy row before answering.

```{r quiz-interpret-fhrol, echo=FALSE}
question("Which statements correctly describe what the output tells us? Select all that apply.",
  correct = "Right: all three correctly describe what the output tells us.",
  answer("For each additional child per woman in the fertility rate, a country's rule of law score is expected to fall by about 0.90 points, controlling for democracy level",
         correct = TRUE,
         message = "The slope on wdi_fertility is -0.902 (p < .001) --- countries with higher fertility rates tend to have weaker rule of law even after accounting for regime type."),
  answer("Full democracies score about 7.38 rule of law points higher than autocracies, on average, controlling for fertility rate",
         correct = TRUE,
         message = "The coefficient for Democracy vs. Autocracy is 7.378 (p < .001) --- a large and statistically significant advantage across 185 countries."),
  answer("Partial democracies score about 2.05 rule of law points higher than autocracies, controlling for fertility rate --- and this difference is statistically significant",
         correct = TRUE,
         message = "Unlike in the governance model from Example 2, partial democracies do significantly outperform autocracies on rule of law (p = .008). The full democracy advantage is still much larger (7.38 vs. 2.05), but the partial democracy finding is statistically significant here."),
  answer("For every one-point increase in the rule of law score, we expect the fertility rate to decrease by 0.902, controlling for democracy level",
         message = "This reverses the dependent and independent variables. The slope tells us the expected change in fh_rol for a one-unit increase in wdi_fertility, not the other way around."),
  answer("For every one-unit increase in democracy level, a country's rule of law score is expected to increase by about 7.38 points, controlling for fertility rate",
         message = "Democracy level is not a continuous variable in this model --- it is a three-category factor. The coefficient 7.38 is not a per-unit slope; it is the difference in average rule of law between full democracies and autocracies (the reference category). A one-unit increase in democracy level is not a meaningful quantity here."),
  type = "learnr_checkbox",
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: make sure to select all true statements."
)
```

### How Large Are the Effects?

You have seen the recipe in Examples 1 and 2. Now apply it here on your own. Calculate the standard deviation of **wdi_fertility** and the distribution of **fh_rol** before assessing the size of each effect. For the continuous variable, use $c * \hat{\beta}$ with one standard deviation as $c$. For the factor variable, compare the coefficients to the standard deviation of the outcome. One hint is available. Click "Submit Answer" to check your work.

**Practice: Calculate the distributions needed to assess effect sizes**

```{r exercise-effectsize-fhrol, exercise=TRUE, exercise.cap="Practice: Calculate the distributions needed to assess effect sizes"}

```

```{r exercise-effectsize-fhrol-hint-1}
# Hint 1 of 1
# You need three pieces of information: the standard deviation of
# the continuous explanatory variable, and the summary statistics
# and standard deviation of the outcome. You already know which
# functions to use from Examples 1 and 2.
```

```{r exercise-effectsize-fhrol-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("sd\\s*\\(", code_no_comments)) {
    fail("Use sd() to calculate the standard deviation of at least one variable.")
  }
  if (!grepl("wdi_fertility", code_no_comments)) {
    fail("Calculate the standard deviation of wdi_fertility --- the continuous explanatory variable.")
  }
  if (!grepl("fh_rol", code_no_comments)) {
    fail("Calculate summary statistics or the standard deviation of fh_rol --- the outcome variable.")
  }

  pass("Right. With sd(wdi_fertility) ≈ 1.3 and sd(fh_rol) ≈ 4.0, you can now assess the effects: a one-sd shift in fertility rate is associated with a change of about 1.3 * (-0.902) = -1.17 points in rule of law --- about 0.29 standard deviations of the outcome. The Democracy vs. Autocracy gap of 7.38 points is about 1.8 standard deviations of fh_rol --- a large conditional difference. The Partial Democracy gap of 2.05 points is about 0.5 standard deviations --- modest but statistically distinguishable from zero here, unlike in Example 2.")
})
```

### Conduct Hypothesis Tests

**Practice: Test the hypothesis for each predictor**

Find the p-values for **wdi_fertility**, Partial Democracy, and Democracy in your output before answering.

```{r quiz-hypothesistesting-fhrol, echo=FALSE}
question("Look at the p-values for wdi_fertility, Partial Democracy, and Democracy in your output. Which statements are true? Select all that apply.",
  correct = "Right: and notice the contrast with Example 2 on the partial democracy finding.",
  answer("All three predictors are statistically significant at the .05 level",
         correct = TRUE,
         message = "Partial Democracy (p = .008) is significant here, unlike in Example 2 where it was far from significant (p = .654). Rule of law and overall governance quality are related but not identical outcomes, and democracy's relationship to each turns out to be different."),
  answer("Because Partial Democracy is not significant in Example 2, it cannot be significant in this model either",
         message = "The outcome variable is different. Whether partial democracy outperforms autocracy depends on which dimension of governance you are measuring. The same regime-type differences can produce different results for different outcomes."),
  answer("The p-value for Democracy is far below .05, so we fail to reject the null hypothesis that full democracies have no rule of law advantage over autocracies",
         message = "It is the other way around: a p-value below .05 leads us to reject the null, not fail to reject it. A very small p-value means the data would be highly surprising if the null were true."),
  answer("The significant result for Democracy proves that democratic institutions cause stronger rule of law",
         message = "A significant coefficient tells us the relationship is statistically distinguishable from zero. It does not establish causation --- democracies may differ from autocracies in many other ways that also strengthen rule of law."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the p-value column for each predictor. Which are below .05?"
)
```

### Assess Model Fit

With the coefficients interpreted and tested, how well does this two-predictor model capture the variation in rule of law across countries?

**Practice: Assess the fit of the rule of law model**

Find the Adjusted R-squared and residual standard error in your output before answering.

```{r quiz-modelfit-fhrol, echo=FALSE}
question("Which statement correctly describes what the model fit statistics tell us?",
  correct = "Right. Strong fit across a large sample --- the model explains most of the variation in rule of law and the estimated standard deviation of the residuals is modest.",
  answer("The Multiple R-squared tells us the model explains about 65.3% of the variation in rule of law across the 185 countries with complete data, and under roughly normal errors, about two-thirds of countries have a predicted rule of law score within about 2.87 points of their actual score",
         correct = TRUE,
         message = "That is strong explanatory power for a two-predictor model. The much larger sample here (185 vs. 50 in Example 2) gives us more precise estimates, though it does not resolve whether the 185 countries with complete data are representative of all countries."),
  answer("The Adjusted R-squared tells us the model explains about 64.3% of the variation in rule of law",
         message = "The Adjusted R-squared penalizes for the number of predictors and does not have the clean 'percentage of variation explained' interpretation. For that, use the Multiple R-squared."),
  answer("Because the rule of law score only runs from 0 to 16, a residual standard error of 2.87 means this model's predictions are nearly perfect",
         message = "A residual standard error of 2.87 on a 16-point scale is reasonably precise, but 'nearly perfect' overstates it. Under roughly normal errors, about one-third of countries will have a prediction error larger than 2.87 points --- and with a 16-point range, that is meaningful uncertainty."),
  answer("The Multiple R-squared is always higher than the Adjusted R-squared, so it overstates how well the model fits",
         message = "Being higher does not mean overstating fit. The Multiple R-squared gives the correct 'percentage of variation explained' interpretation for a single model. It is only when comparing models with different numbers of predictors that the Adjusted R-squared is preferred, because the Multiple R-squared would mechanically favor the larger model."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: which statistic gives the clean 'percentage of variation explained' interpretation, and what does the residual standard error tell you?"
)
```

### What We Learned About Democracy and Rule of Law

So does democracy predict rule of law? Yes --- and with a pattern that differs meaningfully from what we found for governance quality in Example 2.

Take a moment before reading this summary to think about what you found. Did partial democracy behave the same way as in Example 2?

Rule of law produces a different pattern from governance quality. Here, partial democracies score significantly higher than autocracies on rule of law --- a finding that was absent in Example 2. The full democracy advantage remains larger (about 7 points on a 16-point scale vs. 2 for partial democracy), but the gradient from autocracy through partial democracy to full democracy is statistically significant and measurable for rule of law, even after controlling for the fertility rate. Comparing Examples 2 and 3 illustrates something important about how we interpret regression results: the association between regime type and governance depends on which outcome we measure. In these data, a researcher studying governance quality would find that partial democracies are not distinguishable from autocracies; a researcher studying rule of law would find that partial democracies score modestly but significantly higher than autocracies, controlling for fertility. Both findings are conditional on the variables included in the model.

## Presenting a Regression Table

The models you have estimated in this tutorial are the kind you would present in a paper or thesis. The `summary()` output you have been reading throughout is the right tool for checking your own work --- but it is not designed for a reader who was not in the room when you ran the code. Variable names come straight from the formula, there is no title, and the layout does not match what journals or thesis committees expect to see. The `stargazer()` function in the [stargazer]{.package-name} package solves all of these problems. This section introduces the basics; Tutorial 18 will build on them substantially.

### A Basic Table

The governance model from Example 2 is available here as **model_gov**. Let's see what stargazer produces with no customization.

**Run and observe: Produce a basic regression table with stargazer**

```{r exercise-stargazer-basic16, exercise=TRUE, exercise.cap="Run and observe: Produce a basic regression table with stargazer"}
library(stargazer)
stargazer(model_gov,
          type = "text")
```

Once a package is loaded with `library()`, all of its functions are available for the rest of your R session --- you do not need to load it again each time you use a function from it. Since [stargazer]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial.

The output is already closer to a publication-ready table: coefficients with standard errors in parentheses, significance stars, and fit statistics at the bottom. But the variable names are still raw formula names. Notice the order in which predictors appear --- top to bottom, exactly as they appear in the model formula. Note also that the Constant (intercept) appears at the bottom. You will need to know both of these things when you add labels.

### The Key Arguments

Here is what you need to produce a labeled, titled table:

**`type`** controls the output format. Use `type = "text"` in these exercises. In your Quarto documents, use `type = "html"` with the chunk option `#| results: asis`.

**`title`** adds a caption above the table.

**`dep.var.labels`** labels the dependent variable column.

**`covariate.labels`** provides one label per predictor row, in the exact order the predictors appear in the basic table output. For factor variables, `levels()` tells you which category is the reference (omitted) and in what order the remaining categories appear. You already checked the levels of **democ_f** earlier in this tutorial: Autocracy is the reference category, followed by Partial Democracy then Democracy. That means **model_gov** needs four labels: one for wdi_fertility, one for bti_eo, one for Partial Democracy, and one for Democracy --- in that order.

::: {.caution}
**Getting the label order wrong puts the wrong label on the wrong coefficient.** Always run the basic table first, note the predictor order, and check `levels()` for any factor variables before writing your labels. Compare the customized table to `summary()` to verify every label landed on the right row.
:::

**`single.row = TRUE`** puts each coefficient and its standard error on one line, making the table more compact and easier to read at a glance.

For a complete list of arguments and customization options, run `?stargazer` in your R console.

### A Customized Table

The prescribed title and dependent variable label are:
- `title = "Predictors of Governance Quality"`
- `dep.var.labels = "Governance Quality Score"`

The covariate labels for this model are: **Fertility Rate**, **Equal Opportunity**, **Partial Democracy**, and **Democracy**. Determine the correct order to supply them from the basic table output above.

**Practice: Produce a labeled, titled stargazer table for the governance model**

Fill in the blanks to complete the table with the prescribed title, dependent variable label, and covariate labels. Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-stargazer-governance, exercise=TRUE, exercise.cap="Practice: Produce a labeled, titled stargazer table for the governance model"}
stargazer(model_gov,
          type = "text",
          title = "___",
          dep.var.labels = "___",
          single.row = TRUE,
          covariate.labels = c("Fertility Rate",
                               "___",
                               "___",
                               "___"))
```

```{r exercise-stargazer-governance-hint-1}
# Hint 1 of 2
# The three remaining labels should cover: Equal Opportunity,
# Partial Democracy, and Democracy --- in the order they appear
# in the basic table output. Check the basic table above and
# confirm the order before filling in the blanks.
```

```{r exercise-stargazer-governance-hint-2}
# Hint 2 of 2
# The title and dep.var.labels are prescribed above.
# covariate.labels needs four entries in formula order:
# Fertility Rate is already filled in as the first.
```

```{r exercise-stargazer-governance-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("stargazer\\s*\\(", code_no_comments)) {
    fail("Use stargazer() to produce the table.")
  }
  if (!grepl("Predictors of Governance Quality", code_no_comments, fixed = TRUE)) {
    fail("Use the prescribed title: \"Predictors of Governance Quality\"")
  }
  if (!grepl("Governance Quality Score", code_no_comments, fixed = TRUE)) {
    fail("Use the prescribed dep.var.labels: \"Governance Quality Score\"")
  }
  if (!grepl("single\\.row\\s*=\\s*TRUE", code_no_comments)) {
    fail("Keep single.row = TRUE to display each coefficient and standard error on one line.")
  }

  labels <- c("Fertility Rate", "Equal Opportunity", "Partial Democracy", "Democracy")
  for (lbl in labels) {
    if (!grepl(lbl, code_no_comments, fixed = TRUE)) {
      fail(paste0("The label \"", lbl, "\" is missing. Use the four prescribed labels in formula order."))
    }
  }

  positions <- sapply(labels, function(lbl) regexpr(lbl, code_no_comments, fixed = TRUE)[1])
  if (!all(diff(positions) > 0)) {
    fail("The labels are not in the correct order. They should appear as: Fertility Rate, Equal Opportunity, Partial Democracy, Democracy --- matching the predictor order in the basic table output.")
  }

  pass("Well done! Compare your customized table to the summary() output to confirm every label landed on the right row. This is the format a reader would see in a paper or report.")
})
```

### A Professional Table in Your Toolkit

You have now produced a regression table in the format that appears in published political science research. Where `summary()` gives you raw R output designed for your own inspection, a stargazer table gives your reader clean labels, standard errors in parentheses below each coefficient, and significance stars that make the key results visible at a glance. Tutorial 18 will build on this foundation --- more models, more customization, and the plots that bring regression results to life for audiences who will never read a table of numbers.

## Check Your Understanding

You have now done the full workflow of a quantitative political scientist: formulated a hypothesis, estimated a model that controls for competing explanations, interpreted each coefficient with the "all else equal" qualifier, tested it for significance, and presented the results in a professional regression table. The three examples also showed something more substantive --- that what you conclude about a relationship depends on what you control for, which outcome you measure, and which comparison your reference category makes possible. That is the real payoff of multiple regression: not just more predictors, but more precise answers to more specific questions.

The exercise below asks you to bring together everything from Examples 1 through 3. Work through it without looking back --- the goal is to see how much of this you can do on your own.

**Apply: Connect the findings across examples**

Think back across all three examples and what changed when additional controls were added before answering.

```{r quiz-cyu-connect, echo=FALSE}
question("Across Examples 1, 2, and 3, which statement best captures what multiple regression adds compared to simple regression?",
  correct = "Right: multiple regression changes both what we estimate and how we interpret it.",
  answer("In multiple regression, each coefficient describes the conditional relationship between a predictor and the outcome among cases with the same values on the other included variables --- this can change both the size and significance of individual coefficients compared to simple regression",
         correct = TRUE,
         message = "In Example 1, wage growth's slope shrank from 0.456 to 0.314 once we controlled for other county characteristics. In Example 2, partial democracy went from appearing significantly worse than full democracy to being indistinguishable from autocracy once we controlled for fertility and equal opportunity. The 'all else equal' qualifier changes both the estimates and their interpretation."),
  answer("Adding controls necessarily brings coefficients closer to their true causal effects, since the remaining variation is unconfounded",
         message = "Controls reduce one source of confounding but do not eliminate all others. Unmeasured variables, measurement error, and other issues can still affect the estimates. Multiple regression produces conditional associations, not causal effects, unless the study design supports that interpretation."),
  answer("Multiple regression separates the unique percentage of variance explained by each predictor, so the R-squared values for individual predictors sum to the model's total R-squared",
         message = "R-squared cannot be partitioned across predictors when they are correlated --- and in observational data they almost always are. The model's R-squared reflects the combined explanatory power of all predictors together, not the sum of their individual contributions."),
  answer("Multiple regression always produces larger R-squared values than simple regression, which proves the added variables matter",
         message = "R-squared does mechanically increase when you add variables, which is exactly why we use the Adjusted R-squared to compare models. An increase in R-squared alone does not prove the added variables matter --- they need significant coefficients too."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what 'holding other variables constant' means for interpretation, and whether adding controls is the same as identifying causal effects."
)
```

## The Takeaways

[**Multiple regression**]{.important-text} extends simple regression to models with more than one independent variable. The formula syntax extends naturally: list each independent variable after the `~`, separated by `+`. The `summary()` output is identical in structure to simple regression, just with more rows in the coefficient table.

The key additions this tutorial introduces:

1. [**All else equal.**]{.important-text} In multiple regression, every coefficient is a partial effect --- the expected change in $Y$ for a one-unit change in that predictor, holding all other predictors constant. This qualifier changes interpretation fundamentally. A significant coefficient in a multiple regression means the variable predicts the outcome *even after accounting for* the other variables in the model.

2. [**Omitted variable bias.**]{.important-text} Omitted-variable bias can occur when an omitted factor helps explain the outcome and is also correlated with an included predictor. In that case, the predictor's coefficient may partly capture the omitted factor's relationship with the outcome. Including relevant controls reduces this problem, but does not eliminate it --- unmeasured confounders, measurement error, and other issues can still affect the estimates.

3. [**Nominal variables with three or more categories.**]{.important-text} A variable with $k$ categories contributes $k - 1$ coefficients to the output, each comparing one non-reference category to the reference. Use `factor()` to create a labeled factor and `relevel()` to choose the reference category deliberately. The choice of reference category does not change the model's fit or the coefficients on interval predictors --- it only changes which comparisons the factor coefficients describe.

4. [**Adjusted $R^2$.**]{.important-text} When comparing models with different numbers of predictors, use the Adjusted $R^2$ rather than the Multiple $R^2$. The Adjusted $R^2$ penalizes for adding predictors and will increase only when the added variable improves in-sample fit enough to outweigh the penalty.

If you take one thing from this tutorial beyond the mechanics, let it be this: multiple regression changes what you are allowed to say. Once you add controls, a coefficient describes the relationship between a predictor and the outcome among cases that have the same values on the other variables in the model. This is often more informative than the raw bivariate relationship, but it does not by itself establish causation. The difference between a simple and a multiple regression is not just statistical; it is the difference between noticing that two things move together and having evidence that they are associated even after accounting for other measured differences. That is worth the extra rows in the coefficients table.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial.

```{r encode-logic-16, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-16, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-16, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You have completed this tutorial and can now estimate and interpret multiple regression models, present results in a professional table, and draw substantive conclusions from what the estimates tell you.">
  <span aria-hidden="true">📊</span>
  Excellent work! You can now estimate and interpret multiple regression models with interval and nominal independent variables --- and present your results in a professional table.
</p>
')
```
