---
title: "Levels of Measurement"
tutorial:
  id: "03-Levels-of-Measurement"
output:
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Define and distinguish the level of measurement of a variable
---

## Overview {#main-content}

Before you can describe a variable, choose a graph, or run a test, you need to know what its values actually mean. Some variables record only which category a case falls into. Others place cases in a meaningful order. Others take ordered values for which the distance between values is itself meaningful. Which of these you are holding determines what you can legitimately do with it --- which summary statistics make sense, which graphs are appropriate, which hypothesis tests apply. That is what a variable's level of measurement tells you, and identifying it is one of the first decisions an analyst makes. Time spent here pays off in every tutorial that follows.

Unlike the other tutorials in this series, this one contains no code at all --- nothing to run and nothing to write. Each section is followed instead by questions that let you practice the distinctions as they are introduced. Read the explanation before answering, use the feedback to work out where a misunderstanding lies, and retry a question as often as you like. The goal is not to memorize a label for each example but to practice asking what the values of a variable actually mean.

In this tutorial you will learn:

* How to define a variable
* How to distinguish nominal, ordinal, and interval levels of measurement
* How to recognize the level of measurement of a variable
* How to distinguish a concept from the variable used to measure it
* How to explain why analysts sometimes treat ordinal variables as interval-level, and what that assumes

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(error = TRUE)
```

## Variables

The purpose of data analysis is to find meaning in data and use it to answer questions. We usually begin with data sets that have already been assembled, but every data set reflects earlier choices about what to measure and how to measure it. Those choices matter. A data set is useful only to the extent that its variables capture the concepts we care about. The variables in these data sets are the building blocks for every analysis we will do.

What is a variable? A [**variable**]{.important-text} records a characteristic that can differ across cases. Analysts often use variables to measure broader concepts they care about. For example, the concept of political participation might be measured with variables such as whether someone voted, how often they attend meetings, or how much they donated to a campaign. If every case in a particular data set has the same value, the variable is constant in those data. The values are often numeric, even when those numbers stand in for labels. Variables may be classified by their level of measurement.


## Levels of Measurement

Across this series we will describe individual variables and analyze relationships between variables. The appropriate method depends on how each variable is measured. The [**level of measurement**]{.important-text} tells us what the values of a variable mean: whether they are unordered categories, ordered categories, or numbers with equal distances between them. In this series we distinguish among three levels of measurement: **nominal**, **ordinal**, and **interval**.

The three levels form a hierarchy, and what accumulates is the set of summaries each one supports. Identifying the most common category requires only labels, so it works for any variable. Identifying a middle category requires order, so it works for ordinal and interval variables but not nominal ones. Computing an arithmetic average requires meaningful, equal distances, which only interval variables provide. So an interval variable can be summarized in any of these ways, an ordinal variable in the first two, and a nominal variable only in the first. Using fewer of a variable's properties than you have is always legitimate --- you are declining information, not misusing it.

Moving up the hierarchy is a different matter. Treating ordered categories as though the distances between them were equal is a claim about the data, not a fact about them, and it has to be justified rather than assumed. We return to that at the end of the tutorial.


## Nominal Variables

A [**nominal variable**]{.important-text} takes qualitative values that denote categories. A nominal variable's values are labels or names that categorize observations with respect to the attribute being measured. *There is no intrinsic order to the categories.* If rearranging the categories would not change their meaning, you are probably looking at a nominal variable.

A variable that measures individual partisan status as "Democrat," "Independent," "Republican," or "Other/None" is an example of a nominal variable. Note that you can code nominal variables with numbers, but the numbers themselves are meaningless, and the order they are assigned is arbitrary: if we labeled the values of partisan status 1, 2, 3, and 4, respectively, it would not imply "Democrat" is less than "Independent" or "Republican" is less than "Other/None."

Two-category variables, such as whether a respondent in a survey voted for president in the most recent election (coded "yes" or "no"), are often referred to as [**binary**]{.important-text} or [**dichotomous**]{.important-text} variables, and are conventionally treated as nominal. Two categories on their own do not create a ranked scale, whatever order you happen to list them in.

R has a special class of variable called a factor that may be used for nominal variables, but not all nominal variables will automatically be assigned this class. When a nominal variable is stored as a factor, R will treat it as categorical, which can be useful when summarizing or modeling the data. (We will discuss factors more in a later tutorial.)

**Practice 1: Identify the nominal-level variables**

```{r nominal-variable-question, echo=FALSE}
question("Which of the following variables are nominal level measures? Select all that apply.",
  correct = "Right. Rearranging these categories would change nothing, which is the test for nominal.",
  answer("Whether an individual has not completed high school, has a high school degree, has attended some college, has a college degree, or has a postgraduate degree.",
         message = "The education variable is not nominal --- its categories are rank-ordered, with each representing a higher level of attainment than the one before it. That makes education level an ordinal variable."),
  answer("Whether a state's electoral votes were assigned to Donald Trump or Hillary Clinton", correct = TRUE,
         message = "The Trump/Clinton electoral vote variable is nominal. Its two values are unordered categories --- there is no sense in which one candidate's outcome is more or less than the other's, only different. Two-category nominal variables like this are also called binary or dichotomous."),
  answer("The number of protest events in the week following George Floyd's death",
         message = "The protest event count variable is not nominal --- its values are counts, not category labels. Count variables are interval-level because the values are ordered and the differences between them are equal."),
  answer("Whether a country has a parliamentary, presidential, or mixed democratic system, or is not democratic", correct = TRUE,
         message = "The democratic system variable is nominal. Its four categories --- parliamentary, presidential, mixed, and not democratic --- are unordered labels. Coding them 1, 2, 3, 4 would be arbitrary; no arrangement of these categories carries any inherent meaning."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: There are two nominal variables in this list. Look for variables whose values are unordered category labels."
)
```

## Ordinal Variables

An [**ordinal variable**]{.important-text} takes values with a natural order or rank, but the differences between those ranks are not equal. The order matters, but the exact size of the steps between categories does not.

An example of an ordinal variable is a variable that measures whether someone views the national economy today compared to a year ago as "better," "about the same," or "worse." We can be confident that "worse" is less positive than "about the same," which is less positive than "better," but we cannot treat those categories like numbers on a scale. We do not know how far apart the categories are, so it would not make sense to say that "better" is twice as good as "about the same" or three times better than "worse."

As another example, consider income categories on a survey where the variable takes a value of 1 for incomes less than $20,000, 2 for incomes $20,000 to $49,999, 3 for incomes $50,000 to $99,999, 4 for incomes $100,000 to $199,999, 5 for incomes $200,000--$499,999, and 6 for incomes $500,000 or more. Higher numbers mean more income, but the interval between a respondent in category 1 and 2 is not the same as between category 4 and 5. This is the key feature of ordinal measures: *the order is meaningful, but the interval between values is not equal*. Put differently, the step from one category to the next does not represent the same amount of income throughout the scale.

Most of the time, an ordinal variable will be coded with numeric rather than character values. R does have an ordered factor class, but we will not use it here.

**Practice 2: Identify the ordinal-level variables**

```{r ordinal-variable-question, echo=FALSE}
question("Which of the following variables are ordinal level measures? Select all that apply.",
  correct = "Right. Both impose a clear order without guaranteeing equal gaps between adjacent ranks.",
  answer("Whether a county's population is majority white, Black, Hispanic, or some other race",
         message = "The racial majority variable is nominal, not ordinal. Its categories cannot be rank-ordered --- there is no sense in which one racial majority is more or less than another."),
  answer("Whether Freedom House labels a country as free, partly free, or not free", correct = TRUE,
         message = "The Freedom House variable is ordinal. Its three categories run in a clear order from not free to partly free to free, but the distance between adjacent categories is not a known equal unit. We cannot say 'partly free' is exactly halfway between 'free' and 'not free.'"),
  answer("A nation's level of inflation",
         message = "Inflation is not ordinal --- it is measured as a percentage, so the values are ordered and the differences between them are equal. One percentage point of inflation is the same size everywhere on the scale. That makes inflation interval-level."),
  answer("How satisfied an individual is with the direction the country is headed, on a five-point scale", correct = TRUE,
         message = "The satisfaction variable is ordinal. Its five response options can be ranked from most to least satisfied, but the psychological distance between adjacent categories is not necessarily equal --- that is the defining feature of an ordinal scale."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: There are two ordinal variables in this list. Look for variables whose values can be ranked but whose gaps between ranks are not necessarily equal."
)
```

Note that both nominal and ordinal level variables are considered [**categorical**]{.important-text} variables because their values fall into categories.


## Interval Variables

An [**interval variable**]{.important-text} takes values that have a natural order **and** the differences between values are equal. Examples include monetary amounts such as GDP; percentages such as unemployment; or quantities that can be counted, such as the number of terrorist acts. For interval variables, subtraction makes sense: the difference between 4 and 5 has the same meaning as the difference between 10 and 11.

A special type of interval variable, often given its own level of measurement, is a [**ratio variable**]{.important-text}. A ratio variable is a type of interval variable that has a natural zero point. For simplicity, we will simply refer to these variables as interval-level.

Interval variables may be [**discrete**]{.important-text} or [**continuous**]{.important-text}. A discrete variable takes separate, countable values, such as 0, 1, 2, or 3 protest events. A continuous variable can in principle take any value within a range. R may describe numeric variables using storage labels such as `int`, `num`, or `dbl`, depending on how the data are stored and displayed. These R storage labels are useful for understanding how R sees the data, but they are not the same thing as a variable's level of measurement.


**Practice 3: Identify the interval-level variables**

```{r interval-variable-question, echo=FALSE}
question("Which of the following variables are interval (or ratio) level measures? Select all that apply.",
  correct = "Right. Both are ordered, and equal differences carry equal meaning anywhere on the scale.",
  answer("Whether a state's governor is a Republican or Democrat or something else",
         message = "The gubernatorial partisanship variable is nominal, not interval. Its values denote categories that cannot be rank-ordered."),
  answer("The level of foreign direct investment in a country measured in dollars", correct = TRUE,
         message = "Monetary amounts are interval-level --- and more specifically ratio-level --- because equal differences have the same meaning throughout the scale and zero represents none of the quantity."),
  answer("Whether an individual approves, neither approves nor disapproves, or disapproves of the president's handling of the economy",
         message = "The presidential approval variable measured this way is ordinal, not interval. Its three categories are ranked, but the psychological distance between adjacent categories is not necessarily equal."),
  answer("The homicide rate in a country in a year", correct = TRUE,
         message = "A rate is a count divided by population, and that division leaves the spacing intact: one additional homicide per 100,000 people means the same thing at 1 per 100,000 as it does at 20."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: There are two interval variables in this list. Look for values that are ordered and for which the same numerical difference carries the same meaning anywhere on the scale."
)
```


::: {.tip}
**Quick classification rule:** Ask three questions. Are the values unordered category labels? If yes, the variable is nominal. If the values are ordered, ask whether the differences between values are meaningful and equal. If the ordered values do not have equal distances, the variable is ordinal. If the ordered values do have meaningful, equal distances, the variable is interval.
:::

## Additional Considerations

There are two important considerations that can lead to confusion in determining the level of measurement of a variable. The first is that the same *concept* can often be measured at multiple levels of measurement. The second is that in some cases ordinal variables are treated as interval-level variables.

### Concept versus Variable

A [**concept**]{.important-text} is an idea such as "support for the president" or "income inequality." When analysts operationalize a concept, they create or select one or more variables to measure it. Analysts may choose to measure concepts in a variety of ways. Income inequality in a country might be measured using the Gini index or the share of total income going to a particular income group, for example. The degree of political polarization in the United States over time might be measured as the difference in the percentage of Democrats and Republicans who support a particular public policy, e.g., the death penalty, gun control, or abortion rights. Or it might be measured as the proportion of survey respondents who agree that the other party is "a threat to the Nation's well-being."

Thus, a concept does not have a level of measurement. Instead, how a concept is measured determines the level of measurement of a variable. This is why the same idea can show up in different statistical forms.

Consider support for Donald Trump in US counties in 2016:

**Practice 4: Identify the measurement level for a binary majority indicator**

```{r majority-measurement-question, echo=FALSE}
question("What level of measurement would a variable be if it was coded as whether or not a majority of the county vote was cast for Donald Trump?",
  correct = "Right. Two unordered categories, so nominal --- and specifically binary.",
  answer("Ordinal",
         message = "The two observed values identify categories rather than positions on a ranked scale, and coding them 0 and 1 does not by itself create an order."),
  answer("Nominal", correct = TRUE,
         message = "Coding the two outcomes 0 and 1 is a labeling convenience, not a ranking. Swap the codes --- 1 for counties without a Trump majority, 0 for counties with one --- and nothing about the variable changes. Numbers attached to categories do not create an order."),
  answer("Interval",
         message = "The values of this variable are categories --- yes or no --- not numbers with equal distances between them. Interval variables require meaningful, equal-sized gaps between values."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Think about whether the two values --- majority for Trump or not --- can be rank-ordered. Is one value higher or lower than the other, or are they just different categories?"
)
```

**Practice 5: Identify the measurement level for a percentage**

```{r percentage-measurement-question-1, echo=FALSE}
question("What level of measurement would a variable be if it was coded as the percentage of the vote cast for Donald Trump?",
  correct = "Right. The same concept can have a different level of measurement when it is measured differently.",
  answer("Nominal",
         message = "The values this variable takes are ordered and the differences between them are meaningful, so it is not nominal. Nominal values are unordered category labels."),
  answer("Ordinal",
         message = "The values this variable takes are ordered and the differences between them are meaningful and equal --- a county at 60% is exactly 10 points higher than one at 50%. That equal-interval property makes it interval-level, not ordinal."),
  answer("Interval", correct = TRUE,
         message = "Vote share is an interval variable. It is ordered and the differences between values are equal --- a one-percentage-point difference in vote share is the same size anywhere on the scale."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Ask whether the values are ordered and whether the differences between them are equal throughout the scale."
)
```

Consider political ideology, which can be measured in more than one way:

**Practice 6: Classify the 1--7 ideology scale**

```{r scale-measurement-question, echo=FALSE}
question("What level of measurement would a variable measuring ideology have if it was coded based on responses to the question: On a scale of 1 to 7, where 1 is very liberal and 7 is very conservative, where would you place yourself?",
  correct = "Right. The scale is ordered, but its steps are not necessarily equal.",
  answer("Nominal",
         message = "The values this variable takes are ordered from most liberal to most conservative, so it is not nominal. Nominal values are unordered category labels."),
  answer("Ordinal", correct = TRUE,
         message = "The values run from most liberal to most conservative, but the differences between adjacent points are not necessarily equal --- a move from 3 to 4 does not necessarily represent the same ideological shift as a move from 6 to 7."),
  answer("Interval",
         message = "The values are ordered, but the differences between adjacent points are not necessarily equal, so this scale is ordinal. This is not a careless answer, though: analysts often do treat scales like this as interval-level, which is defensible so long as the assumption behind it is acknowledged rather than assumed away."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The values are clearly rank-ordered from most liberal to most conservative. Ask whether the differences between adjacent points are necessarily equal."
)
```

**Practice 7: Identify the measurement level for a voting-with-leadership percentage**

```{r percentage-measurement-question-2, echo=FALSE}
question("Suppose an analyst uses the percent of times a member of Congress voted with the Republican leadership as a measure of ideology. Whether that percentage captures ideology well is a separate question --- but what level of measurement is the variable itself?",
  correct = "Right. Ideology again, and again the measure decides the level, not the concept.",
  answer("Nominal",
         message = "A percentage is not a category label. These values run from 0 to 100 in a meaningful order, which rules nominal out immediately."),
  answer("Ordinal",
         message = "Ordinal would be right if the values were merely ranked. Here they are more than ranked: the gap between 60% and 70% is the same quantity as the gap between 80% and 90%, so the variable carries equal-interval information as well as order."),
  answer("Interval", correct = TRUE,
         message = "A percentage is interval-level. A member who votes with leadership 70% of the time is exactly 10 points higher than one who does so 60% of the time --- the differences are equal throughout the scale."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Ask whether the values are ordered and whether a one-point difference carries the same meaning anywhere on the scale."
)
```

### Ordinal Variables Are Sometimes Treated as Interval Variables

One final complication is worth naming directly: analysts sometimes treat an ordinal variable as if it were measured at the interval level. Consider the level of education coded as whether an individual has not completed high school, has a high school degree, has attended some college, has a college degree, or has a postgraduate degree. Often analysts treat the differences between categories *as if* they are equal. This is more common when an ordinal variable has several ordered categories and analysts judge the spacing between them to be roughly even --- a seven-point ideology scale is treated this way far more often than a three-category measure such as "free," "partly free," and "not free." In some cases such treatment may be justified, but it is important to recognize that in treating an ordinal variable as an interval-level variable, an analyst is assuming that the differences between values are equal.

## Check Your Understanding

The following questions ask you to apply what you have learned about levels of measurement. Each question requires you to connect more than one idea --- think carefully about what the variable's values actually represent before choosing an answer. If you get one wrong, read the feedback to identify which part of the distinction tripped you up.

**Practice 8: Distinguish ordinal and interval variables**

```{r apply-ordinal-interval-question, echo=FALSE}
question("What is the key difference between an ordinal variable and an interval variable?",
  correct = "Right. Order separates both from nominal; equal distances separate interval from ordinal.",
  answer("Ordinal variables use words, while interval variables always use numbers.",
         message = "Not necessarily. Ordinal variables are often coded with numbers --- income categories coded 1 through 6, for example. What matters is not the label type but whether the distances between values are meaningful and equal."),
  answer("Ordinal values are ordered; interval values are ordered with meaningful, equal distances.", correct = TRUE,
         message = "This is the key distinction. Both types of variable have ordered values. The difference is whether the distance between adjacent values can be interpreted as equal throughout the scale."),
  answer("Ordinal variables have no order, while interval variables do.",
         message = "Variables with no order are nominal, not ordinal. Ordinal variables do have a meaningful order --- that is what makes them ordinal. The distinction from interval is about equal distances, not the presence of order."),
  answer("Ordinal variables can only have two categories, while interval variables can have many values.",
         message = "Two-category variables are binary or dichotomous --- a type of nominal variable. Ordinal variables can have many categories. The distinction from interval is about equal distances between values, not the number of categories."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Both types have ordered values. Focus on what distinguishes them: ask whether the gaps between adjacent values are equal and meaningful."
)
```

**Apply 9: Compare measures of the same concept**

```{r apply-same-concept-question, echo=FALSE}
question("Suppose the concept is support for stricter gun laws. Which pair correctly classifies two different ways to measure that concept?",
  correct = "Right. A binary item tells you which side of a divide someone falls on; a percentage tells you how much support exists across a population.",
  answer("Support or oppose stricter gun laws: interval; percentage of survey respondents in a state who support stricter gun laws: nominal.",
         message = "This reverses the classifications. A binary support/oppose variable is a two-category nominal variable, not interval. A percentage is interval-level because the differences between values are ordered and equal."),
  answer("Support or oppose stricter gun laws: nominal; percentage of survey respondents in a state who support stricter gun laws: interval.", correct = TRUE,
         message = "Support or oppose is a two-category nominal variable --- the two values are unordered categories. With only two unordered response options, we treat this as binary nominal. A percentage is interval-level because the values are ordered and the differences between them are equal."),
  answer("Strongly oppose, oppose, neither, support, strongly support: nominal; support or oppose: ordinal.",
         message = "The five-category scale is ordinal because the categories can be ranked, but the gaps between them are not necessarily equal. Support or oppose is a two-category nominal variable, not ordinal."),
  answer("Percentage of survey respondents in a state who support stricter gun laws: ordinal because higher percentages mean more support.",
         message = "Higher percentages do mean more support, but that alone does not make the variable ordinal. Percentages are also equal-interval --- a difference of 10 percentage points is the same size anywhere on the scale. That makes this measure interval-level."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Classify each measure separately. For the binary measure, ask whether the two values are unordered categories or a ranked scale. For the percentage, ask whether the values are ordered and whether equal differences carry equal meaning."
)
```

**Apply 10: Recognize an assumption about ordinal variables**

```{r apply-ordinal-as-interval-question, echo=FALSE}
question("When an analyst treats an ordinal variable as if it were interval-level, what assumption is the analyst making?",
  correct = "Right. Treating the categories as interval-level assumes that the gaps between them are close enough to equal.",
  answer("That the categories have no meaningful order.",
         message = "That would describe a nominal variable. Ordinal variables do have a meaningful order --- that order is what makes treating them as interval-level a reasonable starting point rather than a random one."),
  answer("That the distances between categories are close enough to equal.", correct = TRUE,
         message = "This is the assumption. The analyst is saying that even though the distances between ordered categories are not guaranteed to be equal, they are close enough to equal that treating them as such will not seriously distort the analysis."),
  answer("That the variable has only two categories.",
         message = "A two-category variable is binary or dichotomous --- a type of nominal variable. The ordinal-as-interval assumption applies to variables with many ranked categories, not just two."),
  answer("That the numbers assigned to the categories are arbitrary labels.",
         message = "Arbitrary numeric labels describe nominal variables --- for example, coding party affiliation as 1, 2, 3. For ordinal variables, the numbers carry order information. The ordinal-as-interval assumption goes further, treating those ordered numbers as if they also carry equal-distance information."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The variable is already acknowledged to be ordinal, so the analyst knows the values are rank-ordered. What additional property is the analyst assuming exists?"
)
```


## The Takeaways

The [**level of measurement**]{.important-text} of a variable determines how we describe its central features and how we analyze relationships between variables. Every method in this series begins with this step: identifying what kind of variable we are working with.

A [**variable**]{.important-text} records a characteristic that can differ across cases. Variables with unordered or ordered category values are [**categorical**]{.important-text}. A two-category nominal variable is often called [**binary**]{.important-text} or [**dichotomous**]{.important-text}.

Three levels cover everything you will meet in these tutorials. A [**nominal variable**]{.important-text} takes unordered category labels --- rearranging them changes nothing. An [**ordinal variable**]{.important-text} takes values with a meaningful order, but with no guarantee that adjacent steps are the same size. An [**interval variable**]{.important-text} takes ordered values whose differences are equal throughout the scale, which is what makes subtraction meaningful. A [**ratio variable**]{.important-text} is an interval variable with a meaningful zero point; these tutorials include ratio variables within the interval category. Interval variables may be [**discrete**]{.important-text}, taking separate countable values, or [**continuous**]{.important-text}, potentially taking any value within a range. R storage labels such as `int`, `num`, and `dbl` describe how values are stored, not their level of measurement.

Analysts sometimes treat an ordinal variable as if it were interval-level, particularly when it has several ordered categories whose spacing looks roughly even --- a seven-point ideology scale far more often than a three-category rating. That may be defensible, but it is an assumption rather than a property of the data: it asserts that the distance between adjacent categories is the same everywhere on the scale.

The quick classification rule from the end of the interval section is worth keeping in mind as a practical checklist: start with order, then ask about equal distances. When in doubt, slow down and ask what the values actually represent --- are they labels, ranks, or numbers with equal gaps between them?

The hardest cases are often not about memorizing definitions, but about noticing how a [**concept**]{.important-text} has been measured. Support for a candidate, satisfaction with democracy, or political ideology can be measured in more than one way. Your task is always to classify the variable in front of you, not the broader concept behind it.

One complication is worth carrying forward. Analysts frequently treat an ordinal variable as though it were interval-level, particularly when the scale has many ordered points --- a seven-point ideology scale, or income recorded in six brackets. This is common practice and often defensible, but it is a choice rather than a fact about the variable. What the analyst is assuming is that the steps between adjacent categories are close enough to equal that treating them as equal will not distort the result. That assumption may be reasonable; it should still be stated rather than left implicit.

The table below summarizes the three levels of measurement, gives a short description of each, and provides examples.

::: {.qpa-infographic}
<table class="qpa-table narrow-label">
  <caption>Levels of Measurement</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Level of measurement</span></th>
      <th scope="col">Ordered?</th>
      <th scope="col">Equal Differences?</th>
      <th scope="col">Examples</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Nominal</th>
      <td>No --- the values are labels, and rearranging them would change nothing</td>
      <td>Not applicable</td>
      <td>Partisan identification; racial majority of a county; marital status</td>
    </tr>
    <tr>
      <th scope="row">Ordinal</th>
      <td>Yes --- the categories have a natural rank</td>
      <td>No --- the steps between adjacent ranks are not equal</td>
      <td>Freedom House ratings of not free, partly free, and free; a seven-point ideology scale</td>
    </tr>
    <tr>
      <th scope="row">Interval (Ratio)</th>
      <td>Yes</td>
      <td>Yes --- subtraction is meaningful, so a difference of 5 means the same anywhere on the scale</td>
      <td>GDP and other monetary amounts; the unemployment rate; a count of protest events</td>
    </tr>
  </tbody>
</table>
:::

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-03, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-03, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-03, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Nicely done! You can now identify and distinguish levels of measurement, a foundation you will use throughout the series.">
  <span aria-hidden="true">🎉</span>
  Nicely done! You can now identify and distinguish levels of measurement, a foundation you will use throughout the series.
</p>
')
```
