---
title: "From Coefficients to Predicted Values"
tutorial:
  id: "17-Regression-Predictions"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Turning coefficients into predicted values and plots a reader can picture.
---

## Overview {#main-content}

A statistically significant coefficient does not always tell a reader whether a relationship is substantively important. How much higher is turnout predicted to be in a wealthy community than in a poor one? How different are the predicted outcomes for democracies and autocracies? What does a model imply for realistic cases rather than for a one-unit change in isolation? Answering questions like these requires translating regression coefficients into predicted values that can be compared and visualized.

This tutorial shows how to turn an additive regression model into concrete predictions and prediction plots. It builds on Tutorials 15 and 16, where you learned to estimate models, interpret coefficients, test hypotheses, assess model fit, and present results in a regression table. Here, you will use those results to communicate what the model predicts across substantively meaningful values of the predictors.

In this tutorial you will learn:

* How to read a predicted value as a model-implied comparison rather than a forecast for an individual case
* How to obtain specific predicted values using `predict()` and how to visualize predicted values across a range using `plot_model(type = "pred")` from the [sjPlot]{.package-name} package, including how to vary two predictors at once and how to choose which variable goes on the x-axis
* How to use the `condition` argument to hold specific predictors at substantively meaningful values rather than the defaults, and why that choice matters for interpretation
* How to interpret any predicted value comparison using a systematic framework covering the pattern, held values, specific predicted values, uncertainty, and the hypothesis connection
* How to recognize parallel prediction lines as the signature of an additive model

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(gradethis)
library(stargazer)
library(sjPlot)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
options(lifecycle_verbosity = "quiet")
knitr::opts_chunk$set(error = TRUE)
data("states", package = "QPATutorialsCourse")
states$p.liberal <- states$liberal / 100
states$region_censusf <- factor(states$region_census,
                                 levels = c(1, 2, 3, 4),
                                 labels = c("South", "West", "Midwest", "Northeast"))
fit1 <- lm(hs_dem_prop_all ~ nonwhite + p.liberal + povrate + region_censusf,
           data = states)
data("ipu", package = "QPATutorialsCourse")
ipu$electoral_f <- factor(ipu$electoral_system,
                          levels = c("plurality_majority",
                                     "proportional_representation",
                                     "mixed_system"),
                          labels = c("Plurality/Majority", "Proportional", "Mixed"))
ipu$quota_f <- factor(ipu$gender_quota, levels = c(FALSE, TRUE),
                      labels = c("No quota", "Quota"))
ipu$years_c <- ipu$years_since_first_woman -
               mean(ipu$years_since_first_woman, na.rm = TRUE)
model_rep <- lm(women_percent ~ quota_f + electoral_f + years_c, data = ipu)
```

## From Model to Meaning: Choosing and Interpreting Predictions

In Tutorials 15 and 16, you learned to estimate regression models, interpret their coefficients, test hypotheses, and assess model fit. That work produces a table of numbers --- and a good table is worth having. Each coefficient tells you the direction and size of a predictor's relationship with the outcome, holding everything else constant --- the [**effect**]{.important-text} of that predictor, in the statistical sense the term carries throughout these tutorials, without any claim that one variable causes the other. But a table of numbers does not always make that relationship easy to picture in substantively meaningful terms.

Think about what you actually want to know. If you are studying what predicts voter turnout across counties, you might ask: how much higher is the predicted turnout in a wealthy county compared to a poor one, holding unemployment constant? What does the model predict for a county with average income and average unemployment --- as opposed to one with high income but also high unemployment? A coefficient tells you the direction and per-unit size of the income relationship, but it does not tell you how large that difference is across realistic values represented in your data, in units a reader can picture. That is what predictions are for.

Predicted values translate the fitted regression into concrete [**model-implied comparisons**]{.important-text}. You specify a realistic profile of predictor values and ask what outcome the model expects, on average, for cases with those characteristics. This is an [**expected outcome**]{.important-text}: it describes the average outcome predicted for cases with that profile, not the exact outcome that every individual case will have. Compare two profiles that differ in one predictor while holding the others fixed, and you can see how the expected outcome changes across substantively meaningful values of that predictor. These are model-based predictions. They describe how expected outcomes differ across specified predictor values. With the observational data used in these tutorials, they do not show what would happen if we deliberately changed one of those predictors.

When your model has several predictors, which ones should you examine? The answer comes from your research questions, not from the software. You vary the predictor --- or predictors --- whose relationship with the outcome you are trying to communicate. If your hypothesis concerns voter turnout and income, you produce predictions across the range of income. If your hypothesis also concerns the role of unemployment, you produce a second set of predictions varying unemployment. If your research question involves several predictors of interest, look at predictions for each one. Everything else gets held constant at values you choose and report.

In this tutorial you will use two complementary tools. `predict()` answers one what-if question at a time: you specify a complete set of predictor values, and it returns the model's expected outcome for cases with that profile. Running it for two carefully chosen profiles gives you concrete, legible numbers in the units of the outcome --- numbers that anchor your understanding of what the model is saying. `plot_model(type = "pred")` then shows many such predictions at once, across the full observed range of a predictor, with uncertainty visible as a shaded band. If you have already computed two predicted values for a specific comparison, you know exactly what the y-axis means when you see those values on a plot. The two tools work together --- the numbers anchor your interpretation, and the plot fills in the full picture.

Once you have made a comparison --- whether two numbers from `predict()` or a full prediction plot --- a complete interpretation covers five things:

::: {.important-note}
**The five-component framework**

1. **The pattern.** What does the comparison show overall? Is the relationship positive or negative, steep or shallow, consistent across the range or concentrated at one end?

2. **The [**held values**]{.important-text}.** Every predicted value assumes specific values for all predictors not being varied. State them. In an additive linear regression model, changing the held values does not change the predicted difference produced by varying that predictor --- only the baseline level shifts. When the displayed predictor is interval, this appears as a line shifting up or down without changing its steepness. When a categorical predictor is displayed, the predicted differences among its categories remain the same across values of the other predictors. The conventional defaults --- means for interval variables, reference category for factor variables --- provide a standardized comparison case, though they are not necessarily substantively typical of any real observation.

3. **Specific predicted values.** At least two. "The predicted proportion increases" is weaker than "the predicted proportion increases from 0.58 to 0.67 as the liberal proportion rises from 0.25 to 0.35." Magnitude is part of the finding.

4. **The uncertainty.** For a pair of specific predictions, adding `interval = "confidence"` to your `predict()` call returns lower and upper bounds alongside each predicted value. In a prediction plot, the same uncertainty appears as a shaded band around the line. A [**confidence interval**]{.important-text} shows how precisely the model estimates the expected outcome for cases with that predictor profile. Narrow intervals indicate greater precision; wide intervals indicate greater uncertainty. That should change how firmly you state the finding: with a narrow band you can report the predicted value as a reasonably settled quantity, while with a wide one the honest claim is about the range rather than the point. A prediction whose band spans several points on the outcome's scale has not pinned much down, however confident the line through the middle of it looks. Comparing the intervals also helps you see how clearly the expected outcomes are separated. However, overlap is not an exact significance test of the difference between them. When the difference corresponds directly to a regression coefficient, the test of that coefficient provides the formal evidence about whether the difference is statistically significant.

5. **The connection to the hypothesis.** A predicted value comparison is evidence. Does the pattern support the hypothesis, fail to support it, or complicate it?
:::

Here is what a complete interpretation looks like written as a single paragraph.

A researcher models county voter turnout as a function of median household income, unemployment rate, and urban/rural status, with urban counties as the reference category. She produces a prediction plot showing predicted turnout across the range of income for urban counties at average unemployment. A complete interpretation would read:

::: {.research-example}
*As median household income increases, predicted voter turnout rises --- the relationship is positive and linear across the observed range. Predictions hold the unemployment rate at its mean and assume urban counties; a rural county at the same income level would be predicted to have lower turnout by about 4 percentage points, but the estimated income association would be the same. At $30,000 median income, the model predicts about 54% turnout; at $60,000, about 67% --- a 13-point difference across a realistic range. The confidence band is moderately narrow throughout, reflecting a reasonably large sample. This positive relationship is consistent with the hypothesis that wealthier communities have more resources for civic participation.*
:::

Notice that all five components appear naturally in the prose without being labeled. That is the standard this tutorial asks you to meet.

### The Workflow

Everything in Tutorials 15 and 16 has been building toward this moment: you have estimated a model, interpreted its coefficients, tested the hypotheses, and assessed model fit. This tutorial does not replace that work. It extends it by translating selected parts of the fitted model into concrete comparisons that are easier to communicate.

The workflow this tutorial follows has three steps. We begin with a **regression table** --- not because tables are exciting, but because the table documents the complete model. Every coefficient is there, with its standard error and its significance. Any reader who wants to verify the analysis or build on it starts here. The coefficient table establishes the estimated relationships, their uncertainty, and model fit. Predictions then translate selected parts of that table into comparisons across substantively meaningful values.

We then ask the model for **specific predicted values** --- a comparison for two carefully chosen cases. A coefficient of 0.876 says the slope is positive; a predicted value of 0.58 at **p.liberal** = 0.25 and 0.67 at **p.liberal** = 0.35 says what that slope actually means for the predicted gap between a moderately liberal and a more liberal state. Before you draw a line, you know two of its points.

Finally, we **visualize**. The prediction plot sweeps across the full range of the predictor, showing every prediction at once, with the uncertainty visible as a shaded band. By the time you see the plot, you already understand what you are looking at --- because you just computed it by hand for two specific cases.

This is the sequence to follow whenever you interpret a prediction plot --- in the examples below and in your own work.

## What Is Associated with Democratic Representation in State Legislatures?

State legislatures vary dramatically in their partisan composition. What explains these differences? The theoretical argument runs through electoral representation: in a system where legislators are elected by geographic constituencies, the ideological preferences of a state's voters should be reflected in the partisan composition of its legislature. States where more residents identify as liberal provide a larger pool of potential Democratic voters, create more favorable conditions for Democratic candidates, and shift the median voter in legislative districts toward the left. This suggests that public ideology should be a strong positive predictor of Democratic seat share --- and that the relationship should persist even after accounting for other features of states that might confound it. Racial composition, poverty, and regional political culture are all plausibly correlated with both ideology and legislative outcomes; including them in the model lets us assess whether the association between public opinion and seat share persists after accounting for those potentially confounding features.

::: {.hypothesis}
In a comparison of states, those with more liberal public opinion will have a higher proportion of Democratic state house seats than those with less liberal public opinion, after controlling for racial composition, poverty rate, and census region.
:::

The data frame **states** describes every U.S. state as of 2010. Two of the variables below are built in the tutorial's setup rather than shipped: **p.liberal** rescales **liberal** to a proportion, and **region_censusf** turns the numeric region code into a labeled factor. One state is dropped for missing data, leaving 49 of the 50 in the model below. It contains the following variables:

- **hs_dem_prop_all**: proportion of state house seats held by Democrats
- **p.liberal**: proportion of state residents identifying as liberal
- **nonwhite**: proportion of state residents who are not white
- **povrate**: state poverty rate (percent)
- **region_censusf**: a four-category factor --- South (reference), West, Midwest, Northeast

### Estimate the Model

`lm()`, `summary()`, and `predict()` are part of base R --- no `library()` call is needed for those functions.

**Practice 1: Estimate the regression model**

Save your model as **fit1** and present the results. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-fit1-estimate, exercise=TRUE, exercise.cap="Practice 1"}

```

```{r exercise-fit1-estimate-hint-1}
# Hint 1 of 3
# You need to do three things: specify the regression formula
# (outcome variable and predictors), save the model to an
# object named fit1, and display the results. Think about which
# variable is the outcome we are trying to explain and which
# four variables are the predictors listed in the variable
# descriptions above.
```

```{r exercise-fit1-estimate-hint-2}
# Hint 2 of 3
# Use lm() with hs_dem_prop_all as the outcome variable and
# nonwhite, p.liberal, povrate, and region_censusf as predictors.
# Save to fit1 and call summary(fit1).
```

```{r exercise-fit1-estimate-hint-3}
# Hint 3 of 3
# Your call should assign the model to fit1 using lm(), with
# hs_dem_prop_all as the outcome variable on the left of ~,
# and nonwhite, p.liberal, povrate, and region_censusf on the
# right separated by +, with data = states. Then call summary()
# on fit1 to display the results.
```

```{r exercise-fit1-estimate-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("lm\\s*\\(", code_nc))
    fail("Use lm() to estimate the model.")
  if (!grepl("fit1\\s*<-", code_nc))
    fail("Save your model as fit1.")
  if (!grepl("hs_dem_prop_all", code_nc))
    fail("hs_dem_prop_all should be the outcome variable.")
  if (!grepl("p\\.liberal", code_nc))
    fail("Include p.liberal as a predictor.")
  if (!grepl("nonwhite", code_nc))
    fail("Include nonwhite as a predictor.")
  if (!grepl("povrate", code_nc))
    fail("Include povrate as a predictor.")
  if (!grepl("region_censusf", code_nc))
    fail("Include region_censusf as a predictor.")
  if (!grepl("summary\\s*\\(", code_nc))
    fail("Call summary() on fit1 to display the results.")
  expected <- lm(hs_dem_prop_all ~ nonwhite + p.liberal + povrate + region_censusf,
                 data = states)
  result_coef <- if (inherits(.result, "summary.lm")) {
    setNames(.result$coefficients[, "Estimate"], rownames(.result$coefficients))
  } else {
    coef(.result)
  }
  if (is.null(result_coef))
    fail("Make sure you display the results by calling summary() on fit1.")
  if (!isTRUE(all.equal(
    sort(as.numeric(result_coef)),
    sort(as.numeric(coef(expected))),
    tolerance = 1e-6)))
    fail("Check your formula. The model should regress hs_dem_prop_all on nonwhite, p.liberal, povrate, and region_censusf using the states data frame.")
  pass("Right. Only two predictors are significant at .05: p.liberal and the Northeast dummy. Keep that in mind as you interpret the predictions that follow.")
})
```

### Present the Model in a Table

The `stargazer()` function from the [stargazer]{.package-name} package was introduced in Tutorial 16. The key arguments are summarized below.

::: {.tip}
**Key `stargazer()` arguments**

- **`type = "text"`** --- produces plain-text output readable in the learnr console; use `"latex"` or `"html"` for documents
- **`title =`** --- the table title, as a character string
- **`dep.var.labels =`** --- the label for the outcome-variable column header

- **`covariate.labels =`** --- a character vector of row labels for the predictors, one label per predictor in the order they appear in the model formula; the reference category of a factor variable is omitted from the output and should be omitted from the labels as well, so a three-category factor contributes two labels
:::

**Practice 2: Produce a stargazer table for fit1**

Use the title \"Predictors of Democratic State House Seat Share\" and the outcome-variable label \"Proportion of Democratic House Seats\". The six covariate labels for this model are \"Proportion Non-White\", \"Proportion Liberal\", \"Poverty Rate\", \"West\", \"Midwest\", and \"Northeast\". Determine the correct order from your model output above. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-stargazer-fit1, exercise=TRUE, exercise.cap="Practice 2"}
library(stargazer)
stargazer(fit1,
          ___ = "text",
          title = "___",
          dep.var.labels = "___",
          ___ = c("Proportion Non-White",
                  "___",
                  "___",
                  "___",
                  "___",
                  "___"))
```

```{r exercise-stargazer-fit1-hint-1}
# Hint 1 of 3
# A presentation table needs four things beyond the numbers: what
# format the output should be in, a title, a name for the outcome
# column, and a readable name for each predictor row. Two of the
# blanks are argument names; the rest is the text you supply.
```

```{r exercise-stargazer-fit1-hint-2}
# Hint 2 of 3
# The two argument name blanks are: type and covariate.labels.
# Covariate labels follow formula order; South is the reference
# category and is omitted, so region labels are West, Midwest, Northeast.
```

```{r exercise-stargazer-fit1-hint-3}
# Hint 3 of 3
# The six covariate labels follow formula order. The first three are
# Proportion Non-White, Proportion Liberal, and Poverty Rate.
# The last three are the non-reference region categories in the order
# they appear in the model output: West, Midwest, and Northeast.
# South is the reference category and is omitted from the labels.
```

```{r exercise-stargazer-fit1-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("stargazer\\s*\\(", code_nc))
    fail("Use stargazer() to produce the table.")
  if (!grepl("type\\s*=", code_nc))
    fail("Include the type = argument.")
  if (!grepl("covariate\\.labels\\s*=", code_nc))
    fail("Include the covariate.labels = argument.")
  if (!grepl("Predictors of Democratic State House Seat Share", code_nc, fixed = TRUE))
    fail("Use the prescribed title.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use the prescribed dep.var.labels.")

  labels <- c("Proportion Non-White","Proportion Liberal","Poverty Rate",
              "West", "Midwest", "Northeast")
  for (lbl in labels[-1])
    if (!grepl(lbl, code_nc, fixed = TRUE))
      fail(paste0("The label \"", lbl, "\" is missing."))
  pos <- sapply(labels, function(l) regexpr(l, code_nc, fixed=TRUE)[1])
  if (!all(diff(pos) > 0))
    fail("The labels are out of order. Follow formula order: Proportion Non-White, Proportion Liberal, Poverty Rate, West, Midwest, Northeast.")
  pass("Well done. The table gives your reader the complete picture; the predictions and plots that follow translate the coefficients into model-implied comparisons.")
})
```

Once a package is loaded with `library()`, all of its functions are available for the rest of your R session --- you do not need to load it again each time you use a function from it. Since [stargazer]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial.

### Obtain Specific Predicted Values

The hypothesis concerns liberalism. The most direct way to make the liberalism coefficient concrete is to compare the model's prediction for a state with a lower liberal proportion to one with a higher liberal proportion, holding everything else constant. The code below does exactly that.

Two things in the code below are worth reading before you run it. Interval predictors that are not being varied are held at their means, which is what `mean(..., na.rm = TRUE)` does. A factor predictor cannot be held at a mean, so it is set to one of its levels --- and it must be supplied as a factor carrying **all** of the original levels, not as a bare string, or `predict()` will not recognize it.

**Example code:**

```{r predict-factor-template, exercise=FALSE, eval=FALSE}
predict(yourmodel,
        newdata = data.frame(
          interval_predictor = mean(yourdata$interval_predictor, na.rm = TRUE),
          factor_predictor   = factor("level you want",
                                      levels = levels(yourdata$factor_predictor))
        ))
```

The `levels =` argument is what makes the factor complete. Without it, R builds a one-level factor and `predict()` fails.

**Run and observe 3: Compute predicted Democratic seat share at two levels of p.liberal**

```{r exercise-predict-fit1, exercise=TRUE, exercise.cap="Run and observe 3"}
# A Southern state at average nonwhite and poverty, with p.liberal = 0.25
predict(fit1,
        newdata = data.frame(
          p.liberal      = 0.25,
          nonwhite       = mean(states$nonwhite, na.rm = TRUE),
          povrate        = mean(states$povrate,  na.rm = TRUE),
          region_censusf = factor("South", levels = levels(states$region_censusf))
        ))

# Same characteristics, but p.liberal = 0.35
predict(fit1,
        newdata = data.frame(
          p.liberal      = 0.35,
          nonwhite       = mean(states$nonwhite, na.rm = TRUE),
          povrate        = mean(states$povrate,  na.rm = TRUE),
          region_censusf = factor("South", levels = levels(states$region_censusf))
        ))
```

**How this code works.** `predict()` takes the model object and a `newdata` argument --- a data frame describing the exact case you want to predict. Every predictor in the model must appear in `newdata`, named and typed exactly as in the formula. Here that means four variables: `p.liberal`, `nonwhite`, and `povrate` set to specific values, and `region_censusf`. Factor variables in R carry information about all their possible levels, not just the value you are supplying --- they are not simply character strings. The recommended way to specify a factor predictor in `newdata` is `factor("South", levels = levels(`states$region_censusf`))`, which creates a one-element factor with the correct level set explicitly attached. This makes the type unambiguous and protects against edge cases where R cannot determine the levels from context alone. The output is in the same units as the outcome variable: the proportion of Democratic house seats, on a 0-to-1 scale.

The two predictions differ in exactly one way --- the value of `p.liberal`. Everything else is held constant. That difference is the liberalism slope made concrete. A second useful comparison is the regional gap: the West coefficient is also significant, so comparing an otherwise identical Southern and Western state gives you a concrete statement of what that coefficient means.

**Apply 4: Interpret the difference**

```{r quiz-predict-fit1, echo=FALSE}
question("Look at both outputs. What does the predicted gap between **p.liberal** = 0.25 and **p.liberal** = 0.35 tell you?",
  correct = "Right. The two predictions express the p.liberal slope as a concrete model-implied comparison.",
  answer("All else equal, a state at 0.35 liberal is predicted to have about 9 percentage points more Democratic house seats than one at 0.25",
         correct = TRUE,
         message = "The liberalism slope is 0.876, so a 0.10 increase predicts about 0.088 more Democratic seats --- roughly 9 percentage points. This is the **p.liberal** coefficient expressed as a concrete predicted difference rather than an abstract number."),
  answer("The predictions tell us that 58% of voters in less liberal states and 67% of voters in more liberal states support Democrats",
         message = "The outcome variable is the proportion of state house seats held by Democrats, not the proportion of voters. And these are model predictions, not observed proportions."),
  answer("The gap proves that more liberal electorates cause Democratic victories in state legislative races",
         message = "A predicted difference from an observational regression is not evidence of causation. It is the model-implied difference in predicted outcomes across specified predictor values."),
  answer("The gap tells us that p.liberal = 0.25 and p.liberal = 0.35 are the only values at which we can make predictions",
         message = "These two values were chosen to illustrate the slope, but `predict()` can return a predicted value for any combination of predictor values you supply. The two specific values anchor your interpretation; `plot_model()` then shows the prediction at every value across the full range."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at both predicted values and compute the difference. What predictor changed between the two calls, and what stayed the same?"
)
```

### Visualize the Association with State Liberalism

The `predict()` calls answered one specific what-if at one value of **p.liberal**. The liberalism hypothesis calls for something richer: how does predicted Democratic seat share differ across the full observed range of **p.liberal**? `plot_model(type = "pred")` from the [sjPlot]{.package-name} package answers that. The key arguments:

- **`type = "pred"`** requests predicted values across a range, as opposed to `type = "est"` for a coefficient plot
- **`terms =`** names the predictor(s) to display. Variable names are always **character strings in quotes** --- unlike bare R variable names. Three forms:
    - `terms = "x"` --- one variable on the x-axis; all other predictors are held at their means (interval) or reference category (factor)
    - `terms = c("x", "grouping")` --- second variable produces separate lines; if a **factor**, one line per level colored and labeled by level name; if an **interval variable**, lines at selected values (default: mean minus 1 SD, mean, and mean plus 1 SD; override with bracket notation: `"variable [v1, v2]"`)
    - `terms = c("x", "grouping", "facet")` --- third variable produces separate panels, one per level or selected value
- **`axis.title = c("x label", "y label")`** labels the axes --- x-axis first, then y-axis (the reverse of what ggplot users sometimes expect)
- **`show.data = TRUE/FALSE`** overlays the raw observations; useful when N is small enough for individual points to be readable
- **`labs(caption = "...")`** adds a caption stating the held values; since `plot_model()` returns a ggplot object, any ggplot layer works with `+`. The caption tells the reader what covariate profile the predictions are for --- always list the held values in a complete sentence ending with a period, e.g., `"Predictions are for Southern states with average poverty rate and proportion non-white."`

Run the code below and examine the output --- the direction of the line, the shaded band, the individual points, and the caption --- before reading the explanation that follows.

**Run and observe 5: Produce a prediction plot for state liberalism**

```{r exercise-predplot-liberal-tryit, exercise=TRUE, exercise.cap="Run and observe 5"}
library(sjPlot)
plot_model(fit1,
           type = "pred",
           terms = "p.liberal",
           title = "Predicted Democratic House Seat Share by State Liberalism",
           axis.title = c("Proportion Liberal",
                          "Proportion of Democratic House Seats"),
           show.data = TRUE) +
  labs(caption = "Predictions are for Southern states with average poverty rate and proportion non-white.")
```

Since [sjPlot]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial. In your own scripts, call `library(sjPlot)` once at the top before using `plot_model()`. You may also see a note about overlapping data points --- with only 49 states this is not a concern, and the `jitter` argument is optional. Jitter adds a small random offset to each data point's position so that overlapping points spread apart and remain individually visible. With sparse data like these 49 states it is unnecessary, but in datasets where many observations share similar x-axis values, `show.data = TRUE` combined with `jitter = 0.1` (or a similar small value) prevents points from stacking on top of each other.

Notice what this plot is doing: it is sweeping across the full range of **p.liberal**, computing a predicted value at every point --- exactly what `predict()` did for one specific value, but repeated across the entire range and displayed as a continuous line. The two specific values you computed earlier (about 0.58 and 0.67) are points on this line. The shaded band is the 95% confidence region around those predictions. Its width is not constant: it is narrower in the middle of the **p.liberal** range, where more states are concentrated and the model has more information, and wider at the tails, where fewer states have very high or very low liberalism levels. Predictions are often more precise where data are dense and less precise where data are sparse, though precision also depends on the model's overall fit and the covariance among predictors.

### Interpret the Liberalism Plot

The five questions below walk through each component of the interpretation framework. The prediction plot is reproduced here for reference.

```{r fit1-predplot-ref, echo=FALSE, message=FALSE}
plot_model(fit1,
           type = "pred",
           terms = "p.liberal",
           title = "Predicted Democratic House Seat Share by State Liberalism",
           axis.title = c("Proportion Liberal",
                          "Proportion of Democratic House Seats"),
           show.data = FALSE) +
  labs(caption = "Predictions are for Southern states with average poverty rate and proportion non-white.")
```

**Apply 6: Component 1 --- The pattern**

Look at the direction and shape of the line before answering.

```{r quiz-liberal-pattern, echo=FALSE}
question("Look at the prediction plot above. Which statement best describes the overall pattern?",
  correct = "Right. The line slopes upward from left to right --- positive and exactly linear.",
  answer("As the proportion of liberal residents increases, the predicted proportion of Democratic house seats increases",
         correct = TRUE,
         message = "The upward slope matches the hypothesis. In a linear regression model, the prediction line is a straight line by construction --- the relationship is exactly, not approximately, linear."),
  answer("As the proportion of liberal residents increases, the predicted proportion of Democratic house seats decreases",
         message = "The line slopes upward from left to right --- a positive relationship."),
  answer("There is no clear pattern --- the line is essentially flat",
         message = "The line has a clearly positive slope from left to right."),
  answer("The relationship is positive but curved --- it rises steeply at low liberalism levels and flattens at higher levels",
         message = "The prediction line is straight, not curved. In a linear regression model, the line is exactly linear by construction --- the slope is constant across the full range of the predictor."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: trace the line from left to right --- does the y-axis value increase, decrease, or stay flat? And is the line straight or curved?"
)
```

**Apply 7: Component 2 --- The held values**

The caption tells you what was held constant. Think about what that means for how the slope generalizes.

```{r quiz-liberal-held, echo=FALSE}
question("The caption states that predictions are for Southern states with average poverty rate and proportion non-white. What does this mean for interpreting the slope?",
  correct = "Right. Changing the baseline shifts the line up or down but never changes its slope.",
  answer("The slope is the same regardless of which region or values of poverty and **nonwhite** we hold constant",
         correct = TRUE,
         message = "This is a key property of linear regression without interactions: the slope on **p.liberal** is the same at any baseline --- changing the held values shifts the line up or down without changing its steepness. The caption tells you which level of the line you are looking at, not which slope."),
  answer("The slope only applies to Southern states --- the liberalism association is completely different in other regions",
         message = "Because this model does not include an interaction between **p.liberal** and **region_censusf**, the slope is identical in every region. Regional differences shift the line up or down but do not change its slope."),
  answer("Changing the held values would change both the slope and the level of the line",
         message = "In an additive model without interactions involving **p.liberal**, only the level of the line changes when you change the held values --- not the slope. The slope on **p.liberal** is the same at any baseline."),
  answer("The slope only applies when poverty and proportion non-white are at their means --- at other values, the liberalism slope would be different",
         message = "The choice of held values for the other predictors does not affect the slope on **p.liberal** in an additive model. It determines which level of the line you are looking at, but the steepness is constant."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what an interaction term would be needed to make the slope differ across groups or baseline values."
)
```

**Apply 8: Component 3 --- Specific predicted values**

Read two values from the plot --- one at each end of a realistic range --- and compute the difference.

```{r quiz-liberal-values, echo=FALSE}
question("Using the prediction plot above, approximately what is the predicted Democratic seat share at **p.liberal** = 0.25 and at **p.liberal** = 0.35 for a Southern state at average poverty and **nonwhite** values?",
  correct = "Right. About 0.58 at p.liberal = 0.25 and about 0.67 at p.liberal = 0.35.",
  answer("About 0.58 at **p.liberal** = 0.25 and about 0.67 at **p.liberal** = 0.35",
         correct = TRUE,
         message = "The slope on **p.liberal** is 0.876, so a 0.10 increase predicts about 0.088 more Democratic seats --- roughly 9 percentage points. You could verify this by plugging these values into the regression equation by hand; the plot makes it visible at a glance."),
  answer("About 0.25 at **p.liberal** = 0.25 and about 0.35 at **p.liberal** = 0.35",
         message = "Read the y-axis value at each x position, not the x-axis value itself."),
  answer("About 0.50 at both values --- the line is flat",
         message = "The line is not flat. Trace from x = 0.25 up to the line and read the y-axis value; repeat at x = 0.35."),
  answer("About 0.40 at **p.liberal** = 0.25 and about 0.49 at **p.liberal** = 0.35 --- a 9-percentage-point increase",
         message = "The gap of 9 percentage points is correct, but the absolute values are not --- read the y-axis carefully at x = 0.25 and x = 0.35. The line sits higher on the y-axis than 0.40 at those liberalism levels."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find x = 0.25 on the horizontal axis, trace up to the line, and read the y-axis value. Repeat at x = 0.35."
)
```

**Apply 9: Component 4 --- The uncertainty**

```{r quiz-liberal-band, echo=FALSE}
question("Look at the shaded confidence band in the plot above. Which statement best describes what it tells you?",
  correct = "Right. The band is moderately wide and widens at the extremes.",
  answer("The band is moderately wide and widens at the extremes, but stays consistent with a positive slope",
         correct = TRUE,
         message = "With only 49 states, there is meaningful uncertainty around each prediction. The widening at the tails is expected. The band never crosses in a way that would reverse the directional conclusion."),
  answer("The band is so narrow that predictions can be read to three decimal places",
         message = "With 49 states, the band is visible and meaningful."),
  answer("The band is so wide that no conclusions can be drawn about direction",
         message = "The band limits precision but not directional conclusions. It is consistent with a positive slope throughout."),
  answer("The band is narrower at the tails because that is where the model has the most reliable predictions",
         message = "It is the opposite: the band widens at the tails, where fewer states have very high or very low liberalism levels. Predictions are more precise where data are dense --- in the middle of the range --- and less precise where data are sparse, at the extremes."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the width of the shaded band and whether it is wider at the tails than in the middle."
)
```

**Apply 10: Component 5 --- The hypothesis connection**

A coefficient and p-value tell you the relationship is statistically significant. The plot translates that coefficient into a substantively interpretable pattern across the observed range.

```{r quiz-liberal-hypothesis, echo=FALSE}
question("Look at the **p.liberal** coefficient and its p-value in your output. What does the prediction plot add to that conclusion?",
  correct = "Right. The plot adds substantive magnitude to what the p-value can only confirm as non-zero.",
  answer("The plot shows how large the difference is, not just that it is nonzero",
         correct = TRUE,
         message = "A significant p-value tells you the relationship is statistically distinguishable from zero. The plot tells you what it looks like in the units and range that matter --- about 9 percentage points between a less liberal and a more liberal state across a realistic range."),
  answer("The plot adds nothing the p-value has not already established",
         message = "The p-value tells you the relationship is unlikely to be chance. The plot tells you its magnitude across a realistic range --- substantively different information."),
  answer("The plot proves liberal voters cause Democratic victories",
         message = "A prediction plot is evidence of an association, not proof of causation."),
  answer("The plot shows the same information as the p-value, just in visual form",
         message = "The p-value tells you whether the relationship is statistically distinguishable from zero. The plot tells you how large the predicted difference is across a realistic range --- about 9 percentage points for a 10-point shift in the liberal proportion. Statistical significance and substantive magnitude are different things."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: think about what a p-value tells you and what it cannot tell you about the size or practical meaning of an effect."
)
```

### Visualize Predicted Differences across Census Regions

The `predict()` calls showed a regional gap. Seeing all four regions at once makes the comparison easier --- `plot_model()` shows the predicted Democratic seat share for each region as a point with a confidence interval, using the same held values.

**Practice 11: Produce a prediction plot for census region**

Produce the plot and examine it carefully before answering the interpretation question that follows. Use the title \"Predicted Democratic House Seat Share by Census Region\", the x-axis label \"Census Region\", the y-axis label \"Proportion of Democratic House Seats\", and the caption "Predictions are for states with average poverty rate, proportion liberal, and proportion non-white.\" Three hints are available. Click \"Submit Answer" to check your work.

```{r exercise-predplot-region, exercise=TRUE, exercise.cap="Practice 11"}
___(fit1,
    ___ = "pred",
    ___ = "region_censusf",
    title = "___",
    axis.title = c("___", "___")) +
  labs(caption = "___")
```

```{r exercise-predplot-region-hint-1}
# Hint 1 of 3
# A prediction plot sweeps one predictor across its range and shows
# what the model expects at each point along it. When the predictor
# that varies is a factor there is no range to sweep --- only a few
# separate categories, each with its own predicted value.
```

```{r exercise-predplot-region-hint-2}
# Hint 2 of 3
# The argument that controls what kind of plot you get is type =.
# For a prediction plot, its value is "pred".
# The argument that controls the x-axis variable is terms =.
# Set it to the name of the factor variable, in quotes.
# Prescribed axis labels: x = "Census Region", y = "Proportion of Democratic House Seats"
# Prescribed caption: "Predictions are for states with average poverty rate,
#   proportion liberal, and proportion non-white."
```

```{r exercise-predplot-region-hint-3}
# Hint 3 of 3
# Your call should be plot_model(fit1, type = "pred",
# terms = "region_censusf", with the prescribed title, axis labels,
# and a labs(caption = "...") layer. The prescribed caption is:
# "Predictions are for states with average poverty rate, proportion
# liberal, and proportion non-white."
```

```{r exercise-predplot-region-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("terms\\s*=", code_nc))
    fail("Name the argument that decides which variable goes on the x-axis.")
  if (!grepl("Census Region", code_nc, fixed = TRUE))
    fail("Use \"Census Region\" as the x-axis label.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use \"Proportion of Democratic House Seats\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = \"...\") layer stating the held values.")
  if (!grepl("Predictions are for states with average poverty rate, proportion liberal, and proportion non-white.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with average poverty rate, proportion liberal, and proportion non-white.\"")
  pass("Well done. The plot is built and captioned --- and the caption is what tells a reader which values everything else was held at.")
})
```

When `terms` is a factor, `plot_model()` produces a point with a confidence interval for each level rather than a continuous line --- there is no meaningful range to sweep across for a categorical variable, only discrete levels. Overlapping confidence intervals show that the predictions are estimated with substantial uncertainty; they do not by themselves determine whether a contrast is statistically significant --- that requires the relevant coefficient test.

**Apply 12: Interpret the region plot**

For each region, `plot_model()` displays a point showing the predicted seat share with a confidence interval around it.

```{r quiz-region-interpret, echo=FALSE}
question("Look at the regional prediction plot. Which statement best describes what the predicted values and confidence intervals show?",
  correct = "Right. The Northeast is highest, but overlapping intervals show substantial uncertainty --- significance requires the coefficient test.",
  answer("The predicted Democratic seat share is highest for Northeastern states, but the confidence intervals overlap across all four regions --- the intervals show substantial uncertainty around each prediction",
         correct = TRUE,
         message = "The Northeast has the highest predicted seat share, but the confidence intervals around the separate predictions overlap. That overlap shows the predictions are estimated with considerable uncertainty; it does not by itself determine whether the Northeast-South contrast is statistically significant. For that conclusion, consult the Northeast coefficient and its hypothesis test."),
  answer("The predicted Democratic seat share is nearly identical across all four regions, with overlapping confidence intervals suggesting the predicted values do not vary meaningfully across regions",
         message = "The predicted values are not identical --- the Northeast is clearly higher than the other three. The overlapping intervals limit what conclusions you can draw about significance, but the predicted values themselves do differ across regions."),
  answer("The predicted Democratic seat share is highest for the Northeast and second highest for the West, and the West confidence interval does not overlap with South --- the two expected outcomes show clear visual separation",
         message = "The West is the LOWEST of the four, at about 0.53, just below South and Midwest --- and its coefficient is nowhere near significant. Regional intuition is not a substitute for reading the plot."),
  answer("The predicted Democratic seat share for Northeastern states has a narrow confidence interval that does not overlap with any other region, making it the only prediction with clear visual separation from the others",
         message = "The Northeast confidence interval is not narrow --- it is wide and does overlap with the other regions. Wide intervals with overlap are the visual signature of uncertainty about the size of a difference."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at where each region's predicted value falls on the y-axis and whether the confidence intervals overlap with each other."
)
```

### Visualize Two Predictors at Once

The liberalism and region plots each varied one predictor. A two-variable plot extends this: it shows whether predicted Democratic seat share is systematically higher or lower in some regions than others after accounting for liberalism, and whether the estimated association of liberalism with seat share is the same size across regions.

Setting `terms` to a vector of two variable names produces this plot: the first variable goes on the x-axis, and the second variable's levels appear as separate colored lines.

**Apply 13: Produce a two-variable prediction plot for liberalism and region**

Produce the plot and notice whether the four lines are parallel --- that parallelism is a finding about the model, not just a visual feature. Use the title \"Predicted Democratic House Seat Share by Liberalism and Region\", the x-axis label \"Proportion Liberal\", the y-axis label \"Proportion of Democratic House Seats\", and the caption \"Predictions are for states with average poverty rate and proportion non-white.\" `plot_model()` maps the grouping variable to the color aesthetic, so without a `color =` argument in `labs()`, the variable name appears as the legend title. Use `labs(color = "Census Region", caption = "...")` to set both. Four hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-twovar, exercise=TRUE, exercise.cap="Apply 13"}
___(fit1,
    ___ = "pred",
    ___ = c("p.liberal", "region_censusf"),
    title = "___",
    ___ = c("___", "___")) +
  labs(color = "___", caption = "___")
```

```{r exercise-predplot-twovar-hint-1}
# Hint 1 of 4
# You have produced prediction plots for one variable at a time.
# This plot varies two variables simultaneously. Think about which
# variable should be on the x-axis (the continuous predictor your
# hypothesis is about) and which should produce separate colored
# lines (the grouping variable). The only change in the code
# structure is how the terms argument is specified.
```

```{r exercise-predplot-twovar-hint-2}
# Hint 2 of 4
# The blank argument names are: the function name (plot_model),
# type =, terms =, and axis.title =.
# terms takes a character vector of two variable names.
# In labs(), color = sets the legend title and caption = sets the caption.
```

```{r exercise-predplot-twovar-hint-3}
# Hint 3 of 4
# Prescribed title: "Predicted Democratic House Seat Share by Liberalism and Region"
# Prescribed axis labels: x = "Proportion Liberal", y = "Proportion of Democratic House Seats"
# Prescribed legend title (color =): "Census Region"
# Prescribed caption: "Predictions are for states with average poverty rate and proportion non-white."
```

```{r exercise-predplot-twovar-hint-4}
# Hint 4 of 4
# Your call should be plot_model(fit1, type = "pred",
# terms = c("p.liberal", "region_censusf"), with the prescribed
# title, axis labels, and a labs(color = "Census Region", caption = "...")
# layer with the prescribed caption.
```

```{r exercise-predplot-twovar-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("terms\\s*=", code_nc))
    fail("Name the argument that takes the two variable names.")
  if (!grepl("Predicted Democratic House Seat Share by Liberalism and Region", code_nc, fixed = TRUE))
    fail("Use the prescribed title: \"Predicted Democratic House Seat Share by Liberalism and Region\".")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Proportion Liberal", code_nc, fixed = TRUE))
    fail("Use \"Proportion Liberal\" as the x-axis label.")
  if (!grepl("Proportion of Democratic House Seats", code_nc, fixed = TRUE))
    fail("Use \"Proportion of Democratic House Seats\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs() layer with color = and caption = arguments.")
  if (!grepl("Census Region", code_nc, fixed = TRUE))
    fail("Use labs(color = \"Census Region\") to set the legend title.")
  if (!grepl("Predictions are for states with average poverty rate and proportion non-white.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for states with average poverty rate and proportion non-white.\"")
  pass("Well done. Passing two names to `terms` puts the first on the x-axis and draws one line for each level of the second.")
})
```

::: {.caution}
**The y-axis runs past 1.0, and the outcome cannot.** **hs_dem_prop_all** is a proportion, so it is bounded at 0 and 1 --- yet the Northeast line approaches 1.0 at the top of the liberalism range and its confidence band goes beyond it. Nothing has gone wrong with your code. A linear model fits a straight line and knows nothing about the bounds of the variable it is predicting, so extended far enough that line will leave the range the outcome can actually occupy. Two things follow. Predictions near the edge of the observed data are the least trustworthy part of any plot, and predictions outside the outcome's possible range are not predictions at all. And when an outcome is bounded --- a proportion, or a yes/no --- a method that respects the bound is the better tool. That is what Tutorial 20 introduces.
:::

**Apply 14: Interpret the liberalism-by-region plot**

```{r quiz-twovar-interpret, echo=FALSE}
question("Look at the two-variable plot. Which statement best describes what the predicted values and confidence bands show about regional differences?",
  correct = "Right. The Northeast stands clearly apart; the other three regions are nearly indistinguishable in the plot.",
  answer("The Northeast line is clearly above the other three --- at a liberal proportion of 0.20, the model predicts about 0.67 for Northeastern states and about 0.53 for the other three, holding poverty rate and proportion non-white at their means.",
         correct = TRUE,
         message = "The vertical gap between the Northeast line and the others is visible across the full range of liberalism --- that gap is the Northeast coefficient, 0.128, and it is the only regional difference the model distinguishes from zero. The South, West and Midwest lines cluster so closely, with heavily overlapping bands, that the plot cannot separate them."),
  answer("The Northeast line is clearly above the other three --- at a liberal proportion of 0.20, the model predicts about 0.60 for Northeastern states and about 0.40 for the other three regions, holding poverty rate and proportion non-white at their means. The South, West, and Midwest lines cluster closely together with overlapping bands.",
         message = "The Northeast advantage is real, but read the y-axis more carefully at p.liberal = 0.20. The other three lines sit just above 0.53, not near 0.40, and the Northeast is closer to 0.67 than 0.60."),
  answer("The Northeast line is clearly above the other three, and the West line is clearly above the South and Midwest --- at a liberal proportion of 0.20, the West is predicted to have meaningfully higher Democratic seat share than Southern states, with confidence intervals that do not overlap.",
         message = "The West is the lowest of the four, marginally below South and Midwest, and all three track almost identically across the full range with heavily overlapping bands. Reputation is not what the plot shows."),
  answer("All four regional lines are clustered closely together with heavily overlapping confidence bands across the full range of liberalism --- the plot suggests the model cannot distinguish predicted Democratic seat share across any of the four regions.",
         message = "The Northeast line is clearly separated from the other three. Look at the vertical gap between the Northeast and South lines at any given liberalism level --- that gap is visible throughout the plot."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find the highest line and read its predicted value at p.liberal = 0.20. Then look at where the other three fall at the same point and how closely their confidence bands overlap."
)
```

The four lines are parallel. That is not a coincidence --- it is what a model without an interaction produces. Because there is no interaction term between **p.liberal** and **region_censusf** in this model, the slope of liberalism is mathematically identical in every region. The vertical distances between lines are the region coefficients: the Northeast line runs consistently above the others, and South, West, and Midwest cluster closely together. Parallel lines tell you that the x-axis predictor has the same estimated slope in every group.

### What We Learned About Ideology, Region, and Democratic Representation

The model confirms the hypothesis: states with more liberal public opinion do have higher proportions of Democratic state house seats, even after controlling for racial composition, poverty, and region. A 10-percentage-point increase in the liberal share of the electorate predicts about 0.09 more in the proportion of Democratic seats --- a substantive association, visible in the slope of the prediction plot, and significant (p = .013).

The regional finding is worth dwelling on. The Northeast stands apart from the other three even after controlling for ideology, and it is the only region whose coefficient the model distinguishes from zero (0.128, p = .047). The West and Midwest are not significantly different from the South --- and the West is in fact predicted lowest of the four, marginally below the South, despite a reputation that would suggest otherwise. What makes the Northeast distinct here is not just a more liberal electorate, since the gap persists with liberalism held constant.

A complete interpretation of this plot reads like this:

::: {.research-example}
*As the proportion of liberal state residents increases, the predicted proportion of Democratic house seats rises in a straight line across the full observed range of the variable. Predictions are for Southern states with average poverty rate and proportion non-white; a Northeastern state at the same liberalism level would be predicted to have about 0.13 higher proportion of Democratic seats, but the liberalism slope would be identical. At a liberal proportion of 0.25, the model predicts about 0.58; at 0.35, about 0.67 --- a difference of about 0.09 for a 10-point shift in the liberal electorate. The confidence band is moderately wide, narrower in the middle of the range where most states are concentrated, and widens at the extremes. This is consistent with the hypothesis that public ideology shapes legislative representation, even after controlling for racial composition, poverty, and region.*
:::

## What Do Electoral Rules Predict for Women's Representation?

In Tutorial 16 you estimated a model asking whether proportional electoral systems are associated with higher women's representation, controlling for gender quotas and how long women have served in parliament. That model is available here as **model_rep**. In Tutorial 16 the focus was on the coefficients --- their direction, size, and significance. Here the focus is on what those coefficients predict for specific country profiles, and on visualizing the relationships across the full range of each predictor.

The data frame **ipu** contains one row for each of 193 national legislatures --- the lower house where a parliament is bicameral, the single chamber where it is not. Chamber composition is current as of August 2026, when these data were collected from the Inter-Parliamentary Union. Of the 193 legislatures, **electoral_system** is missing for 15, **women_percent** for 8, **gender_quota** for 1, and **years_since_first_woman** for none. The variables in this model are:

- **women_percent**: percentage of seats held by women (0--100)
- **quota_f**: a two-category factor --- **No quota** (reference category) and **Quota**
- **electoral_f**: a three-category factor --- **Plurality/Majority** (reference category), **Proportional**, and **Mixed**
- **years_c**: years since the first woman entered that parliament, centered at its mean of 60.3 years

The theoretical argument is institutional. Proportional systems let a party balance its list, putting women in winnable positions without displacing anyone in particular; plurality systems ask the same party to nominate a woman in a specific district against a specific opponent. Gender quotas and the length of women's parliamentary history are included as controls because both are plausibly related to electoral rules and to representation.

::: {.hypothesis}
In a comparison of countries, those using proportional electoral systems will have a higher percentage of women in the national legislature than those using plurality electoral systems, controlling for electoral gender quotas and how long women have served in that parliament.
:::

### Present the Model in a Table

The `summary()` output below shows the model coefficients and their order --- you will need both to produce a correctly labeled table. Read it before writing your `stargazer()` call.

```{r model-rep-summary, echo=FALSE, comment=NA}
summary(model_rep)
```

**Practice 15: Produce a labeled `stargazer()` table for the representation model**

Write the complete `stargazer()` call from scratch, passing **model_rep** as the model object. Use the title \"Electoral Rules, Quotas, History, and Women's Representation\" and the outcome-variable label \"Percent Women\". The four covariate labels are \"Gender Quota\", \"Proportional System\", \"Mixed System\", and \"Years Since First Woman\" --- determine their order from the `summary()` output above. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-stargazer-rep, exercise=TRUE, exercise.cap="Practice 15"}

```

```{r exercise-stargazer-rep-hint-1}
# Hint 1 of 3
# The same table as before, for a different model. You supply the
# format, a title, a name for the outcome column, and one readable
# label per predictor row --- and a factor contributes one row for
# each category beyond its reference, not one row in total.
```

```{r exercise-stargazer-rep-hint-2}
# Hint 2 of 3
# The function name is stargazer(). The model object is model_rep.
# Set type = "text". The title and dep.var.labels are given in the
# exercise prompt. covariate.labels takes a character vector of
# four labels in formula order: the quota indicator first, then the
# two non-reference levels of electoral_f, then years_c.
```

```{r exercise-stargazer-rep-hint-3}
# Hint 3 of 3
# Your call is stargazer(model_rep, type = "text", followed by
# the title and dep.var.labels from the exercise prompt, then
# covariate.labels = c(...) with four labels in formula order:
# "Gender Quota", "Proportional System", "Mixed System",
# "Years Since First Woman".
```

```{r exercise-stargazer-rep-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("stargazer\\s*\\(", code_nc))
    fail("Use stargazer() to produce the table.")
  if (!grepl("model_rep", code_nc))
    fail("Pass model_rep as the model object.")
  if (!grepl("type\\s*=\\s*['\"]text['\"]", code_nc))
    fail("Set type = \"text\".")
  if (!grepl("Electoral Rules, Quotas, History, and Women's Representation", code_nc, fixed = TRUE))
    fail("Use the prescribed title: \"Electoral Rules, Quotas, History, and Women's Representation\".")
  if (!grepl("Percent Women", code_nc, fixed = TRUE))
    fail("Use the prescribed dep.var.labels: \"Percent Women\".")
  if (!grepl("covariate\\.labels\\s*=", code_nc))
    fail("Include a covariate.labels = argument.")
  labels <- c("Gender Quota", "Proportional System", "Mixed System", "Years Since First Woman")
  for (lbl in labels)
    if (!grepl(lbl, code_nc, fixed = TRUE))
      fail(paste0("The label \"", lbl, "\" is missing."))
  cov_match <- regexpr("covariate\\.labels\\s*=.*", code_nc)
  cov_str   <- if (cov_match > 0) substring(code_nc, cov_match) else code_nc
  pos_quota <- regexpr("Gender Quota",            cov_str, fixed = TRUE)[1]
  pos_prop  <- regexpr("Proportional System",     cov_str, fixed = TRUE)[1]
  pos_mixed <- regexpr("Mixed System",            cov_str, fixed = TRUE)[1]
  pos_years <- regexpr("Years Since First Woman", cov_str, fixed = TRUE)[1]
  if (!all(diff(c(pos_quota, pos_prop, pos_mixed, pos_years)) > 0))
    fail("The labels are out of order. They should follow formula order: Gender Quota, Proportional System, Mixed System, Years Since First Woman.")
  pass("Right. The coefficient values and significance levels in this table will anchor your interpretation of the predictions that follow.")
})
```

### Obtain Specific Predicted Values

The Proportional coefficient tells you the direction and size of the electoral-system gap in abstract units. A predicted value makes it concrete: what percentage of women does the model predict for a plurality country with no quota and an average length of parliamentary history, and how much higher is the predicted percentage for an otherwise identical proportional country?

**Apply 16: Compute predicted percentages for two electoral systems**

Fill in the blanks to complete the first `predict()` call for a plurality country, then write the second call for a proportional country from scratch. For both calls, set **quota_f** to \"No quota\" and **years_c** to 0, which is the average length of history because the variable is centered. Remember that both **quota_f** and **electoral_f** are factors. The recommended approach is to use `factor("level", levels = levels(ipu$electoral_f))` so that the factor levels are specified explicitly. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predict-rep, exercise=TRUE, exercise.cap="Apply 16"}
# Call 1: Plurality/Majority, no quota, average history
predict(model_rep,
        newdata = data.frame(
          quota_f     = factor("___", levels = levels(ipu$quota_f)),
          electoral_f = factor("___", levels = levels(ipu$electoral_f)),
          years_c     = ___
        ))

# Call 2: Proportional --- write from scratch

```

```{r exercise-predict-rep-hint-1}
# Hint 1 of 3
# A predicted value answers a what-if: for a country with this
# particular combination of characteristics, what does the model
# expect? So the country has to be described completely --- every
# predictor in the model needs a value, including the ones you are
# holding constant rather than varying.
```

```{r exercise-predict-rep-hint-2}
# Hint 2 of 3
# For each blank in Call 1: quota_f takes the level name for a
# country without a quota, electoral_f takes the level name for
# plurality systems, and years_c takes the value that puts a
# centred variable at its mean.
```

```{r exercise-predict-rep-hint-3}
# Hint 3 of 3
# Call 2 is a second predict() call with the same three variables
# as Call 1. Only the electoral_f level changes --- to the
# proportional one. quota_f and years_c stay where they were.
```

```{r exercise-predict-rep-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("predict\\s*\\(", code_nc))
    fail("Use predict() to obtain predicted values.")
  if (!grepl("model_rep", code_nc))
    fail("Pass model_rep as the model object.")
  if (!grepl("newdata", code_nc))
    fail("Include a newdata = argument specifying the predictor values.")
  if (!grepl("quota_f", code_nc))
    fail("Include quota_f in your newdata data frame.")
  if (!grepl("electoral_f", code_nc))
    fail("Include electoral_f in your newdata data frame.")
  if (!grepl("years_c", code_nc))
    fail("Include years_c in your newdata data frame.")
  if (!grepl("Plurality/Majority", code_nc, fixed = TRUE))
    fail("Include a prediction for plurality/majority countries.")
  if (!grepl("Proportional", code_nc, fixed = TRUE))
    fail("Include a prediction for proportional countries.")
  if (!grepl("No quota", code_nc, fixed = TRUE))
    fail("Set quota_f to \"No quota\" in both calls.")
  n_calls <- lengths(regmatches(code_nc, gregexpr("predict\\s*\\(", code_nc)))
  if (n_calls < 2)
    fail("Write two predict() calls: one for a plurality system, one for a proportional system.")
  pass("Right. Two `predict()` calls identical except for the electoral system isolate exactly the comparison the hypothesis is about.")
})
```

**Apply 17: Interpret the predicted gap**

```{r quiz-predict-rep-interpret, echo=FALSE}
question("Your two predicted values show the representation gap between a plurality country and a proportional one, both with no quota and an average length of parliamentary history. Which statement best describes what this gap tells you?",
  correct = "Right. The predicted gap translates the Proportional coefficient into concrete units.",
  answer("With no quota and average history, the model predicts proportional systems to seat about 7.8 percentage points more women than plurality systems",
         correct = TRUE,
         message = "The Proportional coefficient of 7.766 is the expected gap between proportional and plurality systems holding the other variables constant. The two predict() calls anchor this comparison: about 17.8% for plurality and about 25.6% for proportional, a gap of about 7.8 percentage points in the units of the outcome."),
  answer("The gap proves that proportional representation causes countries to elect more women, since the model controls for quotas and history",
         message = "Controlling for other variables reduces confounding but does not establish causation. Countries do not adopt electoral systems at random, and whatever produced a proportional system may also have produced a politics more open to women candidates."),
  answer("The gap of about 7.8 points simply restates the Proportional coefficient, so the predict() calls added no new information beyond what was already in the table",
         message = "The gap does match the coefficient --- that is the point. But the predict() calls express it in the units of the outcome for a specific, realistic country profile, which is more interpretable than an abstract coefficient. The coefficient tells you the direction and size; the predicted values tell you what that means in practice."),
  answer("The predictions apply to any country, since setting years_c to zero removes the effect of parliamentary history from the calculation",
         message = "Setting a centered variable to zero does not remove it --- it holds it at its mean, about 60 years. The predictions describe a country with an average length of women's parliamentary history, not one with none."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the Proportional coefficient gives the expected gap between proportional and plurality systems. How does the difference between your two predicted values compare to that coefficient?"
)
```

### Visualize the Association with Parliamentary History

`predict()` answered one specific what-if. `plot_model()` shows how predicted representation differs across the full range of **years_c** --- the same tool you used for **p.liberal** in Example 1, now applied to a different model and predictor.

A `plot_model()` call has four parts, and they always appear in the same order. The model object comes first. `type = "pred"` asks for predicted values rather than coefficients. `terms` names the predictor to sweep across the x-axis --- one name for a single line, or a vector of two when a second variable should produce separate lines. Everything after that is labelling: `title` and `axis.title` belong inside the call, while the legend title and caption are added afterwards with a `labs()` layer joined by `+`.

**Example code:**

```{r plot-model-template, exercise=FALSE, eval=FALSE}
plot_model(yourmodel,
           type = "pred",
           terms = "predictor_for_the_x_axis",
           title = "What the plot shows",
           axis.title = c("x-axis label", "y-axis label")) +
  labs(caption = "What the other predictors were held at.")
```

`axis.title` takes both labels as a vector, x first. The caption is not decoration: it is the only place a reader learns which values everything else was held at, and a prediction plot without one cannot be interpreted.

**Practice 18: Produce a prediction plot for parliamentary history**

Write the complete `plot_model()` call from scratch, using **model_rep** as the model object. Use the title \"Predicted Women's Representation by Length of Parliamentary History\", the x-axis label \"Years Since First Woman (centered)\", the y-axis label \"Predicted Percent Women\", and the caption \"Predictions are for plurality/majority countries with no gender quota.\" Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-years, exercise=TRUE, exercise.cap="Practice 18"}

```

```{r exercise-predplot-years-hint-1}
# Hint 1 of 3
# A prediction plot shows what the model expects across the whole
# observed range of one predictor, with everything else fixed. Two
# decisions follow from that: which predictor varies, and what the
# reader is told about the values the others were held at.
```

```{r exercise-predplot-years-hint-2}
# Hint 2 of 3
# The function is plot_model(). The model object is model_rep.
# Set type = "pred" and terms = "years_c".
# axis.title takes a character vector: c("x label", "y label").
# Chain labs(caption = "...") after the plot_model() call with +.
```

```{r exercise-predplot-years-hint-3}
# Hint 3 of 3
# Your call should be plot_model(model_rep, type = "pred",
# terms = "years_c", with title = "Predicted Women's Representation
# by Length of Parliamentary History", axis.title = c("Years Since
# First Woman (centered)", "Predicted Percent Women"), and a
# labs(caption = "...") layer. The prescribed caption is:
# "Predictions are for plurality/majority countries with no gender quota."
```

```{r exercise-predplot-years-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("years_c", code_nc))
    fail("Set terms = \"years_c\" to put parliamentary history on the x-axis.")
  if (!grepl("model_rep", code_nc))
    fail("Pass model_rep as the model object.")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Years Since First Woman \\(centered\\)", code_nc))
    fail("Use \"Years Since First Woman (centered)\" as the x-axis label.")
  if (!grepl("Predicted Percent Women", code_nc, fixed = TRUE))
    fail("Use \"Predicted Percent Women\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = \"...\") layer stating the held values.")
  if (!grepl("Predictions are for plurality/majority countries with no gender quota.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for plurality/majority countries with no gender quota.\"")
  pass("Well done. The upward slope confirms that a longer history of women in parliament predicts higher representation today --- but note how gentle it is next to the electoral-system gaps you calculated above.")
})
```

**Apply 19: Interpret the parliamentary history plot**

```{r quiz-predplot-years-interpret, echo=FALSE}
question_checkbox(
  "Which of the following statements should be included in a complete interpretation of the parliamentary history prediction plot? Select all that apply.",
  answer("As the length of women's parliamentary history increases, predicted representation rises in a straight line across the observed range",
         correct = TRUE,
         message = "This is the pattern component --- direction and shape of the relationship."),
  answer("Predictions are for plurality/majority countries with no gender quota",
         correct = TRUE,
         message = "This is the held values component --- essential for a complete interpretation."),
  answer("At the low end of the range, about 55 years below the mean, predicted representation is roughly 12.6%; at the high end, about 59 years above it, roughly 23.3% --- a difference of about 11 percentage points across the observed range",
         correct = TRUE,
         message = "This is the specific predicted values component --- magnitude makes the pattern concrete. Note that this 11-point spread takes the full 114-year range of the variable, while switching from a plurality to a proportional system moves the prediction 7.8 points on its own."),
  answer("The confidence band widens at the extremes, reflecting fewer countries with very short or very long histories of women in parliament",
         correct = TRUE,
         message = "This is the uncertainty component --- the band's behavior tells you where predictions are more or less precise."),
  answer("The upward slope is consistent with the argument that representation accumulates over time, even after controlling for electoral system and quotas",
         correct = TRUE,
         message = "This is the hypothesis connection component --- linking the visual back to the research question."),
  answer("The slope would be steeper for proportional systems than for plurality systems, because proportional systems elect more women overall",
         correct = FALSE,
         message = "In an additive model, the slope on years_c is the same regardless of electoral system. Changing the held value from Plurality to Proportional shifts the line up but does not change its steepness."),
  answer("Because the x-axis is centered, a value of zero means a country with no history of women in parliament",
         correct = FALSE,
         message = "Zero on a centered variable is the mean, not the absence of the thing measured. Here zero means a parliament that seated its first woman about 60 years ago."),
  correct = "Right. All five components are present.",
  try_again = "Not quite. Five of the seven statements belong in a complete interpretation --- check which components are missing and which claims the plot does not support.",
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

### Visualize Predicted Differences across Electoral Systems

The two `predict()` calls showed the representation gap between a plurality and a proportional country at no quota and average history. But the model has three electoral-system categories --- where does Mixed fall, and how precisely can we place all three? `plot_model()` answers this by showing the predicted percentage for each system alongside its confidence interval.

**Practice 20: Produce a prediction plot for electoral system**

Write the complete `plot_model()` call from scratch, using **model_rep** as the model object. Use the title \"Predicted Women's Representation by Electoral System\", the x-axis label \"Electoral System\", the y-axis label \"Predicted Percent Women\", and the caption "Predictions are for countries with no gender quota and average parliamentary history.\" Three hints are available. Click \"Submit Answer" to check your work.

```{r exercise-predplot-electoral, exercise=TRUE, exercise.cap="Practice 20"}

```

```{r exercise-predplot-electoral-hint-1}
# Hint 1 of 3
# The same idea as the previous plot, except that the predictor
# being varied is categorical. A category has no range to sweep
# across, so what the plot shows is one predicted value per
# category with its uncertainty, rather than a line.
```

```{r exercise-predplot-electoral-hint-2}
# Hint 2 of 3
# Set terms = "electoral_f". Everything else follows the same
# structure as the parliamentary history plot: type = "pred",
# axis.title = c(...), and a labs(caption = ...) layer.
# Prescribed x-axis label: "Electoral System"
# Prescribed y-axis label: "Predicted Percent Women"
# Prescribed caption: "Predictions are for countries with no
#   gender quota and average parliamentary history."
```

```{r exercise-predplot-electoral-hint-3}
# Hint 3 of 3
# Your call should be plot_model(model_rep, type = "pred",
# terms = "electoral_f", with title = "Predicted Women's
# Representation by Electoral System", axis.title = c("Electoral
# System", "Predicted Percent Women"), and a labs(caption = "...")
# layer. The prescribed caption is: "Predictions are for countries
# with no gender quota and average parliamentary history."
```

```{r exercise-predplot-electoral-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("electoral_f", code_nc))
    fail("Set terms = \"electoral_f\" to put electoral system on the x-axis.")
  if (!grepl("model_rep", code_nc))
    fail("Pass model_rep as the model object.")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Electoral System", code_nc, fixed = TRUE))
    fail("Use \"Electoral System\" as the x-axis label.")
  if (!grepl("Predicted Percent Women", code_nc, fixed = TRUE))
    fail("Use \"Predicted Percent Women\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs(caption = \"...\") layer stating the held values.")
  if (!grepl("Predictions are for countries with no gender quota and average parliamentary history.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for countries with no gender quota and average parliamentary history.\"")
  pass("Well done. When terms is a factor, plot_model() produces a point with a confidence interval for each level rather than a continuous line --- there is no meaningful range to sweep across for a categorical variable, only discrete levels to compare.")
})
```

**Apply 21: Interpret the electoral system plot**

```{r quiz-predplot-electoral-interpret, echo=FALSE}
question("Look at the electoral system prediction plot. Which statement best describes the key finding?",
  correct = "Right. Proportional's interval clears plurality's, while mixed overlaps both --- and overlap is not the same as non-significance.",
  answer("Proportional systems have the highest predicted representation, and their interval does not overlap plurality's. Mixed falls between, overlapping both.",
         correct = TRUE,
         message = "Proportional sits at about 25.7% with an interval running from about 21.9 to 29.5, clear of plurality's 15.0 to 20.9. Mixed sits at about 23.5% with an interval from about 18.9 to 28.0, which overlaps both of the others."),
  answer("All three electoral systems are predicted to have similar representation --- the intervals overlap throughout, so the model does not distinguish any of them from the others.",
         message = "Look again at plurality and proportional. Plurality's interval ends at about 20.9 and proportional's begins at about 21.9, so those two do not overlap. It is mixed that overlaps both."),
  answer("Because the mixed-system interval overlaps the plurality interval, the coefficient on mixed systems must not be statistically significant.",
         message = "This is the trap. The Mixed coefficient is significant at p = .026 in the table, even though the two predicted intervals overlap. Overlapping confidence intervals on two predicted values do not imply a non-significant difference between them --- the test of the difference is a different quantity with a smaller standard error than either prediction. The coefficient test is the formal evidence."),
  answer("Mixed systems have the widest confidence interval because they have the largest coefficient standard error, which means the estimate is unreliable.",
         message = "Mixed does have the widest interval, and that reflects having the fewest countries of the three --- 36 against 65 and 77. Wide is not the same as unreliable; it means the prediction is less precise, which is exactly what the interval is there to show."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: check which pairs of intervals actually overlap, then compare what you see with the p-values in the table."
)
```

### Choosing Held Values: The `condition` Argument

By default, `plot_model()` holds all predictors not on the x-axis at their means (for interval variables) or reference category (for factor variables). The parliamentary history plot above used No quota as the reference category for **quota_f** --- that is what the caption says. But what if the research question is specifically about countries that have adopted a quota? The predicted level of representation will differ when the model is evaluated for a quota country, even though the history slope remains the same in this additive model.

The `condition` argument lets you override the defaults explicitly:

```r
plot_model(model_rep, type = "pred", terms = "years_c",
           condition = list(quota_f = "Quota"))
```

`condition` takes a **named list**: `condition = list(variable_name = value)`. Note the contrast with `terms =`: in `terms`, variable names are **quoted strings** (e.g., `"quota_f"`); in `condition`, variable names are **unquoted list element names** (e.g., `quota_f = ...`). The value specifies what to hold that variable at:

- For an **interval variable**: supply a number, e.g., `condition = list(years_c = 0)` or `condition = list(years_c = 20)`
- For a **factor variable**: supply a character string in quotes matching a factor level exactly, e.g., `condition = list(quota_f = "Quota")`
- For **multiple variables**: `condition = list(var1 = value1, var2 = value2)`

::: {.tip}
**Choose held values deliberately.** The defaults (means for interval variables, reference category for factor variables) provide a standardized comparison case, but they are not necessarily substantively typical of any real observation. If you are studying how representation accumulates in countries that have adopted quotas, conditioning on No quota produces a plot that answers a different question. The caption is your contract with the reader --- it must accurately state what the plot is showing, and the values you choose should reflect the context the research question concerns.
:::

**Run and observe 22: Parliamentary history in quota countries**

The plot below shows predicted representation across the range of **years_c**, this time conditioning on countries with a gender quota rather than the default No quota. Run the code and compare the line to the no-quota version you produced above before answering.

```{r exercise-predplot-years-quota, exercise=TRUE, exercise.cap="Run and observe 22"}
plot_model(model_rep,
           type = "pred",
           terms = "years_c",
           condition = list(quota_f = "Quota"),
           title = "Predicted Women's Representation by Length of Parliamentary History",
           axis.title = c("Years Since First Woman (centered)",
                          "Predicted Percent Women")) +
  labs(caption = "Predictions are for plurality/majority countries with a gender quota.")
```

**Apply 23: Interpret the shift**

```{r quiz-condition-shift-rep, echo=FALSE}
question("Compare the parliamentary history plot for quota countries to the one for no-quota countries you produced earlier. Which statement best describes what changes and what stays the same?",
  correct = "Right. The slope is the same; the level shifts.",
  answer("The predicted representation line is higher throughout for quota countries, but the slope of the history relationship is the same",
         correct = TRUE,
         message = "This is the key property of an additive model: the slope on years_c is the same regardless of which quota status you condition on. What changes is the intercept --- quota countries start higher and end higher, but the rate of increase is identical. The vertical gap between the two lines is the Quota coefficient, about 7.3 percentage points."),
  answer("The slope is steeper for quota countries, because a quota accelerates the accumulation of women's representation over time",
         message = "In an additive model, the slope on years_c is the same in every group --- that is what additive means. If you believed a quota should change how representation accumulates, you would need to include an interaction term between years_c and quota_f."),
  answer("The predicted representation is lower throughout for quota countries, because the model is controlling for their shorter parliamentary histories",
         message = "Quota countries have higher predicted representation at every level of history --- the line is higher, not lower. The Quota coefficient is positive, about 7.3, so conditioning on Quota shifts the line upward."),
  answer("Both the slope and the level are the same --- the condition argument only affects the caption, not the predictions",
         message = "The condition argument changes the held values for the predictors not on the x-axis, which directly changes the predicted values. Conditioning on Quota rather than No quota shifts the line up by the Quota coefficient --- about 7.3 percentage points."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: in an additive model, which aspects of a prediction plot change when you change the held values, and which stay the same?"
)
```

### Visualize Two Predictors Together

The single-variable plots show parliamentary history and electoral system separately. A two-variable plot combines them: **years_c** on the x-axis, **electoral_f** producing three separate lines, one per electoral system. This lets you see all three trajectories at once and compare the level differences across the full range of history.

**Apply 24: Produce a two-variable prediction plot for history and electoral system**

Write the complete `plot_model()` call from scratch, using **model_rep** as the model object. Put **years_c** on the x-axis and use **electoral_f** to produce separate lines. Use the title \"Predicted Women's Representation by History and Electoral System\", the x-axis label \"Years Since First Woman (centered)\", the y-axis label \"Predicted Percent Women\", and the caption \"Predictions are for countries with no gender quota.\" `plot_model()` maps the grouping variable to the color aesthetic, so without a `color =` argument in `labs()`, the variable name appears as the legend title. Use `labs(color = "Electoral System", caption = "...")` to set both. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-twovar-rep, exercise=TRUE, exercise.cap="Apply 24"}

```

```{r exercise-predplot-twovar-rep-hint-1}
# Hint 1 of 3
# This plot answers two questions at once: how predicted
# representation changes across the history scale, and whether it
# changes the same way in every electoral system. That means one
# variable along the x-axis and a separate line for each category
# of the other.
```

```{r exercise-predplot-twovar-rep-hint-2}
# Hint 2 of 3
# The function is plot_model(). The model object is model_rep.
# Set type = "pred" and terms = c("years_c", "electoral_f").
# axis.title takes a character vector: c("x label", "y label").
# Chain labs(color = "Electoral System", caption = "...") after with +.
```

```{r exercise-predplot-twovar-rep-hint-3}
# Hint 3 of 3
# Your call should be plot_model(model_rep, type = "pred",
# terms = c("years_c", "electoral_f"), with the prescribed title,
# axis labels, and a labs(color = "Electoral System", caption = "...")
# layer. The prescribed caption is:
# "Predictions are for countries with no gender quota."
```

```{r exercise-predplot-twovar-rep-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use plot_model() to produce the prediction plot.")
  if (!grepl("model_rep", code_nc))
    fail("Pass model_rep as the model object.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("years_c", code_nc))
    fail("Include years_c as the x-axis variable in terms.")
  if (!grepl("electoral_f", code_nc))
    fail("Include electoral_f as the grouping variable in terms.")
  if (!grepl("c\\s*\\(", code_nc))
    fail("Use terms = c(\"years_c\", \"electoral_f\") to specify both variables.")
  cov_pos     <- regexpr("c\\s*\\(", code_nc)[1]
  years_pos   <- regexpr("years_c",     code_nc, fixed = TRUE)[1]
  electra_pos <- regexpr("electoral_f", code_nc, fixed = TRUE)[1]
  if (!(years_pos > cov_pos && electra_pos > cov_pos))
    fail("Both years_c and electoral_f should appear inside the terms = c(...) argument.")
  if (years_pos > electra_pos)
    fail("years_c should come first in terms (x-axis), and electoral_f second (separate lines). Check the order inside c(...).")
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Years Since First Woman \\(centered\\)", code_nc))
    fail("Use \"Years Since First Woman (centered)\" as the x-axis label.")
  if (!grepl("Predicted Percent Women", code_nc, fixed = TRUE))
    fail("Use \"Predicted Percent Women\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a labs() layer with color = and caption = arguments.")
  if (!grepl("Electoral System", code_nc, fixed = TRUE))
    fail("Use labs(color = \"Electoral System\") to set the legend title.")
  if (!grepl("Predictions are for countries with no gender quota.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for countries with no gender quota.\"")
  pass("Right. Three electoral systems, three lines, all from a single `plot_model()` call.")
})
```

**Apply 25: Interpret the two-variable plot for history and electoral system**

```{r quiz-twovar-rep-interpret, echo=FALSE}
question("Look at the two-variable prediction plot. Which statement best describes what the three lines show?",
  correct = "Right. Parallel lines, equal slopes, vertical gaps equal to the electoral-system coefficients.",
  answer("The three lines are parallel --- the history slope is the same in every electoral system --- and the vertical gaps between them are the electoral-system coefficients",
         correct = TRUE,
         message = "Proportional systems are predicted highest at every level of history, mixed systems fall between, and plurality systems lowest. Parallel lines are the visual signature of an additive model, because the slope on years_c is identical across electoral systems when there is no interaction term. The vertical gaps are the electoral-system coefficients --- about 7.8 points for proportional and about 5.5 for mixed, the same values you read from the table."),
  answer("The lines are not parallel --- the history slope is steeper for plurality systems, showing that time matters more where the electoral rules help less",
         message = "That is a real hypothesis, and Tutorial 18 tests it. But this model is additive, so the lines here are parallel by construction. Non-parallel lines are the visual signature of an interaction, which this model does not include."),
  answer("The vertical gaps between lines grow as parliamentary history lengthens, showing that the advantage of proportional systems compounds over time",
         message = "In an additive linear model, parallel lines mean constant vertical gaps --- the distance between the lines is the same at every level of history. Growing gaps would require an interaction term."),
  answer("The three lines converge at long histories, suggesting that electoral-system differences disappear once women have served in parliament for long enough",
         message = "Parallel lines do not converge --- they maintain a constant vertical gap throughout. Converging lines would require an interaction between history and electoral system."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at whether the three lines have the same slope and whether the vertical gaps between them are constant or changing across the range of history."
)
```

### What We Learned About Electoral Rules and Women's Representation

The prediction plots translate the coefficient table into a visual argument.

::: {.research-example}
Parliamentary history is a real but gentle predictor: across the full 114-year range of the centered variable, predicted representation rises from about 12.6% to about 23.3%, roughly 11 percentage points. The electoral-system plot shows that this gradual accumulation coexists with a sharp institutional pattern: with no quota and an average length of history, proportional systems are predicted to seat about 25.6% women against about 17.8% for plurality systems, and the two confidence intervals do not overlap. Mixed systems fall between at about 23.3%, with an interval overlapping both. A change of electoral system is worth roughly as much as three-quarters of a century of accumulated history.

These two plots together show something the coefficient table could only summarize in rows of numbers: institutional rules and elapsed time both predict representation, but not on the same scale. The mixed-system case is also a useful caution. Its predicted interval overlaps plurality's, yet its coefficient is statistically significant at p = .026 --- overlapping intervals on two predicted values do not imply that the difference between them is indistinguishable from zero. The coefficient test, not the visual overlap, is the formal evidence.
:::


## Check Your Understanding

The two examples showed that `predict()` translates coefficients into concrete model-implied comparisons, that `plot_model()` makes those comparisons visible across the full range of a predictor, and that a complete interpretation requires more than noting the direction of a relationship. The questions below ask you to apply these ideas to new scenarios without looking back at the examples.

**Apply 26: Identify the predictor and held values**

A researcher estimates a model of county-level turnout as a function of median income, unemployment rate, and college graduation rate. Her research question is: "Do counties with higher college graduation rates turn out at higher rates, holding economic conditions constant?" She wants to produce a prediction plot.

```{r quiz-cyu-predictor, echo=FALSE}
question("Which predictor should vary on the x-axis, and what values should she use for the others?",
  correct = "Right. The predictor the research question is about goes on the x-axis; the controls are held at standardized values.",
  answer("College graduation rate on the x-axis; median income and unemployment rate held at their means",
         correct = TRUE,
         message = "The research question is about college graduation rate, so that goes on the x-axis. The other two predictors --- median income and unemployment rate --- are economic conditions the question asks to hold constant. Means are the conventional default for interval predictors."),
  answer("Median income on the x-axis; unemployment rate and college graduation rate held at their means",
         message = "The research question is about college graduation rate, not median income. The predictor the hypothesis is about is what should vary on the x-axis."),
  answer("All three predictors should vary simultaneously to show the full multivariate relationship",
         message = "A useful prediction plot varies the predictor the research question is about while holding the remaining predictors at interpretable values. Varying all three simultaneously would make it difficult to isolate and communicate the specific relationship the hypothesis concerns."),
  answer("College graduation rate on the x-axis; median income and unemployment rate each at their maximum values to show the best-case scenario",
         message = "Choosing maximum values to produce a best-case scenario is not a principled default. Means are the conventional choice for held values unless there is a specific substantive reason to choose otherwise --- and that reason should be stated in the caption."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: the research question identifies which predictor the hypothesis is about. That variable goes on the x-axis."
)
```

**Apply 27: Interpret parallel lines**

A researcher estimates an additive model of policy liberalism as a function of state median income and a three-category region variable (South, Northeast, West). She produces a two-variable prediction plot with median income on the x-axis and three separate lines for each region. The lines are parallel.

```{r quiz-cyu-parallel, echo=FALSE}
question("What do the parallel lines tell you about the relationship between income and policy liberalism?",
  correct = "Right. Parallel lines mean equal slopes --- the income relationship is the same in every region.",
  answer("The association between median income and policy liberalism is the same in every region",
         correct = TRUE,
         message = "Parallel lines are the visual signature of an additive model. The income slope is identical across regions because there is no interaction term between income and region. The vertical gaps between lines are the region coefficients --- they shift the lines up or down but do not change their steepness."),
  answer("The three regions have identical predicted policy liberalism at every income level --- the lines are indistinguishable",
         message = "Parallel lines can still be vertically separated. Parallel means equal slopes, not equal intercepts. The gaps between lines are the region coefficients."),
  answer("The income association is stronger in some regions than others, which is why the lines run in different directions",
         message = "If the income association differed across regions, the lines would not be parallel --- they would have different slopes and would cross or diverge. Parallel lines mean exactly equal slopes in every group."),
  answer("The parallel lines indicate that region is not a significant predictor of policy liberalism once income is controlled",
         message = "Parallelism is about the interaction between income and region, not about whether region is significant. The lines can be parallel and still be vertically separated by large, significant region coefficients."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: what does it mean geometrically for lines to be parallel? And what would non-parallel lines indicate about the model?"
)
```

**Apply 28: Diagnose an incomplete interpretation**

A student writes the following interpretation of a prediction plot showing the relationship between poverty rate and Democratic vote share across US counties:

*"As poverty rate increases, predicted Democratic vote share decreases. This proves that economic hardship causes counties to vote Republican. The effect is large."*

```{r quiz-cyu-diagnose, echo=FALSE}
question("Which of the following best identifies what is wrong or missing from this interpretation?",
  correct = "Right. Multiple problems: causal claim, missing held values, missing specific values, missing uncertainty.",
  answer("It claims causation, omits the held values, gives no specific predicted values, and does not describe the confidence band",
         correct = TRUE,
         message = "A complete interpretation avoids causal language (uses 'associated with' rather than 'causes'), states the held values so readers know what covariate profile the plot shows, gives at least two specific predicted values to make 'large' concrete, and describes the confidence band. All four components are missing or wrong here."),
  answer("It is correct --- poverty rate is a valid predictor of vote share and the interpretation covers the key finding",
         message = "The interpretation has at least two serious problems: 'causes' is causal language inappropriate for an observational regression, and 'large' is vague without specific predicted values to anchor it."),
  answer("The only problem is the causal language --- changing 'causes' to 'is associated with' would make the interpretation complete",
         message = "Removing the causal claim is necessary but not sufficient. The interpretation is also missing the held values for other predictors, specific predicted values to quantify 'large', and any description of the confidence band."),
  answer("The only problem is that 'large' is vague --- adding a specific predicted value would complete the interpretation",
         message = "Adding a specific value would help, but the causal language ('causes'), missing held values, and missing confidence band description are also problems."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: check the student's interpretation against all five components of the framework and the causal language convention."
)
```

## The Takeaways

Predicted values translate the model's estimated associations into concrete [**model-implied comparisons**]{.important-text}. They describe how expected outcomes differ across specified predictor values. With the observational data used in these tutorials, they do not show what would happen if we deliberately changed one of those predictors.

The key things to carry forward:

1. **Predictions extend the regression analysis you learned in Tutorials 15 and 16.** After estimating the model, interpreting its coefficients, testing the hypotheses, and assessing fit, use the workflow **table → specific predictions → plot**. The table documents the full model. The `predict()` calls translate selected coefficients into concrete comparisons in the units of the outcome --- each one an [**expected outcome**]{.important-text}, the average predicted for cases with that profile rather than the value any single case will have. The plot shows how predicted outcomes change across the full range of the predictor, with uncertainty visible as a shaded band.

2. **`predict()` requires a `newdata` data frame with every predictor named and typed exactly as in the model.** For factor variables, the recommended approach is `factor("level", levels = levels(data$variable))` --- this explicitly attaches the full level set and avoids edge cases where R cannot determine the levels from context. The output is in the units of the outcome variable.

3. **`plot_model(type = "pred", terms = "variable")` sweeps across the full range of the specified predictor**, holding all others at their means or reference category. `axis.title` takes x first, then y. The variable name in `terms` must be in quotes --- unlike most R arguments. Always add `labs(caption = "...")` to state the [**held values**]{.important-text}.

4. **The `condition` argument overrides the default held values.** `condition = list(variable_name = value)` holds the named variable at a specific value. Variable names in `condition` are **unquoted** list element names (unlike `terms =`, where names are quoted strings). Values are numbers for interval variables or quoted strings for factor variable levels. Use it when the default held values do not correspond to the context your research question concerns. The caption must always state what values were used.

5. **`terms = c("x", "grouping_variable")` shows the relationship for multiple groups simultaneously.** In an additive model, the lines are [**parallel lines**]{.important-text} --- the x-variable has the same estimated slope in every group. Parallel lines are the visual signature of an additive model, not a coincidence. `plot_model()` has additional capabilities not covered here --- a third variable in `terms` produces faceted panels, and bracket notation like `[meansd]` or `[value1, value2]` controls which specific values are shown for any term. These features become useful when visualizing interaction models and logistic regression predictions --- Tutorials 19 and 21 build on them.

6. **A slope is the effect of its predictor.** [**Effect**]{.important-text} is used here in the statistical sense --- the expected change in the outcome for a one-unit increase, holding the others constant. It carries no claim that one variable causes the other.

7. **A complete interpretation covers five things:** the pattern, the held values, specific predicted values and their difference, the [**confidence interval**]{.important-text} (the shaded band around the line, showing uncertainty in the expected outcome), and the connection to the hypothesis. Wide or overlapping confidence intervals show that the expected outcomes are estimated less precisely or less distinctly at the displayed predictor profiles; they are not an exact significance test of the differences between predictions.

Together, Tutorials 15--17 provide the core workflow for linear regression. You can now estimate simple and multiple regression models, interpret coefficients, test hypotheses, assess model fit, and translate the results into substantively meaningful predictions and plots.

Tutorials 18 and 19 are an optional extension of that core workflow. Tutorial 18 introduces interaction models, where the slope of one predictor depends on the value of another. Tutorial 19 then extends these prediction tools to interaction models, where prediction plots make conditional relationships visible. Tutorials 20 and 21 carry the same workflow to binary outcomes: logistic regression coefficients tell you direction and significance but not effect size, making prediction plots especially important.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial.

```{r encode-logic-17, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-17, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-17, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now translate regression coefficients into concrete predicted comparisons — and that changes what it means to understand a model.">
  <span aria-hidden="true">📊</span>
  Excellent work! You can now translate regression coefficients into concrete predicted comparisons &mdash; and that changes what it means to understand a model.
</p>
')
```
