---
title: "Bivariate Description: Two Interval Variables"
tutorial:
  id: "09-Bivariate-Description-Interval-Data"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Scatter plots and the correlation coefficient.
---

## Overview {#main-content}

Political arguments often connect one quantity to another: whether wealthier countries are more democratic, whether education levels track turnout, or whether economic conditions move with presidential approval. Those relationships may sound straightforward, but the pattern in the data can be strong or weak, positive or negative, tightly clustered or scattered, and sometimes driven by a few unusual cases.

This tutorial covers the case where both variables are interval. With no groups to compare, the question becomes whether the two variables move together across their full ranges, in which direction, and how closely. The tools are the scatter plot and the correlation coefficient. Each section opens with a research question and a hypothesis, then asks whether the pattern in the data is consistent with what the hypothesis predicts. Tutorial 6 is where to look if you need the mean or the standard deviation again --- the correlation coefficient is built from both.

In this tutorial you will learn:

* How to visualize the relationship between two interval variables with a scatter plot, using `ggplot()` from the [ggplot2]{.package-name} package with `geom_point()` and `geom_smooth()`
* How to summarize the direction and strength of a linear relationship with Pearson's correlation coefficient, using `cor()` from base R
* How to evaluate whether the pattern in a scatter plot and a correlation coefficient is consistent with a hypothesis


```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)
data("counties", package = "QPATutorialsCourse")
data("qog", package = "QPATutorialsCourse")
data("world", package = "QPATutorialsCourse")
```


## Describing the Relationship Between Two Interval Variables

Does economic development tend to go hand in hand with longer life expectancy? Do citizens of more democratic countries express more support for democratic governance? Do counties that experienced stronger wage growth vote differently than those that did not? Each of these questions asks whether two interval variables move together in a systematic way --- and each is exactly the kind of question this tutorial addresses.

Examining two variables at a time to see whether the values of one differ systematically across the values of the other is [**bivariate description**]{.important-text}, in which both variables are interval. Where one variable is categorical, cases can be divided into groups and compared across them; where both are interval, there are no groups to compare. In Tutorials 7 and 8, one variable was always categorical, which meant we could divide cases into groups and compare across them. When both variables are interval, there are no groups to compare. Instead, we ask whether the two variables tend to move together across their full range of values --- and if so, in which direction and how strongly.

A [**positive relationship**]{.important-text} means that higher values of one variable tend to go with higher values of the other. Countries with more effective governments tend to have longer life expectancies. Counties with higher median incomes tend to have higher educational attainment. The variables move in the same direction.

A [**negative relationship**]{.important-text} means that higher values of one variable tend to go with lower values of the other. Countries with higher GDP per capita tend to have lower fertility rates. The variables move in opposite directions.

[**No linear relationship**]{.important-text} means that higher or lower values of one variable do not systematically go with higher or lower values of the other in a straight-line pattern.

Two tools let us describe these patterns. A **scatter plot** shows the relationship visually --- each point represents one case, positioned by its values on the two variables. A **correlation coefficient** summarizes the same pattern with a single number that captures both direction and strength.


The most direct way to see the relationship between two interval variables is to plot one against the other. A [**scatter plot**]{.important-text} does exactly that: each case becomes a point, placed by its value on the explanatory variable along the horizontal axis and its value on the outcome along the vertical. The scatter plot below shows the relationship between **prop_college_grad**, the proportion of a county's residents age 25 or older who have completed a bachelor's degree or higher, and **dem2p_percent**, the percent of the 2016 presidential two-party vote cast for Hillary Clinton. Each dot is one county. The vertical and horizontal lines mark the mean values of each variable, dividing the plot into four quadrants. Four counties are labeled in color --- one in each quadrant.

```{r scatterplot-clintonvote-by-education-labeled, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', fig.alt="Scatter plot of U.S. counties showing proportion of adults with a college degree on the x-axis and percent of two-party vote for Clinton in 2016 on the y-axis. Each point represents one county. Vertical and horizontal reference lines mark the means of the two variables, dividing the plot into four quadrants. Four counties are labeled in color: Arlington VA in the upper right (high education, high Clinton vote), Bronx NY in the upper left (low education, high Clinton vote), McDowell WV in the lower left (low education, low Clinton vote), and Williamson TN in the lower right (high education, low Clinton vote)."}
library(ggplot2)
library(dplyr)
library(ggrepel)

counties_plot <- counties %>%
  filter(!is.na(prop_college_grad))

mean_educ <- mean(counties_plot$prop_college_grad)
mean_vote <- mean(counties_plot$dem2p_percent)

labeled <- counties_plot %>%
  filter(
    (county_name == "Arlington"  & state_name == "Virginia")      |
    (county_name == "Mcdowell"   & state_name == "West Virginia") |
    (county_name == "Bronx"      & state_name == "New York")      |
    (county_name == "Williamson" & state_name == "Tennessee")
  ) %>%
  mutate(
    label = case_when(
      county_name == "Arlington"  ~ "Arlington, VA",
      county_name == "Mcdowell"   ~ "McDowell, WV",
      county_name == "Bronx"      ~ "Bronx, NY",
      county_name == "Williamson" ~ "Williamson, TN"
    ),
    point_color = case_when(
      county_name == "Arlington"  ~ "#2166ac",
      county_name == "Mcdowell"   ~ "#d73027",
      county_name == "Bronx"      ~ "#1a9850",
      county_name == "Williamson" ~ "#f46d43"
    )
  )

ggplot(counties_plot, aes(x = prop_college_grad, y = dem2p_percent)) +
  geom_point(color = "steelblue", alpha = 0.4, size = 1.2) +
  geom_vline(xintercept = mean_educ, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = mean_vote, color = "black", linewidth = 0.6) +
  geom_point(data = labeled, aes(color = I(point_color)), size = 3) +
  geom_label_repel(data = labeled,
                   aes(label = label, color = I(point_color)),
                   fill = "white", size = 3.2, fontface = "bold",
                   box.padding = 0.5, point.padding = 0.3,
                   segment.color = "grey50", show.legend = FALSE) +
  labs(
    title    = "Relationship between Education and Vote for Clinton",
    subtitle = "US Counties in 2016",
    x        = "Proportion of Adults with a College Degree",
    y        = "Percent of Two-Party Vote for Clinton",
    caption  = "Source: Linn, Nagler, Zilinsky"
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title    = element_text(size = 10),
    axis.text     = element_text(size = 9),
    plot.caption  = element_text(size = 8, hjust = 1)
  )
```

The four labeled counties illustrate the quadrant logic. **Arlington, VA** (blue, upper right) has above-average college graduation and above-average Clinton vote --- above the mean on both variables. **McDowell, WV** (red, lower left) has below-average college graduation and below-average Clinton vote --- below the mean on both. Both counties sit in quadrants consistent with a positive relationship: cases that are high on one variable tend to be high on the other, and cases that are low on one tend to be low on the other.

The two remaining counties show a different pattern. **Bronx, NY** (green, upper left) has below-average college graduation but above-average Clinton vote. **Williamson, TN** (orange, lower right) has above-average college graduation but below-average Clinton vote. These counties sit in quadrants where the variables move in opposite directions --- which is what a negative relationship looks like. The fact that there are far fewer counties in these two quadrants than in the lower-left and upper-right is consistent with the overall pattern being positive: most counties follow the direction of the relationship, with education and Clinton vote moving together.

A negative relationship would show the opposite pattern overall --- most points clustering in the upper-left and lower-right quadrants, with above-average values on one variable going with below-average values on the other.

The visual pattern we have been describing has a name: [**correlation**]{.important-text}. Pearson's correlation describes the direction and strength of the **linear relationship** between two interval variables --- how consistently the two variables move together in a straight-line pattern.

[**Pearson's correlation coefficient**]{.important-text}, denoted $r$, summarizes this pattern with a single number that captures both its direction --- positive or negative --- and its strength: how closely the two variables move together along a straight line.

Pearson's correlation coefficient is given by the following formula:

$$r=\frac{1}{n-1} \Sigma \frac{ (X_i-\bar{X}) (Y_i-\bar{Y}) }{S_X S_Y}$$

where

* $X_i,Y_i$ are the individual observations of $X$ and $Y$ for unit $i$,
* $\bar{X},\bar{Y}$ are the sample means of $X$ and $Y$,
* $S_X,S_Y$ are the sample standard deviations of $X$ and $Y$, and
* $n$ is the sample size.


You do not need to memorize this formula, and you will never calculate it by hand in this course. Its job here is to show the intuition behind correlation. For each case, the formula asks whether that case is above or below the mean on each variable. If a case is above the mean on both variables, or below the mean on both variables, it contributes positively to $r$. If a case is above the mean on one variable and below the mean on the other, it contributes negatively to $r$.

The value of $r$ ranges between -1 and +1.

* $r>0$ indicates a positive relationship: higher values of one variable tend to occur with higher values of the other variable.
* $r<0$ indicates a negative relationship: higher values of one variable tend to occur with lower values of the other variable.
* $r=0$ indicates no linear relationship.

Below is a picture illustrating the relationship between the pattern in a scatter plot and Pearson's $r$. A line is drawn through the cloud of points to summarize the overall linear pattern. If the line slopes downward, as on the left, $r$ is negative. If the line is flat, as in the middle, $r$ is close to zero. If the line slopes upward, as on the right, $r$ is positive. The strength of the relationship depends on how tightly the points cluster around a straight-line pattern: tighter clustering means a stronger relationship, while more scatter means a weaker relationship. Note that the slope of the fitted line tells us the direction of the relationship but not by itself how strong it is --- a steep line does not necessarily mean a strong correlation, because the slope depends on the scales of the two variables. Strength is always about how closely the points follow a linear pattern, not how steep the line appears.

```{r scatterplot-correlation, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, out.width='80%', fig.alt="Three side-by-side scatter plots illustrating different linear relationships. The left panel shows a strong negative relationship with r = -0.86, points clustered around a downward-sloping line. The middle panel shows no relationship with r = 0.04, points scattered randomly around a nearly flat line. The right panel shows a strong positive relationship with r = 0.86, points clustered around an upward-sloping line."}
library(patchwork)

set.seed(123)
n <- 150

neg <- MASS::mvrnorm(n, mu = c(0,0),
                     Sigma = matrix(c(1, -0.85, -0.85, 1), 2, 2))
df_neg <- data.frame(x = neg[,1], y = neg[,2])

set.seed(789)
zero <- MASS::mvrnorm(n, mu = c(0,0),
                      Sigma = matrix(c(1, 0.05, 0.05, 1), 2, 2))
df_zero <- data.frame(x = zero[,1], y = zero[,2])

set.seed(123)
pos <- MASS::mvrnorm(n, mu = c(0,0),
                     Sigma = matrix(c(1, 0.85, 0.85, 1), 2, 2))
df_pos <- data.frame(x = pos[,1], y = pos[,2])

r_neg  <- round(cor(df_neg$x,  df_neg$y),  2)
r_zero <- round(cor(df_zero$x, df_zero$y), 2)
r_pos  <- round(cor(df_pos$x,  df_pos$y),  2)

panel_theme <- theme_bw() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 11, hjust = 0.5, color = "#444444"))

p_neg <- ggplot(df_neg, aes(x, y)) +
  geom_point(color = "steelblue", alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d73027", linewidth = 1) +
  labs(title = "Negative relationship", subtitle = paste0("r = ", r_neg)) +
  panel_theme

p_zero <- ggplot(df_zero, aes(x, y)) +
  geom_point(color = "steelblue", alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d73027", linewidth = 1) +
  labs(title = "No linear relationship", subtitle = paste0("r = ", r_zero)) +
  panel_theme

p_pos <- ggplot(df_pos, aes(x, y)) +
  geom_point(color = "steelblue", alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d73027", linewidth = 1) +
  labs(title = "Positive relationship", subtitle = paste0("r = ", r_pos)) +
  panel_theme

p_neg + p_zero + p_pos + plot_layout(ncol = 3)
```

How do we translate a specific value of $r$ into a description like weak, moderate, or strong? When describing strength, we look at the **absolute value** of $r$, written $|r|$, which means its distance from zero regardless of whether the sign is positive or negative --- a correlation of -.7 is just as strong as one of +.7, just in the opposite direction. There are no sharp cutoffs, but a common rough guide is: values of $|r|$ below about .3 are usually described as weak, values from about .3 to .6 as moderate, and values above about .6 as strong. These thresholds are only a starting point --- what counts as strong can depend on the variables involved and the research question, so we will always describe both the sign and the approximate size of $r$ rather than mechanically applying fixed labels.

This is still descriptive analysis. A scatter plot and Pearson's $r$ can tell us whether the pattern in the data is consistent with what a hypothesis predicts --- whether the relationship points in the predicted direction and whether it is strong or weak. They cannot tell us whether the observed relationship is larger than we would expect by chance --- that requires a formal test, covered in Tutorial 14. For now, our question is the same as in Tutorials 7 and 8: does the descriptive pattern point in the direction the hypothesis predicts?

### Building a Scatter Plot with `ggplot()`

Building a scatter plot follows the same basic steps as the plots you built in earlier tutorials. Filter out the missing values in whichever of the two variables has any, before passing the data to `ggplot()` --- this keeps missing cases out of the display and avoids ggplot dropping observations and warning about them. A variable with nothing missing needs no condition, so check the counts before writing the filter. Then map the explanatory variable to `x` and the outcome variable to `y` inside `aes()`. Add `geom_point()` to draw one point per case, and `geom_smooth(se = FALSE, method = "lm")` to add a straight fitted line that summarizes the overall pattern. The `method = "lm"` argument draws a straight line; `se = FALSE` removes a shaded band that ggplot would otherwise draw around it. Finish with `labs()` for an informative title and axis labels. You will see this structure in the first run-and-observe below.

When you use `geom_smooth(method = "lm")`, R prints a message saying `geom_smooth() using formula = 'y ~ x'`. That confirms the type of line being fit --- it is not a warning or an error.

### Calculating Pearson's $r$ with `cor()`

To calculate Pearson's $r$, use the `cor()` function in base R. Give it the two variables as `x` and `y` (as `data$variable`), set `use = "pairwise.complete.obs"` so R uses only cases with observed values on both variables, and set `method = "pearson"` to request Pearson's $r$. The order of the two variables makes no difference here: a correlation measures how tightly two variables move together, not how one acts on the other, so `cor(x, y)` and `cor(y, x)` return the same number.

We will use this same two-tool structure throughout the tutorial: inspect the scatter plot first, then calculate Pearson's $r$. Let's turn to three examples.

## Do More Effective Governments Have Higher Life Expectancy?

Why do people live longer in some countries than in others? One possibility is that government effectiveness plays a role --- countries where governments function well may be better positioned to support population health.

::: {.hypothesis}
In a comparison of [countries], those with [more effective governments] will tend to have [higher life expectancy] than will those with [less effective governments].
:::

The theoretical explanation is that effective governments deliver the public goods that directly sustain population health --- clean water, sanitation, vaccination programs, and reliable healthcare systems --- while less effective governments leave these needs unmet.

The data frame **world** contains the variable **effectiveness**, a measure of government effectiveness running from 0 to 100 in principle, though the observed values run from about 7.8 to 100, where 100 is most effective. It is stored as a numeric variable. It also contains **lifeex_total**, life expectancy at birth, also stored as a numeric variable. **effectiveness** is our explanatory variable and **lifeex_total** is our outcome, so **effectiveness** goes on the `x`-axis and **lifeex_total** goes on the `y`-axis. **effectiveness** is missing for 14 countries and **lifeex_total** for 2, and because a point needs both values to be drawn, 153 of the 167 countries appear in the plot.

### Create the Scatter Plot

**Run and observe 1: Build a scatter plot of government effectiveness and life expectancy**

Load [dplyr]{.package-name} and [ggplot2]{.package-name} at the top of the code --- once loaded, both are available for all subsequent exercises in this tutorial.

```{r scatterplot-effectiveness-lifeexpectancy, exercise=TRUE, exercise.cap="Run and observe 1", fig.alt="Scatter plot of countries showing government effectiveness on the x-axis and life expectancy at birth on the y-axis. Each point represents one country. The points form an upward-sloping band from the lower left to the upper right, and a fitted line runs upward through them. The points are more widely spread at lower and middle levels of government effectiveness than at higher levels."}
library(dplyr)
library(ggplot2)
world %>%
  filter(!is.na(effectiveness), !is.na(lifeex_total)) %>%
  ggplot(mapping = aes(x = effectiveness, y = lifeex_total)) +
  geom_point(color = "steelblue") +
  geom_smooth(color = "darkorange", se = FALSE, method = "lm") +
  labs(title = "Government Effectiveness and Life Expectancy",
       x = "Government Effectiveness",
       y = "Life Expectancy at Birth",
       caption = "Source: Kaufmann (2002); CIA")
```

### Interpret the Scatter Plot

**Practice 2: Read the scatter plot**

```{r quiz-interpret-scatterplot-effectiveness-lifeexpectancy, echo=FALSE}
question("Based on the scatter plot, which statement best describes the relationship between government effectiveness and life expectancy?",
  correct = "Right.",
  answer("The points follow an upward-sloping pattern with moderate clustering around the fitted line",
         correct = TRUE,
         message = "The line slopes upward and the points cluster reasonably closely around it, but there is still meaningful scatter --- some countries have much higher or lower life expectancy than the line would predict."),
  answer("The points are widely scattered with no clear pattern, and the fitted line is nearly flat",
         message = "Look at the direction of the fitted line. It slopes clearly upward, not flat, and the points follow that trend more closely than random scatter would produce."),
  answer("The points cluster tightly along a steep upward-sloping line, indicating a very strong positive relationship",
         message = "The direction is positive but the clustering is moderate, not tight. There is meaningful scatter around the line --- some countries have much higher or lower life expectancy than the line predicts."),
  answer("The fitted line slopes downward, suggesting countries with more effective governments tend to have lower life expectancy",
         message = "Look at the slope of the fitted line. It rises from left to right, indicating that higher government effectiveness is associated with higher, not lower, life expectancy."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Two separate things describe a scatter plot: which way the fitted line slopes, and how closely the points follow it. Check each one before choosing."
)
```

### Calculate Pearson's Correlation

We use the same `cor()` syntax introduced above, with **effectiveness** and **lifeex_total** as the two interval variables whose relationship we want to summarize. Either variable can go first; the result is the same.

**Run and observe 3: Calculate Pearson's r for government effectiveness and life expectancy**

```{r correlation-effectiveness-lifeexpectancy, exercise=TRUE, exercise.cap="Run and observe 3"}
cor(x      = world$effectiveness,
    y      = world$lifeex_total,
    use    = "pairwise.complete.obs",
    method = "pearson")
```

### Is the Pattern Consistent with the Hypothesis?

**Apply 4: Evaluate the government effectiveness/life expectancy hypothesis**

```{r quiz-hypothesis-effectiveness-lifeexpectancy, echo=FALSE}
question("Based on the scatter plot and the correlation coefficient, which statement best describes the pattern?",
  correct = "Right. The relationship is positive and moderate-to-strong, consistent with the hypothesis.",
  answer("The descriptive pattern is consistent with the hypothesis --- the relationship is positive and moderate-to-strong",
         correct = TRUE,
         message = "A moderate-to-strong, correctly-signed correlation is a descriptive pattern consistent with the hypothesis. It does not tell us whether this relationship is bigger than we would expect by chance, or rule out other explanations for why effective governments and longer lives go together."),
  answer("The descriptive pattern is consistent with the hypothesis, and the correlation confirms that effective government causes longer life expectancy",
         message = "A correlation tells us about the direction and strength of the descriptive association. It cannot establish causation or confirm a hypothesis --- it can only show whether the pattern points in the predicted direction."),
  answer("The descriptive pattern is not consistent with the hypothesis because r = .63 is not large enough to matter",
         message = "A value of r around .63 is moderate-to-strong by the rough guide introduced earlier --- it is far from negligible. The pattern points in the predicted direction and is moderate-to-strong, which is consistent with the hypothesis."),
  answer("The descriptive pattern is not consistent with the hypothesis because the scatter around the fitted line is too wide",
         message = "Some scatter around the line is expected and does not mean the hypothesis is wrong. The question is whether the overall pattern points in the predicted direction --- and here, the line slopes upward and r is positive, which is consistent with the hypothesis."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Consistency with the hypothesis is a question about direction. Then ask separately what a correlation can and cannot establish about why the two go together."
)
```


## Is Income Inequality Associated with Democracy?

A long tradition in political science argues that economic inequality and democracy are in tension. Highly unequal societies may struggle to sustain strong democratic institutions because concentrated wealth can be converted into concentrated political power.

::: {.hypothesis}
In a comparison of [countries], those with [higher income inequality] will have [lower levels of democracy] than will those with [lower income inequality].
:::

The theoretical explanation is that when wealth is highly concentrated, the wealthy can translate their economic power into political power --- shaping policy, funding campaigns, and capturing institutions --- which crowds out the broad and equal participation that democracy depends on.

The data frame **qog**, from the Quality of Government project, contains **wdi_gini**, the World Bank's Gini index of income inequality (higher values mean more unequal), and **vdem_polyarchy**, the V-Dem electoral democracy index, which runs from 0 to 1 with higher values indicating more democratic (more free and fair elections, broader suffrage, and stronger protections for political participation). Both are stored as numeric variables. **wdi_gini** is our explanatory variable and **vdem_polyarchy** is our outcome. **wdi_gini** is missing for 80 countries and **vdem_polyarchy** for 21, leaving 110 of the 194 countries with both values recorded --- a much larger share dropped than in the previous example, so keep in mind that any pattern here describes those 110.

### Create the Scatter Plot

Now we can ask whether countries with higher income inequality tend to have lower democracy scores --- and whether any relationship between the two is strong enough to stand out above the scatter.

**Practice 5: Build a scatter plot of income inequality and democracy**

Fill in the blanks in the starter code. The `filter()` call takes two conditions separated by a comma, one for each variable, and you write each condition yourself. Decide which variable is explanatory and which is the outcome before you begin. Two hints are available. Click "Submit Answer" to check your work.

```{r scatterplot-gini-polyarchy, exercise=TRUE, exercise.cap="Practice 5", fig.alt="Scatter plot showing income inequality on the x-axis and the electoral democracy index on the y-axis across countries, with dark teal points and a black fitted line. The points are spread widely across the plot, and the fitted line runs nearly horizontally with only a slight downward tilt."}
qog %>%
  filter(___, ___) %>%
  ggplot(mapping = aes(x = ___, y = ___)) +
  geom_point(color = "darkcyan") +
  geom_smooth(color = "black", se = FALSE, method = "lm") +
  labs(title = "Income Inequality and Democracy",
       x = "Income Inequality (Gini Index)",
       y = "Electoral Democracy Index",
       caption = "Source: Quality of Government project")
```

```{r scatterplot-gini-polyarchy-hint-1}
# Hint 1 of 2
# Think about which variable in the hypothesis is doing
# the explaining and which is being explained.
# The explanatory variable goes on the x-axis;
# the outcome variable goes on the y-axis.
# Both variables have missing values, so both need
# a condition of their own in the filter step.
```

```{r scatterplot-gini-polyarchy-hint-2}
# Hint 2 of 2
# Each filter condition takes the form !is.na(variable), and the two are
# separated by a comma. The same two variable names then appear again inside
# aes(). wdi_gini measures income inequality (explanatory); vdem_polyarchy
# measures democracy (outcome).
```

```{r scatterplot-gini-polyarchy-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing `+` at the end of a line, unmatched parentheses, a label string missing its quotation marks, or a missing `%>%` in the pipeline.")
  }
  if (!grepl("\\bqog\\b", code_no_comments)) {
    fail("Use the **qog** data frame.")
  }

  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Use `filter()` before `ggplot()` to remove rows with missing values, as shown in the starter code.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*wdi_gini\\s*\\)", code_no_comments)) {
    fail("In `filter()`, remove rows where **wdi_gini** is missing with `!is.na(wdi_gini)`.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*vdem_polyarchy\\s*\\)", code_no_comments)) {
    fail("In `filter()`, remove rows where **vdem_polyarchy** is missing with `!is.na(vdem_polyarchy)`.")
  }
  if (!grepl("ggplot\\s*\\(", code_no_comments)) {
    fail("Use `ggplot()` to build the scatter plot.")
  }
  if (!grepl("x\\s*=\\s*wdi_gini", code_no_comments)) {
    fail("The explanatory variable belongs on the x-axis inside `aes()`. The hypothesis above tells you which of the two is doing the explaining.")
  }
  if (!grepl("y\\s*=\\s*vdem_polyarchy", code_no_comments)) {
    fail("The outcome variable belongs on the y-axis inside `aes()` --- the one the hypothesis says is being explained.")
  }
  if (!grepl("geom_point\\s*\\(", code_no_comments)) {
    fail("Add `geom_point()` to plot the data as a scatter plot.")
  }
  if (!grepl("geom_smooth\\s*\\(", code_no_comments)) {
    fail("Add `geom_smooth()` to show the fitted line.")
  }
  if (!grepl("method\\s*=\\s*['\"]lm['\"]", code_no_comments)) {
    fail('Set `method = "lm"` inside `geom_smooth()` to draw a straight fitted line.')
  }
  if (!grepl("se\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set `se = FALSE` inside `geom_smooth()` to remove the shaded confidence band.")
  }

  pass("Exactly! You filtered out missing values and mapped **wdi_gini** to x and **vdem_polyarchy** to y, matching their roles as explanatory variable and outcome.")
})
```

### Interpret the Scatter Plot

**Practice 6: Read the scatter plot**

```{r quiz-interpret-scatterplot-gini-polyarchy, echo=FALSE}
question("Based on the scatter plot, which statement best describes the relationship between income inequality and democracy?",
  correct = "Right. The points are widely scattered and the fitted line is nearly flat.",
  answer("The points cluster tightly along a steep downward-sloping line, indicating a strong negative relationship",
         message = "Look again at how widely scattered the points are. A strong relationship would show the points clustered close to the fitted line, not spread across the whole plot."),
  answer("The points are widely scattered, and the fitted line is nearly flat with a slight downward tilt",
         correct = TRUE,
         message = "The line tilts down only slightly, and the points don't cluster closely around it --- both are signs of a weak relationship."),
  answer("The points cluster tightly along an upward-sloping line, indicating a strong positive relationship",
         message = "The fitted line does not slope upward --- if anything it tilts slightly downward. And the points are too widely scattered to indicate a strong relationship in either direction."),
  answer("The scatter plot cannot tell us about direction or strength --- we need Pearson's r first",
         message = "A scatter plot is itself informative about direction and strength. The direction of the fitted line shows the relationship's direction, and how closely the points cluster around it shows its strength. Pearson's r gives a precise number, but the plot already tells you something."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "A fitted line always has a slope, even when the points barely follow it. Judge the direction and the spread separately."
)
```

### Calculate Pearson's Correlation

The scatter plot gave us a visual impression of the relationship. Now we can quantify it.

**Practice 7: Calculate the correlation between income inequality and democracy**

Use **wdi_gini** and **vdem_polyarchy** from the **qog** data frame, following the same `cor()` syntax introduced earlier. Two hints are available. Click "Submit Answer" to check your work.

```{r correlation-gini-polyarchy, exercise=TRUE, exercise.cap="Practice 7"}
cor(x      = qog$___,
    y      = qog$___,
    use    = ___,
    method = "pearson")
```

```{r correlation-gini-polyarchy-hint-1}
# Hint 1 of 2
# You need both variables, plus the two arguments that say how missing
# values should be handled and which correlation to compute. The order you list
# the two variables in makes no difference to the answer.
```

```{r correlation-gini-polyarchy-hint-2}
# Hint 2 of 2
# Recall from the section introducing cor(): if either variable has missing values, cor() will
# return NA unless you tell R how to handle them. The government effectiveness example
# used use = "pairwise.complete.obs" --- the same value works here.
```

```{r correlation-gini-polyarchy-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number. Check that your `cor()` call is complete and uses the correct variable names and arguments.")
  }
  if (!grepl("cor\\s*\\(", code_no_comments)) {
    fail("Use the `cor()` function to calculate Pearson's r.")
  }
  if (!grepl("qog\\\\$wdi_gini", code_no_comments) ||
      !grepl("qog\\\\$vdem_polyarchy", code_no_comments)) {
    fail("Include both `qog$wdi_gini` and `qog$vdem_polyarchy` in your `cor()` call. Either order works.")
  }
  if (!grepl("use\\s*=\\s*['\"]pairwise\\.complete\\.obs['\"]", code_no_comments)) {
    fail('Set `use = "pairwise.complete.obs"` so `cor()` handles missing values correctly.')
  }
  if (!grepl("method\\s*=\\s*['\"]pearson['\"]", code_no_comments)) {
    fail('Set `method = "pearson"` since we are calculating Pearson\'s r.')
  }

  expected <- cor(x = qog$wdi_gini,
                  y = qog$vdem_polyarchy,
                  use = "pairwise.complete.obs",
                  method = "pearson")

  if (!isTRUE(all.equal(as.numeric(.result), as.numeric(expected), tolerance = 1e-6))) {
    fail("Check your two variable names, `use = \"pairwise.complete.obs\"`, and `method = \"pearson\"`. Either order works for the two variables.")
  }

  pass("Well done! **wdi_gini** and **vdem_polyarchy** are both in the call, with the missing-value and method arguments set correctly.")
})
```

### Is the Pattern Consistent with the Hypothesis?


**Apply 8: Evaluate the inequality/democracy hypothesis**

```{r quiz-hypothesis-gini-polyarchy, echo=FALSE}
question("Based on the scatter plot and the correlation coefficient you just calculated, is this pattern consistent with the hypothesis?",
  correct = "Right. The pattern is directionally consistent but very weak.",
  answer("Yes --- the correlation is negative as predicted, but so weak that it provides very little support",
         correct = TRUE,
         message = "A correlation of -0.12 points in the predicted direction, so the pattern is consistent with the hypothesis. But the relationship is so weak that it provides very little support. This is a good example of why direction and strength both matter: a correctly signed but near-zero correlation is very different from a strong negative relationship."),
  answer("Yes --- the correlation is negative and strong, exactly as the hypothesis predicts",
         message = "The direction is correct, but r = -0.12 is not strong --- it falls well below the rough threshold of .3 for a moderate relationship."),
  answer("No --- the correlation is negative, which is the opposite of what the hypothesis predicted",
         message = "The hypothesis predicted a negative relationship --- countries with higher inequality should have lower democracy scores. A negative correlation points in exactly that direction. The issue is not the sign but the strength: r = -0.12 is very close to zero."),
  answer("Yes --- even though the correlation is weak, any negative correlation fully confirms the hypothesis",
         message = "A negative correlation is consistent with the hypothesis direction, but 'fully confirms' goes too far. A near-zero correlation means the relationship is barely present. Direction and strength both matter when evaluating descriptive evidence."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Check the sign of r against what the hypothesis predicts, then check its size against the rough guide. Both bear on how much support the data actually give."
)
```

There is a long-standing argument in comparative politics that high inequality undermines democratic institutions. A result like this one is a reminder that a well-motivated expectation is not the same as a pattern in the data --- and that finding little is an answer, not a failed analysis.


## Is Wage Growth Associated with Democratic Vote Percentage?

Voters often reward or punish the party in power based on how the economy has performed. Wage growth is one concrete sign of how the local economy is doing, and counties may differ in how much they benefited during the years leading up to the 2016 election.

::: {.hypothesis}
In a comparison of [counties], those with [higher wage growth] will have [a higher Democratic vote percentage] than will those with [lower wage growth].
:::

The theoretical explanation is that voters vote retrospectively: they credit the incumbent party for a strong local economy, so counties experiencing stronger wage growth should be more likely to reward that party's candidate at the polls.

The data frame **counties** contains **wage_growth**, the percent change in a county's average weekly wage from the third quarter of 2015 to the third quarter of 2016, and **dem2p_percent**, the Democratic candidate's percent of the two-party presidential vote in 2016. Both are stored as numeric variables. **wage_growth** is missing for 2 counties while every county has a recorded **dem2p_percent**, so 3,110 of the 3,112 counties appear in the plot.

### Create the Scatter Plot

The question is whether counties that experienced stronger wage growth rewarded the incumbent party's candidate with a higher Democratic vote percentage.

**Apply 9: Build a scatter plot of wage growth and Democratic vote percentage**

Build this one from scratch, with no starter code. Very little here is actually new: you have written `filter()` and `labs()` many times, and the `labs()` layer works exactly as it did for every bar chart, histogram, and box plot you have built. The genuinely new part is mapping your two interval variables inside `aes()` and adding the two scatter plot layers you learned earlier in this tutorial. Focus your attention there, then fill in the parts you already know well.

Filter out the missing **wage_growth** values before plotting --- **dem2p_percent** needs no condition of its own. Make the points steelblue and the fitted line darkorange. Use the following labels: title \"Wage Growth and Democratic Vote Percentage\", axis labels \"Wage Growth\" and \"Percent of Two-Party Vote\" (you decide which goes where), and caption \"Source: Linn, Nagler, Zilinsky\". Five hints are available. Click "Submit Answer" to check your work.

```{r scatterplot-wagegrowth-demvoteshare, exercise=TRUE, exercise.cap="Apply 9", fig.alt="Scatter plot showing wage growth on the x-axis and Democratic vote percentage on the y-axis across U.S. counties in 2016. Most points form a dense vertical cluster near zero wage growth, spanning nearly the entire range of Democratic vote percentage from close to 0 to close to 100 percent. A smaller number of counties with higher or lower wage growth are more sparsely scattered to the left and right. A fitted line runs from the lower left to the upper right of the plot."}

```

```{r scatterplot-wagegrowth-demvoteshare-hint-1}
# Hint 1 of 5
# You are showing how one measured quantity varies with another across
# counties, one point per county, with a straight line summarizing the
# overall pattern. Which of the two variables does the hypothesis treat as
# doing the explaining, and which as being explained?
```

```{r scatterplot-wagegrowth-demvoteshare-hint-2}
# Hint 2 of 5
# The explanatory variable goes on the x-axis and the outcome on the y-axis,
# as in the previous example. Only one of the two needs a filter condition
# here --- the paragraph above says which --- and the filtering comes before
# the plot is built.
```

```{r scatterplot-wagegrowth-demvoteshare-hint-3}
# Hint 3 of 5
# After the filter step, pipe into ggplot().
# Inside aes(), map the explanatory variable to x
# and the outcome variable to y.
# Use the same variable names as in the filter step.
```

```{r scatterplot-wagegrowth-demvoteshare-hint-4}
# Hint 4 of 5
# Add two geom layers after ggplot():
# geom_point() draws the scatter plot.
# geom_smooth() adds the fitted line --- it needs
# method to specify a straight line and se to
# suppress the shaded band.
```

```{r scatterplot-wagegrowth-demvoteshare-hint-5}
# Hint 5 of 5
# labs() takes title, x, y, and caption.
# The axis labels are "Wage Growth" and
# "Percent of Two-Party Vote" --- the hypothesis
# tells you which variable is explanatory (x)
# and which is the outcome (y).
```

```{r scatterplot-wagegrowth-demvoteshare-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Make sure your code produces a plot the grader can check.")
  }
  if (!inherits(.result, "ggplot")) {
    fail("Your code should return a ggplot object. Check for a missing `+` at the end of a line, unmatched parentheses, a label string missing its quotation marks, or a missing `%>%` in the pipeline.")
  }
  if (!grepl("\\bcounties\\b", code_no_comments)) {
    fail("Use the **counties** data frame.")
  }
  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Filter out missing values with `filter()` before plotting, as in the previous example.")
  }
  if (!grepl("!\\s*is\\.na\\s*\\(\\s*wage_growth\\s*\\)", code_no_comments)) {
    fail("In `filter()`, remove rows where **wage_growth** is missing with `!is.na(wage_growth)`.")
  }
  if (!grepl("ggplot\\s*\\(", code_no_comments)) {
    fail("Use `ggplot()` to build the scatter plot.")
  }

  plot_mapping <- .result$mapping
  x_mapping <- rlang::as_label(plot_mapping$x)
  y_mapping <- rlang::as_label(plot_mapping$y)

  if (!identical(x_mapping, "wage_growth")) {
    fail("The explanatory variable belongs on the x-axis inside `aes()`. The hypothesis above tells you which of the two is doing the explaining.")
  }
  if (!identical(y_mapping, "dem2p_percent")) {
    fail("The outcome variable belongs on the y-axis inside `aes()` --- the one the hypothesis says is being explained.")
  }
  if (!grepl("geom_point\\s*\\(", code_no_comments)) {
    fail("Add `geom_point()` to plot the data as a scatter plot.")
  }
  if (!grepl("geom_smooth\\s*\\(", code_no_comments)) {
    fail("Add `geom_smooth()` to add a fitted line.")
  }
  if (!grepl("method\\s*=\\s*['\"]lm['\"]", code_no_comments)) {
    fail('Set `method = "lm"` inside `geom_smooth()` to draw a straight line.')
  }
  if (!grepl("se\\s*=\\s*FALSE", code_no_comments)) {
    fail("Set `se = FALSE` inside `geom_smooth()` to remove the shaded band.")
  }
  layer_geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  pt <- which(layer_geoms == "GeomPoint")
  sm <- which(layer_geoms == "GeomSmooth")

  if (length(pt) == 0 || !identical(.result$layers[[pt[1]]]$aes_params$colour, "steelblue")) {
    fail("The prompt asks for steelblue points --- set `color = \"steelblue\"` inside `geom_point()`.")
  }
  if (length(sm) == 0 || !identical(.result$layers[[sm[1]]]$aes_params$colour, "darkorange")) {
    fail("The prompt asks for a darkorange fitted line --- set `color = \"darkorange\"` inside `geom_smooth()`.")
  }
  if (!grepl("labs\\s*\\(", code_no_comments)) {
    fail("Add `labs()` to give the plot a title, axis labels, and caption.")
  }

  plot_labels <- .result$labels
  if (!identical(plot_labels$title, "Wage Growth and Democratic Vote Percentage")) {
    fail('Use `title = "Wage Growth and Democratic Vote Percentage"` inside `labs()`.')
  }
  if (!identical(plot_labels$x, "Wage Growth")) {
    fail('Set `x = "Wage Growth"` inside `labs()`.')
  }
  if (!identical(plot_labels$y, "Percent of Two-Party Vote")) {
    fail('Set `y = "Percent of Two-Party Vote"` inside `labs()`.')
  }
  if (!identical(plot_labels$caption, "Source: Linn, Nagler, Zilinsky")) {
    fail('Set `caption = "Source: Linn, Nagler, Zilinsky"` inside `labs()`.')
  }

  pass("Good work! **wage_growth** is mapped to x and **dem2p_percent** is mapped to y, with missing values filtered out and the plot properly labeled and cited.")
})
```

### Interpret the Scatter Plot

**Practice 10: Read the scatter plot**

```{r quiz-wagegrowth-scatter-read, echo=FALSE}
question("Based on the scatter plot, which statement best describes the relationship between wage growth and Democratic vote percentage?",
  correct = "Right. The points cluster near zero wage growth across a wide range of Democratic vote percentage, and the line slopes upward only modestly.",
  answer("The points cluster tightly along a steep upward-sloping line, indicating a strong positive relationship",
         message = "Look again at how widely the points are scattered, especially the dense cluster near zero wage growth that spans almost the full range of Democratic vote percentage. A strong relationship would show much tighter clustering around the line."),
  answer("Most counties cluster near zero wage growth while Democratic vote percentage varies widely",
         correct = TRUE,
         message = "Counties with stronger wage growth show, on average, a slightly higher Democratic vote percentage, but most counties sit close to zero wage growth and vary enormously in Democratic vote percentage --- a sign that wage growth is only weakly related to Democratic vote percentage"),
  answer("The handful of counties with the highest and lowest wage growth show a clear, strong pattern, so the relationship is strong overall",
         message = "Those extreme counties are a small minority of the data. Most counties cluster densely near zero wage growth with a wide spread of Democratic vote percentage, and that dense cluster --- not the few outliers --- is what the correlation mostly reflects."),
  answer("The points show a clear negative relationship: higher wage growth is associated with lower Democratic vote percentage",
         message = "The fitted line slopes upward, not downward. Higher wage growth is associated with slightly higher, not lower, Democratic vote percentage."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Look at where the bulk of the counties sit along the horizontal axis, and how much they vary vertically at that one position."
)
```

### Calculate Pearson's Correlation

Now quantify what you observed.

**Practice 11: Calculate the correlation between wage growth and Democratic vote percentage**

Use **wage_growth** and **dem2p_percent** from the **counties** data frame, following the same `cor()` structure as the previous examples. Three hints are available. Click "Submit Answer" to check your work.

```{r correlation-wagegrowth-demvoteshare, exercise=TRUE, exercise.cap="Practice 11"}

```

```{r correlation-wagegrowth-demvoteshare-hint-1}
# Hint 1 of 3
# This call takes the same four arguments as the two earlier examples. Either
# order works for the two variables --- a correlation is symmetric.
```

```{r correlation-wagegrowth-demvoteshare-hint-2}
# Hint 2 of 3
# Look back at the cor() calls from the government effectiveness and democracy
# examples. The function always takes the same four arguments: x, y, use, and method.
```

```{r correlation-wagegrowth-demvoteshare-hint-3}
# Hint 3 of 3
# The hardest piece is remembering all four arguments and their values:
# x and y are the two variables, each written as counties$variable
# use = "pairwise.complete.obs" handles the missing values
# method = "pearson" requests Pearson's r
```

```{r correlation-wagegrowth-demvoteshare-check}
grade_this({
  code_lines <- gsub("#[^\r\n]*", "", .user_code)
  code_no_comments <- paste(code_lines, collapse = "\n")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Your result should be a single number. Check that your `cor()` call is complete and uses the correct variable names and arguments.")
  }
  if (!grepl("cor\\s*\\(", code_no_comments)) {
    fail("Use the `cor()` function to calculate Pearson's r.")
  }
  if (!grepl("counties\\\\$wage_growth", code_no_comments) ||
      !grepl("counties\\\\$dem2p_percent", code_no_comments)) {
    fail("Include both `counties$wage_growth` and `counties$dem2p_percent` in your `cor()` call. Either order works.")
  }
  if (!grepl("use\\s*=\\s*['\"]pairwise\\.complete\\.obs['\"]", code_no_comments)) {
    fail('Set `use = "pairwise.complete.obs"` so `cor()` handles missing values correctly.')
  }
  if (!grepl("method\\s*=\\s*['\"]pearson['\"]", code_no_comments)) {
    fail('Set `method = "pearson"` since we are calculating Pearson\'s r.')
  }

  expected <- cor(x = counties$wage_growth,
                  y = counties$dem2p_percent,
                  use = "pairwise.complete.obs",
                  method = "pearson")

  if (!isTRUE(all.equal(as.numeric(.result), as.numeric(expected), tolerance = 1e-6))) {
    fail("Check your two variable names, `use = \"pairwise.complete.obs\"`, and `method = \"pearson\"`. Either order works for the two variables.")
  }

  pass("Correct! **wage_growth** and **dem2p_percent** are both in the call, with the missing-value and method arguments set correctly.")
})
```

### Is the Pattern Consistent with the Hypothesis?


**Apply 12: Evaluate the wage growth/Democratic vote percentage hypothesis**

```{r quiz-hypothesis-wagegrowth-demvoteshare, echo=FALSE}
question("Based on the scatter plot and the correlation coefficient you just calculated, is this pattern consistent with the hypothesis?",
  correct = "Right. The correlation is positive but weak --- consistent in direction but small in magnitude.",
  answer("Yes --- the correlation is positive, matching the direction predicted by the hypothesis, though it is quite weak",
         correct = TRUE,
         message = "The positive correlation matches the direction retrospective voting predicts, but the relationship is weak: wage growth is only weakly related to Democratic vote percentage. Reporting this honestly means saying both things together --- the sign is in the predicted direction, but the descriptive relationship is small. A weak relationship provides much less descriptive evidence for a hypothesis than a strong one, even when the sign is correct. This is a common outcome in the social sciences, where any single variable usually explains only part of what we observe."),
  answer("No --- the correlation is too weak to be consistent with any hypothesis",
         message = "The correlation is weak, but it does point in the direction the hypothesis predicted, so the pattern is not flatly inconsistent. The honest description is that the sign matches but the relationship is small --- weak descriptive evidence in the predicted direction, rather than none at all."),
  answer("Yes --- the correlation is positive and strong, which strongly supports the hypothesis",
         message = "The direction is right, but at this size the relationship is weak, not strong. A correct sign does not mean the relationship is large."),
  answer("No --- the relationship is positive, which is the opposite of what retrospective voting would predict",
         message = "Retrospective voting predicts that stronger local economic performance should be rewarded with higher support for the incumbent party's candidate --- a positive relationship. That is exactly the direction we see here."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Check the sign of r against what retrospective voting predicts, then ask how much weight a correlation of that size can carry."
)
```

Whatever separates American counties politically, recent wage growth is only a small part of it. Retrospective voting is a claim about what moves voters at the margin, not a claim that economic experience is the main thing telling one county from another.

## Check Your Understanding

Now let’s step back and practice the main ideas without writing code. Focus on the logic: identify whether a relationship is positive or negative, translate the size of Pearson’s $r$ into words, and use a scatter plot to describe the direction and strength of a relationship.

**Apply 13: Connect the pattern to the sign of r**

```{r quiz-negative-relationship-sign, echo=FALSE}
question("Consider this pattern: large values of the explanatory variable tend to occur with small values of the outcome variable, and small values of the explanatory variable tend to occur with large values of the outcome variable. What does this pattern imply about the sign of r?",
  correct = "Right. Variables that move in opposite directions have a negative correlation.",
  answer("r would be positive",
         message = "A positive r describes the opposite pattern: large values of one variable occurring together with large values of the other, and small with small. This description is about values moving in opposite directions, which is negative."),
  answer("r would be negative",
         correct = TRUE,
         message = "Recall the logic behind the formula: when a case is above the mean on one variable and below the mean on the other, it contributes negatively to r. A pattern where large values of one variable systematically pair with small values of the other is exactly that --- which is why this pattern corresponds to a negative r."),
  answer("r would be zero",
         message = "r is close to zero when there is no systematic tendency in either direction. This description is describing a clear, systematic tendency --- just in the negative direction, not the absence of one."),
  answer("r would be negative only if the relationship is also strong",
         message = "Sign and strength are separate things. The sign of r says which direction the two variables move in; the size of |r| says how tightly they track each other. A negative relationship can be weak, moderate, or strong --- the pattern described here fixes the direction and says nothing about the strength."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The pattern described in the question tells you how the two variables move relative to each other, and that is all the sign depends on."
)
```

**Apply 14: Translate a correlation into words**

Use the rough guide from the tutorial to decide where 0.6 falls on the weak-moderate-strong scale.

```{r quiz-magnitude-r, echo=FALSE}
question("Suppose two variables have a correlation of r = 0.6. Which statement best describes this relationship?",
  correct = "Right. Under the rough guide, r = 0.6 sits at the top of the moderate range, near the boundary with strong.",
  answer("A weak positive relationship",
         message = "Weak relationships sit closer to zero, generally below about .3. At 0.6, this relationship is considerably stronger than that."),
  answer("A moderate positive relationship, near the boundary with strong",
         correct = TRUE,
         message = "Recall the rough guide from earlier: values of |r| below about .3 are usually described as weak, from about .3 to .6 as moderate, and above about .6 as strong. At 0.6, this relationship sits at the top of the moderate range, right at the boundary with strong --- a meaningful, fairly substantial positive relationship. Because the thresholds are only rough, values near a boundary can reasonably be described either way."),
  answer("A strong positive relationship, since .6 is much closer to 1 than to 0",
         message = "Being nearer 1 than 0 is not the test. The rough guide puts values above about .6 in the strong range, so 0.6 sits right at that boundary rather than inside it --- moderate is the better description, with the caveat that it is close."),
  answer("No meaningful relationship, since r is far from 1",
         message = "r does not need to be close to 1 to be meaningful. Only values close to 0 indicate a weak or negligible relationship; 0.6 is well above that range."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Look back at the rough guide for weak, moderate, and strong, and ask which band 0.6 falls in --- and how close it sits to the boundary."
)
```

**Apply 15: Predict a correlation's sign**

Think about which direction the two variables move relative to each other before answering.

```{r quiz-pressfreedom-correlation, echo=FALSE}
question("Suppose x is a country's level of press freedom and y is the number of journalists imprisoned in that country. What sign would you expect for the correlation between x and y?",
  correct = "Right. Greater press freedom goes with fewer imprisoned journalists --- opposite directions, negative correlation.",
  answer("Positive, because countries with more press freedom have more journalists overall",
         message = "The question is about the relationship between press freedom and journalists *imprisoned*, not the total number of journalists. Countries with greater press freedom tend to imprison fewer journalists, so the two variables move in opposite directions."),
  answer("Negative, because greater press freedom goes with fewer journalists imprisoned",
         correct = TRUE,
         message = "This is the same logic from earlier in the tutorial: when higher values of one variable systematically occur with lower values of the other, the correlation is negative. Greater press freedom goes with fewer imprisoned journalists, so we'd expect r to be negative."),
  answer("Zero, because press freedom and imprisonment of journalists are unrelated",
         message = "Press freedom and the imprisonment of journalists are closely related --- that's exactly why we'd expect a relationship, just a negative one rather than no relationship at all."),
  answer("Positive, because both variables measure how prominent the press is in a country",
         message = "Both variables do involve the press, but they measure opposing things: one counts how free journalists are to work, the other counts how many have been jailed for it. Variables about the same subject can still move in opposite directions, and the sign of a correlation follows the direction they move, not the topic they share."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Picture two countries, one with a free press and one without. Which would you expect to have more journalists in prison?"
)
```

Why do families have fewer children in some countries than in others? One possibility is that it depends on what the state provides.

::: {.hypothesis}
In a comparison of [countries], those with [more effective governments] will have [lower fertility rates] than will those with [less effective governments].
:::

The theoretical explanation is that effective governments deliver schooling, health care, and support in old age, which weaken the economic reasons for having many children and widen the opportunities available to women outside the home.

The data frame **world** contains **effectiveness**, a measure of government effectiveness running from 0 to 100 in principle, though the observed values run from about 7.8 to 100, where 100 is most effective, and **fertility**, the average number of children born per woman. **effectiveness** is our explanatory variable and **fertility** is our outcome. **effectiveness** is missing for 14 countries and **fertility** for 2, so 153 of the 167 countries are plotted.

```{r scatterplot-governmenteffectiveness-fertility, echo=FALSE, warning=FALSE, message=FALSE, fig.alt="Scatter plot showing government effectiveness on the x-axis and fertility rate on the y-axis across countries, with purple points and a firebrick fitted line. The line runs from the upper left of the plot down to the lower right. The points are spread around the line, with a wider vertical spread at lower levels of government effectiveness."}
world %>%
  filter(!is.na(effectiveness), !is.na(fertility)) %>%
  ggplot(mapping = aes(x = effectiveness, y = fertility)) +
  geom_point(color = "purple4") +
  geom_smooth(color = "firebrick", method = "lm", se = FALSE) +
  labs(title = "Government Effectiveness and Fertility Around the World",
       x = "Government Effectiveness Score",
       y = "Number of Children",
       caption = "Source: Kaufmann (2002); CIA")
```

**Apply 16: Interpret a new scatter plot**

```{r quiz-governmenteffectiveness-fertility, echo=FALSE}
question("Based on the scatter plot, which statement best describes the relationship between government effectiveness and fertility?",
  correct = "Right. The line slopes clearly downward, so the relationship is negative.",
  answer("A negative relationship: more effective governments tend to have lower fertility",
         correct = TRUE,
         message = "The fitted line slopes downward, so higher government effectiveness goes with lower fertility --- a negative relationship. There is a fair amount of scatter around the line, so the relationship is moderate rather than very strong, but the downward direction is clear."),
  answer("A positive relationship: more effective governments tend to have higher fertility",
         message = "Look again at the direction of the points and the fitted line --- they trend downward, not upward, as government effectiveness increases."),
  answer("A very strong negative relationship, with the points falling tightly along the line",
         message = "The direction is negative, but the points do not fall tightly along the line --- at most levels of government effectiveness, fertility varies quite a bit. The relationship is clearly negative but only moderately strong, not tight."),
  answer("No meaningful relationship; the points are randomly scattered",
         message = "The points are not randomly scattered --- they follow a clear downward trend, which is the signature of a negative relationship, not the absence of one."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Direction first: does the line run up or down as you move to the right? Then ask how tightly the points hold to it."
)
```

Pearson's r for these two variables is -.55.

**Apply 17: Evaluate the government effectiveness/fertility hypothesis**

```{r quiz-hypothesis-effectiveness-fertility, echo=FALSE}
question("Based on the scatter plot and a correlation of -.55, is this pattern consistent with the hypothesis?",
  correct = "Right. The direction matches the prediction, and the relationship is moderate in size.",
  answer("Yes --- the correlation is negative as the hypothesis predicts, and moderate in size",
         correct = TRUE,
         message = "The hypothesis predicted that more effective governments would go with lower fertility, and a negative correlation is exactly that. At -.55 it sits near the top of the moderate range, so the pattern is consistent and reasonably strong --- though a descriptive correlation still cannot tell us whether it is larger than chance."),
  answer("No --- the correlation is negative, which is the opposite of what the hypothesis predicted",
         message = "Read the hypothesis again. It predicts that higher effectiveness goes with LOWER fertility, so a negative correlation is the predicted direction, not the opposite of it."),
  answer("Yes --- and a correlation this size shows that better government causes families to have fewer children",
         message = "The direction is right, but no correlation can establish that one variable causes the other. Wealthier countries tend to have both more effective governments and lower fertility, and that alone would produce this pattern."),
  answer("Partly --- the direction matches, but at -.55 the relationship is too weak to count as support",
         message = "Check -.55 against the rough guide: below about .3 is weak, .3 to .6 moderate. At -.55 this sits near the top of the moderate range, which is not a weak relationship."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Compare the sign of the correlation with the direction the hypothesis predicts, then check its size against the rough guide."
)
```

## The Takeaways

The table below completes the set, adding two interval variables to the two combinations introduced in Tutorials 7 and 8, and everything that follows explains why the entries differ and how to produce them in R.

::: {.qpa-infographic}
<table class="qpa-table">
  <caption>Bivariate Description</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Aspect of the relationship</span></th>
      <th scope="col">Two Categorical Variables</th>
      <th scope="col">One Categorical, One Interval Variable</th>
      <th scope="col">Two Interval Variables</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Numerical Summary</th>
      <td>Cross-tabulation with Column Proportions</td>
      <td>Group Means and Group Sizes</td>
      <td>Correlation Coefficient</td>
    </tr>
    <tr>
      <th scope="row">Visualization</th>
      <td>Grouped Bar Plot of Proportions</td>
      <td>Faceted Histogram, Box Plot</td>
      <td>Scatter Plot</td>
    </tr>
    <tr>
      <th scope="row">What to Look For</th>
      <td>Differences in Column Proportions Across Categories</td>
      <td>Differences in Center, Spread, and Shape Across Categories</td>
      <td>Direction, Form, and Strength of the Relationship</td>
    </tr>
    <tr>
      <th scope="row">R Functions</th>
      <td><code>CrossTable()</code>, <code>geom_bar()</code></td>
      <td><code>group_by()</code>, <code>summarise()</code>, <code>geom_histogram()</code>, <code>facet_wrap()</code>, <code>geom_boxplot()</code></td>
      <td><code>cor()</code>, <code>geom_point()</code>, <code>geom_smooth()</code></td>
    </tr>
  </tbody>
</table>
:::

With this tutorial, you have completed the [**bivariate description**]{.important-text} considered here. The question has been the same throughout --- does the descriptive pattern point in the direction the hypothesis predicts? --- and the answer is always the same kind of answer: consistent or inconsistent with what we expected, but not yet a formal test. Tutorial 14 takes that step for two interval variables.

Two interval variables can be related in three ways. A [**positive relationship**]{.important-text} means that higher values of one tend to go with higher values of the other. A [**negative relationship**]{.important-text} means that higher values of one tend to go with lower values of the other. [**No linear relationship**]{.important-text} means there is no systematic straight-line tendency for the variables to move together. A nonlinear relationship may still exist, which is why the scatter plot matters. The name for this kind of pattern is [**correlation**]{.important-text}, and the two tools below are how we describe it.

The [**scatter plot**]{.important-text} is the visual tool for describing the relationship between two interval variables. Build one with `ggplot()`, mapping the explanatory variable to `x` and the outcome variable to `y` inside `aes()`. Add `geom_point()` to draw each case as a point and `geom_smooth(method = "lm", se = FALSE)` to add a straight fitted line without the shaded band. Before plotting, check both variables for missing values and use `filter()` only on the ones that have any --- a variable with nothing missing needs no condition.

[**Pearson's correlation coefficient**]{.important-text}, $r$, summarizes the scatter plot's pattern in a single number: its sign gives the direction (positive means the variables move together; negative means they move in opposite directions), and its size gives the strength (below about .3 is weak; from about .3 to .6 is moderate; above about .6 is strong). Calculate it with `cor(x = data$var1, y = data$var2, use = "pairwise.complete.obs", method = "pearson")`. These thresholds are a starting point --- always report both sign and size rather than mechanically applying cutoffs.

One important limitation: $r$ only captures linear relationships. If a scatter plot reveals a clear curve or some other non-linear pattern, the correlation coefficient can be misleading or meaningless even when a real relationship exists. Always look at the scatter plot before interpreting $r$.

In R, load each package once at the top of your script or Quarto document with `library(ggplot2)` and `library(dplyr)` before any code that uses functions from those packages. You do not need to repeat the `library()` call every time you use a function from it.

From here the course moves to inference --- asking not just whether the observed pattern is consistent with a hypothesis, but whether it would be unlikely to observe a pattern this strong if there were no relationship in the population. Tutorial 10 introduces the logic of inference. You have the descriptive foundation. The rest builds on it.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial.

```{r encode-logic-09, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-09, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-09, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Superb work! You can now describe relationships between two interval variables using scatter plots and Pearson&apos;s correlation coefficient.">
  <span aria-hidden="true">📈</span>
  Superb work! You can now describe relationships between two interval variables using scatter plots and Pearson&apos;s correlation coefficient.
</p>
')
```
