---
title: "Prediction Plots for Interaction Models"
tutorial:
  id: "19-Regression-Predictions-Interactions"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Prediction plots when the lines are not parallel.
---

## Overview {#main-content}

Interaction coefficients show whether a relationship changes across values or categories of another predictor, but they do not always make the resulting pattern easy to see. How much does a predicted outcome differ across realistic cases? Where do group differences widen, narrow, or reverse? Which parts of the pattern are estimated most precisely? Predicted values and plots turn the interaction into comparisons that can be interpreted directly.

This tutorial shows how to calculate and visualize predictions from interaction models. It builds on Tutorial 18, which covers interpreting interaction coefficients and conditional slopes; here, you will use those results to show what the model predicts across substantively meaningful combinations of the predictors.

In this tutorial you will learn:

* How to choose a prediction strategy appropriate to the type of interaction in your model --- when both interacting predictors are interval variables versus when one is categorical
* How to use `predict()` to calculate model-implied outcomes at strategically chosen combinations of the interacting predictors
* How to use `plot_model()` from the [sjPlot]{.package-name} package to visualize conditional relationships for both interval × interval and categorical × interval interactions
* How to interpret non-parallel prediction lines, including changing predicted differences, group gaps, crossings, and the uncertainty around predicted outcomes shown by the confidence bands

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(gradethis)
library(stargazer)
library(sjPlot)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
options(lifecycle_verbosity = "quiet")
knitr::opts_chunk$set(error = TRUE)
data("qog", package = "QPATutorialsCourse")
qog$rights[qog$fh_pr <= 2] <- 0
qog$rights[qog$fh_pr >  2] <- 1
qog$rights_f <- factor(qog$rights, labels = c("Few Rights", "Many Rights"))
data("counties", package = "QPATutorialsCourse")
model_int <- lm(dem2p_percent ~ mortality_risk_25_45 * wage_growth +
                  college_grad_percent + hispanic_percent +
                  white_percent + Black_percent, data = counties)
model_env <- lm(epi_epi ~ rights_f * ipi_tradeopen + hf_efiscore, data = qog)
data("ipu", package = "QPATutorialsCourse")
ipu$electoral_f <- factor(ipu$electoral_system,
                          levels = c("plurality_majority",
                                     "proportional_representation",
                                     "mixed_system"),
                          labels = c("Plurality/Majority", "Proportional", "Mixed"))
ipu$quota_f <- factor(ipu$gender_quota, levels = c(FALSE, TRUE),
                      labels = c("No quota", "Quota"))
ipu$years_c <- ipu$years_since_first_woman -
               mean(ipu$years_since_first_woman, na.rm = TRUE)
model_rep_int <- lm(women_percent ~ quota_f + electoral_f * years_c, data = ipu)
```

## The Prediction Strategy for Interaction Models

Every prediction plot involves a choice: which predictor goes on the x-axis, and what values do the others take? Tutorial 17 makes that choice for additive models. For an additive model, the second choice was mostly a communication decision. Predictions for Southern states at average poverty and proportion non-white, or for Western states at the same values, differ in where the line sits but not in how steep it is. The association between the x-axis predictor and the outcome is the same regardless of the baseline. That is the defining feature of an additive model: one set of estimated associations, visible from any angle.

Interaction models change that. When two variables interact, the estimated association between one of them and the outcome genuinely differs depending on the value of the other. It is not that the line shifts up or down --- the line itself is steeper or shallower. Choosing held values is no longer just a communication decision; it is a substantive one. A prediction plot at low wage growth shows a different association between mortality risk and Clinton's vote than a plot at high wage growth --- not because the baseline differs, but because the model says the association differs there.

That is the new challenge. To communicate what an interaction model implies, you cannot just vary one predictor and plot the result. You need to show how the [predicted difference]{.important-text} --- the model-implied change in the outcome when one predictor moves from low to high --- itself changes as the other interacting variable changes. The right way to do that depends on what kinds of variables are interacting.

Note that interactions are symmetric: either variable can go on the x-axis or serve as the grouping variable. The choice is a [predicted comparison]{.important-text} decision --- which presentation makes the changing association most visible for the research question at hand.

### When Both Interacting Predictors Are Interval Variables

Both variables vary continuously, so there are no natural groups to compare across. Instead, you choose specific, meaningful values of one variable to use as anchors --- typically the 25th percentile, mean, and 75th percentile. At each anchor, you ask: when the other variable moves from low to high, how much does the predicted outcome change? That difference is the model's answer to "how strong is the association here?" Comparing those differences across your chosen anchors reveals whether and how the interaction changes the relationship.

For the plot, you put one variable on the x-axis and show the other at selected values as separate lines. A shorthand explained in the "From Strategy to Code" section automatically draws three lines at the mean minus 1 SD, the mean, and the mean plus 1 SD. If the lines are not parallel --- if the association is steeper for one line than for another --- the interaction is visible directly.

### When One Interacting Predictor Is Categorical

When one variable is categorical, natural groups exist. But in an interaction model, the group difference is not fixed --- it changes across values of the interval predictor. That means you need two things to tell the full story: the group difference at a representative level of the interval predictor, and the predicted difference within each group as the interval predictor varies. Neither alone is enough.

- **Comparison 1** fixes the interval predictor at its mean and computes one predicted value per category. This shows who leads at a typical level of the interval predictor and by how much.
- **Comparison 2** fixes the categorical variable at the category of greatest theoretical interest and varies the interval predictor from low to high. This shows how much the interval predictor matters within that group.

Together, Comparison 1 establishes the starting position and Comparison 2 establishes the trajectory. For the plot, you put the interval variable on the x-axis and use the categorical variable to produce separate colored lines --- one per category.

## Interpreting Interaction Plots

Once you have your predicted differences and your plot, the five-component framework from Tutorial 17 applies: describe the pattern (direction and steepness of each line), state the held values (what covariate profile the plot assumes), give specific predicted values (at least two concrete numbers), assess the confidence band (how certain are the predictions), and connect to the hypothesis (does the pattern support the research question). Interaction plots require one additional element: describe how the predicted difference itself changes as the other interacting variable changes --- which group has the steeper line, whether the lines converge, diverge, or cross.

::: {.tip}
**Non-parallel lines do not confirm that the interaction is statistically significant.** Non-parallel lines appear whenever an interaction term is in the model, whether or not the interaction reaches conventional significance --- they reflect the fitted point estimates, not the strength of evidence. Non-parallel lines also only appear when the plot displays the two variables that actually interact; plotting a control variable against one of the interacting variables would still produce parallel lines. The coefficient test, not the visual appearance of the lines, indicates how strongly the data support a nonzero interaction.
:::

## From Strategy to Code

Now that the strategy is clear, here is the syntax from Tutorial 17 that puts it into practice. Two functions do the work: `predict()` computes specific model-implied outcomes for cases you define, and `plot_model()` from the [sjPlot]{.package-name} package sweeps across the full range of a predictor and draws those predictions as a continuous line. The two functions are complementary --- `predict()` gives you the concrete numbers that anchor your interpretation, and `plot_model()` makes the full conditional relationship visible. If you need a reminder of how they were introduced, the examples in Tutorial 17 walk through them in detail.

### Computing Predicted Values with `predict()`

`predict()` computes what the model predicts for a specific hypothetical case. You supply the model object and a `newdata` argument --- a data frame with one row describing the case, with every predictor named exactly as it appears in the model formula:

```r
predict(model_object, newdata = data.frame(
  predictor1 = value1,
  predictor2 = value2,
  ...
))
```

The output is a single number in the units of the outcome variable.

How you specify each predictor's value depends on its type. For an interval variable, supply a number directly --- `wage_growth = 4.3` --- or use `mean(data$variable, na.rm = TRUE)` to hold it at the sample mean. For a categorical variable coded as a factor, use `factor("Level Name", levels = levels(data$variable))`. Factor variables in R carry metadata about all their possible levels; the explicit `factor()` construction attaches the full level set that the model expects, which avoids edge cases where R cannot determine the levels from context alone.

To compare two cases, write two separate `predict()` calls, changing only the values that differ between them. Everything else stays identical.

### Visualizing with `plot_model()`

`plot_model()` from the [sjPlot]{.package-name} package sweeps across the full range of a predictor and draws the model's predicted values as a continuous line or a set of points with confidence intervals. Because `plot_model()` returns a ggplot object, any ggplot layer --- including `labs()`, `theme()`, and others --- can be chained to it with `+`.

The `type = "pred"` argument requests a prediction plot rather than a coefficient plot. The `terms =` argument names the predictor(s) to display; unlike most R arguments, variable names here are **character strings in quotes**. `axis.title = c("x label", "y label")` labels the axes, x first then y.

For a single predictor on the x-axis --- `terms = "x"` --- all other predictors are held at their means (for interval variables) or reference category (for factor variables). The caption should always state these held values explicitly: `labs(caption = "Predictions are for...")`, ending with a period.

For two predictors --- `terms = c("x", "grouping")` --- a second variable produces separate lines. If the grouping variable is categorical, one line appears per category, colored and labeled by category name; use `labs(color = "Label")` to rename the legend from the variable name to something readable. If the grouping variable is interval, bracket notation controls which values appear as separate lines: `"variable [meansd]"` automatically draws three lines at the mean minus 1 SD, the mean, and the mean plus 1 SD --- a convenient way to show how the relationship changes across a realistic range without choosing arbitrary values. You can also supply specific values: `"variable [v1, v2, v3]"`.

To override any default held value, use the `condition` argument, introduced in Tutorial 17: `condition = list(variable_name = value)`. Example 3 below puts it to work.

## Example 1: Mortality Risk, Wage Growth, and Clinton's Vote

Does the estimated association between county-level mortality risk and Clinton's vote depend on local economic conditions? Tutorial 18's first example estimates exactly that: an interaction between two interval variables --- **mortality_risk_25_45** and **wage_growth** --- predicting Clinton's share of the two-party vote (**dem2p_percent**). The model is available here as **model_int**.

The theoretical argument from Tutorial 18: in counties where wages are growing, residents may see economic opportunity as available to those willing to work and attribute poor health outcomes to individual choices --- weakening the connection between community health and political backlash. In counties where wages are stagnant or declining, those same health outcomes may feel more like a structural failure --- producing a stronger political response against the incumbent party's candidate. If this reasoning is right, the estimated association between mortality risk and Clinton's vote should be more negative where wage growth is lower, and less negative where wage growth is higher. Tutorial 18 found the opposite; the plots below show what that reversal looks like.

The data frame **counties** describes every U.S. county as of 2016. Of its 3,112 counties, **dem2p_percent** has no missing values, **mortality_risk_25_45** is missing for 1, and **wage_growth** for 2. The four demographic controls are each missing for one county except **Black_percent**, which is missing for two. It contains:

- **dem2p_percent**: Clinton's percent of the two-party vote
- **mortality_risk_25_45**: the estimated percentage risk that a person who has reached age 25 will die before reaching age 45
- **wage_growth**: recent wage growth (percent; negative = declining wages)
- **college_grad_percent**, **hispanic_percent**, **white_percent**, **Black_percent**: control variables

::: {.hypothesis}
In a comparison of counties, the negative association between mortality risk and Hillary Clinton's percent of the two-party vote is stronger where wage growth is lower.
:::

The model output is reproduced below. Before reading on, locate the interaction coefficient (`mortality_risk_25_45:wage_growth`) and recall the two rules from Tutorial 18 for computing conditional slopes from it.

```{r model-int-table, echo=TRUE}
stargazer(model_int,
          type = "text",
          title = "Mortality Risk, Wage Growth, and Clinton's Vote",
          dep.var.labels = "Clinton Two-Party Vote (%)",
          covariate.labels = c("Mortality Risk",
                               "Wage Growth",
                               "College Graduates (%)",
                               "Hispanic (%)",
                               "White (%)",
                               "Black (%)",
                               "Mortality Risk x Wage Growth"))
```

**Reading the predicted differences from Tutorial 18.** Tutorial 18 computes the predicted change in Clinton's vote for a one-unit increase in mortality risk at specific wage growth values:

- The mortality risk association **when wage growth = 0**: −2.494 (the main effect)
- The interaction coefficient: −0.139 (how much the mortality risk association changes per unit of wage growth)
- At any specific wage growth value: association = −2.494 + (−0.139) × wage_growth
- At the mean wage growth (≈ 4.3): −2.494 + (−0.139)(4.3) ≈ **−3.09**
- At low wage growth (≈ −0.35, one standard deviation below the mean): −2.494 + (−0.139)(−0.35) ≈ **−2.45**
- At high wage growth (≈ 9.0, one standard deviation above the mean): −2.494 + (−0.139)(9.0) ≈ **−3.74**

These are per-unit figures from the coefficient table. In Tutorial 18, these conditional slopes described how sharply Clinton's predicted vote changes with mortality risk at different wage growth values. Here, we translate those slopes into predicted vote shares and predicted differences for realistic county profiles.

### Obtain Specific Predicted Values

Following the interval × interval strategy: we compute two predicted values at each of two wage growth levels --- holding all other predictors at their means --- to make the changing predicted difference concrete.

**Comparison 1** fixes wage growth at a low value (≈ −4.7, 25th percentile) and varies mortality risk from 2.0 (near its 25th percentile) to 5.0 (near its 75th percentile), holding all controls at their means. The difference between the two predicted values is the model-implied change in Clinton's predicted vote when mortality risk moves from low to high in counties with weak or declining wages.

**Comparison 2** does the same at a high wage growth value (≈ 12.7, 75th percentile). Comparing the two predicted differences --- one from Comparison 1, one from Comparison 2 --- shows directly how much the interaction changes the relationship between mortality risk and Clinton's vote across different economic contexts.

**Run and observe 1: Compute four predicted values for model_int**

Run the code and note the difference between the two predicted values within each comparison before answering the question below.

```{r exercise-predict-int, exercise=TRUE, exercise.cap="Run and observe 1"}
# Comparison 1: mortality risk slope at low wage growth (-4.7)
predict(model_int, newdata = data.frame(
  mortality_risk_25_45  = 2.0,
  wage_growth           = -4.7,
  college_grad_percent  = mean(counties$college_grad_percent,  na.rm = TRUE),
  hispanic_percent      = mean(counties$hispanic_percent,       na.rm = TRUE),
  white_percent         = mean(counties$white_percent,          na.rm = TRUE),
  Black_percent         = mean(counties$Black_percent,          na.rm = TRUE)
))
predict(model_int, newdata = data.frame(
  mortality_risk_25_45  = 5.0,
  wage_growth           = -4.7,
  college_grad_percent  = mean(counties$college_grad_percent,  na.rm = TRUE),
  hispanic_percent      = mean(counties$hispanic_percent,       na.rm = TRUE),
  white_percent         = mean(counties$white_percent,          na.rm = TRUE),
  Black_percent         = mean(counties$Black_percent,          na.rm = TRUE)
))

# Comparison 2: mortality risk slope at high wage growth (12.7)
predict(model_int, newdata = data.frame(
  mortality_risk_25_45  = 2.0,
  wage_growth           = 12.7,
  college_grad_percent  = mean(counties$college_grad_percent,  na.rm = TRUE),
  hispanic_percent      = mean(counties$hispanic_percent,       na.rm = TRUE),
  white_percent         = mean(counties$white_percent,          na.rm = TRUE),
  Black_percent         = mean(counties$Black_percent,          na.rm = TRUE)
))
predict(model_int, newdata = data.frame(
  mortality_risk_25_45  = 5.0,
  wage_growth           = 12.7,
  college_grad_percent  = mean(counties$college_grad_percent,  na.rm = TRUE),
  hispanic_percent      = mean(counties$hispanic_percent,       na.rm = TRUE),
  white_percent         = mean(counties$white_percent,          na.rm = TRUE),
  Black_percent         = mean(counties$Black_percent,          na.rm = TRUE)
))
```

**Apply 2: Interpret the mortality risk comparisons**

Compute the difference between the two values within each comparison before answering.

```{r quiz-predict-int, echo=FALSE}
question("Your four predicted values give you two predicted differences: the model-implied change in Clinton's vote when moving from low to high mortality risk at low wage growth, and the same at high wage growth. Which statement best describes what those two predicted differences together reveal?",
  correct = "Right. The predicted difference is larger at high wage growth --- mortality risk predicts a bigger drop where wages are growing faster.",
  answer("The predicted drop in Clinton's vote as mortality risk rises is larger in counties with high wage growth than in counties with low wage growth",
         correct = TRUE,
         message = "Subtracting within each comparison: in Comparison 1 (low wage growth, ≈ −4.7), moving from mortality risk 2.0 to 5.0 predicts a drop of about 5.5 percentage points. In Comparison 2 (high wage growth, ≈ 12.7), the same move predicts a drop of about 12.7 percentage points. The interaction changes the predicted difference by about 7 percentage points across those two wage growth levels --- exactly what the interaction coefficient of −0.139 quantifies per unit of wage growth."),
  answer("The predicted change in Clinton's vote is nearly identical in both comparisons, suggesting wage growth does not meaningfully change the relationship between mortality risk and Clinton's vote",
         message = "Compute the within-comparison differences: subtract the first predicted value from the second in each comparison. The two differences are substantially different, reflecting the significant interaction from Tutorial 18."),
  answer("Comparison 1 shows the predicted change for low-mortality-risk counties and Comparison 2 shows it for high-mortality-risk counties --- the comparisons are about different types of counties, not different wage growth contexts",
         message = "Each comparison holds the wage growth value fixed and varies mortality risk. Comparison 1 is for low-wage-growth counties; Comparison 2 is for high-wage-growth counties. Within each, mortality risk varies from 2.0 to 5.0."),
  answer("The predicted change in Clinton's vote is larger at low wage growth, meaning mortality risk is associated with a bigger drop in Clinton support where wages are declining",
         message = "Check the direction: the interaction coefficient is negative (−0.139), meaning the predicted difference becomes more negative as wage growth increases, not decreases. Subtract within each comparison to verify which gives the larger predicted drop."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: subtract the first predicted value from the second in each comparison. Which comparison gives a larger drop? That group has the steeper mortality risk slope."
)
```

### Visualize the Interaction

After calculating the changing predicted difference numerically, the plot makes it visible across the full range of mortality risk. Using `[meansd]` bracket notation, `plot_model()` automatically draws three lines: one for counties at the mean minus 1 SD of wage growth, one at the mean, and one at the mean plus 1 SD. Non-parallel lines show the interaction directly.

**Practice 3: Produce the interaction plot for model_int**

Fill in the blanks to produce the three-line interaction plot, using **model_int** as the model object. The bracket notation `"wage_growth [meansd]"` tells `plot_model()` to show lines at the mean minus 1 SD, mean, and mean plus 1 SD of wage growth. Use `"Wage Growth (%)"` as the color legend label. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-int, exercise=TRUE, exercise.cap="Practice 3", fig.alt="Predicted Clinton two-party vote against mortality risk, with one line for each of three wage-growth values. The horizontal axis runs from 0 to 10 and the vertical axis from about 10 to 50. All three lines fall from left to right, each with a shaded band, and a legend gives the three wage-growth values."}
library(sjPlot)
___(model_int,
    type = "pred",
    terms = c("mortality_risk_25_45", "___"),
    title = "Predicted Clinton Vote by Mortality Risk and Wage Growth",
    axis.title = c("Mortality Risk (%)",
                   "Clinton Two-Party Vote (%)")) +
  labs(color = "___",
       caption = "Predictions hold all controls at their means.")
```

```{r exercise-predplot-int-hint-1}
# Hint 1 of 3
# This is a two-variable prediction plot using the interval x interval
# strategy from the strategy section. The focal predictor (mortality_risk_25_45)
# goes on the x-axis. The second term uses bracket notation to show
# wage_growth at selected values: "wage_growth [meansd]" produces
# three lines at mean - 1SD, mean, and mean + 1SD automatically.
# The function name and color legend label are the two remaining blanks.
```

```{r exercise-predplot-int-hint-2}
# Hint 2 of 3
# The function is plot_model(). The second term in terms = is
# "wage_growth [meansd]" --- include the brackets and meansd exactly
# as shown, inside the quotes with the variable name.
# The color legend should be labeled "Wage Growth (%)" to match the
# variable's units.
```

```{r exercise-predplot-int-hint-3}
# Hint 3 of 3
# Both variable names go inside terms = c(...), the interval one
# first. The bracket notation attaches to the second name, inside
# its quotes. The legend title is set with color = in labs(),
# alongside the caption.
```

```{r exercise-predplot-int-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("wage_growth", code_nc))
    fail("Include **wage_growth** as the grouping variable in terms.")
  if (!grepl("\\[meansd\\]", code_nc))
    fail("Use bracket notation [meansd] with **wage_growth** to show lines at mean ± 1 SD.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a `labs()` layer with color = and caption = arguments.")
  if (!grepl("Wage Growth", code_nc, fixed = TRUE))
    fail("Use labs(color = \"Wage Growth (%)\") to label the legend.")
  pass("Well done. The plot draws one line per wage-growth value, with the controls held at their means as the caption states.")
})
```

Since [sjPlot]{.package-name} was loaded in the exercise above, you can assume it is already loaded for all subsequent exercises in this tutorial. In your own scripts, call `library(sjPlot)` once at the top before using `plot_model()`.

**Apply 4: Interpret the mortality risk interaction plot**

Read the plot carefully --- look at the direction and steepness of each line, the specific predicted values you can read at the ends of the lines, the confidence bands, and whether the lines keep the same order from one side to the other.

```{r quiz-predplot-int-pattern, echo=FALSE}
question("Which statement best describes how the predicted drop in Clinton's vote when mortality risk moves from low to high changes as wage growth increases?",
  correct = "Right. The predicted drop is largest where wage growth is highest --- steep enough that the lines cross.",
  answer("The predicted drop is largest in high-wage-growth counties and smallest in low-wage-growth counties",
         correct = TRUE,
         message = "The green line (highest wage growth ≈ 9%) falls most steeply, the red line (lowest wage growth ≈ −0.35%) falls least steeply. Because the green line starts highest and falls fastest, the three lines converge and cross at about 5% mortality risk, then reopen with the order reversed. Below that point high-wage-growth counties are predicted highest; above it, lowest."),
  answer("The predicted drop is the same in all three wage growth groups --- the lines are parallel",
         message = "Look at how much each line falls from the left to the right edge of the plot. The three falls are different sizes --- the lines are not parallel, which is the whole point of the interaction."),
  answer("The predicted drop is largest in low-wage-growth counties --- economic hardship amplifies the association between mortality risk and Clinton's vote",
         message = "Check which line is steepest. The red line (lowest wage growth) is the shallowest; the green line (highest wage growth) is the steepest. The interaction coefficient is negative, meaning higher wage growth produces a more negative predicted difference in the fitted model."),
  answer("The lines converge as mortality risk increases, showing that the wage growth difference in predicted Clinton vote disappears at high mortality risk",
         message = "They do converge at first --- but only until about 5% mortality risk, where they cross. Past that point the gap reopens with the ranking reversed, so the difference does not disappear. It changes sign."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: which line falls most steeply from left to right? And check whether the lines stay in the same order across the whole plot."
)
```

**Apply 5: Read predicted values off the interaction plot**

```{r quiz-predplot-int-values, echo=FALSE}
question("Using the mean wage growth line, approximately what is the predicted Clinton vote at mortality risk = 2.5 and at mortality risk = 7.5? What does the caption tell you is held constant?",
  correct = "Right. About 35% at low mortality risk and about 20% at high --- a drop of roughly 15 points. All controls at their means.",
  answer("At mortality risk 2.5 the predicted vote is about 35% and at 7.5 about 20% --- a drop of roughly 15 points",
         correct = TRUE,
         message = "Reading from the blue (mean wage growth) line: at mortality risk 2.5 the predicted vote is near 35%, and at 7.5 it is near 20%. The 15-point drop across a realistic range of mortality risk makes the predicted difference concrete. The caption confirms that college graduation rate, Hispanic, white, and Black percentages are all held at their means."),
  answer("The plot does not show specific values --- only the direction of the relationship can be read from it",
         message = "The y-axis has a numerical scale and the lines are continuous --- you can read approximate predicted values at any mortality risk level by finding the x-axis value, tracing up to the line, and reading the y-axis. Specific predicted values are one of the five required components of a complete interpretation."),
  answer("At mortality risk 2.5 the predicted vote is approximately 47% and at 7.5 approximately 20% --- a drop of about 27 percentage points; the caption states all control variables are held at their means",
         message = "Check the y-axis scale more carefully at those two x-axis positions for the mean wage growth line. The line sits closer to 35% at the low end and 20% at the high end --- a drop of about 15 points, not 27."),
  answer("The predicted vote is approximately 30% at both values --- the mean wage growth line is essentially flat",
         message = "All three lines slope downward --- none is flat. Trace the mean wage growth line (blue) from mortality risk 2.5 to 7.5 and read the y-axis values at each point."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: find mortality risk = 2.5 on the x-axis, trace up to the blue (mean) line, and read the y-axis. Repeat at mortality risk = 7.5."
)
```

**Apply 6: Read the confidence bands**

```{r quiz-predplot-int-uncertainty, echo=FALSE}
question("Look at the confidence bands in the plot. Where are the expected vote shares estimated more and less precisely?",
  correct = "Right. Bands are narrower at low mortality risk and wider at high mortality risk --- the expected vote shares are estimated more precisely at the low end.",
  answer("The bands are narrower at low mortality risk and wider at high, so predictions are more precise at the low end",
         correct = TRUE,
         message = "Band width reflects uncertainty around the expected outcome for counties with each predictor profile. At low mortality risk, the bands are narrow, so the model estimates the average vote share for those counties more precisely. At high mortality risk, the bands widen, indicating greater uncertainty."),
  answer("The bands are equally wide across the full mortality risk range, so the uncertainty in the predicted values is the same at every point",
         message = "Look at the left and right ends of the plot. The bands are visibly narrower at low mortality risk and wider at high mortality risk --- confidence bands for prediction plots are typically narrower where data are denser and wider where data are sparser."),
  answer("The bands are widest at low mortality risk where the lines are most separated, confirming the interaction is most reliable at the high end of the scale",
         message = "This reverses what the plot shows. The bands are narrowest at low mortality risk --- where the lines are most separated --- and widest at high mortality risk, past the crossing point, where the lines sit close together in reversed order. Look at the shaded regions at the far left and far right of the plot."),
  answer("The bands tell you only about overall model fit --- wider bands mean lower R-squared and narrower bands mean higher R-squared",
         message = "Confidence bands in a prediction plot reflect uncertainty around the expected outcome at each specific predictor profile, not overall model fit. R-squared is a single number summarizing overall fit; the bands vary across the x-axis depending on data density and the covariance structure of the estimates."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the left side of the plot versus the right side. Where are the bands narrowest, indicating more precise estimates of the expected vote share? Where do the bands widen, indicating less precision?"
)
```

### What We Learned About Mortality Risk, Wage Growth, and the Vote

The interaction plot makes visible what the coefficient table only hints at: the predicted change in Clinton's vote as mortality risk increases is not the same everywhere --- it varies with local economic conditions. Whether this reflects the political meaning of health outcomes differing under different economic conditions is a substantive question the regression cannot answer; it establishes the estimated pattern, not the mechanism.

Here is what a complete interpretation looks like, working through all six components from the Interpreting Interaction Plots section above:

::: {.research-example}
*As mortality risk increases, predicted Clinton two-party vote share declines in all three wage growth groups, but the steepness of that decline differs. All control variables are held at their means throughout. At low mortality risk (≈ 2.5%), the three lines are clearly separated and ordered by wage growth: counties at high wage growth (≈ 9%) are predicted to give Clinton about 37% of the two-party vote, counties at the mean (≈ 4.3%) about 35%, and counties at low wage growth (≈ −0.35%) about 33%. As mortality risk rises to approximately 7.5%, all three fall to around 18 to 21% --- and the order reverses. The high-wage-growth line, highest at low mortality risk, is now the lowest; the low-wage-growth line is now the highest. That reversal is what the interaction amounts to: the lines are not merely converging on a common value, they are crossing, because the mortality-risk slope is steeper where wage growth is stronger. The confidence bands also widen substantially, indicating greater uncertainty about the expected vote shares at high mortality risk. The interaction is statistically significant, but it runs against the argument that motivated it: the predicted decline is steepest where wage growth is strongest, not weakest. One possibility is that rising wages sharpen the contrast between apparent prosperity and persistent poor health; another is that county wage growth is an average, so high mortality alongside rising wages may mark places where the gains are unevenly shared. Distinguishing those would require evidence this model does not contain.*
:::

## Example 2: Political Rights, Trade Openness, and Environmental Performance

Does the estimated association between international trade and environmental performance depend on a country's political context? Tutorial 18's second example estimates exactly that: a categorical × interval interaction between trade openness (**ipi_tradeopen**) and political rights status (**rights_f**), predicting environmental performance (**epi_epi**). The model is available here as **model_env**.

The theoretical argument from Tutorial 18: in countries with many political rights, civil society, free media, and political competition give citizens tools to hold governments and firms accountable for environmental outcomes regardless of trade levels. In countries with few political rights, those domestic accountability mechanisms are weak or absent. International economic integration may serve as one of the few external pressures that can improve environmental behavior --- making the trade-environment association stronger where domestic accountability is limited.

::: {.hypothesis}
The estimated association between trade openness and environmental performance is stronger in countries with few political rights than in countries with many political rights.
:::

The data frame **qog** is the January 2020 cross-section of the Quality of Government Basic Dataset, in which each variable takes its most recent available observation as of that release. Of its 194 countries, **ipi_tradeopen** is missing for 78, **hf_efiscore** for 17, **epi_epi** for 14, and **fh_pr** for none. It contains:

- **epi_epi**: the Environmental Performance Index, scaled 0 to 100 and observed from 27.4 to 87.4, with higher values indicating better environmental performance
- **rights_f**: two-category factor --- **Few Rights** (reference) and **Many Rights**
- **ipi_tradeopen**: the trade-openness component of the Index of Public Integrity, scored 1 to 10, with higher values indicating greater openness
- **hf_efiscore**: the Heritage Foundation Economic Freedom Index, on a theoretical scale of 0 to 100 and observed from 2.3 to 87.8 --- included as a control

The model output is reproduced below.

```{r model-env-table, echo=TRUE}
stargazer(model_env,
          type = "text",
          title = "Trade Openness, Political Rights, and Environmental Performance",
          dep.var.labels = "Environmental Performance Index",
          covariate.labels = c("Many Rights (vs. Few Rights)",
                               "Trade Openness",
                               "Economic Freedom Score",
                               "Many Rights x Trade Openness"))
```

**Reading the conditional slopes from the table.** Tutorial 18 computes:

- **Few Rights slope** (reference group, main effect only): approximately **4.75** per unit of trade openness
- **Many Rights slope** (main effect + interaction): 4.75 + (−4.63) ≈ **0.12** per unit of trade openness

In Tutorial 18, these conditional slopes showed how the trade-openness association differs by rights status. Here, predicted values show what those different slopes imply for countries at substantively meaningful levels of trade openness.

### Obtain Specific Predicted Values

Following the categorical × interval strategy, two comparisons anchor the interpretation.

**Comparison 1** asks: at average trade openness (**ipi_tradeopen** ≈ 7.77) and average economic freedom (**hf_efiscore** ≈ 60.26), how much does **epi_epi** differ between Few Rights and Many Rights countries? This shows which group leads at a typical level of trade openness.

**Comparison 2** asks: within Few Rights countries at average economic freedom, how much does **ipi_tradeopen** matter? Following the categorical × interval strategy from the Prediction Strategy section, this comparison fixes the categorical variable at the category of greatest theoretical interest --- Few Rights, where the interaction hypothesis predicts the steepest association --- and varies the interval predictor from low (6.9, the 25th percentile) to high (9.4, the 75th percentile). The difference between the two predicted values captures the estimated association within that group across a realistic range of trade openness.

The first call has partial starter code with a blank to fill in; the remaining three are yours to write from scratch.

**Practice 7: Compute four predicted values**

Write the four `predict()` calls described above, using **model_env**. Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predict-env, exercise=TRUE, exercise.cap="Practice 7"}
# Comparison 1: Few Rights vs. Many Rights at average trade openness
predict(model_env, newdata = data.frame(
  rights_f      = factor("___", levels = levels(qog$rights_f)),
  ipi_tradeopen = mean(qog$ipi_tradeopen, na.rm = TRUE),
  hf_efiscore   = mean(qog$hf_efiscore,   na.rm = TRUE)
))

# Comparison 1: Many Rights

# Comparison 2: Few Rights at low trade openness (ipi_tradeopen = 6.9)

# Comparison 2: Few Rights at high trade openness (ipi_tradeopen = 9.4)

```

```{r exercise-predict-env-hint-1}
# Hint 1 of 3
# A predicted value answers a what-if: for a country with this
# particular combination of characteristics, what does the model
# expect? Every predictor in the model needs a value, including
# the ones being held fixed rather than varied.
```

```{r exercise-predict-env-hint-2}
# Hint 2 of 3
# All four calls use predict() with model_env and a newdata = data.frame(...)
# containing three predictors: rights_f, ipi_tradeopen, and hf_efiscore.
# For rights_f, use factor("level name", levels = levels(qog$rights_f)).
# Comparison 1: ipi_tradeopen and hf_efiscore stay at their means across both calls;
#   only the rights_f level changes ("Few Rights" then "Many Rights").
# Comparison 2: rights_f = "Few Rights" for both calls;
#   hf_efiscore at its mean; ipi_tradeopen = 6.9 then 9.4.
```

```{r exercise-predict-env-hint-3}
# Hint 3 of 3
# Comparison 1 values:
#   First call:  rights_f = "Few Rights", ipi_tradeopen = mean(...), hf_efiscore = mean(...)
#   Second call: rights_f = "Many Rights", ipi_tradeopen = mean(...), hf_efiscore = mean(...)
# Comparison 2 values:
#   Third call:  rights_f = "Few Rights", ipi_tradeopen = 6.9, hf_efiscore = mean(...)
#   Fourth call: rights_f = "Few Rights", ipi_tradeopen = 9.4, hf_efiscore = mean(...)
```

```{r exercise-predict-env-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("predict\\s*\\(", code_nc))
    fail("Use `predict()` to obtain predicted values.")
  if (!grepl("model_env", code_nc))
    fail("Pass **model_env** as the model object.")
  if (!grepl("rights_f", code_nc))
    fail("Include **rights_f** in your newdata data frame.")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Include **ipi_tradeopen** in your newdata data frame.")
  if (!grepl("hf_efiscore", code_nc))
    fail("Include **hf_efiscore** in your newdata data frame.")
  if (!grepl("Few Rights", code_nc, fixed = TRUE))
    fail("Include predictions for Few Rights countries.")
  if (!grepl("Many Rights", code_nc, fixed = TRUE))
    fail("Include a prediction for Many Rights countries (Comparison 1).")
  # Parse the submission into top-level expressions so each call can be graded on
  # its own. `.result` holds only the last one, and all four are being asked for.
  exprs <- tryCatch(as.list(parse(text = code_nc)), error = function(e) NULL)
  if (is.null(exprs))
    fail("Your code could not be parsed. Check for a missing comma or bracket.")
  is_pred <- function(x) {
    if (is.call(x) && identical(deparse(x[[1]]), "<-") && length(x) == 3L) x <- x[[3]]
    is.call(x) && identical(deparse(x[[1]]), "predict")
  }
  as_pred <- function(x) {
    if (is.call(x) && identical(deparse(x[[1]]), "<-") && length(x) == 3L) x[[3]] else x
  }
  pcalls <- lapply(Filter(is_pred, exprs), as_pred)
  if (length(pcalls) != 4)
    fail(paste0("Write four `predict()` calls, one per predicted value: two for Comparison 1 and ",
                "two for Comparison 2. Your code has ", length(pcalls), "."))

  mn   <- mean(qog$hf_efiscore, na.rm = TRUE)
  prof <- data.frame(
    rights_f      = factor(c("Few Rights", "Many Rights", "Few Rights", "Few Rights"),
                           levels = levels(qog$rights_f)),
    ipi_tradeopen = c(7.77, 7.77, 6.9, 9.4),
    hf_efiscore   = mn)
  want <- as.numeric(predict(model_env, newdata = prof))

  # Evaluate each call rather than matching its text, so 7.77 and
  # mean(qog$ipi_tradeopen, na.rm = TRUE) are both accepted.
  got <- suppressWarnings(vapply(pcalls, function(cl) {
    v <- tryCatch(as.numeric(eval(cl))[1], error = function(e) NA_real_)
    if (length(v) == 0L) NA_real_ else v
  }, numeric(1)))
  labels <- c("Few Rights at average trade openness",
              "Many Rights at average trade openness",
              "Few Rights at **ipi_tradeopen** 6.9",
              "Few Rights at **ipi_tradeopen** 9.4")
  for (i in seq_len(4)) {
    if (is.na(got[i]) || abs(got[i] - want[i]) > 0.05)
      fail(paste0("Call ", i, " does not give the predicted value for ", labels[i],
                  ", with **hf_efiscore** held at its mean. Check the profile in that call."))
  }

  pass("Right. Four predicted values, one for each combination of rights status and trade openness.")
})
```

**Apply 8: Interpret the trade openness comparisons**

```{r quiz-predict-env, echo=FALSE}
question("Your four predicted values show the interaction pattern concretely. Which statement best describes what Comparison 1 and Comparison 2 together reveal?",
  correct = "Right. Who leads at the mean, and how steep the Few Rights slope is.",
  answer("At mean trade openness Few Rights countries are predicted to outperform Many Rights; within Few Rights, moving from low to high trade openness predicts a large further increase",
         correct = TRUE,
         message = "Comparison 1 is a snapshot of the group difference at the mean --- it shows who leads and by how much at a typical trade level. Comparison 2 isolates the Few Rights slope across a realistic range of trade openness. Together they tell the story the interaction plot will show continuously."),
  answer("The two comparisons show the same pattern --- both Comparison 1 and Comparison 2 describe the group difference between Few Rights and Many Rights countries at different trade openness levels",
         message = "They answer different questions. Comparison 1 shows the predicted group difference at one trade openness value. Comparison 2 shows how much trade openness itself matters within the Few Rights group. Both are needed to characterize the full interaction pattern."),
  answer("Comparison 1 is a formal significance test of whether the **rights_f** main effect is statistically significant, while Comparison 2 tests whether trade openness matters overall",
         message = "Neither comparison is a significance test. Both are model-implied predicted differences at specific covariate values. The statistical evidence for the interaction comes from the coefficient p-value."),
  answer("Many Rights countries are always predicted to outperform Few Rights countries on environmental performance regardless of the trade openness level",
         message = "Comparison 1 shows Few Rights countries are predicted to outperform Many Rights countries at mean trade openness. The advantage of political rights is largest at low trade openness; as trade rises, the steep Few Rights slope closes and reverses the gap."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Comparison 1 shows position (who leads at the mean); Comparison 2 shows trajectory (how steep is the Few Rights slope). What does each tell you?"
)
```

### Visualize the Few Rights Conditional Slope

Before looking at the full interaction, let's see the Few Rights predicted association on its own. Because **Few Rights** is the reference category, a single-variable prediction plot for **ipi_tradeopen** shows that group's fitted relationship while holding **hf_efiscore** at its mean.

**Run and observe 9: Few Rights predicted association**

```{r exercise-predplot-trade, exercise=TRUE, exercise.cap="Run and observe 9", fig.alt="Predicted environmental performance against trade openness. The horizontal axis runs from about 2 to 10 and the vertical axis from about 20 to 70. A straight line rises from left to right, with a shaded band that is widest at the left."}
plot_model(model_env,
           type = "pred",
           terms = "ipi_tradeopen",
           title = "Predicted Environmental Performance by Trade Openness",
           axis.title = c("Trade Openness (1-10)",
                          "Environmental Performance Index"),
           show.data = FALSE) +
  labs(caption = "Predictions are for Few Rights countries with average economic freedom.")
```

**Apply 10: Interpret the Few Rights predicted association**

```{r quiz-trade-interpret, echo=FALSE}
question("The prediction plot shows a steep upward line for Few Rights countries. Which statement best describes what it shows?",
  correct = "Right. A steep positive estimated association for the Few Rights reference group.",
  answer("For Few Rights countries, the model predicts a large rise in environmental performance across the range",
         correct = TRUE,
         message = "The Few Rights estimated association is approximately 4.75 EPI points per unit of trade openness, against a near-flat 0.12 for Many Rights countries."),
  answer("The upward line means trade openness causes better environmental outcomes in Few Rights countries",
         message = "A positive estimated association is not evidence of causation. Trade openness may be correlated with other factors that also predict better environmental performance."),
  answer("This plot shows the trade openness association averaged across both rights groups, not just Few Rights",
         message = "Because Few Rights is the reference category, `plot_model()` holds **rights_f** at Few Rights by default. This shows the Few Rights association only, not an average across both groups."),
  answer("The line would be identical for Many Rights countries because the model estimates one trade openness coefficient",
         message = "The interaction term creates a different estimated association for each group, which is why the two groups do not share a single slope."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: what does the caption tell you about which group this plot shows?"
)
```

### Compare Two Orientations of the Interaction Plot

There are two ways to display the interaction: categorical variable on the x-axis (showing predicted values for each group at selected levels of the interval variable), or interval variable on the x-axis (showing continuous lines for each group across the full range). Both are informative; the preferred orientation depends on what you want to communicate.

**Run and observe 11: Rights status on the x-axis**

```{r exercise-predplot-int1, exercise=TRUE, exercise.cap="Run and observe 11", fig.alt="Predicted environmental performance with political rights status on the horizontal axis, in two categories labeled Few Rights and Many Rights. At each category three predicted values are plotted as points with vertical confidence intervals, one for each of the three trade-openness values named in the legend. The vertical axis spans roughly 45 to 72."}
plot_model(model_env,
           type = "pred",
           terms = c("rights_f", "ipi_tradeopen"),
           title = "Environmental Performance by Rights Status",
           axis.title = c("Political Rights Status",
                          "Environmental Performance Index")) +
  labs(caption = "Predictions are for countries with average economic freedom.")
```

You may see a short message above the plot beginning `Ignoring unknown labels`, naming `linetype` and `shape`. It is not an error, and nothing in your code caused it. `plot_model()` does not know in advance which visual channel will distinguish the groups --- color, line type, or point shape --- so it prepares a legend label for all three. This plot puts a categorical variable on the x-axis, so it draws points with intervals rather than lines; `linetype` and `shape` therefore go unused, and recent versions of [ggplot2]{.package-name} report labels supplied for aesthetics a plot does not use. Whether the message appears at all depends on your [ggplot2]{.package-name} version.

**Apply 12: Evaluate this orientation**

```{r quiz-int1-interpret, echo=FALSE}
question("The plot shows predicted values for Few Rights and Many Rights countries at three selected trade openness levels. What can this orientation show, and what can it not show?",
  correct = "Right. Discrete snapshots at three trade values, but not how the estimated association evolves continuously.",
  answer("It shows the group gap at three selected trade openness values, not how it evolves across the full range",
         correct = TRUE,
         message = "Three selected values give you snapshots of the group difference at those points. You can see the reversal, but you cannot read the predicted differences directly across the full range. For that, trade openness needs to be on the x-axis."),
  answer("This orientation is always preferred when one variable is categorical, because it makes the group comparison easiest to read",
         message = "The choice depends on what you want to communicate. Putting the interval variable on the x-axis shows the full continuous range and is more informative for an interaction interpretation."),
  answer("Because the confidence intervals overlap at the middle trade openness value, the interaction is not statistically significant",
         message = "Each confidence interval shows uncertainty around its own expected outcome. Whether the interaction is statistically distinguishable from zero comes from the interaction-coefficient test, not from whether these separate intervals overlap."),
  answer("This orientation and the continuous-axis orientation show exactly the same information in different visual formats",
         message = "They show related but not identical information. This orientation gives snapshots at three trade values; the continuous orientation shows the full range for each group."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: how many trade openness values can you read from this plot? What would you need to see the slope across the full range?"
)
```

### The Preferred Orientation: Trade Openness on the x-axis

With **ipi_tradeopen** on the x-axis and **rights_f** producing separate lines, the full conditional relationship becomes visible. Non-parallel lines make the interaction visible directly.

**Practice 13: Produce the interaction plot**

Write the complete `plot_model()` call from scratch, using **model_env** as the model object. Use the title \"Predicted Environmental Performance\\nby Trade Openness and Rights Status\" --- the `\n` breaks it across two lines so it is not cut off at the width of the plot, the x-axis label \"Trade Openness (1-10)\", the y-axis label \"Environmental Performance Index\", and the caption \"Predictions are for countries with average economic freedom.\" `plot_model()` maps the grouping variable to the color aesthetic, so without a `color =` argument in `labs()`, the variable name appears as the legend title. Use `labs(color = "Rights Status", caption = "...")` to set both. Four hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-int2, exercise=TRUE, exercise.cap="Practice 13", fig.alt="Predicted environmental performance against trade openness, with one line for each rights status. The horizontal axis runs from about 2 to 10 and the vertical axis from about 20 to 70. Each line has a shaded band, and a legend names the two categories, Few Rights and Many Rights."}

```

```{r exercise-predplot-int2-hint-1}
# Hint 1 of 4
# This is a two-variable prediction plot --- the same structure introduced
# in Tutorial 17. The interval variable (ipi_tradeopen) goes on the x-axis
# and the categorical variable (rights_f) produces separate colored lines.
# The interval variable is the one that goes first.
```

```{r exercise-predplot-int2-hint-2}
# Hint 2 of 4
# Use plot_model() with type = "pred".
# Give terms both variable names in a character vector, interval first.
# No bracket notation is needed here, because the second variable is a
# two-category factor rather than an interval variable.
# Chain labs(color = "Rights Status", caption = "...") after with +.
```

```{r exercise-predplot-int2-hint-3}
# Hint 3 of 4
# Prescribed title: "Predicted Environmental Performance\\nby Trade Openness and Rights Status"
# The \\n is part of the string -- it breaks the title across two lines
# so it is not cut off at the width of the plot.
# Prescribed x-axis label: "Trade Openness (1-10)"
# Prescribed y-axis label: "Environmental Performance Index"
# Prescribed legend title (color =): "Rights Status"
# Prescribed caption: "Predictions are for countries with average economic freedom."
```

```{r exercise-predplot-int2-hint-4}
# Hint 4 of 4
# One plot_model() call with the model, type, terms, title and
# axis.title, then a labs() layer chained with + that sets both the
# legend title (color =) and the caption. Every string comes from the
# prescribed list above and must be typed exactly.
```

```{r exercise-predplot-int2-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use `plot_model()` to produce the prediction plot.")
  if (!grepl("model_env", code_nc))
    fail("Pass **model_env** as the model object.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("ipi_tradeopen", code_nc))
    fail("Include **ipi_tradeopen** in terms (as the first variable --- x-axis).")
  if (!grepl("rights_f", code_nc))
    fail("Include **rights_f** in terms (as the second variable --- separate lines).")
  if (regexpr("ipi_tradeopen", code_nc) > regexpr("rights_f", code_nc))
    fail("**ipi_tradeopen** should come first in terms (x-axis), and **rights_f** second. Check the order inside c(...).")
  if (!grepl("Predicted Environmental Performance\\nby Trade Openness and Rights Status", code_nc, fixed = TRUE)) {
    if (grepl("Predicted Environmental Performance by Trade Openness and Rights Status", code_nc, fixed = TRUE)) {
      fail("The title is right but it needs the line break. Type `\\n` where the prescribed title has it --- without it the title is wider than the plot and gets cut off.")
    }
    fail("Use the prescribed title, including the `\\n` that breaks it across two lines.")
  }
  if (!grepl("axis\\.title\\s*=", code_nc))
    fail("Include the axis.title = argument.")
  if (!grepl("Trade Openness \\(1-10\\)|Trade Openness \\(1\u201310\\)", code_nc))
    fail("Use \"Trade Openness (1-10)\" as the x-axis label.")
  if (!grepl("Environmental Performance Index", code_nc, fixed = TRUE))
    fail("Use \"Environmental Performance Index\" as the y-axis label.")
  if (!grepl("labs\\s*\\(", code_nc))
    fail("Add a `labs()` layer with color = and caption = arguments.")
  if (!grepl("Rights Status", code_nc, fixed = TRUE))
    fail("Use labs(color = \"Rights Status\") to set the legend title.")
  if (!grepl("Predictions are for countries with average economic freedom.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption ending with a period.")

  # The checks above find the required pieces somewhere in the submission. This
  # one establishes that they belong to the plot actually produced: the built
  # plot's own labels and data are read off `.result`.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument is inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one. Check that it is set inside this plot rather than elsewhere in your code."))
  }
  chk("title", "Predicted Environmental Performance\nby Trade Openness and Rights Status", "title")
  chk("x", "Trade Openness (1-10)", "x-axis label")
  chk("y", "Environmental Performance Index", "y-axis label")
  chk("colour", "Rights Status", "legend title")
  chk("caption", "Predictions are for countries with average economic freedom.", "caption")
  if (length(unique(bl$data[[1]]$group)) != nlevels(qog$rights_f))
    fail("The plot should draw one line per rights category. Check that **rights_f** is the second variable in `terms`.")

  pass("Well done. Your plot shows the predicted relationship across trade openness separately for the two rights categories, with economic freedom held at its mean.")
})
```

**Apply 14: Interpret the trade openness interaction plot**

```{r quiz-int2-interpret, echo=FALSE}
question_checkbox(
  "Look at the interaction plot with trade openness on the x-axis. Which statements correctly describe what it shows?",
  answer("The two lines are not parallel --- the visual signature of an interaction",
         correct = TRUE,
         message = "Non-parallel lines are exactly what an interaction looks like in a prediction plot. The interaction coefficient quantifies the difference in the two estimated associations; the plot makes it visible across the full range."),
  answer("The Few Rights line rises steeply while the Many Rights line is nearly flat",
         correct = TRUE,
         message = "The Few Rights estimated association is approximately 4.75 EPI points per unit of trade openness; the Many Rights association is approximately 0.12. This is the interaction finding made visible."),
  answer("The lines cross at a trade openness of about 6.3, and the ranking of the two groups reverses there",
         correct = TRUE,
         message = "The crossover at approximately 6.3 is where the two predicted values are equal based on the point estimates. The exact location is estimated with uncertainty and should not be reported as a precisely determined threshold."),
  answer("Because the adjusted R-squared is below 0.65, the model's predictions are unreliable",
         correct = FALSE,
         message = "Adjusted R-squared summarizes overall explanatory fit. It does not determine whether the interaction coefficient or specific predictions are reliable."),
  correct = "Right. The central finding is the contrast between the steep Few Rights slope and the near-flat Many Rights slope. The fitted lines also cross at approximately 6.3.",
  try_again = "Not quite. Compare the steepness of the two lines and check what the caption says is held constant.",
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

### What We Learned About Political Rights, Trade, and the Environment

The interaction coefficient of −4.63 quantifies the difference between the two estimated associations. The plot makes that difference visible: the Few Rights line sweeps steeply upward while the Many Rights line is nearly flat.

As an additional feature of the fitted point predictions, the lines cross at approximately **ipi_tradeopen** = 6.3: below that value, Many Rights countries have higher predicted environmental performance, while above it, Few Rights countries do. The exact location is estimated with uncertainty and should not be treated as a precise threshold.

Here is what a complete interpretation looks like, working through all six components from the Interpreting Interaction Plots section above:

::: {.research-example}
*As trade openness increases, predicted environmental performance rises steeply in Few Rights countries and barely changes in Many Rights countries. Predictions hold the economic freedom score at its mean throughout. At a trade openness score of about 5, Few Rights countries are predicted to score approximately 46 on the Environmental Performance Index while Many Rights countries are predicted to score approximately 52 --- Many Rights are ahead. At a score of about 9, Few Rights countries are predicted to reach approximately 65 while Many Rights countries remain near 53 --- Few Rights have overtaken them by a wide margin. The prediction lines and their confidence bands overlap near the middle of the trade-openness range, where the fitted values are closest. As an additional feature of the point estimates, the lines cross near 6.3, where the predicted advantage shifts from Many Rights to Few Rights countries; this should not be treated as a precisely determined threshold. This is consistent with the hypothesis that the estimated association between trade openness and environmental performance is stronger in countries with limited political rights.*
:::

## Example 3: Electoral Systems, Parliamentary History, and Women's Representation

Tutorial 18's third example finds that the relationship between how long a parliament has been seating women and its representation today depends on the electoral system. The three conditional slopes were 0.194 under plurality rules, 0.091 under proportional rules, and -0.038 under mixed rules. Those are three numbers in a table. What they mean is easiest to see as three lines.

Example 1 already showed you a crossing: the three wage-growth lines meet at about 5% mortality risk and swap order. But there the conditional slopes all had the same sign, so every line still fell --- what changed was which fell fastest. Here one slope is positive and another negative, so the mixed line runs the opposite way from the other two and is crossed by both. A crossing produced by slopes of opposite sign is a qualitatively different pattern from one produced by slopes of the same sign that happen to meet --- not better supported, but describing something different about the world.

The data frame **ipu** describes 193 national parliaments. **women_percent** is missing for 8 of them and **gender_quota** for one; **electoral_system** leaves 15 countries unclassified into the three families used here, and **years_since_first_woman** --- the source of the centred **years_c** --- is recorded for all 193.

The model is **model_rep_int**, already estimated:

```r
lm(women_percent ~ quota_f + electoral_f * years_c, data = ipu)
```

**years_c** is centered, so zero is a parliament that first seated a woman around 1966. Negative values are parliaments that seated their first woman more recently, positive values are ones that did so earlier. **quota_f** is a control and sits outside the interaction.

### Obtain Specific Predicted Values

Three lines are easier to read once you have a few points on them. Take three positions on the history scale --- 40 years below the average, at the average, and 40 years above it --- and compute the predicted representation for each electoral system at each.

**Practice 15: Compute predicted values across the three electoral systems**

Complete the `newdata` frame so that it covers all three levels of **electoral_f** at each of the three values of **years_c**, holding **quota_f** at "No quota". Two hints are available. Click "Submit Answer" to check your work.

```{r exercise-predict-ipu, exercise=TRUE, exercise.cap="Practice 15"}
nd <- expand.grid(
  quota_f     = factor("___", levels = levels(ipu$quota_f)),
  electoral_f = factor(levels(ipu$electoral_f), levels = levels(ipu$electoral_f)),
  years_c     = c(___, ___, ___))

cbind(nd, fit = round(predict(model_rep_int, newdata = nd), 2))
```

```{r exercise-predict-ipu-hint-1}
# Hint 1 of 2
# You need nine predicted values --- each electoral system at each
# of three points on the history scale --- with quota status held
# constant throughout. Think about how to build that set of
# combinations without writing nine separate calls.
```

```{r exercise-predict-ipu-hint-2}
# Hint 2 of 2
# The three history values are described in the paragraph above:
# forty years below the average, at the average, and forty above.
# Remember that the variable is centered, so the average is zero.
```

```{r exercise-predict-ipu-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("No quota", code_nc, fixed = TRUE))
    fail("Hold **quota_f** at \"No quota\".")
  if (!grepl("-40", code_nc) || !grepl("40", code_nc))
    fail("The three history values are -40, 0 and 40 --- forty years below the average, at it, and forty above.")
  if (is.null(.result) || !is.data.frame(.result))
    fail("Your last line should display the data frame of predictions.")
  if (nrow(.result) != 9)
    fail("The result should have nine rows: three electoral systems at each of three history values.")

  want <- expand.grid(electoral_f = levels(ipu$electoral_f),
                      years_c     = c(-40, 0, 40),
                      quota_f     = factor("No quota", levels = levels(ipu$quota_f)))
  want$predicted <- as.numeric(predict(model_rep_int, newdata = want))
  got <- .result
  key <- function(d) {
    cols <- c("electoral_f", "years_c")
    if (!all(cols %in% names(d))) return(NULL)
    paste(as.character(d$electoral_f), round(as.numeric(d$years_c), 6))
  }
  if (is.null(key(got)))
    fail("The frame needs an **electoral_f** column and a **years_c** column, one row per combination.")
  if (!setequal(key(got), key(want)))
    fail("The nine rows are not the nine combinations asked for: each of the three levels of **electoral_f** at **years_c** of -40, 0 and 40.")
  num <- vapply(got, function(cl) is.numeric(cl) && !all(cl %in% c(-40, 0, 40)), logical(1))
  if (!any(num))
    fail("The frame has no column of predicted values. Add the predictions from **model_rep_int** to the grid.")
  ord <- match(key(want), key(got))
  pr  <- as.numeric(got[[which(num)[1]]])[ord]
  if (!isTRUE(all.equal(pr, want$predicted, tolerance = 1e-4)))
    fail("The predicted values do not match **model_rep_int** at those profiles. Check that **quota_f** is held at \"No quota\" and that the model is **model_rep_int**.")
  pass("Right. The grid holds quota status fixed and varies electoral system and centred history, so every row is a country profile the model can predict for.")
})
```

Read the nine values by electoral system rather than by row. Under plurality rules the predictions rise from 10.83 to 18.59 to 26.34: the earlier a parliament first seated a woman, the more women it is predicted to seat now. Under mixed rules they fall, from 25.86 to 24.35 to 22.84. At 40 years below the average --- a parliament whose first woman arrived around 2006 --- mixed systems lead plurality systems by 15 percentage points. At 40 above, around 1926, plurality systems lead by 3.5.

### Visualize the Three Conditional Slopes

**Practice 16: Produce the three-line interaction plot**

Write the complete `plot_model()` call from scratch, using **model_rep_int**. Put **years_c** on the x-axis and use **electoral_f** to produce the three lines. Use the title \"Predicted Women's Representation by History and Electoral System\", the x-axis label \"Years Since First Woman (centered)\", the y-axis label \"Predicted Percent Women\", the legend title \"Electoral System\", and the caption \"Predictions are for countries with no gender quota.\" Three hints are available. Click "Submit Answer" to check your work.

```{r exercise-predplot-ipu, exercise=TRUE, exercise.cap="Practice 16", fig.alt="Predicted percent women against centered years since the first woman entered parliament, with one line for each electoral system. The horizontal axis runs from about -60 to 60 and the vertical axis from 0 to about 40. Each line has a shaded band, and a legend names the three systems."}

```

```{r exercise-predplot-ipu-hint-1}
# Hint 1 of 3
# This is the same two-variable structure as in Tutorial 17.
# The terms argument takes a character vector of two names: the
# x-axis variable first, then the grouping variable that produces
# separate lines.
```

```{r exercise-predplot-ipu-hint-2}
# Hint 2 of 3
# Set type = "pred", and give terms both variable names in a character
# vector rather than one name.
# axis.title takes a character vector: c("x label", "y label").
# Chain a labs() layer after the call with + to set the legend
# title and the caption.
```

```{r exercise-predplot-ipu-hint-3}
# Hint 3 of 3
# In labs(), the legend title is set with color = because
# plot_model() maps the grouping variable to the color aesthetic.
# The prescribed caption is:
# "Predictions are for countries with no gender quota."
```

```{r exercise-predplot-ipu-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_nc <- gsub("#[^\r\n]*", "", code)
  if (!grepl("plot_model\\s*\\(", code_nc))
    fail("Use `plot_model()` to produce the plot.")
  if (!grepl("model_rep_int", code_nc))
    fail("Pass **model_rep_int** as the model object --- the interaction model, not the additive one.")
  if (!grepl("type\\s*=\\s*['\"]pred['\"]", code_nc))
    fail("Set type = \"pred\".")
  if (!grepl("years_c", code_nc) || !grepl("electoral_f", code_nc))
    fail("Include both **years_c** and **electoral_f** in terms.")
  yc <- regexpr("years_c", code_nc, fixed = TRUE)[1]
  ef <- regexpr("electoral_f", code_nc, fixed = TRUE)[1]
  if (yc > ef)
    fail("**years_c** should come first in terms (x-axis) and **electoral_f** second (separate lines).")
  if (!grepl("Years Since First Woman \\(centered\\)", code_nc))
    fail("Use \"Years Since First Woman (centered)\" as the x-axis label.")
  if (!grepl("Predicted Percent Women", code_nc, fixed = TRUE))
    fail("Use \"Predicted Percent Women\" as the y-axis label.")
  if (!grepl("Electoral System", code_nc, fixed = TRUE))
    fail("Use labs(color = \"Electoral System\") to set the legend title.")
  if (!grepl("Predictions are for countries with no gender quota.", code_nc, fixed = TRUE))
    fail("Use the prescribed caption: \"Predictions are for countries with no gender quota.\"")

  # The checks above find the required pieces somewhere in the submission. This
  # one establishes that they belong to the plot actually produced: the built
  # plot's own labels and data are read off `.result`.
  if (!inherits(.result, "ggplot"))
    fail("Your last line should produce the plot itself, so `plot_model()` is what prints.")
  bl <- tryCatch(ggplot2::ggplot_build(.result), error = function(e) NULL)
  if (is.null(bl))
    fail("The plot could not be built. Check that every argument is inside the `plot_model()` call.")
  lb <- .result$labels
  chk <- function(field, want, what) {
    got <- lb[[field]]
    if (is.null(got) || !identical(gsub("\\s+", " ", as.character(got)),
                                   gsub("\\s+", " ", want)))
      fail(paste0("The plot's ", what, " is not the prescribed one. Check that it is set inside this plot rather than elsewhere in your code."))
  }
  chk("colour", "Electoral System", "legend title")
  chk("caption", "Predictions are for countries with no gender quota.", "caption")
  if (length(unique(bl$data[[1]]$group)) != nlevels(ipu$electoral_f))
    fail("The plot should draw one line per electoral system. Check that **electoral_f** is the second variable in `terms`.")

  pass("Well done. One line per electoral system, each with its own slope, and quota status held at the value the caption names.")
})
```

### Changing a Held Value

**quota_f** is a control sitting outside the interaction, so the plot above holds it at "No quota", its reference category. Conditioning on "Quota" instead answers a different question: what these three trajectories look like among countries that have adopted a quota. The `condition` argument, introduced in Tutorial 17, does this.

**Run and observe 17: The same plot for quota countries**

```{r exercise-predplot-ipu-quota, exercise=TRUE, exercise.cap="Run and observe 17", fig.alt="The same three electoral-system lines for countries with a gender quota. The horizontal axis runs from about -60 to 60 and the vertical axis from about 10 to 45, with a shaded band around each line and a legend naming the three systems."}
plot_model(model_rep_int,
           type = "pred",
           terms = c("years_c", "electoral_f"),
           condition = list(quota_f = "Quota"),
           title = "Predicted Women's Representation by History and Electoral System",
           axis.title = c("Years Since First Woman (centered)",
                          "Predicted Percent Women")) +
  labs(color = "Electoral System",
       caption = "Predictions are for countries with a gender quota.")
```

Compare this to the plot you produced. All three lines sit about 6.9 percentage points higher --- the Quota coefficient. Nothing else moves: the three slopes are unchanged, and the lines still cross at the same place. That is what it means for **quota_f** to be outside the interaction. It shifts the level of every line and changes the relationship of none of them.

### Interpreting the Crossing

**Apply 18: Interpret the three-line plot**

```{r quiz-ipu-crossing, echo=FALSE}
question("The plurality and mixed lines cross inside the plot. Which statement best describes what that crossing means?",
  correct = "Right. The two lines run in opposite directions, so the crossing is a reversal.",
  answer("The two lines run in opposite directions, so which system leads depends on where you look",
         correct = TRUE,
         message = "Under plurality rules the predicted percentage rises the earlier a parliament first seated a woman; under mixed rules it falls. Among parliaments that seated their first woman recently, mixed systems are predicted highest. Among those that did so long ago, plurality systems are. The crossing point sits at a parliament whose first woman arrived around 1941, which is inside the observed data --- the file runs back to 1907."),
  answer("The two groups converge, so electoral system stops mattering once a parliament has been seating women long enough",
         message = "Converging lines approach a common value and stay there. These two pass through each other and keep going in opposite directions, so the gap reopens on the other side with the ranking reversed."),
  answer("The crossing shows that the interaction is statistically significant",
         message = "A crossing is a feature of the point estimates, not a test. Fitted lines with different slopes will always cross somewhere if you extend them far enough --- the question is whether the crossing falls inside the data, and whether the slope difference is distinguishable from zero. Tutorial 18 tested that separately."),
  answer("The mixed line's negative slope means mixed systems are seating fewer women over time",
         message = "The x-axis is not time passing. It is how long ago each parliament first seated a woman, compared across countries. A negative slope means that among mixed-system countries, those that seated their first woman longer ago have fewer women now --- a comparison between countries, not a trend within any one of them."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: look at the direction of each line, not just where they meet."
)
```

### What We Learned About Electoral Systems, History, and Representation

::: {.research-example}
The three-line plot shows what a table of interaction coefficients cannot. Under plurality rules, the earlier a parliament first seated a woman, the more women it is predicted to seat today --- from about 11% for a parliament whose first woman arrived around 2006 to about 26% for one dating back to 1926. Under proportional rules the same relationship holds but is shallower, running from about 22% to about 30%. Under mixed rules it runs the other way: from about 26% down to about 23%.

The consequence is that the mixed line is overtaken twice rather than a gap simply widening. Among parliaments that first seated a woman recently, mixed systems are predicted to have the highest representation and plurality systems the lowest, a spread of about 15 percentage points. Proportional systems pass mixed at a parliament whose first woman arrived around 1978, and plurality systems pass it around 1941. Both crossings fall well inside the data, which reaches back to 1907. Among parliaments with the longest histories the ordering has fully reversed: mixed systems are predicted lowest, about 3.5 points below plurality and 7 below proportional.

What the plot does not do is explain the mixed-system pattern. Tutorial 18 found that this is the one contrast the data distinguish clearly, and also the one the institutional argument says least about. Seeing it as a line makes the finding harder to overlook, which is a reason to plot predictions rather than only report coefficients --- but a picture of an unexplained pattern is still an unexplained pattern.
:::


## Check Your Understanding

A researcher estimates a model of government responsiveness (0--100 scale, higher = more responsive) across countries: responsiveness ~ foreign aid ($ per capita) * GDP per capita (in $10,000s) + institutional quality. Both foreign aid and GDP per capita are interval variables.

**Apply 19: Identify the right prediction strategy**

```{r quiz-cyu-strategy, echo=FALSE}
question("Given that both interacting predictors are interval variables, which prediction strategy is appropriate?",
  correct = "Right. Choose specific values of one variable and compute predicted differences across the range of the other at each.",
  answer("Choose two or three meaningful values of one variable and compute predicted outcomes at low and high values of the other at each level",
         correct = TRUE,
         message = "This is the interval x interval strategy: with no discrete groups, you choose specific values of one variable to show how the predicted difference changes. The plot uses bracket notation to show all three lines simultaneously."),
  answer("Use the group-difference strategy: compute one predicted value per category at the mean of the other variable, then vary the interval predictor within the key category",
         message = "The group-difference strategy applies when one variable is categorical. With two interval variables there are no discrete groups --- you choose specific values of one to show how the predicted difference varies."),
  answer("Produce a single prediction plot with foreign aid on the x-axis --- an interval x interval interaction is interpreted the same way as an additive model plot",
         message = "An interaction changes the interpretation: the predicted difference produced by changing foreign aid varies with GDP per capita. A single plot shows only one slice. Bracket notation is needed to show how the relationship changes across GDP levels."),
  answer("Use a joint F-test first; if non-significant, revert to the additive model prediction strategy",
         message = "Prediction plots are informative regardless of significance --- they show the estimated pattern from the fitted model. The joint F-test informs how strongly you can interpret the pattern, not whether to plot it."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: with two interval variables, what replaces the 'group' concept from the categorical x interval strategy?"
)
```

**Apply 20: Read a bracket-notation plot**

The researcher produces: `plot_model(model, type = "pred", terms = c("foreign_aid", "gdp_pc [meansd]"))`. The plot shows three lines labeled "−8.2", "0.0", and "8.2" in the legend (mean − 1 SD, mean, mean + 1 SD of GDP per capita in $10,000s).

```{r quiz-cyu-bracket, echo=FALSE}
question("What does the `[meansd]` notation produce, and how should the three lines be interpreted?",
  correct = "Right. Three lines at mean ± 1 SD of GDP per capita, one per chosen GDP level.",
  answer("`[meansd]` draws three lines at mean minus 1 SD, mean, and mean plus 1 SD of GDP per capita",
         correct = TRUE,
         message = "Each line shows the predicted foreign aid association at that GDP level, and non-parallel lines reveal the interaction. The bracket notation is the interval x interval analogue of the factor grouping: instead of one line per category, you get one line per chosen value of the continuous grouping variable. Mean ± 1 SD covers a realistic range without requiring arbitrary value choices."),
  answer("`[meansd]` plots the mean and standard deviation of the outcome variable as separate uncertainty bounds around the prediction",
         message = "`[meansd]` applies to the grouping variable in terms =, not the outcome. It specifies which values of GDP per capita to use for the three separate lines."),
  answer("The three lines show the same relationship at three different time points, since meansd is a time series decomposition shorthand",
         message = "`[meansd]` is sjPlot bracket notation for a continuous grouping variable. It selects mean minus 1 SD, mean, and mean plus 1 SD of the named variable --- nothing to do with time."),
  answer("The legend labels -8.2, 0.0, and 8.2 are the predicted responsiveness values, not GDP per capita levels",
         message = "The legend labels are the GDP per capita values at which the three lines are drawn. Each line then shows the full foreign aid relationship at that GDP level."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: what does the bracket notation tell `plot_model()` to do with the second variable in terms = ?"
)
```

**Apply 21: Connect the plot to the coefficient test**

The three lines in the responsiveness plot are not perfectly parallel --- the line for high-GDP countries is somewhat steeper than the line for low-GDP countries. A colleague says the non-parallel lines confirm the interaction is statistically significant.

```{r quiz-cyu-significance, echo=FALSE}
question("How would you correct this claim?",
  correct = "Right. Non-parallel fitted lines appear with any nonzero estimated interaction --- significance requires the coefficient test.",
  answer("Non-parallel fitted lines appear whenever an interaction term is in the model, significant or not",
         correct = TRUE,
         message = "A nonzero estimated interaction always produces non-parallel point-estimate lines. Statistical significance is determined by the t-statistic and p-value on the interaction term, not by the visual appearance of the lines."),
  answer("The colleague is correct --- non-parallel lines can only arise when the interaction coefficient is significantly different from zero",
         message = "Non-parallel lines result from any nonzero estimated interaction, regardless of significance. A non-significant interaction still produces non-parallel lines; they just reflect sampling variation that could plausibly be zero in the population."),
  answer("The colleague is correct if the confidence bands for the three lines do not overlap, since that confirms a statistically significant separation",
         message = "These are confidence bands for separate predicted outcomes, not a confidence interval for their difference. Whether they overlap does not establish whether the interaction coefficient is statistically significant."),
  answer("The plot is uninformative about the interaction --- only the coefficient table can say whether the interaction is clearly supported by the data",
         message = "The plot is informative about the estimated pattern --- it just cannot establish statistical significance. The right correction is to clarify what the plot shows and what the coefficient test adds."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: what produces non-parallel lines in a prediction plot? Does the interaction need to be significant for that to happen?"
)
```

## The Takeaways

The same two tools, `predict()` and `plot_model()`, work for additive and interaction models alike. The tools are identical; what changes is what they reveal.

1. **The workflow is: table → specific predictions → plot.** The `predict()` calls give you concrete numbers before you see the full picture. The plot shows how predicted outcomes change across the full range of the predictor, with uncertainty visible as a shaded band.

2. **`predict()` requires a `newdata` data frame with every predictor named and typed exactly as in the model.** For categorical variables coded as factors, use `factor("level", levels = levels(data$variable))`. The output is in the units of the outcome variable.

3. **`plot_model(type = "pred")` sweeps across the full range of the x-axis predictor.** Variable names in `terms` are always quoted strings. Always add `labs(caption = "...")` to state the held values. Use `condition = list(variable_name = value)` to override defaults --- note that variable names in `condition` are unquoted, and values are numbers or quoted strings depending on the predictor type.

4. **For interval × interval interactions:** choose two or three meaningful values of one variable and for each, compute the [predicted difference]{.important-text} between a low and a high value of the other variable. The within-pair difference is the model-implied predicted difference at that level of the other variable. Comparing those differences across chosen values shows how the interaction changes the relationship. In the plot, use `terms = c("x_axis_var", "other_var [meansd]")` to show three lines automatically at mean ± 1 SD. Non-parallel lines make the changing predicted difference directly visible.

5. **For categorical × interval interactions:** the [predicted comparison]{.important-text} runs two ways --- across categories at a representative value of the interval predictor, and across the range of the interval predictor within each category. In the plot, put the interval variable on the x-axis and use the categorical variable to produce colored lines --- one per category, however many there are. Non-parallel lines make the changing predicted differences directly visible. Use `labs(color = "Label")` to rename the color legend.

6. **Non-parallel fitted lines appear whenever an interaction term is in the model**, whether or not the interaction is statistically significant. The plot shows the estimated pattern; the coefficient test indicates how strongly the data support a nonzero interaction. A predictor that is in the model but **not** part of the interaction behaves differently: changing its held value with `condition` shifts every line by the same amount and changes none of their slopes.

7. **A complete interpretation of an interaction plot covers:** the pattern for each group or line (direction and steepness), the held values, specific predicted values and their differences, the confidence bands, and the connection to the hypothesis. When the fitted lines cross, check where. A crossing inside the observed range of the x-axis variable means the ordering of the groups actually reverses within the data --- a substantive finding. A crossing that would occur only beyond the observed range is an extrapolation rather than a finding. Either way it is a feature of the point estimates, and the confidence bands show how uncertain the predicted outcomes are near the crossing --- they are not a test of it. Whether the interaction itself is supported comes from the coefficient test, or the joint test when a factor contributes several interaction terms.

Tutorial 20 changes the outcome. Everything so far has predicted a quantity --- a vote share, a percentage, an index score --- with a straight line. When the outcome is binary, whether a country ratified a treaty or a respondent turned out to vote, that line no longer works, and logistic regression takes its place.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below.

```{r encode-logic-19, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-19, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-19, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now translate interaction model coefficients into concrete predicted comparisons — and make the conditional relationship visible across the full observed range.">
  <span aria-hidden="true">📊</span>
  Excellent work! You can now translate interaction model coefficients into concrete predicted comparisons &mdash; and make the conditional relationship visible across the full observed range.
</p>
')
```
