---
title: "Hypothesis Tests: Categorical and Interval Variables"
tutorial:
  id: "13-Hypotheses-Categorical-Interval"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  One-way ANOVA and the two-sample, one-sample, and paired t-tests.
---

## Overview {#main-content}

Two groups almost never have exactly the same mean in a sample, even when they have the same mean in the population. Draw a different sample and the gap between them changes. So a difference in group means, however suggestive it looks, leaves open the question that matters: whether a gap that size is larger than chance would ordinarily produce.

This tutorial covers the tests that settle it when the outcome is interval and the groups are categorical: a one-way ANOVA, which asks whether any of the group means differ; a two-sample t-test, which sharpens that into a directional claim about two specific groups; a one-sample t-test against a fixed value; and a paired t-test for two measurements taken on the same cases. All four are worked through the same five explicit steps. If you need to review that procedure or the decision rule, go back to Tutorial 11.

In this tutorial, you will learn how to test whether a categorical and an interval variable are related in the population. Specifically, we will cover:

* How to test whether the means of an interval variable differ across categories of a categorical variable using a one-way ANOVA, with the `oneway.test()` function
* How to sharpen a general ANOVA result into a directional two-sample t-test using `t.test()`, when a hypothesis predicts which of two specific groups will have the higher mean
* How to use `filter()` with the `%in%` operator to narrow a multi-category variable to the two groups a hypothesis names before running a t-test
* How to test whether a single sample mean differs from a fixed hypothesized value using a one-sample t-test
* How to test whether the average difference between two variables measured on the same cases differs from zero using a paired t-test
* How to organize any of these tests into the same five explicit steps, from stating the hypotheses to reporting what the result means


```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(lifecycle_verbosity = "quiet")
options(cli.num_colors = 1)
data("counties", package = "QPATutorialsCourse")
data("world", package = "QPATutorialsCourse")

# A simple random sample of counties, used by the one-sample t-test example.
# The reported figures depend on this seed and this n: change either and
# the mean, the t-statistic and the interval in the prose all change.
set.seed(2026)
county_sample_t <- counties[sample(nrow(counties), 150), ]
```

## Testing the Relationship Between a Categorical Variable and an Interval Variable

Tutorial 8 compares means of an interval variable across categories of a categorical variable. But observed group means can differ even when the underlying process contains no systematic group difference --- a different realization could have produced somewhat different means. We test this kind of relationship using two complementary tools. A **one-way ANOVA** considers every category at once and asks whether the population means could plausibly all be equal. A **two-sample t-test** asks a narrower but sharper question: comparing exactly two specific groups, it can test a specific directional prediction in a way the ANOVA cannot. Since computing an ANOVA requires nothing beyond what we already have, our usual approach will be to test the general claim first, then sharpen to a directional two-group t-test when the hypothesis supports one. We also look at two related but different questions: whether a single sample mean differs from a fixed hypothesized value, and whether the average difference between two variables measured on the same cases differs from zero.

Tutorial 12 organizes the building blocks of a hypothesis test --- introduced conceptually in Tutorials 10 and 11 --- into five explicit steps:

1. State the null hypothesis, $H_0$ --- a claim about a population parameter, not a sample statistic.

2. State the alternative hypothesis, $H_A$. Like the null, it is a claim about the population.

3. Run the test and obtain a test statistic.

4. Find the p-value associated with that test statistic.

5. Decide whether to reject or fail to reject $H_0$, and report what that means substantively.

We will follow these same five steps to test relationships between a categorical and an interval variable, and for the one-sample and paired t-tests that follow.

A word about the data frames used below. Neither **counties** nor **world** is a random sample from a larger population the way a survey sample is. For the ANOVA and the two-sample t-tests, inference rests instead on treating the observed cases as realizations of a broader process that could have produced different outcomes. Under that reading the p-value asks how surprising group differences this large would be if that broader process contained no systematic difference between the groups. The one-sample t-test later in the tutorial is built differently, on a sample drawn at random from the county data --- the contrast is worth watching for.

## The One-Way ANOVA

The [**one-way ANOVA**]{.important-text} compares two kinds of variability: [**between-group variability**]{.important-text}, how far the group means are from each other, against [**within-group variability**]{.important-text}, how spread out individual cases are around their own group's mean --- variability that has nothing to do with group membership at all. ANOVA summarizes the relationship between these two quantities in a single test statistic, the [**F-statistic**]{.important-text}:

$$F=\frac{\text{between-group variability}}{\text{within-group variability}}$$

If the population means really are all equal, the differences among the observed group means reflect nothing but chance, so $F$ should be close to 1. If the population means differ, the group means will tend to be far apart relative to the ordinary spread within each group, and $F$ will be large. As with the test statistics in earlier tutorials, you do not need to remember the formula behind $F$ --- understand the logic, and let R compute it for you.

The formula above captures the intuition. In practice, R's `oneway.test()` uses a Welch adjustment that handles unequal variances across groups --- the same default it applies to the two-sample t-test. This modifies how the denominator is calculated but preserves the same broad logic: large between-group spread relative to within-group spread produces a large $F$ and a small p-value.

The figure below makes this concrete. Both panels show three groups with exactly the same three means. The only thing that differs is how spread out the individual values are within each group.

```{r concept-f-statistic-illustration, echo=FALSE, fig.alt="Two side-by-side panels of dot plots with box overlays, each showing three groups --- Group A, Group B, and Group C --- with identical group means of 10, 15, and 20, marked by black diamonds. In the left panel, labeled 'Low within-group spread (F about 718),' each group's points are tightly clustered around its mean with little overlap between groups. In the right panel, labeled 'High within-group spread (F about 29),' each group's points are widely spread with substantial overlap between adjacent groups, even though the group means remain at the same three positions."}
set.seed(13)
groups <- c("Group A", "Group B", "Group C")
means  <- c(10, 15, 20)
n <- 30

# Generate a single set of mean-zero deviations, then scale for each panel
# This guarantees the sample means are exactly 10, 15, 20 in both panels
raw_devs <- lapply(seq_along(means), function(i) {
  d <- rnorm(n)
  d - mean(d)   # center so sample mean is exactly 0
})

low_spread <- data.frame(
  group = rep(groups, each = n),
  value = c(means[1] + raw_devs[[1]] * 1,
            means[2] + raw_devs[[2]] * 1,
            means[3] + raw_devs[[3]] * 1),
  panel = "Low within-group spread\n(F ≈ 718)"
)

high_spread <- data.frame(
  group = rep(groups, each = n),
  value = c(means[1] + raw_devs[[1]] * 5,
            means[2] + raw_devs[[2]] * 5,
            means[3] + raw_devs[[3]] * 5),
  panel = "High within-group spread\n(F ≈ 29)"
)

f_illustration <- rbind(low_spread, high_spread)
# F values are the Welch statistics oneway.test() returns for these two panels,
# rounded. They depend on set.seed(13), n = 30 and the 1x / 5x scaling below --
# recompute them if any of those change.
f_illustration$panel <- factor(f_illustration$panel,
                                levels = c("Low within-group spread\n(F ≈ 718)",
                                           "High within-group spread\n(F ≈ 29)"))

ggplot(f_illustration, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 4, color = "black") +
  facet_wrap(~panel) +
  labs(title = "Same Group Means, Different Within-Group Spread",
       subtitle = "Black diamonds mark the group means",
       x = NULL,
       y = "Value",
       caption = "Hypothetical data") +
  guides(fill = "none")
```

Both panels have exactly the same group means --- 10, 15, and 20, marked by the black diamonds --- but the left panel's tightly clustered values produce an $F$ of about 718, while the right panel's widely spread, overlapping values produce an $F$ of about 29. The means are identical; only the within-group variability differs. That is precisely what $F$ is measuring.

The two numbers are worth comparing. The values in the right panel are spread five times as widely as those on the left, and 718 divided by 29 is very close to 25 --- five squared. That is the formula doing exactly what it says: within-group variability sits in the denominator, so multiplying the spread by five divides $F$ by twenty-five, while the between-group distances in the numerator never move.

In Tutorials 10 and 11, the sampling distribution developed was the normal curve --- symmetric and bell-shaped. In Tutorial 12, $\chi^2$ introduced a different shape: bounded at 0 and stretched into a long right tail, because it is built entirely from squared terms. $F$ follows yet another distribution, for a closely related reason --- it is a ratio of two variability measures, each built from squared terms, so $F$ itself can never be negative either. Like $\chi^2$, its distribution starts at 0 and skews to the right rather than spreading symmetrically. The exact shape depends on two degrees-of-freedom values --- the numerator and denominator df R reports with the test --- rather than just one, but as before, the precise calculation is not something we will dwell on.

What does not change is the underlying logic. Just as with the normal and chi-square distributions, a hypothesis test locates your observed $F$ on this distribution and asks how far into the tail it falls. An $F$ value far out in that long right tail is one that would be rare if the population means were really all equal --- exactly the kind of evidence a hypothesis test is built to evaluate. R converts $F$ into a p-value this way automatically, and we apply the same decision rule as always: a small p-value (conventionally below .05) leads us to reject the null hypothesis that the population means are equal, while a large p-value means we fail to reject it.

### Running the Test with `oneway.test()`

To run a one-way ANOVA in R, we use the `oneway.test()` function: a formula, with the interval (outcome) variable on the left of a tilde (`~`) and the categorical (explanatory) variable on the right, plus a `data` argument: `oneway.test(outcome ~ group, data = df)`.

Notice that the formula uses bare variable names --- `outcome` and `group` --- not `df$outcome` and `df$group`. That is what the `data` argument is for, exactly as in `ggplot()`: it tells R where to look up the names. You will see the same pattern in `t.test()` later in this tutorial.

By default, `oneway.test()` does not assume the population variances are equal across groups --- the same default stance R takes for the two-sample t-test you will meet later in this tutorial. The output reports the F-statistic, two degrees-of-freedom values, and a p-value. Because of how the unequal-variance correction is calculated, the denominator degrees of freedom is often not a whole number --- that is expected, not an error, and it is not something we will focus on in this course. The test statistic and the p-value are what we use to reach a conclusion.

Putting those together, every one-way ANOVA in this tutorial has the same shape: the interval outcome, a tilde, the categorical variable that defines the groups, and the data frame both live in.

**Example code:**

```{r onewaytest-template, exercise=FALSE, eval=FALSE}
oneway.test(outcome_variable ~ grouping_variable, data = yourdata)
```

One thing about this test is worth noting before we formalize the procedure. Its null states that all population group means are equal --- $H_0: \mu_1 = \mu_2 = \ldots = \mu_k$ --- and its alternative that at least one differs from the others. Notice what the alternative does not say: not which mean, and not in which direction. That vagueness is not sloppiness; it is what an omnibus test claims, and it is the reason a directional hypothesis needs a second test.

## Is College Attainment Lower in Rural Counties?

College attainment was one of the cleaner patterns in Tutorial 8, where we calculated the group means: urban counties averaged well above the other two categories, which sat close to each other. The argument is intuitive --- rural areas have fewer local jobs requiring a degree and less access to higher education, so college graduates tend to concentrate in cities. Now we ask whether gaps as consistent as those across the three rurality categories would be surprising if the null hypothesis were true, then sharpen to the specific urban-versus-rural contrast the hypothesis predicts.

The hypothesis, unchanged from Tutorial 8:

::: {.hypothesis}
In a comparison of [counties], those that are [more rural] will have a [lower share of college graduates] than will those that are [more urban].
:::

The data frame **counties** contains the ordinal variable **rural**, which takes the value 0 for urban counties, 1 for counties that are neither distinctly urban nor rural, and 2 for rural counties. It also contains the interval variable **prop_college_grad**, the proportion of a county's residents age 25 or older who have completed a bachelor's degree or higher. Every county has a recorded value of **rural**, while **prop_college_grad** is missing for 1 county, so any summary you compute from it describes the 3,111 counties with a recorded value rather than all 3,112 in the data. **rural** is our explanatory variable and **prop_college_grad** is our outcome.

### Testing the General Claim

The hypotheses below use $\mu$ (mu), the population mean --- the Greek-letter counterpart to the sample mean $\bar{x}$, following the convention from Tutorial 10.

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{urban}=\mu_{neither}=\mu_{rural}$$
where $\mu$ is the mean proportion of college graduates in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

**Run and observe 1: The one-way ANOVA**

```{r anova-collegegrad-by-rural, exercise=TRUE, exercise.cap="Run and observe 1"}
oneway.test(prop_college_grad ~ rural, data = counties)
```

The output reports, in order: the F-statistic (labeled `F`), the numerator and denominator degrees of freedom (`num df` and `denom df`), and the p-value. As noted earlier, the denominator degrees of freedom will likely not be a whole number --- that is expected.

#### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 2: Interpret the ANOVA result**

Find the p-value in your output and compare it to .05 before answering.

```{r quiz-collegegrad-general-result, echo=FALSE}
question("Find the p-value in your output. Which statement best summarizes what this test tells us?",
  correct = "Right. An ANOVA p-value carries no direction.",
  answer("We reject the null --- evidence that the three means are not all equal --- but that evidence carries no direction",
         correct = TRUE,
         message = "At F = 239.35 the p-value is far below .05, so group means this far apart would be very unlikely if the three population means were equal. Notice what F was built from, though: a ratio of between-group spread to within-group spread, both of them sums of squared distances. Squaring discards sign, so a table where college attainment falls with rurality and one where it rises produce the same F. The test tells you the means differ somewhere; it has no way to say which group is highest, or even which two differ. That is why the hypothesis needs the t-test that follows."),
  answer("We reject the null hypothesis, which provides strong evidence that urban counties have a higher mean proportion of college graduates than rural counties",
         message = "Rejecting this null is evidence that the three population means are not all equal, and no more. It names no group and no direction, so it cannot support a claim about urban counties in particular --- that is what the t-test is for."),
  answer("We fail to reject the null hypothesis, so there is no relationship between rural and prop_college_grad",
         message = "A p-value this small leads us to reject the null, not fail to reject it. Failing to reject would require a p-value above .05 --- and even then it would not show that no relationship exists, only that these data gave no strong evidence against that possibility."),
  answer("We reject the null hypothesis, and since the relationship runs across all three rurality categories, the directional claim is also established",
         message = "Running across all three categories is not the same as running in a particular direction. F is built from squared distances, so it would be just as large if the pattern ran the other way. The direction is visible in the group means --- but it is the t-test, not this one, that weighs evidence for it."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Find the p-value and compare it to .05. Then ask what F was built from, and whether that arithmetic could tell a rising pattern from a falling one."
)
```

The general claim is settled. Testing the directional claim requires a second tool, which the next section introduces.

## The Two-Sample t-Test

The ANOVA has answered the general question. The directional claim needs a different tool --- one that compares exactly two groups and can test which way the difference runs. That is a [**two-sample t-test**]{.important-text}, run with the `t.test()` function you will also use for the one-sample and paired tests later in this tutorial.

### The Logic of the Test

A two-sample t-test states a null that two specific population means are equal, and a directional alternative that one is greater or less than the other --- a sharper pair of claims than the ANOVA's, and about two groups rather than all of them.

Unlike chi-square and F, which are right-skewed and bounded at zero, t is symmetric and bell-shaped --- the same general shape as the normal curve. Tutorials 10 and 11 named it in passing as the small-sample alternative to that curve; here is what that means. The difference shows up in the tails: with smaller samples, individual estimates of variability are themselves less reliable, so extreme values of the test statistic are a bit more likely than the normal curve alone would suggest, and the t-distribution's wider tails account for that. The exact shape depends on degrees of freedom --- few degrees of freedom means noticeably heavier tails, and as degrees of freedom grow, the t-distribution converges to the normal curve. As with chi-square and F, the logic does not change: locate your observed statistic on this distribution, see how far into the tail it falls, and let R compute the resulting p-value --- whichever version of `t.test()` you are running: two-sample, one-sample, or paired.

### Running the Test with `t.test()`

`t.test()` compares the means of exactly two groups. When a categorical variable has exactly two categories, we can use formula notation directly: `t.test(outcome ~ group, data = df)`. Several of the categorical variables in this tutorial have three or four categories, so before running a directional t-test we will first use `filter()` with the `%in%` operator to narrow the data down to the two specific categories the hypothesis compares, then use formula notation on that filtered data. You will see this pattern in every directional t-test in this tutorial.

As with `prop.test()` in Tutorial 12, the `alternative` argument controls whether the test is one-tailed or two-tailed. The default, `alternative = "two.sided"`, tests whether the two means differ in either direction. When a hypothesis predicts a specific direction, we use `alternative = "less"` or `alternative = "greater"`. The key rule: `alternative = "less"` tests whether the *first* group's mean is lower than the *second's*; `alternative = "greater"` tests whether the first group's mean is higher. For numeric grouping variables, R orders groups by their numeric value. For character variables, R orders groups alphabetically. Always check which group will come first before choosing `alternative` --- the labels attached to the group estimates in the output are an easy way to confirm R's ordering.

Alongside the p-value, `t.test()` reports a confidence interval, as `prop.test()` did in Tutorial 12. What it is an interval *for* depends on which version you are running --- the difference in means here, and something different for the one-sample and paired tests later.

Putting those together, a directional two-sample t-test takes two steps: narrow the data to the two groups the hypothesis names, then run the test on what is left.

**Example code:**

```{r ttest-twosample-template, exercise=FALSE, eval=FALSE}
two_groups <- yourdata %>%
  filter(grouping_variable %in% c("first group", "second group"))

t.test(outcome_variable ~ grouping_variable,
       data = two_groups,
       alternative = "greater")
```

## College Attainment: Testing the Directional Claim

Our hypothesis does not just predict that the three group means differ --- it predicts a specific direction: the mean proportion of college graduates should be higher among urban counties than among rural counties specifically. The group means from Tutorial 8 run in that direction. But the ANOVA p-value does not directly test that specific one-tailed contrast. To do that, we compare the two endpoint categories --- urban and rural --- using `filter()` to narrow the data to just those two groups before running a two-sample t-test.

### Step 1: State the Null Hypothesis

$$H_0: \mu_{urban}=\mu_{rural}$$
where $\mu$ is the mean proportion of college graduates in the population.

### Step 2: State the Alternative Hypothesis

$$H_A: \mu_{urban} > \mu_{rural}$$

### Step 3: Run the Test

We set `alternative = "greater"` because the hypothesis predicts that the urban mean is *higher* than the rural mean --- not just different from it. The `filter()` call keeps only counties where **rural** equals 0 (urban) or 2 (rural), dropping those equal to 1 (neither), which leaves exactly the two groups `t.test()`'s formula notation requires --- 1,162 urban counties and 627 rural ones. **prop_college_grad** is recorded for every one of them, so all 1,789 enter the test.

This is a new use of `filter()`. Rather than piping a filtered result straight into a plot or a summary function, here we save the filtered data frame as a new object, `counties_urban_rural`, and then pass it to `t.test()` separately via the `data =` argument --- because `t.test()` does not accept piped input the way `ggplot()` does. The two steps are: filter down to the right cases, then run the test on that filtered data.

Load [dplyr]{.package-name} with `library(dplyr)` before using `filter()` and the pipe. Once loaded, it is available for all subsequent exercises in this tutorial.

**Run and observe 3: A one-tailed two-sample t-test**

```{r ttest-collegegrad-urban-rural, exercise=TRUE, exercise.cap="Run and observe 3"}
library(dplyr)
counties_urban_rural <- counties %>%
  filter(rural %in% c(0, 2))

t.test(prop_college_grad ~ rural, data = counties_urban_rural, alternative = "greater")
```

**Practice 4: Match the output's group labels to the categories**

Look at how **rural** is coded and match the numeric labels in the output to the categories they represent.

```{r quiz-collegegrad-group-labels, echo=FALSE}
question("In the `t.test()` output above, the alternative hypothesis line refers to 'group 0' and 'group 2', and the sample estimates report 'mean in group 0' and 'mean in group 2'. Which counties do these labels refer to?",
  correct = "Right. R prints the stored codes, not the category names.",
  answer("Group 0 refers to urban counties, and group 2 refers to rural counties",
         correct = TRUE,
         message = "**rural** is stored as a number --- 0 for urban, 1 for neither, 2 for rural --- and R labels the groups with whatever values it finds, so the output says 0 and 2 rather than \"urban\" and \"rural.\" Nothing in the printout tells you which is which; you have to carry the coding with you. This is the same hazard `CrossTable()` presents in Tutorial 12, where the columns printed as 0, 1 and 2 for the same reason. Translate the codes back into categories before you interpret any result, or it is easy to report the comparison backwards."),
  answer("Group 0 refers to rural counties, and group 2 refers to urban counties",
         message = "This reverses the coding. rural is coded 0 for urban counties and 2 for rural counties --- not the other way around."),
  answer("Group 0 and group 2 refer to two different values of prop_college_grad, not to categories of rural",
         message = "prop_college_grad is the outcome variable being averaged within each group --- it is not itself split into 'group 0' and 'group 2'. Those labels come from the values of rural, the explanatory variable."),
  answer("It is not possible to tell which counties group 0 and group 2 refer to from this output alone",
         message = "It is possible --- R reports the numeric codes from the explanatory variable directly. You need to recall, from how rural was defined, which code corresponds to which category."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Look back at how rural was defined when we introduced this example --- which numeric code was assigned to urban counties, and which to rural counties?"
)
```

### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 5: Evaluate the rural/college-attainment hypothesis**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-collegegrad-directional-result, echo=FALSE}
question("Find the p-value in your output. Does this result provide statistically significant evidence in support of the hypothesis?",
  correct = "Right. Significant, and in the direction the hypothesis predicted.",
  answer("Yes --- the p-value is far below .05 and urban counties have the higher mean, as predicted",
         correct = TRUE,
         message = "Urban counties average 0.255 against 0.177 for rural ones, and the p-value is below .001, so a gap that size would be very unlikely if the two population means were equal. The confidence interval agrees, excluding 0. This is where the ANOVA's limitation gets repaired: that test gave evidence that the three means are not all equal without saying which or in which direction, and this one weighs evidence for the specific contrast the hypothesis names. Neither result proves the hypothesis; together they are what support for it looks like."),
  answer("Yes, but only weakly --- the p-value is small because the sample has over 3,000 counties, not because the effect is meaningful",
         message = "Sample size does affect p-values, but the group means themselves tell the story too: urban counties average 0.255 --- 25.5% of residents age 25 or older --- compared with 0.177 for rural counties, a gap of nearly 8 percentage points. Both the statistical significance and the magnitude of the difference support the hypothesis."),
  answer("No --- the p-value is below .05, so we fail to reject the null hypothesis",
         message = "This reverses the decision rule. A p-value *below* .05 leads us to *reject* the null hypothesis, not fail to reject it. We fail to reject the null only when the p-value is *above* .05."),
  answer("We cannot conclude anything about direction from this test because the ANOVA already established the relationship",
         message = "The ANOVA provides evidence that the means are not all equal across the three groups, but its p-value does not directly test whether urban counties specifically have a higher mean than rural counties. This t-test is what directly evaluates that contrast. The two results answer different questions."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value and compare it to .05. Then check whether the sample estimates match the direction the hypothesis predicted: which group (0 or 2) has the higher mean?"
)
```

The output also reports a 95% confidence interval --- here (0.072, Inf), for the difference in means, group 0 minus group 2. One bound is the theoretical extreme rather than a value calculated from the data, because this is a one-tailed test; for a difference in means that extreme is Inf, since a mean difference has no fixed upper limit the way a proportion does. The bound that matters is the calculated one, 0.072, and what to ask of it is whether the interval excludes 0. When a confidence interval and a hypothesis test are matched --- same significance level, same one- or two-tailed specification --- they agree, and here both point the same way.

## Is the Democratic Vote Percentage Related to a County's Racial Majority?

The racial majority example from Tutorial 8 produced one of the starkest contrasts in the tutorial: Black Majority counties averaged a Democratic vote percentage more than 30 percentage points above White Majority counties. The argument connects to decades of party coalition-building --- the Democratic Party's positions on civil rights and voting rights have made it the party more closely associated with addressing Black voters' political concerns, while the Republican Party has drawn disproportionate support from White voters over the same period. Formally testing this hypothesis introduces a new challenge: the variable has four categories, but the hypothesis names only two of them. We need to filter before we can test.

The hypothesis, unchanged from Tutorial 8:

::: {.hypothesis}
In a comparison of [counties], those with a [Black racial majority] will have a [higher mean Democratic vote percentage] than will those with a [White racial majority].
:::

The data frame **counties** contains the nominal variable **racial_majority**, stored as a character variable, which takes the values "White Majority," "Black Majority," "Hispanic Majority," and "Other/No Majority." It also contains the interval variable **dem2p_percent**, the Democratic candidate's share of the two-party presidential vote in the county in 2016 (as a percent), stored as a numeric variable. Neither variable has any missing values, so summaries computed from them describe all 3,112 counties.

As in the previous example, we start with the general claim that the population means differ across all four categories, then sharpen to the specific comparison the hypothesis actually makes: Black Majority counties against White Majority counties.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{White}=\mu_{Black}=\mu_{Hispanic}=\mu_{Other}$$
where $\mu$ is the mean Democratic vote percentage in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

The comparison of group means in Tutorial 8 showed that Black Majority counties had a noticeably higher mean Democratic vote percentage than White Majority counties. Now we test whether a difference that large across all four categories would be surprising if the population means were actually equal.

**Practice 6: Add the ANOVA for racial majority**

Fill in the explanatory variable on the right side of the tilde. Two hints are available. Click "Submit Answer" to check your work.

```{r anova-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice 6"}
oneway.test(dem2p_percent ~ ___, data = counties)
```

```{r anova-dem2p-by-racialmajority-hint-1}
# Hint 1 of 2
# The test compares mean Democratic vote percentage across groups of
# counties. Only the grouping variable is blank. Which variable in this
# hypothesis defines the groups being compared?
```

```{r anova-dem2p-by-racialmajority-hint-2}
# Hint 2 of 2
# The groups are defined by racial_majority, so that is what goes on the
# right side of the tilde.
```

```{r anova-dem2p-by-racialmajority-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("oneway\\.test\\s*\\(", code_no_comments)) {
    fail("Use `oneway.test()` to run the one-way ANOVA.")
  }
  if (!grepl("dem2p_percent\\s*~\\s*racial_majority", code_no_comments)) {
    fail("The outcome goes on the left of the tilde and the explanatory variable on the right. Which variable defines the groups this test compares?")
  }
  if (!grepl("data\\s*=\\s*counties", code_no_comments)) {
    fail("Set `data = counties`.")
  }

  expected <- oneway.test(dem2p_percent ~ racial_majority, data = counties)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the `oneway.test()` output. If you're seeing an error, check for a missing tilde (`~`) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check the variable you put on the right of the tilde --- it should be the one whose categories define the groups.")
  }

  pass("Exactly! **dem2p_percent** is the outcome and **racial_majority** is the explanatory variable, so it tests whether the mean Democratic vote percentage differs across all four racial-majority categories at once.")
})
```

#### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 7: Diagnose what a four-group ANOVA leaves open**

```{r quiz-dem2p-general-result, echo=FALSE}
question("Find the p-value in your output. What does this result tell us?",
  correct = "Right. A significant F says the means differ somewhere, and no more.",
  answer("We reject the null --- evidence that the four means are not all equal --- but not which groups differ, or how",
         correct = TRUE,
         message = "This ANOVA pooled four racial-majority categories, and F is a single ratio built from all of them, so a large value can arise many ways. It could come from Black Majority counties standing apart, as the hypothesis predicts --- or from Hispanic Majority and Other/No Majority counties differing from each other, or from any mixture. The one p-value cannot separate these, because every group fed into it. Reading the group means will show you where the action is; the t-test is what weighs evidence for a claim about one specific pair."),
  answer("We fail to reject the null hypothesis --- the p-value is too small to provide reliable evidence of a relationship",
         message = "A very small p-value is stronger evidence against the null, not weaker. A p-value this small leads us to reject the null, not fail to reject it."),
  answer("We reject the null hypothesis, and since the result is highly significant, we can conclude that every racial-majority group mean differs significantly from every other",
         message = "Rejecting this null says only that the four means are not all alike. It does not say each differs from every other. A significant F is consistent with three groups being nearly identical while one stands apart, and with a good many other arrangements besides."),
  answer("We reject the null hypothesis, which establishes that Black Majority counties specifically have a higher mean Democratic vote percentage than White Majority counties",
         message = "Two problems. A test never establishes a claim; it gives evidence for or against one. And this test asked whether all four means could be equal, which names neither Black Majority counties nor a direction."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "The hypothesis names two of the four categories. Ask whether one F built from all four can be traced back to any particular pair."
)
```

### Testing the Directional Claim

Our hypothesis is specifically about Black Majority counties compared to White Majority counties --- not about comparing either group to Hispanic Majority or Other/No Majority counties, and not about the four categories generally. To test that directly, we filter the data down to just those two categories before running a two-sample t-test.

#### Steps 1 and 2: State the Hypotheses

The null and alternative have been formalized once already. Before building this test, identify the correct `alternative` argument on your own.

**Practice 8: Choose the correct alternative**

```{r quiz-dem2p-alternative, echo=FALSE}
question("**racial_majority** is a plain character variable. Which value of `alternative` correctly tests the directional claim, and why?",
  correct = "Right. Alphabetical order puts Black Majority first, and the hypothesis predicts it higher.",
  answer('`alternative = "greater"`, because R sorts character variables alphabetically and "Black Majority" comes before "White Majority"',
         correct = TRUE,
         message = "Two things have to line up, and only one of them is in the hypothesis. R decides the group order itself --- alphabetically for a character variable --- so \"Black Majority\" is the first group whatever you intended. The hypothesis then says that first group has the higher mean, and `greater` is how you ask R whether the first mean exceeds the second. Had the categories been named so that White sorted first, the same hypothesis would need `less`. The setting follows R's ordering, not the order the hypothesis names the groups in. The formal statement is $H_0: \\mu_{Black} = \\mu_{White}$ against $H_A: \\mu_{Black} > \\mu_{White}$, where $\\mu$ is the mean Democratic vote percentage in the population."),
  answer('`alternative = "less"`, because R sorts character variables alphabetically and "Black Majority" comes before "White Majority"',
         message = 'The alphabetical ordering is right --- "Black Majority" does sort first. But the hypothesis predicts Black Majority counties have a *higher* mean Democratic vote percentage, not a lower one. `alternative = "less"` would test the wrong direction.'),
  answer('`alternative = "greater"`, because Black Majority counties have more Democratic voters and therefore go first in the formula',
         message = 'The substantive reasoning is backwards --- R does not order groups by the size of their means. For a plain character variable, R uses alphabetical ordering regardless of the group means.'),
  answer('`alternative = "two.sided"`, because we cannot know in advance which group R will treat as first',
         message = 'We can know which group comes first --- for a plain character variable, R sorts alphabetically. "Black Majority" sorts before "White Majority," making Black the first group. Since the hypothesis predicts a specific direction, a one-tailed test is appropriate.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: How does R order groups for a plain character variable? Which group comes first alphabetically, and does the hypothesis predict that group has a higher or lower mean?"
)
```

#### Step 3: Run the Test

The ANOVA provided statistically significant evidence that the means are not all equal across the four racial majority categories. Now we test the specific directional claim the hypothesis makes: that Black Majority counties have a higher mean Democratic vote percentage than White Majority counties.

**Practice 9: Build the two-sample t-test for racial majority**

Filter the data down to just the two categories the hypothesis compares, then run the t-test with the correct outcome, grouping variable, and direction. When filtering on a character variable, the category labels must be quoted inside `c()`. You can name the filtered data frame anything you like. Three hints are available. Click "Submit Answer" to check your work.

```{r ttest-dem2p-black-white, exercise=TRUE, exercise.cap="Practice 9"}

```

```{r ttest-dem2p-black-white-hint-1}
# Hint 1 of 3
# Which two categories does the hypothesis specifically
# compare? You will need to narrow the data down to just
# those two categories before comparing their means.
```

```{r ttest-dem2p-black-white-hint-2}
# Hint 2 of 3
# The structure mirrors the urban/rural t-test in the previous
# example: filter() down to the two categories the hypothesis
# compares, then run t.test() with the outcome on the left of
# the tilde and the grouping variable on the right. Check the
# hypothesis to confirm which two categories to keep and which
# direction to test.
```

```{r ttest-dem2p-black-white-hint-3}
# Hint 3 of 3
# Filter to "Black Majority" and "White Majority" using %in%, then
# run t.test() with dem2p_percent as the outcome and racial_majority
# as the grouping variable. You identified the correct alternative
# in the question above.
```

```{r ttest-dem2p-black-white-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Use `filter()` to keep only the two categories the hypothesis compares.")
  }
  if (!grepl("racial_majority\\s*%in%\\s*c\\(", code_no_comments)) {
    fail('Filter with `%in%` and a `c()` holding the two category labels, quoted exactly as they appear in the data.')
  }
  if (!grepl('["\']Black Majority["\']', code_no_comments)) {
    fail('One of the two categories the hypothesis names is missing from your filter.')
  }
  if (!grepl('["\']White Majority["\']', code_no_comments)) {
    fail('The other category the hypothesis names is missing from your filter.')
  }
  if (!grepl("t\\.test\\s*\\(", code_no_comments)) {
    fail("Use `t.test()` to run the two-sample t-test.")
  }
  if (!grepl("dem2p_percent\\s*~\\s*racial_majority", code_no_comments)) {
    fail("The formula puts the outcome on the left of the tilde and the grouping variable on the right.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts which of the two groups has the higher mean. Set `alternative =` to test that direction rather than a two-sided difference.')
  }

  expected_data <- counties %>%
    filter(racial_majority %in% c("Black Majority", "White Majority"))

  expected <- t.test(dem2p_percent ~ racial_majority,
                     data = expected_data,
                     alternative = "greater")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the `t.test()` output. If you're seeing an error, check for a missing tilde (`~`) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail('Check three things: that you filtered to the two categories the hypothesis compares, that the outcome is on the left of the tilde, and that `alternative =` matches the direction predicted.')
  }

  pass("Well done! Filtering to Black Majority and White Majority counties, then comparing their means with `alternative = \"greater\"`, directly tests the hypothesis's specific comparison.")
})
```

#### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 10: Read the confidence interval**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-dem2p-directional-result, echo=FALSE}
question("The t-test output reports a 95% confidence interval of (32.09, Inf) for the difference in means (Black Majority minus White Majority). What does this interval tell us?",
  correct = "Right. The interval excludes 0, agreeing with the p-value.",
  answer("It excludes 0, evidence that the population mean difference is positive",
         correct = TRUE,
         message = "The calculated bound is 32.09, so at this confidence level a difference of zero --- or anything below about 32 percentage points --- is not consistent with these data. The p-value said the same thing: below .001, so we rejected the null. That is not a coincidence, and it is not a second opinion. For a matched pair --- same significance level, same one- or two-tailed specification --- a 95% interval excludes the null value exactly when the p-value falls below .05. They are one statement in two forms, so if you ever find them disagreeing, something is mismatched. The Inf is not a result; it is R declining to estimate a side the test never asked about."),
  answer("The true mean Democratic vote percentage in Black Majority counties falls between 32.09% and infinity",
         message = "The interval is for the difference in means between the two groups, not for the mean of Black Majority counties alone. A 95% confidence interval for the difference tells us the plausible range of values for μ_Black minus μ_White, not for μ_Black by itself."),
  answer("The interval is unreliable because one bound is Inf --- a valid confidence interval should have two finite bounds",
         message = "For a one-tailed test, R fixes one bound at the theoretical extreme rather than estimating it from the data. For a difference in means, that extreme is Inf (no finite upper limit exists for a mean difference). This is expected and correct, not a sign of an error."),
  answer("We cannot interpret this interval because it does not exclude 0",
         message = "The interval (32.09, Inf) excludes 0 entirely --- 0 falls below the lower bound of 32.09. An interval that excludes the null value of 0 points to the same conclusion as rejecting the null hypothesis, and this one does."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: For a one-tailed test, R fixes one bound at a theoretical extreme rather than estimating it. Which bound matters for the conclusion, and does the interval exclude 0?"
)
```

## Is Democracy Associated with Economic Development?

The democracy and GDP example from Tutorial 8 showed a clear pattern at the extremes --- Full Democracies averaged far higher GDP per capita than Authoritarian regimes --- but the middle categories complicated the picture. The argument is that democratic accountability gives leaders an incentive to pursue broadly beneficial economic policy, since they face electoral consequences for poor performance, while non-democratic leaders face fewer such constraints. We start by asking whether the population means could plausibly all be equal across all four regime types, then sharpen to the specific directional contrast the hypothesis predicts.

The hypothesis, unchanged from Tutorial 8:

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have [higher GDP per capita] than will those that are [less democratic].
:::

The data frame **world** contains the ordinal variable **dem_level4**, stored as a factor variable, which takes the values "Full Democ," "Part Democ," "Hybrid," and "Authoritarian," ordered from most to least democratic. It also contains the interval variable **gdppcap08**, GDP per capita in 2008 (World Bank), measured in 2008 US dollars, stored as a numeric variable. Every country has a recorded regime type, while **gdppcap08** is missing for 15 countries, so any summary you compute from it describes the 152 countries with a recorded value rather than all 167 in the data. Narrowing to the two end categories leaves 24 Full Democracies and 52 Authoritarian regimes, and **gdppcap08** is recorded for 23 and 45 of them respectively --- so 68 of those 76 countries enter the test.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{Full Democ}=\mu_{Part Democ}=\mu_{Hybrid}=\mu_{Authoritarian}$$
where $\mu$ is the mean GDP per capita in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

Tutorial 8 found that Full Democracies had a much higher mean GDP per capita than Authoritarian regimes, but the means table alone could not tell us whether a gap that large would be unlikely if the population means were truly all equal. Now we test whether the population means could plausibly all be equal across all four regime types.

**Practice 11: Build the ANOVA for democracy level**

Write the full `oneway.test()` call from scratch. Three hints are available. Click "Submit Answer" to check your work.

```{r anova-gdp-by-demlevel, exercise=TRUE, exercise.cap="Practice 11"}

```

```{r anova-gdp-by-demlevel-hint-1}
# Hint 1 of 3
# This test treats all four regime types at once, asking
# whether the population means could plausibly all be equal.
# Which variable is the outcome you are averaging, and which
# variable defines the groups?
```

```{r anova-gdp-by-demlevel-hint-2}
# Hint 2 of 3
# Use oneway.test(). Its formula puts the interval outcome on the left
# of the tilde and the categorical grouping variable on the right, and
# the data argument tells R which data frame holds them. Both variables
# are in the world data frame.
```

```{r anova-gdp-by-demlevel-hint-3}
# Hint 3 of 3
# The outcome is gdppcap08 and the grouping variable is dem_level4.
```

```{r anova-gdp-by-demlevel-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("oneway\\.test\\s*\\(", code_no_comments)) {
    fail("Use `oneway.test()` to run the one-way ANOVA.")
  }
  if (!grepl("gdppcap08\\s*~\\s*dem_level4", code_no_comments)) {
    fail("The outcome goes on the left of the tilde and the explanatory variable on the right. Which of the two is the quantity whose mean might differ?")
  }
  if (!grepl("data\\s*=\\s*world", code_no_comments)) {
    fail("Set `data = world`.")
  }

  expected <- oneway.test(gdppcap08 ~ dem_level4, data = world)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the `oneway.test()` output. If you're seeing an error, check for a missing tilde (`~`) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check the formula and the data argument --- the outcome on the left of the tilde, the grouping variable on the right, and the data frame these two variables live in.")
  }

  pass("Good work! **gdppcap08** is the outcome and **dem_level4** is the explanatory variable, so it tests whether the mean GDP per capita differs across all four regime types at once.")
})
```

#### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 12: Interpret an ordering hypothesis**

```{r quiz-gdp-general-result, echo=FALSE}
question("Find the p-value in your output. What does this result tell us?",
  correct = "Right. Somewhere among the four, but the test does not say where.",
  answer("We reject the null --- at least one regime-type mean differs, but not which or how",
         correct = TRUE,
         message = "This is the third time the same limitation has come up, and here it bites hardest: the hypothesis predicts an *ordering* across all four regime types, and a single F cannot speak to an ordering at all. It could be large because Full Democracies stand apart from the other three, or because Authoritarian regimes sit far below, or because the middle two differ from each other. Rejecting the null rules out the possibility that all four means are equal, and that is the whole of it. Anything more specific --- including the direction the hypothesis names --- takes a test aimed at particular groups."),
  answer("We fail to reject the null hypothesis --- the p-value is too small to provide reliable evidence of a relationship",
         message = "A very small p-value is stronger evidence against the null, not weaker. A p-value this small leads us to reject the null hypothesis, not fail to reject it."),
  answer("We reject the null hypothesis, which establishes that Full Democracies have significantly higher GDP per capita than Authoritarian regimes",
         message = "Two problems. A test never establishes a claim; it gives evidence for or against one. And this test asked whether all four means could be equal, which names neither Full Democracies nor a direction."),
  answer("We reject the null hypothesis, which provides evidence that GDP per capita rises monotonically with democracy level",
         message = "A monotonic ordering across all four regime types is exactly the kind of specific directional claim the ANOVA p-value cannot evaluate. It tests whether all four means are equal --- it cannot tell us whether the means increase steadily from Authoritarian through to Full Democracy."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value and compare it to .05. Then ask: what exactly does the null hypothesis of a one-way ANOVA state, and what does rejecting it tell you --- and not tell you?"
)
```

### Testing the Directional Claim

Our hypothesis makes a broad prediction about the ordering of all four democracy categories. A two-sample t-test cannot evaluate that complete ordering --- it compares exactly two groups. Here we test one focused implication of the hypothesis: whether countries at the most democratic end of the scale have higher mean GDP per capita than countries at the least democratic end. That is a particularly clear implication of the broader hypothesis and one we can test directly with the two-sample t-test.

#### Steps 1 and 2: State the Hypotheses

This time, identify the complete null and alternative hypothesis pair before building the test.

**Practice 13: State the hypotheses**

```{r quiz-gdp-hypotheses, echo=FALSE}
question("Given that \"Full Democ\" comes before \"Authoritarian\" in R's ordering of **dem_level4**, which pair of hypotheses correctly formalizes the directional claim?",
  correct = "Right. Equal means in the null, the predicted direction in the alternative.",
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} > \\mu_{Authoritarian}$",
         correct = TRUE,
         message = "The null is the no-difference baseline --- the two population means are equal --- and the alternative carries the prediction, that Full Democracies have the higher mean GDP per capita. Both are written with $\\mu$, the population mean, because both are claims about the population rather than about the 152 countries with a recorded value. Notice the stem had to tell you which group R puts first: **dem_level4** is a factor whose levels run from most to least democratic, so R's ordering follows the levels rather than the alphabet --- unlike **racial_majority**, which is a plain character variable. Which group comes first decides whether the alternative uses $>$ or $<$."),
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} < \\mu_{Authoritarian}$",
         message = "The null is right, but the alternative points in the wrong direction. The hypothesis predicts Full Democracies have *higher* GDP per capita than Authoritarian regimes, so the alternative should use $>$, not $<$."),
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} \\neq \\mu_{Authoritarian}$",
         message = "The null is right, but a two-sided alternative does not capture the directional prediction the hypothesis makes. The hypothesis specifically predicts Full Democracies are richer --- that is a one-tailed claim."),
  answer("$H_0: \\mu_{Full Democ} > \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} = \\mu_{Authoritarian}$",
         message = "The null and alternative are reversed. In the hypothesis-testing framework used in this course, the null is the no-difference baseline --- equal population means --- and the alternative expresses the directional prediction we are testing."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: For the tests in this tutorial, what does the null hypothesis say about the two population means? Then check which direction the hypothesis predicts, given that Full Democ comes first."
)
```

#### Step 3: Run the Test

The ANOVA provides evidence that the means are not all equal across the four regime types. The variable description above tells us that **dem_level4**'s factor levels run from most to least democratic --- so "Full Democ" comes before "Authoritarian" in R's ordering, and the difference R computes is Full Democ minus Authoritarian. The hypothesis predicts that difference is positive, so `alternative = "greater"` is the correct choice.

**Practice 14: Build the two-sample t-test for democracy level**

Filter the data to the two categories at opposite ends of the democracy scale, then run the t-test with the correct outcome, grouping variable, and direction. Three hints are available. Click "Submit Answer" to check your work.

```{r ttest-gdp-fulldemoc-authoritarian, exercise=TRUE, exercise.cap="Practice 14"}

```

```{r ttest-gdp-fulldemoc-authoritarian-hint-1}
# Hint 1 of 3
# Which two categories does the hypothesis compare most
# directly? You will need to narrow the data down to just
# those two categories before comparing their means.
```

```{r ttest-gdp-fulldemoc-authoritarian-hint-2}
# Hint 2 of 3
# Use filter() with %in% to keep only the two categories, then run
# t.test() on the result. Its formula takes the outcome on the left of
# the tilde and the grouping variable on the right; data names the
# filtered object and alternative sets the direction. The outcome is
# gdppcap08 and the grouping variable is dem_level4.
# "Full Democ" comes before "Authoritarian" in dem_level4's factor level
# order --- use the hypothesis to determine whether alternative = "less" or "greater".
```

```{r ttest-gdp-fulldemoc-authoritarian-hint-3}
# Hint 3 of 3
# Filter to "Full Democ" and "Authoritarian" using %in%, then run
# t.test() with gdppcap08 as the outcome and dem_level4 as the
# grouping variable. The hypothesis pair you just identified tells
# you the correct alternative.
```

```{r ttest-gdp-fulldemoc-authoritarian-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#[^\r\n]*", "", code)

  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Use `filter()` to keep only the two categories the hypothesis compares.")
  }
  if (!grepl("dem_level4\\s*%in%\\s*c\\(", code_no_comments)) {
    fail('Filter with `%in%` and a `c()` holding the two category labels, quoted exactly as they appear in the data.')
  }
  if (!grepl('["\']Full Democ["\']', code_no_comments)) {
    fail('One of the two categories at the ends of the democracy scale is missing from your filter.')
  }
  if (!grepl('["\']Authoritarian["\']', code_no_comments)) {
    fail('The other category at the ends of the democracy scale is missing from your filter.')
  }
  if (!grepl("t\\.test\\s*\\(", code_no_comments)) {
    fail("Use `t.test()` to run the two-sample t-test.")
  }
  if (!grepl("gdppcap08\\s*~\\s*dem_level4", code_no_comments)) {
    fail("The formula puts the outcome on the left of the tilde and the grouping variable on the right.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts which of the two groups has the higher mean. Set `alternative =` to test that direction rather than a two-sided difference.')
  }

  expected_data <- world %>%
    filter(dem_level4 %in% c("Full Democ", "Authoritarian"))

  expected <- t.test(gdppcap08 ~ dem_level4,
                     data = expected_data,
                     alternative = "greater")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the `t.test()` output. If you're seeing an error, check for a missing tilde (`~`) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail('Check three things: that you filtered to the two categories at the ends of the scale, that the outcome is on the left of the tilde, and that `alternative =` matches the direction predicted.')
  }

  pass("Correct! Filtering to Full Democ and Authoritarian countries, then comparing their means with `alternative = \"greater\"`, directly tests the hypothesis's specific comparison.")
})
```

#### Steps 4 and 5: Find the p-value, Decide, and Report the Result

**Apply 15: Weigh the evidence for the democracy hypothesis**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-gdp-directional-result, echo=FALSE}
question("The t-test result is statistically significant (t = 7.0635, p < .001), with Full Democracies averaging $35,891 in GDP per capita compared to $9,412 for Authoritarian regimes. Which conclusion is best supported by this result combined with the ANOVA?",
  correct = "Right. One implication of the hypothesis, tested; the rest untouched.",
  answer("The endpoint comparison supports one focused implication of the broader hypothesis",
         correct = TRUE,
         message = "$35,891 against $9,412 is a very large gap, and at p < .001 it would be most unlikely if the two population means were equal. But look at what was compared: two of four categories, the endpoints. The hypothesis predicted an ordering across all four, and the two middle categories took no part in this test --- Part Democracies and Hybrid regimes could sit anywhere relative to each other and nothing here would notice. That is not a flaw in the test; it is what a two-sample test is. The honest report is that the sharpest implication of the hypothesis holds, and the rest of the ordering is untested."),
  answer("More democratic countries have significantly higher GDP per capita than less democratic countries, fully establishing the hypothesis",
         message = "The t-test compared only the two endpoint categories --- Full Democracies and Authoritarian regimes. That result supports one focused implication of the hypothesis, but it does not establish the complete ordering across all four regime types. For example, it is consistent with Hybrid regimes having higher GDP than Partial Democracies, which would contradict the overall ordering the hypothesis predicts."),
  answer("The result is only meaningful if all four regime-type comparisons are individually significant",
         message = "Testing every pairwise comparison is not required to learn something meaningful from the hypothesis. The ANOVA provides evidence that not all group means are equal; the targeted t-test evaluated the clearest endpoint implication of the hypothesis. Together they provide evidence for that focused implication, while leaving the complete ordering of all four categories unresolved."),
  answer("The result provides evidence only that Full Democracies and Authoritarian regimes differ --- the hypothesis is fully tested only once all four regime types are compared pairwise",
         message = "The first half is right and the second overstates. A significant endpoint comparison is real evidence for the hypothesis, not a preliminary to be discounted until every pair is tested. What it does not do is establish the full ordering --- so the result is neither the whole hypothesis nor nothing."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The hypothesis predicts a full ordering across all four categories. The t-test compared only two. What does the significant result support, and what does it leave open?"
)
```

## The One-Sample t-Test

Every test in this tutorial so far has asked whether a categorical and an interval variable are related, or whether two groups differ. A different kind of question asks whether a single sample mean differs from some fixed, hypothesized value --- no second variable involved at all. That is what a [**one-sample t-test**]{.important-text} does. For example, we might ask whether the average county-level Democratic vote percentage --- the same **dem2p_percent** variable from the second example --- differs from 50%. A value of 50% represents an evenly split two-party vote; testing against it tells us whether the typical county leaned Democratic or Republican on average in 2016.

The other tests in this tutorial rest on treating the observed cases as realizations of a broader process. This one is built differently, and the contrast is worth seeing. Here the full **counties** data serve as a *population*, and we draw a simple random sample of 150 counties from it. The code below fixes that sample so everyone works from the same 150 counties --- reading it is enough, and nothing later depends on writing it yourself.

```{r ttest-onesample-draw, eval=FALSE}
set.seed(2026)
county_sample_t <- counties[sample(nrow(counties), 150), ]
```

Because the full data happen to be available, we could compute the population mean directly. We sample deliberately so you can see a one-sample t-test doing the job it was designed for: using a sample mean to learn about a population mean you do not know. That is the sampling logic Tutorial 10 develops, now driving an actual test.

We use the same `t.test()` function, but with a different call structure than the formula notation used in the two-sample tests above. Instead of `outcome ~ group`, we supply the variable directly using `data$variable` notation, and use `mu` to specify the hypothesized population mean we are comparing against. There is no `data =` argument here --- the variable is passed in directly.

**Example code:**

```{r ttest-onesample-template, exercise=FALSE, eval=FALSE}
t.test(yourdata$variable, mu = hypothesized_mean)
```

### Step 1: State the Null Hypothesis

$H_0: \mu = 50$, where $\mu$ is the mean Democratic vote percentage across all U.S. counties --- an evenly split two-party vote.

### Step 2: State the Alternative Hypothesis

$H_A: \mu \ne 50$, the population mean differing from an even split in either direction.

### Step 3: Run the Test

**Run and observe 16: A one-sample t-test**

```{r ttest-onesample-dem2p, exercise=TRUE, exercise.cap="Run and observe 16"}
t.test(county_sample_t$dem2p_percent, mu = 50)
```

### Steps 4 and 5: Find the p-value, Decide, and Report the Result

The sample mean is 33.07%, well below the hypothesized value of 50% (t = -14.109, df = 149, p < .001), so we reject the null hypothesis: the mean county-level Democratic vote percentage differs significantly from an even two-party split. The average county in 2016 leaned Republican, not Democratic. That conclusion makes sense given the geography of American politics: Democratic support is concentrated in a relatively small number of densely populated urban counties, while most of the nation's 3,112 counties are smaller and more Republican-leaning.

The output's 95% confidence interval, (30.70, 35.44), gives the same answer a different way: since it excludes the hypothesized value of 50 entirely, it points to the same conclusion as the p-value --- and because this test is two-tailed, both bounds here are calculated from the data rather than fixed at a theoretical extreme.

Because the full county data happen to be available, we can check what the sample was estimating: across all 3,112 counties the mean is 33.34%. The sample of 150 gave 33.07%, and the interval built from it contains the population value. In the ordinary case that value is unknown, and the sample mean, the confidence interval and the test are the evidence you have, which is why the point estimate is normally reported together with its uncertainty.

## The Paired t-Test

The two-sample t-tests in this tutorial compared different groups of cases --- urban counties against rural counties, Full Democracies against Authoritarian regimes, and so on. A [**paired t-test**]{.important-text} asks a different kind of question: when the *same* cases are measured twice, does the value change, on average, between the two measurements?

The data frame **world** contains **gini04** and **gini08**, a country's Gini coefficient --- a standard measure of income inequality, where 0 represents perfect equality and 100 represents perfect inequality --- in 2004 and 2008. Both are stored as numeric variables, and both have substantial gaps: **gini04** is missing for 46 of the 167 countries and **gini08** for 45. A paired test needs both values for the same country, and 120 countries have both, so that is the number the test describes. We might ask whether income inequality changed, on average, across those four years.

The logic is more straightforward than it might look: for each country, subtract its 2004 Gini coefficient from its 2008 Gini coefficient, giving one difference score per country. Then take the mean of those difference scores across all countries. That mean difference --- not the mean of **gini08** alone, and not the mean of **gini04** alone --- is what the test evaluates. A small illustration with three hypothetical countries makes this concrete:

| Country | gini04 | gini08 | Difference (gini08 $-$ gini04) |
|:--|:--:|:--:|:--:|
| Country A | 35.2 | 36.0 | 0.8 |
| Country B | 28.4 | 27.1 | -1.3 |
| Country C | 41.6 | 42.5 | 0.9 |

For these three countries, the mean difference is $(0.8 + (-1.3) + 0.9)/3 = 0.13$. A paired t-test asks whether a mean difference like this is far enough from zero to be surprising under the null hypothesis --- the same logic as every other test in this tutorial, just applied to difference scores instead of raw values.

### Steps 1 and 2: State the Hypotheses

The 120 countries with both measurements are not a random sample of countries. Reading the p-value inferentially therefore means treating these observed changes as realizations of a broader process that could have produced different country-level changes, since inequality in those countries genuinely could have evolved differently between 2004 and 2008. The test then asks whether a mean change this far from zero would be surprising if that process had a mean change of zero.

Let $d = \text{gini08} - \text{gini04}$, the within-country change over the period.

$$H_0: \mu_d = 0$$
$$H_A: \mu_d \ne 0$$

where $\mu_d$ is the mean within-country difference in the broader process. Writing the hypotheses this way matches what the test does: it reduces each country to a single difference score and asks whether the mean of those differences is far from zero. It is not a comparison of two independent population means with `paired = TRUE` added on.

We use the same `t.test()` function again, but like the one-sample test above, the paired t-test uses a different call structure than the formula notation used in the two-sample tests. Instead of `outcome ~ group`, we supply both variables directly --- `t.test(x, y, paired = TRUE)` --- and set `paired = TRUE` to tell R to compute the difference for each country first, exactly as in the table above, and then run a test on the mean of those differences, rather than treating **gini08** and **gini04** as two separate, unrelated groups of countries.

**Example code:**

```{r ttest-paired-template, exercise=FALSE, eval=FALSE}
t.test(yourdata$second_measure, yourdata$first_measure, paired = TRUE)
```

### Step 3: Run the Test

**Run and observe 17: A paired t-test**

```{r ttest-paired-gini, exercise=TRUE, exercise.cap="Run and observe 17"}
t.test(world$gini08, world$gini04, paired = TRUE)
```

### Steps 4 and 5: Find the p-value, Decide, and Report the Result

Notice that the output reports a single mean difference, not two separate group means the way an unpaired t-test would --- a direct reflection of the logic above. Here, the mean difference is 0.41 (t = 0.952, df = 119, p = .3431). Because the p-value is well above .05, we fail to reject the null hypothesis: we do not have evidence that the mean Gini coefficient changed between 2004 and 2008.

The 95% confidence interval, (-0.440, 1.255), tells the same story. In the earlier tests that rejected the null, the confidence interval excluded the relevant null value --- usually 0 for a difference in means, and 50 for the one-sample test above. Here, 0 falls comfortably inside the interval, right alongside a wide range of other plausible values. That is exactly what a confidence interval looks like when a test fails to reject: not evidence that there is no difference, but evidence that we cannot rule out 0, along with a great many other values, as the true mean difference.

## Check Your Understanding

The questions below test whether you have internalized the key ideas from this tutorial: how to set up the right comparison when a hypothesis requires combining categories, what a significant ANOVA result can and cannot tell you, when a paired t-test is appropriate, and how to translate a hypothesis into the correct `alternative` argument given R's group ordering rules.

**Apply 18: Set up the right comparison**

```{r cyu-recognize-combining, echo=FALSE}
question("A researcher hypothesizes that countries with consensus-oriented electoral systems --- either proportional representation or mixed --- have higher average voter turnout than countries with majoritarian systems --- either plurality or runoff. Electoral system is measured with a four-category variable taking the values 'Proportional,' 'Mixed,' 'Plurality,' and 'Runoff.' Which approach correctly sets up a two-sample t-test of this hypothesis?",
  correct = "Right. Neither group exists as a category, so both have to be built.",
  answer("Create a two-group variable combining 'Proportional' with 'Mixed' and 'Plurality' with 'Runoff', then run a two-sample t-test",
         correct = TRUE,
         message = "The hypothesis names two groups and the variable has four categories, and neither group matches a category --- so both sides have to be assembled before any test can run. This is a step beyond the examples in this tutorial, where you filtered down to two categories that already existed. Filtering cannot help here: dropping to any two categories would leave out cases the hypothesis includes. Recoding is the general move, and the rule for it is the same one that governed filtering --- the groups come from the hypothesis, not from what the variable happens to offer."),
  answer("Filter to 'Proportional' and 'Plurality' only, since those are the most representative categories on each side of the comparison",
         message = "Choosing categories based on representativeness rather than what the hypothesis predicts changes the question being asked. The hypothesis defines both comparison groups explicitly --- consensus systems include both 'Proportional' and 'Mixed,' and majoritarian systems include both 'Plurality' and 'Runoff.' Dropping either pair leaves out cases the hypothesis includes."),
  answer("Run a one-way ANOVA on all four categories, since the hypothesis involves more than two groups",
         message = "The hypothesis does not ask whether the four categories differ from each other --- it predicts a specific directional contrast between two theoretically defined groups, each of which combines two categories. ANOVA treats all four categories simultaneously and cannot test that specific grouping."),
  answer("Recode or create a new grouping variable that combines 'Proportional' and 'Plurality' into one group and 'Mixed' and 'Runoff' into another, then run a two-sample t-test on that variable",
         message = "The grouping has to follow the hypothesis, not split the categories evenly. The hypothesis contrasts consensus systems ('Proportional' and 'Mixed') against majoritarian systems ('Plurality' and 'Runoff') --- combining 'Proportional' with 'Plurality' mixes categories from opposite sides of that contrast and would test a comparison the hypothesis never made."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: How many groups does the hypothesis define, and do those groups match existing categories in the variable? What do you need to do when neither side of the comparison maps to a single category?"
)
```

**Apply 19: Choose the right test**

```{r cyu-choose-the-test, echo=FALSE}
question("A researcher measures political trust in the same 500 survey respondents before and after a national election, and wants to know whether trust changed, on average, between the two measurements. Which test is most appropriate?",
  correct = "Right. Same cases measured twice.",
  answer("A paired t-test, because the same respondents are measured at both time points",
         correct = TRUE,
         message = "The question is not whether two groups differ but whether one group changed, and that difference decides the test. With the same 500 people measured twice, each respondent supplies a before and an after, so you can subtract one from the other and get a change score per person --- which is what a paired t-test does before testing anything. Pairing is also what makes the test sensitive: a respondent who is distrustful at both time points contributes a difference near zero rather than adding noise at both ends. Treating the measurements as two independent samples throws that away."),
  answer("A two-sample t-test, because there are two groups of measurements --- before and after",
         message = "The before and after measurements are not two independent groups --- they are two measurements on the same 500 people. A two-sample t-test assumes the two groups are independent, which is not the case here. When the same cases are measured twice, a paired t-test is the appropriate tool."),
  answer("A one-way ANOVA, because the researcher wants to compare means across two time points",
         message = "ANOVA is used when comparing means across categories of a grouping variable across independent groups. Here the same respondents are measured twice, which calls for a paired t-test rather than an omnibus comparison across groups."),
  answer("A one-sample t-test, because the question is about a single mean --- the mean change in trust",
         message = "A one-sample t-test compares a single sample mean to a fixed hypothesized value, with no second measurement involved. While the paired t-test does ultimately test whether a mean difference differs from zero, it works by first computing a difference score for each respondent and then testing that. The one-sample t-test cannot account for the paired structure of the data."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Are the two groups of measurements independent of each other, or do they come from the same cases? That distinction determines which test is appropriate."
)
```

**Apply 20: Predict the right direction for a new hypothesis**

```{r cyu-direction-prediction, echo=FALSE}
question('A hypothesis predicts that countries with high judicial independence have lower rates of corruption than countries with low judicial independence. The grouping variable **judicial_independence** is stored as a plain character variable and takes the values "High" and "Low". If you ran `t.test(corruption ~ judicial_independence, data = ..., alternative = ...)`, which value of `alternative` would test this hypothesis directly?',
  correct = "Right. \"High\" sorts first, and the hypothesis puts it lower.",
  answer('`alternative = "less"`',
         correct = TRUE,
         message = "Two facts have to be combined, and neither alone gives the answer. R sorts a plain character variable alphabetically, so \"High\" is the first group --- a fact about your data, not about your hypothesis. The hypothesis then says that group has the *lower* mean corruption, and `less` is how you ask R whether the first mean falls below the second. Had the categories been labeled so that \"Low\" sorted first, the same hypothesis would need `greater`. This is the last of three places in this tutorial where group order decided the setting, and the rule has been the same each time: find out which group R puts first, then translate the hypothesis into a claim about that group."),
  answer('`alternative = "greater"`',
         message = 'With "High" sorted first, `alternative = "greater"` would test whether the High group\'s mean is higher than the Low group\'s --- the opposite of what this hypothesis predicts.'),
  answer('`alternative = "two.sided"`, since we cannot know the direction in advance',
         message = "We can know the direction in advance --- it is what the hypothesis predicts. A two-sided test would be appropriate if we had no directional prediction, but this hypothesis makes one explicitly."),
  answer('The value of `alternative` does not depend on which group R lists first --- it only depends on the direction of the hypothesis',
         message = 'The value of `alternative` depends on both: the direction the hypothesis predicts AND which group R lists first. `alternative = "less"` tests whether the first group\'s mean is lower than the second\'s. If R listed Low before High, you would need `alternative = "greater"` to test the same hypothesis. Always check the group order before choosing `alternative`.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: How does R order groups for a plain character variable? Which value comes first alphabetically --- High or Low? The hypothesis predicts that first group has a lower mean. Which `alternative` matches?"
)
```

## The Takeaways

The table below adds a categorical and an interval variable to the combination introduced in Tutorial 12.

::: {.qpa-infographic}
<table class="qpa-table">
  <caption>Testing Categorical and Interval Variables</caption>
  <thead>
    <tr>
      <th class="qpa-corner" scope="col"><span class="visually-hidden">Aspect of the test</span></th>
      <th scope="col">Two Categorical Variables</th>
      <th scope="col">One Categorical, One Interval Variable</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">What the Test Asks</th>
      <td>Whether the variables are related at all; then whether two specific proportions differ</td>
      <td>Whether any group means differ; then whether two specific means differ</td>
    </tr>
    <tr>
      <th scope="row">Null Hypothesis</th>
      <td>The variables are independent; or &pi;<sub>1</sub> = &pi;<sub>2</sub></td>
      <td>&mu;<sub>1</sub> = &mu;<sub>2</sub> = &hellip; = &mu;<sub>k</sub>; or &mu;<sub>1</sub> = &mu;<sub>2</sub></td>
    </tr>
    <tr>
      <th scope="row">Test Statistic</th>
      <td>Chi-square, for both tests</td>
      <td>F for the ANOVA; t for the t-tests</td>
    </tr>
    <tr>
      <th scope="row">R Functions</th>
      <td><code>CrossTable(chisq = TRUE)</code>, <code>prop.test()</code></td>
      <td><code>oneway.test()</code>, <code>t.test()</code></td>
    </tr>
  </tbody>
</table>
:::

This tutorial applied the general-to-directional pattern from Tutorial 12 to a categorical and an interval variable, subjecting the group-mean relationships from Tutorial 8 to formal **hypothesis tests**. A one-way ANOVA asked whether the population means could plausibly all be equal across every category at once; a two-sample t-test sharpened that to a specific directional contrast between two groups. Two additional tests asked related but different questions: whether a single mean differs from a fixed value, and whether the average difference between two paired measurements differs from zero.

- The [**one-way ANOVA**]{.important-text} asks the general question: could the population means be equal across all categories? Run it with `oneway.test(outcome ~ group, data = df)`, which compares [**between-group variability**]{.important-text} --- how far the group means sit from one another --- against [**within-group variability**]{.important-text}, the spread of cases around their own group's mean, and reports the ratio as the [**F-statistic**]{.important-text}. By default, R does not assume equal variances across groups. Because ANOVA considers every category simultaneously, its p-value has no direction built into it --- a significant result tells you at least one mean differs from at least one other, not which specific groups drive that difference or in what direction.

- The [**two-sample t-test**]{.important-text} asks the sharper, directional question: is the mean higher (or lower) in one specific group than in another? Run it with `t.test(outcome ~ group, data = df, alternative = ...)`.
  - When the categorical variable has more than two categories, use `filter()` with `%in%` to narrow the data to the two categories the hypothesis names before running `t.test()`. When the hypothesis defines groups that each combine multiple categories, recode or create a new grouping variable so each group maps to a single value before running the test.
  - R determines which group comes first by how the variable is stored: numerically for numeric variables, alphabetically for plain character variables, and by stored level order for factor variables. Always check which group comes first before choosing `alternative = "less"` or `"greater"`.
  - A planned directional t-test can also stand on its own when a hypothesis specifies the comparison from the start --- the two-step ANOVA-then-t-test pattern is a teaching structure, not a requirement.

- The [**one-sample t-test**]{.important-text} asks a different question: whether a single sample mean differs from a fixed hypothesized value, with no second group involved. Run it with `t.test(x, mu = ...)`, supplying the variable directly rather than using formula notation.

- The [**paired t-test**]{.important-text} asks whether the average difference between two variables measured on the same cases differs from zero. Run it with `t.test(x, y, paired = TRUE)`, supplying both variables directly. The key distinction: use a paired t-test when the same cases are measured twice, and a two-sample t-test when the two groups are independent.

Every test follows the same **five-step procedure**: state $H_0$, state $H_A$, run the test, find the p-value, and decide whether to reject or fail to reject --- then report what that decision means substantively. Rejecting a null hypothesis --- even with a very small p-value --- never proves a hypothesis is true, and failing to reject it never proves it is false. A p-value tells us how unusual a test statistic at least as extreme as the one observed would be if the null hypothesis were true; a small p-value is **statistically significant evidence in support of a hypothesis**, not certainty that it is right.

Tutorial 14 completes the sequence by applying the same logic to two interval variables. The tool changes again --- a correlation test instead of ANOVA and t-tests --- but the procedure is identical, and the same five steps carry over unchanged.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-13, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-13, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-13, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You can now test relationships between categorical and interval variables using ANOVA and t-tests — and you know exactly what those tests can, and cannot, tell you.">
  <span aria-hidden="true">🎉</span>
  Excellent work! You can now test relationships between categorical and interval variables using ANOVA and t-tests &mdash; and you know exactly what those tests can, and cannot, tell you.
</p>
')
```
