---
title: "Hypothesis Tests: Categorical and Interval Variables"
tutorial:
  id: "13-Hypotheses-Categorical-Interval"
output: 
  QPATutorialsCourse::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to conduct a hypothesis test of the relationship between a categorical and an interval variable.
---

## Overview {#main-content}

Tutorial 8 described the relationship between a categorical and an interval variable by comparing group means --- but every example ended with the same caution: a difference in sample means cannot tell us whether that difference is larger than we would expect by chance. In Tutorials 10 and 11, you built the tools a formal test requires: the sampling distribution, the p-value, and the decision rule. This tutorial takes the group-mean relationships from Tutorial 8 and subjects them to formal [**hypothesis tests**]{.important-text} --- using sample data to assess whether observed differences in means are strong enough to rule out chance. Tutorial 13 does the same for the group-mean relationships from Tutorial 8, and Tutorial 14 for the correlations from Tutorial 9.

In this tutorial, you will learn how to test whether a categorical and an interval variable are related in the population. Specifically, we will cover:

* How to test whether the means of an interval variable differ across categories of a categorical variable using a one-way ANOVA, with the `oneway.test()` function
* How to decide when an ANOVA is the right tool, when a two-sample t-test is, and how comparing group means lets you ask the general question before deciding whether to sharpen to a specific one
* How to organize a hypothesis test into five explicit steps, building on the null hypothesis, alternative hypothesis, p-value, and decision rule from Tutorial 11 and the five-step procedure introduced in Tutorial 12
* How to interpret a p-value and apply the decision rule to determine whether a relationship is statistically significant
* How to test whether a single sample mean differs from a hypothesized population value
* How to test whether the mean difference between two variables measured on the same cases differs from zero, using a paired t-test


```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr,
                 exercise.completion = FALSE)
knitr::opts_chunk$set(error = TRUE)
counties$dem2p_percent <- counties$dem2p_vote_share*100
counties$rural <- 1
counties$rural[counties$rural_urban <= 3] <- 0
counties$rural[counties$rural_urban >= 8] <- 2
```

## Testing the Relationship Between a Categorical Variable and an Interval Variable

In Tutorial 8, you compared means of an interval variable across categories of a categorical variable. But a sample is just one sample --- if we drew a different sample from the same population, the means would differ at least a little, just by chance. A [**one-way ANOVA**]{.important-text} evaluates how surprising the differences among our sample's group means would be if the corresponding population means were actually all equal.

The logic compares two kinds of variability. [**Between-group variability**]{.important-text} is how far the group means are from each other. [**Within-group variability**]{.important-text} is how spread out individual cases are around their own group's mean --- variability that has nothing to do with group membership at all. ANOVA summarizes the relationship between these two quantities in a single test statistic, the [**F-statistic**]{.important-text}:

$$F=\frac{\text{between-group variability}}{\text{within-group variability}}$$

If the population means really are all equal, the differences among our sample's group means reflect nothing but chance, so $F$ should be close to 1. If the population means differ, the group means will tend to be far apart relative to the ordinary spread within each group, and $F$ will be large. As with the test statistics in earlier tutorials, you do not need to remember the formula behind $F$ --- understand the logic, and let R compute it for you.

The formula above captures the intuition. In practice, R's `oneway.test()` uses a Welch adjustment that handles unequal variances across groups --- the same default it applies to the two-sample t-test. This modifies how the denominator is calculated but preserves the same broad logic: large between-group spread relative to within-group spread produces a large $F$ and a small p-value.

The figure below makes this concrete. Both panels show three groups with exactly the same three means. The only thing that differs is how spread out the individual values are within each group.

```{r concept-f-statistic-illustration, echo=FALSE, fig.alt="Two side-by-side panels of dot plots with box overlays, each showing three groups --- Group A, Group B, and Group C --- with identical group means of 10, 15, and 20, marked by black diamonds. In the left panel, labeled 'Low within-group spread (large F),' each group's points are tightly clustered around its mean with little overlap between groups. In the right panel, labeled 'High within-group spread (small F),' each group's points are widely spread with substantial overlap between adjacent groups, even though the group means remain at the same three positions."}
set.seed(13)
groups <- c("Group A", "Group B", "Group C")
means  <- c(10, 15, 20)
n <- 30

# Generate a single set of mean-zero deviations, then scale for each panel
# This guarantees the sample means are exactly 10, 15, 20 in both panels
raw_devs <- lapply(seq_along(means), function(i) {
  d <- rnorm(n)
  d - mean(d)   # center so sample mean is exactly 0
})

low_spread <- data.frame(
  group = rep(groups, each = n),
  value = c(means[1] + raw_devs[[1]] * 1,
            means[2] + raw_devs[[2]] * 1,
            means[3] + raw_devs[[3]] * 1),
  panel = "Low within-group spread\n(large F)"
)

high_spread <- data.frame(
  group = rep(groups, each = n),
  value = c(means[1] + raw_devs[[1]] * 5,
            means[2] + raw_devs[[2]] * 5,
            means[3] + raw_devs[[3]] * 5),
  panel = "High within-group spread\n(small F)"
)

f_illustration <- rbind(low_spread, high_spread)
f_illustration$panel <- factor(f_illustration$panel,
                                levels = c("Low within-group spread\n(large F)",
                                           "High within-group spread\n(small F)"))

ggplot(f_illustration, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 4, color = "black") +
  facet_wrap(~panel) +
  labs(title = "Same Group Means, Different Within-Group Spread",
       subtitle = "Black diamonds mark the group means",
       x = NULL,
       y = "Value",
       caption = "Hypothetical data") +
  guides(fill = "none")
```

Both panels have exactly the same group means --- 10, 15, and 20, marked by the black diamonds --- but the left panel's tightly clustered values produce a large $F$, while the right panel's widely spread, overlapping values produce a small $F$. The means are identical; only the within-group variability differs. That is precisely what $F$ is measuring.

In Tutorials 10 and 11, the sampling distribution you built intuition for was the normal curve --- symmetric and bell-shaped. In Tutorial 12, $\chi^2$ introduced a different shape: bounded at 0 and stretched into a long right tail, because it is built entirely from squared terms. $F$ follows yet another distribution, for a closely related reason --- it is a ratio of two variability measures, each built from squared terms, so $F$ itself can never be negative either. Like $\chi^2$, its distribution starts at 0 and skews to the right rather than spreading symmetrically. The exact shape depends on two degrees-of-freedom values --- the numerator and denominator df R reports with the test --- rather than just one, but as before, the precise calculation is not something we will dwell on.

What does not change is the underlying logic. Just as with the normal and chi-square distributions, a hypothesis test locates your observed $F$ on this distribution and asks how far into the tail it falls. An $F$ value far out in that long right tail is one that would be rare if the population means were really all equal --- exactly the kind of evidence a hypothesis test is built to evaluate. R converts $F$ into a p-value this way automatically, and we apply the same decision rule as always: a small p-value (conventionally below .05) leads us to reject the null hypothesis that the population means are equal, while a large p-value means we fail to reject it.

### Running the Test with `oneway.test()`

To run a one-way ANOVA in R, we use the `oneway.test()` function: a formula, with the interval (outcome) variable on the left of a tilde (`~`) and the categorical (explanatory) variable on the right, plus a `data` argument: `oneway.test(outcome ~ group, data = df)`.

Notice that the formula uses bare variable names --- `outcome` and `group` --- not `df$outcome` and `df$group`. This is the same `data` argument you first saw all the way back in Tutorial 4's very first `ggplot()` call: whenever a function includes a `data` argument, that argument tells R which data frame to search for the variable names that appear elsewhere in the call, so you never need the `dataframe$variable` notation alongside it. You will see this same pattern in `t.test()` below.

By default, `oneway.test()` does not assume the population variances are equal across groups --- the same default stance R takes for the two-sample t-test you will see next. The output reports the F-statistic, two degrees-of-freedom values, and a p-value. Because of how the unequal-variance correction is calculated, the denominator degrees of freedom is often not a whole number --- that is expected, not an error, and it is not something we will focus on in this course. The test statistic and the p-value are what we use to reach a conclusion.

`oneway.test()`, `t.test()`, and `filter()` are all available without any `library()` call beyond what is already loaded in this tutorial --- you do not need to load any additional packages.

### Running the Test with `t.test()` on Two Specific Groups

A one-way ANOVA treats every category at once, and its p-value does not directly test a specific directional contrast. When a hypothesis predicts a specific direction between two groups, we can sharpen to a [**two-sample t-test**]{.important-text}, using the same `t.test()` function you will also use for the one-sample test later in this tutorial.

Tutorials 10 and 11 already named this one: the t-distribution is the small-sample alternative to the normal curve they mentioned in passing. Now that you are actually using it, here is what that means. Unlike chi-square (Tutorial 12) and F (above), which are right-skewed and bounded at zero, t is symmetric and bell-shaped --- the same general shape as the normal curve itself. The difference shows up in the tails: with smaller samples, individual estimates of variability are themselves less reliable, so extreme values of the test statistic are a bit more likely than the normal curve alone would suggest, and the t-distribution's wider tails account for that. The exact shape depends on degrees of freedom --- few degrees of freedom means noticeably heavier tails, and as degrees of freedom grow, the t-distribution converges to the normal curve. As with chi-square and F, the logic does not change: locate your observed statistic on this distribution, see how far into the tail it falls, and let R compute the resulting p-value --- whichever version of `t.test()` you are running: two-sample, one-sample, or paired.

`t.test()` compares the means of exactly two groups. When a categorical variable has exactly two categories, we can use formula notation directly: `t.test(outcome ~ group, data = df)`. Several of the categorical variables in this tutorial have three or four categories, so before running a directional t-test we will first use `filter()` to narrow the data down to the two specific categories the hypothesis compares, then use formula notation on that filtered data. The `%in%` operator, used inside `filter()`, checks whether each case's value matches any value in the list that follows --- `filter(group %in% c("A", "B"))` keeps only cases where `group` equals "A" or "B" and drops all others. You will see this pattern in every directional t-test in this tutorial.

As with `prop.test()` in Tutorial 12, the `alternative` argument controls whether the test is one-tailed or two-tailed. The default, `alternative = "two.sided"`, tests whether the two means differ in either direction. When a hypothesis predicts a specific direction, we use `alternative = "less"` or `alternative = "greater"`. The key rule: `alternative = "less"` tests whether the *first* group's mean is lower than the *second's*; `alternative = "greater"` tests whether the first group's mean is higher. For numeric grouping variables, R orders groups by their numeric value. For character variables, R orders groups alphabetically. Always check which group will come first before choosing `alternative` --- the labels attached to the group estimates in the output are an easy way to confirm R's ordering.

This is also a good moment to reconnect with Tutorial 11's other major tool, and with what you just saw in Tutorial 12. Alongside the p-value, `t.test()` reports a confidence interval. For a two-sample t-test, that interval is for the difference in means. For a one-sample t-test, it is for the population mean. For a paired t-test, it is for the mean of the paired differences.

When the confidence interval and hypothesis test are matched --- same significance level and same one-tailed or two-tailed specification --- they point to the same conclusion. Here, the hypothesized difference is usually 0. If the interval excludes 0, that matches a p-value small enough to reject the null. For a one-tailed test, just as with `prop.test()`, one side of the interval will appear as a theoretical extreme rather than a value calculated from the data. For t-tests, that extreme is $\pm\infty$, since a mean or mean difference has no fixed upper or lower limit. The bound that matters is the calculated one: does the interval exclude the hypothesized value?

### Applying the Five-Step Hypothesis Testing Procedure

In Tutorial 12, we organized the building blocks of a hypothesis test --- introduced conceptually in Tutorials 10 and 11 --- into five explicit steps:

1. State the null hypothesis, $H_0$.
2. State the alternative hypothesis, $H_A$.
3. Run the test and obtain a test statistic.
4. Find the p-value associated with that test statistic.
5. Decide whether to reject or fail to reject $H_0$, and report what that means substantively.

We will follow these same five steps to test relationships between a categorical and an interval variable.

## Is College Attainment Lower in Rural Counties?

College attainment was one of the cleaner patterns in Tutorial 8: the means dropped steadily from urban to neither to rural, with urban counties averaging well above the other two categories. The argument is intuitive --- rural areas have fewer local jobs requiring a degree and less access to higher education, so college graduates tend to concentrate in cities. Now we ask whether gaps as consistent as those across the three rurality categories would be surprising by chance, then sharpen to the specific urban-versus-rural contrast the hypothesis predicts.

::: {.hypothesis}
In a comparison of [counties], those that are [more rural] will have a [lower share of college graduates] than will those that are [more urban].
:::

The data frame **counties** contains the nominal variable **rural**, which takes the value 0 for urban counties, 1 for counties that are neither distinctly urban nor rural, and 2 for rural counties. It also contains the interval variable **prop_college_grad**, the proportion of a county's population that has completed a college degree. **rural** is our explanatory variable and **prop_college_grad** is our outcome.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{urban}=\mu_{neither}=\mu_{rural}$$
where $\mu$ is the mean proportion of college graduates in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

**Run and observe: Run the one-way ANOVA**

```{r anova-collegegrad-by-rural, exercise=TRUE, exercise.cap="Run and observe: One-way ANOVA for college graduation rates by rurality"}
oneway.test(prop_college_grad ~ rural, data = counties)
```

The output reports, in order: the F-statistic (labeled `F`), the numerator and denominator degrees of freedom (`num df` and `denom df`), and the p-value. As noted earlier, the denominator degrees of freedom will likely not be a whole number --- that is expected.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value in your output and compare it to .05 before answering.

```{r quiz-collegegrad-general-result, echo=FALSE}
question("Find the p-value in your output. Which statement best summarizes what this test tells us?",
  correct = "Right. The ANOVA establishes that the population means are not all equal, but its p-value does not directly test whether urban counties specifically have a higher mean than rural counties. We can see that direction in the group means; the t-test evaluates it directly.",
  answer("We reject the null hypothesis --- providing evidence that the means are not all equal across the three rurality categories --- but the ANOVA p-value does not directly test whether urban counties specifically have a higher mean than rural counties",
         correct = TRUE),
  answer("We reject the null hypothesis, which provides strong evidence that urban counties have a higher mean proportion of college graduates than rural counties",
         message = "Rejecting this null hypothesis only provides evidence that the population means are not all equal across the three categories (F = 239.35, p < .001). It does not, by itself, evaluate the specific direction our hypothesis predicts --- for that we need the t-test that follows."),
  answer("We fail to reject the null hypothesis, so there is no relationship between rural and prop_college_grad",
         message = "A p-value this small (p < .001) leads us to reject the null hypothesis, not fail to reject it. Failing to reject would require a p-value above .05."),
  answer("We reject the null hypothesis, and since the relationship runs across all three rurality categories, the directional claim is also established",
         message = "An ANOVA p-value tests all categories simultaneously and has no direction built into it. Rejecting the null only tells us the population means are not all equal --- it does not tell us whether urban counties specifically have a higher mean than rural counties. A targeted directional test is still needed."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value in your output. A one-way ANOVA treats all categories at once --- what can it tell you, and what can it not tell you about the specific direction?"
)
```

### Testing the Directional Claim

Our hypothesis does not just predict that the three group means differ --- it predicts a specific direction: the mean proportion of college graduates should be higher among urban counties than among rural counties specifically. Looking at the group means, we can see that direction in the data. But the ANOVA p-value does not directly test that specific one-tailed contrast. To do that, we compare the two endpoint categories --- urban and rural --- using `filter()` to narrow the data to just those two groups before running a two-sample t-test.

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{urban}=\mu_{rural}$$
where $\mu$ is the mean proportion of college graduates in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \mu_{urban} > \mu_{rural}$$

#### Step 3: Run the Test

**Run and observe: Run a one-tailed two-sample t-test**

We set `alternative = "greater"` because the hypothesis predicts that the urban mean is *higher* than the rural mean --- not just different from it. The `%in%` operator checks whether each county's value matches *any* value in the list that follows --- here, keeping only counties where `rural` equals 0 or 2, and dropping those equal to 1. Filtering this way leaves exactly the two groups `t.test()`'s formula notation requires.

Load [dplyr]{.package-name} with `library(dplyr)` before using `filter()` and the pipe. Once loaded, it is available for all subsequent exercises in this tutorial.

```{r ttest-collegegrad-urban-rural, exercise=TRUE, exercise.cap="Run and observe: One-tailed two-sample t-test for college graduation by rurality"}
library(dplyr)
counties_urban_rural <- counties %>%
  filter(rural %in% c(0, 2))

t.test(prop_college_grad ~ rural, data = counties_urban_rural, alternative = "greater")
```

**Practice: Match the output's group labels to the categories they represent**

Look at how **rural** is coded and match the numeric labels in the output to the categories they represent.

```{r quiz-collegegrad-group-labels, echo=FALSE}
question("In the t.test() output above, the alternative hypothesis line refers to 'group 0' and 'group 2', and the sample estimates report 'mean in group 0' and 'mean in group 2'. Which counties do these labels refer to?",
  correct = "Right. Rural is coded 0 for urban and 2 for rural --- R reports those numeric codes directly, not category names. Always translate numeric codes back into substantive labels before interpreting t.test() output.",
  answer("Group 0 refers to urban counties, and group 2 refers to rural counties",
         correct = TRUE),
  answer("Group 0 refers to rural counties, and group 2 refers to urban counties",
         message = "This reverses the coding. rural is coded 0 for urban counties and 2 for rural counties --- not the other way around."),
  answer("Group 0 and group 2 refer to two different values of prop_college_grad, not to categories of rural",
         message = "prop_college_grad is the outcome variable being averaged within each group --- it is not itself split into 'group 0' and 'group 2'. Those labels come from the values of rural, the explanatory variable."),
  answer("It is not possible to tell which counties group 0 and group 2 refer to from this output alone",
         message = "It is possible --- R reports the numeric codes from the explanatory variable directly. You need to recall, from how rural was defined, which code corresponds to which category."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Look back at how rural was defined when we introduced this example --- which numeric code was assigned to urban counties, and which to rural counties?"
)
```

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-collegegrad-directional-result, echo=FALSE}
question("Find the p-value in your output. Does this result provide statistically significant evidence in support of the hypothesis?",
  correct = "Right. The p-value is far below .05 --- we reject the null hypothesis. Urban counties have a significantly higher mean proportion of college graduates than rural counties, matching the predicted direction.",
  answer("Yes --- the p-value is far below .05, and the mean proportion of college graduates is significantly higher among urban counties than rural counties, matching the predicted direction",
         correct = TRUE),
  answer("Yes, but only weakly --- the p-value is small because the sample has over 3,000 counties, not because the effect is meaningful",
         message = "Sample size does affect p-values, but the group means themselves tell the story too: urban counties average 25.5% college graduates compared to 17.7% for rural counties --- a nearly 8-percentage-point gap. Both the statistical significance and the magnitude of the difference support the hypothesis."),
  answer("No --- the p-value is below .05, so we fail to reject the null hypothesis",
         message = "This reverses the decision rule. A p-value *below* .05 leads us to *reject* the null hypothesis, not fail to reject it. We fail to reject the null only when the p-value is *above* .05."),
  answer("We cannot conclude anything about direction from this test because the ANOVA already established the relationship",
         message = "The ANOVA established that the means are not all equal across the three groups, but its p-value does not directly test whether urban counties specifically have a higher mean than rural counties. This t-test is what directly evaluates that contrast. The two results answer different questions."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Find the p-value and compare it to .05. Then check whether the sample estimates match the direction the hypothesis predicted: which group (0 or 2) has the higher mean?"
)
```

The output also reports a 95% confidence interval of (0.072, Inf) for the difference in means (group 0 minus group 2). For a one-tailed test, R fixes one bound at the theoretical extreme --- here, Inf, since a mean difference has no fixed upper limit the way a proportion does (which is bounded between -1 and 1). The bound that matters is the calculated one: 0.072. Because the interval excludes 0, it points to the same conclusion as the p-value.

## Is Democratic Vote Share Related to a County's Racial Majority?

The racial majority example from Tutorial 8 produced one of the starkest contrasts in the tutorial: Black Majority counties averaged a Democratic vote share more than 30 percentage points above White Majority counties. The argument connects to decades of party coalition-building --- the Democratic Party's positions on civil rights and voting rights have made it the party more closely associated with addressing Black voters' political concerns, while the Republican Party has drawn disproportionate support from White voters over the same period. Formally testing this hypothesis introduces a new challenge: the variable has four categories, but the hypothesis names only two of them. We need to filter before we can test.

::: {.hypothesis}
In a comparison of [counties], those with a [Black racial majority] will have a [higher mean Democratic vote share] than will those with a [White racial majority].
:::

The data frame **counties** contains the nominal variable **racial_majority**, which takes the values "White Majority," "Black Majority," "Hispanic Majority," and "Other/No Majority." It also contains the interval variable **dem2p_percent**, the Democratic candidate's share of the two-party vote in the county (as a percent).

As in the previous example, we start with the general claim that the population means differ across all four categories, then sharpen to the specific comparison the hypothesis actually makes: Black Majority counties against White Majority counties.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{White}=\mu_{Black}=\mu_{Hispanic}=\mu_{Other}$$
where $\mu$ is the mean Democratic vote share in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

The comparison of group means in Tutorial 8 showed that Black Majority counties had a noticeably higher mean Democratic vote share than White Majority counties. Now we test whether a difference that large across all four categories would be surprising if the population means were actually equal.

**Practice: Run the one-way ANOVA**

Fill in the explanatory variable on the right side of the tilde. One hint is available. Click Submit Answer to check your work.

```{r anova-dem2p-by-racialmajority, exercise=TRUE, exercise.cap="Practice: One-way ANOVA for Democratic vote share by racial majority"}
oneway.test(dem2p_percent ~ ___, data = counties)
```

```{r anova-dem2p-by-racialmajority-hint-1}
# Hint 1 of 1
# The structure mirrors the Run and observe above --- same function,
# same data argument. Which variable defines the groups you are
# comparing here? That goes on the right side of the tilde.
```

```{r anova-dem2p-by-racialmajority-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("oneway\\.test\\s*\\(", code_no_comments)) {
    fail("Use oneway.test() to run the one-way ANOVA.")
  }
  if (!grepl("dem2p_percent\\s*~\\s*racial_majority", code_no_comments)) {
    fail("The formula should be dem2p_percent ~ racial_majority --- the outcome on the left of the tilde, the explanatory variable on the right.")
  }
  if (!grepl("data\\s*=\\s*counties", code_no_comments)) {
    fail("Set data = counties.")
  }

  expected <- oneway.test(dem2p_percent ~ racial_majority, data = counties)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the oneway.test() output. If you're seeing an error, check for a missing tilde (~) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check that you are running oneway.test(dem2p_percent ~ racial_majority, data = counties).")
  }

  pass("Right! dem2p_percent is the outcome and racial_majority is the explanatory variable, so it tests whether the mean Democratic vote share differs across all four racial-majority categories at once.")
})
```

The p-value appears at the bottom of the output, labeled `p-value`. Find it, compare it to .05, and then ask what this test --- which treats all four racial majority categories simultaneously --- can and cannot tell you about the hypothesis.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value in your output and compare it to .05 before answering.

```{r quiz-dem2p-general-result, echo=FALSE}
question("The ANOVA for Democratic vote share by racial majority produces F = 382.25 with p < .001. Which null hypothesis does this result give us strong evidence to reject?",
  correct = "Right. Rejecting the ANOVA null tells us at least one group mean differs --- not which groups, not the direction, and not whether any specific comparison is significant.",
  answer("That the four racial-majority group means are all equal in the population",
         correct = TRUE),
  answer("That every racial-majority group mean differs significantly from every other group mean",
         message = "Rejecting the ANOVA null tells us that at least one population mean differs from at least one other --- not that every pairwise difference is statistically significant. A significant overall F is consistent with, say, three groups having similar means while one stands apart, with no significant difference among the three similar groups."),
  answer("That Black Majority counties specifically have a lower mean Democratic vote share than White Majority counties",
         message = "The ANOVA does not test any specific pairwise comparison. It asks whether all four group means could plausibly be equal, not whether any particular pair differs in a specific direction."),
  answer("That Democratic vote share is unrelated to any aspect of a county's demographic composition",
         message = "The ANOVA only tests whether the means differ across the four categories of racial_majority specifically. It says nothing about other demographic variables or about racial composition more broadly."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: What exactly does the null hypothesis of a one-way ANOVA state? Then ask what rejecting it tells you --- and what it does not."
)
```

### Testing the Directional Claim

Our hypothesis is specifically about Black Majority counties compared to White Majority counties --- not about comparing either group to Hispanic Majority or Other/No Majority counties, and not about the four categories generally. To test that directly, we filter the data down to just those two categories before running a two-sample t-test.

#### Steps 1 and 2: State the Hypotheses

You've seen the null and alternative formalized once. Before building this test, identify the correct `alternative` argument on your own.

```{r quiz-dem2p-alternative, echo=FALSE}
question("**racial_majority** is a plain character variable. Which value of `alternative` correctly tests the directional claim, and why?",
  correct = "Right. R sorts character variables alphabetically, putting Black Majority before White Majority --- so `alternative = \"greater\"` tests whether the first group's mean exceeds the second's, matching the predicted direction. With a plain character grouping variable, R sorts alphabetically. \"Black Majority\" sorts before \"White Majority,\" so Black is the first group and White is the second. The hypothesis predicts Black Majority counties have a higher mean Democratic vote share, so `alternative = \"greater\"` tests whether the first group's mean is greater than the second's --- the correct direction.",
  answer('`alternative = "greater"`, because R sorts character variables alphabetically and "Black Majority" comes before "White Majority"',
         correct = TRUE),
  answer('`alternative = "less"`, because R sorts character variables alphabetically and "Black Majority" comes before "White Majority"',
         message = 'The alphabetical ordering is right --- "Black Majority" does sort first. But the hypothesis predicts Black Majority counties have a *higher* mean Democratic vote share, not a lower one. `alternative = "less"` would test the wrong direction.'),
  answer('`alternative = "greater"`, because Black Majority counties have more Democratic voters and therefore go first in the formula',
         message = 'The substantive reasoning is backwards --- R does not order groups by the size of their means. For a plain character variable, R uses alphabetical ordering regardless of the group means.'),
  answer('`alternative = "two.sided"`, because we cannot know in advance which group R will treat as first',
         message = 'We can know which group comes first --- for a plain character variable, R sorts alphabetically. "Black Majority" sorts before "White Majority," making Black the first group. Since the hypothesis predicts a specific direction, a one-tailed test is appropriate.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: How does R order groups for a plain character variable? Which group comes first alphabetically, and does the hypothesis predict that group has a higher or lower mean?"
)
```

The null and alternative hypotheses follow from what you just identified:

$$H_0: \mu_{Black}=\mu_{White}$$
where $\mu$ is the mean Democratic vote share in the population.

$$H_A: \mu_{Black} > \mu_{White}$$

#### Step 3: Run the Test

The ANOVA established that the means are not all equal across the four racial majority categories. Now we test the specific directional claim the hypothesis makes: that Black Majority counties have a higher mean Democratic vote share than White Majority counties.

**Practice: Build the two-sample t-test**

Filter the data down to just the two categories the hypothesis compares, then run the t-test with the correct outcome, grouping variable, and direction. Three hints are available. Click Submit Answer to check your work.

```{r ttest-dem2p-black-white, exercise=TRUE, exercise.cap="Practice: Two-sample t-test for Democratic vote share by racial majority"}

```

```{r ttest-dem2p-black-white-hint-1}
# Hint 1 of 3
# Which two categories does the hypothesis specifically
# compare? You will need to narrow the data down to just
# those two categories before comparing their means.
```

```{r ttest-dem2p-black-white-hint-2}
# Hint 2 of 3
# The structure mirrors the urban/rural t-test in the previous
# example: filter() down to the two categories the hypothesis
# compares, then run t.test() with the outcome on the left of
# the tilde and the grouping variable on the right. Check the
# hypothesis to confirm which two categories to keep and which
# direction to test.
```

```{r ttest-dem2p-black-white-hint-3}
# Hint 3 of 3
# Filter to "Black Majority" and "White Majority" using %in%, then
# run t.test() with dem2p_percent as the outcome and racial_majority
# as the grouping variable. You identified the correct alternative
# in the question above.
```

```{r ttest-dem2p-black-white-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Use filter() to keep only Black Majority and White Majority counties.")
  }
  if (!grepl("racial_majority\\s*%in%\\s*c\\(", code_no_comments)) {
    fail('Filter using racial_majority %in% c("Black Majority", "White Majority").')
  }
  if (!grepl('["\']Black Majority["\']', code_no_comments)) {
    fail('Your filter should include "Black Majority".')
  }
  if (!grepl('["\']White Majority["\']', code_no_comments)) {
    fail('Your filter should include "White Majority".')
  }
  if (!grepl("t\\.test\\s*\\(", code_no_comments)) {
    fail("Use t.test() to run the two-sample t-test.")
  }
  if (!grepl("dem2p_percent\\s*~\\s*racial_majority", code_no_comments)) {
    fail("The formula should be dem2p_percent ~ racial_majority.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts the Black Majority mean is higher than the White Majority mean. Set alternative = "greater" to test that specific direction.')
  }

  expected_data <- counties %>%
    filter(racial_majority %in% c("Black Majority", "White Majority"))

  expected <- t.test(dem2p_percent ~ racial_majority,
                     data = expected_data,
                     alternative = "greater")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the t.test() output. If you're seeing an error, check for a missing tilde (~) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail('Check that you filtered to Black Majority and White Majority counties, then ran t.test(dem2p_percent ~ racial_majority, data = ..., alternative = "greater").')
  }

  pass("Right! Filtering to Black Majority and White Majority counties, then comparing their means with alternative = \"greater\", directly tests the hypothesis's specific comparison.")
})
```

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-dem2p-directional-result, echo=FALSE}
question("The t-test output reports a 95% confidence interval of (32.09, Inf) for the difference in means (Black Majority minus White Majority). What does this interval tell us?",
  correct = "Right. The calculated lower bound (32.09) excludes 0 --- the interval tells us the difference is at least 32 percentage points in the predicted direction.",
  answer("The mean Democratic vote share in Black Majority counties is at least 32.09 percentage points higher than in White Majority counties, with 95% confidence, and the upper bound is not estimated because the one-tailed test only constrains one side",
         correct = TRUE),
  answer("The true mean Democratic vote share in Black Majority counties falls between 32.09% and infinity",
         message = "The interval is for the difference in means between the two groups, not for the mean of Black Majority counties alone. A 95% confidence interval for the difference tells us the plausible range of values for μ_Black minus μ_White, not for μ_Black by itself."),
  answer("The interval is unreliable because one bound is Inf --- a valid confidence interval should have two finite bounds",
         message = "For a one-tailed test, R fixes one bound at the theoretical extreme rather than estimating it from the data. For a difference in means, that extreme is Inf (no finite upper limit exists for a mean difference). This is expected and correct, not a sign of an error."),
  answer("We cannot interpret this interval because it does not exclude 0",
         message = "The interval (32.09, Inf) excludes 0 entirely --- 0 falls below the lower bound of 32.09. An interval that excludes the null value of 0 points to the same conclusion as rejecting the null hypothesis, and this one does."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: For a one-tailed test, R fixes one bound at a theoretical extreme rather than estimating it. Which bound matters for the conclusion, and does the interval exclude 0?"
)
```

## Does Democracy Predict Economic Outcomes?

The democracy and GDP example from Tutorial 8 showed a clear pattern at the extremes --- Full Democracies averaged far higher GDP per capita than Authoritarian regimes --- but the middle categories complicated the picture. The argument is that democratic accountability gives leaders an incentive to pursue broadly beneficial economic policy, since they face electoral consequences for poor performance, while non-democratic leaders face fewer such constraints. Formally testing this hypothesis requires a choice: the hypothesis predicts a full ordering across all four categories, but a two-sample t-test can only compare two at a time. We test the sharpest implication of the hypothesis --- the endpoint contrast --- and note what that does and does not establish.

::: {.hypothesis}
In a comparison of [countries], those that [are more democratic] will have [higher GDP per capita] than will those that are [less democratic].
:::

The data frame **world** contains the ordinal variable **dem_level4**, which takes the values "Full Democ," "Part Democ," "Hybrid," and "Authoritarian," ordered from most to least democratic. It also contains the interval variable **gdppcap08**, GDP per capita in 2008 (World Bank), measured in 2008 US dollars.

As in the previous two examples, we start with the general claim that the population means differ across all four categories, then sharpen to the directional claim the hypothesis actually makes.

### Testing the General Claim

#### Step 1: State the Null Hypothesis

$$H_0: \mu_{Full Democ}=\mu_{Part Democ}=\mu_{Hybrid}=\mu_{Authoritarian}$$
where $\mu$ is the mean GDP per capita in the population.

#### Step 2: State the Alternative Hypothesis

$$H_A: \text{at least one population mean differs from the others}$$

#### Step 3: Run the Test

Tutorial 8 found that Full Democracies had a much higher mean GDP per capita than Authoritarian regimes, but the means table alone could not tell us whether a gap that large would be surprising by chance. Now we test whether the population means could plausibly all be equal across all four regime types.

**Practice: Build the one-way ANOVA**

Write the full `oneway.test()` call from scratch. Three hints are available. Click Submit Answer to check your work.

```{r anova-gdp-by-demlevel, exercise=TRUE, exercise.cap="Practice: One-way ANOVA for GDP per capita by democracy level"}

```

```{r anova-gdp-by-demlevel-hint-1}
# Hint 1 of 3
# This test treats all four regime types at once, asking
# whether the population means could plausibly all be equal.
# Which variable is the outcome you are averaging, and which
# variable defines the groups?
```

```{r anova-gdp-by-demlevel-hint-2}
# Hint 2 of 3
# Use oneway.test() with a formula: outcome ~ group, data = dataframe.
# The outcome goes on the left of the tilde, the grouping variable
# on the right. Both variables are in the world data frame.
```

```{r anova-gdp-by-demlevel-hint-3}
# Hint 3 of 3
# The outcome is gdppcap08 and the grouping variable is dem_level4,
# both from the world data frame. The function is oneway.test() with
# a formula: outcome ~ group, data = dataframe.
```

```{r anova-gdp-by-demlevel-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("oneway\\.test\\s*\\(", code_no_comments)) {
    fail("Use oneway.test() to run the one-way ANOVA.")
  }
  if (!grepl("gdppcap08\\s*~\\s*dem_level4", code_no_comments)) {
    fail("The formula should be gdppcap08 ~ dem_level4 --- the outcome on the left of the tilde, the explanatory variable on the right.")
  }
  if (!grepl("data\\s*=\\s*world", code_no_comments)) {
    fail("Set data = world.")
  }

  expected <- oneway.test(gdppcap08 ~ dem_level4, data = world)

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the oneway.test() output. If you're seeing an error, check for a missing tilde (~) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail("Check that you are running oneway.test(gdppcap08 ~ dem_level4, data = world).")
  }

  pass("Right! gdppcap08 is the outcome and dem_level4 is the explanatory variable, so it tests whether the mean GDP per capita differs across all four regime types at once.")
})
```

The p-value appears at the bottom of the output. Find it, compare it to .05, and ask what this test --- which considers all four regime types at once --- can and cannot tell you about the hypothesis.

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value in your output and compare it to .05 before answering.

```{r quiz-gdp-general-result, echo=FALSE}
question("The ANOVA for GDP per capita across four regime types produces F = 45.921 with p < .001. Which conclusion follows directly from this result?",
  correct = "Right. A significant ANOVA establishes that at least one mean differs from at least one other --- nothing more specific than that.",
  answer("At least one regime-type population mean differs from at least one other",
         correct = TRUE),
  answer("All four regime-type population means differ from one another",
         message = "A significant ANOVA tells you that at least one mean differs from at least one other, not that every pair of groups differs. The result is consistent with a single outlying group while three others remain similar."),
  answer("GDP per capita rises monotonically with democracy level",
         message = "A monotonic ordering across all four regime types is exactly the kind of specific directional claim the ANOVA p-value cannot evaluate. It tests whether all four means are equal --- it cannot tell us whether the means increase steadily from Authoritarian through Hybrid, Partial Democracy, to Full Democracy."),
  answer("Full Democracies have significantly higher GDP per capita than Authoritarian regimes",
         message = "This is a specific pairwise comparison. The ANOVA does not test individual pairs --- it only evaluates whether all four means could plausibly be equal. The endpoint comparison requires a separate targeted test, which is what comes next."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: A one-way ANOVA tests a single null hypothesis about all four group means simultaneously. What exactly does rejecting that null hypothesis tell you --- and what does it leave open?"
)
```

### Testing the Directional Claim

Our hypothesis makes a broad prediction about the ordering of all four democracy categories. A two-sample t-test cannot evaluate that complete ordering --- it compares exactly two groups. Here we test one focused implication of the hypothesis: whether countries at the most democratic end of the scale have higher mean GDP per capita than countries at the least democratic end. That is a particularly clear implication of the broader hypothesis and one we can test directly with the two-sample t-test.

#### Steps 1 and 2: State the Hypotheses

This time, identify the complete null and alternative hypothesis pair before building the test.

```{r quiz-gdp-hypotheses, echo=FALSE}
question("The Run and observe above confirmed that R puts Full Democ before Authoritarian. Which pair of hypotheses correctly formalizes the directional claim?",
  correct = "Right. For the tests in this tutorial, the null states that the population means are equal; the alternative matches the predicted direction given which group comes first.",
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} > \\mu_{Authoritarian}$",
         correct = TRUE),
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} < \\mu_{Authoritarian}$",
         message = "The null is right, but the alternative points in the wrong direction. The hypothesis predicts Full Democracies have *higher* GDP per capita than Authoritarian regimes, so the alternative should use $>$, not $<$."),
  answer("$H_0: \\mu_{Full Democ} = \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} \\neq \\mu_{Authoritarian}$",
         message = "The null is right, but a two-sided alternative does not capture the directional prediction the hypothesis makes. The hypothesis specifically predicts Full Democracies are richer --- that is a one-tailed claim."),
  answer("$H_0: \\mu_{Full Democ} > \\mu_{Authoritarian}$ and $H_A: \\mu_{Full Democ} = \\mu_{Authoritarian}$",
         message = "The null and alternative are reversed. Here, the null is the no-difference baseline --- equal population means --- and the alternative is the directional claim we are testing. A null hypothesis never contains an inequality."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: For the tests in this tutorial, what does the null hypothesis say about the two population means? Then check which direction the hypothesis predicts, given that Full Democ comes first."
)
```

The null and alternative hypotheses follow from what you just identified:

$$H_0: \mu_{Full Democ}=\mu_{Authoritarian}$$
where $\mu$ is the mean GDP per capita in the population.

$$H_A: \mu_{Full Democ} > \mu_{Authoritarian}$$

#### Step 3: Run the Test

The ANOVA established that the means are not all equal across the four regime types. The hypothesis already tells us the direction to test --- Full Democracies should have higher GDP per capita than Authoritarian regimes. Before running the t-test, let's confirm which group R will treat as the first group in the formula, so we can translate that directional prediction into the correct `alternative` argument.

**Run and observe: Check the group order**

```{r gdp-group-means, exercise=TRUE, exercise.cap="Run and observe: Group means for Full Democ and Authoritarian countries"}
world_extremes <- world %>%
  filter(dem_level4 %in% c("Full Democ", "Authoritarian"))

world_extremes %>%
  group_by(dem_level4) %>%
  summarise(mean_gdp = mean(gdppcap08, na.rm = TRUE))
```

The output lists the groups in the order R will use them in the formula: **dem_level4** is a factor, and its level order puts "Full Democ" before "Authoritarian." So the difference R computes is Full Democ minus Authoritarian. The hypothesis predicts that difference is positive --- meaning `alternative = "greater"` is the correct choice. Now we run the test.

**Practice: Build the two-sample t-test**

Filter the data to the two categories at opposite ends of the democracy scale, then run the t-test with the correct outcome, grouping variable, and direction. Three hints are available. Click Submit Answer to check your work.

```{r ttest-gdp-fulldemoc-authoritarian, exercise=TRUE, exercise.cap="Practice: Two-sample t-test for GDP per capita by democracy level"}

```

```{r ttest-gdp-fulldemoc-authoritarian-hint-1}
# Hint 1 of 3
# Which two categories does the hypothesis compare most
# directly? You will need to narrow the data down to just
# those two categories before comparing their means.
```

```{r ttest-gdp-fulldemoc-authoritarian-hint-2}
# Hint 2 of 3
# Use filter() with %in% to keep only the two categories, then
# run t.test(outcome ~ group, data = filtered_data, alternative = ...).
# The outcome is gdppcap08 and the grouping variable is dem_level4.
# The Run and observe above shows which group R treats as first ---
# use the hypothesis to determine whether alternative = "less" or "greater".
```

```{r ttest-gdp-fulldemoc-authoritarian-hint-3}
# Hint 3 of 3
# Filter to "Full Democ" and "Authoritarian" using %in%, then run
# t.test() with gdppcap08 as the outcome and dem_level4 as the
# grouping variable. The hypothesis pair you just identified tells
# you the correct alternative.
```

```{r ttest-gdp-fulldemoc-authoritarian-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("filter\\s*\\(", code_no_comments)) {
    fail("Use filter() to keep only Full Democ and Authoritarian countries.")
  }
  if (!grepl("dem_level4\\s*%in%\\s*c\\(", code_no_comments)) {
    fail('Filter using dem_level4 %in% c("Full Democ", "Authoritarian").')
  }
  if (!grepl('["\']Full Democ["\']', code_no_comments)) {
    fail('Your filter should include "Full Democ".')
  }
  if (!grepl('["\']Authoritarian["\']', code_no_comments)) {
    fail('Your filter should include "Authoritarian".')
  }
  if (!grepl("t\\.test\\s*\\(", code_no_comments)) {
    fail("Use t.test() to run the two-sample t-test.")
  }
  if (!grepl("gdppcap08\\s*~\\s*dem_level4", code_no_comments)) {
    fail("The formula should be gdppcap08 ~ dem_level4.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts the Full Democ mean is higher than the Authoritarian mean. Set alternative = "greater" to test that specific direction.')
  }

  expected_data <- world %>%
    filter(dem_level4 %in% c("Full Democ", "Authoritarian"))

  expected <- t.test(gdppcap08 ~ dem_level4,
                     data = expected_data,
                     alternative = "greater")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the t.test() output. If you're seeing an error, check for a missing tilde (~) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail('Check that you filtered to Full Democ and Authoritarian countries, then ran t.test(gdppcap08 ~ dem_level4, data = ..., alternative = "greater").')
  }

  pass("Right! Filtering to Full Democ and Authoritarian countries, then comparing their means with alternative = \"greater\", directly tests the hypothesis's specific comparison.")
})
```

#### Steps 4 and 5: Find the P-value, Decide, and Report the Result

**Practice: Evaluate the result**

Find the p-value and the sample estimates in your output before answering.

```{r quiz-gdp-directional-result, echo=FALSE}
question("The t-test result is statistically significant (t = 7.0635, p < .001), with Full Democracies averaging $35,891 in GDP per capita compared to $9,412 for Authoritarian regimes. Which conclusion is best supported by this result combined with the ANOVA?",
  correct = "Right. The ANOVA and t-test together support one focused implication of the broader hypothesis --- not the full four-category ordering prediction.",
  answer("Countries at the most democratic end of the scale have significantly higher mean GDP per capita than those at the least democratic end, supporting a focused implication of the broader hypothesis",
         correct = TRUE),
  answer("More democratic countries have significantly higher GDP per capita than less democratic countries, fully establishing the hypothesis",
         message = "The t-test compared only the two endpoint categories --- Full Democracies and Authoritarian regimes. That result supports one focused implication of the hypothesis, but it does not establish the complete ordering across all four regime types. For example, it is consistent with Hybrid regimes having higher GDP than Partial Democracies, which would contradict the overall ordering the hypothesis predicts."),
  answer("The result is only meaningful if all four regime-type comparisons are individually significant",
         message = "Testing every pairwise comparison is not required to learn something meaningful from the hypothesis. The ANOVA established that not all group means are equal; the targeted t-test evaluated the clearest endpoint implication of the hypothesis. Together they provide evidence for that focused implication, while leaving the complete ordering of all four categories unresolved."),
  answer("The result only establishes that Full Democracies and Authoritarian regimes differ --- the hypothesis is fully tested only once all four regime types are compared pairwise",
         message = "The result does establish specifically that Full Democracies have higher mean GDP than Authoritarian regimes --- that is its strength. Whether this fully tests the hypothesis depends on how specifically the hypothesis was stated. Our hypothesis made a broad ordering prediction; we tested its sharpest focused implication. That is a principled and informative test, not an incomplete one."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: The hypothesis predicts a full ordering across all four categories. The t-test compared only two. What does the significant result establish, and what does it leave open?"
)
```

## Testing a Single Mean

Every test in this tutorial so far has asked whether a categorical and an interval variable are related, or whether two groups differ. A different kind of question asks whether a single sample mean differs from some fixed, hypothesized value --- no second variable involved at all. For example, we might ask whether the average county-level Democratic two-party vote share in our data --- the same **dem2p_percent** variable from the second example --- differs from 50%. A value of 50% represents an evenly split two-party vote; testing against it tells us whether the typical county leaned Democratic or Republican on average in 2016.

We use the same `t.test()` function, but with a different call structure than the formula notation used in the two-sample tests above. Instead of `outcome ~ group`, we supply the variable directly using `data$variable` notation, and use `mu` to specify the hypothesized population mean we are comparing against. There is no `data =` argument here --- the variable is passed in directly.

**Run and observe: Run a one-sample t-test**

```{r ttest-onesample-dem2p, exercise=TRUE, exercise.cap="Run and observe: One-sample t-test for mean Democratic vote share"}
t.test(counties$dem2p_percent, mu = 50)
```

The logic of the five steps still applies here: a null hypothesis ($H_0: \mu=50$), an alternative ($H_A: \mu \ne 50$), a test statistic, a p-value, and a decision --- but with only one sample mean involved, not a comparison between two groups or a relationship between two variables.

The sample mean is 33.34%, well below the hypothesized value of 50% (t = -55.527, df = 3111, p < .001), so we reject the null hypothesis: the average county-level Democratic vote share differs significantly from an even two-party split. The average county in 2016 leaned Republican, not Democratic. That conclusion makes sense given the geography of American politics: Democratic support is concentrated in a relatively small number of densely populated urban counties, while the majority of the nation's 3,112 counties are smaller and more Republican-leaning.

The output's 95% confidence interval, (32.77, 33.91), gives the same answer a different way: since it excludes the hypothesized value of 50 entirely, it points to the same conclusion as the p-value --- and because this test is two-tailed, both bounds here are calculated from the data rather than fixed at a theoretical extreme.

## Paired t-test

The two-sample t-tests in this tutorial compared different groups of cases --- urban counties against rural counties, Full Democracies against Authoritarian regimes, and so on. A [**paired t-test**]{.important-text} asks a different kind of question: when the *same* cases are measured twice, does the value change, on average, between the two measurements?

The data frame **world** contains **gini08** and **gini10**, a country's Gini coefficient --- a standard measure of income inequality, where 0 represents perfect equality and 100 represents perfect inequality --- in 2008 and 2010. We might ask whether income inequality changed, on average, across those two years.

The logic is more straightforward than it might look: for each country, subtract its 2008 Gini coefficient from its 2010 Gini coefficient, giving one difference score per country. Then take the mean of those difference scores across all countries. That mean difference --- not the mean of **gini10** alone, and not the mean of **gini08** alone --- is what the test evaluates. A small illustration with three hypothetical countries makes this concrete:

| Country | gini08 | gini10 | Difference (gini10 $-$ gini08) |
|:--|:--:|:--:|:--:|
| Country A | 35.2 | 36.0 | 0.8 |
| Country B | 28.4 | 27.1 | -1.3 |
| Country C | 41.6 | 42.5 | 0.9 |

For these three countries, the mean difference is $(0.8 + (-1.3) + 0.9)/3 = 0.13$. A paired t-test asks whether a mean difference like this is far enough from zero to be surprising under the null hypothesis --- the same logic as every other test in this tutorial, just applied to difference scores instead of raw values.

$$H_0: \mu_{gini10}-\mu_{gini08}=0$$
where $\mu_{gini10}$ and $\mu_{gini08}$ are the mean Gini coefficients in the population in 2010 and 2008, respectively.
$$H_A: \mu_{gini10}-\mu_{gini08} \ne 0$$

We use the same `t.test()` function again, but like the one-sample test above, the paired t-test uses a different call structure than the formula notation used in the two-sample tests. Instead of `outcome ~ group`, we supply both variables directly --- `t.test(x, y, paired = TRUE)` --- and set `paired = TRUE` to tell R to compute the difference for each country first, exactly as in the table above, and then run a test on the mean of those differences, rather than treating **gini10** and **gini08** as two separate, unrelated groups of countries.

**Run and observe: Run a paired t-test**

```{r ttest-paired-gini, exercise=TRUE, exercise.cap="Run and observe: Paired t-test for Gini coefficients 2008 vs. 2010"}
t.test(world$gini10, world$gini08, paired = TRUE)
```

Notice that the output reports a single mean difference, not two separate group means the way an unpaired t-test would --- a direct reflection of the logic above. Here, the mean difference is $-0.54$ (t = -1.495, df = 120, p = .1375). Because the p-value is well above .05, we fail to reject the null hypothesis: we do not have evidence that the mean Gini coefficient changed between 2008 and 2010.

The 95% confidence interval, (-1.254, 0.175), tells the same story. In the earlier tests that rejected the null, the confidence interval excluded the relevant null value --- usually 0 for a difference in means, and 50 for the one-sample test above. Here, 0 falls comfortably inside the interval, right alongside a wide range of other plausible values. That is exactly what a confidence interval looks like when a test fails to reject: not evidence that there is no difference, but evidence that we cannot rule out 0, along with a great many other values, as the true mean difference.

## Check Your Understanding

The exercises below give you a new hypothesis, new variables, and a new data frame --- but the same test structure you used throughout this tutorial. The first exercise asks you to build a two-sample t-test from scratch on a fresh scenario. The remaining questions test whether you have internalized the key ideas: when filtering is needed before a t-test, what a significant ANOVA result can and cannot tell you, and how to translate a hypothesis into the correct `alternative` argument given R's group ordering rules.

State supreme courts use several different methods to select judges. Under merit selection (sometimes called the "Missouri Plan"), a nonpartisan commission nominates candidates and the governor appoints from that list; judges then face periodic retention elections rather than contested races. Under partisan election, judges run for office on a party ticket, the same way many other elected officials do. Because partisan election ties a judge's seat directly to electoral performance, while merit selection insulates judges from that pressure, political scientists have hypothesized that merit-selected judges serve longer, more stable tenures.

::: {.hypothesis}
In a comparison of [state supreme court justices], those selected through [merit selection] will serve [longer average tenures] than will those selected through [partisan election].
:::

Suppose a researcher collected data on 50 state supreme court justices, stored in a data frame called **judges_df**: the nominal variable **selection_method**, which takes the values "Merit Selection" and "Partisan Election," and the interval variable **years_served**, the number of years each justice served on the bench.

```{r cyu-prepare-judges, include=FALSE}
set.seed(13)
judges_df <- data.frame(
  selection_method = rep(c("Merit Selection", "Partisan Election"), each = 25),
  years_served = c(rnorm(25, mean = 14, sd = 4),
                    rnorm(25, mean = 9, sd = 3.5))
)
```

The scenario above gives you a new variable, a new data frame, and a directional hypothesis. The structure of the test is identical to the examples in this tutorial --- the only thing that changes is which variables and data frame you supply.

**Apply: Build a two-sample t-test from a fresh scenario**

Use **years_served**, **selection_method**, and the **judges_df** data frame. Two hints are available. Click Submit Answer to check your work.

```{r cyu-ttest-judges, exercise=TRUE, exercise.setup="cyu-prepare-judges", exercise.cap="Apply: Two-sample t-test for judicial tenure by selection method"}

```

```{r cyu-ttest-judges-hint-1}
# Hint 1 of 2
# Use t.test() with a formula: outcome ~ group, data = dataframe,
# alternative = ... . Which variable is the outcome, and which
# defines the two groups? The hypothesis predicts a specific
# direction --- which group should have the higher mean?
```

```{r cyu-ttest-judges-hint-2}
# Hint 2 of 2
# "Merit Selection" sorts before "Partisan Election" alphabetically,
# so Merit Selection is the first group in the formula. The hypothesis
# predicts that group has a higher mean --- which value of alternative
# tests whether the first group's mean is greater than the second's?
```

```{r cyu-ttest-judges-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  code_no_comments <- gsub("#.*", "", code)

  if (!grepl("t\\.test\\s*\\(", code_no_comments)) {
    fail("Use t.test() to run the two-sample t-test.")
  }
  if (!grepl("years_served\\s*~\\s*selection_method", code_no_comments)) {
    fail("The formula should be years_served ~ selection_method --- the outcome on the left of the tilde, the explanatory variable on the right.")
  }
  if (!grepl("data\\s*=\\s*judges_df", code_no_comments)) {
    fail("Set data = judges_df.")
  }
  if (!grepl("alternative\\s*=\\s*['\"]greater['\"]", code_no_comments)) {
    fail('The hypothesis predicts merit-selected justices serve longer tenures than partisan-elected justices. Since "Merit Selection" sorts before "Partisan Election" alphabetically, set alternative = "greater" to test that direction.')
  }

  expected <- t.test(years_served ~ selection_method,
                     data = judges_df,
                     alternative = "greater")

  if (is.null(.result)) {
    fail("Make sure your code produces a result the grader can check.")
  }
  if (!inherits(.result, "htest")) {
    fail("Your code should return the t.test() output. If you're seeing an error, check for a missing tilde (~) in the formula, a misspelled variable name, or a missing data argument.")
  }
  if (!isTRUE(all.equal(as.numeric(.result$p.value),
                        as.numeric(expected$p.value),
                        tolerance = 1e-6))) {
    fail('Check that you are running t.test(years_served ~ selection_method, data = judges_df, alternative = "greater").')
  }

  pass("Right! Merit-selected justices serve substantially longer average tenures (14.9 years) than partisan-elected justices (8.0 years), a difference that is statistically significant (t = 6.81, p < .001) and matches the direction the hypothesis predicts.")
})
```

**Apply: Recognize when filtering is needed**

```{r cyu-recognize-filtering, echo=FALSE}
question("Suppose a hypothesis predicts that countries with proportional representation electoral systems have higher voter turnout than countries with majoritarian electoral systems, and electoral system is measured with a 4-category variable: proportional representation, mixed/semi-proportional, majoritarian, and other. To test this hypothesis with a two-sample t-test, what would you need to do first?",
  correct = "Right. t.test() requires exactly two groups --- with four categories, filter down to the two the hypothesis names before running it. Use filter() with %in% to keep only the two categories the hypothesis compares, then run t.test() on that filtered data --- the same approach used in every two-sample t-test in this tutorial.",
  answer("Filter the data down to just the proportional representation and majoritarian categories, dropping the other two, before running t.test()",
         correct = TRUE),
  answer("Run t.test() directly on all four categories at once",
         message = "t.test()'s formula method requires exactly two groups. With four categories, R will return an error rather than a result --- you have to narrow the data down to the two categories the hypothesis compares first."),
  answer("Combine all four categories into two groups --- proportional representation and mixed systems in one, majoritarian and other in another --- before running t.test()",
         message = "Combining categories is sometimes necessary, but the grouping has to follow the hypothesis, not convenience. The hypothesis specifically compares proportional representation with majoritarian systems --- mixed and other are not part of either comparison group. Adding them in would change the question being asked."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: t.test() compares exactly two groups. How many groups does this 4-category variable give you? What do you need to do first?"
)
```

**Apply: Choose the right test**

```{r cyu-general-vs-directional, echo=FALSE}
question("A researcher measures political trust in the same 500 survey respondents before and after a national election, and wants to know whether trust changed, on average, between the two measurements. Which test is most appropriate?",
  correct = "Right. The same cases are measured twice --- that is exactly when a paired t-test is appropriate.",
  answer("A paired t-test, because the same respondents are measured at both time points",
         correct = TRUE),
  answer("A two-sample t-test, because there are two groups of measurements --- before and after",
         message = "The before and after measurements are not two independent groups --- they are two measurements on the same 500 people. A two-sample t-test assumes the two groups are independent, which is not the case here. When the same cases are measured twice, a paired t-test is the appropriate tool."),
  answer("A one-way ANOVA, because the researcher wants to compare means across two time points",
         message = "ANOVA is used when comparing means across categories of a grouping variable across independent groups. Here the same respondents are measured twice, which calls for a paired t-test rather than an omnibus comparison across groups."),
  answer("A one-sample t-test, because the question is about a single mean --- the mean change in trust",
         message = "A one-sample t-test compares a single sample mean to a fixed hypothesized value, with no second measurement involved. While the paired t-test does ultimately test whether a mean difference differs from zero, it works by first computing a difference score for each respondent and then testing that. The one-sample t-test cannot account for the paired structure of the data."),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: Are the two groups of measurements independent of each other, or do they come from the same cases? That distinction determines which test is appropriate."
)
```

**Apply: Predict the right direction for a new hypothesis**

```{r cyu-direction-prediction, echo=FALSE}
question('A hypothesis predicts that countries with high judicial independence have lower rates of corruption than countries with low judicial independence. The grouping variable judicial_independence takes the values "High" and "Low". If you ran t.test(corruption ~ judicial_independence, data = ..., alternative = ...), which value of alternative would test this hypothesis directly?',
  correct = "Right. High sorts before Low alphabetically, so the High group comes first in the formula --- and the hypothesis predicts it has the lower mean. R sorts character variables alphabetically: \"High\" before \"Low\". With High listed first, alternative = \"less\" tests whether the first group's mean is lower than the second's --- exactly matching the predicted direction.",
  answer('alternative = "less"',
         correct = TRUE),
  answer('alternative = "greater"',
         message = 'With "High" sorted first, alternative = "greater" would test whether the High group\'s mean is higher than the Low group\'s --- the opposite of what this hypothesis predicts.'),
  answer('alternative = "two.sided", since we cannot know the direction in advance',
         message = "We can know the direction in advance --- it is what the hypothesis predicts. A two-sided test would be appropriate if we had no directional prediction, but this hypothesis makes one explicitly."),
  answer('It depends on whether judicial_independence is stored as a factor or a plain character variable',
         message = 'For a plain character variable, R always sorts alphabetically. For a factor, R uses the stored factor level order. Since the question does not specify the storage class, you need to check --- but the alphabetical default for a character variable is "High" before "Low", and alternative = "less" would test the predicted direction. Always check how a variable is stored before choosing alternative.'),
  allow_retry = TRUE,
  random_answer_order = TRUE,
  try_again = "Hint: How does R order groups for a character variable? Which value comes first alphabetically --- High or Low? The hypothesis predicts that first group has a lower mean. Which alternative matches?"
)
```

## The Takeaways

This tutorial completed the same general-to-directional pattern Tutorial 12 used for two categorical variables, now applied to a categorical and an interval variable. A one-way ANOVA, run with `oneway.test(outcome ~ group, data = df)`, treats every category of the explanatory variable at once and asks whether the population means could plausibly all be equal. By default, R does not assume the population variances are equal across groups --- the same default stance it takes for the two-sample t-test. Because ANOVA considers every category simultaneously, it has no way to encode a specific, directional prediction; in this tutorial, we sharpened to a two-sample t-test, run with `t.test(outcome ~ group, data = df, alternative = ...)`, which compares exactly two groups and can test a specific direction with `alternative = "less"` or `"greater"`. A planned directional t-test can also stand on its own when a hypothesis specifies the comparison from the start --- the two-step pattern here is a teaching structure, not a requirement.

When the categorical variable has more than two categories, a directional t-test cannot run on the whole variable at once --- `t.test()`'s formula method requires exactly two groups. The fix throughout this tutorial was to use `filter()` with the `%in%` operator to narrow the data down to the two specific categories a hypothesis names, before running the t-test on that filtered data. This also matters for choosing `alternative`: R determines which group's mean comes first by the stored values of the grouping variable --- numerically for numeric codes, alphabetically for plain character variables --- not by the order you type the values inside `filter()`. Always check which group comes first before choosing `alternative = "less"` or `"greater"`.

We also looked at two related but different questions. A one-sample t-test, using `t.test(x, mu = ...)`, asks whether a single sample mean differs from a fixed, hypothesized value, with no second variable or comparison group involved at all. A paired t-test, using `t.test(x, y, paired = TRUE)`, asks whether the average difference between two variables measured on the same cases --- rather than two different groups of cases --- differs from zero. Both use a different call structure than formula notation: variables are supplied directly rather than as `outcome ~ group`. Both follow the same five-step logic as every other test in this tutorial.

That decision is worth being precise about. Rejecting a null hypothesis --- even with a very small p-value --- never proves a hypothesis is true, and failing to reject a null hypothesis never proves it is false. A p-value tells us how surprising our result would be if the null hypothesis were exactly correct; a small p-value is [**statistically significant evidence in support of a hypothesis**]{.important-text}, not certainty that it is right. That is not a small distinction. It is the most a hypothesis test can honestly tell us.

Tutorial 14 completes the sequence by applying the same logic to two interval variables. The tool changes again --- a correlation test instead of ANOVA and t-tests --- but the procedure is identical, and by now you should find the five steps feel familiar.

## Submit Your Work

When you have completed all exercises in this tutorial, generate your submission code using the button below. Copy the full code that appears and paste it into the Canvas submission box for this tutorial. The code can be long --- make sure you have copied all of it before submitting.

```{r encode-logic-13, context="server"}
learnrhash::encoder_logic()
```

```{r encode-ui-13, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    shiny::p("Click the button below to generate your submission code.
              Copy the full code and paste it into Canvas.")
  )
)
```

```{r tutorial-success-banner-13, results="asis", echo=FALSE}
cat('
<p class="tutorial-complete"
   role="status"
   aria-label="Excellent work! You have completed this tutorial and can now test relationships between categorical and interval variables using ANOVA and t-tests, and you know exactly what those tests can, and cannot, tell you.">
  <span aria-hidden="true">🎉</span>
  Excellent work! You can now test relationships between categorical and interval variables using ANOVA and t-tests --- and you know exactly what those tests can, and cannot, tell you.
</p>
')
```
